---
title: "Analysis & Visualization with iQualityR.doe"
author: "iQualityR Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analysis & Visualization with iQualityR.doe}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  fig.align = "center"
)
```

## 1. Introduction

Once an experimental design is generated and response data collected, the
analysis phase answers: **Which factors matter? How large are the effects?
Is the model adequate?** `iQualityR.doe` provides an integrated analysis
and visualization toolkit covering ANOVA, effect estimation, model
diagnostics, model selection, VIF, uncoded equations, fold-over
augmentation, botched-run handling, and 16 plot types.

This vignette uses a **welding-strength case study** to demonstrate the
full analysis workflow.

```{r setup}
library(iQualityR.doe)
set.seed(42)
```

## 2. Case Study: Welding Process Optimization

### Problem

A manufacturing engineer investigates 4 factors affecting weld shear
strength: Current (A), Voltage (B), Gas Flow (C), and Travel Speed (D).
A $2^4$ full factorial with 3 center points was run.

```{r data}
factors <- lapply(LETTERS[1:4], function(nm) {
  list(name = nm, type = "continuous", levels = c(-1, 1))
})

plan <- IqrDoePlan$new(
  task_tag      = "welding",
  design_type   = "factorial",
  factors       = factors,
  center_points = 3,
  randomize     = FALSE,
  response_vars = "Strength"
)

task <- IqrDoeTask$new(plan = plan, theme = "academic")
analyzer <- DoeAnalyzer$new()
results <- analyzer$run(plan = plan)
design <- results$design_info

# Simulate response: A and C are active, with a mild AC interaction.
design$Strength <- 400 + 25 * design$A - 15 * design$C +
  10 * design$A * design$C +
  rnorm(nrow(design), 0, 3)

task <- IqrDoeTask$new(plan = plan, data = design, theme = "academic")
task$compute()
```

## 3. Model Summary & ANOVA

### 3.1 Summary output

```{r summary}
task$summary()
```

The summary displays the design metadata, ANOVA table, coefficient
estimates, and model fit statistics ($R^2$, adjusted $R^2$).

### 3.2 Accessing results programmatically

All results are stored in `task$results`:

```{r results-structure}
names(task$results)
```

## 4. Effect Estimation & Unreplicated Designs

### 4.1 Lenth's PSE for unreplicated factorials

When no replication exists, standard ANOVA cannot estimate error.
Lenth's (1989) PSE method identifies active effects. The half-normal and
Pareto plots (below) internally compute Lenth's PSE and draw reference
lines. For a direct computation, extract the effects from the model:

```{r lenth}
# Fit a model with all main effects and two-factor interactions.
model_full <- lm(Strength ~ (A + B + C + D)^2, data = design)
beta <- coef(model_full)
effects <- 2 * beta[names(beta) != "(Intercept)"]
analyzer <- DoeAnalyzer$new()
lenth_out <- analyzer$compute_lenth_pse(effects)
lenth_out$pse
lenth_out$active_me
```

### 4.2 Half-normal probability plot

The half-normal plot uses Lenth's PSE internally to draw ME/SME
reference lines. It requires effects from the task's results:

```{r half-normal}
# The half-normal plot is rendered by the plotter from effects stored
# in the task results. Use the full model for effect extraction.
print(task$plot(type = "half_normal"))
```

### 4.3 Normal effects plot (signed)

```{r normal-effects}
print(task$plot(type = "normal_effects"))
```

### 4.4 Pareto chart of effects

```{r pareto}
print(task$plot(type = "pareto_effects"))
```

### Decision

The half-normal and Pareto plots confirm that factors **A** (Current)
and **C** (Gas Flow), along with the **AC** interaction, are the active
effects. All other effects fall below the Lenth ME reference line.

## 5. Model Diagnostics

### 5.1 Residual diagnostics (3-panel)

```{r residual}
print(task$plot(type = "residual"))
```

### 5.2 Residuals vs predictors

To check whether residuals exhibit curvature against individual
predictors:

```{r resid-vs-pred}
print(task$plot(type = "residuals_vs_predictors"))
```

### Decision

If the Residuals-vs-Fitted panel shows a funnel (heteroscedasticity) or
the Normal Q-Q deviates at the tails, a transformation (e.g. Box-Cox)
or additional quadratic terms should be considered.

## 6. Main Effects & Interaction Plots

### 6.1 Main effects plot

```{r main-effects}
print(task$plot(type = "main_effects"))
```

### 6.2 Interaction plot

```{r interaction}
print(task$plot(type = "interaction"))
```

### 6.3 Full diagnostic panel

```{r full}
print(task$plot(type = "full"))
```

## 7. Model Adequacy Metrics

### 7.1 R-squared (predicted) and PRESS

`compute_model_fit_extended()` returns PRESS, $R^2_{pred}$, VIF, and a
multicollinearity flag:

```{r model-fit}
fit_ext <- analyzer$compute_model_fit_extended(model_full)
fit_ext$PRESS
fit_ext$r_squared_pred
fit_ext$max_vif
```

### 7.2 Variance Inflation Factors (VIF)

Standalone VIF computation matches Minitab's Coefficients table VIF column:

```{r vif}
analyzer$compute_vif(model_full)
```

VIF > 10 flags harmful multicollinearity. In this orthogonal design, all
VIFs are 1.

### 7.3 Curvature test

For designs with center points, the curvature test checks for
quadratic curvature:

```{r curvature}
curv <- analyzer$test_curvature(
  model       = model_full,
  data        = design,
  response_var = "Strength",
  factor_vars  = c("A", "B", "C", "D")
)
curv$p_value
curv$interpretation
```

## 8. Uncoded Coefficient Equation

By default, coefficients are in coded units ($\pm 1$). To obtain the
regression equation in **actual engineering units**:

```{r uncoded}
factors_actual <- list(
  list(name = "A", type = "continuous", levels = c(180, 220)),  # Current (A)
  list(name = "B", type = "continuous", levels = c(20, 30)),     # Voltage (V)
  list(name = "C", type = "continuous", levels = c(15, 25)),     # Gas Flow (L/min)
  list(name = "D", type = "continuous", levels = c(5, 15))       # Travel Speed (mm/s)
)
uncoded <- analyzer$get_uncoded_equation(model_full, factors_actual)
cat(uncoded$equation, "\n")
```

### Decision

The uncoded equation allows the engineer to predict weld strength directly
from the machine settings (Current in Amps, Gas Flow in L/min) without
converting to coded units.

## 9. Model Selection

When the full model contains inert terms, `select_model()` performs
stepwise / AICc / BIC selection:

```{r select-model}
selected <- analyzer$select_model(model_full, direction = "backward",
                                  criterion = "aicc")
selected$n_terms_initial
selected$n_terms_final
```

### Decision

Backward elimination using AICc removes inert terms, yielding a parsimonious
model with fewer parameters and improved prediction accuracy.

## 10. Design Augmentation

### 10.1 Fold-over for de-aliasing

When a fractional factorial reveals aliased active effects, a fold-over
adds mirror runs to break the aliasing:

```{r foldover}
frac_task <- screening_design(
  factors = lapply(LETTERS[1:5], function(nm) {
    list(name = nm, type = "continuous", levels = c(-1, 1))
  }),
  resolution = "IV",
  seed = 1
)
frac_design <- frac_task$results$design_info[, 1:5]

folded <- analyzer$fold_over(
  design    = frac_design,
  factors   = lapply(LETTERS[1:5], function(nm)
    list(name = nm, type = "continuous", levels = c(-1, 1))),
  fold_factor = NULL   # full fold-over
)
nrow(folded)
table(folded$Foldover)
```

### 10.2 Augment factorial to CCD

To fit a second-order model, augment the factorial with axial points:

```{r augment-ccd}
augmented <- analyzer$augment_to_ccd(
  design      = design[, 1:4],
  factors     = factors_actual,
  alpha_type  = "rotatable",
  n_center_points = 4
)
nrow(augmented)
table(augmented$PointType)
```

### Decision

Fold-over doubles the 16-run fractional to 32 runs, fully de-aliasing all
effects. Augmenting to CCD adds axial points enabling response-surface
analysis (see the *RSM & Optimization* vignette).

## 11. Botched Runs Handling

In practice, planned levels sometimes deviate. The `handle_botched_runs()`
method replaces planned levels with actual measured levels and re-fits the
model:

```{r botched}
# Simulate: actual A for run 3 was 185 instead of 180 (coded -1).
actual <- design[, 1:4]
actual$A[3] <- -0.75   # Deviates from planned -1

botched_result <- analyzer$handle_botched_runs(
  design       = design,
  actual_data  = cbind(actual, Strength = design$Strength),
  factor_names = c("A", "B", "C", "D"),
  response_name = "Strength"
)
botched_result$changed_cols
botched_result$design_corrected$A[3]
```

### Decision

The corrected design reflects what truly happened on the shop floor. The
re-fitted model uses actual levels, ensuring the analysis is valid.

## 12. Power & Sample Size

Before running an experiment, verify the design has adequate power:

```{r power}
power_out <- analyzer$compute_power(
  n_factors     = 4,
  n_replicates  = 1,
  effect_size   = 15,
  sigma         = 5,
  n_center_points = 3
)
power_out$power
power_out$interpretation
```

### Power curve visualization

```{r power-curve}
curve_df <- analyzer$plot_power_curve(
  n_factors     = 4,
  n_replicates  = 1,
  sigma         = 5,
  n_points      = 20,
  max_effect    = 20
)
# Render as a ggplot.
library(ggplot2)
ggplot(curve_df, aes(Effect_Size, Power)) +
  geom_line(color = "#2C7BB6", linewidth = 1) +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "gray50") +
  labs(title = "Power Curve (4 factors, sigma=5)",
       x = "Effect Size", y = "Power") +
  theme_minimal()
```

### Sample size calculation

```{r sample-size}
ss <- analyzer$compute_sample_size(
  n_factors    = 4,
  effect_size  = 15,
  sigma        = 5,
  target_power = 0.90
)
ss$n_replicates
ss$achieved_power
```

## 13. Reporting

### 13.1 Export to Excel

```{r report-excel, eval=FALSE}
tmp <- tempfile(fileext = ".xlsx")
reporter <- DoeReporter$new()
reporter$output(task$results, plan, format = "excel", path = tmp)
```

### 13.2 Export to Word / PowerPoint

When the `officer` package is available:

```{r report-office, eval=FALSE}
# Word
reporter$output(task$results, plan, format = "word",
                path = tempfile(fileext = ".docx"))
# PowerPoint
reporter$output(task$results, plan, format = "powerpoint",
                path = tempfile(fileext = ".pptx"))
```

## 14. Conclusion

This vignette demonstrated the complete analysis workflow:

* **Effect estimation** via Lenth PSE, half-normal, Pareto plots
* **Model diagnostics** via residual plots and residuals-vs-predictors
* **Model adequacy** via $R^2_{pred}$, PRESS, VIF, curvature test
* **Uncoded equations** for engineering-unit interpretation
* **Model selection** via stepwise / AICc / BIC
* **Design augmentation** via fold-over and CCD conversion
* **Botched-run handling** for real-world deviation correction
* **Power & sample size** for design planning
* **Reporting** to Excel, Word, and PowerPoint

The next vignette, *Response Surface & Optimization*, extends the analysis
to second-order models and response-surface methodology.
