---
title: "Design Generation with iQualityR.doe"
author: "iQualityR Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Design Generation with iQualityR.doe}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  fig.align = "center"
)
```

## 1. Introduction

Experimental design is the foundation of any data-driven quality
improvement initiative. `iQualityR.doe` provides a unified interface for
generating more than a dozen design families --- from classical full
factorials to modern Definitive Screening Designs (DSD), mixture designs,
and computer-experiment space-filling designs.

This vignette walks through the **major design-generation APIs** using
realistic engineering case studies. After reading this vignette you will
know how to:

* Create full, fractional, and orthogonal (Taguchi) factorial designs
* Build response-surface designs (CCD, Box-Behnken, DSD)
* Generate mixture designs (simplex centroid / lattice / extreme vertices)
* Construct split-plot designs with hard-to-change factors
* Build D/A/I-optimal designs via coordinate exchange
* Generate Latin Hypercube and Maximin space-filling designs

```{r setup}
library(iQualityR.doe)
set.seed(42)
```

## 2. API Overview

### 2.1 The `IqrDoePlan` configurator

Every experiment starts with an `IqrDoePlan` object that declares the
factors, design type, and optional parameters (replication, center
points, blocking, alpha, resolution, randomization seed).

```{r plan-example}
factors <- list(
  list(name = "Temperature", type = "continuous", levels = c(80, 120)),
  list(name = "Pressure",    type = "continuous", levels = c(2, 6)),
  list(name = "Time",        type = "continuous", levels = c(10, 30))
)

plan <- IqrDoePlan$new(
  task_tag    = "injection_molding",
  design_type = "factorial",
  factors     = factors,
  replication = 2,
  center_points = 4,
  randomize   = TRUE,
  seed        = 123
)
plan$validate()
```

### 2.2 Factory functions

For convenience, `iQualityR.doe` exports one-call factory functions that
build a plan, generate the design, and return a ready-to-use
`IqrDoeTask`:

| Function | Design family |
|---|---|
| `factorial_design()` | Full / fractional factorial |
| `screening_design()` | Fractional screening (Res III/IV/V) |
| `rsm_design()` | CCD / Box-Behnken |
| `orthogonal_design()` | Taguchi orthogonal arrays |
| `taguchi_design()` | Taguchi robust design (control + noise) |

### 2.3 Direct generation via `DoeAnalyzer`

For advanced users, `DoeAnalyzer$new()$run(plan)` returns a results list
containing `design_info` (the generated design matrix) and, when response
data is supplied, ANOVA tables and model objects.

---

## 3. Case Study: Injection Molding Screening

### Problem

An injection-molding team wants to identify which of **5 process
factors** most affect part shrinkage: Temperature, Pressure, Holding Time,
Cooling Time, and Screw Speed. Budget allows 16 runs.

### Design: Fractional factorial (Resolution IV)

A $2^{5-1}_{IV}$ design uses 16 runs and keeps main effects clear of
two-factor interactions.

```{r screening}
factors <- lapply(LETTERS[1:5], function(nm) {
  list(name = nm, type = "continuous", levels = c(-1, 1))
})

task <- screening_design(factors = factors, resolution = "IV", seed = 2024)
head(task$results$design_info, 4)
nrow(task$results$design_info)
```

### Decision

With 16 runs and Resolution IV, the team can estimate all 5 main effects
free of two-factor-interaction aliasing. If interactions are suspected
active, the design can be **folded over** to de-alias them (see the
*Analysis & Visualization* vignette).

---

## 4. Case Study: Chemical Yield Optimization (RSM)

### Problem

After screening, three factors remain: Temperature, Time, and
Catalyst concentration. The team needs a second-order model to locate the
optimal yield region.

### Design: Central Composite Design (CCD)

```{r ccd}
factors <- list(
  list(name = "Temp",     type = "continuous", levels = c(80, 120)),
  list(name = "Time",     type = "continuous", levels = c(20, 40)),
  list(name = "Catalyst", type = "continuous", levels = c(1, 5))
)

task <- rsm_design(
  factors       = factors,
  design_type   = "ccd",
  center_points = 6,
  alpha         = "rotatable",
  randomize     = FALSE,
  seed          = 7
)
nrow(task$results$design_info)
table(task$results$design_info$PointType)
```

### Alternative: Box-Behnken Design

When axial points are infeasible (e.g. extreme temperatures), a
Box-Behnken design avoids the corners of the cube:

```{r bbd}
task_bbd <- rsm_design(
  factors     = factors,
  design_type = "box_behnken",
  randomize   = FALSE
)
nrow(task_bbd$results$design_info)
```

### Decision

A rotatable CCD with 6 center points provides 20 runs capable of fitting
a full second-order model. The team proceeds with data collection.

---

## 5. Case Study: Definitive Screening Design

### Problem

For 4 factors where curvature is possible but runs are expensive, a DSD
(Jones & Nachtsheim 2011) provides 3 levels per factor in only $2k + 1$
runs, with main effects not aliased with second-order effects.

```{r dsd}
factors <- lapply(paste0("X", 1:4), function(nm) {
  list(name = nm, type = "continuous", levels = c(-1, 0, 1))
})

plan <- IqrDoePlan$new(
  task_tag = "dsd_screening",
  design_type = "dsd",
  factors = factors,
  randomize = FALSE
)
analyzer <- DoeAnalyzer$new()
results <- analyzer$run(plan = plan)
nrow(results$design_info)
```

### Decision

A 4-factor DSD uses only 9 runs yet can detect curvature without
confounding main effects --- ideal for early-stage screening when
second-order behavior is uncertain.

---

## 6. Case Study: Formulation Optimization (Mixture Design)

### Problem

A food scientist needs to optimize a 3-component formulation (Water,
Sugar, Fat) where the proportions must sum to 1.

### Simplex Centroid Design

```{r mixture}
factors <- lapply(c("Water", "Sugar", "Fat"), function(nm) {
  list(name = nm, type = "continuous", levels = c(0, 1))
})

plan <- IqrDoePlan$new(
  task_tag = "formulation",
  design_type = "simplex_centroid",
  factors = factors,
  randomize = FALSE
)
results <- DoeAnalyzer$new()$run(plan = plan)
nrow(results$design_info)   # 2^3 - 1 = 7 points
```

### Simplex Lattice Design

For a denser grid, a {3,2} simplex lattice provides 6 blend points:

```{r lattice}
plan_lat <- IqrDoePlan$new(
  task_tag = "lattice",
  design_type = "simplex_lattice",
  factors = factors,
  alpha = 2,   # lattice degree m
  randomize = FALSE
)
results_lat <- DoeAnalyzer$new()$run(plan = plan_lat)
nrow(results_lat$design_info)
```

### Extreme Vertices Design

When each component has lower/upper bounds (not 0 to 1), the
McLean-Anderson extreme-vertices algorithm enumerates the feasible
polytope:

```{r extreme}
factors <- list(
  list(name = "A", type = "continuous", levels = c(0.2, 0.5)),
  list(name = "B", type = "continuous", levels = c(0.3, 0.6)),
  list(name = "C", type = "continuous", levels = c(0.1, 0.4))
)
plan_ev <- IqrDoePlan$new(
  task_tag = "extreme_vertices",
  design_type = "extreme_vertices",
  factors = factors,
  randomize = FALSE
)
results_ev <- DoeAnalyzer$new()$run(plan = plan_ev)
nrow(results_ev$design_info)
```

### Decision

For an unconstrained 3-component blend, the simplex centroid (7 runs)
is the most efficient starting point. If the resulting model shows
inadequate fit, augment with a simplex lattice.

---

## 7. Case Study: Split-Plot Design

### Problem

In a semiconductor process, **Oven Temperature** is hard to change
(each adjustment costs 2 hours), while **Gas Flow** and **Time** are
easy to change within an oven run.

```{r splitplot}
factors <- list(
  list(name = "OvenTemp", type = "continuous", levels = c(-1, 1),
       hard_to_change = TRUE),
  list(name = "GasFlow",  type = "continuous", levels = c(-1, 1)),
  list(name = "Time",     type = "continuous", levels = c(-1, 1))
)

plan <- IqrDoePlan$new(
  task_tag = "split_plot",
  design_type = "split_plot",
  factors = factors,
  randomize = TRUE,
  seed = 99
)
results <- DoeAnalyzer$new()$run(plan = plan)
head(results$design_info, 4)
# WholePlot column groups runs sharing the same hard-to-change setting.
table(results$design_info$WholePlot)
```

### Decision

The split-plot design respects the nested randomization structure: oven
temperature changes only twice (2 whole plots), while gas flow and time
vary within each whole plot.

---

## 8. Case Study: Optimal Design Construction

### Problem

A team needs a 12-run design for 4 factors with a specific model
including interactions, but standard designs do not fit. A
**D-optimal** design maximizes the determinant of $X'X$.

```{r optimal}
factors <- lapply(paste0("F", 1:4), function(nm) {
  list(name = nm, type = "continuous", levels = c(-1, 1))
})

analyzer <- DoeAnalyzer$new()
opt <- analyzer$create_optimal_design(
  factors     = factors,
  n_runs      = 12,
  model_order = "interaction",
  criterion   = "D",
  n_starts    = 5,
  seed        = 42
)
nrow(opt$design)
opt$criterion_name
```

### Comparing D / A / I optimality

```{r optimal-compare}
for (crit in c("D", "A", "I")) {
  res <- analyzer$create_optimal_design(
    factors = factors, n_runs = 12,
    model_order = "main", criterion = crit,
    n_starts = 5, seed = 42
  )
  cat(sprintf("  %s-optimal criterion value: %.4f\n", crit, res$criterion))
}
```

### Decision

When standard catalogs do not fit the model or run-count constraint,
optimal design construction via coordinate exchange (Meyer & Nachtsheim
1995) provides a customized, near-optimal design.

---

## 9. Space-Filling Designs for Computer Experiments

### Problem

When running physical experiments is impossible (e.g. simulation models),
**Latin Hypercube Sampling** (LHS) and **Maximin** designs fill the
factor space uniformly.

```{r lhs}
factors <- lapply(paste0("V", 1:3), function(nm) {
  list(name = nm, type = "continuous", levels = c(0, 1))
})
plan_lhs <- IqrDoePlan$new(
  task_tag = "sim_lhs",
  design_type = "lhs",
  factors = factors,
  randomize = FALSE,
  seed = 42
)
results_lhs <- DoeAnalyzer$new()$run(plan = plan_lhs)
# LHS generates max(10*k, 20) points by default.
head(results_lhs$design_info, 4)
```

### Decision

LHS is the workhorse for surrogate-model fitting and Bayesian
optimization initial designs (see the *AI/ML Advanced Features* vignette).

---

## 10. Design Evaluation

After generating a design, `evaluate_design()` reports D/A/G/I-optimality
and the condition number:

```{r evaluate}
# Evaluate the optimal design from Section 8.
analyzer <- DoeAnalyzer$new()
opt_factors <- lapply(paste0("F", 1:4), function(nm) {
  list(name = nm, type = "continuous", levels = c(-1, 1))
})
opt <- analyzer$create_optimal_design(
  factors = opt_factors, n_runs = 12,
  model_order = "main", criterion = "D",
  n_starts = 5, seed = 42
)
eval_out <- analyzer$evaluate_design(opt$design, model_order = "main")
eval_out$D_eff
eval_out$interpretation
```

For fractional factorials, `get_alias_structure()` reveals the aliasing
pattern:

```{r alias}
alias_info <- analyzer$get_alias_structure(n_factors = 5, resolution = "IV")
alias_info$generators
```

---

## 11. Conclusion

`iQualityR.doe` covers the full spectrum of experimental design
generation --- from classical factorials to modern DSDs, mixture designs,
and optimal designs. Key takeaways:

* Use `factorial_design()` / `screening_design()` for 2-level factorials
* Use `rsm_design()` for CCD / Box-Behnken response-surface designs
* Use `IqrDoePlan` with `design_type = "dsd"` for definitive screening
* Use mixture designs (`simplex_centroid` / `simplex_lattice` /
  `extreme_vertices`) for formulation problems
* Use `create_optimal_design()` when standard designs do not fit your
  constraints
* Use `lhs` / `maximin` for computer experiments and surrogate fitting

The next vignette, *Analysis & Visualization*, shows how to fit models,
compute effects, and generate diagnostic plots using the designs created
here.
