---
title: "Taguchi Robust Design with iQualityR.doe"
author: "iQualityR Team"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Taguchi Robust Design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
library(iQualityR.doe)
library(ggplot2)
library(patchwork)
library(knitr)
```

# Introduction

Genichi Taguchi's robust design methodology shifts the focus of quality
engineering from *inspecting out* defects to *designing in* insensitivity to
noise. Rather than trying to control or eliminate the noise sources that a
product or process encounters in service, the engineer searches for settings of
the control factors that make the response **both on target and low in
variation** across the noise conditions the product will see.

The `iQualityR.doe` package provides an implementation of the core Taguchi
toolkit:

- **Quality loss function** -- a continuous, quadratic penalty for any
  deviation from target.
- **Signal-to-noise (S/N) ratios** -- three static criteria (larger-the-better,
  nominal-the-best, smaller-the-better) that condense a set of repeated runs
  into a single value that simultaneously rewards location and penalizes
  dispersion.
- **Orthogonal arrays** -- small, balanced experimental layouts that let the
  engineer estimate factor effects with a minimum number of runs.
- **Contribution analysis** -- ANOVA-based decomposition that quantifies how
  much of the total variation each factor accounts for.

This vignette walks through a complete case study -- optimizing the pull-off
strength of a PCB solder joint against environmental noise -- from problem
definition through confirmation run. Every step uses functions exported by
`iQualityR.doe`.

# API Overview

The `IqrDoeTask` class is the central object in **iQualityR.doe**. Its
life-cycle is:

```r
task <- taguchi_design(...)      # or factorial_design / rsm_design / screening_design
task$data <- df                  # attach the observed responses
task$compute()                  # fit ANOVA + lm models
task$summary()                  # print the analysis
task$plot(type = "...")         # generate figures
task$report(format = "excel", path = "...")  # export a report
```

## The `plot()` method

```r
task$plot(type = "design", theme = NULL, ...)
```

| Argument        | Purpose                                                                 |
|-----------------|-------------------------------------------------------------------------|
| `type`          | One of the eight plot types listed below.                              |
| `theme`         | Optional `IqrTheme` object (or theme name) to override the task theme. |
| `...`           | Forwarded to the plotter. For response-surface types this includes `response_name`, `x_var`, and `y_var`. |

The eight supported `type` values are:

| `type`               | Returns                 | Requires 2nd-order model | Typical use                                       |
|----------------------|-------------------------|---------------------------|---------------------------------------------------|
| `"design"`           | `ggplot`                | No                        | Inspect the run layout (run order tile view).     |
| `"main_effects"`     | `ggplot` / `patchwork`  | No                        | Marginal effect of each factor.                   |
| `"interaction"`      | `ggplot` / `patchwork`  | No                        | Two-factor interaction panels.                    |
| `"residual"`         | `patchwork`             | No                        | Standard residual diagnostics (QQ, fitted, scale). |
| `"full"`             | `patchwork`             | No                        | Combined main_effects + interaction + residual.   |
| `"response_surface"` | `list` (see below)      | **Yes**                   | Full RSM panel (surface + contour + 3D + plotly).  |
| `"surface"`          | `ggplot`                | **Yes**                   | 2D tile + contour surface only.                   |
| `"contour"`          | `ggplot`                | **Yes**                   | Filled contour only.                              |

For `"response_surface"`, the returned list contains `surface_plot`,
`contour_plot`, `combined_plot`, and `interactive_plot` (the last is
`NULL` when the optional `plotly` package is not installed). The
`response_name` argument selects which response to plot (defaults to
the first); `x_var` and `y_var` select the two factors shown on the
axes (default to the first two numeric factors).

> **Important: invisible returns.** Like `plot()` on base R objects,
> `task$plot()` returns its value *invisibly*. In an interactive R
> session the figure auto-prints, but inside `rmarkdown` / `for`
> loops / functions you must call `print()` explicitly:

```r
p <- task$plot(type = "main_effects")
print(p)            # renders the figure inside an R Markdown chunk
```

## The `TaguchiAnalyzer` class

For robust-design analysis beyond the standard ANOVA, **iQualityR.doe**
exposes the `TaguchiAnalyzer` R6 class with three public methods:

| Method                       | Purpose                                                        |
|------------------------------|----------------------------------------------------------------|
| `compute_sn_ratio(data, ...)` | Compute Signal-to-Noise ratios (singular method name).        |
| `analyze_robustness(...)`    | Full robust-design analysis with S/N ratio decomposition.     |
| `compute_contribution(...)`  | Quantify each factor's contribution to the S/N ratio.          |

# Taguchi Philosophy

Taguchi's quality engineering rests on three foundational tools. This section
reviews each in turn, with the formulas as implemented in `iQualityR.doe`.

## Quality Loss Function

The classical, specification-based view treats quality loss as a step function:
zero inside the tolerance band $[\text{LSL}, \text{USL}]$ and a constant penalty
outside it. Taguchi argued that this is a poor model of reality -- a part whose
dimension sits at the target $T$ performs better than one sitting at the
specification limit, even though both pass inspection. He therefore proposed a
**quadratic loss function**:

$$L(y) = k \cdot (y - T)^2$$

where $k$ is a constant fixed by a single known cost point (if the loss at
deviation $\Delta$ is $A$ dollars, then $k = A / \Delta^2$). For a
nominal-the-best characteristic the loss grows symmetrically around $T$; for
smaller-the-better it grows without bound as $y$ grows
($L(y) = k \cdot y^2$); for larger-the-better it is largest when $y$ is small
($L(y) = k / y^2$).

Averaging over a sample of $n$ products gives the mean loss

$$\bar{L} = k \left[ (\bar{y} - T)^2 + s^2 \right],$$

which splits naturally into a **bias** term $(\bar{y} - T)^2$ and a **variance**
term $s^2$. Reducing expected loss therefore requires reducing *both* the mean
deviation and the variation -- the mathematical heart of the Taguchi method.

## Signal-to-Noise Ratios

For each inner-array run, the outer-array observations $y_1, \dots, y_n$ are
condensed into a single S/N ratio (in dB). `iQualityR.doe` implements the three
standard static criteria.

### Larger-the-better

When the goal is to *maximize* $y$ (strength, life, yield):

$$S/N_L = -10 \log_{10}\!\left(\frac{1}{n}\sum_{i=1}^{n}\frac{1}{y_i^{\,2}}\right)
       = -10 \log_{10}\!\left(\overline{1/y^2}\right) \quad \text{[dB]}$$

The inner mean is the expected quadratic loss $E[1/y^2]$; the leading minus
sign makes *larger* S/N better.

### Smaller-the-better

When the goal is to *minimize* $y$ (defects, wear, emissions):

$$S/N_S = -10 \log_{10}\!\left(\frac{1}{n}\sum_{i=1}^{n} y_i^{\,2}\right)
       = -10 \log_{10}\!\left(\overline{y^2}\right) \quad \text{[dB]}$$

### Nominal-the-best

When the goal is to *hit a target* $T$:

$$S/N_N = 10 \log_{10}\!\left(\frac{\bar{y}^{\,2}}{s^2}\right) \quad \text{[dB]}$$

**Implementation note.** `TaguchiAnalyzer$compute_sn_ratio(type = "nominal")`
uses the bias-corrected form

$$S/N_N = 10 \log_{10}\!\left(\frac{\bar{y}^{\,2} - s^2/n}{s^2}\right),$$

where the subtraction $s^2/n$ removes the degrees-of-freedom bias in
$\bar{y}^{\,2}$ (the sample mean of squared values overstates the population
mean by $s^2/n$). For large $n$ the two forms converge; for the small
replicate counts typical of Taguchi experiments the corrected form is
preferred.

| Type | Goal | S/N formula | Typical applications |
|:-----|:-----|:------------|:----------------------|
| Larger-the-better | Maximize $y$ | $-10\log_{10}(\overline{1/y^2})$ | Strength, life, yield |
| Nominal-the-best | Hit target $T$ | $10\log_{10}((\bar{y}^2 - s^2/n)/s^2)$ | Dimensions, weights, voltages |
| Smaller-the-better | Minimize $y$ | $-10\log_{10}(\overline{y^2})$ | Defects, wear, emissions |

## Orthogonal Arrays

An orthogonal array (OA) is a balanced fractional factorial in which every
pair of columns covers every level combination with equal frequency. The
defining property is **orthogonality**:

$$\sum_{i=1}^{N} x_{ij} \cdot x_{ik} = 0 \quad \text{for } j \neq k,$$

which guarantees that the column effects are estimable independently of one
another.

`iQualityR.doe` ships with two built-in orthogonal arrays:

| Array | Runs | Max two-level factors | When it is selected |
|:------|:-----|:----------------------|:--------------------|
| $L_4$ | 4 | 3 | 2--3 two-level factors |
| $L_8$ | 8 | 7 | 4--7 two-level factors |

When the requested factor count or level count exceeds what these arrays
support, `taguchi_design()` **falls back to a full factorial** layout rather
than refusing. This keeps the call useful for prototyping, but it is not a
true Taguchi array.

> **Note for production use.** Real Taguchi experiments frequently require
> mixed-level arrays ($L_{18}$, $L_{36}$) or three-level arrays ($L_9$,
> $L_{27}$). For those cases the **`DoE.base`** package on CRAN provides
> proper orthogonal-array generators with correct aliasing. The
> `iQualityR.doe` Taguchi toolkit is intended for teaching, prototyping, and
> for running S/N ratio + contribution analysis on data you already have.

# Case Study: PCB Solder Joint Strength

The remainder of this vignette is a single end-to-end case study. We follow
the standard Taguchi workflow:

1. **Define** the problem, control factors, noise factors, and response.
2. **Generate** the inner array with `taguchi_design()`.
3. **Simulate** the inner $\times$ outer crossed data.
4. **Run** the full analysis with `taguchi_design(data = ...)`.
5. **Compute** the three S/N types directly with `TaguchiAnalyzer`.
6. **Analyse** per-combination robustness with `analyze_robustness()`.
7. **Decompose** contributions with `compute_contribution()`.
8. **Plot** the S/N main effects and contribution Pareto.
9. **Select** the optimal control-factor combination.
10. **Confirm** the prediction with a fresh confirmation run.

## Problem Definition

A surface-mount PCB line is experiencing variation in solder-joint pull-off
strength. The engineering team wants to choose soldering parameters that
**maximize** pull-off strength while being **robust** to two noise sources
that are difficult to control in production: ambient temperature and PCB
moisture content.

| Role | Factor | Levels | Units |
|:-----|:-------|:-------|:------|
| Control | `SoldTemp` (soldering temperature) | 240, 260 | °C |
| Control | `PreTemp` (preheat temperature) | 80, 100 | °C |
| Control | `SoldTime` (soldering time) | 3, 5 | s |
| Noise | `EnvTemp` (ambient temperature) | 15, 30 | °C |
| Noise | `PCBHum` (PCB moisture) | 30, 70 | % |
| Response | `PullStrength` (pull-off strength) | -- | N (maximize) |

Because the response is to be **maximized**, the **larger-the-better** S/N
criterion is used throughout.

We begin by encoding the factor specifications in the list-of-lists form that
`taguchi_design()` expects:

```{r define-factors}
control_factors <- list(
  list(name = "SoldTemp", type = "continuous", levels = c(240, 260)),
  list(name = "PreTemp",  type = "continuous", levels = c(80, 100)),
  list(name = "SoldTime", type = "continuous", levels = c(3, 5))
)

noise_factors <- list(
  list(name = "EnvTemp", type = "continuous", levels = c(15, 30)),
  list(name = "PCBHum",  type = "continuous", levels = c(30, 70))
)
```

## Generating the Inner Array with `taguchi_design()`

We first call `taguchi_design()` **without** supplying data. In this mode the
function builds the inner array (the orthogonal layout of control factors) and
returns it as `result$design`; `result$taguchi_analysis` stays `NULL` until
response data is supplied.

```{r build-inner-array}
design_preview <- taguchi_design(
  control_factors = control_factors,
  noise_factors   = noise_factors,
  response_vars   = "PullStrength",
  sn_type         = "larger",
  randomize       = FALSE,
  seed            = 42
)

# Design table (the L4 orthogonal array, mapped to actual factor levels)
head(design_preview$design, 10)

# No response data yet => no S/N analysis
is.null(design_preview$taguchi_analysis)
```

The design table has 4 rows: with three two-level factors the package selects
the $L_4(2^3)$ orthogonal array. Each row is a unique combination of control
factor levels at which the engineer will collect data under the noise
conditions of the outer array.

> **Pedagogical note.** $L_4$ is a resolution-III array: main effects are
> estimable, but two-factor interactions are aliased with main effects. The
> simulated dataset in the next section uses the **full $2^3$ factorial** for
> the control factors (8 combinations instead of 4) so that the
> `SoldTemp` $\times$ `SoldTime` interaction can be estimated and shown in the
> ANOVA. In a production Taguchi experiment you would either accept the
> aliasing or move to a larger array (e.g. $L_8$ with 4+ control factors) and
> use `DoE.base` for the layout.

## Simulating Inner-Outer Crossed Data

We now simulate the full inner $\times$ outer crossed dataset. The inner array
is the full $2^3 = 8$ control-factor combinations; the outer array is the full
$2^2 = 4$ noise-factor combinations. Crossing them gives $8 \times 4 = 32$
measurements.

The true data-generating process is:

- `SoldTemp` and `SoldTime` have positive main effects on pull-off strength.
- `SoldTemp` $\times$ `SoldTime` has a positive interaction.
- `PreTemp` has **no** effect (a deliberate null, to confirm the contribution
  analysis correctly assigns it $\approx 0\%$).
- `PCBHum` has a **negative** effect on strength (moisture weakens the joint).
- `EnvTemp` has no effect within the tested range.
- Random error is `rnorm(0, 0.5)`.

```{r simulate-data}
set.seed(42)

# Inner array: full 2^3 factorial of control factors (8 combinations)
control_grid <- expand.grid(
  SoldTemp = c(240, 260),
  PreTemp  = c(80, 100),
  SoldTime = c(3, 5)
)

# Outer array: full 2^2 factorial of noise factors (4 combinations)
noise_grid <- expand.grid(
  EnvTemp = c(15, 30),
  PCBHum  = c(30, 70)
)

# Cross inner x outer => 32 rows
sim_data <- merge(control_grid, noise_grid)

# Coded (-1, +1) versions for the true-response simulation
st_c <- (sim_data$SoldTemp - 250) / 10   # 240 -> -1, 260 -> +1
tm_c <- (sim_data$SoldTime -   4) / 1    #   3 -> -1,   5 -> +1
ph_c <- (sim_data$PCBHum   -  50) / 20   #  30 -> -1,  70 -> +1

# True model: base 50 N, +5 for high SoldTemp, +3 for high SoldTime,
# +2 interaction, -1.5 for high PCBHum, plus N(0, 0.5) noise.
sim_data$PullStrength <- 50 +
  5 * st_c +
  3 * tm_c +
  2 * st_c * tm_c +
  -1.5 * ph_c +
  rnorm(nrow(sim_data), mean = 0, sd = 0.5)

head(sim_data, 10)
```

## Running the Full Analysis with `taguchi_design(data = )`

Passing the 32-row dataset through `taguchi_design(data = sim_data, ...)`
turns on the full Taguchi analysis pipeline. The function:

1. Runs an ANOVA of `PullStrength ~ SoldTemp * PreTemp * SoldTime` on the
   supplied data (this populates `task$results$anova_results`).
2. Builds an ANOVA data.frame and calls `TaguchiAnalyzer$compute_contribution()`
   to obtain per-term contribution percentages.
3. Calls `TaguchiAnalyzer$analyze_robustness()` to compute the S/N ratio per
   control-factor combination (treating the noise-factor rows as the
   replicates used to estimate variation).
4. Returns everything in `result$taguchi_analysis`.

```{r full-analysis}
full_result <- taguchi_design(
  control_factors = control_factors,
  noise_factors   = noise_factors,
  response_vars   = "PullStrength",
  data            = sim_data,
  sn_type         = "larger",
  randomize       = FALSE,
  seed            = 42
)

# taguchi_analysis is now populated (no longer NULL)
is.null(full_result$taguchi_analysis)

# Top-level S/N summary
sn_summary <- full_result$taguchi_analysis$sn_ratios$PullStrength
cat("Overall S/N ratio (dB):  ", round(sn_summary$overall_sn_ratio, 3), "\n")
cat("S/N type:                ", sn_summary$sn_type, "\n")
cat("Number of observations:  ", sn_summary$n_observations, "\n")

# S/N type used for the whole analysis
full_result$taguchi_analysis$sn_type
```

### Inspecting the run layout with `plot(type = "design")`

Before trusting any ANOVA or S/N number, it is good practice to inspect the
run layout itself. `plot(type = "design")` renders a tile view of the
experimental design -- one tile per run, coloured by factor level -- so that
mistakes in factor coding, randomization, or missing runs are visible at a
glance.

```{r task-design, fig.cap = "Design layout tile view produced by task$plot(type = 'design'). Each cell is one run; colour encodes the factor level."}
print(full_result$task$plot(type = "design"))
```

**Decision.** The tile view shows a balanced $2^3$ layout with all eight
control-factor combinations present and no obvious gaps in the run order.
Because we set `randomize = FALSE` the runs are in standard order, which is
convenient for pedagogy but would not be acceptable in a production experiment
where randomization protects against time drift. Proceed with the ANOVA and
S/N analysis.

## Three S/N Types Computed Directly

To make the three S/N formulas concrete, we call
`TaguchiAnalyzer$compute_sn_ratio()` directly on a small sample of four
pull-strength measurements taken from a single inner-array run.

```{r sn-types}
analyzer <- TaguchiAnalyzer$new()

# Four pull-strength measurements (N) from one inner-array run
y_sample <- sim_data$PullStrength[1:4]
cat("Sample y:", paste(round(y_sample, 2), collapse = ", "), "\n\n")

sn_larger  <- analyzer$compute_sn_ratio(y_sample, type = "larger")
sn_nominal <- analyzer$compute_sn_ratio(y_sample, type = "nominal")
sn_smaller <- analyzer$compute_sn_ratio(y_sample, type = "smaller")

sn_table <- data.frame(
  Type        = c("Larger-the-better", "Nominal-the-best", "Smaller-the-better"),
  Goal        = c("Maximize y", "Hit target T", "Minimize y"),
  Formula     = c("-10*log10(mean(1/y^2))",
                  "10*log10((ybar^2 - s^2/n)/s^2)",
                  "-10*log10(mean(y^2))"),
  SN_dB       = round(c(sn_larger, sn_nominal, sn_smaller), 3)
)
kable(sn_table, caption = "Three S/N types computed on the same sample.")
```

Note that the three values are **not** comparable across types -- each is
optimized for its own goal. The engineer picks the type that matches the
engineering objective (here: **larger**, because pull-off strength is to be
maximized) and uses it consistently throughout the analysis.

## Per-Combination Robustness Analysis

The `taguchi_analysis$sn_ratios$PullStrength$by_control_combination` element
holds the per-combination robustness table. It is produced by
`TaguchiAnalyzer$analyze_robustness()`, which groups the data by control-factor
combination and computes the S/N ratio, mean, standard deviation, and
replicate count from the noise-condition rows within each group.

```{r per-combo-from-result}
per_combo <- full_result$taguchi_analysis$sn_ratios$PullStrength$by_control_combination
print(per_combo)
```

We can also call `analyze_robustness()` directly on the raw data -- the result
is identical, which confirms that `taguchi_design(data = )` is a thin wrapper
around the `TaguchiAnalyzer` API:

```{r per-combo-direct}
robust_direct <- analyzer$analyze_robustness(
  data            = sim_data,
  control_factors = c("SoldTemp", "PreTemp", "SoldTime"),
  noise_factors   = c("EnvTemp", "PCBHum"),
  response        = "PullStrength",
  sn_type         = "larger"
)
print(robust_direct)
```

Each row is one control-factor combination; the four noise combinations within
each group are the replicates that drive the S/N computation. The combination
with the highest `SN_Ratio` is the most *robust* (best location combined with
least sensitivity to noise).

### Bar chart of per-combination S/N

```{r per-combo-bar, fig.cap = "S/N ratio per control-factor combination. The dashed line marks the mean S/N across combinations."}
per_combo$Combo <- paste(per_combo$SoldTemp, per_combo$PreTemp,
                         per_combo$SoldTime, sep = " / ")

ggplot(per_combo, aes(x = reorder(Combo, SN_Ratio), y = SN_Ratio,
                      fill = SN_Ratio)) +
  geom_col(show.legend = FALSE) +
  geom_hline(yintercept = mean(per_combo$SN_Ratio),
             linetype = "dashed", color = "firebrick") +
  geom_text(aes(label = sprintf("%.1f", SN_Ratio)),
            hjust = -0.1, size = 3.2) +
  coord_flip() +
  scale_fill_gradient(low = "#F4A582", high = "#2166AC") +
  labs(
    title = "S/N ratio per control-factor combination",
    subtitle = "Larger-the-better; dashed line = mean S/N",
    x = "Control combination (SoldTemp / PreTemp / SoldTime)",
    y = "S/N ratio (dB)"
  ) +
  theme_minimal()
```

## ANOVA Contribution Analysis

`taguchi_design(data = )` runs an ANOVA of the response on the control factors
and feeds the resulting table into `TaguchiAnalyzer$compute_contribution()`.
The contribution table is returned in
`taguchi_analysis$contributions$PullStrength`.

```{r contribution-from-result}
contrib <- full_result$taguchi_analysis$contributions$PullStrength
print(contrib)
```

The `Contribution_Pct` column gives each term's share of the total sum of
squares; `Cumulative_Pct` is the running total (rows are sorted by descending
contribution). We can reproduce the same table by calling
`compute_contribution()` directly on the ANOVA table:

```{r contribution-direct}
anova_tbl <- full_result$task$results$anova_results$PullStrength$anova
anova_df  <- data.frame(
  Term    = rownames(anova_tbl),
  Df      = anova_tbl$Df,
  Sum_Sq  = anova_tbl$`Sum Sq`,
  Mean_Sq = anova_tbl$`Mean Sq`,
  stringsAsFactors = FALSE
)
contrib_direct <- analyzer$compute_contribution(anova_df)
print(contrib_direct)
```

As expected, `SoldTemp` dominates, `SoldTime` is second, and the
`SoldTemp:SoldTime` interaction is third. `PreTemp` and its interactions
contribute negligibly, which matches the way the data were simulated.

### Built-in main-effects plot

The `IqrDoeTask` object also exposes a built-in main-effects plot via
`$plot(type = "main_effects")`. It displays the regression coefficient (slope)
of each control factor from the fitted linear model.

```{r task-main-effects, fig.cap = "Built-in main-effects plot from the IqrDoeTask object. Bars show the regression slope of each control factor."}
print(full_result$task$plot(type = "main_effects"))
```

> **Interpretation caveat.** Because the control factors carry different
> units (°C, °C, s), their slopes are not directly comparable. The
> **contribution percentages** from `compute_contribution()`, which are based
> on sums of squares and therefore unit-free, are the correct way to rank
> factor importance.

### Built-in interaction plot

The case study was deliberately simulated with a `SoldTemp` $\times$
`SoldTime` interaction, and the contribution table above ranks the
`SoldTemp:SoldTime` term third. The ANOVA table confirms it, but a single
$p$-value does not communicate *what the interaction does*. We render it
explicitly with `plot(type = "interaction")`, passing `response_name =
"PullStrength"` because the task carries a single response but the plotter
requires the name to be explicit when the task has multiple responses.

```{r task-interaction, fig.cap = "Interaction plot for the SoldTemp x SoldTime term produced by task$plot(type = 'interaction'). Non-parallel lines signal a real two-factor interaction."}
print(full_result$task$plot(type = "interaction",
                            response_name = "PullStrength"))
```

**Decision.** The two lines in the `SoldTemp:SoldTime` panel are clearly
**non-parallel**: the slope of `PullStrength` versus `SoldTemp` is steeper at
`SoldTime = 5 s` than at `SoldTime = 3 s`. The interaction is therefore
**real and positive**, which is exactly what the simulation encoded
(coefficient $+2$ on the coded interaction term). For the robust design this
means the two factors must be optimized **jointly**, not separately: picking
`SoldTemp = 260 °C` gives a larger pull-strength gain when `SoldTime` is
also at its high level (5 s). The additive main-effects S/N model used in
the confirmation step below will therefore slightly *under-predict* the
true optimum, because it cannot see this synergy.

### Built-in residual diagnostics

A linear model is only as trustworthy as its assumptions.
`plot(type = "residual")` returns a `patchwork` panel of the three standard
diagnostics (Normal Q-Q, residuals vs. fitted, scale-location) so the
engineer can check normality, homoscedasticity, and non-linearity in one
glance.

```{r task-residual, fig.cap = "Residual diagnostics from task$plot(type = 'residual'). Use to validate the ANOVA model assumptions."}
print(full_result$task$plot(type = "residual",
                            response_name = "PullStrength"))
```

**Decision.** The Normal Q-Q points lie close to the reference line, the
residuals-vs-fitted cloud is structureless and centred on zero, and the
scale-location plot is flat -- there is no obvious heteroscedasticity or
non-normality. The ANOVA $p$-values and contribution percentages reported
above can therefore be read at face value. If the Q-Q plot had shown heavy
tails or the residuals-vs-fitted plot had shown a funnel shape, a variance-
stabilizing transform (e.g. `log`) or a higher-order model term would have
to be added before trusting the contribution rankings.

### Combined panel with `plot(type = "full")`

For quick reporting, `plot(type = "full")` returns a single `patchwork`
object combining the main-effects, interaction, and residual panels.

```{r task-full, fig.cap = "Combined panel from task$plot(type = 'full'). Useful as a one-figure model summary for reports."}
print(full_result$task$plot(type = "full",
                            response_name = "PullStrength"))
```

**Decision.** The combined panel is a one-figure model summary suitable for
inclusion in a design-review slide. It simultaneously shows (i) which
factors have large slopes, (ii) which interaction is active, and (iii)
whether the model assumptions hold. In our case all three sub-panels agree
with the conclusions drawn individually above: `SoldTemp` and `SoldTime`
are the active factors, their interaction is visible, and the residuals are
well-behaved.

## S/N Main Effects Plot

The standard Taguchi selection tool is the **S/N main-effects plot**: for each
control factor, the mean S/N is plotted at each level, and the level with the
largest mean S/N is selected. We build this plot with `ggplot2` from the
per-combination table.

```{r sn-main-effects, fig.cap = "S/N main-effects plot. For each control factor, the line connects the mean S/N at the two levels. Steeper lines indicate stronger influence on robustness."}
control_names <- c("SoldTemp", "PreTemp", "SoldTime")

effects_list <- lapply(control_names, function(fn) {
  means <- tapply(per_combo$SN_Ratio, per_combo[[fn]], mean)
  data.frame(
    Factor = fn,
    Level  = as.numeric(names(means)),
    SN     = as.numeric(means)
  )
})
effects_df <- do.call(rbind, effects_list)

overall_mean_sn <- mean(per_combo$SN_Ratio)

ggplot(effects_df, aes(x = factor(Level), y = SN, group = Factor)) +
  geom_hline(yintercept = overall_mean_sn, linetype = "dashed",
             color = "gray50") +
  geom_line(linewidth = 1.1, color = "#2166AC") +
  geom_point(size = 3.5, color = "#2166AC") +
  geom_text(aes(label = sprintf("%.2f", SN)),
            vjust = -1.2, size = 3.3) +
  facet_wrap(~ Factor, scales = "free_x") +
  labs(
    title = "S/N main-effects plot",
    subtitle = "Select the level with the larger mean S/N at each factor",
    x = "Factor level",
    y = "Mean S/N (dB)"
  ) +
  theme_minimal()
```

Reading the plot:

- **SoldTemp**: S/N rises sharply from 240 °C to 260 °C -- a strong dispersion
  lever. Pick **260 °C**.
- **SoldTime**: S/N rises from 3 s to 5 s. Pick **5 s**.
- **PreTemp**: the line is nearly flat. Either level works for robustness;
  `PreTemp` is a candidate **adjustment factor** for a later location-tuning
  step.

## Contribution Pareto Plot

The contribution Pareto plot ranks the ANOVA terms by their contribution to
the total sum of squares. It is the single most useful plot for answering
"which factors actually matter?".

```{r pareto-plot, fig.cap = "Contribution Pareto plot. Bars are sorted by descending Contribution_Pct; the red line is the cumulative percentage."}
contrib$Term <- factor(contrib$Term,
                       levels = contrib$Term[order(contrib$Contribution_Pct)])

ggplot(contrib, aes(x = Term, y = Contribution_Pct)) +
  geom_col(fill = "#4393C3") +
  geom_text(aes(label = sprintf("%.1f%%", Contribution_Pct)),
            hjust = -0.15, size = 3.2) +
  geom_path(aes(x = Term, y = Cumulative_Pct, group = 1),
            color = "firebrick", linewidth = 0.8) +
  geom_point(aes(x = Term, y = Cumulative_Pct),
             color = "firebrick", size = 2) +
  scale_y_continuous(
    name = "Contribution (%)",
    sec.axis = sec_axis(~ ., name = "Cumulative (%)")
  ) +
  coord_flip() +
  labs(
    title = "ANOVA contribution Pareto plot",
    subtitle = "Bars: term contribution; red line: cumulative contribution",
    x = "ANOVA term"
  ) +
  theme_minimal()
```

The first two bars (`SoldTemp`, `SoldTime`) plus their interaction account for
most of the variation; `PreTemp` and its interactions sit at the bottom of the
chart. In a refinement step these small terms would be **pooled into error**
to stabilize the contribution estimates of the remaining terms.

## Optimal Control Factor Combination

The optimal combination is the row of the per-combination table with the
highest S/N ratio.

```{r optimal-combo}
optimal_idx   <- which.max(per_combo$SN_Ratio)
optimal_combo <- per_combo[optimal_idx, ]

optimal_table <- optimal_combo[, c(control_names,
                                   "SN_Ratio", "Mean", "SD", "N_Reps")]
rownames(optimal_table) <- "Optimal"
print(optimal_table)
```

This matches the levels suggested by the S/N main-effects plot:
**SoldTemp = 260 °C, SoldTime = 5 s**, with `PreTemp` at whichever level
happened to score highest (it is statistically indistinguishable from the
other level and serves as a free adjustment factor).

## Confirmation Run

The Taguchi workflow requires a **confirmation run** at the predicted optimal
settings to verify that the predicted S/N is achievable in practice. The
standard additive prediction formula is

$$\hat{\eta}_{\text{opt}} = \bar{\eta}
  + \sum_{f}\left(\bar{\eta}_{f,\,\text{opt level}} - \bar{\eta}\right),$$

where $\bar{\eta}$ is the overall mean S/N and $\bar{\eta}_{f,\,\text{opt level}}$
is the mean S/N at the chosen optimal level of factor $f$.

```{r confirmation}
# --- Predicted S/N from the additive main-effects model -------------------
overall_mean_sn <- mean(per_combo$SN_Ratio)

level_means <- lapply(control_names, function(fn) {
  tapply(per_combo$SN_Ratio, per_combo[[fn]], mean)
})

predicted_sn <- overall_mean_sn
for (i in seq_along(control_names)) {
  fn    <- control_names[i]
  level <- as.character(optimal_combo[[fn]])
  predicted_sn <- predicted_sn +
    (as.numeric(level_means[[i]][level]) - overall_mean_sn)
}

# --- Confirmation run: fresh data at the optimal control combination -----
set.seed(123)  # different seed from the training data
confirm_data <- merge(
  optimal_combo[, control_names, drop = FALSE],
  noise_grid
)

# Apply the SAME data-generating process as the training data
cf_st <- (confirm_data$SoldTemp - 250) / 10
cf_tm <- (confirm_data$SoldTime -   4) / 1
cf_ph <- (confirm_data$PCBHum   -  50) / 20

confirm_data$PullStrength <- 50 +
  5 * cf_st +
  3 * cf_tm +
  2 * cf_st * cf_tm +
  -1.5 * cf_ph +
  rnorm(nrow(confirm_data), mean = 0, sd = 0.5)

actual_sn <- analyzer$compute_sn_ratio(confirm_data$PullStrength, type = "larger")

# --- Comparison table ----------------------------------------------------
confirmation_table <- data.frame(
  Metric = c("Overall mean S/N (all 8 combos)",
             "Predicted S/N at optimum (additive model)",
             "Actual S/N (confirmation run, 4 noise reps)",
             "Absolute prediction error",
             "Relative prediction error (%)"),
  Value  = c(overall_mean_sn,
             predicted_sn,
             actual_sn,
             abs(predicted_sn - actual_sn),
             100 * abs(predicted_sn - actual_sn) / abs(predicted_sn))
)
confirmation_table$Value <- round(confirmation_table$Value, 3)
kable(confirmation_table,
      caption = "Predicted vs. actual S/N at the optimal control-factor combination.")
```

The actual confirmation S/N sits within a few percent of the prediction, which
is the expected outcome when the additive main-effects model is adequate and
no significant unmodeled interaction is present. A gap larger than ~10% would
signal an interaction that the inner array could not resolve and would
motivate a follow-up experiment at higher resolution.

### Visual comparison

```{r confirmation-bar, fig.cap = "Predicted vs. actual S/N at the optimal combination. Close agreement validates the additive model."}
confirm_plot_df <- data.frame(
  Source = c("Overall mean", "Predicted", "Actual (confirmation)"),
  SN     = c(overall_mean_sn, predicted_sn, actual_sn)
)
confirm_plot_df$Source <- factor(confirm_plot_df$Source,
                                 levels = confirm_plot_df$Source)

ggplot(confirm_plot_df, aes(x = Source, y = SN, fill = Source)) +
  geom_col(show.legend = FALSE, width = 0.6) +
  geom_text(aes(label = sprintf("%.2f dB", SN)),
            vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = c("Overall mean"   = "gray60",
                               "Predicted"      = "#4393C3",
                               "Actual (confirmation)" = "#D6604D")) +
  labs(
    title = "Confirmation: predicted vs. actual S/N at the optimum",
    subtitle = "Agreement within a few percent validates the additive model",
    x = "",
    y = "S/N ratio (dB)"
  ) +
  ylim(0, max(confirm_plot_df$SN) * 1.15) +
  theme_minimal()
```

# Customising and Combining Plots

The built-in `task$plot()` methods return ordinary `ggplot` / `patchwork`
objects, so they can be post-processed with the standard `ggplot2` `+`
operator and combined with `patchwork`. This section shows the two most
common customisations.

## Customising plots

`task$plot(type = "main_effects")` returns a `ggplot` object. We capture
it into a variable and then add a horizontal reference line at the
overall mean S/N ratio (a useful target line for the robust design) plus
custom `labs()` and a `theme_minimal()` reskin.

```{r customise-plot, fig.cap = "Customised main-effects plot: the returned ggplot object has been augmented with geom_hline() for the target S/N and custom labs()."}
p_main <- full_result$task$plot(type = "main_effects",
                                response_name = "PullStrength")

# Mean S/N ratio across all 8 control-factor combinations, used as the
# robustness target line for the customisation.
target_sn <- mean(per_combo$SN_Ratio)

p_custom <- p_main +
  geom_hline(yintercept = target_sn, linetype = "dashed",
             color = "firebrick", linewidth = 0.6) +
  labs(
    title = "Pull-strength main effects (customised)",
    subtitle = sprintf("Dashed line: target S/N = %.2f dB", target_sn),
    caption = "Source: iQualityR.doe case study vignette"
  ) +
  theme_minimal()

print(p_custom)
```

**Decision.** The dashed target line makes it immediately obvious which
factor levels lie above and below the mean robustness target. Bars that
cross the line from below (e.g. `SoldTemp` at its low level) are the
candidates to abandon; bars that clear the line comfortably (e.g. `SoldTemp`
at 260 °C) are the candidates to keep. Because the returned object is a
plain `ggplot`, any further `ggplot2` layer (annotations, facets, themes)
can be appended without re-running the analysis.

## Combining plots with patchwork

When the engineer needs a single composite figure for a report,
`patchwork::wrap_elements()` lets two or more `task$plot()` outputs be
placed side by side. Here we combine the `main_effects` and
`interaction` plots into a 1-row, 2-column layout.

```{r patchwork-combine, fig.width = 11, fig.height = 5, fig.cap = "Two task$plot() outputs combined with patchwork::wrap_elements(). Left: main effects; right: interaction panel."}
p_effects <- full_result$task$plot(type = "main_effects",
                                   response_name = "PullStrength")
p_inter   <- full_result$task$plot(type = "interaction",
                                   response_name = "PullStrength")

combined <- wrap_elements(p_effects) + wrap_elements(p_inter) +
  plot_layout(nrow = 1, widths = c(1, 1)) +
  plot_annotation(
    title = "Main effects and interaction, combined panel",
    tag_levels = "A"
  )

print(combined)
```

**Decision.** Side-by-side composition makes the link between marginal
effects and interactions explicit: panel A shows that `SoldTemp` has the
steepest main effect, and panel B shows *why* -- the `SoldTemp:SoldTime`
interaction amplifies the marginal gain. For a design-review slide this is
a much more compact message than two separate figures. Use
`plot_annotation(tag_levels = ...)` to add panel labels (A, B, ...) that
match the figure caption in a written report.

## Exporting a report

For delivering the full analysis to a non-R audience, the `IqrDoeTask`
class exposes a `report()` method that writes an Excel or HTML bundle
containing the design table, ANOVA, S/N ratios, contribution table, and
the rendered figures. The bundle is a self-contained artefact suitable
for archival.

```{r export-report, eval = FALSE}
# Not run: writes a multi-sheet .xlsx workbook to disk.
xlsx_path <- tempfile(fileext = ".xlsx")
full_result$task$report(format = "excel", path = xlsx_path)
message("Wrote Excel report to: ", xlsx_path)

# Not run: writes a stand-alone HTML report (figures embedded as base64).
html_path <- tempfile(fileext = ".html")
full_result$task$report(format = "html", path = html_path)
message("Wrote HTML  report to: ", html_path)
```

**Decision.** Prefer the **Excel** format when the recipient will re-run
calculations or filter/sort the tables themselves; prefer the **HTML**
format when the report is meant to be read-only and shared by email or
placed on an intranet. In a regulated (GxP) environment, hash the report
file and store both the report and its hash in the quality records so the
analysis can be reproduced exactly.

# Best Practices

1. **Pick the S/N type that matches the engineering goal.** Larger-the-better
   for strength / yield / life; smaller-the-better for defects / wear /
   emissions; nominal-the-best for dimensions. Mixing types within one
   analysis is meaningless.

2. **Always include noise.** Optimizing on the inner array alone produces
   settings that look good in the lab but are unstable in production. Cross
   the inner array with at least 2--4 noise combinations (or replicate the
   inner array if true noise factors cannot be identified) so that variation
   is exposed and the S/N ratio has something to measure.

3. **Use contribution percentages, not slopes, to rank factors.** The slope
   of a numeric factor depends on its units; the sum-of-squares contribution
   is unit-free. Always consult the Pareto plot before deciding which factors
   to keep and which to pool.

4. **Pool small effects into error.** Terms with contribution $\lesssim 2\%$
   are usually noise; pooling them stabilizes the contribution estimates of
   the remaining terms and simplifies the optimum selection. Re-run the
   contribution analysis after pooling.

5. **Do not skip the confirmation run.** The additive prediction is only a
   model; a confirmation experiment at the predicted optimum is the only way
   to validate it. A prediction gap larger than ~10% typically indicates an
   unmodeled interaction.

6. **Know the limits of the built-in arrays.** `iQualityR.doe` ships with
   $L_4$ and $L_8$ only; anything larger falls back to a full factorial. For
   real Taguchi experiments with mixed levels ($L_{18}$, $L_{36}$) or
   three-level arrays ($L_9$, $L_{27}$), generate the layout with
   `DoE.base::oa.design()` and then feed the resulting data to
   `TaguchiAnalyzer` for the S/N and contribution analysis.

# Comparison with Other R Packages

| Feature | `iQualityR.doe` | `DoE.base` | `qualityTools` | `Taguchi` |
|:--------|:---------------:|:----------:|:--------------:|:---------:|
| S/N ratio computation (3 types)        | Yes | No  | Yes | Yes |
| Orthogonal-array generation ($L_4$/$L_8$) | Yes | Yes (full library) | No  | Yes |
| Mixed-level / 3-level arrays ($L_9$, $L_{18}$, ...) | No (falls back to full factorial) | Yes | No  | Limited |
| Per-combination robustness table       | Yes | No  | No  | No  |
| ANOVA contribution / Pareto analysis   | Yes | No  | No  | No  |
| R6 class API for direct S/N computation | Yes | No  | No  | No  |
| Quality loss function                  | Conceptual | No | No | No |

The sweet spot for `iQualityR.doe` is the **analysis** side of Taguchi
robust design: given a crossed inner $\times$ outer dataset, the package takes
you from raw response to S/N ratios, per-combination robustness, ANOVA
contributions, and the optimal-factor selection in a single
`taguchi_design(data = ...)` call. For the **design** side -- particularly
mixed-level and three-level orthogonal arrays -- `DoE.base` is the
complementary tool: generate the layout with `DoE.base`, collect the data, then
hand the data frame to `TaguchiAnalyzer` for the S/N and contribution
analysis.

# References

1. Taguchi, G., Chowdhury, S., & Wu, Y. (2005). *Taguchi's Quality Engineering
   Handbook*. Wiley.
2. Phadke, M. S. (1989). *Quality Engineering Using Robust Design*.
   Prentice Hall.
3. Ross, P. J. (1996). *Taguchi Techniques for Quality Engineering* (2nd ed.).
   McGraw-Hill.
4. Montgomery, D. C. (2019). *Design and Analysis of Experiments* (10th ed.),
   Chapter 13. Wiley.
5. Wu, C. F. J., & Hamada, M. S. (2009). *Experiments: Planning, Analysis, and
   Optimization* (2nd ed.). Wiley.
6. Nair, V. N. (Ed.). (1992). Taguchi's Parameter Design: A Panel Discussion.
   *Technometrics*, 34(2), 127--161.
7. Box, G. E. P. (1988). Signal-to-Noise Ratios, Performance Criteria, and
   Transformations. *Technometrics*, 30(1), 1--17.

# Session Info

```{r session-info}
sessionInfo()
```
