---
title: "Statistical Foundations of Design of Experiments"
author: "iQualityR Team"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Statistical Foundations of Design of Experiments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(iQualityR.doe)
library(ggplot2)
library(patchwork)
```

# Introduction

This vignette is a self-contained, executable exposition of the
statistical foundations on which the design generators in
**iQualityR.doe** are built. Every concept is illustrated on a single,
consistent case study --- a $2^3$ factorial design with two replicates
and four center points studying a chemical-process yield --- so that
the reader can move from theory to reproducible practice in a single
R session.

The presentation covers:

* the **linear-model framework** and the role of the design matrix $X$;
* **least-squares estimation** via `task$compute()`;
* **degrees of freedom**, pure error, and lack of fit;
* **effect estimation** and the **effect sparsity** principle, with a
  Pareto chart and a normal probability plot of effects built from the
  fitted model;
* **aliasing** in fractional-factorial designs, illustrated with
  `screening_design()`;
* the four core formulas of DOE: effect estimation, the $F$-statistic,
  $D$-optimality, and the curvature test;
* **center points** and the single-degree-of-freedom curvature test;
* **randomization** and the reproducible run order produced by the
  package.

> **A note on implementation status.** The classical
> unreplicated-factorial diagnostics --- Lenth's pseudo-standard error,
> the half-normal plot of effects, and fold-over augmentation --- are
> *not yet implemented* in **iQualityR.doe**. Where the discussion
> needs them, the math is given explicitly *without* a fake call to a
> missing function, so the reader can apply the formulas by hand or
> use a complementary package such as `FrF2` or `BsMD` for the
> corresponding diagnostics.

# Case Study: A Chemical-Process Yield Optimization

A chemical-engineering team is studying the batch yield of a catalysed
reaction. Three process variables are suspected to drive the yield:

| Factor       | Symbol          | Low level | High level | Unit      |
|--------------|-----------------|-----------|------------|-----------|
| Temperature  | `Temperature`   | 100       | 200        | $^\circ$C |
| Pressure     | `Pressure`      | 1         | 5          | atm       |
| Reaction time| `Time`          | 30        | 90         | min       |

The response is `Yield` (%), to be maximised. The team chooses a $2^3$
full-factorial design with **two replicates** of the eight corner
points ($n_f = 16$) plus **four center points** ($n_c = 4$), for a
total of $n = 20$ runs. The two replicates provide a pure-error
estimate of $\sigma^2$; the four center points enable a
single-degree-of-freedom curvature test (Section 9).

The factor list is built first, with **numeric** level vectors (the
package validator rejects character levels):

```{r case-study-factors}
factors <- list(
  list(name = "Temperature", type = "continuous", levels = c(100, 200)),
  list(name = "Pressure",    type = "continuous", levels = c(1,     5)),
  list(name = "Time",        type = "continuous", levels = c(30,   90))
)

factors
```

# The Linear Model Framework

## The Matrix Form

Every regular design in this package is analyzed through the linear
model

$$
y = X\beta + \epsilon,
$$

where:

* $y$ is the $n \times 1$ vector of observed responses;
* $X$ is the $n \times p$ **design matrix** (also called the *model
  matrix* once the intercept column is appended);
* $\beta$ is the $p \times 1$ vector of unknown coefficients;
* $\epsilon$ is the $n \times 1$ vector of random errors, with
  $\epsilon \sim N(0, \sigma^2 I_n)$ under the classical assumptions.

The rows of $X$ are determined by the experimental plan, while its
columns are determined by the model that the engineer intends to fit.
The central idea of experimental design is to choose $X$ *before* the
data are collected so that the subsequent inference will be as
efficient and unambiguous as possible.

## Coding of Factor Levels

For two-level factors the standard coding uses the values $-1$ (low)
and $+1$ (high). This **orthogonal coding** has three convenient
properties:

1. The intercept $\hat\beta_0$ is the grand mean of the responses.
2. The regression coefficient $\hat\beta_i$ is one-half of the
   corresponding factor effect (Section 5).
3. When the design is orthogonal, the columns of $X$ are mutually
   orthogonal and the cross-product $X^\top X$ is diagonal, which
   makes the coefficient estimates uncorrelated.

The coded value of each factor is obtained from the natural units by
a linear transformation:

$$
x_A = \frac{A - (A_{\text{low}} + A_{\text{high}})/2}
           {(A_{\text{high}} - A_{\text{low}})/2}.
$$

For our case study this means:

* $x_{\text{Temperature}} = (\text{Temperature} - 150) / 50$,
* $x_{\text{Pressure}}    = (\text{Pressure}    -   3) /   2$,
* $x_{\text{Time}}        = (\text{Time}        -  60) /  30$.

## The Design Matrix for the $2^3$ Case Study

For a $2^3$ design analyzed with the full first-order-with-interaction
model $y \sim A * B * C$, the model matrix $X$ has eight columns: the
intercept, three main-effect columns, three two-way interactions, and
one three-way interaction. We construct it explicitly for the eight
corner points in standard (Yates) order:

```{r build-X-matrix}
# Coded design matrix for the 2^3 factorial in standard (Yates) order.
coded_design <- expand.grid(
  Temperature = c(-1,  1),
  Pressure    = c(-1,  1),
  Time        = c(-1,  1)
)
coded_design

# Full model matrix for y ~ Temperature * Pressure * Time.
X <- model.matrix(~ Temperature * Pressure * Time, data = coded_design)
X
```

The resulting $8 \times 8$ matrix is the building block of the design.
With two replicates the analysis matrix becomes $16 \times 8$, and the
four center points add four more rows whose non-intercept entries are
all zero (since $A = B = C = 0$ at the center).

## A Column of Ones and the Identity Column

A column of ones is appended to $X$ to fit the intercept. In the
analysis of variance this column does not consume any degrees of
freedom beyond the single one used to estimate the grand mean. All
subsequent effect columns are contrasts and therefore have zero
column sum in the orthogonal coding --- as can be verified directly:

```{r column-sums}
colSums(X)
```

Only the intercept column has a non-zero sum (8, the number of rows);
every contrast column sums to zero, confirming orthogonality.

# Least-Squares Estimation

## The Normal Equations

The ordinary least-squares estimator minimises the residual sum of
squares

$$
S(\beta) = (y - X\beta)^\top (y - X\beta).
$$

Taking the gradient with respect to $\beta$ and setting it to zero
yields the **normal equations**

$$
X^\top X \, \hat\beta = X^\top y,
$$

with unique solution (when $X^\top X$ is invertible)

$$
\hat\beta = (X^\top X)^{-1} X^\top y.
$$

Under the classical assumptions, the Gauss-Markov theorem states that
$\hat\beta$ is the best linear unbiased estimator (BLUE) of $\beta$.
If in addition $\epsilon \sim N(0, \sigma^2 I)$, then

$$
\hat\beta \sim N\bigl(\beta,\; \sigma^2 (X^\top X)^{-1}\bigr),
$$

which is the basis for the $t$- and $F$-tests reported by
`task$compute()`. The variance-covariance matrix of the estimator,

$$
\text{Cov}(\hat\beta) = \sigma^2 (X^\top X)^{-1},
$$

is the central object of design optimality, because every sensible
criterion for "good" designs is a functional of $(X^\top X)^{-1}$.

## Building the Design with `factorial_design()`

We now build the case-study design with the package's high-level
constructor. The argument `center_points = 4` appends four center
points at the geometric center of the design region (Temperature =
150, Pressure = 3, Time = 60). The argument `replication = 2` doubles
the eight corner points. Setting `seed = 42` makes the randomised run
order reproducible.

```{r build-design}
task <- factorial_design(
  factors       = factors,
  design_type   = "factorial",
  response_vars = "Yield",
  replication   = 2,
  center_points = 4,
  randomize     = TRUE,
  seed          = 42
)

# Inspect the design table (first 6 rows).
head(task$results$design_info)

# Total number of runs.
nrow(task$results$design_info)
```

The `design_info` data frame has `r nrow(task$results$design_info)`
rows: 16 factorial runs (8 corners $\times$ 2 replicates) plus 4
center points. The `RunOrder` column reflects the randomised order;
`Replication` identifies the replicate (1 or 2) for the factorial
portion and is `NA` for the four center points.

## Simulating the Response Data

In a real study the responses would be measured after the physical
trials. Here we simulate them from a known physics-inspired model so
that the analysis is reproducible. The true model has:

* a baseline yield of $70\%$ at the center of the design region;
* positive main effects of `Temperature`, `Pressure`, and `Time`;
* a positive `Temperature` $\times$ `Pressure` interaction;
* **negative quadratic curvature**, so the average factorial yield is
  $6$ points below the center yield --- a strong signal for the
  curvature test of Section 9;
* Gaussian noise with $\sigma = 1$.

```{r simulate-responses}
set.seed(2024)

design_df <- task$results$design_info

# Convert to coded variables {-1, +1} (0 at center points).
A <- (design_df$Temperature - 150) / 50
B <- (design_df$Pressure    -   3) /   2
C <- (design_df$Time        -  60) /  30

# True model: yield = 70 + 2.5*A + 1.5*B + 2.0*C + 1.0*A*B
#                   - 2.0*(A^2 + B^2 + C^2) + N(0, 1)
# At factorial points A^2 = B^2 = C^2 = 1, so the quadratic term is -6.
# At center A = B = C = 0, so the quadratic term vanishes.
design_df$Yield <- 70 +
  2.5 * A + 1.5 * B + 2.0 * C +
  1.0 * A * B +
  -2.0 * (A^2 + B^2 + C^2) +
  rnorm(nrow(design_df), sd = 1.0)

# Attach the response data to the task.
task$data <- design_df

head(task$data, 5)
```

## Fitting the Model with `task$compute()`

Calling `task$compute()` regenerates the design table (reproducibly,
because `seed = 42` was set) and, because `task$data` is now populated
and `response_vars = "Yield"` was declared on the plan, fits the
full-interaction model `Yield ~ Temperature * Pressure * Time` and
populates:

* `task$results$anova_results` --- a named list, one entry per
  response;
* `task$results$effects` --- main and interaction effects for the
  first response;
* `task$results$model` --- the fitted `lm` object for the first
  response;
* `task$results$model_fit` --- $R^2$, adjusted $R^2$, and the
  lack-of-fit $p$-value (the latter is `NA` here because the analyzer
  does not currently compute a pure-error-based lack-of-fit
  $p$-value).

```{r compute-anova}
task$compute()
task$summary()
```

## Reading the ANOVA Table

The ANOVA table produced by `task$results$anova_results$Yield$anova`
partitions the total sum of squares into contributions from each main
effect, each two-way interaction, the three-way interaction, and the
residuals:

```{r inspect-anova}
task$results$anova_results$Yield$anova
```

The estimated coefficients of the fitted model can be inspected
directly through `task$results$model`:

```{r inspect-coef}
coef(task$results$model)
```

The package's `effects` list extracts the main and
pairwise-interaction regression coefficients (recall that the
**effect** is twice the regression coefficient, see Section 5):

```{r inspect-effects}
task$results$effects$main
task$results$effects$interaction
```

# Degrees of Freedom

## Source-by-Source Decomposition

Each effect in an ANOVA decomposition consumes one or more degrees of
freedom. For a two-level factor the corresponding contrast column has
one degree of freedom. For a three-level factor the corresponding
effect consumes two degrees of freedom (linear plus quadratic).

The total degrees of freedom in a dataset of $n$ observations is
$n - 1$ (one is lost to the grand mean). The model degrees of freedom
is $p - 1$, where $p$ is the number of estimated parameters including
the intercept. The error degrees of freedom is

$$
df_{\text{Error}} = n - p.
$$

For our case study with $n = 20$ observations and $p = 8$ parameters
(intercept, three main effects, three two-way interactions, one
three-way interaction), the error degrees of freedom is

$$
df_{\text{Error}} = 20 - 8 = 12.
$$

These 12 residual degrees of freedom decompose into **lack-of-fit**
and **pure-error** components, as we now show.

## Pure Error and Lack of Fit

When replication is present, the error sum of squares splits further:

$$
\text{SS}_{\text{Error}}
  = \text{SS}_{\text{LOF}} + \text{SS}_{\text{PE}},
$$

where $\text{SS}_{\text{PE}}$ is the **pure error** sum of squares
computed from replicate-to-replicate variation at the same design
point, and $\text{SS}_{\text{LOF}}$ is the **lack of fit** sum of
squares. The associated test statistic is

$$
F_{\text{LOF}}
  = \frac{\text{MS}_{\text{LOF}}}{\text{MS}_{\text{PE}}}
  \sim F_{df_{\text{LOF}},\, df_{\text{PE}}}.
$$

For our design we can enumerate the degrees of freedom directly from
the ANOVA table:

```{r enumerate-df}
anova_tbl <- task$results$anova_results$Yield$anova
anova_tbl

# Total df: n - 1
n_total <- nrow(task$data)
df_total <- n_total - 1

# Model df: sum of effect df (excluding the residual row).
df_model <- sum(anova_tbl$Df[-nrow(anova_tbl)])

# Error df: residual df from the ANOVA table.
df_error <- anova_tbl$Df[nrow(anova_tbl)]

# Pure-error df: 8 factorial points (1 df each from 2 reps) +
#                4 center points (3 df) = 8 + 3 = 11.
df_pe <- 8 * (2 - 1) + (4 - 1)

# Lack-of-fit df: residual df - pure-error df.
df_lof <- df_error - df_pe

data.frame(
  Source = c("Total", "Model", "Residual",
             "  Pure error", "  Lack of fit"),
  Df     = c(df_total, df_model, df_error, df_pe, df_lof)
)
```

The single lack-of-fit degree of freedom is exactly the curvature
degree of freedom --- the only model misspecification in our
simulation is the unmodelled quadratic curvature, so all lack-of-fit
signal is attributable to curvature. Section 9 computes the curvature
$F$-statistic directly.

> **Note.** The package's `DoeAnalyzer` does not currently split the
> residual sum of squares into pure-error and lack-of-fit components
> automatically; the lack-of-fit $p$-value reported by `model_fit` is
> `NA`. The curvature $F$-test in Section 9 closes this gap for
> designs with center points.

# Effect Estimation and Effect Sparsity

## The Effect Formula

For a two-level design coded in the standard $\{-1, +1\}$ convention,
the **effect** of factor $A$ is the difference between the mean
response at the high level and the mean response at the low level:

$$
\text{Effect}_A = \bar{y}_{A+} - \bar{y}_{A-}.
$$

Equivalently, if $c_i \in \{-1, +1\}$ is the coded value of factor
$A$ for run $i$ and $n$ is the number of runs over which the contrast
is computed (here $n_f = 16$ factorial runs, since center points have
$c_i = 0$ and contribute nothing to the contrast),

$$
\text{Effect}_A = \frac{2}{n}\, \sum_{i=1}^{n} c_i\, y_i.
$$

The ordinary-least-squares regression coefficient $\hat\beta_A$ is
half the effect, because the coded variable changes by two units when
moving from $-1$ to $+1$:

$$
\hat\beta_A = \frac{\text{Effect}_A}{2}.
$$

We can verify the equivalence directly on the case-study data:

```{r effect-formula}
# Use only the 16 factorial runs (exclude center points where A=B=C=0).
factorial_only <- task$data[!is.na(task$data$Replication), ]

# Coded values for the factorial points.
A <- (factorial_only$Temperature - 150) / 50
y <- factorial_only$Yield
n_f <- nrow(factorial_only)

# Effect of Temperature using the contrast formula (2/n_f) * sum(c * y).
effect_Temp_contrast <- (2 / n_f) * sum(A * y)
effect_Temp_contrast

# Effect of Temperature using 2 * beta_hat from the fitted lm.
effect_Temp_coef <- 2 * coef(task$results$model)["Temperature"]
effect_Temp_coef
```

Both quantities agree to numerical precision. The same computation
applies to every main effect and every interaction; the contrast
column $c$ for an interaction $AB$ is simply the elementwise product
$A \cdot B$.

## The F-Statistic for an Effect

Each contrast has one degree of freedom, so its mean square equals
its sum of squares:

$$
\text{SS}_{\text{Effect}}
  = \frac{(\text{Contrast})^2}{n}
  = \frac{n}{4}\, (\text{Effect})^2,
\qquad
\text{MS}_{\text{Effect}} = \text{SS}_{\text{Effect}} / 1.
$$

The test statistic for the null hypothesis "the effect is zero" is

$$
F_0 = \frac{\text{MS}_{\text{Effect}}}{\text{MS}_{\text{Error}}}
  \sim F_{1,\, df_{\text{Error}}}.
$$

This is exactly the $F$-statistic reported in the `Mean Sq` and
`F value` columns of `task$results$anova_results$Yield$anova`. An
effect is declared significant at level $\alpha$ when $F_0$ exceeds
$F_{\alpha, 1, df_{\text{Error}}}$, equivalently when the $p$-value
is below $\alpha$.

## The Effect Sparsity Principle

Empirical experience across many fields of industrial experimentation
shows that, in a saturated or near-saturated design, only a small
minority of the estimable effects are truly active. This is the
**effect sparsity** principle, sometimes called the *Pareto principle
of effects* or the *key few and trivial many*.

The principle has two consequences:

1. **It justifies the use of saturated screening designs.** If only
   a few effects are active, a small design is sufficient to identify
   them.
2. **It enables the use of inert effects as a noise estimate.** When
   replication is unavailable, the many inert effects provide a
   pseudo-error estimate (Section 6.5).

## Pareto Chart of Effects

The Pareto chart orders the absolute effects from largest to smallest
and displays them as bars. The chart is the simplest visual
representation of effect sparsity: a sharp drop separates the *vital
few* active effects from the *trivial many* inert effects.

The package's `task$plot(type = "main_effects")` renders a bar chart
of the **regression coefficients** for the main effects only. To
visualise **all** effects (main + two-way + three-way interactions)
as a true Pareto chart, we build one explicitly with `ggplot2`,
multiplying the regression coefficients by $2$ to obtain the effects:

```{r pareto-chart, fig.width = 8, fig.height = 4.5}
# Collect all coefficients from the fitted model and convert to effects.
all_coefs <- coef(task$results$model)
effect_df <- data.frame(
  Term   = names(all_coefs),
  Effect = 2 * as.numeric(all_coefs)   # Effect = 2 * beta_hat
)
effect_df <- effect_df[effect_df$Term != "(Intercept)", ]
effect_df$AbsEffect <- abs(effect_df$Effect)

# Order by absolute effect (descending) for the Pareto chart.
effect_df <- effect_df[order(effect_df$AbsEffect, decreasing = TRUE), ]
effect_df$Term <- factor(effect_df$Term, levels = effect_df$Term)

ggplot(effect_df, aes(x = Term, y = Effect, fill = AbsEffect)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  scale_fill_gradient(low = "#9ecae1", high = "#08519c", guide = "none") +
  labs(
    title = "Pareto Chart of Factorial Effects",
    subtitle = "Chemical yield case study (2^3 with 2 replicates + 4 center points)",
    x = "Effect (ordered by |effect|)",
    y = "Effect estimate (Yield %)"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The chart shows the *vital few* --- `Temperature`, `Time`, `Pressure`,
and the `Temperature:Pressure` interaction --- clearly separated from
the *trivial many* (`Temperature:Time`, `Pressure:Time`,
`Temperature:Pressure:Time`), whose estimates are dominated by noise.
This is the effect-sparsity pattern in graphical form.

## Normal Probability Plot of Effects

Daniel (1959) introduced the use of **normal probability plots of
effects** for unreplicated factorial designs. If all effects were
inert, the ordered contrasts would lie approximately on a straight
line through the origin with slope $\sigma$. Active effects appear as
outliers at the extremes of the plot.

> **Implementation status.** The `task$plot()` API in **iQualityR.doe**
> exposes eight plot types: `design`, `main_effects`, `interaction`,
> `residual`, `full`, `response_surface`, `surface`, and `contour`
> (see the *API Overview* section of the package-level vignettes for
> the full signature, including the `response_name`, `x_var`, and
> `y_var` arguments used by the response-surface types). A built-in
> **normal-probability plot of effects** and **half-normal plot** are
> *not yet implemented* as a dedicated `task$plot()` type, and
> **Lenth's pseudo-standard error** is likewise not yet exposed. The
> plot below is therefore built directly with `ggplot2`, using the
> effect estimates from `task$results$model`. Lenth (1989) proposed a
> robust pseudo-standard error
> $\text{PSE} = 1.5 \cdot \text{median}\{|c_i| : |c_i| < 2.5 s_0\}$
> with $s_0 = 1.5 \cdot \text{median}(|c_i|)$; the reader can compute
> the PSE by hand from `effect_df$Effect` and apply it as a reference
> line.

```{r npp-effects, fig.width = 7, fig.height = 5}
# Sort effects and assign theoretical normal scores.
n_eff <- nrow(effect_df)
effect_df_sorted <- effect_df[order(effect_df$Effect), ]
effect_df_sorted$Rank <- seq_len(n_eff)
effect_df_sorted$p <- (effect_df_sorted$Rank - 0.5) / n_eff
effect_df_sorted$z <- qnorm(effect_df_sorted$p)

# Reference line through the origin; slope estimated from the three
# smallest |effects| (treated as inert, in the spirit of Lenth's method).
inert_idx <- order(abs(effect_df_sorted$Effect))[1:3]
slope <- sd(effect_df_sorted$Effect[inert_idx])

ggplot(effect_df_sorted, aes(x = z, y = Effect)) +
  geom_abline(intercept = 0, slope = slope, color = "gray50",
              linetype = "dashed") +
  geom_point(size = 3, color = "#08519c") +
  geom_text(aes(label = Term), vjust = -0.8, hjust = 0.5, size = 3) +
  labs(
    title = "Normal Probability Plot of Effects",
    subtitle = "Active effects deviate from the straight line",
    x = "Theoretical normal quantile",
    y = "Effect estimate (Yield %)"
  ) +
  theme_minimal(base_size = 12)
```

The four active effects (`Temperature`, `Time`, `Pressure`,
`Temperature:Pressure`) depart visibly from the straight line, while
the three inert effects (`Temperature:Time`, `Pressure:Time`,
`Temperature:Pressure:Time`) lie close to it. In a real
unreplicated-screening analysis the inert effects would be used to
estimate $\sigma$ via Lenth's method --- but, as noted above, that
diagnostic is not yet provided by the package.

# Aliasing in Fractional Designs

## Generators and the Defining Relation

A $2^{k-p}$ fractional factorial design is constructed by:

1. selecting a base $2^{k-p}$ full factorial in $k - p$ *independent*
   factors (the *generators*);
2. equating the remaining $p$ factors to specific products of the
   generators.

Each equation of the form $E = ABC$ gives rise to a **defining word**
$I = ABCE$. The collection of all such words, including their
products, is the **defining relation**. The length of the shortest
word is the **resolution** of the design.

## The Alias Pattern

For every defining word $W$, two effects $E_1$ and $E_2$ are *aliased*
(confounded) if and only if $E_1 E_2 = W$. The alias structure
therefore follows directly from the defining relation.

| Resolution | Main effects aliased with | 2FI aliased with        |
|------------|---------------------------|-------------------------|
| III        | 2FI                       | Main effects and other 2FI |
| IV         | 3FI                       | Other 2FI               |
| V          | 4FI                       | 3FI                     |

A small worked example is the $2^{7-4}$ resolution-III design with
defining relation $I = ABD = ACE = BCF = DEG$. The alias of the main
effect $A$ is

$$
A = A \cdot I = A \cdot ABD = BD = A \cdot ACE = CE = \dots
$$

so $A$ is aliased with $BD$, $CE$, and longer products. Whenever such
an alias is encountered, the engineer uses the effect-sparsity
principle to decide which member of the alias chain is the most
plausible physical explanation.

## A Fractional Design with `screening_design()`

To illustrate the API for fractional designs, we build a six-factor
resolution-IV screening design. The `screening_design()` constructor
is a convenience wrapper around
`factorial_design(design_type = "fractional")` that defaults to
resolution IV.

```{r screening-design}
factors_6 <- lapply(1:6, function(i) {
  list(name = paste0("X", i), type = "continuous", levels = c(-1, 1))
})

screen_task <- screening_design(
  factors    = factors_6,
  resolution = "IV",
  randomize  = TRUE,
  seed       = 2024
)

screen_task$summary()
head(screen_task$results$design_info, 8)
nrow(screen_task$results$design_info)
```

The design has `r nrow(screen_task$results$design_info)` runs. A
textbook $2^{6-2}_{\text{IV}}$ design with 16 runs would have defining
relation $I = ABD = ACE = BCDE$, aliasing each main effect with
three-factor interactions (e.g., $A = BD = CE$) and each two-factor
interaction with one other two-factor interaction (e.g., $AB = CD$,
$AC = BE$, etc.).

> **Implementation status --- fractional generator.** The
> fractional-factorial generator currently shipped with
> **iQualityR.doe** is a *simplified* one: it subsets a full factorial
> by systematic sampling rather than constructing a proper alias
> structure from independent generators. The resulting design is
> adequate for teaching the API and for rough screening, but the
> alias pattern is not guaranteed to match the textbook
> resolution-IV aliasing above. For production work that depends on
> the alias structure, prefer `DoE.base::fac.design()` or
> `FrF2::FrF2()`, then attach the resulting design table to an
> `IqrDoeTask` for analysis and reporting.

## Fold-Over Designs (Not Yet Implemented)

When a screening design identifies an alias that cannot be resolved by
argument, a **fold-over** can be run. The fold-over duplicates the
original design with all signs of one or more factors reversed. This
breaks specific alias chains and lifts the design to a higher
resolution.

> **Implementation status.** Fold-over augmentation is *not yet
> implemented* in **iQualityR.doe**. The math is straightforward:
> for a full fold-over, append a copy of the original design matrix
> with the sign of *every* factor reversed; for a single-factor
> fold-over on factor $A$, append a copy with only the sign of $A$
> reversed. The combined design is then re-analyzed with
> `task$compute()` after attaching the augmented data. The reader
> can implement this by hand with `rbind(design, -design)` (full
> fold-over) or
> `rbind(design, transform(design, A = -A))` (single-factor
> fold-over).

# Design Optimality Criteria

## The Information Matrix

All classical optimality criteria are functions of the **information
matrix** $M = X^\top X$ (or equivalently its inverse, the
**dispersion matrix** $(X^\top X)^{-1}$). A design is **optimal** in
a given sense if it optimises the corresponding criterion over the
design space.

For our $2^3$ case-study design, the information matrix is diagonal
because the design is orthogonal:

```{r information-matrix}
# Rebuild X for the 16 factorial runs (2 replicates of 2^3 = 8 corners).
coded_full <- rbind(coded_design, coded_design)
X_full <- model.matrix(~ Temperature * Pressure * Time, data = coded_full)

# X'X and its inverse.
M <- t(X_full) %*% X_full
M
solve(M)
```

The diagonal structure of $M$ confirms that, for an orthogonal
design, all coefficient estimates are uncorrelated. The off-diagonal
entries of $(X^\top X)^{-1}$ (here, all zero) give the covariances
$\text{Cov}(\hat\beta_i, \hat\beta_j)$.

## D-Optimality

A design is **D-optimal** if it maximises the determinant of the
information matrix,

$$
\max_{X}\; \det(X^\top X).
$$

**Geometric interpretation.** The joint confidence ellipsoid for
$\beta$ has volume proportional to
$\det\bigl((X^\top X)^{-1}\bigr)^{1/2} = \det(X^\top X)^{-1/2}$.
**Maximising** $\det(X^\top X)$ therefore **minimises** the volume of
the joint confidence ellipsoid, i.e., it makes the joint estimate of
$\beta$ as precise as possible in the volume sense. D-optimal designs
are widely used because the criterion is invariant under
reparameterisation and is easy to compute numerically.

For our case-study design the determinant is

```{r d-optimality}
det(M)
```

Adding a replicate doubles every diagonal entry of $M$, which
multiplies the determinant by $2^p$, where $p$ is the number of
parameters. For our $2^3$ design with $p = 8$ parameters, going from
$1$ replicate to $2$ replicates multiplies the determinant by
$2^8 = 256$, shrinking the joint confidence ellipsoid by a factor of
$256^{1/2} = 16$ in volume:

```{r d-optimality-replication}
# Single-replicate information matrix (8 corner runs only).
M1 <- t(X) %*% X
det(M1)
det(M) / det(M1)
2^8
```

## A-, G-, V-, and I-Optimality

Several other criteria are in routine use:

* **A-optimality**:
  $\min_X\, \text{tr}\bigl((X^\top X)^{-1}\bigr)$
  --- minimises the *sum* of the variances of the estimated
  coefficients.
* **G-optimality**:
  $\min_X\, \max_{x \in \mathcal{X}} x^\top (X^\top X)^{-1} x$
  --- minimises the *worst-case* prediction variance over the design
  region.
* **V-optimality**: minimises the *average* prediction variance over
  a finite set of candidate points.
* **I-optimality**: minimises the *average* prediction variance
  integrated over the continuous design region; particularly relevant
  for response-surface work.

The **Kiefer-Wolfowitz equivalence theorem** (1960) states that, in
the approximate-design framework, a design is D-optimal if and only
if it is G-optimal. This equivalence justifies the practical use of
D-optimal computer-generated designs as a proxy for G-optimal
prediction-focused designs.

## Practical Implementation in iQualityR.doe

The `IqrDoePlan` class exposes an `optimality` field (default `"D"`)
so that the user can declare the intended criterion.
**iQualityR.doe** uses regular designs by default because they are
orthogonal and hence simultaneously D-, A-, G-, V-, and I-optimal
within the regular-design family. The `optimality` field is recorded
on the plan for downstream use but is not currently enforced by the
analyzer when irregular or computer-generated designs are requested.

```{r optimality-field}
task$plan$optimality   # "D" by default
task$plan$alpha        # NULL for non-RSM designs
task$plan$resolution   # NULL for full factorial
```

# Center Points and the Curvature Test

## Why Center Points?

A two-level factorial design assumes that the response is locally
linear in every factor. When the response is actually curved, the
linear model is biased and the optimum may be missed. To detect
curvature without committing to a full second-order design, the
engineer appends **center points** at the geometric center of the
design region. In our case study, the four center points sit at
`Temperature = 150`, `Pressure = 3`, `Time = 60` (natural units),
i.e., $A = B = C = 0$ in coded units.

## The Curvature Contrast

Let $\bar y_f$ be the average of the $n_f$ factorial runs and
$\bar y_c$ the average of the $n_c$ center points. The curvature
contrast is

$$
C = \bar y_f - \bar y_c,
$$

with associated sum of squares

$$
\text{SS}_{\text{Curvature}}
  = \frac{n_f\, n_c}{n_f + n_c}\, (\bar y_f - \bar y_c)^2,
$$

and test statistic

$$
F_{\text{Curv}}
  = \frac{\text{SS}_{\text{Curvature}} / 1}{\text{MS}_{\text{PE}}}
  \sim F_{1,\, df_{\text{PE}}}.
$$

Here $\text{MS}_{\text{PE}}$ is the pure-error mean square obtained
from the replicate-to-replicate variation at each design point. With
$n_c = 4$ center points (3 df) and 8 factorial points each replicated
twice (8 df), the pure-error degrees of freedom is
$df_{\text{PE}} = 3 + 8 = 11$.

## Computing the Curvature Test for Our Case Study

We extract the factorial and center-point subsets from `task$data`
and compute the curvature $F$-statistic directly from the formula:

```{r curvature-test}
# Split the data into factorial and center-point subsets.
is_center <- is.na(task$data$Replication)
yf <- task$data$Yield[!is_center]
yc <- task$data$Yield[is_center]

n_f <- length(yf)
n_c <- length(yc)

ybar_f <- mean(yf)
ybar_c <- mean(yc)

# Curvature contrast and sum of squares.
C <- ybar_f - ybar_c
SS_curv <- (n_f * n_c / (n_f + n_c)) * C^2
df_curv <- 1

# Pure-error SS from the 8 factorial points (2 reps each -> 1 df each)
# plus 4 center points (3 df).
factorial_pts <- task$data[!is_center,
                           c("Temperature", "Pressure", "Time", "Yield")]
factorial_pts$Key <- paste(factorial_pts$Temperature,
                           factorial_pts$Pressure,
                           factorial_pts$Time, sep = "|")

factorial_pe <- 0
for (key in unique(factorial_pts$Key)) {
  yy <- factorial_pts$Yield[factorial_pts$Key == key]
  factorial_pe <- factorial_pe + sum((yy - mean(yy))^2)
}
factorial_df_pe <- length(unique(factorial_pts$Key)) * (2 - 1)  # 8 df

# Center pure error: 4 center points -> 3 df.
center_pe <- sum((yc - mean(yc))^2)
center_df_pe <- n_c - 1  # 3 df

SS_pe <- factorial_pe + center_pe
df_pe <- factorial_df_pe + center_df_pe
MS_pe <- SS_pe / df_pe

# Curvature F-statistic.
F_curv <- (SS_curv / df_curv) / MS_pe
p_curv <- 1 - pf(F_curv, df_curv, df_pe)
F_crit <- qf(0.95, df_curv, df_pe)

curvature_results <- data.frame(
  Quantity = c("n_f (factorial runs)", "n_c (center points)",
               "y_bar_f", "y_bar_c",
               "Curvature contrast C = y_bar_f - y_bar_c",
               "SS_Curvature", "df_Curvature",
               "SS_PureError", "df_PureError",
               "MS_PureError",
               "F_Curvature", "p-value",
               "F_crit (0.05, 1, 11)"),
  Value = c(n_f, n_c,
            round(ybar_f, 4), round(ybar_c, 4),
            round(C, 4),
            round(SS_curv, 4), df_curv,
            round(SS_pe, 4), df_pe,
            round(MS_pe, 4),
            round(F_curv, 4),
            signif(p_curv, 4),
            round(F_crit, 4))
)

curvature_results
```

The curvature $F$-statistic is highly significant ($p \ll 0.05$),
confirming that the response surface is curved and that a
first-order-with-interaction model is inadequate. The natural next
step is to augment the design with axial runs to form a Central
Composite Design and fit a full second-order model --- the topic of
the `doe_rsm` vignette.

## Visualising the Curvature

The contrast between $\bar y_f$ and $\bar y_c$ can be visualised
directly:

```{r curvature-plot, fig.width = 7, fig.height = 4}
plot_df <- data.frame(
  Group = c(rep("Factorial points", n_f), rep("Center points", n_c)),
  Yield = c(yf, yc)
)

ggplot(plot_df, aes(x = Group, y = Yield, fill = Group)) +
  geom_boxplot(width = 0.4, outlier.shape = NA) +
  geom_jitter(width = 0.05, size = 2, alpha = 0.7) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 4,
               color = "black", fill = "white") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Curvature: factorial mean vs. center mean",
    subtitle = sprintf("y_bar_f = %.2f,  y_bar_c = %.2f,  F_Curv = %.2f",
                       ybar_f, ybar_c, F_curv),
    x = "",
    y = "Yield (%)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

The center-point distribution sits visibly above the factorial
distribution, exactly as engineered by the simulation.

## ANOVA Partition of the Total Sum of Squares

The full ANOVA partition for our case study, including the curvature
component, is summarised below. The model sums of squares come from
`task$results$anova_results$Yield$anova`; the pure-error and
curvature sums of squares are computed above.

```{r anova-partition, fig.width = 8, fig.height = 4.5}
# Build a data frame describing the ANOVA partition.
anova_rows <- task$results$anova_results$Yield$anova
ss_total <- sum(anova_rows$`Sum Sq`)
ss_model <- sum(anova_rows$`Sum Sq`[-nrow(anova_rows)])
ss_resid <- anova_rows$`Sum Sq`[nrow(anova_rows)]

partition_df <- data.frame(
  Source = c("Model", "Residual (total)",
             "  Curvature (1 df)", "  Pure error (11 df)"),
  SS = c(ss_model, ss_resid, SS_curv, SS_pe),
  df = c(df_model, df_error, df_curv, df_pe)
)
partition_df$MS <- partition_df$SS / partition_df$df
partition_df$Proportion <- partition_df$SS / ss_total

# Bar chart of the SS partition.
partition_plot <- partition_df[partition_df$Source != "Residual (total)", ]
partition_plot$Source <- factor(partition_plot$Source,
                                levels = partition_plot$Source)

ggplot(partition_plot, aes(x = Source, y = SS, fill = Source)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = sprintf("%.1f\n(%.0f%%)", SS, 100 * Proportion)),
            vjust = -0.3, size = 3.5) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "ANOVA Partition of Total Sum of Squares",
    subtitle = sprintf("Total SS = %.2f  (model + curvature + pure error)",
                       ss_total),
    x = "",
    y = "Sum of Squares"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 15, hjust = 1))

partition_df
```

The partition shows that the model effects (mostly `Temperature`,
`Time`, `Pressure`, and their `Temperature:Pressure` interaction)
account for the bulk of the variation, with curvature contributing a
non-trivial share that is highly significant relative to pure error.

# Randomization

## The Three Principles of Experimental Design

The three fundamental principles of experimental design, due to
Fisher (1935), are:

1. **Randomization.** The order of runs must be randomised so that
   the assumption of independent errors is approximately valid.
2. **Replication.** Replicating runs provides an estimate of
   experimental error and increases the precision of effect
   estimates.
3. **Blocking.** Grouping runs into homogeneous blocks reduces the
   error variance by removing known nuisance sources.

Randomization converts systematic, possibly unknown, sources of
variability --- time trends, operator drift, material lot effects ---
into unsystematic noise that satisfies the independence assumption
of the linear model. Without randomization, such sources can produce
spurious significant effects.

## Reproducible Randomization in iQualityR.doe

The **iQualityR.doe** package randomizes by default
(`randomize = TRUE`) and exposes the `seed` argument so that the
random order is reproducible. The randomized `RunOrder` column is
stored in `task$results$design_info`:

```{r randomized-run-order}
# Show the first 10 rows of the randomized run order.
head(task$results$design_info[, c("RunOrder", "Temperature", "Pressure",
                                  "Time", "Replication")], 10)

# Compare with the *standard* (unrandomized) order: rebuild with
# randomize = FALSE and the same seed.
task_std <- factorial_design(
  factors       = factors,
  design_type   = "factorial",
  response_vars = "Yield",
  replication   = 2,
  center_points = 4,
  randomize     = FALSE,
  seed          = 42
)

head(task_std$results$design_info[, c("RunOrder", "Temperature", "Pressure",
                                      "Time", "Replication")], 10)
```

The first run of the randomized design differs from the first run of
the unrandomized design. Following the randomized run order on the
shop floor is essential --- it is the physical act that protects the
validity of the $F$-tests reported by `task$compute()`.

## Restricted Randomization

In some situations fully unrestricted randomization is impractical
(for example, when changing a factor level is expensive). In such
cases the engineer uses **restricted randomization** by blocking and
randomizing within blocks. The principle is the same: randomize as
much as the experimental constraints allow, and block the rest.
Blocking support in **iQualityR.doe** is currently limited to the
`blocking` field on `IqrDoePlan`; full blocked-design generation is
not yet implemented, and the engineer is expected to combine multiple
tasks by hand when blocks are needed.

# Visualising Results with `task$plot()`

The theory in the previous sections was illustrated with hand-built
`ggplot2` charts so that every formula could be inspected directly.
In day-to-day practice, however, the engineer should reach for the
package's built-in plotting API, which produces consistent, themed
diagnostic figures from any fitted `IqrDoeTask`. This section is a
short, practical reference.

## API Overview

The `IqrDoeTask` class exposes a single `plot()` method whose
signature is:

```r
task$plot(type = "design", theme = NULL, ...)
```

| Argument        | Purpose                                                                 |
|-----------------|-------------------------------------------------------------------------|
| `type`          | One of the eight plot types listed below.                              |
| `theme`         | Optional `IqrTheme` object (or theme name) to override the task theme. |
| `...`           | Forwarded to the plotter. For response-surface types this includes `response_name`, `x_var`, and `y_var`. |

The eight supported `type` values are:

| `type`               | Returns                 | Requires 2nd-order model | Typical use                                       |
|----------------------|-------------------------|---------------------------|---------------------------------------------------|
| `"design"`           | `ggplot`                | No                        | Inspect the run layout (run order tile view).     |
| `"main_effects"`     | `ggplot` / `patchwork`  | No                        | Marginal effect of each factor.                   |
| `"interaction"`      | `ggplot` / `patchwork`  | No                        | Two-factor interaction panels.                    |
| `"residual"`         | `patchwork`             | No                        | Standard residual diagnostics (QQ, fitted, scale). |
| `"full"`             | `patchwork`             | No                        | Combined main_effects + interaction + residual.   |
| `"response_surface"` | `list` (see below)      | **Yes**                   | Full RSM panel (surface + contour + 3D + plotly).  |
| `"surface"`          | `ggplot`                | **Yes**                   | 2D tile + contour surface only.                   |
| `"contour"`          | `ggplot`                | **Yes**                   | Filled contour only.                              |

For `"response_surface"`, the returned list contains `surface_plot`,
`contour_plot`, `combined_plot`, and `interactive_plot` (the last is
`NULL` when the optional `plotly` package is not installed). The
`response_name` argument selects which response to plot (defaults to
the first); `x_var` and `y_var` select the two factors shown on the
axes (default to the first two numeric factors).

> **Important: invisible returns.** Like `plot()` on base R objects,
> `task$plot()` returns its value *invisibly*. In an interactive R
> session the figure auto-prints, but inside `rmarkdown` / `for`
> loops / functions you must call `print()` explicitly:

```r
p <- task$plot(type = "main_effects")
print(p)            # renders the figure inside an R Markdown chunk
```

## Demonstration on the Yield Case Study

The `task` built earlier (Section *Building the Design with
`factorial_design()`*) already has its response data attached and
fitted, so we can directly call `task$plot()`.

### Design layout

```{r plot-design, fig.width = 7, fig.height = 3.2, fig.cap = "Built-in design layout produced by task$plot(type = 'design')."}
print(task$plot(type = "design"))
```

The tile view makes the run order explicit: each tile is one
experimental run, coloured by its position in the randomised
sequence. This is the figure to share with the lab technician before
the trials start.

### Main effects

```{r plot-main-effects, fig.width = 7, fig.height = 3.2, fig.cap = "Main effects of the three factors on Yield, from task$plot(type = 'main_effects')."}
print(task$plot(type = "main_effects"))
```

**Decision.** `Temperature` has the steepest main-effect slope, so
it is the dominant lever on `Yield`. `Time` has a moderate positive
effect; `Pressure` is nearly flat. If the budget only allows one
factor to be re-tuned, it should be `Temperature`.

### Interaction

```{r plot-interaction, fig.width = 7, fig.height = 3.5, fig.cap = "Two-factor interaction panels from task$plot(type = 'interaction')."}
print(task$plot(type = "interaction"))
```

**Decision.** The near-parallel lines for `Temperature:Time` confirm
the absence of a strong two-factor interaction --- the additive model
is adequate. The slight non-parallelism of `Temperature:Pressure`
near the high-`Pressure` corner motivates the confirmation run
discussed below.

### Residual diagnostics

```{r plot-residual, fig.width = 7, fig.height = 4.5, fig.cap = "Residual diagnostics from task$plot(type = 'residual')."}
print(task$plot(type = "residual"))
```

**Decision.** The QQ plot is linear in the middle range, the
residuals-vs-fitted plot shows no funnel shape, and the scale-location
plot is flat. The model assumptions are satisfied; the inference on
the ANOVA table (Section *Reading the ANOVA Table*) can be trusted.

### Combined `full` panel

The `"full"` type composes main-effects, interaction, and residual
panels into a single `patchwork` object. It is the quickest
one-liner for a slide or a report.

```{r plot-full, fig.width = 9, fig.height = 8, fig.cap = "Combined diagnostic panel from task$plot(type = 'full')."}
print(task$plot(type = "full"))
```

### Customising the returned `ggplot` object

Because `task$plot()` returns a standard `ggplot` (or `patchwork`)
object, you can post-process it with additional `ggplot2` layers. For
example, to add a horizontal spec limit to the main-effects plot:

```{r plot-customise, fig.width = 7, fig.height = 3.2, fig.cap = "Customised main-effects plot: a user-added horizontal line marks the target Yield of 75%."}
library(ggplot2)

p_main <- task$plot(type = "main_effects")
p_main +
  geom_hline(yintercept = 75, linetype = "dashed", color = "firebrick") +
  labs(subtitle = "Red line: target Yield = 75%")
```

### Combining multiple `task$plot()` outputs with `patchwork`

Several `task$plot()` outputs can be composed into a custom layout
with `patchwork`:

```{r plot-patchwork, fig.width = 9, fig.height = 6, fig.cap = "Custom patchwork composition: main effects (left) and interaction (right) combined with patchwork::wrap_elements()."}
library(patchwork)

p1 <- task$plot(type = "main_effects")
p2 <- task$plot(type = "interaction")

wrap_elements(p1) + wrap_elements(p2) +
  plot_layout(ncol = 2, widths = c(1, 1.2)) +
  plot_annotation(
    title = "Yield case study: effects and interactions",
    tag_levels = "A"
  )
```

### Exporting a report

In addition to `plot()`, `IqrDoeTask$report()` exports the full
analysis (design table, ANOVA, coefficients, model summary) to an
Excel or HTML file:

```r
# Not run here to keep the vignette self-contained.
task$report(format = "excel", path = tempfile(fileext = ".xlsx"))
task$report(format = "html",  path = tempfile(fileext = ".html"))
```

`$report()` returns the task invisibly, so it can be chained.

# Summary

The statistical foundations of design of experiments rest on three
pillars:

1. The **linear model** $y = X\beta + \epsilon$, whose estimation and
   inference are well understood under the classical assumptions.
2. **Orthogonality**, which makes the effect estimates uncorrelated
   and the information matrix $X^\top X$ diagonal.
3. **Principled experimentation** through randomization, replication,
   and blocking, which together guarantee that the assumptions of
   the linear model are approximately met in practice.

The four core formulas that recur throughout DOE are:

$$
\text{Effect}_A = \bar y_{A+} - \bar y_{A-}
                = \frac{2}{n}\sum_i c_i y_i,
\qquad
F_0 = \frac{\text{MS}_{\text{Effect}}}{\text{MS}_{\text{Error}}},
$$

$$
\max_X \det(X^\top X) \;\;\text{(D-optimality)},
\qquad
F_{\text{Curv}}
  = \frac{(n_f n_c / (n_f + n_c))\,(\bar y_f - \bar y_c)^2}
         {\text{MS}_{\text{PE}}}.
$$

The **iQualityR.doe** package implements these foundations through a
small set of constructors that share a common `IqrDoeTask` interface,
so that the engineer can move from theory to reproducible practice in
a single R session. The features that are *not yet implemented* ---
Lenth's method, the half-normal plot, fold-over augmentation, and
proper fractional-factorial alias-structure generation --- are
clearly flagged in the corresponding sections and replaced by their
mathematical treatment so that the reader can apply them by hand or
via complementary R packages.

# References

1. Fisher, R. A. (1935). *The Design of Experiments*. Oliver and
   Boyd.
2. Montgomery, D. C. (2019). *Design and Analysis of Experiments*
   (10th ed.). Wiley.
3. Box, G. E. P., Hunter, W. G., & Hunter, J. S. (2005).
   *Statistics for Experimenters* (2nd ed.). Wiley.
4. Wu, C. F. J., & Hamada, M. S. (2009). *Experiments: Planning,
   Analysis, and Optimization* (2nd ed.). Wiley.
5. Kiefer, J., & Wolfowitz, J. (1960). The equivalence of two
   extremum problems. *Canadian Journal of Mathematics*, 12,
   363-366.
6. Daniel, C. (1959). Use of half-normal plots in interpreting
   factorial two-level experiments. *Technometrics*, 1(4), 311-341.
7. Lenth, R. V. (1989). Quick and easy analysis of unreplicated
   factorials. *Technometrics*, 31(4), 469-473.
8. Myers, R. H., Montgomery, D. C., & Anderson-Cook, C. M. (2016).
   *Response Surface Methodology: Process and Product Optimization
   Using Designed Experiments* (4th ed.). Wiley.

# sessionInfo

```{r sessioninfo}
sessionInfo()
```
