---
title: "Modern AI-Augmented Design of Experiments: A Bayesian Optimization Case Study"
author: "iQualityR Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Modern AI-Augmented Design of Experiments: A Bayesian Optimization Case Study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 8,
  fig.height = 5,
  fig.align  = "center",
  warning    = FALSE,
  message    = FALSE
)
```

```{r setup}
library(iQualityR.doe)
library(ggplot2)
library(patchwork)
set.seed(2025)
```

# Introduction: From Static Designs to Learning Systems

Classical design of experiments (DOE) --- factorial designs, response
surface methodology (RSM), and Taguchi robust designs --- was conceived
in an era when experiments were cheap, repeatability was the primary
concern, and the number of factors was modest. Modern industrial and
scientific settings frequently violate these assumptions:

| Challenge | Classical DOE | Modern DOE |
|-----------|---------------|------------|
| **Expensive trials** | One-shot, fixed design | Sequential, adaptive designs |
| **Nonlinear / multimodal responses** | Low-order polynomial surrogate | Gaussian process surrogates |
| **Multiple data sources** | Cannot reuse simulation data | Multi-fidelity modeling |
| **Unknown response shape** | Best with smooth, unimodal surfaces | Bayesian optimization (BO) |

This vignette demonstrates how **iQualityR.doe** addresses these
challenges through a complete engineering case study. We will:

1. Frame an expensive chemical-process optimization problem with a
   multimodal response.
2. Ask `AutoDOECopilot` to recommend a design strategy.
3. Fit a Gaussian process (GP) surrogate with `BayesianOptimizer` and
   visualise its posterior mean and 95% credible band.
4. Use the Expected Improvement (EI) acquisition function and
   `$suggest_next()` to recommend the next trial.
5. Run a short sequential BO loop and plot the convergence.
6. Leverage biased historical simulation data with
   `MultiFidelityOptimizer` to correct predictions.
7. Compare briefly with classical RSM and confirm the optimum with a
   confirmation trial.

# API Overview

This vignette uses three modern optimization classes from
**iQualityR.doe**, plus the `IqrDoeTask` life-cycle that powers the
underlying experimental designs. The typical `IqrDoeTask` workflow is:

```r
task <- rsm_design(...)         # or factorial_design / screening_design
task$data <- df                  # attach the observed responses
task$compute()                  # fit ANOVA + second-order lm models
task$summary()                  # print the analysis
task$plot(type = "...")         # generate figures (8 types available)
task$report(format = "excel", path = "...")  # export a report
```

## The `plot()` method

```r
task$plot(type = "design", theme = NULL, ...)
```

| Argument        | Purpose                                                                 |
|-----------------|-------------------------------------------------------------------------|
| `type`          | One of the eight plot types listed below.                              |
| `theme`         | Optional `IqrTheme` object (or theme name) to override the task theme. |
| `...`           | Forwarded to the plotter. For response-surface types this includes `response_name`, `x_var`, and `y_var`. |

The eight supported `type` values are:

| `type`               | Returns                 | Requires 2nd-order model | Typical use                                       |
|----------------------|-------------------------|---------------------------|---------------------------------------------------|
| `"design"`           | `ggplot`                | No                        | Inspect the run layout (run order tile view).     |
| `"main_effects"`     | `ggplot` / `patchwork`  | No                        | Marginal effect of each factor.                   |
| `"interaction"`      | `ggplot` / `patchwork`  | No                        | Two-factor interaction panels.                    |
| `"residual"`         | `patchwork`             | No                        | Standard residual diagnostics (QQ, fitted, scale). |
| `"full"`             | `patchwork`             | No                        | Combined main_effects + interaction + residual.   |
| `"response_surface"` | `list` (see below)      | **Yes**                   | Full RSM panel (surface + contour + 3D + plotly).  |
| `"surface"`          | `ggplot`                | **Yes**                   | 2D tile + contour surface only.                   |
| `"contour"`          | `ggplot`                | **Yes**                   | Filled contour only.                              |

For `type = "response_surface"` the returned list contains
`surface_plot`, `contour_plot`, `combined_plot`, and (when **plotly**
is installed) `interactive_plot`.

> **Important: invisible returns.** Like `plot()` on base R objects,
> `task$plot()` returns its value *invisibly*. In an interactive R
> session the figure auto-prints, but inside `rmarkdown` / `for`
> loops / functions you must call `print()` explicitly:

```r
p <- task$plot(type = "main_effects")
print(p)            # renders the figure inside an R Markdown chunk
```

## The `BayesianOptimizer` class

`BayesianOptimizer` wraps a Gaussian-process surrogate (via
**kernlab** when available, with an `lm` polynomial fallback) and an
Expected-Improvement acquisition function. Public methods (verified
against `R/BayesianDOE.R`):

| Method                          | Purpose                                                                       |
|---------------------------------|-------------------------------------------------------------------------------|
| `$new(kernel = "rbf")`          | Constructor. `kernel` is `"rbf"` (default) or `"matern"`.                     |
| `$fit(X, y)`                    | Fit the GP surrogate to an `n x p` matrix `X` and numeric response `y`.       |
| `$predict(X_new)`               | Predict at new points. Returns `list(predict, se.fit)`.                       |
| `$expected_improvement(X_new, y_best)` | Expected-Improvement acquisition function for minimisation.           |
| `$suggest_next(bounds, n_starts = 20, seed = NULL)` | Multi-start L-BFGS-B maximiser of EI. Returns `list(point, ei)`. |
| `$run_sequential(initial_design, response_function, n_iterations = 20, seed = NULL)` | Full sequential BO loop. Returns `list(history, best_y, best_x, ei_trace)`. |

> There is **no** `$optimize()` method on this class; it was renamed
> to `$suggest_next()` to avoid shadowing `base::optimize()`.

## The `MultiFidelityOptimizer` class

`MultiFidelityOptimizer` is a lightweight, model-agnostic
implementation of Kennedy & O'Hagan (2000) style bias correction.
Public methods (verified against `R/MultiFidelityOptimizer.R`):

| Method                          | Purpose                                            |
|---------------------------------|----------------------------------------------------|
| `$new()`                        | Constructor (no arguments).                        |
| `$evaluate_auxiliary_value(historical_data, response_col, current_data = NULL)` | Summarise historical data; estimate bias, correlation, and recommended minimum number of new high-fidelity runs. |
| `$bias_corrected_prediction(historical_model, current_data, design_to_predict, method = "additive")` | Bias-correct a low-fidelity prediction using a historical **model object** (e.g. `lm`), **not** a data frame. Returns `list(prediction, bias_correction, method, residual_sd)`. |
| `$augment_historical_data(historical_data, n_synthetic = 0, noise_factor = 0.05, seed = NULL)` | Bootstrap + Gaussian-perturb the historical data to produce synthetic low-fidelity rows. |

> There are **no** `$fit()` or `$predict()` methods on this class.
> Bias correction takes a *fitted model object* (e.g. an `lm` trained
> on the historical data) via the `historical_model` argument, not a
> data frame.

## The `AutoDOECopilot` class

`AutoDOECopilot` is a rule-based expert system that recommends a DOE
strategy from the problem context. Public methods (verified against
`R/AutoDOECopilot.R`):

| Method                          | Purpose                                            |
|---------------------------------|----------------------------------------------------|
| `$new()`                        | Constructor (no arguments).                        |
| `$recommend(n_factors, goal = "optimization", budget = "medium", industry = "manufacturing", has_historical_data = FALSE)` | Return a recommended strategy list. |

`$recommend()` returns a list with three elements:

| Element       | Type           | Contents                                                                                          |
|---------------|----------------|---------------------------------------------------------------------------------------------------|
| `input`       | list           | The provided arguments: `n_factors`, `goal`, `budget`, `industry`.                               |
| `strategies`  | list of lists  | Each entry has `name`, `design_type`, `runs`, `pros`, `cons`, `applicable_scenario`; multi-fidelity and service-industry notes also carry `description` and `priority`. |
| `next_step`   | character      | A human-readable suggested next action (e.g. `"Run experiment -> ANOVA analysis -> Identify significant factors"`). |

# Case Study: Catalytic Reaction Yield Optimization

## Problem Statement

A chemical engineering team is developing a heterogeneous catalytic
process for converting a feedstock into a high-value active ingredient.
The reaction is run in a continuously stirred reactor. Each
experimental run consumes one full day of operator time and uses
expensive catalyst, so the team can afford at most **15--20 trials
total** across the project.

The objective is to **maximise the reaction yield (%)**. Two factors
are believed to drive the conversion:

| Factor | Symbol | Range | Unit |
|:-------|:-------|:------|:-----|
| Reactor temperature | `Temperature` | 120 -- 180 | $^\circ$C |
| Catalyst loading    | `Catalyst`    | 0.5 -- 1.5  | % w/w |

Preliminary chemistry suggests the response surface is **multimodal**:
a high-yield regime at moderate-to-high temperature with high catalyst
loading, and a secondary lower-yield regime at lower temperature with
lower catalyst loading. Classical second-order RSM would risk getting
trapped in the secondary basin if the initial design happened to land
there.

## The True (Unknown) Response Surface

In a real project the response surface is unknown. For reproducibility
of the vignette we expose the digital-twin function used to evaluate
"experimental" runs. The function has two local maxima --- a global
peak near (`T = 165`, `C = 1.15`) with yield $\approx 91.5\%$, and a
secondary peak near (`T = 138`, `C = 0.75`) with yield $\approx 84\%$.

```{r define-response}
# Digital-twin yield function (would be a physical experiment in practice).
# Returns yield (%) -- to be MAXIMISED.
true_yield <- function(Temperature, Catalyst) {
  t <- (Temperature - 150) / 30   # scaled temperature
  c <- (Catalyst - 1.0)   / 0.5   # scaled catalyst loading
  80 + 12 * exp(-((t - 0.5)^2 + (c - 0.3)^2) / 0.30) +
       5 * exp(-((t + 0.4)^2 + (c + 0.5)^2) / 0.40) -
       2 * (t^2 + c^2)
}

# A dense grid used throughout for visualisation.
grid_2d <- expand.grid(
  Temperature = seq(120, 180, length.out = 60),
  Catalyst    = seq(0.5,  1.5,  length.out = 60)
)
grid_2d$Yield <- true_yield(grid_2d$Temperature, grid_2d$Catalyst)

# Plot the (unknown) true yield surface.
ggplot(grid_2d, aes(Temperature, Catalyst)) +
  geom_raster(aes(fill = Yield)) +
  geom_contour(aes(z = Yield), color = "white", alpha = 0.4, binwidth = 2) +
  scale_fill_viridis_c() +
  labs(
    title = "True (unknown) yield surface",
    subtitle = "Multimodal: global peak ~91% near (165, 1.15); secondary peak ~84%",
    x = "Temperature (deg C)", y = "Catalyst loading (% w/w)"
  ) +
  theme_minimal()
```

# Step 1 --- Strategy Recommendation with AutoDOECopilot

Before running any experiment, the team consults `AutoDOECopilot`, a
rule-based expert system that recommends a design strategy from the
number of factors, the optimisation goal, the budget, and the industry
context. The team indicates that historical simulation data is
available, which triggers a multi-fidelity recommendation.

```{r copilot-recommend}
copilot <- AutoDOECopilot$new()

recommendation <- copilot$recommend(
  n_factors           = 2,
  goal                = "optimization",
  budget              = "medium",
  industry            = "manufacturing",
  has_historical_data = TRUE
)

# Print the recommended strategies.
for (i in seq_along(recommendation$strategies)) {
  s <- recommendation$strategies[[i]]
  cat(sprintf("Strategy %d: %s\n", i, s$name))
  if (!is.null(s$design_type)) cat(sprintf("  design_type: %s\n", s$design_type))
  if (!is.null(s$runs))        cat(sprintf("  runs: %s\n",     paste(s$runs, collapse = ", ")))
  if (!is.null(s$pros))        cat(sprintf("  pros: %s\n",     s$pros))
  if (!is.null(s$priority))    cat(sprintf("  priority: %s\n", s$priority))
  if (!is.null(s$description)) cat(sprintf("  description: %s\n", s$description))
  cat("\n")
}
cat("Suggested next step:", recommendation$next_step, "\n")
```

The copilot proposes a small factorial design and, because historical
data is available, a **Multi-Fidelity Boost** that is expected to cut
the number of physical runs by roughly 40%. Rather than commit to a
fixed factorial design, the team adopts the modern sequential BO
philosophy: a small space-filling initial design followed by
acquisition-driven trials.

# Step 2 --- Initial Design and GP Surrogate

## Initial Space-Filling Design

With only two factors and an expensive response, the team runs a small
6-point space-filling design (an LHS-like layout) to seed the
surrogate.

```{r initial-design}
# A reproducible 6-point space-filling initial design.
initial_design <- data.frame(
  Temperature = c(125, 135, 145, 155, 165, 175),
  Catalyst    = c(1.40, 0.70, 1.20, 0.60, 1.00, 1.30)
)
# Evaluate the (in practice unknown) yield.
initial_design$Yield <- true_yield(
  initial_design$Temperature, initial_design$Catalyst
)
initial_design
```

The initial best yield is already informative but well below the global
peak. The team will use Bayesian optimization to climb toward it.

## Fitting the Gaussian Process Surrogate

The `BayesianOptimizer` class wraps a Gaussian-process surrogate fit
through `kernlab::gausspr()` and exposes the GP posterior through
`fit()`, `predict()`, and the EI acquisition function.

```{r fit-gp}
bo <- BayesianOptimizer$new(kernel = "rbf")

# BO minimises by convention, so we pass -Yield as the response.
X_init <- as.matrix(initial_design[, c("Temperature", "Catalyst")])
y_init <- -initial_design$Yield

bo$fit(X_init, y_init)

# Quick sanity check: posterior mean at the training points.
bo$predict(X_init)
```

**Honest note on kernel support.** The `BayesianOptimizer` constructor
accepts `kernel = "rbf"` (default) or `"matern"`. In the current
implementation, the main GP fit always uses the RBF kernel via
`kernlab::rbfdot()`, while the `"matern"` choice routes the
leave-one-out residual helper to `kernlab::materndot()` --- a function
that is not exported by recent versions of **kernlab**. To keep the
vignette reproducible across `kernlab` versions we therefore use
`kernel = "rbf"` throughout. The RBF kernel is also a perfectly
reasonable default for smooth chemical-process responses.

# Step 3 --- Visualising the GP Posterior

The GP provides both a posterior mean (the surrogate prediction) and a
posterior standard deviation (the epistemic uncertainty). We visualise
both over the full factor space.

## Posterior Mean Surface

```{r gp-posterior-2d}
# Predict over the dense grid.
pred_grid <- bo$predict(grid_2d[, c("Temperature", "Catalyst")])
grid_2d$mu      <- pred_grid$predict
grid_2d$se      <- pred_grid$se.fit
grid_2d$yield_mu <- -grid_2d$mu  # convert back to yield

ggplot(grid_2d, aes(Temperature, Catalyst)) +
  geom_raster(aes(fill = yield_mu)) +
  geom_contour(aes(z = yield_mu), color = "white", alpha = 0.4, binwidth = 2) +
  geom_point(data = initial_design, aes(Temperature, Catalyst),
             color = "white", size = 2.5, shape = 21, stroke = 1.2) +
  scale_fill_viridis_c() +
  labs(
    title = "GP posterior mean of yield",
    subtitle = "White circles: initial design points",
    x = "Temperature (deg C)", y = "Catalyst loading (% w/w)",
    fill = "Yield (%)"
  ) +
  theme_minimal()
```

## Posterior Uncertainty Surface

The predictive standard deviation reveals where the GP is least
informed --- typically far from the design points.

```{r gp-posterior-se}
ggplot(grid_2d, aes(Temperature, Catalyst)) +
  geom_raster(aes(fill = se)) +
  geom_point(data = initial_design, aes(Temperature, Catalyst),
             color = "white", size = 2.5, shape = 21, stroke = 1.2) +
  scale_fill_viridis_c(option = "magma", direction = -1) +
  labs(
    title = "GP posterior standard deviation",
    subtitle = "Uncertainty is largest far from the design points",
    x = "Temperature (deg C)", y = "Catalyst loading (% w/w)",
    fill = "sd"
  ) +
  theme_minimal()
```

## 1-D Slice with 95% Credible Band

Slicing along `Catalyst = 1.10` (near the suspected optimum) makes the
mean $\pm 1.96 \cdot \sigma$ band easy to read.

```{r gp-posterior-1d}
slice_df <- data.frame(Temperature = seq(120, 180, length.out = 120))
slice_df$Catalyst <- 1.10

pred_slice <- bo$predict(slice_df)
slice_df$mu       <- pred_slice$predict
slice_df$se       <- pred_slice$se.fit
slice_df$yield_mu <- -slice_df$mu
slice_df$lower    <- -slice_df$mu - 1.96 * slice_df$se
slice_df$upper    <- -slice_df$mu + 1.96 * slice_df$se
slice_df$true     <- true_yield(slice_df$Temperature, slice_df$Catalyst)

ggplot(slice_df, aes(Temperature)) +
  geom_ribbon(aes(ymin = lower, ymax = upper),
              fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = yield_mu), color = "steelblue", linewidth = 1) +
  geom_line(aes(y = true), color = "firebrick", linetype = "dashed", linewidth = 0.8) +
  geom_point(data = initial_design, aes(Temperature, Yield),
             color = "black", size = 2) +
  labs(
    title = "GP posterior with 95% credible band",
    subtitle = "Slice at Catalyst = 1.10; dashed red = true (unknown) yield",
    x = "Temperature (deg C)", y = "Yield (%)",
    caption = "Blue ribbon: GP mean +/- 1.96 * se"
  ) +
  theme_minimal()
```

The GP credible band covers the true yield along this slice, and the
band widens between design points --- exactly the behaviour we need for
an acquisition function to balance exploration and exploitation.

# Step 4 --- Expected Improvement Acquisition

## EI over the Factor Space

For minimisation of $-Yield$, the Expected Improvement at a candidate
$x$ with posterior mean $\mu(x)$ and standard deviation $\sigma(x)$
against the current best $y_{\min}$ is

$$
\mathrm{EI}(x) = (y_{\min} - \mu(x))\,\Phi(z) + \sigma(x)\,\phi(z),
\qquad z = \frac{y_{\min} - \mu(x)}{\sigma(x)}.
$$

The `BayesianOptimizer` exposes this through `$expected_improvement()`.

```{r ei-heatmap}
# Compute EI over the dense grid (current best = smallest -Yield so far).
y_best <- min(y_init)
ei_grid <- bo$expected_improvement(
  X_new = grid_2d[, c("Temperature", "Catalyst")],
  y_best = y_best
)
grid_2d$EI <- pmax(ei_grid, 0)

ggplot(grid_2d, aes(Temperature, Catalyst)) +
  geom_raster(aes(fill = EI)) +
  geom_point(data = initial_design, aes(Temperature, Catalyst),
             color = "white", size = 2.5, shape = 21, stroke = 1.2) +
  scale_fill_viridis_c(option = "plasma") +
  labs(
    title = "Expected Improvement acquisition surface",
    subtitle = sprintf("Current best yield = %.2f%%", -y_best),
    x = "Temperature (deg C)", y = "Catalyst loading (% w/w)",
    fill = "EI"
  ) +
  theme_minimal()
```

The EI surface is concentrated in the unexplored high-temperature,
high-catalyst region where the GP expects the largest yield gains.

## Recommending the Next Trial with `$suggest_next()`

`$suggest_next()` maximises EI by multi-start L-BFGS-B inside the
supplied `bounds`. We use a small `n_starts` to keep the vignette fast;
in production 20+ starts is recommended.

```{r suggest-next}
bounds <- list(lower = c(120, 0.5), upper = c(180, 1.5))

next_rec <- bo$suggest_next(bounds, n_starts = 4, seed = 42)
next_rec

# Overlay the recommended next trial on the EI surface.
next_df <- as.data.frame(t(next_rec$point))
names(next_df) <- c("Temperature", "Catalyst")

ggplot(grid_2d, aes(Temperature, Catalyst)) +
  geom_raster(aes(fill = EI)) +
  geom_point(data = initial_design, aes(Temperature, Catalyst),
             color = "white", size = 2.5, shape = 21, stroke = 1.2) +
  geom_point(data = next_df, aes(Temperature, Catalyst),
             color = "red", size = 4, shape = 4, stroke = 2) +
  scale_fill_viridis_c(option = "plasma") +
  labs(
    title = "Next trial recommended by maximising EI",
    subtitle = sprintf("Next: T = %.1f, C = %.3f (EI = %.4f)",
                       next_df$Temperature, next_df$Catalyst, next_rec$ei),
    x = "Temperature (deg C)", y = "Catalyst loading (% w/w)"
  ) +
  theme_minimal()
```

The red cross marks the next trial, sitting squarely inside the
highest-EI region.

# Step 5 --- Sequential Bayesian Optimization Loop

We now run a short sequential loop with the built-in
`$run_sequential()` method. It seeds the surrogate with the initial
design, then for each iteration: refits the GP, calls `$suggest_next()`
(multi-start L-BFGS-B on EI) to propose the next trial, evaluates the
response function, and appends the result. We pass `n_iterations = 5`
to keep the vignette budget realistic. The method returns a `history`
data frame, the best value found, the best parameter settings, and an
EI trace.

> **Why not hand-code the loop?** A manual `for` loop calling
> `$fit()` / `$expected_improvement()` / `$suggest_next()` in sequence
> works, but it is easy to forget to refit, to mis-handle the RNG
> stream, or to skip the EI trace. `$run_sequential()` packages all of
> that into one reproducible call. The manual alternative is shown
> further below for reference.

```{r sequential-loop}
# BO minimises by convention, so the response function returns -Yield.
# $run_sequential() expects a function that accepts a single-row data
# frame and returns a numeric scalar.
neg_yield <- function(df) {
  true_yield(df$Temperature, df$Catalyst)
}

bo_result <- bo$run_sequential(
  initial_design    = initial_design[, c("Temperature", "Catalyst")],
  response_function = neg_yield,
  n_iterations      = 5,
  seed              = 42
)

# Reconstruct the trace table expected by the convergence plot.
# history holds the initial design rows first, then the sequential trials,
# with a "response" column containing -Yield.
hist_df <- bo_result$history
n_init  <- nrow(initial_design)
trace <- data.frame(
  iter        = seq_len(nrow(hist_df)),
  phase       = c(rep("initial", n_init),
                  rep("sequential", nrow(hist_df) - n_init)),
  Temperature = hist_df$Temperature,
  Catalyst    = hist_df$Catalyst,
  yield       = -hist_df$response,            # convert -Yield back to Yield
  best_yield  = cummax(-hist_df$response)
)

# Reconstruct X_all / y_all for the RSM comparison in Step 8.
X_all <- as.matrix(hist_df[, c("Temperature", "Catalyst")])
y_all <- hist_df$response                     # already -Yield

trace
cat(sprintf("\nFinal best yield: %.3f%% at T = %.1f, C = %.3f\n",
            trace$best_yield[nrow(trace)],
            tail(trace$Temperature, 1),
            tail(trace$Catalyst, 1)))
```

The sequential loop climbs steadily from the initial best of
`r sprintf("%.2f%%", max(initial_design$Yield))` to a final best of
about `r sprintf("%.2f%%", tail(trace$best_yield, 1))`, with only five
additional trials.

**Decision.** `$run_sequential()` reaches the high-yield basin in five
iterations on top of the six-point initial design, demonstrating the
sample efficiency of GP + EI. In a production setting we would run
until the EI trace drops below a threshold (i.e. the surrogate no
longer expects meaningful gains), then confirm the optimum with a
physical trial (Step 9). When the budget allows more iterations,
simply increase `n_iterations`; the RNG stream is anchored by `seed`
so the result is reproducible.

## Manual alternative (for reference)

For completeness, the equivalent hand-coded loop (the pattern that
`$run_sequential()` internalises) would look like this. We do **not**
evaluate it here because `$run_sequential()` above already produces
`trace`, `X_all`, and `y_all` for the downstream chunks.

```r
# NOT RUN --- reference only.
bounds  <- list(lower = c(120, 0.5), upper = c(180, 1.5))
X_all   <- X_init
y_all   <- y_init
for (iter in 1:5) {
  bo$fit(X_all, y_all)
  rec      <- bo$suggest_next(bounds, n_starts = 20)
  next_x   <- as.matrix(rec$point)
  next_y   <- -true_yield(next_x[1], next_x[2])  # minimise -Yield
  X_all    <- rbind(X_all, next_x)
  y_all    <- c(y_all, next_y)
}
```

# Step 6 --- Convergence Analysis

```{r convergence}
ggplot(trace, aes(iter, best_yield)) +
  geom_step(color = "steelblue", linewidth = 1) +
  geom_point(aes(color = phase), size = 3) +
  geom_hline(yintercept = 91.45, linetype = "dashed", color = "firebrick") +
  annotate("text", x = 1, y = 91.6, color = "firebrick",
           hjust = 0, label = "True global max (~91.5%)") +
  scale_color_manual(values = c(initial = "black", sequential = "red")) +
  labs(
    title = "BO convergence: best yield vs. trial number",
    subtitle = "Five sequential trials (red) on top of the 6-point initial design (black)",
    x = "Trial number", y = "Best yield observed so far (%)",
    color = "Phase"
  ) +
  theme_minimal()
```

The dashed red line is the true (unknown) global maximum. After only
11 total trials the BO loop has closed most of the gap to the global
optimum, demonstrating the sample efficiency of the GP + EI approach.

# Step 7 --- Multi-Fidelity Optimization with Historical Data

## The Multi-Fidelity Setting

The team also has access to a **coarse kinetic simulation** that
approximates the reactor. The simulation is fast but biased: it
systematically under-predicts the true yield. We treat the simulation
as the **low-fidelity** source and the physical experiments as
**high-fidelity**.

`MultiFidelityOptimizer` implements the Kennedy & O'Hagan (2000)
autoregressive bias-correction idea in a lightweight, model-agnostic
way: it learns a global additive bias between a historical
low-fidelity model and a few high-fidelity runs, then applies that
bias correction to predict the high-fidelity response at any new
design point.

## Historical (Low-Fidelity) Data

```{r mf-historical}
set.seed(7)
historical <- data.frame(
  Temperature = runif(40, 120, 180),
  Catalyst    = runif(40, 0.5,  1.5)
)
# Low-fidelity simulation: biased (under-predicts) and slightly noisy.
historical$Yield <- 0.85 * true_yield(historical$Temperature, historical$Catalyst) -
                    4.0 + rnorm(40, 0, 0.5)

ggplot() +
  geom_point(data = historical, aes(Temperature, Catalyst, color = Yield),
             size = 2.5) +
  scale_color_viridis_c() +
  labs(
    title = "Historical low-fidelity simulation data",
    subtitle = "Biased: under-predicts true yield by ~20 percentage points",
    x = "Temperature (deg C)", y = "Catalyst loading (% w/w)"
  ) +
  theme_minimal()
```

## Assessing the Auxiliary Value

`$evaluate_auxiliary_value()` summarises the historical data and, when
a few high-fidelity runs are available, estimates the bias and a
coarse correlation level (High / Medium / Low) based on how many
historical SDs the high-fidelity mean is shifted by.

```{r mf-auxiliary}
mf <- MultiFidelityOptimizer$new()

# Three high-fidelity physical trials near the suspected optimum.
new_trials <- data.frame(
  Temperature = c(155, 165, 158),
  Catalyst    = c(1.10, 1.20, 1.10)
)
new_trials$Yield <- true_yield(new_trials$Temperature, new_trials$Catalyst)

# Auxiliary-value assessment with the few high-fidelity runs.
aux <- mf$evaluate_auxiliary_value(
  historical_data = historical,
  response_col    = "Yield",
  current_data    = new_trials
)
aux
```

The historical mean is $\approx 66\%$ while the few high-fidelity
trials average $\approx 90\%$, so the bias estimate is large and the
heuristic reports `r aux$correlation` correlation. Even so, the
historical data is informative about the *shape* of the response; we
just need to correct the bias.

## Bias-Corrected Prediction

We first fit a low-fidelity model on the historical data (a quadratic
`lm`), then ask `$bias_corrected_prediction()` to estimate and apply
the additive bias using the few high-fidelity trials.

```{r mf-bias-correction}
# Low-fidelity model trained on the historical simulation data.
historical_model <- lm(
  Yield ~ Temperature + Catalyst +
          I(Temperature^2) + I(Catalyst^2) +
          Temperature:Catalyst,
  data = historical
)

# Grid of design points at which we want a high-fidelity prediction.
design_to_predict <- expand.grid(
  Temperature = seq(120, 180, length.out = 30),
  Catalyst    = seq(0.5,  1.5,  length.out = 30)
)

corrected <- mf$bias_corrected_prediction(
  historical_model  = historical_model,
  current_data      = new_trials,
  design_to_predict = design_to_predict,
  method            = "additive"
)

cat(sprintf("Estimated mean bias: %.3f\n", corrected$bias_correction))
cat(sprintf("Residual sd:         %.3f\n", corrected$residual_sd))
```

We now compare three surfaces over the same grid: the **raw
low-fidelity** prediction, the **bias-corrected** prediction, and the
**true** (unknown in practice) yield.

```{r mf-comparison-plot}
pred_df <- design_to_predict
pred_df$low_fidelity <- as.numeric(predict(historical_model, design_to_predict))
pred_df$corrected    <- corrected$prediction
pred_df$true_yield   <- true_yield(pred_df$Temperature, pred_df$Catalyst)

# Long form for faceting.
pred_long <- rbind(
  data.frame(pred_df[, c("Temperature", "Catalyst")],
             surface = "Low-fidelity (raw)",     yield = pred_df$low_fidelity),
  data.frame(pred_df[, c("Temperature", "Catalyst")],
             surface = "Bias-corrected",         yield = pred_df$corrected),
  data.frame(pred_df[, c("Temperature", "Catalyst")],
             surface = "True (unknown in practice)", yield = pred_df$true_yield)
)
pred_long$surface <- factor(pred_long$surface, levels = c(
  "Low-fidelity (raw)", "Bias-corrected", "True (unknown in practice)"
))

ggplot(pred_long, aes(Temperature, Catalyst)) +
  geom_raster(aes(fill = yield)) +
  geom_contour(aes(z = yield), color = "white", alpha = 0.3, binwidth = 3) +
  facet_wrap(~ surface, nrow = 1) +
  scale_fill_viridis_c() +
  labs(
    title = "Multi-fidelity bias correction",
    subtitle = sprintf("Adding 3 high-fidelity trials shifts the surface by +%.1f on average",
                       corrected$bias_correction),
    x = "Temperature (deg C)", y = "Catalyst loading (% w/w)"
  ) +
  theme_minimal() +
  theme(legend.position = "right")
```

Visually, the bias-corrected surface closely matches the true yield
surface, while the raw low-fidelity surface misses it by a roughly
constant offset. A single batch of three high-fidelity trials is
enough to recalibrate the historical model.

## Augmenting Historical Data with `$augment_historical_data()`

When the historical log is sparse, `$augment_historical_data()`
generates additional synthetic low-fidelity rows by bootstrap
resampling of the existing rows followed by a small Gaussian
perturbation on each numeric column. The perturbation magnitude is
controlled by `noise_factor` as a fraction of each column's standard
deviation. This is a cheap form of data augmentation that can
stabilise the fitted low-fidelity model without running any new
simulations.

```{r mf-augment}
augmented <- mf$augment_historical_data(
  historical_data = historical,
  n_synthetic     = 20,
  noise_factor    = 0.05,
  seed            = 7
)

cat(sprintf("Original rows:  %d\n", nrow(historical)))
cat(sprintf("Augmented rows: %d\n", nrow(augmented)))
cat(sprintf("Synthetic rows added: %d\n", nrow(augmented) - nrow(historical)))

# Compare the marginal distribution of Yield before / after augmentation.
summary_original  <- summary(historical$Yield)
summary_augmented <- summary(augmented$Yield)
cat("\nYield summary (original):\n");  print(summary_original)
cat("\nYield summary (augmented):\n"); print(summary_augmented)
```

**Decision.** The augmented data frame keeps the original 40 rows
intact and appends 20 synthetic rows whose `Yield` distribution
closely matches the original (the five-number summary is preserved up
to the small `noise_factor = 0.05` perturbation). Use this when the
historical log is too small to fit a stable low-fidelity model --- for
example, when the quadratic `lm` in the bias-correction step is
rank-deficient. Set `n_synthetic = 0` to get the input back unchanged,
which is a convenient no-op for unit tests and conditional pipelines.

# Step 8 --- Comparison with Classical RSM

A natural question is whether classical RSM would have done as well.
The honest answer is: **it depends on the initial design**. RSM fits a
single second-order polynomial

$$
y = \beta_0 + \sum_i \beta_i x_i + \sum_i \beta_{ii} x_i^2 + \sum_{i<j}\beta_{ij} x_i x_j + \varepsilon,
$$

which is unimodal in $x$ and therefore **cannot represent the
secondary peak** at all. To illustrate, we fit the RSM second-order
model to the same data accumulated by the BO loop and compare its
predictions to the GP posterior mean.

```{r rsm-comparison}
# RSM second-order polynomial fit on all BO data so far.
rsm_data <- as.data.frame(X_all)
rsm_data$Yield <- -y_all
rsm_fit <- lm(
  Yield ~ Temperature + Catalyst +
          I(Temperature^2) + I(Catalyst^2) +
          Temperature:Catalyst,
  data = rsm_data
)

# Compare predictions on the dense grid.
compare_df <- grid_2d[, c("Temperature", "Catalyst")]
compare_df$GP_posterior_mean <- -bo$predict(compare_df)$predict
compare_df$RSM_prediction    <- predict(rsm_fit, newdata = compare_df)
compare_df$True_yield        <- true_yield(compare_df$Temperature, compare_df$Catalyst)

# RMSE of each surrogate against the true surface.
gp_rmse  <- sqrt(mean((compare_df$GP_posterior_mean - compare_df$True_yield)^2))
rsm_rmse <- sqrt(mean((compare_df$RSM_prediction    - compare_df$True_yield)^2))

cat(sprintf("RMSE vs. true surface:\n"))
cat(sprintf("  GP posterior mean : %.3f\n", gp_rmse))
cat(sprintf("  RSM polynomial    : %.3f\n", rsm_rmse))
```

```{r rsm-comparison-plot}
compare_long <- rbind(
  data.frame(compare_df[, c("Temperature", "Catalyst")],
             surface = "GP posterior mean",
             yield = compare_df$GP_posterior_mean),
  data.frame(compare_df[, c("Temperature", "Catalyst")],
             surface = "RSM polynomial",
             yield = compare_df$RSM_prediction),
  data.frame(compare_df[, c("Temperature", "Catalyst")],
             surface = "True yield",
             yield = compare_df$True_yield)
)
compare_long$surface <- factor(compare_long$surface, levels = c(
  "GP posterior mean", "RSM polynomial", "True yield"
))

ggplot(compare_long, aes(Temperature, Catalyst)) +
  geom_raster(aes(fill = yield)) +
  facet_wrap(~ surface, nrow = 1) +
  scale_fill_viridis_c() +
  labs(
    title = "Surrogate comparison: GP vs RSM polynomial",
    subtitle = sprintf("GP RMSE = %.2f,  RSM RMSE = %.2f  (vs. true surface)", gp_rmse, rsm_rmse),
    x = "Temperature (deg C)", y = "Catalyst loading (% w/w)"
  ) +
  theme_minimal() +
  theme(legend.position = "right")
```

The RSM polynomial smooths over the secondary peak and tilts the
global basin, while the GP reproduces both peaks. With the same
experimental budget, the BO workflow delivers a markedly more
faithful surrogate of the true (multimodal) response.

# Step 9 --- Confirmation Trial

A surrogate is only a model; the recommended optimum must always be
confirmed by a physical trial at the proposed setting. We take the
best point found by the BO loop and confirm it.

```{r confirmation}
best_idx <- which.max(trace$yield)
confirmed <- trace[best_idx, c("Temperature", "Catalyst", "yield")]
confirmed$true_yield <- true_yield(confirmed$Temperature, confirmed$Catalyst)
confirmed
```

The confirmed yield matches the BO prediction, validating the
surrogate. The team has located a high-yield operating regime using
only 11 physical trials plus 3 calibration trials, well within the
budget.

# Step 10 --- Visualising and Reporting with IqrDoeTask

Throughout the case study we have used direct **ggplot2** code to
inspect the GP posterior, the EI surface, and the multi-fidelity
correction. For the underlying experimental design itself, however,
`iQualityR.doe` provides a dedicated plotting API on the `IqrDoeTask`
class. This section builds a small CCD RSM task on the case-study
factors and demonstrates the eight `task$plot()` types, plot
customisation, patchwork composition, and report export.

## Building an IqrDoeTask from the case-study factors

We declare the two case-study factors with their low/center/high
levels, build a CCD design, attach yields simulated from the
digital-twin function, and call `task$compute()` to fit the
second-order RSM model.

```{r task-build}
factors <- list(
  list(name = "Temperature", type = "continuous", levels = c(120, 150, 180)),
  list(name = "Catalyst",    type = "continuous", levels = c(0.5, 1.0, 1.5))
)

task <- rsm_design(
  factors       = factors,
  design_type   = "ccd",
  response_vars = "Yield",
  center_points = 4,
  randomize     = FALSE,
  seed          = 2025,
  auto_compute  = TRUE
)

# Attach observed yields from the digital-twin function and refit.
design_df <- task$results$design_info
design_df$Yield <- true_yield(design_df$Temperature, design_df$Catalyst)
task$data <- design_df
task$compute()

task$summary()
```

The task now holds a fitted second-order model on `Yield`, so all
eight `task$plot()` types are available.

## The `design` plot

```{r task-plot-design, fig.cap = "Run layout tile view produced by task$plot(type = 'design'). Each cell is one run; colour encodes the factor level."}
print(task$plot(type = "design"))
```

**Decision.** The design tile view confirms that the CCD covers the
factor space symmetrically: factorial corner points, axial
star points, and four centre points. Use this plot up front to catch
mistakes in factor coding or run order before any model fitting.

## The `main_effects` plot

```{r task-plot-main-effects, fig.cap = "Marginal effect of each factor on Yield, produced by task$plot(type = 'main_effects')."}
print(task$plot(type = "main_effects"))
```

**Decision.** The marginal-effect panels show that `Temperature` has a
strong non-monotonic effect on yield (rising then plateauing toward
the high-yield basin), while `Catalyst` has a gentler, roughly monotone
effect. Both factors matter; neither can be dropped in a subsequent
screening round.

## The `interaction` plot

```{r task-plot-interaction, fig.cap = "Two-factor interaction panel for Temperature x Catalyst, produced by task$plot(type = 'interaction')."}
print(task$plot(type = "interaction", response_name = "Yield"))
```

**Decision.** The interaction panel reveals that the yield response to
`Temperature` depends on the level of `Catalyst` (the lines are not
parallel). This is consistent with the multimodal true surface and
justifies keeping the `Temperature:Catalyst` term in the second-order
model.

## The `residual` plot

```{r task-plot-residual, fig.cap = "Standard residual diagnostics (QQ, fitted-vs-residual, scale-location) produced by task$plot(type = 'residual')."}
print(task$plot(type = "residual", response_name = "Yield"))
```

**Decision.** The residual diagnostics show no systematic structure
(the QQ points hug the line, the fitted-vs-residual cloud is
centered on zero, and the scale-location band is roughly flat). The
second-order RSM model is a valid baseline for inference, even though
--- as Step 8 showed --- the GP surrogate is more faithful to the
multimodal true surface.

## The `full` combined plot

```{r task-plot-full, fig.cap = "Combined main_effects + interaction + residual panel produced by task$plot(type = 'full')."}
print(task$plot(type = "full", response_name = "Yield"))
```

**Decision.** The `full` panel is a convenient one-figure summary for
reports: it bundles the three diagnostic views above into a single
`patchwork` object. Use it when you need a compact model-validation
figure rather than three separate plots.

## The `surface`, `contour`, and `response_surface` plots

Because the task is a second-order RSM task, the response-surface plot
family is also available. `type = "surface"` returns the 2D tile +
contour surface; `type = "contour"` returns the filled contour only;
`type = "response_surface"` returns a list with `surface_plot`,
`contour_plot`, `combined_plot`, and (if **plotly** is installed)
`interactive_plot`.

```{r task-plot-surface, fig.cap = "2D tile + contour surface of the fitted RSM model, produced by task$plot(type = 'surface')."}
print(task$plot(type = "surface", response_name = "Yield",
                x_var = "Temperature", y_var = "Catalyst"))
```

```{r task-plot-contour, fig.cap = "Filled contour of the fitted RSM model, produced by task$plot(type = 'contour')."}
print(task$plot(type = "contour", response_name = "Yield",
                x_var = "Temperature", y_var = "Catalyst"))
```

```{r task-plot-rsm, fig.cap = "Combined response-surface panel produced by task$plot(type = 'response_surface')$combined_plot."}
rsm_panel <- task$plot(type = "response_surface", response_name = "Yield",
                       x_var = "Temperature", y_var = "Catalyst")
print(rsm_panel$combined_plot)
```

**Decision.** The RSM contour shows a single basin near the centre of
the factor space --- this is the unimodal approximation that the
second-order polynomial can represent. Compare it with the GP
posterior mean in Step 3: the GP captures the secondary peak that the
RSM polynomial smooths over. When the engineer sees this discrepancy,
the right next step is to switch to a GP surrogate (as we did) rather
than trust the RSM optimum blindly.

## Customising plots

`task$plot()` returns ordinary `ggplot` / `patchwork` objects, so you
can post-process them with the usual `+` syntax. A common pattern is
to overlay the next BO trial point on top of the RSM contour to
visualise where the sequential optimiser wants to experiment next.

```{r task-plot-customise, fig.cap = "Customised contour: the RSM contour from task$plot(type = 'contour') with the next BO trial point overlaid in red."}
# Ask the BO surrogate (still fitted on all 11 trials) for the next trial.
next_rec_new <- bo$suggest_next(bounds, n_starts = 4, seed = 42)
next_pt <- as.data.frame(t(next_rec_new$point))
names(next_pt) <- c("Temperature", "Catalyst")

p_contour <- task$plot(type = "contour", response_name = "Yield",
                       x_var = "Temperature", y_var = "Catalyst")

# Overlay the next BO trial point as a red cross.
p_custom <- p_contour +
  geom_point(data = next_pt, aes(Temperature, Catalyst),
             color = "red", size = 4, shape = 4, stroke = 2) +
  labs(subtitle = sprintf("Next BO trial: T = %.1f, C = %.3f (EI = %.4f)",
                          next_pt$Temperature, next_pt$Catalyst,
                          next_rec_new$ei))

print(p_custom)
```

**Decision.** The customised contour shows that the next BO trial
lands in the high-EI region just outside the RSM model's centre basin
--- exactly where the GP expects the largest yield gain. Overlaying
acquisition-function suggestions on top of the classical RSM contour
is a powerful way to communicate to non-expert stakeholders *why* the
sequential optimiser is proposing a point that the polynomial model
would not.

## Combining plots with patchwork

When `task$plot()` returns multiple panels (e.g. `main_effects` and
`contour`), you can compose them into a single figure with
**patchwork**. `wrap_elements()` is the safe way to embed any plot
(including `patchwork` objects returned by `main_effects` / `full`)
inside a larger composition.

```{r task-plot-patchwork, fig.cap = "Patchwork composition of the main-effects panel and the RSM contour, assembled with wrap_elements()."}
p_main  <- task$plot(type = "main_effects")
p_cont  <- task$plot(type = "contour", response_name = "Yield",
                     x_var = "Temperature", y_var = "Catalyst")

combined <- wrap_elements(full = p_main) |
  wrap_elements(full = p_cont)

print(combined)
```

**Decision.** Composing the marginal-effect panel with the
response-surface contour in a single figure gives a one-glance summary
of *both* the per-factor effects and the joint factor behaviour. Use
`|` for side-by-side layouts and `/` for stacked layouts; use
`wrap_elements(full = ...)` whenever one of the inputs is itself a
`patchwork` object (as `main_effects` often is).

## Exporting a report

Finally, `task$report()` exports a self-contained report with the
design table, ANOVA results, model-fit statistics, and embedded
figures. Two formats are supported: `"excel"` (a multi-sheet
`.xlsx`) and `"html"` (a single self-contained `.html`).

```r
# Not run:
tmp_xlsx <- tempfile(fileext = ".xlsx")
tmp_html  <- tempfile(fileext = ".html")

task$report(format = "excel", path = tmp_xlsx)
task$report(format = "html",  path = tmp_html)

cat("Excel report written to:", tmp_xlsx, "\n")
cat("HTML  report written to:", tmp_html, "\n")
```

**Decision.** Use the Excel format when the report needs to be
inspected by engineers in a spreadsheet tool (the design table and
ANOVA tables land on separate sheets). Use the HTML format when the
report is meant to be emailed or embedded in a wiki page; the figures
are inlined as base64, so the file is fully self-contained. In both
cases, call `task$compute()` before `task$report()` to ensure the
results are up to date.

# Summary

This vignette walked through a complete modern DOE workflow on a
realistic chemical-process case study:

1. **Strategy recommendation** with `AutoDOECopilot` provided a
   starting point and flagged the multi-fidelity opportunity.
2. **Gaussian-process surrogate** fitted by `BayesianOptimizer$fit()`
   gave both a posterior mean and a credible band over the response
   surface.
3. **Expected Improvement** computed by
   `BayesianOptimizer$expected_improvement()` identified the most
   informative next trial; `$suggest_next()` found the EI maximiser
   directly.
4. **Sequential BO** driven by `$run_sequential()` (5 iterations)
   closed most of the gap to the global optimum with only 11 total
   trials.
5. **Multi-fidelity correction** with `MultiFidelityOptimizer`
   recalibrated a biased historical simulation using just three
   high-fidelity runs; `$augment_historical_data()` was demonstrated
   as a cheap way to stabilise the low-fidelity model when the
   historical log is sparse.
6. **Comparison with RSM** showed that the GP faithfully captures the
   multimodal response that a second-order polynomial cannot.
7. **Confirmation trial** validated the recommended optimum.
8. **Visualising and reporting** with `IqrDoeTask$plot()` (all eight
   plot types), `patchwork` composition, and `task$report()` export
   closed the loop from analysis to deliverable.

The key insight: when experiments are expensive and the response
surface may be multimodal, a learning system (GP + EI + multi-fidelity
correction) is dramatically more sample-efficient than a fixed
classical design.

# References

1. Frazier, P. I. (2018). *A Tutorial on Bayesian Optimization.*
   arXiv:1807.02811.
2. Jones, D. R., Schonlau, M., & Welch, W. J. (1998). Efficient Global
   Optimization of Expensive Black-Box Functions. *Journal of Global
   Optimization*, 13(4), 455--492.
3. Kennedy, M. C., & O'Hagan, A. (2000). Predicting the Output from a
   Complex Computer Code When Fast Approximations Are Available.
   *Biometrika*, 87(1), 1--13.
4. Rasmussen, C. E., & Williams, C. K. I. (2006). *Gaussian Processes
   for Machine Learning*. MIT Press.
5. Santner, T. J., Williams, B. J., & Notz, W. I. (2018). *The Design
   and Analysis of Computer Experiments* (2nd ed.). Springer.
6. Montgomery, D. C. (2019). *Design and Analysis of Experiments*
   (10th ed.). Wiley.

---

*This vignette was generated by the iQualityR.doe package.*
