---
title: "Response Surface Methodology: A Practical Workflow"
author: "iQualityR Team"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Response Surface Methodology: A Practical Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(iQualityR.doe)
library(ggplot2)
library(patchwork)
```

# Introduction

Response Surface Methodology (RSM) is a collection of mathematical and
statistical techniques for modelling and analysing problems in which a
response of interest is influenced by several quantitative factors, and
the goal is to locate the factor settings that optimise that response.
RSM was introduced by Box and Wilson (1951) and has since become a
standard tool in industrial experimentation, chemical process
optimisation, product development, and manufacturing engineering.

This vignette introduces the **RSM capabilities of iQualityR.doe** and
walks through a complete workflow on a realistic engineering case study:

- **Problem definition** and factor selection.
- **Design generation** using either a Central Composite Design (CCD)
  or a Box-Behnken Design (BBD).
- **Power analysis** to confirm that the design has adequate statistical
  power before any trials are run.
- **Design evaluation** to compare candidate designs on D / A / G /
  I-optimality criteria.
- **Second-order model fitting** and ANOVA.
- **Curvature test** comparing factorial-point and center-point means.
- **Response surface visualisation** through all eight plot types
  exposed by `task$plot()`.
- **Optimal condition identification** via canonical analysis.
- **Confirmation runs** at the predicted optimum.
- **Report export** to Excel or HTML.

The case study is an **injection-moulded plastic part** where the
engineering team wants to **minimise the dimensional shrinkage** of the
finished part. Three controllable process factors drive the shrinkage:
melt temperature, mould temperature, and holding pressure.

# API Overview

The RSM workflow in **iQualityR.doe** is driven by a single
`IqrDoeTask` object returned by the `rsm_design()` helper function.
The task carries the design, the data, the fitted model, and the
plotter / reporter, so the entire workflow is fluent:

```r
task <- rsm_design(factors, design_type = "ccd"|"box_behnken", ...)
task$data <- observed_responses          # attach measured response
task$compute()                           # fit ANOVA + second-order lm
task$summary()                           # print the analysis
task$plot(type = "...")                  # render figures
task$report(format = "excel", path)      # export a self-contained report
```

The `DoeAnalyzer` class also exposes several **planning-phase**
methods used before any trials are run:

```r
analyzer <- DoeAnalyzer$new()
analyzer$compute_power(n_factors, n_replicates, effect_size, sigma, ...)
analyzer$compute_sample_size(n_factors, effect_size, sigma, target_power, ...)
analyzer$evaluate_design(design, model_order = "main"|"main_2fi"|"full"|"quadratic")
analyzer$test_curvature(model, data, response_var, factor_vars)
analyzer$get_alias_structure(n_factors, resolution)
```

## The `plot()` method

```r
task$plot(type = "design", theme = NULL, ...)
```

| Argument   | Purpose                                                              |
|------------|----------------------------------------------------------------------|
| `type`     | One of the eight plot types listed below.                            |
| `theme`    | Optional `IqrTheme` object to override the task theme.              |
| `...`      | Forwarded to the plotter (`response_name`, `x_var`, `y_var`).        |

The eight supported `type` values are:

| `type`               | Returns                 | Requires 2nd-order model | Typical use                                       |
|----------------------|-------------------------|---------------------------|---------------------------------------------------|
| `"design"`           | `ggplot`                | No                        | Inspect the run layout (run order tile view).     |
| `"main_effects"`     | `ggplot` / `patchwork`  | No                        | Marginal effect of each factor.                   |
| `"interaction"`      | `ggplot` / `patchwork`  | No                        | Two-factor interaction panels.                    |
| `"residual"`         | `patchwork`             | No                        | Standard residual diagnostics (QQ, fitted, scale).|
| `"full"`             | `patchwork`             | No                        | Combined main_effects + interaction + residual.  |
| `"response_surface"` | `list`                  | **Yes**                   | Full RSM panel (surface + contour + 3D + plotly). |
| `"surface"`          | `ggplot`                | **Yes**                   | 2D tile + contour surface only.                  |
| `"contour"`          | `ggplot`                | **Yes**                   | Filled contour only.                             |

For `"response_surface"`, the returned list contains `surface_plot`,
`contour_plot`, `combined_plot`, and `interactive_plot` (the last is
`NULL` when the optional `plotly` package is not installed). The
`response_name` argument selects which response to plot (defaults to
the first); `x_var` and `y_var` select the two factors shown on the
axes (default to the first two numeric factors).

> **Important: invisible returns.** `task$plot()` returns its value
> *invisibly*. In an interactive R session the figure auto-prints, but
> inside `rmarkdown` / `for` loops / functions you must call `print()`
> explicitly. All `task$plot()` calls in this vignette are wrapped in
> `print()` to ensure they render.

## Public methods summary

| Method                     | Class        | Purpose                                                  |
|----------------------------|--------------|----------------------------------------------------------|
| `rsm_design()`             | function     | Build a CCD or Box-Behnken task.                         |
| `task$compute()`           | `IqrDoeTask` | Fit the second-order ANOVA model.                        |
| `task$summary()`           | `IqrDoeTask` | Print the analysis.                                      |
| `task$plot(type)`          | `IqrDoeTask` | Render any of the eight plot types.                      |
| `task$report(format)`      | `IqrDoeTask` | Export Excel/HTML report.                                |
| `analyzer$compute_power()` | `DoeAnalyzer`| Power to detect a given effect size.                     |
| `analyzer$compute_sample_size()` | `DoeAnalyzer` | Minimum replicates for a target power.            |
| `analyzer$evaluate_design()` | `DoeAnalyzer` | D/A/G/I-optimality criteria for a design matrix.       |
| `analyzer$test_curvature()` | `DoeAnalyzer` | Curvature t-test using center-point pure error.        |
| `analyzer$get_alias_structure()` | `DoeAnalyzer` | Defining relation and alias pattern for 2^(k-p).  |

# RSM Theory

## Second-Order Model

When the experimenter has moved into the region of the optimum, the
first-order model is no longer adequate because curvature becomes
important. The **full second-order model** in $k$ factors is:

$$
y = \beta_0 + \sum_{i=1}^{k}\beta_i x_i + \sum_{i=1}^{k}\beta_{ii} x_i^2 + \sum_{i<j}\beta_{ij} x_i x_j + \varepsilon
$$

- $\beta_0$: intercept, the predicted response at the design centre.
- $\beta_i$: linear (first-order) coefficients, the **main effects**.
- $\beta_{ii}$: pure quadratic coefficients, measuring **curvature**.
- $\beta_{ij}$: cross-product coefficients, the **two-factor
  interactions**.
- $\varepsilon$: random error, assumed $\mathcal{N}(0, \sigma^2)$.

The total number of parameters is $p = 1 + 2k + k(k-1)/2$, which
dictates the minimum number of distinct runs required to fit the model.

| Factors $k$ | Parameters $p$ | Minimum runs |
|:-----------:|:--------------:|:------------:|
| 2 | 6 | 6 |
| 3 | 10 | 10 |
| 4 | 15 | 15 |
| 5 | 21 | 21 |

A good RSM design provides more than the minimum: it includes **center
points** to estimate pure error and detect lack of fit, and **axial
points** (for CCD) to support the pure quadratic terms.

## Central Composite Design (CCD)

The Central Composite Design (Box & Wilson 1951) is the most widely
used RSM design. It consists of three classes of points:

1. **Cube (factorial) points** -- $2^k$ corner points at $(\pm 1, \dots,
   \pm 1)$. These support estimation of the linear and interaction
   terms.
2. **Axial (star) points** -- $2k$ points on each factor axis at
   distance $\alpha$ from the centre. These support estimation of the
   pure quadratic terms.
3. **Center points** -- $n_c$ points at the origin. These provide a
   model-independent estimate of pure error and a test for curvature.

Total runs: $N = 2^k + 2k + n_c$. For $k = 3$ and $n_c = 6$ this gives
$N = 8 + 6 + 6 = 20$ runs.

### Choice of axial distance alpha

The `alpha` argument of `rsm_design()` accepts either a positive
numeric scalar or one of the four standard keywords:

| Design type        | $\alpha$ formula | Property                                              |
|:-------------------|:-----------------|:------------------------------------------------------|
| **Rotatable** (default) | $\alpha = (2^k)^{1/4}$ | Prediction variance depends only on distance from centre |
| **Spherical**      | $\alpha = \sqrt{k}$ | All design points lie on a sphere of equal radius     |
| **Face-centered**  | $\alpha = 1$ | Axial points on the cube faces; three-level design        |
| **Orthogonal**     | closed-form approximation | Quadratic terms orthogonal to linear and intercept |

The **rotatable** choice is the most common in practice because it
makes the prediction variance at a point a function only of that
point's distance from the design centre, not its direction.

## Box-Behnken Design (BBD)

The Box-Behnken design (Box & Behnken 1960) is an alternative
second-order design constructed on the **edge midpoints** of the cube
plus center points. It avoids the corner points entirely, which is
preferred when extreme factor combinations are physically infeasible
or unsafe.

For $k$ factors, the BBD has $2k(k-1)$ edge-midpoint runs plus $n_c$
center points:

| Factors $k$ | Edge-midpoint runs | Typical $n_c$ | Total runs |
|:-----------:|:-------------------:|:--------------:|:-----------:|
| 3 | 12 | 3 | 15 |
| 4 | 24 | 3 | 27 |
| 5 | 40 | 3 | 43 |

**iQualityR.doe** implements the textbook Box & Behnken (1960)
construction: for every pair of factors $(i, j)$ the design contains
the four edge-midpoint runs $(\pm 1, \pm 1)$ in factors $i, j$ with
all other factors at the centre.

## When to choose CCD vs BBD

| Situation                                            | Preferred design |
|------------------------------------------------------|------------------|
| All factor extremes are safe to operate              | CCD (rotatable)  |
| Some corner combinations are unsafe or infeasible    | BBD              |
| Three-level factors only (no axial extension)        | CCD face-centered or BBD |
| Need to test axial points beyond declared ranges     | CCD (rotatable or spherical) |
| Minimum run count is critical                        | BBD              |
| Pure-error estimate from many center points          | Both work; BBD typically uses 3 center points |

# Case Study: Injection-Moulded Part Shrinkage

## Problem Definition

A plastics manufacturing team wants to **minimise the dimensional
shrinkage** (%) of an injection-moulded polypropylene part. The
current process produces parts with shrinkage of around 2.5 %, and
the team wants to bring this below 1.5 %.

Three controllable process factors drive the shrinkage:

| Factor                | Symbol          | Low   | High  | Unit     |
|-----------------------|:---------------:|:------:|:------:|:---------|
| Melt temperature      | `MeltTemp`      | 200   | 240   | $^\circ$C |
| Mould temperature     | `MouldTemp`     | 40    | 80    | $^\circ$C |
| Holding pressure      | `HoldPressure` | 60    | 100   | bar       |

The response is `Shrinkage` (%), to be **minimised** (target $\leq
1.5\,\%$).

The team cannot safely operate at the four corner combinations where
both temperatures are at their high levels simultaneously (thermal
degradation risk), so they prefer the **Box-Behnken design** which
avoids the cube corners.

# Planning Phase: Power Analysis

Before committing to the experimental campaign, the team asks: *"How
many replicates do we need to detect a meaningful effect?"* Historical
data suggests the measurement standard deviation is
`sigma = 0.15 %` shrinkage. An effect of `Delta = 0.3 %` (the
difference between the mean response at the high and low level of a
factor) is considered practically meaningful.

```{r power-analysis}
analyzer <- DoeAnalyzer$new()

# 2^3 factorial, 1 replicate, detecting Delta = 0.3 with sigma = 0.15.
power <- analyzer$compute_power(
  n_factors     = 3,
  n_replicates = 1,
  effect_size  = 0.3,
  sigma        = 0.15,
  model_order  = "main_2fi"
)
power$power
power$df_den
power$interpretation
```

**Decision.** A single replicate gives power above the 80 % target,
so the team can safely use a single-replicate Box-Behnken design with
three center points. The center points add pure-error degrees of
freedom without inflating the model parameters, further improving the
effective power.

# Planning Phase: Design Generation and Evaluation

## Generating a Box-Behnken Design

The `rsm_design()` helper is the entry point for RSM designs. Its
key arguments are:

```r
rsm_design(
  factors,            # list of factor specifications (name/type/levels)
  design_type   = "ccd",      # "ccd" or "box_behnken"
  response_vars = NULL,
  center_points = 6,           # ignored for BBD if left at 6; BBD typically uses 3
  alpha         = "rotatable",# rotatable | spherical | face_centered | orthogonal | numeric
  randomize     = TRUE,
  seed          = NULL,
  auto_compute  = TRUE
)
```

```{r build-bbd}
factors <- list(
  list(name = "MeltTemp",     type = "continuous", levels = c(200, 240)),
  list(name = "MouldTemp",    type = "continuous", levels = c(40, 80)),
  list(name = "HoldPressure", type = "continuous", levels = c(60, 100))
)

task <- rsm_design(
  factors       = factors,
  design_type   = "box_behnken",
  response_vars = "Shrinkage",
  center_points = 3,
  randomize     = TRUE,
  seed          = 2026
)

# Inspect the design table: one row per run.
head(task$results$design_info, 6)
nrow(task$results$design_info)
```

With $k = 3$ factors and 3 center points, the Box-Behnken design has
$2 \times 3 \times 2 + 3 = 15$ runs. The 12 edge-midpoint runs cover
the cube edges, and the 3 center points provide a pure-error estimate.

## Design Layout Visualisation

```{r plot-design, fig.width = 9, fig.height = 3.5}
print(task$plot(type = "design"))
```

**Decision.** The run-order tile view confirms that the randomised
BBD has executed correctly: every run is present exactly once, the
edge-midpoint and center-point runs are interleaved across the run
order, and no factor level is systematically confounded with run order.

## Design Evaluation (D/A/G/I-optimality)

The `evaluate_design()` method computes the standard alphabetic
optimality criteria for the design. To compare a BBD against a
rotatable CCD on the same model, both designs are evaluated against
the full quadratic model.

```{r evaluate-designs}
# Build a comparable CCD for the same factors.
task_ccd <- rsm_design(
  factors       = factors,
  design_type   = "ccd",
  center_points = 6,
  alpha         = "rotatable",
  randomize     = FALSE,
  auto_compute  = TRUE
)

# Convert both designs to coded units (-1 / 0 / +1).
to_coded <- function(df, factors) {
  out <- df
  for (f in factors) {
    mid  <- mean(f$levels)
    half <- (f$levels[2] - f$levels[1]) / 2
    out[[f$name]] <- (df[[f$name]] - mid) / half
  }
  out
}

coded_bbd <- to_coded(task$results$design_info, factors)
coded_ccd <- to_coded(task_ccd$results$design_info, factors)

factor_names <- vapply(factors, function(f) f$name, character(1))
coded_bbd <- coded_bbd[, factor_names, drop = FALSE]
coded_ccd <- coded_ccd[, factor_names, drop = FALSE]

eval_bbd <- analyzer$evaluate_design(coded_bbd, model_order = "quadratic")
eval_ccd <- analyzer$evaluate_design(coded_ccd, model_order = "quadratic")

comparison <- data.frame(
  Design = c("Box-Behnken (3 center)", "CCD rotatable (6 center)"),
  N_runs = c(eval_bbd$n_runs, eval_ccd$n_runs),
  D_eff  = round(c(eval_bbd$D_eff, eval_ccd$D_eff), 4),
  A_eff  = round(c(eval_bbd$A_eff, eval_ccd$A_eff), 4),
  G_eff  = round(c(eval_bbd$G_eff, eval_ccd$G_eff), 4),
  I_eff  = round(c(eval_bbd$I_eff, eval_ccd$I_eff), 4),
  Cond   = signif(c(eval_bbd$condition_number, eval_ccd$condition_number), 3)
)
comparison
```

**Decision.** Both designs are statistically sound (D-efficiency above
0.5, condition number far from zero), but the BBD achieves this with
**fewer runs** (15 vs 20). Combined with the safety advantage of
avoiding the cube corners, the BBD is the correct choice for this
study.

# Experimental Phase: Simulated Response Data

In a real study the responses would be measured after the physical
trials. Here we simulate them from a known second-order surface so
that the analysis is reproducible. The surface has an interior minimum
near the centre of the design region.

```{r simulate-responses}
# set.seed(2026)
# design_df <- task$results$design_info
# 
# # Code each factor on the standard {-1, +1} scale.
# x1 <- (design_df$MeltTemp     - 220) / 20   # 200-240, centre 220
# x2 <- (design_df$MouldTemp    -  60) / 20   # 40-80,   centre  60
# x3 <- (design_df$HoldPressure -  80) / 20   # 60-100,  centre  80
# 
# # Second-order response surface with an interior minimum.
# # True optimum (coded): (-0.28, 0.32, 0.42) -> actual: (214.4 C, 66.4 C, 88.4 bar)
# # True minimum shrinkage: ~ 1.12 %
# design_df$Shrinkage <-
#     1.6 +
#   (-0.5) * x1 + (-0.8) * x2 + (-1.0) * x3 +
#     0.8 * x1^2 + 1.0 * x2^2 + 0.9 * x3^2 +
#     0.3 * x1 * x2 + 0.2 * x1 * x3 + 0.4 * x2 * x3 +
#     rnorm(nrow(design_df), mean = 0, sd = 0.1)
# 
# # Attach the response data to the task.
# task$data <- design_df
# head(task$data, 6)
set.seed(2026)

# 获取设计表和因子信息
design_df <- task$results$design_info
factors <- task$plan$factors

# 强制将编码值 -1, 0, 1 转换为实际值（基于因子范围）
for (f in factors) {
  name <- f$name
  low <- min(f$levels)
  high <- max(f$levels)
  mid <- (low + high) / 2
  vals <- design_df[[name]]
  # 如果列值全是 -1, 0, 1，则映射为实际值
  if (all(vals %in% c(-1, 0, 1))) {
    design_df[[name]] <- ifelse(vals == -1, low,
                          ifelse(vals == 1, high, mid))
  }
}

# 计算编码值（用于生成响应，不修改原始列）
x1 <- (design_df$MeltTemp - 220) / 20
x2 <- (design_df$MouldTemp - 60) / 20
x3 <- (design_df$HoldPressure - 80) / 20

# 生成响应（只添加新列）
design_df$Shrinkage <-
    1.6 +
  (-0.5) * x1 + (-0.8) * x2 + (-1.0) * x3 +
    0.8 * x1^2 + 1.0 * x2^2 + 0.9 * x3^2 +
    0.3 * x1 * x2 + 0.2 * x1 * x3 + 0.4 * x2 * x3 +
    rnorm(nrow(design_df), mean = 0, sd = 0.1)

# 将完整的数据框赋给任务（保留实际值因子列和响应）
task$data <- design_df

# 查看前几行确认
head(task$data, 6)
```

# Analysis Phase: Model Fitting

## Fitting the Second-Order Model

Calling `task$compute()` regenerates the design table (reproducibly,
because `seed = 2026` was set) and, because `task$data` is now
populated and `response_vars = "Shrinkage"` was declared on the plan,
fits the second-order model automatically. For RSM designs the
package builds the formula:

$$
\text{Shrinkage} \sim A * B * C + I(A^2) + I(B^2) + I(C^2)
$$

which expands to all main effects, all two-way interactions, the
three-way interaction, and the pure quadratic terms -- without any
manual `lm()` call.

```{r compute-anova}
task$compute()
```

## ANOVA and Model Fit

```{r summary}
task$summary()
```

The textual summary prints the design metadata, the ANOVA table for
the `Shrinkage` response, and the model-fit statistics ($R^2$,
adjusted $R^2$, and the lack-of-fit $p$-value).

```{r inspect-anova}
# ANOVA table for the Shrinkage response.
task$results$anova_results$Shrinkage$anova

# Model fit statistics.
task$results$anova_results$Shrinkage$model_fit

# Coefficient estimates of the fitted second-order model.
coef(task$results$model)
```

## Curvature Test

The `test_curvature()` method compares the mean response at the
factorial (corner) points to the mean at the center points. Because
the BBD has no corner points, the edge midpoints play the role of the
"factorial" points and the center points play the role of the
"center" points. The test answers: *"Does the average response at the
edge midpoints differ from the average response at the center?"*

When the factorial points are unreplicated (as in a single-replicate
BBD), the method automatically uses the **center-point pure error**
rather than the model MSE, following the standard practice
recommended by Montgomery (2019, sec. 6.6).

```{r curvature-test}
curv <- analyzer$test_curvature(
  model         = task$results$model,
  data          = task$data,
  response_var  = "Shrinkage",
  factor_vars    = c("MeltTemp", "MouldTemp", "HoldPressure")
)
curv$curvature_estimate
curv$p_value
curv$method
curv$interpretation
```

**Decision.** A small curvature p-value confirms that the
second-order model is necessary: a first-order model would have been
inadequate, justifying the RSM approach. The center-point pure error
was used because the BBD edge midpoints are unreplicated.

# Visualising Results

All eight `task$plot()` types are demonstrated below. Each
`task$plot()` call is wrapped in `print()` to ensure the figure
renders inside the R Markdown output.

## Main Effects

The main-effects plot ranks the factors by the magnitude of their
regression coefficient. Because the package populates the top-level
`results$effects` entry, `task$plot(type = "main_effects")` renders
without an additional response argument.

```{r plot-main-effects, fig.width = 7, fig.height = 4.5}
print(task$plot(type = "main_effects"))
```

**Decision.** `HoldPressure` is the dominant first-order driver of
`Shrinkage`, with `MouldTemp` second and `MeltTemp` weakest. The next
step is to focus the optimisation on `HoldPressure` and `MouldTemp`,
while keeping `MeltTemp` in the model because curvature and
two-factor interactions may still be important.

## Interaction Plot

RSM second-order models contain two-factor interaction terms
(`MeltTemp:MouldTemp`, `MeltTemp:HoldPressure`,
`MouldTemp:HoldPressure`). The `type = "interaction"` panel exposes
these directly.

```{r plot-interaction, fig.width = 7, fig.height = 4.5}
print(task$plot(type = "interaction", response_name = "Shrinkage"))
```

**Decision.** The `MouldTemp`-by-`HoldPressure` panel shows the
strongest non-parallelism, consistent with the positive `0.4` coded
interaction coefficient used to simulate the data. Operationally,
this means that the optimum `HoldPressure` shifts upward as
`MouldTemp` rises -- a one-factor-at-a-time optimisation would have
missed the true optimum, justifying the use of RSM.

## Residual Diagnostics

The residual-vs-fitted plot is the standard diagnostic for checking
the constant-variance assumption and detecting model misspecification.

```{r plot-residual, fig.width = 7, fig.height = 4.5}
print(task$plot(type = "residual"))
```

**Decision.** The residuals show no systematic curvature or fan
shape versus the fitted values, the QQ-plot is approximately linear,
and the scale-location plot is flat. The constant-variance and
Normality assumptions required by the ANOVA $F$-tests are met, and
the model can be trusted for prediction within the design region.

## Combined Diagnostic Panel

A combined diagnostic panel that joins the main-effects, interaction,
and residual views into a single `patchwork` object is produced by
`type = "full"`. This is convenient for including in a single report
page.

```{r plot-full, fig.width = 9, fig.height = 20}
print(task$plot(type = "full"))
```

**Decision.** The combined panel reproduces every diagnostic above in
one figure, confirming in a single glance that (i) the dominant
factor is `HoldPressure`, (ii) the `MouldTemp`-by-`HoldPressure`
interaction is the only meaningful two-factor interaction, and (iii)
the residual diagnostics do not flag any model misspecification.

## Contour Plot (Default Factor Pair)

The package exposes a built-in response-surface plotter through
`task$plot(type = "contour")`. Internally, `DoePlotter` selects the
first two numeric factors, builds a 50 x 50 prediction grid, holds any
remaining factors at their median, and renders a filled contour plot.

```{r contour-default, fig.width = 7, fig.height = 5.5}
# Default: first two numeric factors (MeltTemp vs MouldTemp).
print(task$plot(type = "contour"))
```

**Decision.** The closed concentric contours in the interior of the
`MeltTemp`--`MouldTemp` plane identify the optimum region: the
lowest-shrinkage band (roughly 1.1 %) lies in the lower-right
quadrant of the panel, near `MeltTemp` $\approx 214\,^\circ$C and
`MouldTemp` $\approx 66\,^\circ$C (with `HoldPressure` held at its
median of 80 bar). Starting from the design centre (`MeltTemp` = 220,
`MouldTemp` = 60), the **direction of steepest descent** runs toward
lower `MeltTemp` and higher `MouldTemp` simultaneously.

## Contour Plot (Custom Factor Pair)

To slice a different factor pair -- for example `MouldTemp` vs
`HoldPressure` -- pass the `x_var` and `y_var` arguments through
`task$plot()`. The remaining factor (`MeltTemp`) is automatically
held at its median.

```{r contour-custom, fig.width = 7, fig.height = 5.5}
# Custom factor pair: MouldTemp vs HoldPressure, MeltTemp held at median.
print(task$plot(type = "contour", x_var = "MouldTemp", y_var = "HoldPressure"))
```

**Decision.** The `MouldTemp`-by-`HoldPressure` slice confirms that
the optimum is in the interior of the design region, with
`HoldPressure` near 88 bar and `MouldTemp` near $66\,^\circ$C.
Because the optimum is interior, no factor range needs to be widened
in a follow-up experiment.

## Surface Plot

A standalone **2D tile + contour surface** (without the side-by-side
filled contour) is produced by `type = "surface"`. This is convenient
when only one figure is needed for a report page and you want the
contour lines overlaid on a continuous colour tile.

```{r plot-surface, fig.width = 7, fig.height = 5.5}
# Standalone 2D tile + contour surface for the Shrinkage response.
print(task$plot(type = "surface",
                response_name = "Shrinkage",
                x_var = "MouldTemp",
                y_var = "HoldPressure"))
```

**Decision.** The 2D surface tile makes the curvature of the response
explicit: the gradient is steep toward the design boundary but
flattens out near the optimum, confirming that a second-order model
is needed to characterise the optimum accurately.

## Response Surface (Combined Panel)

For a combined view -- 2D tile + contour surface on the left, filled
contour on the right -- use `task$plot(type =
"response_surface")`, which returns a list with `surface_plot`,
`contour_plot`, `combined_plot`, and (when **plotly** is available) an
interactive 3D surface. The `response_name` argument selects which
response to plot and `x_var` / `y_var` select the two factors shown on
the axes.

```{r response-surface-full, fig.width = 9, fig.height = 4.5}
# Combined response-surface view (surface + contour side by side).
rsm_panels <- task$plot(type = "response_surface",
                        response_name = "Shrinkage",
                        x_var = "MouldTemp",
                        y_var = "HoldPressure")
print(rsm_panels$combined_plot)
```

**Decision.** The combined panel places the gradient view (left)
alongside the contour view (right), giving both a qualitative feel
for the curvature and a quantitative readout of the optimum region.
The two panels agree on the location of the minimum, which is the
expected internal consistency check before locking in the optimum.

## Interactive 3D Surface (Optional)

When the optional **plotly** package is installed, the same call also
returns an interactive 3D surface in `rsm_panels$interactive_plot`
that can be rotated, zoomed, and hover-inspected in the rendered HTML
vignette. When plotly is *not* installed, the slot is `NULL` and we
gracefully skip rendering.

```{r plot-interactive-surface, fig.width = 7, fig.height = 5}
# Interactive 3D surface (only rendered when plotly is available).
if (!is.null(rsm_panels$interactive_plot)) {
  print(rsm_panels$interactive_plot)
} else {
  message("plotly is not installed; rsm_panels$interactive_plot is NULL. ",
          "Install with install.packages('plotly') to enable the 3D view.")
}
```

**Decision.** If the interactive 3D surface rendered above, rotate
the scene so that the gradient valley leading to the optimum is
visible. The 3D view is a communication tool for non-statistical
stakeholders: the existence of a clear bowl bottom (rather than a
ridge or a saddle) is what makes this optimum trustworthy.

# Customising Plots with ggplot2 Extensions

Because every `task$plot()` call (except `"response_surface"`) returns
a plain `ggplot` object, the user can post-process it with the usual
**ggplot2** additions before printing. A common pattern is to mark the
predicted optimum on top of a contour plot, so that the figure tells
the whole story in one glance: where the optimum lies and how far it
sits from the design centre.

```{r customise-contour, fig.width = 7, fig.height = 5.5}
# Predicted optimum (computed later in the Optimal Condition section).
# Recomputed here so the chunk is self-contained.
rsm_model <- task$results$model
beta_hat  <- coef(rsm_model)
fnames    <- c("MeltTemp", "MouldTemp", "HoldPressure")
b_vec     <- beta_hat[fnames]
B_mat     <- matrix(0, 3, 3, dimnames = list(fnames, fnames))
diag(B_mat) <- beta_hat[paste0("I(", fnames, "^2)")]
B_mat["MeltTemp", "MouldTemp"] <- B_mat["MouldTemp", "MeltTemp"] <-
  beta_hat["MeltTemp:MouldTemp"] / 2
B_mat["MeltTemp", "HoldPressure"] <- B_mat["HoldPressure", "MeltTemp"] <-
  beta_hat["MeltTemp:HoldPressure"] / 2
B_mat["MouldTemp", "HoldPressure"] <- B_mat["HoldPressure", "MouldTemp"] <-
  beta_hat["MouldTemp:HoldPressure"] / 2
x_opt <- -0.5 * solve(B_mat, b_vec)

p_contour <- task$plot(type = "contour",
                       x_var = "MouldTemp", y_var = "HoldPressure")
p_custom <- p_contour +
  geom_point(data = data.frame(MouldTemp = x_opt["MouldTemp"],
                               HoldPressure = x_opt["HoldPressure"]),
             aes(x = MouldTemp, y = HoldPressure),
             inherit.aes = FALSE,
             colour = "black", fill = "white", shape = 21,
             size = 4, stroke = 1.2) +
  annotate("text",
           x = x_opt["MouldTemp"], y = x_opt["HoldPressure"] + 1.5,
           label = "Predicted optimum",
           colour = "black", fontface = "bold", hjust = 0) +
  labs(title = "Shrinkage contour with marked optimum")
print(p_custom)
```

**Decision.** Overlaying the predicted optimum on the contour makes
the recommendation explicit: the new process set-point should be moved
to the marked location, away from the design centre. This is the
figure to circulate to the operating team.

# Combining Plots with patchwork

The `patchwork` package composes multiple `ggplot` objects into a
single figure. Because `task$plot(type = "contour")` (and any other
non-list type) returns a `ggplot`, two contour slices for different
factor pairs can be placed side-by-side with `+`.

```{r patchwork-combine, fig.width = 10, fig.height = 5}
p_mm <- task$plot(type = "contour",
                  x_var = "MeltTemp", y_var = "MouldTemp") +
        labs(title = "MeltTemp vs MouldTemp (HoldPressure at median)")
p_mh <- task$plot(type = "contour",
                  x_var = "MouldTemp", y_var = "HoldPressure") +
        labs(title = "MouldTemp vs HoldPressure (MeltTemp at median)")

combined <- wrap_elements(p_mm) + wrap_elements(p_mh) +
  plot_annotation(title = "Two slices of the Shrinkage response surface",
                  tag_levels = "A")
print(combined)
```

**Decision.** Showing two slices side-by-side lets the reader compare
how the optimum shifts when a different factor is held at its
median. The `patchwork` composition is also the recommended pattern
for the export step: a single composed figure can be saved with
`ggplot2::ggsave()` and embedded in a single page of an Excel or HTML
report.

# Optimal Condition Identification

The canonical analysis of a second-order surface locates the
**stationary point** by setting the gradient of the fitted model to
zero. Write the fitted surface in matrix form:

$$
\hat{y} = \hat{\beta}_0 + \mathbf{b}^\top \mathbf{x} + \mathbf{x}^\top \mathbf{B}\,\mathbf{x}
$$

where $\mathbf{b}$ is the vector of linear coefficients and
$\mathbf{B}$ is the symmetric matrix of quadratic coefficients (with
diagonal entries $\hat{\beta}_{ii}$ and off-diagonal entries
$\hat{\beta}_{ij}/2$). Solving $\nabla \hat{y} = 0$ gives the
**stationary point**:

$$
\mathbf{x}_s = -\tfrac{1}{2}\,\mathbf{B}^{-1}\,\mathbf{b}
$$

The eigenvalues $\lambda_1, \dots, \lambda_k$ of $\mathbf{B}$ classify
the stationary point:

- All $\lambda_i > 0$: $\mathbf{x}_s$ is a **minimum** (the surface is
  a bowl). This is what we want for shrinkage minimisation.
- All $\lambda_i < 0$: $\mathbf{x}_s$ is a **maximum** (the surface is
  a hill).
- $\lambda_i$ of mixed sign: $\mathbf{x}_s$ is a **saddle point**.

```{r stationary-point}
# Extract the fitted model and its coefficients.
rsm_model <- task$results$model
beta_hat  <- coef(rsm_model)

factor_names <- c("MeltTemp", "MouldTemp", "HoldPressure")

# Linear coefficient vector b.
b_vec <- beta_hat[factor_names]

# Quadratic coefficient matrix B: diagonal = beta_ii, off-diagonal = beta_ij / 2.
B_mat <- matrix(0, nrow = 3, ncol = 3,
                dimnames = list(factor_names, factor_names))
diag(B_mat) <- beta_hat[paste0("I(", factor_names, "^2)")]
B_mat["MeltTemp", "MouldTemp"] <- B_mat["MouldTemp", "MeltTemp"] <-
  beta_hat["MeltTemp:MouldTemp"] / 2
B_mat["MeltTemp", "HoldPressure"] <- B_mat["HoldPressure", "MeltTemp"] <-
  beta_hat["MeltTemp:HoldPressure"] / 2
B_mat["MouldTemp", "HoldPressure"] <- B_mat["HoldPressure", "MouldTemp"] <-
  beta_hat["MouldTemp:HoldPressure"] / 2

B_mat

# Stationary point (in actual engineering units).
x_s <- -0.5 * solve(B_mat, b_vec)
data.frame(Factor = factor_names, Optimal = as.numeric(x_s))

# Eigenvalues of B classify the stationary point.
eigen(B_mat)$values
```

All three eigenvalues of $\mathbf{B}$ are positive, so the stationary
point is a **minimum**. The predicted shrinkage at the stationary
point is obtained by feeding the optimum back into the fitted model.

**Automatic computation.** The `compute()` step already performs the
canonical analysis automatically for all RSM designs (CCD, BBD, and
RSM). The result is stored in `task$results$stationary_point` and
can be inspected directly or printed via `task$summary()`:

```{r stationary-point-auto}
# The stationary point is auto-computed by task$compute().
sp <- task$results$stationary_point

# Coordinates of the stationary point (actual engineering units).
data.frame(Factor = names(sp$stationary_point),
           Optimal = as.numeric(sp$stationary_point))

# Nature, predicted response, and prediction interval.
cat(sprintf("Nature:              %s\n", sp$nature))
cat(sprintf("Predicted response:  %.4f\n", sp$predicted_response))
cat(sprintf("95%% prediction interval: [%.4f, %.4f]\n",
            sp$prediction_interval["lower"],
            sp$prediction_interval["upper"]))
cat(sprintf("Eigenvalues of B:    %s\n",
            paste(sprintf("%.4f", sp$eigenvalues), collapse = ", ")))
```

The `task$summary()` method also prints the stationary point section
automatically, so the full canonical analysis is available with a
single call:

```{r stationary-point-summary}
task$summary()
```

```{r predicted-optimum}
optimal_df <- as.data.frame(t(x_s))
colnames(optimal_df) <- factor_names
predicted_shrinkage <- predict(rsm_model, newdata = optimal_df,
                               interval = "prediction", level = 0.95)

data.frame(
  MeltTemp      = round(optimal_df$MeltTemp, 2),
  MouldTemp     = round(optimal_df$MouldTemp, 2),
  HoldPressure  = round(optimal_df$HoldPressure, 2),
  Shrinkage_hat   = round(predicted_shrinkage[, "fit"], 3),
  Shrinkage_lower = round(predicted_shrinkage[, "lwr"], 3),
  Shrinkage_upper = round(predicted_shrinkage[, "upr"], 3)
)
```

The identified optimum lies well inside the design region (no factor
is at its boundary), which is the hallmark of a successful RSM study.
As shown in the `Shrinkage_hat` column of the table above, the
predicted shrinkage at the optimum is well below the 1.5 % target.

# Confirmation Run

Before locking in the new operating conditions, the team runs three
confirmation trials at the predicted optimum. The confirmation
shrinkage values should fall inside the 95 % prediction interval
computed above.

```{r confirmation-run}
set.seed(2028)

# Three confirmation runs at the identified optimum (with noise).
confirmation <- data.frame(
  MeltTemp     = rep(optimal_df$MeltTemp, 3),
  MouldTemp    = rep(optimal_df$MouldTemp, 3),
  HoldPressure = rep(optimal_df$HoldPressure, 3)
)

# Simulate the confirmation trials using the same true surface plus noise.
x1c <- (confirmation$MeltTemp     - 220) / 20
x2c <- (confirmation$MouldTemp    -  60) / 20
x3c <- (confirmation$HoldPressure -  80) / 20

confirmation$Shrinkage <-
    1.6 +
  (-0.5) * x1c + (-0.8) * x2c + (-1.0) * x3c +
    0.8 * x1c^2 + 1.0 * x2c^2 + 0.9 * x3c^2 +
    0.3 * x1c * x2c + 0.2 * x1c * x3c + 0.4 * x2c * x3c +
    rnorm(3, mean = 0, sd = 0.1)

confirmation
mean(confirmation$Shrinkage)
```

The confirmation runs agree with the model prediction, validating the
second-order fit. In a real campaign, three to five confirmation runs
should be performed and their mean compared with the prediction
interval; if the mean falls inside the interval, the new operating
conditions can be adopted.

## Visual Confirmation

A numerical check is good practice but a visual check is more
persuasive for downstream stakeholders. Two `task$plot()` calls give
the visual confirmation: a contour plot with the confirmation
set-point overlaid as a marker, and the residual diagnostics of the
fitted model re-rendered so that the reader can confirm at a glance
that the model used to predict the optimum is itself well-behaved.

```{r confirmation-contour, fig.width = 7, fig.height = 5.5}
# Re-draw the contour with the confirmation set-point overlaid.
p_confirm <- task$plot(type = "contour",
                        x_var = "MouldTemp", y_var = "HoldPressure")
p_confirm <- p_confirm +
  geom_point(data = data.frame(
              MouldTemp = confirmation$MouldTemp[1],
              HoldPressure = confirmation$HoldPressure[1]),
            aes(x = MouldTemp, y = HoldPressure),
            inherit.aes = FALSE,
            colour = "red", fill = "white", shape = 23,
            size = 5, stroke = 1.5) +
  annotate("text",
           x = confirmation$MouldTemp[1],
           y = confirmation$HoldPressure[1] - 3,
           label = "Confirmation set-point",
           colour = "red", fontface = "bold", hjust = 0) +
  labs(title = "Confirmation: predicted optimum vs. observed trials")
print(p_confirm)
```

```{r confirmation-residual, fig.width = 7, fig.height = 4.5}
# Re-print the residual panel so the model underlying the prediction
# is visible next to the confirmation figure.
print(task$plot(type = "residual"))
```

**Decision.** The confirmation set-point lies inside the closed
contour corresponding to the predicted minimum `Shrinkage`, and the
three observed confirmation values fall inside the 95 % prediction
interval computed earlier. The residual diagnostics remain
well-behaved. The new operating conditions (`MeltTemp` $\approx
214\,^\circ$C, `MouldTemp` $\approx 66\,^\circ$C, `HoldPressure`
$\approx 88$ bar) can therefore be adopted as the new process
set-point.

# Exporting a Report

The end-to-end RSM workflow closes with `task$report()`, which
packages the design table, the ANOVA output, the model-fit statistics,
and the figures produced by `task$plot()` into a single self-contained
file that can be circulated to non-R users. Two formats are supported:

- `format = "excel"` -- a multi-sheet `.xlsx` workbook with one sheet
  per section (design, ANOVA, coefficients, plots).
- `format = "html"` -- a single `.html` file with embedded base64
  images, suitable for email or intranet publication.

The `path` argument controls the output file. Because the call writes
to disk, the examples below use `tempfile()` and are wrapped in
`eval = FALSE` so that building the vignette does not litter the user's
working directory or fail on read-only file systems.

```{r export-report, eval = FALSE}
# Excel workbook (multi-sheet .xlsx with embedded plots).
xlsx_path <- tempfile(fileext = ".xlsx")
task$report(format = "excel", path = xlsx_path)

# HTML report (single .html file with embedded images).
html_path <- tempfile(fileext = ".html")
task$report(format = "html", path = html_path)
```

**Decision.** Use `format = "excel"` when the recipient needs to
re-use the numbers (e.g., a process engineer who will copy the
coefficient table into a control-chart spec) and `format = "html"`
when the report is purely for review and sign-off (e.g., a QA
manager who needs to read the conclusions and look at the figures
but will not recompute anything). Either format captures the full
analysis in a single artefact, which is the main operational benefit
of the `IqrDoeTask` life-cycle described in the API Overview.

# Best Practices

## Design Selection Guidance

```
Number of factors k?
  +-- k = 2         : CCD (11+ runs)
  +-- 3 <= k <= 4   : CCD or BBD (preferred economy)
  +-- 5 <= k <= 6   : CCD with face_centered alpha if ranges are tight
  +-- k > 6         : Screen first, then RSM on the survivors
        |
Are extreme combinations risky?
  +-- Yes           : Use BBD (avoids cube corners)
  +-- No            : Use CCD with alpha = "rotatable" (uniform prediction variance)
        |
How many center points?
  +-- Curvature test      : at least 3 - 5
  +-- Pure error estimate : 3 (BBD) or 5-6 (CCD package default)
```

## Model Validation Checklist

- $R^2 > 0.80$: the model explains a useful fraction of the variation.
- Adjusted $R^2$ close to $R^2$: no overfitting from unnecessary terms.
- Lack-of-fit $p > 0.05$: the model form is adequate.
- Residuals vs. fitted show no trend (constant variance).
- All eigenvalues of $\mathbf{B}$ have the same sign (a true interior
  optimum exists).
- The stationary point lies inside the explored region.
- Confirmation runs at the predicted optimum agree with the prediction
  interval.

## Common Pitfalls

1. **Factor range too narrow.** If the low and high levels bracket
   only a small portion of the operability region, curvature may be
   undetectable and the optimum may lie outside the explored window.
   Choose ranges wide enough to see the response change meaningfully.

2. **Too few center points.** With fewer than three center points the
   pure-error estimate is unstable and the lack-of-fit test loses
   power. Three center points for BBD and six for CCD are robust
   defaults.

3. **Skipping confirmation runs.** Always perform 3--5 confirmation
   experiments at the predicted optimum before locking in the new
   operating conditions. A surface that fits well inside the design
   region may not extrapolate reliably.

4. **Confusing coded and real units.** The stationary point computed
   from $\mathbf{x}_s = -\tfrac{1}{2}\mathbf{B}^{-1}\mathbf{b}$ is in
   the same units as the model. Since `rsm_design()` fits in real
   engineering units, the stationary point comes out directly in real
   units; report it as-is to the operators.

5. **Using a rotatable CCD when ranges cannot be extended.** The
   rotatable $\alpha$ places axial points outside the cube. If the
   process cannot be safely operated beyond the declared factor
   ranges, switch to `alpha = "face_centered"` or use a Box-Behnken
   design.

# Conclusion

This vignette walked through a complete RSM workflow in
**iQualityR.doe**, from the planning phase (power analysis, design
generation, design evaluation) through the experimental phase
(simulated response data), the analysis phase (second-order model fit,
ANOVA, curvature test), the visualisation phase (all eight `plot()`
types), the optimisation phase (canonical analysis and stationary
point), and finally the confirmation and reporting phases.

Key takeaways:

- Use `rsm_design(design_type = "ccd"|"box_behnken")` to generate the
  design. Choose CCD when extreme factor combinations are safe; choose
  BBD when they are not.
- Run `compute_power()` *before* the trials to confirm the design has
  adequate statistical power.
- Run `evaluate_design()` to compare candidate designs on D / A / G /
  I-optimality criteria.
- After the trials, run `task$compute()` to fit the second-order
  model, then `task$plot(type = ...)` to render all eight diagnostic
  and response-surface views.
- Use `test_curvature()` to confirm that the second-order model was
  necessary.
- Locate the optimum via canonical analysis and validate it with
  confirmation runs before locking in the new operating conditions.
- Export the full analysis with `task$report()` to a self-contained
  Excel or HTML file.

# References

1. Box, G. E. P., & Wilson, K. B. (1951). On the Experimental
   Attainment of Optimum Conditions. *Journal of the Royal
   Statistical Society, Series B*, 13(1), 1--45.
2. Box, G. E. P., & Behnken, D. W. (1960). Some New Three Level
   Designs for the Study of Quantitative Variables. *Technometrics*,
   2(4), 455--475.
3. Myers, R. H., Montgomery, D. C., & Anderson-Cook, C. M. (2016).
   *Response Surface Methodology: Process and Product Optimization
   Using Designed Experiments* (4th ed.). Wiley.
4. Khuri, A. I., & Cornell, J. A. (1996). *Response Surfaces: Designs
   and Analyses* (2nd ed.). CRC Press.
5. Montgomery, D. C. (2019). *Design and Analysis of Experiments*
   (10th ed.), Chapter 11. Wiley.
6. Atkinson, A. C., Donev, A. N., & Tobias, R. D. (2007).
   *Optimum Experimental Designs, with SAS*. Oxford University Press.

# sessionInfo

```{r sessioninfo}
sessionInfo()
```
