---
title: "Response Surface Methodology"
author: "iQualityR Team"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Response Surface Methodology}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(iQualityR.doe)
library(ggplot2)
library(patchwork)
```

# Introduction

Response Surface Methodology (RSM) is a collection of mathematical and
statistical techniques for modelling and analysing problems in which a
response of interest is influenced by several quantitative factors, and
the goal is to locate the factor settings that optimise that response.
RSM was introduced by Box and Wilson (1951) and has since become a
standard tool in industrial experimentation, chemical process
optimisation, and product development.

This vignette shows how to use **iQualityR.doe** to carry out a complete
RSM study: from problem definition, through central composite design
(CCD) generation and the choice of axial distance $\alpha$, to
second-order model fitting, response surface visualisation, and
optimal-condition identification. The workflow is demonstrated on a
realistic chemical reactor yield-optimisation case study with three
factors.

The package API has been designed so that the entire workflow is driven
by a single `IqrDoeTask` object returned by `rsm_design()`. The user
generates the design, attaches the observed response data, runs the
analysis with `task$compute()`, inspects the results with
`task$summary()` and `task$plot()`, and may export a self-contained
report with `task$report()`.

# API Overview

The `IqrDoeTask` class is the central object in **iQualityR.doe**. Its
life-cycle is:

```r
task <- rsm_design(...)         # or factorial_design / screening_design / taguchi_design
task$data <- df                  # attach the observed responses
task$compute()                  # fit ANOVA + second-order lm models
task$summary()                  # print the analysis
task$plot(type = "...")         # generate figures
task$report(format = "excel", path = "...")  # export a report
```

## The `plot()` method

```r
task$plot(type = "design", theme = NULL, ...)
```

| Argument        | Purpose                                                                 |
|-----------------|-------------------------------------------------------------------------|
| `type`          | One of the eight plot types listed below.                              |
| `theme`         | Optional `IqrTheme` object (or theme name) to override the task theme. |
| `...`           | Forwarded to the plotter. For response-surface types this includes `response_name`, `x_var`, and `y_var`. |

The eight supported `type` values are:

| `type`               | Returns                 | Requires 2nd-order model | Typical use                                       |
|----------------------|-------------------------|---------------------------|---------------------------------------------------|
| `"design"`           | `ggplot`                | No                        | Inspect the run layout (run order tile view).     |
| `"main_effects"`     | `ggplot` / `patchwork`  | No                        | Marginal effect of each factor.                   |
| `"interaction"`      | `ggplot` / `patchwork`  | No                        | Two-factor interaction panels.                    |
| `"residual"`         | `patchwork`             | No                        | Standard residual diagnostics (QQ, fitted, scale). |
| `"full"`             | `patchwork`             | No                        | Combined main_effects + interaction + residual.   |
| `"response_surface"` | `list` (see below)      | **Yes**                   | Full RSM panel (surface + contour + 3D + plotly).  |
| `"surface"`          | `ggplot`                | **Yes**                   | 2D tile + contour surface only.                   |
| `"contour"`          | `ggplot`                | **Yes**                   | Filled contour only.                              |

For `"response_surface"`, the returned list contains `surface_plot`,
`contour_plot`, `combined_plot`, and `interactive_plot` (the last is
`NULL` when the optional `plotly` package is not installed). The
`response_name` argument selects which response to plot (defaults to
the first); `x_var` and `y_var` select the two factors shown on the
axes (default to the first two numeric factors).

> **Important: invisible returns.** Like `plot()` on base R objects,
> `task$plot()` returns its value *invisibly*. In an interactive R
> session the figure auto-prints, but inside `rmarkdown` / `for`
> loops / functions you must call `print()` explicitly:

```r
p <- task$plot(type = "contour")
print(p)            # renders the figure inside an R Markdown chunk
```

# Response Surface Methodology Theory

## Second-Order Model

When the experimenter has moved into the region of the optimum, the
first-order model is no longer adequate because curvature becomes
important. The **full second-order model** in $k$ factors is:

$$
y = \beta_0 + \sum_{i=1}^{k}\beta_i x_i + \sum_{i=1}^{k}\beta_{ii} x_i^2 + \sum_{i<j}\beta_{ij} x_i x_j + \varepsilon
$$

The terms have the following interpretation:

- $\beta_0$: intercept, the predicted response at the design centre.
- $\beta_i$: linear (first-order) coefficients, the **main effects**.
- $\beta_{ii}$: pure quadratic coefficients, measuring **curvature**
  along each axis.
- $\beta_{ij}$: cross-product coefficients, measuring **two-factor
  interactions**.
- $\varepsilon$: random error, assumed $\mathcal{N}(0, \sigma^2)$.

The total number of parameters in the model is:

$$
p = 1 + 2k + \frac{k(k-1)}{2}
$$

This count dictates the minimum number of distinct experimental runs
required to fit the model.

| Factors $k$ | Parameters $p$ | Minimum runs |
|:-----------:|:--------------:|:------------:|
| 2 | 6 | 6 |
| 3 | 10 | 10 |
| 4 | 15 | 15 |
| 5 | 21 | 21 |
| 6 | 28 | 28 |

A good RSM design provides more than the minimum: it includes **center
points** to estimate pure error and detect lack of fit, and **axial
points** to support the pure quadratic terms.

## Central Composite Design (CCD)

The Central Composite Design, introduced by Box and Wilson (1951), is
the most widely used RSM design. It consists of three distinct classes
of points:

1. **Cube (factorial) points** -- $2^k$ points at the corners of the
   coded cube, with coordinates $(\pm 1, \pm 1, \dots, \pm 1)$. These
   support estimation of the linear and interaction terms.

2. **Axial (star) points** -- $2k$ points on each factor axis at
   distance $\alpha$ from the centre:
   $(\pm\alpha, 0, \dots, 0), \dots, (0, \dots, 0, \pm\alpha)$. These
   support estimation of the pure quadratic terms.

3. **Center points** -- $n_c$ points at the origin $(0, \dots, 0)$.
   These provide a model-independent estimate of pure error and a test
   for curvature.

The total number of runs of a CCD is:

$$
N = 2^k + 2k + n_c
$$

For $k = 3$ factors with $n_c = 6$ center points, this gives
$N = 8 + 6 + 6 = 20$ runs.

## Choice of Axial Distance alpha

The value of $\alpha$ controls the geometry of the design and its
statistical properties. Four standard choices are supported by
`rsm_design()`:

| Design type | $\alpha$ formula | Property |
|:------------|:-----------------|:---------|
| **Rotatable** (default) | $\alpha = (2^k)^{1/4}$ | Prediction variance depends only on distance from centre |
| **Spherical** | $\alpha = \sqrt{k}$ | All design points lie on a sphere of equal radius |
| **Face-centered** | $\alpha = 1$ | Axial points on the cube faces; three-level design |
| **Orthogonal** | $\alpha = \left(\frac{4 \cdot 2^k (n_c + 2^k) - 2^k \cdot n_c}{2 \cdot 2^k}\right)^{1/4}$ | Quadratic terms orthogonal to linear and intercept |

The **rotatable** choice is the most common in practice because it
makes the prediction variance at a point a function only of that
point's distance from the design centre, not its direction. Worked
examples of the rotatable $\alpha$:

- $k = 2$: $\alpha = (4)^{1/4} = 1.414$
- $k = 3$: $\alpha = (8)^{1/4} = 1.682$
- $k = 4$: $\alpha = (16)^{1/4} = 2.000$

The **face-centered** choice ($\alpha = 1$) places the axial points on
the cube faces, so all design points lie at three levels ($-1, 0, +1$).
This is preferred when the factor ranges cannot be safely extended
beyond the cube (e.g., a temperature cannot exceed a safety limit).

## Box-Behnken Design

The Box-Behnken design (Box and Behnken, 1960) is an alternative
second-order design constructed on a sphere inscribed within the cube.
It avoids the corner points of the cube and is preferred when extreme
factor combinations are physically infeasible or unsafe.

The `rsm_design()` function accepts `design_type = "box_behnken"` and
generates a proper Box-Behnken layout. For each pair of factors (i, j),
four edge-midpoint runs at (±1, ±1) are created while all remaining
factors are held at the center point (0). Center points are then
appended to provide pure-error estimation. The total run count is
$2k(k-1) + n_c$ where $k$ is the number of factors and $n_c$ is the
number of center points.

```{r bbd-generate}
bbd_factors <- list(
  list(name = "Temperature", type = "continuous", levels = c(120, 180)),
  list(name = "Pressure",    type = "continuous", levels = c(1, 3)),
  list(name = "Catalyst",    type = "continuous", levels = c(0.5, 1.5))
)

# Box-Behnken design with 3 center points.
bbd_task <- rsm_design(
  factors       = bbd_factors,
  design_type   = "box_behnken",
  center_points = 3
)

cat("Box-Behnken design (first 6 runs):\n")
print(head(bbd_task$results$design_info))
cat(sprintf("\nTotal runs: %d\n", nrow(bbd_task$results$design_info)))
```

The `PointType` column tags each run as `"edge"` (edge-midpoint) or
`"center"`, which the design plot uses to visually distinguish the
geometric structure of the BBD layout.

When response data is supplied and `task$compute()` is called, the
stationary point is automatically computed via canonical analysis for
all RSM designs (CCD, BBD, and RSM). The `task$summary()` method
reports the stationary point coordinates, the nature of the response
(minimum / maximum / saddle point), the predicted response at the
stationary point with a 95% prediction interval, and the eigenvalues
of the B matrix that determine the curvature direction.

# Case Study: Chemical Reactor Yield Optimization

## Problem Definition

A chemical engineering team wants to maximise the yield (%) of an
active ingredient produced in a continuously stirred reactor. The
target is a yield of at least 90%. Three process factors are believed
to drive the conversion:

| Factor | Symbol | Low level | High level | Unit |
|:-------|:-------|:---------:|:----------:|:-----|
| Reaction temperature | `Temperature` | 120 | 180 | $^\circ$C |
| Pressure | `Pressure` | 1 | 3 | atm |
| Catalyst concentration | `Catalyst` | 0.5 | 1.5 | % |

The response is `Yield` (%), to be **maximised** (target $\geq 90\%$).

Because all three factors can be safely operated at their extreme
combinations, the team selects a **rotatable CCD** with $\alpha = 1.682$
(the rotatable value for $k = 3$), six center points, and the standard
$2^3 = 8$ cube points plus $2 \times 3 = 6$ axial points, for a total
of $8 + 6 + 6 = 20$ runs.

## Generating a CCD with `rsm_design()`

The function `rsm_design()` is the entry point for constructing RSM
designs. Its key arguments are:

```r
rsm_design(
  factors,            # list of factor specifications (name/type/levels)
  design_type   = "ccd",
  response_vars = NULL,
  center_points = 6,
  alpha         = "rotatable",   # rotatable | spherical | face_centered | orthogonal | numeric
  randomize     = TRUE,
  seed          = NULL,
  auto_compute  = TRUE
)
```

Each factor is a list with `name`, `type = "continuous"`, and `levels`
giving the low and high values in real engineering units. The function
returns an `IqrDoeTask` object whose `$results$design_info` is a
`data.frame` containing one row per run with the factor settings and the
run order.

```{r build-ccd}
# Three continuous factors for the reactor study.
factors <- list(
  list(name = "Temperature", type = "continuous", levels = c(120, 180)),
  list(name = "Pressure",    type = "continuous", levels = c(1, 3)),
  list(name = "Catalyst",    type = "continuous", levels = c(0.5, 1.5))
)

# Rotatable CCD with 6 center points, randomised run order.
task <- rsm_design(
  factors       = factors,
  design_type   = "ccd",
  response_vars = "Yield",
  center_points = 6,
  alpha         = "rotatable",
  randomize     = TRUE,
  seed          = 42
)

# Inspect the design table: one row per run.
head(task$results$design_info, 10)
nrow(task$results$design_info)
```

With $k = 3$ and the rotatable $\alpha = 1.682$, the design has
$2^3 + 2 \times 3 + 6 = 20$ runs. The cube points cover the
rectangular operating window, the axial points extend outward along
each axis, and the six center points provide a stable estimate of
pure error.

The design layout can be visualised with `task$plot(type = "design")`.
The function returns a `ggplot` object invisibly, so we print it
explicitly to embed it in the vignette.

```{r plot-design, fig.width = 9, fig.height = 6}
print(task$plot(type = "design"))
```

**Decision.** The run-order tile view confirms that the randomised
CCD has executed correctly: every run is present exactly once, the
factorial, axial, and center-point runs are interleaved across the
run order rather than blocked, and no factor level is systematically
confounded with run order. The design is therefore safe to analyse
without a block term.

## Comparing alpha Choices (rotatable vs face_centered)

A key practical decision when building a CCD is the choice of
$\alpha$. The `alpha` argument now accepts both numeric values and
the keywords `"rotatable"`, `"spherical"`, `"face_centered"`, and
`"orthogonal"`. Below we generate two CCDs for the same three factors
and compare the axial-point positions.

```{r compare-alpha}
# Rotatable CCD: alpha = (2^3)^(1/4) = 1.682 (axial points outside the cube)
task_rot <- rsm_design(
  factors       = factors,
  design_type   = "ccd",
  center_points = 6,
  alpha         = "rotatable",
  randomize     = FALSE,
  auto_compute  = TRUE
)

# Face-centered CCD: alpha = 1 (axial points on the cube faces)
task_fc <- rsm_design(
  factors       = factors,
  design_type   = "ccd",
  center_points = 6,
  alpha         = "face_centered",
  randomize     = FALSE,
  auto_compute  = TRUE
)

# Resolve the actual alpha values used.
k <- 3
alpha_rot <- (2^k)^(1 / 4)      # 1.681793
alpha_fc  <- 1
c(rotatable = alpha_rot, face_centered = alpha_fc)
```

Both designs have the same total number of runs ($2^3 + 2 \times 3 + 6
= 20$), but the axial points land at very different positions in real
units. The rotatable axial points extend **beyond** the declared
factor ranges, while the face-centered axial points sit **on** the
cube faces.

```{r compare-alpha-table}
# Axial points only (rows where exactly one factor is off-centre).
# For a non-randomised CCD the first 8 rows are factorial, the next 6 are axial,
# the last 6 are centre points.
rot_axial <- head(task_rot$results$design_info, 14)
rot_axial <- tail(rot_axial, 6)
fc_axial  <- head(task_fc$results$design_info, 14)
fc_axial  <- tail(fc_axial, 6)

rot_axial$alpha <- "rotatable (alpha = 1.682)"
fc_axial$alpha  <- "face_centered (alpha = 1)"
rbind(rot_axial, fc_axial)
```

For the `Temperature` factor (range 120--180 $^\circ$C, centre 150):

- Rotatable axial points are at $150 \pm 1.682 \times 30 = 99.5$ and
  $200.5^\circ$C, which lie **outside** the declared range.
- Face-centered axial points are at $150 \pm 1 \times 30 = 120$ and
  $180^\circ$C, sitting **on** the cube faces.

If the reactor cannot be safely operated above $180^\circ$C, the
face-centered CCD is the correct choice. If extended ranges are
permissible, the rotatable CCD gives more uniform prediction variance.
The geometric difference is clearly visible in the side-by-side layout
below.

```{r compare-alpha-plot, fig.width = 9, fig.height = 4.5}
# Classify each design point as Factorial / Axial / Center.
classify_point <- function(df, plan_factors) {
  centers <- vapply(plan_factors, function(f) mean(f$levels), numeric(1))
  half_ranges <- vapply(plan_factors, function(f) (max(f$levels) - min(f$levels)) / 2, numeric(1))
  types <- character(nrow(df))
  for (i in seq_len(nrow(df))) {
    row <- as.numeric(df[i, names(centers)])
    is_center <- all(abs(row - centers) < 1e-6)
    if (is_center) { types[i] <- "Center"; next }
    # Axial: exactly one factor off-centre.
    off_center <- sum(abs(row - centers) > 1e-6)
    types[i] <- if (off_center == 1) "Axial" else "Factorial"
  }
  types
}

design_rot <- task_rot$results$design_info
design_fc  <- task_fc$results$design_info
design_rot$Type <- classify_point(design_rot, factors)
design_fc$Type  <- classify_point(design_fc, factors)
design_rot$Alpha <- "rotatable (alpha = 1.682)"
design_fc$Alpha  <- "face_centered (alpha = 1)"

both <- rbind(design_rot, design_fc)

ggplot(both, aes(x = Temperature, y = Pressure, colour = Type, shape = Type)) +
  geom_point(size = 3) +
  facet_wrap(~ Alpha) +
  geom_hline(yintercept = 2, linetype = "dotted", alpha = 0.4) +
  geom_vline(xintercept = 150, linetype = "dotted", alpha = 0.4) +
  scale_color_manual(values = c("Factorial" = "#2C7BB6",
                                "Axial" = "#D7191C",
                                "Center" = "#1A9641")) +
  labs(
    title = "CCD layout: rotatable vs face_centered alpha",
    subtitle = "Temperature vs Pressure (Catalyst varies similarly)",
    x = "Temperature (deg C)",
    y = "Pressure (atm)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Simulated Response Data

In a real study the responses would be measured after the physical
trials. Here we simulate them from a known second-order surface so
that the analysis is reproducible. The surface has an interior maximum
near the centre of the design region, with positive linear effects,
negative quadratic effects (interior optimum), and two interaction
terms (Temperature-by-Pressure and Temperature-by-Catalyst). Random
noise is added with `set.seed(42)` for reproducibility.

```{r simulate-responses}
set.seed(42)

# Recover the generated design table (reproducible: seed = 42 was set).
design_df <- task$results$design_info

# Code each factor on the standard {-1, +1} scale for the simulation model.
x1 <- (design_df$Temperature - 150) / 30   # Temperature: 120-180, centre 150
x2 <- (design_df$Pressure    - 2)   / 1    # Pressure:    1-3,   centre 2
x3 <- (design_df$Catalyst    - 1.0) / 0.5  # Catalyst:    0.5-1.5, centre 1.0

# Second-order response surface with an interior maximum.
# True optimum (coded): (0.29, 0.41, 0.49) -> actual: (158.8 C, 2.41 atm, 1.24 %)
# True maximum yield  : ~ 90.9 %
design_df$Yield <-
    90.0 +
    0.6 * x1 + 1.8 * x2 + 1.8 * x3 +                # linear terms
  (-2.0) * x1^2 + (-2.5) * x2^2 + (-2.0) * x3^2 +   # quadratic terms (curvature)
    0.8 * x1 * x2 + 0.5 * x1 * x3 +                  # interactions (T x P, T x C)
    rnorm(nrow(design_df), mean = 0, sd = 0.5)       # measurement noise

# Attach the response data to the task. The public `data` field is a
# direct assignment on the IqrTaskBase class.
task$data <- design_df

head(task$data, 6)
```

## Fitting the Second-Order Model

Calling `task$compute()` regenerates the design table (reproducibly,
because `seed = 42` was set) and, because `task$data` is now populated
and `response_vars = "Yield"` was declared on the plan, fits the
second-order model automatically. For CCD and other RSM designs the
package builds the formula:

$$
\text{Yield} \sim A * B * C + I(A^2) + I(B^2) + I(C^2)
$$

which expands to all main effects, all two-way interactions, the
three-way interaction, and the pure quadratic terms -- without any
manual `lm()` call. The fitted model is stored in
`task$results$model`, and top-level `task$results$model` and
`task$results$model_fit` are also populated so that downstream
methods (`task$plot(type = "residual")`) work without an explicit
response argument.

```{r compute-anova}
task$compute()
```

## ANOVA and Model Fit

The textual summary prints the design metadata, the ANOVA table for
each response, and the model-fit statistics ($R^2$, adjusted $R^2$,
and the lack-of-fit $p$-value).

```{r summary}
task$summary()
```

The ANOVA table can also be inspected directly. The linear terms, the
pure quadratic terms, and the engineered two-way interactions should
stand out as significant, confirming that the second-order model is
necessary.

```{r inspect-anova}
# ANOVA table for the Yield response.
task$results$anova_results$Yield$anova

# Model fit statistics.
task$results$anova_results$Yield$model_fit

# Coefficient estimates of the fitted second-order model.
coef(task$results$model)
```

## Main Effects, Interactions, and Residual Diagnostics

The main-effects plot ranks the factors by the magnitude of their
regression coefficient. Because the package populates the top-level
`results$effects` entry, `task$plot(type = "main_effects")` renders
without an additional response argument.

```{r plot-main-effects, fig.width = 7, fig.height = 4.5}
print(task$plot(type = "main_effects"))
```

**Decision.** `Pressure` and `Catalyst` are the dominant first-order
drivers of `Yield`, with `Temperature` a weaker third. The next step
is therefore to spend the model-fitting effort primarily on `Pressure`
and `Catalyst` (and their interactions), while keeping `Temperature`
in the model because curvature and two-factor interactions may still
be important even when the marginal effect is small.

RSM second-order models contain two-factor interaction terms
(`Temperature:Pressure`, `Temperature:Catalyst`, `Pressure:Catalyst`
in this case study). The `type = "interaction"` panel exposes these
directly. Because the case study has a single response (`Yield`) the
`response_name` argument is optional but we pass it explicitly to make
the chunk self-documenting.

```{r plot-interaction, fig.width = 7, fig.height = 9}
print(task$plot(type = "interaction", response_name = "Yield"))
```

**Decision.** The `Temperature`-by-`Pressure` panel shows the
strongest non-parallelism, consistent with the positive `0.8` coded
interaction coefficient used to simulate the data. Operationally,
this means that the optimum `Pressure` shifts upward as `Temperature`
rises -- a one-factor-at-a-time optimisation would therefore have
missed the true optimum, justifying the use of RSM.

The residual-vs-fitted plot is the standard diagnostic for checking
the constant-variance assumption and detecting model misspecification.
Because `task$results$model` is populated, `task$plot(type =
"residual")` is available.

```{r plot-residual, fig.width = 7, fig.height = 4.5}
print(task$plot(type = "residual"))
```

**Decision.** The residuals show no systematic curvature or fan
shape versus the fitted values, the QQ-plot is approximately linear,
and the scale-location plot is flat. The constant-variance and
Normality assumptions required by the ANOVA $F$-tests are therefore
met, and the model can be trusted for prediction within the design
region.

A combined diagnostic panel that joins the main-effects, interaction,
and residual views into a single `patchwork` object is produced by
`type = "full"`. This is convenient for including in a single report
page.

```{r plot-full, fig.width = 9, fig.height = 12}
print(task$plot(type = "full"))
```

**Decision.** The combined panel reproduces every diagnostic above
in one figure, confirming in a single glance that (i) the dominant
factors are `Pressure` and `Catalyst`, (ii) the
`Temperature`-by-`Pressure` interaction is the only meaningful
two-factor interaction, and (iii) the residual diagnostics do not
flag any model misspecification. The model is ready to be used for
response-surface visualisation.

## Response Surface Visualization (contour plot)

The package exposes a built-in response-surface plotter through
`task$plot(type = "contour")`. Internally, `DoePlotter` selects the
first two numeric factors, builds a 50 x 50 prediction grid, holds any
remaining factors at their median, and renders a filled contour plot
with `geom_contour_filled()`. The method returns a `ggplot` object
invisibly, so we print it explicitly.

```{r contour-default, fig.width = 7, fig.height = 5.5}
# Default: first two numeric factors (Temperature vs Pressure).
print(task$plot(type = "contour"))
```

**Decision.** The closed concentric contours in the interior of the
`Temperature`--`Pressure` plane identify the optimum region: the
highest-yield band (roughly 90 %) lies in the upper-right quadrant
of the panel, near `Temperature` $\approx 158\,^\circ$C and
`Pressure` $\approx 2.4$ atm (with `Catalyst` held at its median of
1.0 %). Starting from the design centre (`Temperature` = 150,
`Pressure` = 2), the **direction of steepest ascent** runs toward
higher `Temperature` and higher `Pressure` simultaneously -- exactly
the move a follow-up CCD would have made if the first experiment had
been a screening design.

To slice a different factor pair -- for example `Temperature` vs
`Catalyst` -- pass the `x_var` and `y_var` arguments through
`task$plot()`. The remaining factor (`Pressure`) is automatically held
at its median.

```{r contour-temp-catalyst, fig.width = 7, fig.height = 5.5}
# Custom factor pair: Temperature vs Catalyst, Pressure held at median.
print(task$plot(type = "contour", x_var = "Temperature", y_var = "Catalyst"))
```

**Decision.** The `Temperature`-by-`Catalyst` slice confirms that the
optimum is in the interior of the design region, with `Catalyst` near
$1.25\,\%$ and `Temperature` near $158\,^\circ$C. Because the optimum
is interior, no factor range needs to be widened in a follow-up
experiment -- the second-order model is reliable here.

A standalone **2D tile + contour surface** (without the side-by-side
filled contour) is produced by `type = "surface"`. This is convenient
when only one figure is needed for a report page and you want the
contour lines overlaid on a continuous colour tile.

```{r plot-surface, fig.width = 7, fig.height = 5.5}
# Standalone 2D tile + contour surface for the Yield response.
print(task$plot(type = "surface",
                response_name = "Yield",
                x_var = "Temperature",
                y_var = "Pressure"))
```

**Decision.** The 2D surface tile makes the curvature of the response
explicit: the gradient is steep toward the cube faces but flattens out
near the optimum, confirming that a second-order model -- not a
first-order gradient -- is needed to characterise the optimum
accurately.

For a combined view -- 2D tile + contour surface on the left, filled
contour on the right -- use `task$plot(type = "response_surface")`,
which returns a list with `surface_plot`, `contour_plot`,
`combined_plot`, and (when **plotly** is available) an interactive 3D
surface. The `response_name` argument selects which response to plot
and `x_var` / `y_var` select the two factors shown on the axes.

```{r response-surface-full, fig.width = 9, fig.height = 4.5}
# Combined response-surface view (surface + contour side by side).
rsm_panels <- task$plot(type = "response_surface",
                        response_name = "Yield",
                        x_var = "Temperature",
                        y_var = "Catalyst")
print(rsm_panels$combined_plot)
```

**Decision.** The combined panel places the gradient view (left)
alongside the contour view (right), giving both a qualitative feel
for the curvature and a quantitative readout of the optimum region.
The two panels agree on the location of the maximum, which is the
expected internal consistency check before locking in the optimum.

When the optional **plotly** package is installed, the same call also
returns an interactive 3D surface in `rsm_panels$interactive_plot` that
can be rotated, zoomed, and hover-inspected in the rendered HTML
vignette. When plotly is *not* installed, the slot is `NULL` and we
gracefully skip rendering. We deliberately do **not** add `plotly` to
the setup chunk so that the vignette remains buildable in
environments without plotly.

```{r plot-interactive-surface, eval = TRUE, fig.width = 7, fig.height = 5}
# Interactive 3D surface (only rendered when plotly is available).
if (!is.null(rsm_panels$interactive_plot)) {
  print(rsm_panels$interactive_plot)
} else {
  message("plotly is not installed; rsm_panels$interactive_plot is NULL. ",
          "Install with install.packages('plotly') to enable the 3D view.")
}
```

**Decision.** If the interactive 3D surface rendered above, rotate
the scene so that the gradient ridge leading to the optimum is
visible. The 3D view is a communication tool for non-statistical
stakeholders: the existence of a clear hilltop (rather than a ridge or
a saddle) is what makes this optimum trustworthy.

## Customising plots

Because every `task$plot()` call (except `"response_surface"`) returns
a plain `ggplot` object, you can post-process it with the usual
**ggplot2** additions before printing. A common pattern is to mark the
predicted optimum on top of a contour plot, so that the figure tells
the whole story in one glance: where the optimum lies and how far it
sits from the design centre.

```{r customise-contour, fig.width = 7, fig.height = 5.5}
# Predicted optimum (computed later in the Optimal Condition section).
# Recomputed here so the chunk is self-contained.
rsm_model  <- task$results$model
beta_hat   <- coef(rsm_model)
fnames     <- c("Temperature", "Pressure", "Catalyst")
b_vec      <- beta_hat[fnames]
B_mat      <- matrix(0, 3, 3, dimnames = list(fnames, fnames))
diag(B_mat) <- beta_hat[paste0("I(", fnames, "^2)")]
B_mat["Temperature", "Pressure"] <- B_mat["Pressure", "Temperature"] <-
  beta_hat["Temperature:Pressure"] / 2
B_mat["Temperature", "Catalyst"] <- B_mat["Catalyst", "Temperature"] <-
  beta_hat["Temperature:Catalyst"] / 2
B_mat["Pressure", "Catalyst"] <- B_mat["Catalyst", "Pressure"] <-
  beta_hat["Pressure:Catalyst"] / 2
x_opt <- -0.5 * solve(B_mat, b_vec)

p_contour <- task$plot(type = "contour",
                       x_var = "Temperature", y_var = "Pressure")
p_custom <- p_contour +
  geom_point(data = data.frame(Temperature = x_opt["Temperature"],
                               Pressure = x_opt["Pressure"]),
             aes(x = Temperature, y = Pressure),
             inherit.aes = FALSE,
             colour = "black", fill = "white", shape = 21,
             size = 4, stroke = 1.2) +
  annotate("text",
           x = x_opt["Temperature"], y = x_opt["Pressure"] + 0.15,
           label = "Predicted optimum",
           colour = "black", fontface = "bold", hjust = 0) +
  labs(title = "Yield contour with marked optimum")
print(p_custom)
```

**Decision.** Overlaying the predicted optimum on the contour makes
the recommendation explicit: the new reactor set-point should be
moved to the marked location, away from the design centre. This is
the figure to circulate to the operating team.

## Combining plots with patchwork

The `patchwork` package composes multiple `ggplot` objects into a
single figure. Because `task$plot(type = "contour")` (and any other
non-list type) returns a `ggplot`, two contour slices for different
factor pairs can be placed side-by-side with `+` (or `wrap_elements()`
when the inputs are themselves `patchwork` objects).

```{r patchwork-combine, fig.width = 10, fig.height = 5}
p_tp <- task$plot(type = "contour",
                  x_var = "Temperature", y_var = "Pressure") +
        labs(title = "T vs P (Catalyst at median)")
p_tc <- task$plot(type = "contour",
                  x_var = "Temperature", y_var = "Catalyst") +
        labs(title = "T vs C (Pressure at median)")

combined <- wrap_elements(p_tp) + wrap_elements(p_tc) +
  plot_annotation(title = "Two slices of the Yield response surface",
                  tag_levels = "A")
print(combined)
```

**Decision.** Showing two slices side-by-side lets the reader compare
how the optimum shifts when a different factor is held at its
median. The `patchwork` composition is also the recommended pattern
for the export step: a single composed figure can be saved with
`ggplot2::ggsave()` and embedded in a single page of an Excel or HTML
report.

## Optimal Condition Identification

The canonical analysis of a second-order surface locates the
**stationary point** by setting the gradient of the fitted model to
zero. Write the fitted surface in matrix form:

$$
\hat{y} = \hat{\beta}_0 + \mathbf{b}^\top \mathbf{x} + \mathbf{x}^\top \mathbf{B}\,\mathbf{x}
$$

where $\mathbf{b}$ is the vector of linear coefficients and
$\mathbf{B}$ is the symmetric matrix of quadratic coefficients (with
diagonal entries $\hat{\beta}_{ii}$ and off-diagonal entries
$\hat{\beta}_{ij}/2$). Solving $\nabla \hat{y} = 0$ gives the
**stationary point**:

$$
\mathbf{x}_s = -\tfrac{1}{2}\,\mathbf{B}^{-1}\,\mathbf{b}
$$

The eigenvalues $\lambda_1, \dots, \lambda_k$ of $\mathbf{B}$ classify
the stationary point:

- All $\lambda_i < 0$: $\mathbf{x}_s$ is a **maximum** (the surface is
  a hill).
- All $\lambda_i > 0$: $\mathbf{x}_s$ is a **minimum** (the surface is
  a bowl).
- $\lambda_i$ of mixed sign: $\mathbf{x}_s$ is a **saddle point**.

The **canonical form** of the fitted surface, expressed in the
eigenvector coordinate system $\mathbf{w}$ centred at $\mathbf{x}_s$,
is:

$$
\hat{y} = \hat{y}_s + \sum_{i=1}^{k}\lambda_i\,w_i^2
$$

where $\hat{y}_s$ is the predicted response at the stationary point.

The package does not (yet) expose a one-call canonical-analysis
function, but the formula is straightforward to evaluate from the
fitted `lm` object stored in `task$results$model`.

```{r stationary-point}
# Extract the fitted model and its coefficients.
rsm_model <- task$results$model
beta_hat  <- coef(rsm_model)

factor_names <- c("Temperature", "Pressure", "Catalyst")

# Linear coefficient vector b.
b_vec <- beta_hat[factor_names]

# Quadratic coefficient matrix B: diagonal = beta_ii, off-diagonal = beta_ij / 2.
B_mat <- matrix(0, nrow = 3, ncol = 3,
                dimnames = list(factor_names, factor_names))
diag(B_mat) <- beta_hat[paste0("I(", factor_names, "^2)")]
B_mat["Temperature", "Pressure"] <- B_mat["Pressure", "Temperature"] <-
  beta_hat["Temperature:Pressure"] / 2
B_mat["Temperature", "Catalyst"] <- B_mat["Catalyst", "Temperature"] <-
  beta_hat["Temperature:Catalyst"] / 2
B_mat["Pressure", "Catalyst"] <- B_mat["Catalyst", "Pressure"] <-
  beta_hat["Pressure:Catalyst"] / 2

B_mat

# Stationary point (in actual engineering units).
x_s <- -0.5 * solve(B_mat, b_vec)
data.frame(Factor = factor_names, Optimal = as.numeric(x_s))

# Eigenvalues of B classify the stationary point.
eigen(B_mat)$values
```

All three eigenvalues of $\mathbf{B}$ are negative, so the stationary
point is a **maximum**. The predicted yield at the stationary point
is obtained by feeding the optimum back into the fitted model.

```{r predicted-optimum}
optimal_df <- as.data.frame(t(x_s))
colnames(optimal_df) <- factor_names
predicted_yield <- predict(rsm_model, newdata = optimal_df,
                           interval = "prediction", level = 0.95)

data.frame(
  Temperature = round(optimal_df$Temperature, 2),
  Pressure    = round(optimal_df$Pressure, 2),
  Catalyst    = round(optimal_df$Catalyst, 3),
  Yield_hat   = round(predicted_yield[, "fit"], 2),
  Yield_lower = round(predicted_yield[, "lwr"], 2),
  Yield_upper = round(predicted_yield[, "upr"], 2)
)
```

The identified optimum lies well inside the design region (no factor
is at its boundary), which is the hallmark of a successful RSM study.

## Confirmation Run

Before locking in the new operating conditions, the team runs three
confirmation trials at the predicted optimum. The confirmation yields
should fall inside the 95 % prediction interval computed above.

```{r confirmation-run}
set.seed(2024)

# Three confirmation runs at the identified optimum (with noise).
confirmation <- data.frame(
  Temperature = rep(optimal_df$Temperature, 3),
  Pressure    = rep(optimal_df$Pressure, 3),
  Catalyst    = rep(optimal_df$Catalyst, 3)
)

# Simulate the confirmation trials using the same true surface plus noise.
x1c <- (confirmation$Temperature - 150) / 30
x2c <- (confirmation$Pressure    - 2)   / 1
x3c <- (confirmation$Catalyst    - 1.0) / 0.5

confirmation$Yield <-
    90.0 +
    0.6 * x1c + 1.8 * x2c + 1.8 * x3c +
  (-2.0) * x1c^2 + (-2.5) * x2c^2 + (-2.0) * x3c^2 +
    0.8 * x1c * x2c + 0.5 * x1c * x3c +
    rnorm(3, mean = 0, sd = 0.5)

confirmation
mean(confirmation$Yield)
```

The confirmation runs agree with the model prediction, validating the
second-order fit. In a real campaign, three to five confirmation runs
should be performed and their mean compared with the prediction
interval; if the mean falls inside the interval, the new operating
conditions can be adopted.

### Visual confirmation with `task$plot()`

A numerical check is good practice but a visual check is more
persuasive for downstream stakeholders. Two `task$plot()` calls give
the visual confirmation: a contour plot with the confirmation
set-point overlaid as a marker, and the residual diagnostics of the
fitted model re-rendered so that the reader can confirm at a glance
that the model used to predict the optimum is itself well-behaved.

```{r confirmation-contour, fig.width = 7, fig.height = 5.5}
# Re-draw the contour with the confirmation set-point overlaid.
p_confirm <- task$plot(type = "contour",
                        x_var = "Temperature", y_var = "Pressure")
p_confirm <- p_confirm +
  geom_point(data = data.frame(
              Temperature = confirmation$Temperature[1],
              Pressure    = confirmation$Pressure[1]),
            aes(x = Temperature, y = Pressure),
            inherit.aes = FALSE,
            colour = "red", fill = "white", shape = 23,
            size = 5, stroke = 1.5) +
  annotate("text",
           x = confirmation$Temperature[1],
           y = confirmation$Pressure[1] - 0.18,
           label = "Confirmation set-point",
           colour = "red", fontface = "bold", hjust = 0) +
  labs(title = "Confirmation: predicted optimum vs. observed trials")
print(p_confirm)
```

```{r confirmation-residual, fig.width = 7, fig.height = 4.5}
# Re-print the residual panel so the model underlying the prediction
# is visible next to the confirmation figure.
print(task$plot(type = "residual"))
```

**Decision.** The confirmation set-point lies inside the closed
contour corresponding to the predicted maximum `Yield`, and the three
observed confirmation values fall inside the 95 % prediction interval
computed earlier. The residual diagnostics remain well-behaved. The
new operating conditions (`Temperature` $\approx 158.8\,^\circ$C,
`Pressure` $\approx 2.41$ atm, `Catalyst` $\approx 1.24$ %) can
therefore be adopted as the new reactor set-point.

# Best Practices

## Design Selection Guidance

```
Number of factors k?
  +-- k = 2         : CCD (11+ runs)
  +-- 3 <= k <= 4   : CCD (preferred economy)
  +-- 5 <= k <= 6   : CCD with face_centered alpha if ranges are tight
  +-- k > 6         : Screen first, then RSM on the survivors
        |
Are extreme combinations risky?
  +-- Yes           : Use alpha = "face_centered" (axial on cube faces)
  +-- No            : Use alpha = "rotatable" (uniform prediction variance)
        |
How many center points?
  +-- Curvature test      : at least 3 - 5
  +-- Pure error estimate : 5 - 6 (package default)
```

## Model Validation Checklist

- $R^2 > 0.80$: the model explains a useful fraction of the variation.
- Adjusted $R^2$ close to $R^2$: no overfitting from unnecessary terms.
- Lack-of-fit $p > 0.05$: the model form is adequate.
- Residuals vs. fitted show no trend (constant variance).
- All eigenvalues of $\mathbf{B}$ have the same sign (a true interior
  optimum exists).
- The stationary point lies inside the explored region.
- Confirmation runs at the predicted optimum agree with the prediction
  interval.

## Common Pitfalls

1. **Factor range too narrow.** If the low and high levels bracket
   only a small portion of the operability region, curvature may be
   undetectable and the optimum may lie outside the explored window.
   Choose ranges wide enough to see the response change meaningfully.

2. **Too few center points.** With fewer than five center points the
   pure-error estimate is unstable and the lack-of-fit test loses
   power. Six center points is a robust default.

3. **Skipping confirmation runs.** Always perform 3--5 confirmation
   experiments at the predicted optimum before locking in the new
   operating conditions. A surface that fits well inside the design
   region may not extrapolate reliably.

4. **Confusing coded and real units.** The stationary point computed
   from $\mathbf{x}_s = -\tfrac{1}{2}\mathbf{B}^{-1}\mathbf{b}$ is in
   the same units as the model. Since `rsm_design()` fits in real
   engineering units, the stationary point comes out directly in
   real units; report it as-is to the operators.

5. **Using a rotatable CCD when ranges cannot be extended.** The
   rotatable $\alpha$ places axial points outside the cube. If the
   process cannot be safely operated beyond the declared factor
   ranges, switch to `alpha = "face_centered"`.

# Comparison with Other R Packages

| Feature | iQualityR.doe | rsm | DoE.base | rsm (rsm::rsm) |
|:--------|:-------------:|:---:|:--------:|:--------------:|
| CCD generation              | Yes | Yes | Yes | Yes |
| Box-Behnken generation      | Yes | Yes | Yes | Yes |
| Alpha keywords (rotatable / face_centered / ...) | Yes | Yes | No | No |
| Automatic second-order model fit | Yes | Yes (via rsm object) | No | Yes |
| Stationary point (auto)     | Yes | Yes | No | Yes |
| Contour / surface plots     | Yes (`task$plot`) | Yes | No | Yes |
| Residual diagnostics        | Yes (`task$plot`) | Yes | No | Yes |
| Automated HTML report       | Yes | No | No | No |
| Integration with screening / SPC | Yes | No | Partial | No |

The strengths of **iQualityR.doe** lie in tight integration between
screening, RSM, MSA, and SPC; the convenience of a single
`IqrDoeTask` object that flows through design, fit, plot, and report;
the explicit `alpha` keyword API for CCD generation; automatic
stationary point computation for all RSM designs (CCD, BBD); and
IqrTheme-consistent visualizations that reuse the shared
**iQualityR.plot** infrastructure.

# Exporting a Report

The end-to-end RSM workflow closes with `task$report()`, which packages
the design table, the ANOVA output, the model-fit statistics, and the
figures produced by `task$plot()` into a single self-contained file
that can be circulated to non-R users. Two formats are supported:

- `format = "excel"` -- a multi-sheet `.xlsx` workbook with one sheet
  per section (design, ANOVA, coefficients, plots).
- `format = "html"` -- a single `.html` file with embedded base64
  images, suitable for email or intranet publication.

The `path` argument controls the output file. Because the call writes
to disk, the examples below use `tempfile()` and are wrapped in
`# Not run` so that building the vignette does not litter the user's
working directory or fail on read-only file systems.

```{r export-report, eval = FALSE}
# Not run:

# Excel workbook (multi-sheet .xlsx with embedded plots).
xlsx_path <- tempfile(fileext = ".xlsx")
task$report(format = "excel", path = xlsx_path)
# message("Excel report written to: ", xlsx_path)

# HTML report (single .html file with embedded images).
html_path <- tempfile(fileext = ".html")
task$report(format = "html", path = html_path)
# message("HTML report written to: ", html_path)
```

**Decision.** Use `format = "excel"` when the recipient needs to
re-use the numbers (e.g., a process engineer who will copy the
coefficient table into a control-chart spec) and `format = "html"`
when the report is purely for review and sign-off (e.g., a QA
manager who needs to read the conclusions and look at the figures
but will not recompute anything). Either format captures the full
analysis in a single artefact, which is the main operational benefit
of the `IqrDoeTask` life-cycle described in the API Overview.

# References

1. Box, G. E. P., & Wilson, K. B. (1951). On the Experimental
   Attainment of Optimum Conditions. *Journal of the Royal
   Statistical Society, Series B*, 13(1), 1--45.
2. Box, G. E. P., & Behnken, D. W. (1960). Some New Three Level
   Designs for the Study of Quantitative Variables. *Technometrics*,
   2(4), 455--475.
3. Myers, R. H., Montgomery, D. C., & Anderson-Cook, C. M. (2016).
   *Response Surface Methodology: Process and Product Optimization
   Using Designed Experiments* (4th ed.). Wiley.
4. Khuri, A. I., & Cornell, J. A. (1996). *Response Surfaces: Designs
   and Analyses* (2nd ed.). CRC Press.
5. Montgomery, D. C. (2019). *Design and Analysis of Experiments*
   (10th ed.), Chapter 11. Wiley.
6. Lenth, R. V. (2009). Response-Surface Methods in R, Using **rsm**.
   *Journal of Statistical Software*, 32(7), 1--17.

# sessionInfo

```{r sessioninfo}
sessionInfo()
```
