---
title: "Factorial Designs with iQualityR.doe"
author: "iQualityR Team"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Factorial Designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(iQualityR.doe)
library(ggplot2)
library(patchwork)
```

# Introduction

Injection molding is the most widely used polymer-conversion process in the
plastics industry. A recurring quality problem is **shrinkage**: the molded
part contracts as it cools, and if the contraction is excessive or uneven the
part fails its dimensional specification. The phenomenon is driven by several
process variables that interact strongly with one another, which makes
One-Factor-At-a-Time (OFAT) experimentation ineffective.

This vignette shows how to use **iQualityR.doe** to address a realistic
shrinkage-optimization problem with a full $2^k$ factorial design. The
workflow is the one a quality engineer would follow on the shop floor:

1. Define the problem and the factors to be studied.
2. Generate a randomized experimental design with `factorial_design()`.
3. Collect (here: simulate) the response data and attach them to the task.
4. Run the analysis of variance with `task$compute()`.
5. Visualise main effects, interactions, and residuals.
6. Identify the optimal factor combination.
7. Run confirmation trials at the predicted optimum.

By the end of the vignette the reader will be able to plan, analyse, and
report a two-level factorial experiment entirely from the
`IqrDoeTask` API, without writing any base R model-fitting code by hand.

# API Overview

The `IqrDoeTask` class is the central object in **iQualityR.doe**. Its
life-cycle is:

```r
task <- factorial_design(...)   # or rsm_design / screening_design / taguchi_design
task$data <- df                 # attach the observed responses
task$compute()                 # fit ANOVA + lm models
task$summary()                 # print the analysis
task$plot(type = "...")        # generate figures
task$report(format = "excel", path = "...")  # export a report
```

## The `plot()` method

```r
task$plot(type = "design", theme = NULL, ...)
```

| Argument        | Purpose                                                                 |
|-----------------|-------------------------------------------------------------------------|
| `type`          | One of the eight plot types listed below.                              |
| `theme`         | Optional `IqrTheme` object (or theme name) to override the task theme. |
| `...`           | Forwarded to the plotter. For response-surface types this includes `response_name`, `x_var`, and `y_var`. |

The eight supported `type` values are:

| `type`               | Returns                 | Requires 2nd-order model | Typical use                                       |
|----------------------|-------------------------|---------------------------|---------------------------------------------------|
| `"design"`           | `ggplot`                | No                        | Inspect the run layout (run order tile view).     |
| `"main_effects"`     | `ggplot` / `patchwork`  | No                        | Marginal effect of each factor.                   |
| `"interaction"`      | `ggplot` / `patchwork`  | No                        | Two-factor interaction panels.                    |
| `"residual"`         | `patchwork`             | No                        | Standard residual diagnostics (QQ, fitted, scale). |
| `"full"`             | `patchwork`             | No                        | Combined main_effects + interaction + residual.   |
| `"response_surface"` | `list` (see below)      | **Yes**                   | Full RSM panel (surface + contour + 3D + plotly).  |
| `"surface"`          | `ggplot`                | **Yes**                   | 2D tile + contour surface only.                   |
| `"contour"`          | `ggplot`                | **Yes**                   | Filled contour only.                              |

For `"response_surface"`, the returned list contains `surface_plot`,
`contour_plot`, `combined_plot`, and `interactive_plot` (the last is
`NULL` when the optional `plotly` package is not installed). The
`response_name` argument selects which response to plot (defaults to
the first); `x_var` and `y_var` select the two factors shown on the
axes (default to the first two numeric factors).

> **Important: invisible returns.** Like `plot()` on base R objects,
> `task$plot()` returns its value *invisibly*. In an interactive R
> session the figure auto-prints, but inside `rmarkdown` / `for`
> loops / functions you must call `print()` explicitly:

```r
p <- task$plot(type = "main_effects")
print(p)            # renders the figure inside an R Markdown chunk
```

The same life-cycle applies to `rsm_design()`, `screening_design()`,
and `taguchi_design()` constructors — they all return an
`IqrDoeTask`.

# Factorial Design Theory

## The $2^k$ factorial design

A $2^k$ factorial design studies $k$ factors, each at two levels, in every
possible combination. The total number of runs is $n = 2^k$ (plus any
replicates and center points). For $k = 4$ factors the base design has
$2^4 = 16$ runs; with two replicates it has $32$ runs, which provides enough
degrees of freedom to estimate all main effects and interactions while still
leaving residual degrees of freedom for the error term.

## Effect estimation

For a two-level design coded in the standard $\{-1, +1\}$ convention, the
**effect** of factor $A$ is the difference between the mean response at the
high level and the mean response at the low level:

$$
\text{Effect}_A \;=\; \bar{y}_{A+} \;-\; \bar{y}_{A-}.
$$

The **contrast** is the signed sum of the responses, with signs given by the
contrast column of the design matrix:

$$
\text{Contrast}_A \;=\; \sum_{i=1}^{n} c_i\, y_i, \qquad c_i \in \{-1, +1\}.
$$

The two quantities are related by the number of runs at each level,
$n/2 = 2^{k-1}$:

$$
\text{Contrast}_A \;=\; \frac{n}{2}\,\text{Effect}_A \;=\; 2^{k-1}\,\text{Effect}_A.
$$

For the $2^2$ design ($k = 2$, $n = 4$) this reduces to
$\text{Contrast}_A = 2 \cdot \text{Effect}_A$.

The ordinary-least-squares regression coefficient $\hat\beta_A$ is half the
effect, because the coded variable changes by two units when moving from
$-1$ to $+1$:

$$
\hat\beta_A \;=\; \frac{\text{Effect}_A}{2}.
$$

## The Yates algorithm

The Yates algorithm is a fast, recursive procedure for computing all $2^k - 1$
factorial effects from a standard-order design table. Starting from the
response column in standard (Yates) order, one performs $k$ passes; in each
pass, paired sums and paired differences are placed in the upper and lower
halves of a new column. After $k$ passes the column contains the contrasts,
which are divided by $n/2$ to obtain the effects. The algorithm is the
historical antecedent of the modern regression-based approach used by
`lm()` and by `IqrDoeTask$compute()`.

## ANOVA decomposition

The total sum of squares is partitioned into contributions from each factor
and their interactions, plus the error:

$$
\text{SS}_{T} \;=\; \text{SS}_A + \text{SS}_B + \text{SS}_{AB} + \dots + \text{SS}_{E}.
$$

For an orthogonal two-level design each contrast has one degree of freedom,
and its sum of squares is

$$
\text{SS}_{\text{Effect}} \;=\; \frac{(\text{Contrast})^2}{n}
\;=\; \frac{n}{4}\,(\text{Effect})^2.
$$

The test statistic for the null hypothesis "the effect is zero" is

$$
F_0 \;=\; \frac{\text{MS}_{\text{Effect}}}{\text{MS}_{E}}
\;\sim\; F_{1,\, df_{E}},
$$

where $\text{MS}_{\text{Effect}} = \text{SS}_{\text{Effect}} / 1$ and
$\text{MS}_{E} = \text{SS}_{E} / df_{E}$. An effect is declared significant
at level $\alpha$ when $F_0 > F_{\alpha, 1, df_{E}}$, equivalently when the
$p$-value is below $\alpha$.

# Case Study: Injection Molding Shrinkage

## Problem definition

An automotive supplier manufactures a polypropylene dashboard panel by
injection molding. The customer specification requires the molded-part
shrinkage to remain below $1.5\%$ and the dimensional deviation to remain
below $15\,\mu\text{m}$. The current process averages $2.5\%$ shrinkage and
$30\,\mu\text{m}$ deviation, so both responses need to be reduced.

The engineering team selects four potentially significant factors:

| Factor | Symbol | Low level | High level | Unit |
|--------|--------|-----------|------------|------|
| Mold temperature | `MoldTemp` | 40 | 80 | $^\circ\text{C}$ |
| Holding pressure | `HoldPres` | 40 | 100 | MPa |
| Injection speed | `InjectSpeed` | 50 | 150 | mm/s |
| Cooling time | `CoolTime` | 15 | 45 | s |

Two responses are measured on each run:

* `Shrinkage` — linear shrinkage (%), to be **minimised**.
* `DimensionDev` — dimensional deviation ($\mu\text{m}$), to be **minimised**.

A full $2^4$ design with two replicates is chosen. This gives
$2 \times 16 = 32$ runs, leaving $32 - 16 = 16$ residual degrees of freedom
for the full first-order-with-interaction model
`Shrinkage ~ MoldTemp * HoldPres * InjectSpeed * CoolTime`. The two
replicates also provide a pure-error estimate of $\sigma^2$.

## Design generation with `factorial_design()`

The factor list is built first, with **numeric** level vectors (the package
validator rejects character levels). The design is then generated with
`factorial_design()`, which returns an `IqrDoeTask` object whose
`$results$design_info` is a `data.frame` containing one row per run with the
factor settings and the run order.

```{r build-design}
# Four continuous factors, each at two numeric levels.
factors <- list(
  list(name = "MoldTemp",     type = "continuous", levels = c(40,  80)),
  list(name = "HoldPres",     type = "continuous", levels = c(40, 100)),
  list(name = "InjectSpeed",  type = "continuous", levels = c(50, 150)),
  list(name = "CoolTime",     type = "continuous", levels = c(15,  45))
)

# Full 2^4 factorial design, two replicates, randomized run order.
task <- factorial_design(
  factors       = factors,
  design_type   = "factorial",
  response_vars = c("Shrinkage", "DimensionDev"),
  replication   = 2,
  randomize     = TRUE,
  seed          = 42
)

# The generated design table: one row per run.
head(task$results$design_info, 8)
nrow(task$results$design_info)
```

The design layout can be visualised with `task$plot(type = "design")`. The
function returns a `ggplot` object invisibly, so we print it explicitly to
embed it in the vignette.

```{r plot-design, fig.width = 9, fig.height = 8}
print(task$plot(type = "design"))
```

**Decision.** The tile view confirms that all 32 runs are present and
that the run order is properly randomised across the four-factor grid.
No clustering or diagonal pattern is visible, so the randomisation has
succeeded in protecting the experiment against time-dependent lurking
variables such as tool wear or ambient temperature drift. The design
itself does not depend on `response_name`, because it visualises the
factor settings, not the measured responses.

## Simulated response data

In a real study the responses would be measured after the physical trials.
Here we simulate them from a known physics-inspired model so that the
analysis is reproducible. Higher mold temperature, higher holding pressure,
faster injection, and longer cooling all reduce shrinkage, with a
`MoldTemp`-by-`HoldPres` interaction and a `HoldPres`-by-`CoolTime`
interaction. Random noise is added with `set.seed(42)` for reproducibility.

```{r simulate-responses}
set.seed(42)

design_df <- task$results$design_info

# Code each factor on the standard {-1, +1} scale for the simulation model.
# x1 <- (design_df$MoldTemp    - 60) / 20   # MoldTemp    centered & scaled
# x2 <- (design_df$HoldPres    - 70) / 30   # HoldPres
# x3 <- (design_df$InjectSpeed - 100) / 50  # InjectSpeed
# x4 <- (design_df$CoolTime    -  30) / 15  # CoolTime
# 
# # Shrinkage (%) -- lower is better.
# design_df$Shrinkage <- 2.20
#   - 0.30 * x1 - 0.40 * x2 - 0.25 * x3 - 0.30 * x4
#   - 0.15 * x1 * x2
#   - 0.10 * x2 * x4
#   + rnorm(nrow(design_df), sd = 0.05)

# 重新设计系数
set.seed(42)
x1 <- (design_df$MoldTemp - 60) / 20
x2 <- (design_df$HoldPres - 70) / 30
x3 <- (design_df$InjectSpeed - 100) / 50
x4 <- (design_df$CoolTime - 30) / 15

# 只有 x1, x2, x4 有实际效应，x3 是惰性因子
design_df$Shrinkage <- 2.20
  - 0.20 * x1        # 中等效应
  - 0.35 * x2        # 强效应
  - 0.02 * x3        # 几乎无效应（不显著）
  - 0.20 * x4        # 中等效应
  - 0.10 * x1 * x2   # 交互项
  + rnorm(nrow(design_df), sd = 0.15)   # 更大的噪声

# Dimensional deviation (micrometres) -- lower is better.
# design_df$DimensionDev <- 28.0
#   - 4.0 * x1 - 5.0 * x2 - 3.0 * x3 - 4.0 * x4
#   - 2.5 * x1 * x2
#   - 1.5 * x2 * x4
#   + rnorm(nrow(design_df), sd = 0.8)

design_df$DimensionDev <- 28.0
  - 3.5 * x1
  - 5.0 * x2
  - 0.5 * x3          # 小效应，可能显著或不显著
  - 3.0 * x4
  - 2.0 * x1 * x2
  + rnorm(nrow(design_df), sd = 1.2)

# Attach the response data to the task. The public `data` field is a
# direct assignment on the IqrTaskBase class.
task$data <- design_df
```

A quick look at the simulated data confirms the structure: factor columns,
the run-order columns, and the two measured responses.

```{r preview-data}
head(task$data, 6)
```

## ANOVA analysis with `task$compute()`

Calling `task$compute()` regenerates the design table (reproducibly, because
`seed = 42` was set) and, because `task$data` is now populated and
`response_vars` was declared on the plan, fits the full-interaction model
for each response and populates:

* `task$results$anova_results` --- a named list, one entry per response;
* `task$results$effects` --- main and interaction effects for the first response;
* `task$results$model` --- the fitted `lm` object for the first response;
* `task$results$model_fit` --- $R^2$, adjusted $R^2$, and the lack-of-fit $p$-value.

```{r compute-anova}
task$compute()

# textual summary of the design, the ANOVA tables, and the fit statistics
task$summary()
```

The ANOVA table for `Shrinkage` can also be inspected directly. The four
main effects and the two engineered interactions (`MoldTemp:HoldPres` and
`HoldPres:CoolTime`) should stand out as significant, while the higher-order
interactions should be negligible.

```{r inspect-anova}
# ANOVA table for the Shrinkage response
task$results$anova_results$Shrinkage$anova

# Model fit statistics for both responses
task$results$anova_results$Shrinkage$model_fit
task$results$anova_results$DimensionDev$model_fit
```

## Main effects plot

The main-effects plot ranks the factors by the magnitude of their effect
estimate. Because the underlying physics makes every main effect negative
(high level $\Rightarrow$ lower shrinkage), all four bars should point
downward. By default `task$plot(type = "main_effects")` plots the first
declared response, here `Shrinkage`.

```{r plot-main-effects,fig.width = 9, fig.height = 8}
print(task$plot(type = "main_effects"))
```

**Decision.** The main-effect slope for `HoldPres` is the steepest, so
it is the dominant lever on `Shrinkage`, followed by `MoldTemp`,
`CoolTime`, and `InjectSpeed`. All four slopes point downward,
confirming the engineering expectation that the high level of every
factor reduces shrinkage. The optimisation should therefore push every
factor to its high level.

To plot the second response, pass `response_name = "DimensionDev"`.
Only one response is plotted per call --- the package does not overlay
multiple responses on the same axis, because they live on different
units (% versus $\mu\text{m}$).

```{r plot-main-effects-dimdev,fig.width = 9, fig.height = 8}
print(task$plot(type = "main_effects", response_name = "DimensionDev"))
```

**Decision.** The ranking for `DimensionDev` mirrors the one for
`Shrinkage`, with `HoldPres` again the dominant lever. Because both
responses share the same direction of effect, no trade-off is needed:
a single recipe (all factors high) is optimal for both responses
simultaneously.

## Interaction effects plot

The interaction plot shows, for every pair of factors, the mean response at
each level of one factor stratified by the level of the other. Non-parallel
lines signal an interaction. Two pairs should be visibly non-parallel:
`MoldTemp` $\times$ `HoldPres` and `HoldPres` $\times$ `CoolTime`. The
default call plots the `Shrinkage` response; the `DimensionDev` panel is
produced by passing `response_name`.

```{r plot-interaction,fig.width = 9, fig.height = 8}
print(task$plot(type = "interaction"))
```

**Decision.** The two engineered interaction panels
(`MoldTemp` $\times$ `HoldPres` and `HoldPres` $\times$ `CoolTime`)
show clearly non-parallel lines, confirming that the effect of
`HoldPres` is amplified when `MoldTemp` and `CoolTime` are at their
high levels. The remaining panels are roughly parallel, so the
higher-order interactions can be pooled into the error term if a
reduced model is needed for production.

```{r plot-interaction-dimdev,fig.width = 9, fig.height = 8}
print(task$plot(type = "interaction", response_name = "DimensionDev"))
```

**Decision.** The `DimensionDev` interaction panels reproduce the same
two non-parallel pairs as `Shrinkage`, so the same interaction
structure drives both responses. A single reduced model
(`MoldTemp * HoldPres + HoldPres * CoolTime`) is therefore adequate
for both responses, which simplifies the production monitoring plan.

## Residual diagnostics

The residual-versus-fitted plot is the primary diagnostic for the
regression assumptions. With replicated runs the residuals should be
scattered symmetrically around zero with no systematic curvature or
funnel shape. The default `type = "residual"` call diagnostics the
`Shrinkage` model; pass `response_name` to switch to the second
response.

```{r plot-residual,,fig.width = 9, fig.height = 8}
print(task$plot(type = "residual"))
```

**Decision.** The residual QQ plot for `Shrinkage` is approximately
linear, the fitted-versus-residuals cloud is centred on zero with no
curvature, and the scale-location plot is flat. The regression
assumptions are satisfied, so the ANOVA $p$-values and prediction
intervals for `Shrinkage` are trustworthy.

```{r plot-residual-dimdev,fig.width = 9, fig.height = 8}
print(task$plot(type = "residual", response_name = "DimensionDev"))
```

**Decision.** The `DimensionDev` residuals behave the same way as the
`Shrinkage` residuals, with no visible curvature, funnel, or heavy
tails. Both responses can be modelled with the same
first-order-with-interaction specification, and the prediction
intervals reported below are reliable.

The combined view of main effects and residuals is produced by
`type = "full"`:

```{r plot-full, ,fig.width = 9, fig.height = 12}
print(task$plot(type = "full"))
```

**Decision.** The combined panel brings the main-effects, interaction,
and residual diagnostics for `Shrinkage` into a single figure. The
picture is consistent with the per-type plots above and gives no
reason to revise the model. The full panel is the recommended
single-figure summary for an internal technical report.

```{r plot-full-dimdev, fig.width = 11, fig.height = 12}
print(task$plot(type = "full", response_name = "DimensionDev"))
```

**Decision.** The combined panel for `DimensionDev` confirms the same
conclusions reached for `Shrinkage`. The two responses can be
reported together using the same model structure, with `HoldPres` as
the dominant lever in both cases.

## Optimal condition identification

Because every main effect on `Shrinkage` is negative, the minimum-shrinkage
setting is the high level of every factor. The same conclusion holds for
`DimensionDev`. We can confirm this by predicting the response at the
optimal combination with the fitted model and its $95\%$ prediction
interval.

```{r optimal-condition}
# Main effect signs for the Shrinkage response.
task$results$anova_results$Shrinkage$effects$main

# The optimal setting: all factors at their high level.
optimal_cond <- data.frame(
  MoldTemp    = 80,
  HoldPres    = 100,
  InjectSpeed = 150,
  CoolTime    = 45
)

# Point prediction and 95% prediction interval for Shrinkage.
pred_shrinkage <- predict(
  task$results$anova_results$Shrinkage$model,
  newdata    = optimal_cond,
  interval   = "prediction"
)
pred_shrinkage

# Point prediction and 95% prediction interval for DimensionDev.
pred_dimdev <- predict(
  task$results$anova_results$DimensionDev$model,
  newdata    = optimal_cond,
  interval   = "prediction"
)
pred_dimdev
```

Both predicted means fall comfortably inside the customer specification
(`Shrinkage` < 1.5%, `DimensionDev` < 15 $\mu\text{m}$).

## Confirmation run

A confirmation experiment consists of a small number of trials performed at
the predicted optimal setting. If the model is correct, the mean of the
confirmation runs should fall inside the prediction interval computed
above. We simulate five confirmation runs from the same data-generating
process used in Section 3.3.

```{r confirmation-run}
set.seed(123)

# Five confirmation runs at the optimal condition (all factors high).
confirm_shrinkage <- 2.20
  - 0.30 - 0.40 - 0.25 - 0.30
  - 0.15  # x1 * x2 interaction at (+1, +1)
  - 0.10  # x2 * x4 interaction at (+1, +1)
  + rnorm(5, sd = 0.05)

confirm_dimdev <- 28.0
  - 4.0 - 5.0 - 3.0 - 4.0
  - 2.5
  - 1.5
  + rnorm(5, sd = 0.8)

confirmation <- data.frame(
  Run        = 1:5,
  Shrinkage  = round(confirm_shrinkage, 4),
  DimensionDev = round(confirm_dimdev, 4)
)
confirmation

cat(sprintf(
  "Mean Shrinkage = %.4f %% (95%% PI: [%.4f, %.4f])\n",
  mean(confirm_shrinkage), pred_shrinkage[, "lwr"], pred_shrinkage[, "upr"]
))
cat(sprintf(
  "Mean DimensionDev = %.4f microm (95%% PI: [%.4f, %.4f])\n",
  mean(confirm_dimdev), pred_dimdev[, "lwr"], pred_dimdev[, "upr"]
))
```

Both confirmation means lie inside their $95\%$ prediction intervals and
inside the customer specification, so the model is validated and the
optimal recipe can be released to production.

## Optional: HTML report

The `task$report()` method renders a self-contained HTML report from an
internal R Markdown template. The call below writes the report to a
temporary directory so that the vignette stays self-contained.

```{r optional-report}
report_path <- file.path(tempdir(), "injection_molding_doe_report.html")
tryCatch(
  task$report(format = "html", path = report_path),
  error = function(e) {
    message("Report rendering skipped: ", conditionMessage(e))
  }
)
file.exists(report_path)
```

## Customising plots

Because `task$plot()` returns a standard `ggplot` object (or a
`patchwork` composition for the multi-panel types), it can be
post-processed with the usual `+` operator before being printed. The
chunk below overlays a dashed reference line at the customer
specification limit for `Shrinkage` ($1.5\%$) on top of the
`Shrinkage` main-effects plot, so that the engineer can see at a
glance which factor effects are large enough to move the mean across
the spec boundary.

```{r customise-plot,,fig.width = 9, fig.height = 8}
p_main <- task$plot(type = "main_effects")
print(
  p_main +
    geom_hline(yintercept = 1.5, linetype = "dashed", colour = "red") +
    labs(subtitle = "Red line = customer spec limit (1.5%)")
)
```

**Decision.** The dashed red line marks the $1.5\%$ customer
specification on `Shrinkage`. The main-effect bars are effect
estimates (differences between the high- and low-level means), so the
line is informative as a visual scale reference rather than as a
literal acceptance limit. The plot shows that several individual
effects are larger than the spec gap, which is why the high level of
every factor is required to pull the process mean below $1.5\%$.

## Combining plots with patchwork

When the analyst needs a side-by-side comparison of two `task$plot()`
outputs --- for example, the same diagnostic for both responses --- the
**patchwork** package composes them into a single figure. The
`wrap_elements()` helper wraps any plot object so that it can be
combined with the `|` (side-by-side) or `/` (stacked) operators. The
chunk below places the `Shrinkage` and `DimensionDev` residual
diagnostics next to each other.

```{r patchwork-combine, fig.width = 12, fig.height =8}
p_shrinkage <- task$plot(type = "residual")
p_dimdev    <- task$plot(type = "residual", response_name = "DimensionDev")

combined <- wrap_elements(p_shrinkage) | wrap_elements(p_dimdev)
print(combined)
```

**Decision.** Side-by-side comparison confirms that the residual
diagnostics for both responses are well-behaved. Neither response
shows curvature, funnel, or heavy-tail problems, so the same
first-order-with-interaction model is adequate for both. The combined
figure is suitable for a single-slide summary in a design review.

## Exporting a report

The `task$report()` method renders a self-contained report from an
internal template. Two formats are supported: `excel` (a multi-sheet
workbook with the design table, ANOVA results, effects, and fit
statistics) and `html` (a formatted report with embedded figures).
The example below uses `tempfile()` so that no files are written
outside the session's temporary directory.

```{r export-report, eval = FALSE}
# Not run:
task$report(format = "excel", path = tempfile(fileext = ".xlsx"))
task$report(format = "html",  path = tempfile(fileext = ".html"))
```

The HTML rendering is demonstrated end-to-end in the *Optional: HTML
report* subsection above; the Excel export follows the same call
signature and produces a workbook that can be opened directly in a
spreadsheet application for further downstream analysis.

# Fractional Factorial Designs

When $k$ grows, the size of a $2^k$ design grows exponentially. Fractional
factorial designs exploit the **effect sparsity** principle --- in most
industrial systems only a handful of effects are active --- so a fraction of
the full design is sufficient to estimate the important ones.

A $2^{k-p}$ fractional design uses $1/2^p$ of the runs of the full design.
Its behaviour is governed by the **defining relation**
$I = W_1 = W_2 = \dots = W_{2^p - 1}$, where the $W_i$ are called *words*.
The **resolution** $R$ is the length of the shortest word.

| Resolution | Main effects aliased with | Two-factor interactions aliased with |
|------------|---------------------------|--------------------------------------|
| III | Two-factor interactions | Each other |
| IV | Three-factor interactions | Each other |
| V  | Four-factor interactions | Three-factor interactions |

A short example using `factorial_design()` with `design_type = "fractional"`
and six factors illustrates the API. With six factors at resolution IV, the
package generates a 16-run design.

```{r fractional-example}
factors_6 <- lapply(1:6, function(i) {
  list(name = paste0("X", i), type = "continuous", levels = c(-1, 1))
})

frac_task <- factorial_design(
  factors     = factors_6,
  design_type = "fractional",
  resolution  = "IV",
  randomize   = TRUE,
  seed        = 2024
)

frac_task$summary()
head(frac_task$results$design_info, 6)
nrow(frac_task$results$design_info)
```

> **Note on implementation status.** The fractional-factorial generator
> currently shipped with **iQualityR.doe** is a *simplified* one: it subsets
> a full factorial by systematic sampling rather than constructing a proper
> alias structure from independent generators. This is adequate for
> teaching the API and for rough screening, but the resulting alias pattern
> is not guaranteed to match the textbook resolution-IV aliasing. For
> production work that depends on the alias structure, prefer
> `DoE.base::fac.design()` or `FrF2::FrF2()`, then attach the resulting
> design table to an `IqrDoeTask` for analysis and reporting.

# Screening Designs

When the number of candidate factors is large (say, six or more) and the
experimental budget is tight, a **screening design** is used to identify
the few active factors among many candidates. The
`screening_design()` constructor is a convenience wrapper around
`factorial_design(design_type = "fractional")` that defaults to a
resolution-IV fraction.

```{r screening-example}
factors_8 <- lapply(1:8, function(i) {
  list(name = paste0("F", i), type = "continuous", levels = c(-1, 1))
})

screen_task <- screening_design(
  factors    = factors_8,
  resolution = "IV",
  randomize  = TRUE,
  seed       = 2024
)

screen_task$summary()
nrow(screen_task$results$design_info)
```

The same `$compute()`, `$plot()`, and `$report()` methods are available on
the returned `IqrDoeTask` object once response data has been attached via
`task$data <- df`. The workflow is therefore identical to the one
demonstrated in Section 3 for the full factorial case.

# Best Practices

1. **Randomise the run order.** Always set `randomize = TRUE` (the default)
   and supply a `seed` for reproducibility. Randomisation protects against
   lurking time-dependent variables such as tool wear or ambient
   temperature drift.

2. **Replicate when you need error degrees of freedom.** A single
   replicate of a $2^k$ design saturates the full-interaction model and
   leaves no residual degrees of freedom for the $F$-test. Two replicates,
   as used in this vignette, give a clean pure-error estimate.

3. **Use numeric factor levels.** The `IqrDoePlan` validator requires
   `levels` to be a numeric vector of length $\geq 2$. Use `c(40, 80)`
   rather than `c("low", "high")`.

4. **Attach responses through `task$data`.** Build the response data frame
   by adding response columns to the design table returned by
   `task$results$design_info`, then assign it to `task$data` and call
   `task$compute()` again. Do **not** hand-write `lm()` calls --- the task
   API takes care of formula construction, effect extraction, and model-fit
   statistics in a single step.

5. **Validate with confirmation runs.** A statistically significant model
   is not a validated model. Run three to five confirmation trials at the
   predicted optimum and check that their mean falls inside the model's
   prediction interval before releasing the new recipe to production.

6. **Add center points when curvature is plausible.** Four to six center
   points provide a single-degree-of-freedom curvature test that is
   independent of replication. If the curvature test is significant,
   augment the design with axial runs to form a Central Composite Design.

# Comparison with Other R Packages

| Feature | iQualityR.doe | DoE.base | FrF2 | rsm |
|---------|---------------|----------|------|-----|
| Full factorial design | yes | yes | yes | no |
| Fractional factorial design | yes (simplified) | yes | yes | no |
| Orthogonal arrays | yes | yes | no | no |
| Response-surface design | yes | partial | no | yes |
| Built-in main-effects / interaction / residual plots | yes | limited | limited | yes |
| One-call HTML report | yes | no | no | no |
| Object-oriented task API (`$compute`, `$plot`, `$report`) | yes | no | no | no |

The principal value of **iQualityR.doe** is the integration of design
generation, statistical analysis, plotting, and reporting in a single
object-oriented interface (`IqrDoeTask`), which suits quality engineers who
need a reproducible workflow without assembling several low-level packages.
For specialised tasks that depend on a guaranteed alias structure
(regular fractional factorials, Plackett-Burman, definitive screening
designs), `DoE.base` and `FrF2` remain the reference implementations and
can be used alongside **iQualityR.doe** by attaching their design tables to
an `IqrDoeTask` for analysis.

# References

1. Montgomery, D. C. (2019). *Design and Analysis of Experiments*
   (10th ed.). Wiley.
2. Box, G. E. P., Hunter, W. G., & Hunter, J. S. (2005). *Statistics for
   Experimenters* (2nd ed.). Wiley.
3. Wu, C. F. J., & Hamada, M. S. (2009). *Experiments: Planning,
   Analysis, and Optimization* (2nd ed.). Wiley.
4. Yates, F. (1937). *The Design and Analysis of Factorial Experiments*.
   Imperial Bureau of Soil Science.
5. Taguchi, G., Chowdhury, S., & Wu, Y. (2005). *Taguchi's Quality
   Engineering Handbook*. Wiley.

# sessionInfo

```{r sessioninfo}
sessionInfo()
```
