---
title: "Multi-Response Optimization: A Food Extrusion Case Study"
author: "iQualityR Team"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Multi-Response Optimization: A Food Extrusion Case Study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(iQualityR.doe)
library(ggplot2)
library(patchwork)
```

# Introduction

Industrial processes are almost never characterised by a single quality
attribute. A snack product must be *expansive* (light texture), *firm
enough* (not mushy, not jaw-breaking), and *liked* by the sensory panel.
These responses conflict with one another: the high-temperature,
low-moisture conditions that maximise expansion also drive hardness away
from its target and can burn the product. Improving one response tends to
worsen another, so there is no single setting that is simultaneously
optimal for every response.

`iQualityR.doe` provides three complementary tools for this class of
problem:

| Tool | Method | Output | Best for |
|------|--------|--------|----------|
| `MultiResponseOptimizer$optimize_desirability()` | Derringer--Suich desirability functions aggregated by a weighted geometric mean | A single recommended setting | Engineering practice; one recipe to release |
| `MultiResponseOptimizer$find_pareto_frontier()` | Non-dominated sorting of candidate predictions | A set of trade-off candidates | R&D exploration of the trade-off surface |
| `MultiResponseOptimizer$rank_with_topsis()` | TOPSIS multi-criteria decision making | A ranked short-list | Selecting a defensible winner from the frontier |

This vignette follows a complete engineering workflow on a realistic
food-extrusion case: we design the experiment, simulate the trial data,
fit response-surface models, optimise the desirability index, explore the
Pareto frontier, rank the candidates with TOPSIS, and finally run
confirmation trials at the recommended setting. Every figure is generated
by code the reader can re-run, and no chunk hides its source.

# API Overview

The `IqrDoeTask` class is the central object in **iQualityR.doe**. Its
life-cycle is:

```r
task <- rsm_design(...)         # or factorial_design / screening_design / taguchi_design
task$data <- df                  # attach the observed responses
task$compute()                  # fit ANOVA + second-order lm models
task$summary()                  # print the analysis
task$plot(type = "...")         # generate figures
task$report(format = "excel", path = "...")  # export a report
```

## The `plot()` method

```r
task$plot(type = "design", theme = NULL, ...)
```

| Argument        | Purpose                                                                 |
|-----------------|-------------------------------------------------------------------------|
| `type`          | One of the eight plot types listed below.                              |
| `theme`         | Optional `IqrTheme` object (or theme name) to override the task theme. |
| `...`           | Forwarded to the plotter. For response-surface types this includes `response_name`, `x_var`, and `y_var`. |

The eight supported `type` values are:

| `type`               | Returns                 | Requires 2nd-order model | Typical use                                       |
|----------------------|-------------------------|---------------------------|---------------------------------------------------|
| `"design"`           | `ggplot`                | No                        | Inspect the run layout (run order tile view).     |
| `"main_effects"`     | `ggplot` / `patchwork`  | No                        | Marginal effect of each factor.                   |
| `"interaction"`      | `ggplot` / `patchwork`  | No                        | Two-factor interaction panels.                    |
| `"residual"`         | `patchwork`             | No                        | Standard residual diagnostics (QQ, fitted, scale). |
| `"full"`             | `patchwork`             | No                        | Combined main_effects + interaction + residual.   |
| `"response_surface"` | `list` (see below)      | **Yes**                   | Full RSM panel (surface + contour + 3D + plotly).  |
| `"surface"`          | `ggplot`                | **Yes**                   | 2D tile + contour surface only.                   |
| `"contour"`          | `ggplot`                | **Yes**                   | Filled contour only.                              |

For `"response_surface"`, the returned list contains `surface_plot`,
`contour_plot`, `combined_plot`, and `interactive_plot` (the last is
`NULL` when the optional `plotly` package is not installed). The
`response_name` argument selects which response to plot (defaults to
the first); `x_var` and `y_var` select the two factors shown on the
axes (default to the first two numeric factors).

> **Important: invisible returns.** Like `plot()` on base R objects,
> `task$plot()` returns its value *invisibly*. In an interactive R
> session the figure auto-prints, but inside `rmarkdown` / `for`
> loops / functions you must call `print()` explicitly:

```r
p <- task$plot(type = "contour", response_name = "ExpansionRatio")
print(p)            # renders the figure inside an R Markdown chunk
```

## The `MultiResponseOptimizer` class

Beyond the standard `IqrDoeTask` plots, the `MultiResponseOptimizer`
R6 class exposes six public methods for multi-response work:

| Method                                  | Purpose                                                            |
|-----------------------------------------|--------------------------------------------------------------------|
| `compute_individual_desirability(...)`  | Per-response desirability $d_i$ from spec and goal.                |
| `compute_overall_desirability(...)`     | Aggregate $D = (\prod d_i^{w_i})^{1/\sum w_i}$ over responses.    |
| `optimize_desirability(...)`            | Grid or Nelder-Mead search for the maximum of $D$.                 |
| `find_pareto_frontier(...)`             | Non-dominated sorting of candidate predictions.                    |
| `rank_with_topsis(...)`                 | TOPSIS multi-criteria ranking of candidate solutions.              |
| `plot_pareto_frontier(...)`             | Built-in ggplot visualisation of the Pareto frontier.               |

The rest of this vignette exercises every `type` value and every
`MultiResponseOptimizer` method on a single case study.

# Case Study: Corn Snack Extrusion

A snack-food plant produces a corn-based extruded snack. The quality team
must tune the extruder so that the product is simultaneously light, has the
right bite, and scores well with the sensory panel. Three process factors
are controllable:

| Factor | Symbol | Low | High | Unit |
|--------|--------|-----|------|------|
| Barrel temperature | `Temperature` | 140 | 180 | $^\circ$C |
| Screw speed | `ScrewSpeed` | 150 | 250 | rpm |
| Feed moisture | `Moisture` | 14 | 22 | % |

Three responses are measured on every run:

| Response | Symbol | Goal | Spec range | Target |
|----------|--------|------|------------|--------|
| Expansion ratio | `ExpansionRatio` | **maximise** | 2.0 -- 3.8 | ideal $\geq 3.8$ |
| Hardness | `Hardness` | **target** | 15 -- 45 N | 30 N |
| Overall liking | `OverallLiking` | **maximise** | 4.0 -- 7.5 | ideal $\geq 7.5$ |

Expansion ratio and overall liking both want hot, fast, dry conditions;
hardness, by contrast, is driven *away* from its 30 N target under those
same conditions (the product becomes too soft). The optimisation must
trade the three goals off against one another in a principled way.

# Experimental Design

A **Central Composite Design (CCD)** with three factors is the standard
response-surface layout for this problem. With `alpha = "rotatable"` the
axial points sit on a sphere of radius $(2^k)^{1/4} = 1.682$, which makes
the prediction variance invariant to direction at the centre --- the
property that makes the design "rotatable". The design combines
$2^3 = 8$ factorial points, $2k = 6$ axial points, and 6 centre points,
for 20 runs total.

```{r factors}
factors <- list(
  list(name = "Temperature", type = "continuous", levels = c(140, 180)),
  list(name = "ScrewSpeed", type = "continuous", levels = c(150, 250)),
  list(name = "Moisture",    type = "continuous", levels = c(14,  22))
)
```

```{r build-design}
task <- rsm_design(
  factors       = factors,
  design_type   = "ccd",
  response_vars = c("ExpansionRatio", "Hardness", "OverallLiking"),
  center_points = 6,
  alpha         = "rotatable",
  randomize     = TRUE,
  seed          = 42
)

design_df <- task$results$design_info
nrow(design_df)
head(design_df, 8)
```

The `alpha = "rotatable"` choice is resolved inside the package to
$(2^3)^{1/4} \approx 1.682$, so the axial runs of `Temperature` reach
approximately $160 \pm 1.682 \times 20 = (126.4,\ 193.6)^\circ$C, well
beyond the cube faces. This is intentional: the axial points stabilise the
quadratic fit at the edge of the design region.

A scatter-plot matrix of the design makes the geometry explicit: the
$2^3$ cube, the six star points along each axis, and the six replicated
centre points.

```{r design-plot, fig.width = 7.5, fig.height = 7.5, fig.cap = "Central composite design in the three factor axes. Cube points (grey), axial points (orange), and centre points (blue) span the design region."}
design_df$PointType <- with(design_df, {
  on_axis <- (Temperature == 160 & ScrewSpeed == 200) |
             (Moisture    == 18  & ScrewSpeed == 200) |
             (Temperature == 160 & Moisture    == 18)
  is_center <- Temperature == 160 & ScrewSpeed == 200 & Moisture == 18
  ifelse(is_center, "Center",
    ifelse(on_axis & !is_center, "Axial", "Cube"))
})

panel_plot <- function(df, fx, fy, ...) {
  ggplot(df, aes(.data[[fx]], .data[[fy]], color = PointType)) +
    geom_point(size = 2.6, alpha = 0.85) +
    scale_color_manual(values = c(Cube = "#999999",
                                  Axial = "#E69F00",
                                  Center = "#0072B2")) +
    theme_minimal(base_size = 10) +
    theme(legend.position = "bottom")
}

panel_plot(design_df, "Temperature", "ScrewSpeed") +
  panel_plot(design_df, "Temperature", "Moisture") +
  panel_plot(design_df, "ScrewSpeed",  "Moisture") +
  plot_layout(ncol = 2)
```

# Response Data

In a real study the responses would be measured after the physical trials.
Here we simulate them from a known physics-inspired second-order model so
that the analysis is fully reproducible. The data-generating functions are
defined once and reused later for the confirmation trials.

```{r simulate}
# True process: second-order surfaces in coded variables.
# x1 = (T  - 160) / 20, x2 = (SS - 200) / 50, x3 = (M  - 18) / 4
true_ExpansionRatio <- function(T, SS, M) {
  x1 <- (T  - 160) / 20; x2 <- (SS - 200) / 50; x3 <- (M  - 18) / 4
  3.00 + 0.45*x1 + 0.20*x2 - 0.35*x3
       - 0.12*x1^2 - 0.08*x2^2 - 0.10*x3^2
       + 0.05*x1*x2 - 0.08*x1*x3
}
true_Hardness <- function(T, SS, M) {
  x1 <- (T  - 160) / 20; x2 <- (SS - 200) / 50; x3 <- (M  - 18) / 4
  30.0 - 4.0*x1 - 2.0*x2 + 5.0*x3
        + 1.5*x1^2 + 1.0*x3^2 - 0.5*x1*x3
}
true_OverallLiking <- function(T, SS, M) {
  x1 <- (T  - 160) / 20; x2 <- (SS - 200) / 50; x3 <- (M  - 18) / 4
  6.00 + 0.80*x1 + 0.30*x2 - 0.50*x3
        - 0.35*x1^2 - 0.20*x2^2 - 0.25*x3^2
        + 0.10*x1*x2 - 0.12*x1*x3
}

set.seed(42)
n <- nrow(design_df)
design_df$ExpansionRatio <- true_ExpansionRatio(design_df$Temperature,
                                                design_df$ScrewSpeed,
                                                design_df$Moisture) +
                            rnorm(n, sd = 0.08)
design_df$Hardness       <- true_Hardness(design_df$Temperature,
                                          design_df$ScrewSpeed,
                                          design_df$Moisture) +
                            rnorm(n, sd = 0.8)
design_df$OverallLiking  <- true_OverallLiking(design_df$Temperature,
                                               design_df$ScrewSpeed,
                                               design_df$Moisture) +
                            rnorm(n, sd = 0.25)

# Attach the measured responses to the task and recompute.
task$data <- design_df
```

```{r preview-data}
head(task$data, 6)
```

The simulated responses carry the intended correlation structure:
high-expansion runs (hot, fast, dry) tend to be too soft (hardness below
the 30 N target), which is exactly the conflict the optimiser must
resolve.

# Response Surface Modelling

Calling `task$compute()` re-runs the design (reproducibly, because the
seed was fixed) and, because `task$data` now holds the measured responses,
fits a full second-order response-surface model for each response:

$$
y \sim \text{Temperature} \cdot \text{ScrewSpeed} \cdot \text{Moisture}
      + I(\text{Temperature}^2) + I(\text{ScrewSpeed}^2) + I(\text{Moisture}^2).
$$

```{r compute}
task$compute()
task$summary()
```

```{r fit-stats}
fit_stats <- data.frame(
  Response = c("ExpansionRatio", "Hardness", "OverallLiking"),
  R2       = sapply(task$results$anova_results, function(r) r$model_fit$r_squared),
  R2_adj   = sapply(task$results$anova_results, function(r) r$model_fit$adj_r_squared),
  LoF_p    = sapply(task$results$anova_results, function(r) r$model_fit$lack_of_fit_p)
)
fit_stats
```

All three models fit well (adjusted $R^2$ above $0.95$) and none shows a
significant lack of fit, so the fitted surfaces are reliable enough to
search for an optimum.

The residual diagnostics below confirm the standard regression
assumptions: residuals scattered symmetrically around zero with no
systematic curvature or funnel shape.

```{r residual-plot, fig.width = 9, fig.height = 3.5, fig.cap = "Residuals versus fitted values for the three response-surface models. No structure is visible; the assumptions hold."}
# Extract the per-response models for later use (desirability, Pareto, etc.)
model_ER <- task$results$anova_results$ExpansionRatio$model
model_H  <- task$results$anova_results$Hardness$model
model_OL <- task$results$anova_results$OverallLiking$model

# Residual diagnostics via the package API.
p_resid_ER <- task$plot(type = "residual", response_name = "ExpansionRatio")
p_resid_H  <- task$plot(type = "residual", response_name = "Hardness")
p_resid_OL <- task$plot(type = "residual", response_name = "OverallLiking")

p_resid_ER + p_resid_H + p_resid_OL + plot_layout(ncol = 3)
```

# Desirability Optimisation

## Specifying the desirability functions

Each response is mapped onto $[0, 1]$ with a Derringer--Suich desirability
function. Expansion ratio and overall liking use the *larger-the-better*
form; hardness uses the *nominal-the-best* form with target 30 N. The
weights encode the relative business priority: expansion is the primary
driver of perceived quality, so it receives weight 0.4; hardness and
liking each receive 0.3.

```{r specs}
specs <- list(
  list(type = "max",    lower = 2.0, upper = 3.8, target = NA,  weight = 0.4, s = 1),
  list(type = "target", lower = 15,  upper = 45,  target = 30,  weight = 0.3, s = 1),
  list(type = "max",    lower = 4.0, upper = 7.5, target = NA,  weight = 0.3, s = 1)
)

bounds <- list(
  Temperature = c(140, 180),
  ScrewSpeed  = c(150, 250),
  Moisture    = c(14,  22)
)
```

## Grid search for the overall desirability

`$optimize_desirability()` evaluates the overall desirability index $D$
(the weighted geometric mean of the individual $d_i$) on a 25-by-25-by-25
lattice over the factor cube and returns the best setting. The geometric
mean is the key design choice: a single unacceptable response ($d_i = 0$)
collapses $D$ to zero, so the optimum can never sacrifice one response to
buy gains on the others.

```{r optimize}
mro <- MultiResponseOptimizer$new()

opt <- mro$optimize_desirability(
  models          = list(model_ER, model_H, model_OL),
  specs           = specs,
  bounds          = bounds,
  method          = "grid",
  grid_resolution = 25
)

opt$best_params
opt$best_D
opt$n_evaluations
```

```{r optimal-predictions}
opt_point <- as.data.frame(opt$best_params)
pred_at_opt <- data.frame(
  Response = c("ExpansionRatio", "Hardness", "OverallLiking"),
  Predicted = c(predict(model_ER, opt_point),
                predict(model_H,  opt_point),
                predict(model_OL, opt_point))
)
pred_at_opt
```

The recommended setting pushes temperature and screw speed high (for
expansion and liking) while lifting moisture just enough to keep hardness
within its acceptable band. This is exactly the compromise the geometric
mean was designed to enforce.

## Desirability landscape

To see *why* the optimiser landed here, we slice the overall desirability
surface at the optimal moisture and plot $D$ as a function of temperature
and screw speed. The red star marks the optimum.

```{r desirability-landscape, fig.cap = "Overall desirability D as a function of temperature and screw speed, sliced at the optimal moisture. The red star is the grid-search optimum."}
slice <- expand.grid(
  Temperature = seq(140, 180, length.out = 60),
  ScrewSpeed  = seq(150, 250, length.out = 60)
)
slice$Moisture <- opt$best_params$Moisture

pER <- predict(model_ER, newdata = slice)
pH  <- predict(model_H,  newdata = slice)
pOL <- predict(model_OL, newdata = slice)

dER <- mro$compute_individual_desirability(pER, type = "max",    lower = 2.0, upper = 3.8)
dH  <- mro$compute_individual_desirability(pH,  type = "target", lower = 15,  upper = 45,  target = 30)
dOL <- mro$compute_individual_desirability(pOL, type = "max",    lower = 4.0, upper = 7.5)
slice$D <- mro$compute_overall_desirability(list(dER, dH, dOL), weights = c(0.4, 0.3, 0.3))

opt_pt <- data.frame(
  Temperature = opt$best_params$Temperature,
  ScrewSpeed  = opt$best_params$ScrewSpeed
)

ggplot(slice, aes(Temperature, ScrewSpeed)) +
  geom_tile(aes(fill = D)) +
  geom_contour(aes(z = D), color = "white", alpha = 0.5, bins = 10) +
  geom_point(data = opt_pt, aes(Temperature, ScrewSpeed),
             color = "#D7263D", size = 4, shape = 17) +
  scale_fill_viridis_c(limits = c(0, 1)) +
  labs(title = sprintf("Overall desirability D at Moisture = %.2f%%", opt$best_params$Moisture),
       x = "Temperature (deg C)", y = "Screw speed (rpm)", fill = "D") +
  theme_minimal(base_size = 11)
```

## Response-surface contours

The three panels below show each fitted response surface, again sliced at
the optimal moisture. They make the trade-off visible: expansion and
liking both rise toward the upper-right (hot, fast), while hardness falls
*below* its 30 N target in that same corner --- the product would be too
soft. The optimum (red star) sits at the edge of the high-expansion region
where hardness is still acceptable.

```{r rsm-contours, fig.width = 9, fig.height = 3.7, fig.cap = "Fitted response surfaces at the optimal moisture. Red star: desirability optimum. Dashed line on the Hardness panel: 30 N target."}
# Use the package contour plotter for each response, then overlay the
# desirability optimum (red star) and the hardness target reference line.
# `inherit.aes = FALSE` keeps the overlay layers independent of the
# contour's internal `z = predicted` aesthetic.
p1 <- task$plot(type = "contour",
                response_name = "ExpansionRatio",
                x_var = "Temperature", y_var = "ScrewSpeed") +
  geom_point(data = opt_pt, aes(Temperature, ScrewSpeed),
             color = "#D7263D", size = 3, shape = 17, inherit.aes = FALSE)

p2 <- task$plot(type = "contour",
                response_name = "Hardness",
                x_var = "Temperature", y_var = "ScrewSpeed") +
  geom_point(data = opt_pt, aes(Temperature, ScrewSpeed),
             color = "#D7263D", size = 3, shape = 17, inherit.aes = FALSE) +
  geom_hline(yintercept = 30, linetype = "dashed", color = "grey40")

p3 <- task$plot(type = "contour",
                response_name = "OverallLiking",
                x_var = "Temperature", y_var = "ScrewSpeed") +
  geom_point(data = opt_pt, aes(Temperature, ScrewSpeed),
             color = "#D7263D", size = 3, shape = 17, inherit.aes = FALSE)

p1 + p2 + p3 + plot_layout(ncol = 3)
```

# Plotting with IqrDoeTask

The case-study `task` is now computed and carries three fitted
second-order models, so every `type` value listed in the API Overview
can be exercised on it. The chunks below use the recommended
`print(task$plot(...))` form so they render correctly inside
`rmarkdown`; in an interactive session the bare `task$plot(...)` call
would auto-print the same figure.

## Design layout

```{r plot-design, fig.cap = "Run layout of the central composite design produced by task$plot(type = 'design')."}
print(task$plot(type = "design"))
```

**Decision.** The design tile view confirms that all 20 runs of the CCD
(factorial + axial + centre points) are present and that the run order
was randomised. Use this plot first to spot missing runs or
over-replicated centre points before fitting any model.

## Main effects

```{r plot-main-er, fig.cap = "Main-effect panel for ExpansionRatio."}
print(task$plot(type = "main_effects", response_name = "ExpansionRatio"))
```

**Decision.** Expansion ratio rises steeply with temperature and screw
speed and falls with moisture. The marginal effect of temperature is the
largest, which justifies giving temperature the central role in the
process recipe.

```{r plot-main-h, fig.cap = "Main-effect panel for Hardness."}
print(task$plot(type = "main_effects", response_name = "Hardness"))
```

**Decision.** Hardness is driven in the opposite direction to
expansion: high temperature and low moisture push it below the 30 N
target. This is the first visual confirmation that the expansion goal
and the hardness goal conflict.

```{r plot-main-ol, fig.cap = "Main-effect panel for OverallLiking."}
print(task$plot(type = "main_effects", response_name = "OverallLiking"))
```

**Decision.** Overall liking tracks expansion closely --- hot, fast, dry
conditions are preferred by the sensory panel. Because liking and
expansion agree, the multi-response optimiser can safely weight them
together against hardness.

## Interaction panels

```{r plot-interaction-er, fig.cap = "Two-factor interaction panel for ExpansionRatio from the second-order model."}
print(task$plot(type = "interaction", response_name = "ExpansionRatio"))
```

**Decision.** The Temperature x Moisture interaction is the strongest
for expansion: the gain in expansion from raising temperature is
amplified when moisture is low. Operationally this means the extruder
must control moisture tightly when running hot, otherwise the expansion
benefit is lost.

## Full diagnostics panel

```{r plot-full-er, fig.width = 9, fig.height = 7, fig.cap = "Combined main-effects + interaction + residual panel for ExpansionRatio from task$plot(type = 'full')."}
print(task$plot(type = "full", response_name = "ExpansionRatio"))
```

**Decision.** The `full` plot bundles the three diagnostic families
into a single patchwork. Use it as a one-shot model-validation figure
in reports: it shows the effect structure *and* the residual behaviour
without requiring the reader to cross-reference several plots.

## Response-surface (tile + contour) plot

```{r plot-surface-er, fig.cap = "2D tile + contour surface for ExpansionRatio as a function of Temperature and ScrewSpeed."}
print(task$plot(type = "surface", response_name = "ExpansionRatio",
                x_var = "Temperature", y_var = "ScrewSpeed"))
```

**Decision.** The `surface` plot adds the tile heat-map underneath the
contour lines, making the gradient magnitude easier to read than the
contour-only view. The ridge of high expansion runs along the
upper-right corner (hot, fast), confirming the main-effects reading.

## Filled contour plot

```{r plot-contour-h-moisture, fig.cap = "Filled contour of Hardness as a function of Temperature and Moisture (ScrewSpeed held at its default slice). The 30 N target band is visible in the middle of the plane."}
print(task$plot(type = "contour", response_name = "Hardness",
                x_var = "Temperature", y_var = "Moisture"))
```

**Decision.** Slicing Hardness across Temperature and Moisture (instead
of ScrewSpeed) shows that the 30 N target is reachable only along a
narrow diagonal band. Any drift toward hot-and-dry pushes hardness
below target, which is exactly the regime that maximises expansion.

## Response-surface panel list

`type = "response_surface"` returns a *list* of plots rather than a
single ggplot, so the chunk captures the list and then `print()`s the
combined panel:

```{r plot-rsm-list, fig.width = 9, fig.height = 7, fig.cap = "Combined response-surface panel for OverallLiking produced by task$plot(type = 'response_surface')$combined_plot."}
rsm_panels <- task$plot(type = "response_surface",
                        response_name = "OverallLiking",
                        x_var = "ScrewSpeed", y_var = "Moisture")
print(rsm_panels$combined_plot)
```

**Decision.** The combined panel exposes the surface, the contour, and
the 3-D view of OverallLiking in one figure. When `plotly` is
installed, `rsm_panels$interactive_plot` is also available for an HTML
rotation; otherwise it is `NULL`, in which case `combined_plot` is the
figure to render. The liking surface peaks in the same corner as
expansion, which is why the desirability optimum favours hot, fast, dry
settings.

## Customising plots

Because `task$plot()` returns a standard `ggplot` object (for the
single-figure types), the user can keep adding ggplot2 layers. The
chunk below takes a contour plot of ExpansionRatio and overlays the
desirability optimum found by `optimize_desirability()` in the previous
section.

```{r plot-customise, fig.cap = "Default contour of ExpansionRatio customised with a red star marking the desirability optimum. Adding layers to the returned ggplot is the standard way to annotate package plots."}
p_custom <- task$plot(type = "contour", response_name = "ExpansionRatio",
                      x_var = "Temperature", y_var = "ScrewSpeed")
p_custom <- p_custom +
  geom_point(data = opt_point, aes(Temperature, ScrewSpeed),
             color = "#D7263D", size = 4, shape = 17, inherit.aes = FALSE) +
  labs(subtitle = "Red star: desirability optimum")
print(p_custom)
```

**Decision.** The optimum sits on the high-expansion ridge, just before
the contour curvature flattens. Overlaying the optimum on the response
surface is the cleanest way to communicate *why* the optimiser picked
this setting to a non-statistical audience.

## Combining plots with patchwork

The single-figure types return `ggplot` objects, so they compose
naturally with `patchwork`. The chunk below places the ExpansionRatio
contour next to the Hardness contour using `patchwork::wrap_elements()`,
which is the safe way to embed pre-built ggplot objects into a larger
composition.

```{r plot-patchwork, fig.width = 10, fig.height = 4.5, fig.cap = "Side-by-side comparison of the ExpansionRatio and Hardness contours, composed with patchwork::wrap_elements()."}
p_er <- task$plot(type = "contour", response_name = "ExpansionRatio",
                  x_var = "Temperature", y_var = "ScrewSpeed")
p_h  <- task$plot(type = "contour", response_name = "Hardness",
                  x_var = "Temperature", y_var = "ScrewSpeed")

combined <- patchwork::wrap_elements(p_er) +
  patchwork::wrap_elements(p_h) +
  plot_layout(ncol = 2) +
  plot_annotation(title = "ExpansionRatio vs. Hardness at the optimal moisture",
                  tag_levels = "A")
print(combined)
```

**Decision.** Putting the two contours side-by-side makes the trade-off
visible at a glance: the upper-right corner that maximises expansion is
the same corner where hardness drops below the 30 N target. This
single figure is usually enough to justify the multi-response
optimisation that follows.

# Pareto Frontier Analysis

The desirability optimum is a single point. Before committing to it, the
quality team wants to see the *trade-off surface*: which other settings
are non-dominated, meaning no other setting is simultaneously better on
every response. This is the **Pareto frontier**.

We build a candidate grid over the factor cube, predict the three
responses with the fitted models, and call `$find_pareto_frontier()`.
Because the method accepts only `"max"` and `"min"` directions, the
hardness target is recast as a *minimise-distance-from-target* response
$|H - 30|$, which is mathematically equivalent for dominance purposes.

```{r pareto-candidates}
cand_grid <- expand.grid(
  Temperature = seq(140, 180, length.out = 12),
  ScrewSpeed  = seq(150, 250, length.out = 12),
  Moisture    = seq(14,  22,  length.out = 12)
)
cand_grid$ExpansionRatio <- predict(model_ER, newdata = cand_grid)
cand_grid$Hardness       <- predict(model_H,  newdata = cand_grid)
cand_grid$OverallLiking  <- predict(model_OL, newdata = cand_grid)
cand_grid$HardnessDist   <- abs(cand_grid$Hardness - 30)

cand_responses <- cand_grid[, c("ExpansionRatio", "HardnessDist", "OverallLiking")]
rownames(cand_responses) <- seq_len(nrow(cand_responses))

pareto <- mro$find_pareto_frontier(
  cand_responses,
  directions = c("max", "min", "max")
)

pareto$n_solutions
pareto$n_total
pareto$coverage_pct
```

```{r pareto-recover}
# Recover the factor settings for each Pareto solution by matching the
# preserved rownames back to the candidate grid.
pareto_idx <- as.integer(rownames(pareto$pareto_frontier))
pareto_full <- cbind(
  pareto$pareto_frontier,
  cand_grid[pareto_idx, c("Temperature", "ScrewSpeed", "Moisture", "Hardness")]
)
head(pareto_full[, c("Temperature", "ScrewSpeed", "Moisture",
                     "ExpansionRatio", "Hardness", "OverallLiking")], 6)
```

The frontier is a small fraction of the candidate cloud, which is the
signature of a genuine multi-objective trade-off: most settings are
dominated.

```{r pareto-plot, fig.cap = "Pareto frontier in the Expansion-Ratio vs. Hardness plane, coloured by Overall Liking. Grey points are dominated candidates; red points are non-dominated. The dashed line marks the 30 N hardness target."}
dominated_idx <- setdiff(seq_len(nrow(cand_grid)), pareto_idx)

ggplot() +
  geom_point(data = cand_grid[dominated_idx, ],
             aes(ExpansionRatio, Hardness), color = "#CCCCCC", size = 1.6, alpha = 0.7) +
  geom_point(data = pareto_full,
             aes(ExpansionRatio, Hardness, color = OverallLiking),
             size = 3.2) +
  geom_hline(yintercept = 30, linetype = "dashed", color = "grey40") +
  scale_color_viridis_c(option = "C") +
  labs(title = "Pareto frontier of candidate settings",
       subtitle = sprintf("%d non-dominated solutions out of %d candidates",
                          pareto$n_solutions, pareto$n_total),
       x = "Expansion ratio (maximise)", y = "Hardness (N, target 30)",
       color = "Overall\nLiking") +
  theme_minimal(base_size = 11)
```

The frontier slopes from the upper-left (high expansion, but hardness far
from target) to the lower-right (hardness on target, but lower expansion).
Moving along it always trades one response for another; the desirability
optimum and TOPSIS analysis below are two principled ways to pick a point
on it.

## Built-in Pareto plot

`MultiResponseOptimizer$plot_pareto_frontier()` is the package-provided
convenience plotter for the `find_pareto_frontier()` return value. It
takes the `pareto_result` list, the names of two response columns for
the `x_axis` and `y_axis`, and an optional `z_axis` for the colour
mapping. Like `task$plot()`, it returns its value invisibly, so we
`print()` it.

```{r pareto-plot-builtin, fig.cap = "Pareto frontier rendered by mro$plot_pareto_frontier(). Only the non-dominated candidates are shown; the y-axis is the HardnessDist criterion used by find_pareto_frontier() (lower is better)."}
p_pareto_builtin <- mro$plot_pareto_frontier(
  pareto,
  x_axis = "ExpansionRatio",
  y_axis = "HardnessDist",
  z_axis = "OverallLiking"
)
print(p_pareto_builtin)
```

**Decision.** The built-in plot is the fastest way to inspect the
frontier shape during exploration, but it is intentionally minimal: it
shows only the non-dominated points and uses the response *as supplied
to `find_pareto_frontier()`* --- here `HardnessDist` (the distance from
the 30 N target), not `Hardness` itself. The hand-coded `pareto-plot`
chunk above is therefore preferred for the final report, because it
overlays the dominated cloud for context, plots `Hardness` in
engineering units, and adds the 30 N target reference line. None of
those refinements are exposed as arguments of
`plot_pareto_frontier()`. Use the built-in method for quick iteration
and the hand-coded `ggplot` for stakeholder-facing figures.

# TOPSIS Ranking

TOPSIS (Technique for Order Preference by Similarity to Ideal Solution)
ranks each Pareto candidate by its distance to the *positive ideal* (best
value on every response) and the *negative ideal* (worst value on every
response). The relative closeness

$$C_i = \frac{d_i^-}{d_i^+ + d_i^-} \in [0, 1]$$

is higher for candidates that are simultaneously close to the ideal and
far from the worst. We rank the Pareto set with the same business weights
used for the desirability index.

```{r topsis}
ranking <- mro$rank_with_topsis(
  pareto_full[, c("ExpansionRatio", "HardnessDist", "OverallLiking")],
  weights    = c(0.4, 0.3, 0.3),
  directions = c("max", "min", "max")
)

pareto_ranked <- cbind(pareto_full, ranking$ranking[, c("Closeness", "Rank")])
pareto_ranked <- pareto_ranked[order(pareto_ranked$Rank), ]

head(pareto_ranked[, c("Rank", "Temperature", "ScrewSpeed", "Moisture",
                       "ExpansionRatio", "Hardness", "OverallLiking", "Closeness")], 8)
```

```{r topsis-bar, fig.width = 8, fig.height = 4, fig.cap = "TOPSIS relative closeness C_i for the Pareto-optimal candidates. The rank-1 solution is highlighted."}
winner_idx <- ranking$ranking$Candidate[ranking$ranking$Rank == 1]
bar_df <- pareto_ranked[, c("Temperature", "ScrewSpeed", "Moisture", "Closeness", "Rank")]
bar_df$Label <- sprintf("T=%.0f / SS=%.0f / M=%.1f",
                        bar_df$Temperature, bar_df$ScrewSpeed, bar_df$Moisture)
bar_df$Label <- factor(bar_df$Label, levels = bar_df$Label[order(bar_df$Rank)])

ggplot(bar_df, aes(Label, Closeness, fill = Rank == 1)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = c(`TRUE` = "#D7263D", `FALSE` = "#7FB2D9"),
                    labels = c(`TRUE` = "Rank 1", `FALSE` = "Other")) +
  labs(title = "TOPSIS ranking of the Pareto frontier",
       x = "Candidate setting", y = "Relative closeness C_i") +
  theme_minimal(base_size = 10) +
  theme(legend.title = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 7))
```

# Recommendation and Trade-offs

The table below compares the desirability optimum (the single best $D$
over the grid) with the TOPSIS rank-1 solution (the Pareto candidate
closest to the ideal under the business weights).

```{r compare}
winner <- pareto_full[winner_idx, c("Temperature", "ScrewSpeed", "Moisture")]
winner_point <- as.data.frame(winner)

D_at <- function(df) {
  p1 <- predict(model_ER, df); p2 <- predict(model_H, df); p3 <- predict(model_OL, df)
  d1 <- mro$compute_individual_desirability(p1, "max",    2.0, 3.8)
  d2 <- mro$compute_individual_desirability(p2, "target", 15,  45, target = 30)
  d3 <- mro$compute_individual_desirability(p3, "max",    4.0, 7.5)
  mro$compute_overall_desirability(list(d1, d2, d3), weights = c(0.4, 0.3, 0.3))
}

comparison <- data.frame(
  Method       = c("Desirability optimum", "TOPSIS rank-1"),
  Temperature  = c(opt$best_params$Temperature, winner$Temperature),
  ScrewSpeed   = c(opt$best_params$ScrewSpeed,  winner$ScrewSpeed),
  Moisture     = c(opt$best_params$Moisture,    winner$Moisture),
  ExpansionRatio = c(predict(model_ER, opt_point),    predict(model_ER, winner_point)),
  Hardness       = c(predict(model_H,  opt_point),    predict(model_H,  winner_point)),
  OverallLiking  = c(predict(model_OL, opt_point),    predict(model_OL, winner_point)),
  D              = c(opt$best_D,                     D_at(winner_point))
)
comparison
```

The two methods agree closely: both recommend a hot, fast, moderately dry
recipe. The TOPSIS winner is guaranteed to lie on the Pareto frontier
(no other setting beats it on every response at once), which makes it the
more defensible release candidate. We adopt it as the final recommendation
and verify it with confirmation trials.

```{r final-recommendation}
final_setting <- winner_point
final_setting
```

# Confirmation Trials

A confirmation experiment runs a small number of physical trials at the
recommended setting. If the model is correct, the mean of the confirmation
runs should fall inside the $95\%$ prediction interval computed from the
fitted model. We simulate five confirmation runs from the same
data-generating process used in Section 5.

```{r confirmation}
pred_ER_ci <- predict(model_ER, final_setting, interval = "prediction")
pred_H_ci  <- predict(model_H,  final_setting, interval = "prediction")
pred_OL_ci <- predict(model_OL, final_setting, interval = "prediction")

set.seed(99)
confirm_ER <- true_ExpansionRatio(final_setting$Temperature,
                                  final_setting$ScrewSpeed,
                                  final_setting$Moisture) + rnorm(5, sd = 0.08)
confirm_H  <- true_Hardness(final_setting$Temperature,
                            final_setting$ScrewSpeed,
                            final_setting$Moisture) + rnorm(5, sd = 0.8)
confirm_OL <- true_OverallLiking(final_setting$Temperature,
                                 final_setting$ScrewSpeed,
                                 final_setting$Moisture) + rnorm(5, sd = 0.25)

confirmation <- data.frame(
  Response     = c("ExpansionRatio", "Hardness", "OverallLiking"),
  Model_pred   = c(pred_ER_ci[1], pred_H_ci[1],  pred_OL_ci[1]),
  PI_lower     = c(pred_ER_ci[2], pred_H_ci[2],  pred_OL_ci[2]),
  PI_upper     = c(pred_ER_ci[3], pred_H_ci[3],  pred_OL_ci[3]),
  Confirm_mean = c(mean(confirm_ER), mean(confirm_H), mean(confirm_OL)),
  In_PI        = c(mean(confirm_ER) >= pred_ER_ci[2] & mean(confirm_ER) <= pred_ER_ci[3],
                   mean(confirm_H)  >= pred_H_ci[2]  & mean(confirm_H)  <= pred_H_ci[3],
                   mean(confirm_OL) >= pred_OL_ci[2] & mean(confirm_OL) <= pred_OL_ci[3])
)
confirmation
```

```{r confirmation-plot, fig.width = 8, fig.height = 4, fig.cap = "Model prediction with 95% prediction interval (blue) versus the mean of five confirmation trials (red). All three confirmation means fall inside their prediction intervals."}
conf_plot_df <- data.frame(
  Response = rep(c("ExpansionRatio", "Hardness", "OverallLiking"), each = 2),
  Stat     = rep(c("Model prediction", "Confirmation mean"), 3),
  Value    = c(pred_ER_ci[1], mean(confirm_ER),
               pred_H_ci[1],  mean(confirm_H),
               pred_OL_ci[1], mean(confirm_OL)),
  Lower    = c(pred_ER_ci[2], NA, pred_H_ci[2], NA, pred_OL_ci[2], NA),
  Upper    = c(pred_ER_ci[3], NA, pred_H_ci[3], NA, pred_OL_ci[3], NA)
)
conf_plot_df$Response <- factor(conf_plot_df$Response,
                                levels = c("ExpansionRatio", "Hardness", "OverallLiking"))

ggplot(conf_plot_df, aes(Response, Value, color = Stat)) +
  geom_errorbar(data = subset(conf_plot_df, Stat == "Model prediction"),
                aes(ymin = Lower, ymax = Upper), width = 0.15, size = 0.8) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Model prediction" = "#2C7BB6",
                                "Confirmation mean" = "#D7263D")) +
  labs(title = "Confirmation: model prediction vs. physical trials",
       subtitle = "Error bars are 95% prediction intervals",
       x = NULL, y = "Response value", color = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

All three confirmation means fall inside their $95\%$ prediction intervals
and within the engineering specification (expansion $\geq 2.0$, hardness
in $[15, 45]$ N, liking $\geq 4.0$). The model is validated and the
recommended recipe can be released to production.

# Exporting a Report

The `task$report()` method writes a self-contained report of the entire
analysis to disk. Two formats are supported: `format = "excel"` produces
a multi-sheet `.xlsx` workbook (design table, ANOVA tables, coefficient
tables, model-fit summary), and `format = "html"` produces a single
self-contained HTML file with embedded figures. The `path` argument
controls the destination; `tempfile()` is convenient for throw-away
exports during exploration. The chunk is marked `eval = FALSE` because
writing reports to disk is intentionally side-effectful and should not
run on every build.

```{r report, eval = FALSE}
# Not run: writing reports to disk is intentionally side-effectful.
xlsx_path <- tempfile(fileext = ".xlsx")
html_path  <- tempfile(fileext = ".html")

task$report(format = "excel", path = xlsx_path)
task$report(format = "html",  path = html_path)

list.files(dirname(xlsx_path), pattern = basename(c(xlsx_path, html_path)))
```

For programmatic access without writing to disk, the same information
lives in the `task$results` list, which is populated by `task$compute()`
and is what `task$report()` itself reads from. The most useful entries
are:

| Path                                              | Contents                                                            |
|---------------------------------------------------|---------------------------------------------------------------------|
| `task$results$design_info`                        | Data frame of the design runs (factor settings + run order).        |
| `task$results$anova_results`                      | Named list of per-response ANOVA outputs, keyed by response name.   |
| `task$results$anova_results[[r]]$anova`           | ANOVA table for response `r`.                                       |
| `task$results$anova_results[[r]]$model`           | Fitted `lm` (or `rsm`) model object for response `r`.               |
| `task$results$anova_results[[r]]$model_fit`       | List with `r_squared`, `adj_r_squared`, `lack_of_fit_p`, etc.       |
| `task$results$anova_results[[r]]$effects`         | Named list of main and interaction effects for response `r`.        |
| `task$results$effects`                            | Effects of the first response (convenience alias).                  |
| `task$results$model`                              | Fitted model of the first response (convenience alias).             |
| `task$results$model_fit`                          | Model-fit summary of the first response (convenience alias).        |

```{r results-structure}
# Quick programmatic access to the per-response fit statistics,
# reading directly from task$results without writing a report to disk.
data.frame(
  Response = names(task$results$anova_results),
  R2       = sapply(task$results$anova_results, function(r) r$model_fit$r_squared),
  R2_adj   = sapply(task$results$anova_results, function(r) r$model_fit$adj_r_squared),
  row.names = NULL
)
```

**Decision.** Prefer `task$report()` for stakeholder-facing deliverables
(the workbook is self-contained and includes formatting); prefer
`task$results` for any downstream computation that needs the raw fitted
objects (e.g., feeding the `lm` models into `MultiResponseOptimizer`, as
this vignette does in the Desirability Optimisation section).

# Conclusion

This vignette walked through a complete multi-response optimisation on a
realistic food-extrusion case. The workflow was:

1. **Design** a rotatable CCD with `rsm_design()`.
2. **Fit** a second-order response-surface model per response with
   `IqrDoeTask$compute()`.
3. **Optimise** the weighted-geometric-mean desirability index with
   `MultiResponseOptimizer$optimize_desirability()`.
4. **Explore** the trade-off surface with
   `find_pareto_frontier()`.
5. **Rank** the non-dominated candidates with `rank_with_topsis()`.
6. **Confirm** the recommended setting with physical trials, checking that
   the confirmation means fall inside the model's prediction intervals.

The desirability function gave a single recommended setting; the Pareto
frontier exposed the trade-off structure; TOPSIS picked a defensible
winner from the frontier; and the confirmation trials validated the model.
This is the cycle a quality engineer would repeat every time a process
change or a new product is introduced.

# References

1. Derringer, G., & Suich, R. (1980). Simultaneous Optimization of
   Several Response Variables. *Journal of Quality Technology*, 12(4),
   214--219.
2. Myers, R. H., Montgomery, D. C., & Anderson-Cook, C. M. (2016).
   *Response Surface Methodology: Process and Product Optimization Using
   Designed Experiments* (4th ed.). Wiley.
3. Deb, K. (2001). *Multi-Objective Optimization Using Evolutionary
   Algorithms*. Wiley.
4. Hwang, C.-L., & Yoon, K. (1981). *Multiple Attribute Decision Making:
   Methods and Applications*. Springer.
5. Montgomery, D. C. (2019). *Design and Analysis of Experiments*
   (10th ed.), Chapter 12. Wiley.

---

*This vignette was generated by the iQualityR.doe package.*
