---
title: "Response Surface Methodology & Optimization with iQualityR.doe"
author: "iQualityR Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Response Surface Methodology & Optimization with iQualityR.doe}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  fig.align = "center"
)
```

## 1. Introduction

Response Surface Methodology (RSM) is a collection of mathematical and
statistical techniques for empirical model building and optimization.
When screening experiments have identified the vital few factors, RSM
fits a second-order model to map the response surface, locate the
stationary point, and find optimal operating conditions.

`iQualityR.doe` provides a complete RSM workflow that mirrors and extends
Minitab's capabilities:

* **Path of steepest ascent** to move toward the optimum region
* **Central Composite Designs (CCD)** and **Box-Behnken** for second-order fitting
* **Stationary point** and **canonical analysis** to characterize the surface
* **Surface, contour, wireframe, and cube** visualizations
* **Ridge analysis** for stationary points outside the experimental region
* **Uncoded equation** export for engineering use
* **Desirability optimization** for multi-response problems
* **Pareto frontier** and **overlaid contour plots** for trade-off analysis

This vignette follows a realistic chemical-reactor case study through all
phases of RSM.

```{r setup}
library(iQualityR.doe)
library(ggplot2)
set.seed(42)
```

## 2. API Overview

### 2.1 RSM workflow stages

| Stage | Method | Purpose |
|---|---|---|
| Phase 1 | `compute_steepest_ascent()` | Move toward optimum region |
| Phase 2 | `rsm_design(design_type = "ccd")` | Build second-order design |
| Fit | `task$compute()` | Fit full quadratic model |
| Analyze | `compute_stationary_point()` | Locate & characterize optimum |
| Explore | `task$plot(type = "surface")` | Visualize response surface |
| Robust | `ridge_analysis()` | Constrained optimization |
| Export | `get_uncoded_equation()` | Engineering-units equation |
| Optimize | `optimize_desirability()` | Multi-response optimization |

### 2.2 Plot types for RSM

| `type` argument | Description |
|---|---|
| `"surface"` | Filled 3D-style surface (2D tile + contour) |
| `"contour"` | 2D contour plot with iso-lines |
| `"wireframe"` | 3D wireframe projection (geom_path grid) |
| `"cube"` | 3-factor cube corner display |
| `"overlaid_contour"` | Multi-response overlaid contour with constraints |
| `"response_surface"` | Both surface + contour side by side |

---

## 3. Case Study: Chemical Reactor Yield

### Problem

A chemical engineer is optimizing a batch reactor to maximize **Yield (%)**.
Screening identified three vital factors: **Temperature**, **Pressure**, and
**Catalyst concentration**. The current operating point gives ~70% yield.

### True (unknown) process model

For reproducibility we simulate the true process with known second-order
behavior, but in practice the response is measured experimentally.

---

## 4. Phase 1: Path of Steepest Ascent

Before investing in a second-order design, we run a first-order factorial
near the current operating point and follow the steepest ascent path toward
higher yield.

```{r phase1-design}
factors <- list(
  list(name = "Temp",     type = "continuous", levels = c(120, 180)),
  list(name = "Pressure", type = "continuous", levels = c(2, 6)),
  list(name = "Catalyst", type = "continuous", levels = c(0.5, 2.5))
)

plan <- IqrDoePlan$new(
  task_tag      = "reactor_phase1",
  design_type   = "factorial",
  factors       = factors,
  center_points = 4,
  randomize     = FALSE,
  response_vars = "Yield"
)
task <- IqrDoeTask$new(plan = plan, theme = "academic")
task$compute()

design <- task$results$design_info
# True first-order model (plus small curvature in reality):
#   Yield = 70 + 8*Temp_c + 5*Press_c + 3*Cat_c
design$Yield <- 70 + 8 * design$Temp + 5 * design$Pressure +
  3 * design$Catalyst + rnorm(nrow(design), 0, 1.5)

# Re-fit the model with the simulated response data.
task1 <- IqrDoeTask$new(plan = plan, data = design, theme = "academic")
task1$compute()
```

```{r phase1-model}
model1 <- task1$results$model
summary(model1)$coefficients
```

Now compute the steepest ascent path. The `factors` list is passed
directly --- `compute_steepest_ascent()` matches factor names to the
model's main-effect coefficients internally.

```{r steepest}
ascent <- DoeAnalyzer$new()$compute_steepest_ascent(
  model     = model1,
  factors   = factors,
  step_size = 1.0,
  n_steps   = 5,
  maximize  = TRUE
)
ascent
```

### Decision

The steepest ascent path suggests increasing all three factors
(Temperature shows the largest coefficient, so each coded step translates
to the largest physical move). The team runs confirmatory trials along
this path. When yield peaks and begins to decline, the optimum region has
been reached --- this is where Phase 2 (CCD) is centered.

---

## 5. Phase 2: Central Composite Design

After Phase 1 relocates the operating region to ~88% yield, a CCD is
built around the new center to fit a full second-order model.

```{r ccd-design}
# New factors centered on the improved operating region
factors2 <- list(
  list(name = "Temp",     type = "continuous", levels = c(150, 170)),
  list(name = "Pressure", type = "continuous", levels = c(4, 5)),
  list(name = "Catalyst", type = "continuous", levels = c(1.0, 2.0))
)

plan2 <- IqrDoePlan$new(
  task_tag      = "reactor_phase2",
  design_type   = "ccd",
  factors       = factors2,
  center_points = 6,
  alpha         = "rotatable",
  randomize     = FALSE,
  response_vars = "Yield"
)
task2 <- IqrDoeTask$new(plan = plan2, theme = "academic")
task2$compute()

design2 <- task2$results$design_info
# True second-order model around the new center (coded x1,x2,x3 in [-1.4, 1.4])
design2$Yield <- 90 +
  3.0 * design2$Temp + 2.0 * design2$Pressure + 1.5 * design2$Catalyst -
  4.0 * design2$Temp^2 - 3.0 * design2$Pressure^2 - 2.0 * design2$Catalyst^2 +
  1.5 * design2$Temp * design2$Pressure +
  rnorm(nrow(design2), 0, 1.0)

task2 <- IqrDoeTask$new(plan = plan2, data = design2, theme = "academic")
task2$compute()
```

```{r ccd-anova}
print(task2$plot(type = "design"))
```

### Decision

A rotatable CCD with 6 center points provides 20 runs capable of
estimating the full quadratic model (3 linear + 3 quadratic + 3
interactions + intercept). The design is ready for data collection.

---

## 6. Stationary Point & Canonical Analysis

The stationary point tells us where the gradient of the fitted surface is
zero --- a maximum, minimum, or saddle point.

```{r stationary-point}
model2 <- task2$results$model
factor_names <- vapply(factors2, function(f) f$name, character(1))

sp <- DoeAnalyzer$new()$compute_stationary_point(
  model        = model2,
  factor_names = factor_names
)
sp$stationary_point
sp$nature
```

### Decision

The stationary point nature (`sp$nature`) classifies the surface as a
**maximum**, **minimum**, or **saddle point**. If it is a maximum within
the design region, the team can proceed to validation runs at the
predicted optimum.

---

## 7. Surface & Contour Plots

### 7.1 Filled surface plot

```{r surface}
print(task2$plot(type = "surface", response_name = "Yield",
                 x_var = "Temp", y_var = "Pressure"))
```

### 7.2 Contour plot

```{r contour}
print(task2$plot(type = "contour", response_name = "Yield",
                 x_var = "Temp", y_var = "Pressure"))
```

### 7.3 Response surface (both side by side)

```{r response-surface}
print(task2$plot(type = "response_surface", response_name = "Yield",
                 x_var = "Temp", y_var = "Pressure"))
```

### Decision

The surface plot shows a clear ridge of high yield in the
Temperature-Pressure plane. The contour plot's iso-lines reveal the
sensitivity of yield to factor changes --- tighter contours mean the
process is more sensitive to variation in that direction.

---

## 8. Wireframe Plot

A wireframe provides a 3D-style line projection without tile fill, making
the surface shape easier to read at a glance.

```{r wireframe}
print(task2$plot(type = "wireframe", response_name = "Yield",
                 x_var = "Temp", y_var = "Pressure"))
```

---

## 9. Cube Plot

A cube plot displays the predicted (or mean) response at all 8 corner
combinations of a **2-level factorial** design. It is most informative
for Phase-1 screening designs rather than CCD (which uses axial points
beyond +/-1).

```{r cube}
# Cube plots apply to 2-level factorials with coded levels (+/-1).
factors_coded <- lapply(c("Temp", "Pressure", "Catalyst"), function(nm) {
  list(name = nm, type = "continuous", levels = c(-1, 1))
})
plan_cube <- IqrDoePlan$new(
  task_tag = "cube_demo", design_type = "factorial",
  factors = factors_coded, randomize = FALSE,
  response_vars = "Yield"
)
task_cube <- IqrDoeTask$new(plan = plan_cube, theme = "academic")
task_cube$compute()
design_cube <- task_cube$results$design_info
design_cube$Yield <- 70 + 8 * design_cube$Temp + 5 * design_cube$Pressure +
  3 * design_cube$Catalyst + rnorm(nrow(design_cube), 0, 1.5)
task_cube <- IqrDoeTask$new(plan = plan_cube, data = design_cube, theme = "academic")
task_cube$compute()
print(task_cube$plot(type = "cube", response_name = "Yield"))
```

### Decision

The cube plot quickly shows which corner of the Phase-1 design space
yields the highest response, guiding the engineering team toward the
productive region for Phase-2 centering.

---

## 10. Ridge Analysis

When the stationary point falls **outside** the experimental region (or
the surface is a saddle), ridge analysis finds the path of steepest
ascent **within** a radius constraint, producing a series of constrained
optima.

```{r ridge}
ridge <- DoeAnalyzer$new()$ridge_analysis(
  model        = model2,
  factor_names = factor_names,
  maximize     = TRUE,
  max_radius   = sqrt(3),
  n_radii      = 10
)
head(ridge$path, 4)
```

### Decision

The ridge path shows the best achievable yield at each radius from the
design center. When the radius at which yield is maximized lies inside
the feasible operating window, that point becomes the recommended setpoint.

---

## 11. Uncoded Equation

For engineering handoff, the regression equation in **actual engineering
units** (not coded) is essential.

```{r uncoded-equation}
eq <- DoeAnalyzer$new()$get_uncoded_equation(
  model   = model2,
  factors = factors2
)
cat(eq$equation, "\n")
```

### Decision

The uncoded equation lets process engineers predict yield directly from
Temperature, Pressure, and Catalyst values without converting to coded
units. It can be pasted into a spreadsheet or SOP document.

---

## 12. Multi-Response Desirability Optimization

In practice, multiple responses must be balanced. Suppose alongside
**Yield** (maximize) we also track **Purity** (target 98%) and **Waste**
(minimize).

```{r multi-response-data}
design3 <- design2
# Simulate two more responses on the same CCD runs
design3$Purity <- 98 - 0.5 * design3$Temp^2 - 0.3 * design3$Pressure^2 +
  rnorm(nrow(design3), 0, 0.3)
design3$Waste <- 2.0 + 0.8 * design3$Temp + 0.5 * design3$Pressure +
  0.3 * design3$Catalyst + rnorm(nrow(design3), 0, 0.2)

# Fit a separate second-order model for each response (for desirability).
model_yield  <- lm(Yield ~ (Temp + Pressure + Catalyst)^2 +
                    I(Temp^2) + I(Pressure^2) + I(Catalyst^2), data = design3)
model_purity <- lm(Purity ~ (Temp + Pressure + Catalyst)^2 +
                     I(Temp^2) + I(Pressure^2) + I(Catalyst^2), data = design3)
model_waste  <- lm(Waste ~ (Temp + Pressure + Catalyst)^2 +
                     I(Temp^2) + I(Pressure^2) + I(Catalyst^2), data = design3)

# Build a task with multi-response data for overlaid contour plotting.
plan3 <- IqrDoePlan$new(
  task_tag      = "reactor_multi",
  design_type   = "ccd",
  factors       = factors2,
  center_points = 6,
  alpha         = "rotatable",
  randomize     = FALSE,
  response_vars = c("Yield", "Purity", "Waste")
)
task3 <- IqrDoeTask$new(plan = plan3, data = design3, theme = "academic")
task3$compute()
```

```{r desirability}
opt <- MultiResponseOptimizer$new()

specs <- list(
  list(name = "Yield",  type = "max",    lower = 80, upper = 95),
  list(name = "Purity", type = "target", lower = 95, upper = 99, target = 98),
  list(name = "Waste",  type = "min",    lower = 1,  upper = 5)
)

# Named-list bounds (vignette style): each entry is c(min, max).
bounds <- list(
  Temp     = c(-1.0, 1.0),
  Pressure = c(-1.0, 1.0),
  Catalyst = c(-1.0, 1.0)
)

result <- opt$optimize_desirability(
  models = list(model_yield, model_purity, model_waste),
  specs  = specs,
  bounds = bounds,
  method = "grid",
  grid_resolution = 15
)
result$best_point
result$best_desirability
```

### Decision

The desirability optimizer finds the factor settings that best balance
all three responses according to their individual desirability functions
(maximize yield, hit purity target, minimize waste).

---

## 13. Pareto Frontier & TOPSIS Ranking

When a single desirability aggregate is too restrictive, the **Pareto
frontier** identifies all non-dominated solutions, and **TOPSIS** ranks
them by closeness to the ideal.

```{r pareto}
# Build a grid of candidate solutions and predict responses
grid <- expand.grid(
  Temp = seq(-1, 1, 0.5),
  Pressure = seq(-1, 1, 0.5),
  Catalyst = seq(-1, 1, 0.5)
)
pred_yield <- predict(model_yield, newdata = grid)

# Simple proxy responses for demonstration
candidates <- data.frame(
  Yield  = pred_yield,
  Purity = 98 - 0.5 * grid$Temp^2,
  Waste  = 2.0 + 0.8 * grid$Temp
)

pareto <- opt$find_pareto_frontier(
  responses  = candidates,
  directions = c("max", "max", "min")
)
nrow(pareto$pareto_set)
head(pareto$pareto_set, 3)
```

```{r topsis}
ranked <- opt$rank_with_topsis(
  candidates = candidates,
  weights    = c(0.5, 0.3, 0.2),
  directions = c("max", "max", "min")
)
head(ranked$ranked, 3)
```

### Decision

The Pareto frontier shows the trade-off space: solutions where improving
one response requires sacrificing another. TOPSIS then selects the
solution closest to the ideal point, weighted by engineering priorities.

---

## 14. Overlaid Contour Plot

For two-factor slices of the multi-response problem, an overlaid contour
plot visualizes the **feasible region** where all responses meet their
constraints simultaneously.

```{r overlaid-contour}
response_specs <- list(
  Yield  = list(model = model_yield,  lower = 85,   upper = Inf),
  Purity = list(model = model_purity, lower = 97,   upper = Inf),
  Waste  = list(model = model_waste,  lower = -Inf, upper = 4)
)

print(task3$plot(
  type           = "overlaid_contour",
  response_specs = response_specs,
  x_var          = "Temp",
  y_var          = "Pressure"
))
```

### Decision

The white (unshaded) region in the overlaid contour plot marks the
operating window where Yield >= 85%, Purity >= 97%, and Waste <= 4%.
Any point inside this region is a feasible setpoint satisfying all
three quality requirements.

---

## 15. Conclusion

This vignette demonstrated the complete RSM workflow in `iQualityR.doe`:

* **Phase 1** — `compute_steepest_ascent()` moves the operating region
  toward higher yield before committing to a second-order design.
* **Phase 2** — `rsm_design(design_type = "ccd")` builds a rotatable CCD
  capable of fitting the full quadratic model.
* **Analysis** — `compute_stationary_point()` classifies the surface
  (maximum, minimum, or saddle).
* **Visualization** — `surface`, `contour`, `wireframe`, and `cube` plots
  reveal the shape and sensitivity of the response surface.
* **Robustness** — `ridge_analysis()` handles stationary points outside
  the design region via constrained optimization.
* **Engineering export** — `get_uncoded_equation()` provides the
  regression equation in actual engineering units.
* **Multi-response** — `optimize_desirability()`, `find_pareto_frontier()`,
  `rank_with_topsis()`, and `overlaid_contour` plots support trade-off
  analysis across competing quality requirements.

Together, these tools bring `iQualityR.doe` to feature parity with
Minitab's RSM capabilities while adding modern multi-criteria
optimization (Pareto + TOPSIS) that Minitab lacks natively.
