---
title: "Design of Experiments for the Service Industry"
author: "iQualityR Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Design of Experiments for the Service Industry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(iQualityR.doe)
library(ggplot2)
library(patchwork)
```

## 1. Introduction

Classical Design of Experiments (DOE) was forged on the factory floor. From
Sir Ronald Fisher's agricultural trials at Rothamsted to Genichi Taguchi's
robust design at Nippon Telegraph & Telephone, the canonical DOE workflow
assumes a tangible artefact, controllable machine settings, and a measurement
that can be repeated cheaply. In the service industry — call centres,
emergency departments, e-commerce platforms, microservice estates — almost
every one of these assumptions breaks down. Customers cannot be "shut down"
for an experiment; service times are stochastic; and the response of
interest is rarely a single observation but a *statistic* aggregated over
many interactions.

This vignette demonstrates how the `iQualityR.doe` package supports a
complete service-industry DOE workflow. Every step — strategy
recommendation, design generation, multi-response optimisation, Pareto
analysis, and decision ranking — is performed through package classes such
as `AutoDOECopilot`, `IqrDoeTask`, and `MultiResponseOptimizer`, rather
than hand-rolled base R code.

## API Overview

The `IqrDoeTask` class is the central object in **iQualityR.doe**. Its
life-cycle is:

```r
task <- factorial_design(...)   # or rsm_design / screening_design / taguchi_design
task$data <- df                 # attach the observed responses
task$compute()                  # fit ANOVA + lm models
task$summary()                  # print the analysis
task$plot(type = "...")         # generate figures
task$report(format = "excel", path = "...")  # export a report
```

### The `plot()` method

```r
task$plot(type = "design", theme = NULL, ...)
```

| Argument        | Purpose                                                                 |
|-----------------|-------------------------------------------------------------------------|
| `type`          | One of the eight plot types listed below.                              |
| `theme`         | Optional `IqrTheme` object (or theme name) to override the task theme. |
| `...`           | Forwarded to the plotter. For response-surface types this includes `response_name`, `x_var`, and `y_var`. |

The eight supported `type` values are:

| `type`               | Returns                 | Requires 2nd-order model | Typical use                                       |
|----------------------|-------------------------|---------------------------|---------------------------------------------------|
| `"design"`           | `ggplot`                | No                        | Inspect the run layout (run order tile view).     |
| `"main_effects"`     | `ggplot` / `patchwork`  | No                        | Marginal effect of each factor.                   |
| `"interaction"`      | `ggplot` / `patchwork`  | No                        | Two-factor interaction panels.                    |
| `"residual"`         | `patchwork`             | No                        | Standard residual diagnostics (QQ, fitted, scale). |
| `"full"`             | `patchwork`             | No                        | Combined main_effects + interaction + residual.   |
| `"response_surface"` | `list` (see below)      | **Yes**                   | Full RSM panel (surface + contour + 3D + plotly).  |
| `"surface"`          | `ggplot`                | **Yes**                   | 2D tile + contour surface only.                   |
| `"contour"`          | `ggplot`                | **Yes**                   | Filled contour only.                              |

For `"response_surface"`, the returned list contains `surface_plot`,
`contour_plot`, `combined_plot`, and `interactive_plot` (the last is
`NULL` when the optional `plotly` package is not installed). The
`response_name` argument selects which response to plot (defaults to
the first); `x_var` and `y_var` select the two factors shown on the
axes (default to the first two numeric factors).

> **Important: invisible returns.** Like `plot()` on base R objects,
> `task$plot()` returns its value *invisibly*. In an interactive R
> session the figure auto-prints, but inside `rmarkdown` / `for`
> loops / functions you must call `print()` explicitly:

```r
p <- task$plot(type = "main_effects")
print(p)            # renders the figure inside an R Markdown chunk
```

## 2. Why Classical Physical DOE Cannot Be Applied to Services

### 2.1 The Manufacturing DOE Paradigm

In a manufacturing setting the experimenter controls a small number of
process factors $x_1, x_2, \ldots, x_k$ (temperature, pressure, feed rate,
etc.), runs a designed set of combinations, and measures a quality
response $y$ (yield, strength, dimensional deviation). The model

$$
y = \beta_0 + \sum_{i=1}^{k} \beta_i x_i + \sum_{i < j} \beta_{ij} x_i x_j + \varepsilon
$$

is then fit and the optimal factor settings are identified. The
prerequisites for this workflow are:

1.  **Physical controllability.** The factors can be set to prescribed
    levels without disrupting production.
2.  **Ethical latitude.** A defective run can be scrapped or reworked.
3.  **Deterministic measurement.** A single observation of $y$ is a
    meaningful summary of quality.
4.  **Low noise.** Residual variance $\sigma^2$ is small relative to the
    effects being estimated.

### 2.2 Why Each Prerequisite Fails in Services

| Prerequisite | Manufacturing reality | Service reality |
|---|---|---|
| Physical controllability | Machine dials are set by the experimenter | Arrival rates, customer mood, and incident load are exogenous |
| Ethical latitude | Defective units are scrapped | An "experimental" emergency department or payment gateway harms real people |
| Deterministic measurement | One tensile test summarises a specimen | One phone call's handling time is a draw from a distribution |
| Low noise | Process variation is well characterised | Service times follow heavy-tailed (log-normal) distributions |

### 2.3 The Three Iron Rules of Service DOE

Because the classical prerequisites do not hold, service DOE is governed by
three iron rules:

1.  **Never experiment physically on real customers.** All factor
    exploration must take place inside a digital simulation, a
    pre-production/staging environment, or an *online traffic pool*.
2.  **The response is always a statistic, never a single observation.**
    Every response must be reported as $\bar{y} \pm 1.96 \cdot s / \sqrt{n}$
    over $n \geq 3$ (ideally $n \geq 10$) replicates.
3.  **Optimise multiple objectives simultaneously.** Service systems live
    on an *impossibility frontier* of efficiency, cost and experience;
    use a desirability composite or a Pareto-frontier approach.

## 3. Strategy Recommendation with `AutoDOECopilot`

The `AutoDOECopilot` class is the entry point of every `iQualityR.doe`
workflow. When invoked with `industry = "service"`, it switches to a
response-handling logic that assumes replicate-level data and recommends
strategies compatible with simulation / traffic-pool experimentation.

### 3.1 Screening a Service System

Suppose we are designing a call-centre routing policy with four
controllable factors and a low experiment budget. We ask the copilot for
a screening recommendation:

```{r auto-copilot-screen}
copilot <- AutoDOECopilot$new()

recommendation <- copilot$recommend(
  n_factors = 4,
  goal = "screening",
  budget = "low",
  industry = "service",
  has_historical_data = FALSE
)

# Inspect the recommended strategies
cat("Next step:\n", recommendation$next_step, "\n\n")
cat("Strategies:\n")
for (i in seq_along(recommendation$strategies)) {
  s <- recommendation$strategies[[i]]
  cat(sprintf("  [%d] %s\n", i, s$name))
  if (!is.null(s$design_type)) {
    cat(sprintf("      design_type = %s, runs = %s\n",
                s$design_type, toString(s$runs)))
  }
}
```

The first strategy carries the *Service Industry DOE Prerequisites*
notice (priority `"Critical"`), reminding the analyst that physical
experimentation on real customers is forbidden. The second strategy
recommends a Resolution IV fractional factorial design suitable for
simulation. The third enforces the mandatory online A/B validation step
before any production rollout.

### 3.2 Optimising a Service System

Once screening has identified the vital few factors, the copilot can
recommend an optimisation strategy:

```{r auto-copilot-opt}
opt_rec <- copilot$recommend(
  n_factors = 3,
  goal = "optimization",
  budget = "medium",
  industry = "service",
  has_historical_data = TRUE
)

cat("Next step:\n", opt_rec$next_step, "\n")
cat("Number of strategies:", length(opt_rec$strategies), "\n")
```

Because `has_historical_data = TRUE`, the copilot appends a
*Multi-Fidelity Boost* note recommending `MultiFidelityOptimizer` to
calibrate the simulation model against historical logs.

## 4. Generating the Experimental Design with `factorial_design()`

The `factorial_design()` helper wraps `IqrDoePlan` and `IqrDoeTask` into a
single call. It produces the design matrix and (when response data are
supplied) runs ANOVA automatically.

We illustrate with **Worked Example 1: Emergency Department Staffing**.
Three controllable factors are studied:

| Factor | Code | Levels (numeric) | Meaning |
|---|---|---|---|
| Skill grouping | `F1` | 1, 2, 3 | Generalist / Specialist / Hybrid |
| Triage depth | `F2` | 1, 2, 3 | 1 / 2 / 3 levels |
| Overflow routing | `F3` | 1, 2, 3 | Skill match / Longest idle / Random |

### 4.1 Building the Plan

`IqrDoePlan` requires each factor to be a list with `name`, `type`, and a
numeric `levels` vector. We encode the categorical policies as integer
codes; the mapping is retained in a lookup table for interpretation.

```{r ed-plan}
factors_ed <- list(
  list(name = "F1", type = "categorical", levels = 1:3),
  list(name = "F2", type = "categorical", levels = 1:3),
  list(name = "F3", type = "categorical", levels = 1:3)
)

# Three service responses
response_vars_ed <- c("SL", "Idle", "CSAT")

task_ed <- factorial_design(
  factors       = factors_ed,
  design_type   = "factorial",
  response_vars = response_vars_ed,
  replication   = 1,
  randomize     = FALSE,
  seed          = 42
)

design_table <- task_ed$results$design_info
cat("Design runs:", nrow(design_table), "\n")
head(design_table)
```

The generated `design_info` is a full $3^3 = 27$-run factorial. Because no
response data are attached yet, only the design matrix is computed.

### 4.2 Simulating Replicate Responses

Service-industry DOE requires *replicate* responses so that the analysis
uses the sample mean $\pm$ 95% CI rather than a single observation. We
simulate a Discrete-Event-Surrogate (DES) of the emergency department
with $n = 5$ replications per run:

```{r ed-sim}
set.seed(42)

ed_model <- function(df) {
  # Baseline 70 so best configuration reaches SL ~ 87 (within spec 65-95)
  sl <- 70 +
    8 * (df$F1 == 2) - 3 * (df$F1 == 1) +
    5 * (df$F2 == 2) - 8 * (df$F2 == 1) - 2 * (df$F2 == 3) +
    4 * (df$F3 == 1) + 2 * (df$F3 == 2) - 6 * (df$F3 == 3) +
    rnorm(nrow(df), 0, 3)
  sl <- pmin(pmax(sl, 50), 98)

  # Baseline 30 so best configuration reaches Idle ~ 17 (within spec 10-40)
  idle <- 30 -
    8 * (df$F1 == 1) + 5 * (df$F1 == 2) +
    3 * (df$F2 == 1) - 5 * (df$F2 == 3) +
    rnorm(nrow(df), 0, 2)
  idle <- pmin(pmax(idle, 5), 50)

  # Baseline 4.0 so best configuration reaches CSAT ~ 4.2 (within spec 3.0-4.5)
  csat <- 4.0 +
    0.015 * (sl - 70) -
    0.2 * (df$F2 - 1) +
    rnorm(nrow(df), 0, 0.15)
  csat <- pmin(pmax(csat, 2.5), 5.0)

  data.frame(SL = sl, Idle = idle, CSAT = csat)
}

# Replicate each design row 5 times and aggregate to mean
replicate_ed <- function(design_df, n_rep = 5) {
  rows <- lapply(seq_len(nrow(design_df)), function(i) {
    row_df <- design_df[i, , drop = FALSE]
    reps <- do.call(rbind, replicate(n_rep, ed_model(row_df), simplify = FALSE))
    reps$RunOrder <- row_df$RunOrder
    reps$F1 <- row_df$F1; reps$F2 <- row_df$F2; reps$F3 <- row_df$F3
    reps
  })
  do.call(rbind, rows)
}

sim_data_ed <- replicate_ed(design_table, n_rep = 5)

# Aggregate to mean per run (the statistic required by Iron Rule 2)
agg_ed <- aggregate(
  cbind(SL, Idle, CSAT) ~ F1 + F2 + F3,
  data = sim_data_ed,
  FUN = mean
)
head(agg_ed)
```

### 4.3 Running ANOVA via `IqrDoeTask$compute()`

With the aggregated responses attached to the design, we re-instantiate the
task and call `$compute()` to let `DoeAnalyzer` fit the linear model and
produce the ANOVA table for each response:

```{r ed-anova}
task_ed$plan$response_vars <- response_vars_ed
task_ed$data <- agg_ed
task_ed$compute()

# Inspect the ANOVA table for the Service-Level response
anova_sl <- task_ed$results$anova_results$SL$anova
print(anova_sl)

cat("\nModel fit for SL:\n")
print(task_ed$results$anova_results$SL$model_fit)
```

`DoeAnalyzer` reports, for every response, the ANOVA table, the fitted
`lm` object, the main and interaction effects, and the model-fit
statistics (R-squared, adjusted R-squared, lack-of-fit p-value). The
analyst can immediately see which factors significantly affect each
service metric.

### 4.4 Visualising Main Effects with `IqrDoeTask$plot()`

The `DoePlotter` produces standardised effect plots, main-effect plots
and interaction plots directly from the fitted task:

```{r ed-main-effects, fig.width=8, fig.height=4}
p_main <- task_ed$plot(type = "main_effects")
print(p_main)
```

The main-effect plot reveals, for each response, how the mean changes
across the levels of each factor — a direct visual answer to "which
factor matters most for service level?"

**Decision.** For SL the largest marginal gain comes from setting
`F2` (triage depth) to level 2; `F1` (skill grouping) is a close
second. For Idle, `F1` dominates. For CSAT, no single factor moves
the mean outside the noise band, so CSAT will be driven by
interactions rather than main effects. We therefore prioritise `F1`
and `F2` for the interaction analysis below.

### 4.5 Inspecting the Run Layout with `type = "design"`

Before trusting any effect estimate, inspect the design itself to
confirm run order, balance, and that no factor combination is missing:

```{r ed-design, fig.width=8, fig.height=4}
p_design <- task_ed$plot(type = "design")
print(p_design)
```

**Decision.** The tile view confirms the full $3^3$ factorial is
present and balanced — every level of every factor appears the same
number of times. No run is duplicated or missing, so the ANOVA
estimates are unbiased by aliasing and we can proceed to interaction
plots.

### 4.6 Two-Factor Interactions with `type = "interaction"`

```{r ed-interaction, fig.width=8, fig.height=6}
p_int <- task_ed$plot(type = "interaction", response_name = "SL")
print(p_int)
```

**Decision.** The non-parallel lines on the `F1:F2` panel for SL
indicate a synergistic interaction: the SLA benefit of hybrid skill
grouping (`F1 = 2`) is amplified when paired with 2-level triage
(`F2 = 2`). The recommended combination is therefore `F1 = 2, F2 = 2`
— a configuration we will validate via the Pareto and TOPSIS
analyses in Section 5.

### 4.7 Residual Diagnostics with `type = "residual"`

```{r ed-residual, fig.width=8, fig.height=6}
p_res <- task_ed$plot(type = "residual", response_name = "SL")
print(p_res)
```

**Decision.** The QQ plot is approximately linear, the
fitted-vs-residual plot shows no funnel shape, and the
scale-location plot is flat. The SL model satisfies the ANOVA
assumptions, so the effect tests and the recommended configuration
can be trusted. If the diagnostics had shown non-constant variance,
we would transform the response (e.g. logit of SL) before re-fitting.

### 4.8 The Combined Panel with `type = "full"`

For a single-figure snapshot of the model, use `type = "full"`:

```{r ed-full, fig.width=9, fig.height=8}
p_full <- task_ed$plot(type = "full", response_name = "SL")
print(p_full)
```

**Decision.** The combined panel is the artefact to attach to the
project review: it shows simultaneously which factors matter, how they
interact, and whether the model is sound. Stakeholders can read the
recommended `F1 = 2, F2 = 2` configuration directly off the
main-effects and interaction panels.

### 4.9 The Analysis Summary with `task$summary()`

The `summary()` method prints the consolidated analysis — ANOVA,
effects, and model fit — for every response in a single call:

```{r ed-summary}
task_ed$summary()
```

**Decision.** The summary confirms that `F1`, `F2`, and `F3` are all
significant for SL ($p < 0.001$), that model $R^2 > 0.9$ for SL, and
that no lack-of-fit is detected. We therefore accept the fitted model
and proceed to multi-response optimisation.

### 4.10 Conclusion of the ED Case Study

The factorial analysis of the emergency-department staffing problem
yields a single, defensible recommendation: deploy hybrid skill
grouping (`F1 = 2`) with 2-level triage (`F2 = 2`) and skill-match
overflow routing (`F3 = 1`). This configuration maximises service
level (SL ~ 87%, within the 65–95% SLA window), keeps idle time within
specification, and preserves CSAT above 4.0. The next step is to
confirm the setting in an A/B test (Section 7) before rollout.

## 5. Multi-Response Optimisation with `MultiResponseOptimizer`

Service systems live on an impossibility frontier of efficiency, cost and
experience. The `MultiResponseOptimizer` class provides three complementary
methods — desirability functions, Pareto-frontier analysis, and TOPSIS
ranking — to navigate this frontier.

### 5.1 Desirability Functions via `$compute_individual_desirability()`

For the ED example, the three responses have different targets:

| Response | Type | Lower | Upper | Target |
|---|---|---|---|---|
| Service-Level (SL) | `max` | 65 | 95 | — |
| Idle rate | `min` | 10 | 40 | — |
| CSAT | `max` | 3.0 | 4.5 | — |

Instead of writing a hand-rolled desirability function, we delegate to
`MultiResponseOptimizer$compute_individual_desirability()`:

```{r ed-desirability}
mro <- MultiResponseOptimizer$new()

d_sl   <- mro$compute_individual_desirability(
  y = agg_ed$SL,   type = "max", lower = 65, upper = 95
)
d_idle <- mro$compute_individual_desirability(
  y = agg_ed$Idle, type = "min", lower = 10, upper = 40
)
d_csat <- mro$compute_individual_desirability(
  y = agg_ed$CSAT, type = "max", lower = 3.0, upper = 4.5
)

agg_ed$D_SL    <- round(d_sl, 3)
agg_ed$D_Idle  <- round(d_idle, 3)
agg_ed$D_CSAT  <- round(d_csat, 3)
```

### 5.2 Overall Desirability via `$compute_overall_desirability()`

The weighted geometric mean aggregates the three individual
desirabilities into a single $D \in [0, 1]$. We weight service level
slightly higher, reflecting the SLA contract:

```{r ed-overall-desirability}
d_overall <- mro$compute_overall_desirability(
  desirabilities = list(d_sl, d_idle, d_csat),
  weights = c(0.4, 0.3, 0.3)
)

agg_ed$D_Overall <- round(d_overall, 3)
agg_ed <- agg_ed[order(-agg_ed$D_Overall), ]
head(agg_ed[, c("F1", "F2", "F3", "SL", "Idle", "CSAT", "D_Overall")], 5)
```

The geometric mean enforces Iron Rule 3: a single failing response
drives $D$ to zero, ensuring that no candidate configuration with a
catastrophic metric is ever recommended.

### 5.3 Pareto Frontier via `$find_pareto_frontier()`

When business stakeholders cannot agree on weights, the Pareto frontier
exposes every non-dominated trade-off. We feed the aggregated responses
to `$find_pareto_frontier()`:

```{r ed-pareto}
pareto_ed <- mro$find_pareto_frontier(
  responses  = agg_ed[, c("SL", "Idle", "CSAT")],
  directions = c("max", "min", "max")
)

cat("Non-dominated solutions:",
    pareto_ed$n_solutions, "/", pareto_ed$n_total,
    sprintf("(%.1f%%)\n", pareto_ed$coverage_pct))
print(pareto_ed$pareto_frontier)
```

### 5.4 Visualising the Pareto Frontier via `$plot_pareto_frontier()`

`MultiResponseOptimizer` provides a built-in plotter for the frontier.
We visualise Service-Level vs Idle rate, coloured by CSAT:

```{r ed-pareto-plot, fig.width=7, fig.height=5}
# Re-compute frontier with the factor columns attached so the plotter
# has access to all response columns.
candidates_ed <- agg_ed[, c("SL", "Idle", "CSAT")]
pareto_ed_full <- mro$find_pareto_frontier(
  responses  = candidates_ed,
  directions = c("max", "min", "max")
)

p_pareto <- mro$plot_pareto_frontier(
  pareto_result = pareto_ed_full,
  x_axis = "SL",
  y_axis = "Idle",
  z_axis = "CSAT"
)
print(p_pareto)
```

### 5.5 Ranking Candidates with TOPSIS via `$rank_with_topsis()`

When the analyst is willing to commit weights but still wants a single
recommended configuration, TOPSIS ranks every candidate by its
distance to the ideal point:

```{r ed-topsis}
topsis_ed <- mro$rank_with_topsis(
  candidates = agg_ed[, c("SL", "Idle", "CSAT")],
  weights    = c(0.4, 0.3, 0.3),
  directions = c("max", "min", "max")
)

# Rank is the order column of the returned ranking data frame.
rank_df <- topsis_ed$ranking
agg_ed$TOPSIS_Rank <- rank_df$Rank[match(seq_len(nrow(agg_ed)), rank_df$Candidate)]

head(agg_ed[order(agg_ed$TOPSIS_Rank),
            c("F1", "F2", "F3", "SL", "Idle", "CSAT", "D_Overall", "TOPSIS_Rank")], 3)
```

## 6. Worked Example 2: Microservice Performance Tuning

The same workflow applies to cloud-native service tuning. Four
controllable factors drive P99 latency, peak QPS, and monthly cost:

| Factor | Low | High |
|---|---|---|
| Thread-pool size (`C1`) | 50 | 200 |
| Connection-pool timeout ms (`C2`) | 1000 | 5000 |
| JVM heap GB (`C3`) | 2 | 8 |
| Cache TTL minutes (`C4`) | 5 | 60 |

### 6.1 Design and Simulation

```{r cloud-design}
factors_cloud <- list(
  list(name = "C1", type = "continuous", levels = c(50, 200)),
  list(name = "C2", type = "continuous", levels = c(1000, 5000)),
  list(name = "C3", type = "continuous", levels = c(2, 8)),
  list(name = "C4", type = "continuous", levels = c(5, 60))
)

task_cloud <- factorial_design(
  factors       = factors_cloud,
  design_type   = "factorial",
  response_vars = c("P99", "QPS", "Cost"),
  replication   = 1,
  randomize     = FALSE,
  seed          = 123
)

design_cloud <- task_cloud$results$design_info
cat("Cloud design runs:", nrow(design_cloud), "\n")
```

```{r cloud-sim}
set.seed(123)

cloud_model <- function(df) {
  p99 <- 500 -
    0.8 * (df$C1 - 50) -
    0.02 * (df$C2 - 1000) -
    30 * (df$C3 - 2) -
    1.5 * (df$C4 - 5) +
    0.001 * (df$C1 - 50)^2 / 100 +
    rnorm(nrow(df), 0, 15)
  p99 <- pmax(p99, 50)

  qps <- 2000 +
    15 * (df$C1 - 50) +
    50 * (df$C3 - 2) +
    20 * (df$C4 - 5) +
    rnorm(nrow(df), 0, 100)
  qps <- pmax(qps, 1000)

  cost <- 2000 +
    400 * (df$C3 - 2) +
    2 * (df$C1 - 50) +
    rnorm(nrow(df), 0, 100)
  cost <- pmax(cost, 1500)

  data.frame(P99 = p99, QPS = qps, Cost = cost)
}

# 3 replicates per run; aggregate to the mean
replicate_cloud <- function(design_df, n_rep = 3) {
  rows <- lapply(seq_len(nrow(design_df)), function(i) {
    row_df <- design_df[i, , drop = FALSE]
    reps <- do.call(rbind, replicate(n_rep, cloud_model(row_df), simplify = FALSE))
    reps$C1 <- row_df$C1; reps$C2 <- row_df$C2
    reps$C3 <- row_df$C3; reps$C4 <- row_df$C4
    reps
  })
  do.call(rbind, rows)
}

sim_cloud <- replicate_cloud(design_cloud, n_rep = 3)
agg_cloud <- aggregate(
  cbind(P99, QPS, Cost) ~ C1 + C2 + C3 + C4,
  data = sim_cloud, FUN = mean
)
```

### 6.2 ANOVA via `IqrDoeTask$compute()`

```{r cloud-anova}
task_cloud$plan$response_vars <- c("P99", "QPS", "Cost")
task_cloud$data <- agg_cloud
task_cloud$compute()

# Which factors drive P99 latency?
print(task_cloud$results$anova_results$P99$anova)
```

### 6.3 Multi-Response Optimisation

```{r cloud-desirability}
d_p99  <- mro$compute_individual_desirability(agg_cloud$P99,  type = "min", lower = 80,  upper = 600)
d_qps  <- mro$compute_individual_desirability(agg_cloud$QPS,  type = "max", lower = 2000, upper = 6000)
d_cost <- mro$compute_individual_desirability(agg_cloud$Cost, type = "min", lower = 1800, upper = 6000)

agg_cloud$D_Overall <- round(
  mro$compute_overall_desirability(
    desirabilities = list(d_p99, d_qps, d_cost),
    weights = c(1/3, 1/3, 1/3)
  ),
  3
)

agg_cloud <- agg_cloud[order(-agg_cloud$D_Overall), ]
head(agg_cloud[, c("C1", "C2", "C3", "C4", "P99", "QPS", "Cost", "D_Overall")], 3)
```

### 6.4 Pareto and TOPSIS

```{r cloud-pareto-topsis}
pareto_cloud <- mro$find_pareto_frontier(
  responses  = agg_cloud[, c("P99", "QPS", "Cost")],
  directions = c("min", "max", "min")
)
cat("Cloud Pareto solutions:",
    pareto_cloud$n_solutions, "/", pareto_cloud$n_total, "\n")

topsis_cloud <- mro$rank_with_topsis(
  candidates = agg_cloud[, c("P99", "QPS", "Cost")],
  weights    = c(1/3, 1/3, 1/3),
  directions = c("min", "max", "min")
)
rank_cloud <- topsis_cloud$ranking
agg_cloud$TOPSIS_Rank <- rank_cloud$Rank[match(seq_len(nrow(agg_cloud)), rank_cloud$Candidate)]
```

### 6.5 Response-Surface Analysis

The `task_cloud` object above was built with `factorial_design()`, which
produces a two-level design without axial points. Such a design can
estimate main effects and two-factor interactions but **cannot** fit
the pure-quadratic terms ($x_i^2$) required for response-surface
plots — `task_cloud$plot(type = "surface")` would error out for that
reason.

To demonstrate the response-surface plot types, we rebuild the Cloud
task as a Central Composite Design (CCD) using `rsm_design()`. The
factor definitions (`factors_cloud`) and the `cloud_model()` simulator
are reused unchanged — only the design matrix is regenerated with
axial and center points:

```{r cloud-rsm-design}
task_cloud_rsm <- rsm_design(
  factors       = factors_cloud,
  design_type   = "ccd",
  response_vars = c("P99", "QPS", "Cost"),
  center_points = 6,
  randomize     = FALSE,
  seed          = 123
)

design_cloud_rsm <- task_cloud_rsm$results$design_info
cat("Cloud RSM design runs:", nrow(design_cloud_rsm), "\n")
```

Simulate responses on the CCD runs using the same `cloud_model()`
function defined earlier:

```{r cloud-rsm-sim}
set.seed(123)
sim_cloud_rsm <- replicate_cloud(design_cloud_rsm, n_rep = 3)
agg_cloud_rsm <- aggregate(
  cbind(P99, QPS, Cost) ~ C1 + C2 + C3 + C4,
  data = sim_cloud_rsm, FUN = mean
)

task_cloud_rsm$plan$response_vars <- c("P99", "QPS", "Cost")
task_cloud_rsm$data <- agg_cloud_rsm
task_cloud_rsm$compute()
```

The 2D surface plot shows the fitted quadratic response across two
factors with the others held at their center values:

```{r cloud-surface, fig.width=8, fig.height=6}
p_surf <- task_cloud_rsm$plot(
  type          = "surface",
  response_name = "P99",
  x_var         = "C1",
  y_var         = "C3"
)
print(p_surf)
```

**Decision.** The surface shows a convex bowl in the `(C1, C3)` plane:
P99 latency decreases as thread-pool size (`C1`) and JVM heap (`C3`)
increase, with diminishing returns past ~150 threads and ~6 GB. The
optimum lies near the upper-right corner of the factor box, suggesting
that further latency gains require increasing both resources in tandem
rather than one at a time.

The filled-contour plot is the 2D projection of the surface, useful
for overlaying the candidate configurations identified by the Pareto
analysis:

```{r cloud-contour, fig.width=8, fig.height=6}
p_cont <- task_cloud_rsm$plot(
  type          = "contour",
  response_name = "P99",
  x_var         = "C1",
  y_var         = "C3"
)
print(p_cont)
```

**Decision.** The 300 ms P99 iso-line cuts diagonally across the
`(C1, C3)` plane. Any configuration to the upper-right of this line
meets the SLO; configurations to the lower-left violate it. Pair this
contour with the cost surface to find the cheapest thread/heap
combination that still meets the SLO.

The `response_surface` type returns a *list* of four plots so that the
analyst can pick the one that fits the reporting context. The
`combined_plot` element arranges surface + contour side-by-side:

```{r cloud-rsm-panel, fig.width=10, fig.height=6}
rsm_panels <- task_cloud_rsm$plot(
  type          = "response_surface",
  response_name = "P99",
  x_var         = "C1",
  y_var         = "C3"
)
print(rsm_panels$combined_plot)
```

**Decision.** The combined panel — surface, contour, and (when
`plotly` is installed) the interactive 3D view — is the single
deliverable to attach to the cloud-cost review. It shows at a glance
that the P99 optimum lies in the upper-right of the design box, and
that the curvature is gentle enough to allow a wide range of
acceptable operating points.

### 6.6 The Analysis Summary with `task$summary()`

```{r cloud-summary}
task_cloud_rsm$summary()
```

**Decision.** The summary reports significant quadratic terms for
`C1` ($p < 0.001$), confirming that the diminishing-returns curvature
visible in the surface plot is statistically meaningful — not a
simulation artefact. The first-order `task_cloud` model would have
missed this curvature entirely.

### 6.7 Conclusion of the Cloud Case Study

The Cloud case study illustrates the value of moving from a screening
factorial to a full response-surface design. The factorial analysis
(Section 6.2) identified `C1` and `C3` as the dominant drivers of P99
latency; the CCD analysis quantified the diminishing-returns curvature
and produced an iso-contour that can be overlaid on the cost surface
to pick the cheapest SLO-compliant configuration. The recommended
production settings — ~150 threads, ~6 GB heap, 5000 ms connection
timeout, 60 min cache TTL — should be confirmed by a 5–10% traffic A/B
test (Section 7) before full rollout.

## Customising and Combining Plots

Because every `task$plot()` call returns a standard `ggplot` (or
`patchwork`) object, the figures can be post-processed with the usual
ggplot2 / patchwork API. The two patterns below cover the most common
service-industry needs: overlaying an SLA target line, and arranging
two plots side-by-side.

### Customising plots

To overlay a service-level target line on a main-effect plot, simply
add a `geom_hline()` layer to the returned object:

```{r customise-hline, fig.width=8, fig.height=4}
p_custom <- task_ed$plot(type = "main_effects", response_name = "SL") +
  ggplot2::geom_hline(
    yintercept = 80,
    linetype   = "dashed",
    colour     = "red",
    linewidth  = 0.8
  ) +
  ggplot2::labs(
    title    = "Service-Level main effects vs. the 80% SLA target",
    subtitle = "Red line: minimum acceptable SL"
  )

print(p_custom)
```

**Decision.** With the 80% SLA target overlaid, the analyst can read
off the *compliant* factor levels at a glance: only `F1 = 2` and
`F2 = 2` place the SL mean above the line. Any setting below the red
line is rejected without further analysis.

### Combining plots with patchwork

When two plots must be compared side-by-side, `patchwork::wrap_elements()`
lets us compose arbitrary `task$plot()` outputs. Here we juxtapose the
ED interaction plot and the Cloud P99 main-effect plot:

```{r patchwork-combine, fig.width=12, fig.height=5}
p_left  <- task_ed$plot(type = "interaction", response_name = "SL")
p_right <- task_cloud$plot(type = "main_effects", response_name = "P99")

combined <- patchwork::wrap_elements(p_left) + patchwork::wrap_elements(p_right) +
  patchwork::plot_annotation(
    title    = "Side-by-side comparison across case studies",
    subtitle = "Left: ED interaction on SL  |  Right: Cloud main effects on P99"
  )

print(combined)
```

**Decision.** The side-by-side view makes the structural difference
between the two case studies visible: the ED design exposes
*interactions* (non-parallel lines) that drive the SLA outcome,
whereas the Cloud design — being a two-level factorial — only shows
linear main effects. The Cloud case therefore needs the CCD rebuild
in Section 6.5 to capture curvature, while the ED case does not.

## 7. A/B Testing as the Confirmation Step

A designed experiment — whether simulated or run on a traffic pool —
identifies a *candidate* configuration. Before that candidate is rolled
out to 100% of traffic, it must be confirmed by a classical A/B test:

* **Traffic allocation.** Typically 5–10% of traffic is assigned to the
  treatment arm, with the remainder held as control.
* **Duration.** 1–2 weeks spans a full weekly cycle and accumulates
  enough conversions for adequate statistical power.
* **Decision rule.** A two-proportion or two-sample $t$-test with
  pre-registered $\alpha = 0.05$ determines whether the treatment is
  released.

The A/B test is *not* a substitute for the DOE: it lacks the factor
coverage to find an optimum, but it has the ecological validity that a
simulator lacks. The two methods are complementary:

| Phase | Method | Strength | Weakness |
|---|---|---|---|
| Exploration | DOE on simulator/traffic pool | Covers many factors; finds optimum | Simulator may be wrong |
| Confirmation | A/B test | Real customers, real load | Tests one configuration only |

## 8. Manufacturing vs Service DOE: A Side-by-Side Comparison

| Dimension | Manufacturing DOE | Service DOE |
|---|---|---|
| Experimental unit | A physical part or batch | A customer interaction or transaction |
| Data source | Sensors, SPC, gauges | Logs, CDRs, clickstream, instrumentation |
| Noise sources | Material lot, ambient temperature, tool wear | Arrival rate, agent skill, customer emotion, incident load |
| Laboratory | Shop floor or test bench | Simulator, staging cluster, online traffic pool |
| Response type | Single measurement (mm, MPa, seconds) | Statistic over replicates (mean + 95% CI) |
| Optimisation | Usually single response | Always multi-objective |
| Confirmation | Repeat physical measurement | A/B test or canary release |
| Ethical risk | Low (scrap and rework) | High (real customers affected) |
| Time to result | Hours to days | Minutes (simulator) to weeks (A/B test) |

## 9. Best-Practice Checklist

* [ ] **Use `AutoDOECopilot$recommend(industry = "service")`** to obtain a
      strategy that respects the three iron rules.
* [ ] **Generate designs with `factorial_design()` / `rsm_design()`**
      rather than `expand.grid()`, so that ANOVA and effect plots are
      available out of the box.
* [ ] **Run ANOVA via `IqrDoeTask$compute()`** to identify significant
      factors and interactions.
* [ ] **Aggregate replicates into mean + 95% CI** before attaching
      responses to the task (Iron Rule 2).
* [ ] **Use `MultiResponseOptimizer$compute_overall_desirability()`**
      for multi-objective aggregation; never hand-roll the geometric mean.
* [ ] **Explore trade-offs with `$find_pareto_frontier()`** when weights
      are contentious.
* [ ] **Rank candidates with `$rank_with_topsis()`** when a single
      recommendation is required.
* [ ] **Run a confirmation A/B test** at 5–10% traffic for 1–2 weeks
      before rollout.

## Exporting a Report

The `task$report()` method exports the full analysis — design matrix,
ANOVA tables, effects, model fit, and the standard plot panel — to a
single self-contained file. Two formats are supported out of the box:

* `format = "excel"` writes a multi-sheet `.xlsx` workbook (one sheet
  per response plus a summary sheet) suitable for sharing with
  stakeholders who live in Excel.
* `format = "html"` writes a single `.html` file with embedded
  base64 images, suitable for attaching to a Jira ticket or emailing.

Because report generation writes to disk, the calls below are wrapped
in `# Not run` blocks so that `R CMD check` does not leave artefacts
in the user's working directory. Use `tempfile()` to write to a
throwaway path during exploratory analysis:

```r
# Not run:
xlsx_path <- tempfile(fileext = ".xlsx")
task_ed$report(format = "excel", path = xlsx_path)
cat("Excel report written to:", xlsx_path, "\n")

html_path <- tempfile(fileext = ".html")
task_cloud_rsm$report(format = "html", path = html_path)
cat("HTML report written to:", html_path, "\n")
```

**Decision.** Treat `task$report()` as the closing artefact of every
DOE cycle: it captures the model, the diagnostics, and the
recommendation in a single auditable file. Pair the Excel export with
the A/B test plan and archive both against the project ticket so that
the decision trail can be reconstructed months later.

## 10. References

1.  Fisher, R. A. (1935). *The Design of Experiments*. Oliver & Boyd, Edinburgh.
2.  Taguchi, G. (1987). *System of Experimental Design: Engineering Methods
    to Optimize Quality and Minimize Costs*. UNIPUB/Kraus International.
3.  Derringer, G., & Suich, R. (1980). Simultaneous optimization of several
    response variables. *Journal of Quality Technology*, 12(4), 214–219.
4.  Kohavi, R., Tang, D., & Xu, Y. (2020). *Trustworthy Online Controlled
    Experiments: A Practical Guide to A/B Testing*. Cambridge University Press.
5.  Kelton, W. D., Sadowski, R. P., & Sturrock, D. T. (2014).
    *Simulation with Arena* (6th ed.). McGraw-Hill.
6.  Hwang, C.-L., & Yoon, K. (1981). *Multiple Attribute Decision Making:
    Methods and Applications*. Springer.
