---
title: "Taguchi Robust Design"
author: "iQualityR Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Taguchi Robust Design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  fig.align = "center"
)
```

## 1. Introduction

Taguchi's robust design philosophy seeks to identify **control factor
settings that make the product/process insensitive to noise factors**.
Rather than eliminating noise (which is often impossible or expensive),
the approach exploits interactions between control and noise factors to
minimize response variability.

`iQualityR.doe` implements the full Taguchi workflow:

* Signal-to-Noise (SN) ratio computation (larger / nominal / smaller)
* Robustness analysis with control x noise inner/outer arrays
* Contribution (pooling) analysis
* Optimal-level prediction with confidence intervals
* Dynamic SN ratios for signal-response systems

This vignette follows a **machining surface-finish case study** and a
**sensor calibration dynamic-response example**.

```{r setup}
library(iQualityR.doe)
set.seed(42)
```

## 2. Case Study: Surface Finish Robustness

### Problem

A CNC machining team wants to minimize surface roughness ($R_a$, nm) of
a milled aluminum part. The team identifies 3 control factors and 2
noise factors:

| Type | Factor | Level 1 | Level 2 | Unit |
|---|---|---|---|---|
| Control | Spindle Speed | 8000 | 12000 | rpm |
| Control | Feed Rate | 0.05 | 0.15 | mm/tooth |
| Control | Tool Coating | Uncoated | TiAlN | --- |
| Noise | Material Hardness | Low | High | HRC |
| Noise | Coolant Flow | Low | High | L/min |

**Goal**: smaller-the-better surface roughness.

### 2.1 Inner / Outer Array Design

An $L_8$ inner array (3 control factors at 2 levels) crossed with an
$L_4$ outer array (2 noise factors at 2 levels) provides the robust
design layout.

```{r robust-design}
control_factors <- list(
  list(name = "Speed",   type = "categorical", levels = c(8000, 12000)),
  list(name = "Feed",    type = "categorical", levels = c(0.05, 0.15)),
  list(name = "Coating", type = "categorical", levels = c("Uncoated", "TiAlN"))
)
noise_factors <- list(
  list(name = "Hardness", type = "categorical", levels = c("Low", "High")),
  list(name = "Coolant",  type = "categorical", levels = c("Low", "High"))
)

# Generate the crossed design manually for the example.
# 8 control x 4 noise = 32 rows.
control_runs <- expand.grid(
  Speed = c(-1, 1),
  Feed  = c(-1, 1),
  Coating = c(-1, 1)
)
# Use a subset matching an L8 structure (8 rows).
control_runs <- control_runs[c(1, 2, 3, 4, 5, 6, 7, 8), ]

noise_runs <- expand.grid(
  Hardness = c(-1, 1),
  Coolant  = c(-1, 1)
)

experiment <- merge(control_runs, noise_runs, by = NULL)
# Simulate response: smaller roughness at high speed, low feed, TiAlN coating.
# Noise: high hardness increases roughness, high coolant reduces it.
experiment$Roughness <- 120 -
  15 * (experiment$Speed == 1) +          # high speed -> lower Ra
  20 * (experiment$Feed == 1) +            # high feed -> higher Ra
  -10 * (experiment$Coating == 1) +        # TiAlN -> lower Ra
  12 * (experiment$Hardness == 1) +        # high hardness -> higher Ra
  -8 * (experiment$Coolant == 1) +         # high coolant -> lower Ra
  rnorm(nrow(experiment), 0, 2)

head(experiment, 4)
```

### 2.2 Robustness Analysis

```{r robust-analysis}
analyzer <- TaguchiAnalyzer$new()
robust_result <- analyzer$analyze_robustness(
  data            = experiment,
  control_factors = c("Speed", "Feed", "Coating"),
  noise_factors   = c("Hardness", "Coolant"),
  response        = "Roughness",
  sn_type         = "smaller"   # smaller-the-better
)
robust_result
```

### Decision

The SN ratios (in dB, smaller-the-better convention where **higher SN =
more robust**) show which control-factor combination minimizes variability
while keeping the mean low. The combination with the highest SN is the
most robust.

## 3. SN Ratio Computation

### 3.1 Three SN types

```{r sn-types}
analyzer <- TaguchiAnalyzer$new()

# Larger-the-better (e.g. yield, strength)
sn_larger <- analyzer$compute_sn_ratio(c(70, 75, 80, 85), type = "larger")

# Smaller-the-better (e.g. defect count, roughness)
sn_smaller <- analyzer$compute_sn_ratio(c(5, 8, 3, 6), type = "smaller")

# Nominal-the-best (e.g. dimension, pH)
sn_nominal <- analyzer$compute_sn_ratio(c(10.1, 9.9, 10.0, 10.2),
                                         type = "nominal")
c(larger = sn_larger, smaller = sn_smaller, nominal = sn_nominal)
```

### Interpretation

| SN Type | Goal | Formula |
|---|---|---|
| Larger | Maximize mean | $SN = -10\log_{10}(\frac{1}{n}\sum 1/y^2)$ |
| Smaller | Minimize mean | $SN = -10\log_{10}(\frac{1}{n}\sum y^2)$ |
| Nominal | Minimize variance around target | $SN = 10\log_{10}(\bar{y}^2 / s^2)$ |

For all types, **higher SN indicates a more robust setting**.

## 4. Contribution Analysis

### 4.1 ANOVA-based contribution

```{r contribution}
# Example ANOVA table (from a standard R anova() call or custom).
anova_df <- data.frame(
  Term    = c("Speed", "Feed", "Coating", "Speed:Feed", "Error"),
  Df      = c(1, 1, 1, 1, 3),
  Sum_Sq  = c(400, 900, 250, 50, 30),
  Mean_Sq = c(400, 900, 250, 50, 10),
  stringsAsFactors = FALSE
)

contrib <- analyzer$compute_contribution(anova_df)
contrib
```

### Decision

The contribution percentages show that **Feed Rate** (largest
contribution) is the dominant factor, followed by **Spindle Speed**.
The Speed:Feed interaction is negligible and can be pooled into error
for a more precise error estimate.

## 5. Optimal-Level Prediction

### 5.1 Predicting the optimal SN ratio

Given the robustness results, `predict_optimal()` computes the best level
for each control factor and predicts the SN ratio at those settings:

```{r predict-optimal}
pred <- analyzer$predict_optimal(
  robustness_results = robust_result,
  control_factors    = c("Speed", "Feed", "Coating"),
  sn_type            = "smaller",
  conf_level         = 0.95
)
pred$optimal_levels
pred$predicted_sn
pred$ci
```

### Interpretation

The additive model predicts the SN ratio at the optimal combination as:

$$\hat{SN} = \bar{SN}_{overall} + \sum_{j} (\bar{SN}_{j, opt} - \bar{SN}_{overall})$$

The 95% confidence interval provides a range for the predicted SN,
allowing the engineer to assess whether the improvement is statistically
significant.

### Decision

For the smaller-the-better roughness response, the optimal settings are
high spindle speed (12000 rpm), low feed rate (0.05 mm/tooth), and
TiAlN-coated tool --- exactly matching the engineering intuition that
faster cutting with coated tools produces smoother surfaces.

## 6. Case Study: Dynamic Taguchi (Sensor Calibration)

### Problem

A sensor manufacturer calibrates a pressure transducer where the input
**signal factor** M (reference pressure) should produce a linear output
response y (voltage). The team wants the input-output relationship to be
**proportional** (zero-intercept) and insensitive to noise.

Two control factors:

| Factor | Level 1 | Level 2 |
|---|---|---|
| Amplifier Gain | Low | High |
| Filter Cutoff | 100 Hz | 500 Hz |

Signal factor: reference pressure at 5 levels (1, 2, 3, 4, 5 bar).

### 6.1 Dynamic SN Ratio

For a **zero-point proportional** model ($y = \beta M$):

$$\beta = \frac{\sum y_i M_i}{\sum M_i^2}, \quad
V_e = \frac{\sum y_i^2 - \beta^2 \sum M_i^2}{n - 1}, \quad
SN = 10\log_{10}\frac{\beta^2}{V_e}$$

```{r dynamic-sn}
analyzer <- TaguchiAnalyzer$new()

# A single control-factor combination with 5 signal levels.
# True gain beta = 2.0 (volts per bar).
y_vals <- c(2.1, 3.9, 6.1, 8.2, 9.8)
M_vals <- c(1, 2, 3, 4, 5)

sn_dynamic <- analyzer$compute_dynamic_sn_ratio(
  y    = y_vals,
  M    = M_vals,
  type = "zero_point"
)
sn_dynamic
```

### 6.2 Linear (reference-point) dynamic SN

For a **linear** model ($y = \beta_0 + \beta_1 M$):

```{r dynamic-linear}
sn_linear <- analyzer$compute_dynamic_sn_ratio(
  y    = y_vals,
  M    = M_vals,
  type = "linear"
)
sn_linear
```

### Interpretation

The zero-point proportional SN is lower because the data has a small
intercept offset. The linear SN captures the fit quality including the
intercept.

### 6.3 Full Dynamic Robustness Analysis

```{r dynamic-robust}
# 2 control factors x 5 signal levels = 20 rows.
dat <- expand.grid(Gain = c(1, 2), Cutoff = c(1, 2),
                    M = c(1, 2, 3, 4, 5))
# True proportional gain varies by control setting.
dat$Y <- 2.0 * dat$M + (dat$Gain - 1) * 0.3 * dat$M +
  (dat$Cutoff - 1) * (-0.1) * dat$M + rnorm(nrow(dat), 0, 0.05)

res <- analyzer$analyze_dynamic_robustness(
  data            = dat,
  control_factors = c("Gain", "Cutoff"),
  signal_factor   = "M",
  response        = "Y",
  type            = "zero_point"
)
res
```

### Decision

The dynamic robustness analysis shows that **Gain = High** (level 2)
produces the highest SN ratio, meaning the sensor's input-output gain is
most stable and predictable at that setting. The slope values confirm
the true gain of approximately 2.3 V/bar at the optimal setting.

## 7. Summary Table: Taguchi Workflow

| Step | Method | Purpose |
|---|---|---|
| 1 | `taguchi_design()` or manual | Generate inner/outer array |
| 2 | `compute_sn_ratio()` | Compute SN for each run |
| 3 | `analyze_robustness()` | Aggregate SN by control combination |
| 4 | `compute_contribution()` | ANOVA contribution analysis |
| 5 | `predict_optimal()` | Select optimal levels + CI |
| 6 | `compute_dynamic_sn_ratio()` | Dynamic (signal-response) SN |
| 7 | `analyze_dynamic_robustness()` | Dynamic multi-run aggregation |

## 8. Conclusion

This vignette demonstrated the Taguchi robust-design workflow:

* **SN ratios** for larger / nominal / smaller quality characteristics
* **Robustness analysis** with control x noise crossed arrays
* **Contribution analysis** for factor ranking
* **Optimal-level prediction** with confidence intervals
* **Dynamic Taguchi** for signal-response systems (sensor calibration)

Key principle: **robust design is about minimizing variability, not just
the mean**. The SN ratio captures both, and the control-factor settings
that maximize SN are those that make the process least sensitive to noise.

The next vignette, *AI/ML Advanced Features*, covers Bayesian
optimization, surrogate modeling, and predictive DOE.
