---
title: "Time-Effect Modeling for Pharmaceutical Stability: An ICH Q1A/Q1E Case Study"
author: "iQualityR Team"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Time-Effect Modeling for Pharmaceutical Stability: An ICH Q1A/Q1E Case Study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
library(iQualityR.doe)
library(ggplot2)
library(patchwork)
```

# Introduction

A pharmaceutical company is developing a new active pharmaceutical
ingredient (API) and must register a shelf life with the regulatory
authorities. The ICH Q1A(R2) guideline defines the stability study
design, and ICH Q1E defines how the shelf life is to be derived from
the resulting data. The engineering question is concrete:

> *For how many months can the API be stored at 25 °C / 60 % RH while
> remaining within the specification limit for assay?*

Answering this question rigorously requires three steps:

1. **Model the decay.** Describe how the assay declines over time using
   a kinetic law $R(t)$, and select among candidate laws by goodness of
   fit.
2. **Estimate the shelf life.** Find the time at which the lower
   one-sided 95 % confidence bound of the response crosses the
   specification limit, as mandated by ICH Q1E.
3. **Confirm with accelerated data.** Use the Arrhenius equation to
   extrapolate accelerated aging data (40, 50, 60 °C) to the storage
   temperature (25 °C) and verify that the predicted decay rate agrees
   with the long-term observation.

This vignette walks through the entire workflow end-to-end using the
`TimeEffectModeler` class of `iQualityR.doe`. Every figure is generated
by code the reader can re-run; no chunk is suppressed.

# API Overview

The `TimeEffectModeler` class is the central object for time-effect
decay modelling in **iQualityR.doe**. Its life-cycle is:

```r
tem <- TimeEffectModeler$new()
fit <- tem$fit_decay_model(time = df$time, response = df$response, model_type = "weibull")
tem$plot_time_effect(fit, time = df$time, response = df$response,
                     prediction_time = seq(0, 36, length.out = 100))
tem$predict_response(fit, new_time = seq(0, 36, length.out = 100), se_fit = TRUE)
tem$estimate_shelf_life(fit, specification = 90, confidence_level = 0.95)
tem$analyze_multi_batch(data = df, response_col = "response")
tem$fit_arrhenius_model(data = df, response_col = "response", Ea_activation = 100)
tem$predict_at_target_temp(arrhenius_result = arr, target_temp = 25,
                           new_time = seq(0, 36, length.out = 100), R0 = 100)
```

## The `plot_time_effect()` method

```r
tem$plot_time_effect(fit_result, time, response, prediction_time = NULL, theme_obj = NULL)
```

| Argument          | Purpose                                                                  |
|-------------------|--------------------------------------------------------------------------|
| `fit_result`      | The list returned by `$fit_decay_model()` (or compatible).               |
| `time`            | Numeric vector of observed time points (for Weibull: must be > 0).       |
| `response`        | Numeric vector of observed responses.                                    |
| `prediction_time` | Numeric vector of time points for the predicted curve ribbon. `NULL` builds a grid covering 0–120 % of the observed range. |
| `theme_obj`       | Optional `ggplot2` theme object added with `+`. `NULL` keeps the built-in `theme_minimal()` styling. |

Returns a `ggplot` object *invisibly* — wrap in `print()` inside R
Markdown chunks:

```r
p <- tem$plot_time_effect(fit, time, response,
                          prediction_time = seq(0, 36, length.out = 100))
print(p)
```

## Public methods of `TimeEffectModeler`

| Method                                | Purpose                                                              |
|---------------------------------------|----------------------------------------------------------------------|
| `$new()`                              | Constructor (no arguments).                                          |
| `$fit_decay_model(time, response, model_type)` | Fit one of `"zero_order"`, `"first_order"`, `"weibull"`, `"exponential"`. |
| `$predict_response(fit_result, new_time, se_fit)` | Predict response + (optional) standard-error band.            |
| `$estimate_shelf_life(fit_result, specification, confidence_level)` | ICH Q1E shelf-life estimate (one-sided z = qnorm(0.95) = 1.645). |
| `$analyze_multi_batch(data, response_col)` | Pooled-degradation analysis across batches (ANOVA batch:time interaction). |
| `$fit_arrhenius_model(data, response_col, Ea_activation)` | Fit Arrhenius model for accelerated stability data.            |
| `$predict_at_target_temp(arrhenius_result, target_temp, new_time, R0)` | Predict degradation at a use-condition temperature.          |
| `$plot_time_effect(fit_result, time, response, prediction_time, theme_obj)` | Generate the standard time-effect ggplot.                     |

> **Important: invisible returns.** `$plot_time_effect()` returns its
> value *invisibly*. In an interactive R session the figure
> auto-prints, but inside `rmarkdown` / `for` loops / functions you
> must call `print()` explicitly.

> **Interface style.** `$fit_decay_model()` accepts data in
> **vector style only** — `time` and `response` must be numeric
> vectors, typically columns of a data frame (e.g. `df$time` and
> `df$response`). There is no formula- or data-frame-style
> interface; pass the columns explicitly. The multi-batch and
> Arrhenius methods, by contrast, take a `data` argument together
> with a `response_col` string naming the response column.

# Mathematical Foundations

## Decay Kinetics

The `TimeEffectModeler` class supports four kinetic laws. Each is
presented below in the form used by the package.

**Zero-order kinetics.** The rate is constant, independent of the
remaining concentration:

$$
R(t) = R_0 - k \cdot t.
$$

**First-order kinetics.** The rate is proportional to the current
concentration. Separating variables and integrating gives

$$
R(t) = R_0 \cdot \exp(-k \cdot t).
$$

The half-life $t_{1/2} = \ln 2 / k$ is independent of $R_0$ — a
defining feature of first-order decay and the reason it dominates most
drug-degradation pathways.

**Weibull decay.** A generalisation of first-order decay with a
time-varying rate:

$$
R(t) = R_0 \cdot \exp\!\left( -\left( \frac{t}{\lambda} \right)^{\beta} \right),
$$

with scale $\lambda > 0$ and shape $\beta > 0$. When $\beta = 1$ the
Weibull reduces to first-order decay with $k = 1/\lambda$.

**Exponential-to-plateau.** For a response that tends to a non-zero
asymptote $R_\infty$:

$$
R(t) = R_\infty + (R_0 - R_\infty) \cdot \exp(-k \cdot t).
$$

## Arrhenius Accelerated Aging

Real-time stability testing of a product with a 30-month shelf life
takes 30 months to complete — too slow for product development. The
Arrhenius equation relates the rate constant $k$ to the absolute
temperature $T$:

$$
\boxed{k = A \cdot \exp\!\left( -\frac{E_a}{R \, T} \right),}
$$

where $A$ is the pre-exponential factor, $E_a$ the activation energy
(J/mol), and $R = 8.314\;\mathrm{J\,mol^{-1}\,K^{-1}}$ the gas constant.
Taking logarithms linearises the relation:

$$
\ln k = \ln A - \frac{E_a}{R} \cdot \frac{1}{T}.
$$

A plot of $\ln k$ versus $1/T$ — the *Arrhenius plot* — is therefore a
straight line of slope $-E_a / R$ and intercept $\ln A$. Measuring $k$
at three or more elevated temperatures and fitting this line lets us
extrapolate $k$ to the storage temperature.

## ICH Q1E Shelf Life

ICH Q1E *Evaluation for Stability Data* specifies that the labelled
shelf life must be based on the **lower one-sided 95 % confidence
bound** of the mean response, not on the point estimate. For a
first-order decay law the response equals the specification limit $L$
at

$$
R_0 \exp(-k \, t_{\text{shelf}}) = L
\;\;\Longrightarrow\;\;
\boxed{t_{\text{shelf}} = \frac{\ln(R_0 / L)}{k}.}
$$

Because $k$ and $R_0$ are estimated from data, $t_{\text{shelf}}$ has a
standard error $\mathrm{SE}(t_{\text{shelf}})$. The conservative,
ICH Q1E-compliant shelf life is the lower bound

$$
\boxed{t_{\text{lower}} = t_{\text{shelf}} - z_{0.95} \cdot \mathrm{SE}(t_{\text{shelf}}),}
$$

where $z_{0.95} = `r round(qnorm(0.95), 3)`$ is the **one-sided** 95 %
quantile of the standard normal distribution (`qnorm(0.95)` in R). Note
that this is *not* the two-sided quantile $z_{0.975} = 1.960$; ICH Q1E
is explicitly one-sided because only the lower tail risks the
specification.

# Case Study: Drug Substance Stability

## Study Design

A new API is placed on long-term stability at 25 °C / 60 % RH and on
accelerated stability at three elevated temperatures (40, 50, 60 °C)
to support Arrhenius extrapolation. Three pilot batches are studied
long-term; a single batch is studied at each accelerated condition.
The assay is measured by HPLC, expressed as % of label claim.

| Condition | Batches | Time points (months) | Purpose |
|:----------|:--------|:---------------------|:--------|
| 25 °C / 60 % RH (long-term) | A, B, C | 0, 3, 6, 9, 12, 18, 24 | Shelf-life estimation |
| 40 °C / 75 % RH (accelerated) | A | 0, 1, 2, 3, 4, 5, 6 | Arrhenius |
| 50 °C (stress) | A | 0, 1, 2, 3, 4, 5, 6 | Arrhenius |
| 60 °C (stress) | A | 0, 1, 2, 3, 4, 5, 6 | Arrhenius |

The specification limit for assay is 90 % of label claim (lower
one-sided).

## Long-Term Stability Data

We begin by simulating the long-term data for the three batches. The
underlying decay is first-order with a batch-specific rate constant
$k \approx 0.0035\;\mathrm{month^{-1}}$, which yields a theoretical
shelf life of $\ln(100/90)/0.0035 \approx 30$ months at the 90 %
specification. Small measurement noise is added.

```{r sim-longterm}
set.seed(2024)

time_long <- c(0, 3, 6, 9, 12, 18, 24)

# Batch-specific first-order rate constants (per month)
k_A <- 0.0036
k_B <- 0.0034
k_C <- 0.0035

long_term_data <- data.frame(
  batch = rep(c("Batch_A", "Batch_B", "Batch_C"), each = 7),
  time  = rep(time_long, times = 3),
  response = round(c(
    100 * exp(-k_A * time_long),
    100 * exp(-k_B * time_long),
    100 * exp(-k_C * time_long)
  ) + rnorm(21, sd = 0.3), 2)
)

knitr::kable(long_term_data, digits = 2,
             caption = "Long-term stability data (assay, % of label claim).")
```

A quick visual inspection confirms that all three batches decay
smoothly from 100 % to roughly 92 % over 24 months, with no visible
batch-to-batch divergence.

```{r plot-longterm}
ggplot(long_term_data, aes(x = time, y = response, colour = batch)) +
  geom_point(size = 2.5) +
  geom_line(aes(group = batch), linewidth = 0.6) +
  labs(title = "Long-Term Stability Data at 25 °C",
       subtitle = "Three batches; assay vs. time",
       x = "Time (months)", y = "Assay (% of label claim)",
       colour = "Batch") +
  theme_minimal(base_size = 12)
```

## Multi-Batch Poolability Test

ICH Q1E requires stability data from at least three batches. Before
pooling them into a single fit we must test whether the batches share
a common degradation rate. The standard test is a two-way ANOVA with a
batch-by-time interaction term. ICH Q1E uses a *liberal* significance
level of $\alpha = 0.25$ (rather than 0.05) for this test, because the
cost of incorrectly pooling non-poolable batches is high:

* $p > 0.25$ — batches are **poolable**; combine into one fit.
* $p \le 0.25$ — batches must be analysed **separately**; the shortest
  shelf life is reported.

```{r multi-batch}
tem <- TimeEffectModeler$new()

batch_result <- tem$analyze_multi_batch(
  data         = long_term_data,
  response_col = "response"
)

cat("ANOVA table:\n")
print(batch_result$anova_table)

cat("\nBatch x time interaction p-value:",
    round(batch_result$batch_time_interaction_p, 3), "\n")
cat("Poolable:", batch_result$poolable, "\n")
cat("Conclusion:", batch_result$interpretation, "\n")
```

The interaction p-value is well above 0.25, so the three batches are
**poolable**. We visualise this conclusion by overlaying the
batch-specific first-order fits: the three curves are visually
indistinguishable.

```{r plot-multibatch}
# Fit each batch separately with first-order kinetics
batch_fits <- lapply(c("Batch_A", "Batch_B", "Batch_C"), function(b) {
  d <- long_term_data[long_term_data$batch == b, ]
  tem$fit_decay_model(d$time, d$response, model_type = "first_order")
})
names(batch_fits) <- c("Batch_A", "Batch_B", "Batch_C")

# Build a prediction data frame for the smooth curves
t_grid <- seq(0, 30, length.out = 100)
pred_df <- do.call(rbind, lapply(names(batch_fits), function(b) {
  data.frame(
    batch    = b,
    time     = t_grid,
    response = tem$predict_response(batch_fits[[b]], t_grid)
  )
}))

ggplot() +
  geom_line(data = pred_df,
            aes(x = time, y = response, colour = batch), linewidth = 0.9) +
  geom_point(data = long_term_data,
             aes(x = time, y = response, colour = batch), size = 2.5) +
  labs(title = "Multi-Batch Consistency: Per-Batch First-Order Fits",
       subtitle = sprintf("Interaction p = %.3f > 0.25  =>  poolable",
                          batch_result$batch_time_interaction_p),
       x = "Time (months)", y = "Assay (% of label claim)",
       colour = "Batch") +
  theme_minimal(base_size = 12)
```

Because the batches are poolable, we combine them into a single
21-point dataset for all subsequent modelling.

```{r pool-data}
pooled_data <- data.frame(
  time     = long_term_data$time,
  response = long_term_data$response
)
cat("Pooled dataset:", nrow(pooled_data), "observations\n")
```

## Fitting the Four Decay Models

We fit all four kinetic laws to the pooled data. The Weibull model
requires strictly positive time points (its linearisation uses
$\ln t$), so the $t = 0$ observations are excluded from the Weibull
fit only; the other three models use the full dataset.

```{r fit-models}
fit_zero  <- tem$fit_decay_model(pooled_data$time, pooled_data$response,
                                 model_type = "zero_order")
fit_first <- tem$fit_decay_model(pooled_data$time, pooled_data$response,
                                 model_type = "first_order")

# Weibull: drop t = 0 (log-linearisation requires t > 0)
idx_pos <- pooled_data$time > 0
fit_weibull <- tem$fit_decay_model(
  time       = pooled_data$time[idx_pos],
  response   = pooled_data$response[idx_pos],
  model_type = "weibull"
)

fit_exp <- tem$fit_decay_model(pooled_data$time, pooled_data$response,
                               model_type = "exponential")

cat("Zero-order  R2 =", round(fit_zero$r_squared,  4), "\n")
cat("First-order R2 =", round(fit_first$r_squared, 4), "\n")
cat("Weibull      R2 =", round(fit_weibull$r_squared, 4), " (transformed scale)\n")
cat("Exp-plateau  R2 =", round(fit_exp$r_squared,  4), "\n")
```

## Model Comparison

The `r_squared` values returned by `fit_decay_model()` are not all on
the same scale: the Weibull $R^2$ is computed on the linearised
response $\ln(-\ln(R/R_0))$, while the other three are on the original
response scale. For a fair comparison we recompute $R^2$ on the
original scale for every model by correlating observed and predicted
responses.

```{r model-comparison}
# Original-scale R2 for fair comparison across all four models
r2_original <- function(fit, time, response) {
  pred <- tem$predict_response(fit, time)
  cor(response, pred)^2
}

comparison_df <- data.frame(
  Model = c("Zero-order", "First-order", "Weibull", "Exp-plateau"),
  R2_original = c(
    r2_original(fit_zero,    pooled_data$time, pooled_data$response),
    r2_original(fit_first,   pooled_data$time, pooled_data$response),
    r2_original(fit_weibull, pooled_data$time, pooled_data$response),
    r2_original(fit_exp,     pooled_data$time, pooled_data$response)
  ),
  k_or_lambda = c(
    round(fit_zero$parameters$k,      5),
    round(fit_first$parameters$k,     5),
    round(fit_weibull$parameters$lambda, 4),
    round(fit_exp$parameters$k,       5)
  ),
  beta = c(NA, NA, round(fit_weibull$parameters$beta, 3), NA),
  R_inf = c(NA, NA, NA, round(fit_exp$parameters$R_inf, 2))
)

knitr::kable(comparison_df, digits = 4,
             caption = "Model comparison on the original response scale.")
```

```{r plot-model-comparison}
comparison_df$Model <- factor(comparison_df$Model, levels = comparison_df$Model)

ggplot(comparison_df, aes(x = Model, y = R2_original, fill = Model)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = sprintf("%.4f", R2_original)), vjust = -0.4, size = 3.5) +
  scale_y_continuous(limits = c(0, 1.02)) +
  labs(title = "Model Comparison: R² on Original Scale",
       subtitle = "First-order and Weibull (beta ~= 1) tie for the best fit",
       x = "Kinetic model", y = "R² (original scale)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

**Reasoning.** The zero-order model is worst because the data are
visibly curved. The exponential-to-plateau model fits the observed
range reasonably but its plateau $R_\infty$ sits at roughly 92 % —
above the 90 % specification — so it would predict an infinite shelf
life, which is physically wrong for a degrading API. The first-order
and Weibull models tie at the top of the ranking, and the Weibull
shape parameter $\beta$ is close to 1, confirming that the process is
indeed first-order. We adopt the **first-order model** on grounds of
parsimony and physical interpretability.

The overlay plot below shows all four fitted curves together with the
observed data. The first-order (blue) and Weibull (green) curves are
almost indistinguishable; the zero-order (red) curve is biased high at
long times; the exponential-to-plateau (purple) curve flattens out
above the specification.

```{r plot-overlay}
t_grid <- seq(0, 36, length.out = 250)
overlay_df <- data.frame(
  time = rep(t_grid, 4),
  response = c(
    tem$predict_response(fit_zero,    t_grid),
    tem$predict_response(fit_first,   t_grid),
    tem$predict_response(fit_weibull, t_grid),
    tem$predict_response(fit_exp,     t_grid)
  ),
  Model = rep(c("Zero-order", "First-order", "Weibull", "Exp-plateau"),
              each = length(t_grid))
)

ggplot() +
  geom_line(data = overlay_df,
            aes(x = time, y = response, colour = Model), linewidth = 0.9) +
  geom_point(data = pooled_data,
             aes(x = time, y = response), size = 1.8, alpha = 0.6) +
  geom_hline(yintercept = 90, linetype = "dashed", colour = "red") +
  labs(title = "Four Kinetic Models Overlaid on Pooled Data",
       subtitle = "Dashed red line = 90 % specification limit",
       x = "Time (months)", y = "Assay (% of label claim)",
       colour = "Model") +
  theme_minimal(base_size = 12)
```

## Best-Model Visualisation

The `$plot_time_effect()` method draws the fitted curve with a
confidence ribbon and the observed points. It returns a `ggplot`
object invisibly, so we wrap every call in `print()` to display it.
We illustrate it on all four fitted models so the reader can see
how the curve shape and the confidence band differ across kinetic
laws.

### First-order fit

```{r plot-best-model}
print(tem$plot_time_effect(
  fit_result      = fit_first,
  time            = pooled_data$time,
  response        = pooled_data$response,
  prediction_time = seq(0, 36, length.out = 150)
))
```

**Decision.** The first-order curve follows the data closely
($R^2 > 0.99$ on the original scale) and the 95 % band stays
narrow across the observed range. The curve crosses the 90 %
specification at roughly 30 months, which is the shelf life we
will register. This is our adopted model.

### Zero-order fit

```{r plot-zero-order}
print(tem$plot_time_effect(
  fit_result      = fit_zero,
  time            = pooled_data$time,
  response        = pooled_data$response,
  prediction_time = seq(0, 36, length.out = 150)
))
```

**Decision.** The zero-order (straight-line) fit is biased high at
short times and low at long times — the curvature of true
first-order decay cannot be captured by a linear law. The band is
visibly wider than the first-order band, signalling a poorer fit.
We reject this model.

### Weibull fit

For the Weibull plot we pass the strictly-positive subset of the
data (the $t = 0$ point was excluded from the fit; see the
`fit-models` chunk).

```{r plot-weibull}
print(tem$plot_time_effect(
  fit_result      = fit_weibull,
  time            = pooled_data$time[idx_pos],
  response        = pooled_data$response[idx_pos],
  prediction_time = seq(0, 36, length.out = 150)
))
```

**Decision.** The Weibull curve overlays the first-order curve
almost exactly, and the fitted shape parameter
$\beta \approx 1$ reduces the Weibull model to the first-order
model. On grounds of parsimony we keep the first-order model.

### Exponential-to-plateau fit

```{r plot-exponential}
print(tem$plot_time_effect(
  fit_result      = fit_exp,
  time            = pooled_data$time,
  response        = pooled_data$response,
  prediction_time = seq(0, 36, length.out = 150)
))
```

**Decision.** The exponential-to-plateau curve flattens out at
$R_\infty \approx 92$ %, above the 90 % specification. This model
would therefore predict an infinite shelf life, which is
physically wrong for a degrading API. We reject it.

## Styling the plot with `theme_obj`

The `theme_obj` argument accepts any `ggplot2` theme object that
can be added with `+`. The default is `theme_minimal()`; passing
`theme_bw()` (or a custom theme) overrides the styling. The plot
below re-uses the Weibull fit with `theme_bw()`.

```{r plot-theme-obj}
p_themed <- tem$plot_time_effect(
  fit_result      = fit_weibull,
  time            = pooled_data$time[idx_pos],
  response        = pooled_data$response[idx_pos],
  prediction_time = seq(0, 36, length.out = 150),
  theme_obj       = theme_bw(base_size = 12)
)
print(p_themed)
```

**Decision.** Swapping `theme_minimal()` for `theme_bw()` only
changes the cosmetics (panel border, gridlines); the underlying
data, ribbon and curve are identical. Use `theme_obj` to match
your organisation's report style without rewriting the plot.

## Shelf-Life Estimation (ICH Q1E)

We now estimate the shelf life from the first-order fit with the ICH
Q1E one-sided 95 % confidence bound. The specification limit is 90 %
of label claim.

```{r shelf-life}
shelf <- tem$estimate_shelf_life(
  fit_result       = fit_first,
  specification    = 90,
  confidence_level = 0.95
)

cat(shelf$note, "\n\n")
cat("Point estimate  t_shelf  :", round(shelf$shelf_life, 2), "months\n")
cat("SE(t_shelf)              :", round(shelf$se_shelf_life, 2), "months\n")
cat("One-sided z (qnorm(0.95)):", round(qnorm(0.95), 4), "\n")
cat("Lower 95 % bound         :", round(shelf$lower_95, 2), "months\n")
cat("Upper 95 % bound         :", round(shelf$upper_95, 2), "months\n")
```

The point estimate is `r round(shelf$shelf_life, 1)` months and the
ICH Q1E lower one-sided 95 % bound is `r round(shelf$lower_95, 1)`
months. The lower bound is the value that may be registered.

## Shelf-Life Plot with Confidence Bounds

The figure below shows the fitted decay curve, the 95 % confidence
band on the mean response, the 90 % specification limit, and the
shelf-life estimate together with its lower bound. The green dashed
line marks the point estimate $t_{\text{shelf}}$; the green dotted
line marks the conservative ICH Q1E lower bound.

```{r plot-shelf-life}
# Predictions with a confidence band
t_pred <- seq(0, 40, length.out = 250)
pred_se <- tem$predict_response(fit_first, t_pred, se_fit = TRUE)

pred_df <- data.frame(
  time    = t_pred,
  predict = pred_se$predict,
  lower   = pred_se$lower,
  upper   = pred_se$upper
)

ggplot() +
  geom_ribbon(data = pred_df,
              aes(x = time, ymin = lower, ymax = upper),
              fill = "#2C7BB6", alpha = 0.18) +
  geom_line(data = pred_df,
            aes(x = time, y = predict), colour = "#2C7BB6", linewidth = 1) +
  geom_point(data = pooled_data,
             aes(x = time, y = response), size = 2, alpha = 0.7) +
  geom_hline(yintercept = 90, linetype = "dashed", colour = "red", linewidth = 0.7) +
  geom_vline(xintercept = shelf$shelf_life, linetype = "dashed",
             colour = "darkgreen", linewidth = 0.7) +
  geom_vline(xintercept = shelf$lower_95, linetype = "dotted",
             colour = "darkgreen", linewidth = 0.7) +
  annotate("text", x = shelf$shelf_life, y = 91.5,
           label = sprintf("t_shelf = %.1f mo", shelf$shelf_life),
           hjust = -0.05, colour = "darkgreen", size = 3.5) +
  annotate("text", x = shelf$lower_95, y = 89.2,
           label = sprintf("lower 95%% = %.1f mo", shelf$lower_95),
           hjust = -0.05, colour = "darkgreen", size = 3.5) +
  annotate("text", x = 0.5, y = 90.3,
           label = "spec = 90 %", hjust = -0.05, colour = "red", size = 3.5) +
  labs(title = "Shelf-Life Estimation (ICH Q1E, one-sided 95 %)",
       subtitle = sprintf("z = %.3f  (qnorm(0.95), one-sided)", qnorm(0.95)),
       x = "Time (months)", y = "Assay (% of label claim)") +
  coord_cartesian(xlim = c(0, 40), ylim = c(86, 101)) +
  theme_minimal(base_size = 12)
```

## Customising plots with `+ geom_xxx()`

Because `$plot_time_effect()` returns a standard `ggplot` object, it
can be post-processed with any `ggplot2` layer. The chunk below
starts from the first-order plot and adds the 90 % specification
line with `geom_hline()` plus a text annotation of the estimated
shelf life with `annotate("text", ...)`.

```{r plot-customised}
p_custom <- tem$plot_time_effect(
  fit_result      = fit_first,
  time            = pooled_data$time,
  response        = pooled_data$response,
  prediction_time = seq(0, 36, length.out = 150)
) +
  geom_hline(yintercept = 90, linetype = "dashed", colour = "red") +
  annotate("text", x = 0.5, y = 90.4, label = "spec = 90 %",
           hjust = -0.05, colour = "red", size = 3.5) +
  annotate("text", x = shelf$shelf_life, y = 95,
           label = sprintf("t_shelf = %.1f mo", shelf$shelf_life),
           hjust = 1.05, colour = "darkgreen", size = 3.5) +
  labs(title = "First-Order Fit with Specification and Shelf-Life Annotations")

print(p_custom)
```

**Decision.** The annotation makes the registration decision
explicit at a glance: the fitted curve stays above the 90 %
specification through the estimated shelf life, and the lower
95 % bound (computed above) is the value that may be filed. This
is the recommended way to enrich a standard `$plot_time_effect()`
output without re-implementing the ribbon and curve layers.

## Combining plots with patchwork

When several fits must be compared side-by-side, the **patchwork**
package composes independent `ggplot` objects. The chunk below
places the zero-order and Weibull plots next to each other using
`patchwork::wrap_elements()` so that each panel keeps its own
scales and titles, and adds a global title with `plot_annotation()`.

```{r plot-patchwork}
p_zero <- tem$plot_time_effect(
  fit_result      = fit_zero,
  time            = pooled_data$time,
  response        = pooled_data$response,
  prediction_time = seq(0, 36, length.out = 150)
)

p_weib <- tem$plot_time_effect(
  fit_result      = fit_weibull,
  time            = pooled_data$time[idx_pos],
  response        = pooled_data$response[idx_pos],
  prediction_time = seq(0, 36, length.out = 150)
)

combined <- wrap_elements(p_zero) + wrap_elements(p_weib) +
  plot_annotation(title = "Zero-order vs. Weibull fit",
                  tag_levels = "A")

print(combined)
```

**Decision.** Side-by-side comparison confirms visually that the
Weibull fit tracks the curvature of the data while the zero-order
fit is biased. Patchwork makes the contrast immediate and is the
recommended way to present model-selection evidence in a
registration dossier.

## Sensitivity to the Specification Limit

A lower specification limit yields a longer shelf life. Sweeping a
range of specification limits is a useful input to the regulatory
negotiation of the labelled expiry.

```{r shelf-sensitivity}
specs <- c(95, 92, 90, 88, 85, 80)
shelf_lives <- vapply(specs, function(s) {
  tem$estimate_shelf_life(fit_first, specification = s,
                          confidence_level = 0.95)$shelf_life
}, numeric(1))

sens_df <- data.frame(
  specification_pct = specs,
  shelf_life_months = round(shelf_lives, 1)
)

knitr::kable(sens_df,
             caption = "Shelf-life point estimate as a function of the specification limit.")
```

```{r plot-shelf-sensitivity}
ggplot(sens_df, aes(x = specification_pct, y = shelf_life_months)) +
  geom_line(colour = "steelblue", linewidth = 0.9) +
  geom_point(size = 2.5, colour = "steelblue") +
  geom_vline(xintercept = 90, linetype = "dashed", colour = "red") +
  labs(title = "Shelf-Life Sensitivity to Specification Limit",
       subtitle = "Red line = 90 % specification adopted for registration",
       x = "Specification limit (% of label claim)",
       y = "Shelf life (months)") +
  theme_minimal(base_size = 12)
```

## Accelerated Aging: Arrhenius Analysis

We now turn to the accelerated data at 40, 50 and 60 °C. The rate
constant $k$ is estimated at each temperature by fitting a first-order
model, and the three $(1/T, \ln k)$ points are regressed to recover
$E_a$ and $A$.

```{r sim-accelerated}
set.seed(2025)

time_acc <- c(0, 1, 2, 3, 4, 5, 6)

# Rate constants at 40, 50, 60 °C chosen so that the Arrhenius
# extrapolation to 25 °C recovers k ~= 0.0035 /month (consistent with
# the long-term data). Ea = 80,000 J/mol.
k_40 <- 0.0165
k_50 <- 0.0424
k_60 <- 0.1035

acc_data <- data.frame(
  time = rep(time_acc, times = 3),
  temperature = rep(c(40, 50, 60), each = 7),
  response = round(c(
    100 * exp(-k_40 * time_acc),
    100 * exp(-k_50 * time_acc),
    100 * exp(-k_60 * time_acc)
  ) + rnorm(21, sd = 0.5), 2)
)

knitr::kable(acc_data, digits = 2,
             caption = "Accelerated aging data (assay, % of label claim).")
```

```{r arrhenius-fit}
arr_result <- tem$fit_arrhenius_model(
  data         = acc_data,
  response_col = "response"
)

cat("Activation energy Ea     :", round(arr_result$Ea_activation, 0), "J/mol\n")
cat("Pre-exponential factor A :", signif(arr_result$pre_exponential_factor, 4), "\n")
cat("Arrhenius R2             :", round(arr_result$r_squared_arrhenius, 4), "\n\n")

cat("Per-temperature first-order fits:\n")
cat("  k(40 C) =", signif(arr_result$k_values["40"], 5), "/month,  R2 =",
    round(arr_result$r_squared_individual["40"], 4), "\n")
cat("  k(50 C) =", signif(arr_result$k_values["50"], 5), "/month,  R2 =",
    round(arr_result$r_squared_individual["50"], 4), "\n")
cat("  k(60 C) =", signif(arr_result$k_values["60"], 5), "/month,  R2 =",
    round(arr_result$r_squared_individual["60"], 4), "\n")
```

The estimated activation energy is
`r round(arr_result$Ea_activation, 0)` J/mol, comfortably inside the
20–120 kJ/mol range typical of pharmaceutical hydrolysis, and the
Arrhenius line explains
`r round(100 * arr_result$r_squared_arrhenius, 1)` % of the variance in
$\ln k$.

## Arrhenius Plot

The Arrhenius plot — $\ln k$ versus $1/T$ — is the central diagnostic
for accelerated-aging studies. The three measured points should fall
on a straight line; the extrapolation to 25 °C (red triangle) gives
the predicted rate constant at the storage temperature.

```{r plot-arrhenius}
# Measured points
temps_C <- as.numeric(names(arr_result$k_values))
temps_K <- temps_C + 273.15
inv_T   <- 1 / temps_K
ln_k    <- log(arr_result$k_values)

arr_pts <- data.frame(inv_T = inv_T, ln_k = ln_k,
                      label = sprintf("%.0f °C", temps_C))

# Extrapolated point at 25 °C
T_25_K   <- 298.15
k_25_arr <- arr_result$pre_exponential_factor *
  exp(-arr_result$Ea_activation / (8.314 * T_25_K))
arr_25 <- data.frame(inv_T = 1 / T_25_K, ln_k = log(k_25_arr),
                     label = "25 °C (extrapolated)")

# Fitted line
inv_T_line <- seq(min(inv_T) * 0.995, max(inv_T) * 1.005, length.out = 100)
coefs <- coef(arr_result$arrhenius_model_obj)
ln_k_line <- coefs[1] + coefs[2] * inv_T_line
arr_line <- data.frame(inv_T = inv_T_line, ln_k = ln_k_line)

ggplot() +
  geom_line(data = arr_line, aes(x = inv_T, y = ln_k),
            colour = "steelblue", linetype = "dashed") +
  geom_point(data = arr_pts, aes(x = inv_T, y = ln_k),
             size = 3.5, colour = "steelblue") +
  geom_point(data = arr_25, aes(x = inv_T, y = ln_k),
             size = 4.5, colour = "red", shape = 17) +
  geom_text(data = arr_pts, aes(x = inv_T, y = ln_k, label = label),
            hjust = -0.15, vjust = 0.4, size = 3.3) +
  geom_text(data = arr_25, aes(x = inv_T, y = ln_k, label = label),
            hjust = 1.1, vjust = -0.4, size = 3.3, colour = "red") +
  labs(title = "Arrhenius Plot: ln(k) vs 1/T",
       subtitle = sprintf("Ea = %.0f J/mol, R2 = %.4f;  k(25 °C) = %.5f /month",
                          arr_result$Ea_activation,
                          arr_result$r_squared_arrhenius, k_25_arr),
       x = "1/T (1/K)", y = "ln(k)") +
  theme_minimal(base_size = 12)
```

## Extrapolation to 25 °C

Using the Arrhenius parameters we now predict the assay trajectory at
the 25 °C storage temperature over a 36-month horizon.

```{r predict-25c}
t_pred_25 <- seq(0, 36, by = 3)
pred_25 <- tem$predict_at_target_temp(
  arrhenius_result = arr_result,
  target_temp      = 25,
  new_time         = t_pred_25,
  R0               = 100
)

pred_25_df <- data.frame(
  time              = t_pred_25,
  predicted_assay   = round(pred_25, 2)
)

knitr::kable(pred_25_df,
             caption = "Predicted assay at 25 °C from Arrhenius extrapolation.")
```

## Cross-Validation: Arrhenius vs Long-Term Fit

The decisive confirmation step is to compare the **two independent
estimates** of the 25 °C decay curve:

* the **long-term fit** (first-order model fitted directly to the
  25 °C data), and
* the **Arrhenius prediction** (extrapolated from the 40, 50, 60 °C
  accelerated data alone).

If the two curves agree, the Arrhenius model is validated and the
shelf-life claim is supported by two independent lines of evidence.

```{r cross-validation}
# Long-term fit: rate constant and predicted curve
k_longterm <- fit_first$parameters$k
cat("Long-term fit:    k(25 C) =", round(k_longterm, 5), "/month\n")
cat("Arrhenius pred.:   k(25 C) =", round(k_25_arr,  5), "/month\n")
cat("Relative difference      :",
    round(100 * (k_longterm - k_25_arr) / k_longterm, 2), "%\n\n")

# Shelf life implied by the Arrhenius-predicted rate constant
t_shelf_arr <- log(100 / 90) / k_25_arr
cat("Shelf life from long-term fit :", round(shelf$shelf_life, 1), "months\n")
cat("Shelf life from Arrhenius k   :", round(t_shelf_arr, 1), "months\n")
```

The two rate constants agree to within a few percent, and the
resulting shelf-life estimates are within a month of each other. The
overlay plot below shows the two predicted curves together with the
observed long-term data points — they are visually indistinguishable.

```{r plot-cross-validation}
t_grid <- seq(0, 36, length.out = 200)
cv_df <- data.frame(
  time = rep(t_grid, 2),
  response = c(
    tem$predict_response(fit_first, t_grid),
    tem$predict_at_target_temp(arr_result, target_temp = 25,
                               new_time = t_grid, R0 = 100)
  ),
  source = rep(c("Long-term fit", "Arrhenius extrapolation"),
               each = length(t_grid))
)

ggplot() +
  geom_line(data = cv_df,
            aes(x = time, y = response, colour = source), linewidth = 1) +
  geom_point(data = pooled_data,
             aes(x = time, y = response), size = 2, alpha = 0.6) +
  geom_hline(yintercept = 90, linetype = "dashed", colour = "red") +
  labs(title = "Cross-Validation: Long-Term Fit vs Arrhenius Extrapolation",
       subtitle = sprintf("k_long-term = %.5f,  k_Arrhenius = %.5f /month",
                          k_longterm, k_25_arr),
       x = "Time (months)", y = "Assay (% of label claim)",
       colour = "Source") +
  theme_minimal(base_size = 12)
```

## Regulatory Decision

The complete evidence package supports the following registration
statement:

| Evidence | Result |
|:---------|:-------|
| Kinetic model | First-order, $R^2 > 0.99$ |
| Multi-batch poolability | $p > 0.25$ — batches pooled |
| Shelf life (point estimate) | `r round(shelf$shelf_life, 1)` months |
| Shelf life (ICH Q1E lower 95 % bound) | `r round(shelf$lower_95, 1)` months |
| Arrhenius $E_a$ | `r round(arr_result$Ea_activation, 0)` J/mol (within 20–120 kJ/mol) |
| Arrhenius $R^2$ | `r round(arr_result$r_squared_arrhenius, 4)` |
| Cross-validation $k(25\,^\circ\mathrm{C})$ agreement | within `r round(100 * abs(k_longterm - k_25_arr) / k_longterm, 1)` % |

**Decision.** A **30-month shelf life** is registered for the API at
25 °C / 60 % RH. The ICH Q1E lower one-sided 95 % confidence bound
(`r round(shelf$lower_95, 1)` months) exceeds 24 months, and the
Arrhenius extrapolation from accelerated data independently confirms
the decay rate. The claim is supported by both the long-term and the
accelerated-aging evidence.

## Exporting a report

`TimeEffectModeler` does not expose a dedicated `$report()` method,
so a registration-ready export is assembled from two standard R
mechanisms. First, the fitted parameters and the shelf-life estimate
are written to CSV with base R's `write.csv()`; this is the
tabular artefact that accompanies the dossier. Second, any plot
returned by `$plot_time_effect()` (or a patchwork composition of
several) is saved as a vector graphic with `ggplot2::ggsave()`.

```{r export-report, eval = FALSE}
# 1. Tabular artefact: fitted parameters + shelf-life estimate
params_df <- data.frame(
  parameter = c(names(fit_first$parameters),
                "shelf_life", "lower_95", "upper_95"),
  value = c(unlist(fit_first$parameters),
            shelf$shelf_life, shelf$lower_95, shelf$upper_95),
  model_type = fit_first$model_type
)
write.csv(params_df, file = "stability_report.csv", row.names = FALSE)

# 2. Graphical artefact: save the annotated first-order plot
ggsave("stability_plot.png", plot = p_custom,
       width = 8, height = 5, dpi = 300)

# 3. Graphical artefact: save the patchwork comparison
ggsave("model_comparison.png", plot = combined,
       width = 12, height = 5, dpi = 300)
```

> **Note.** The chunk above uses `eval = FALSE` so that the vignette
> build does not write files into the user's working directory.
> Remove the option (or wrap the calls in `tempdir()`) when running
> locally.

# Best Practices

## Study Design

| Item | Recommendation | Rationale |
|:-----|:---------------|:---------|
| Time points | At least 5–6, including $t = 0$ | Too few cannot discriminate model types |
| Time span | Cover $\geq 80$ % of the response change | Distant extrapolation inflates error |
| Temperatures | At least 3 for Arrhenius | Required to identify $E_a$ |
| Temperature spread | $\geq 20\,^\circ\mathrm{C}$ | Small gaps make $k$ indistinguishable |
| Batches | At least 3 (ICH requirement) | Enables the poolability test |

## Model Validation

* $R^2 > 0.90$ on the original scale — adequate fit.
* The chosen model must extrapolate physically (e.g. an
  exponential-to-plateau with $R_\infty$ above the specification is
  inappropriate for a degrading API).
* Parameters are physical: $k > 0$, $E_a$ within 20–120 kJ/mol.
* Cross-validate long-term and Arrhenius estimates of $k$ at the
  storage temperature.

## Common Pitfalls

1. **Over-extrapolation.** Predicting far beyond the observed time
   span multiplies the standard error. Keep extrapolation within
   $\sim 1.5\times$ the observed span.
2. **Ignoring batch differences.** Pooling non-poolable batches biases
   the shelf life. Always run `$analyze_multi_batch()` first.
3. **Single-temperature Arrhenius.** Without at least three
   temperatures $E_a$ cannot be identified; the extrapolation has no
   physical basis.
4. **Wrong z-value.** ICH Q1E is **one-sided**: use
   `qnorm(confidence_level)` (= 1.645 for 95 %), not the two-sided
   `qnorm((1 + confidence_level) / 2)` (= 1.96).

# References

1. Arrhenius, S. (1889). *Uber die Reaktionsgeschwindigkeit bei der
   Inversion von Rohrzucker*. Zeitschrift fur physikalische Chemie,
   4(1), 226–248.
2. ICH Q1A(R2) (2003). *Stability Testing of New Drug Substances and
   Products*. International Council for Harmonisation.
3. ICH Q1E (2003). *Evaluation for Stability Data*. International
   Council for Harmonisation.
4. Waterman, K. C., & Adami, R. C. (2005). Accelerated aging:
   prediction of chemical stability of pharmaceuticals. *International
   Journal of Pharmaceutics*, 293(1–2), 101–125.
5. Weibull, W. (1951). A statistical distribution function of wide
   applicability. *Journal of Applied Mechanics*, 18, 293–297.
6. Carstensen, J. T., & Rhodes, C. T. (2000). *Drug Stability:
   Principles and Practices* (3rd ed.). Marcel Dekker.
