---
title: "Parametric Reliability Analysis with Weibull, Exponential, and Lognormal Distributions"
author: "iQualityR Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Parametric Reliability Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(iQualityR.reliability)
library(ggplot2)
```

## 1. Introduction

Reliability engineering quantifies how long a component or system performs
its intended function without failure. The cornerstone of parametric
reliability analysis is the **lifetime distribution**: a probability model
that describes the time-to-failure of a population of units. By fitting a
distribution to observed failure times, the analyst obtains estimates of
key metrics such as:

- **MTTF** (Mean Time To Failure): the expected lifetime.
- **B10 life**: the time by which 10% of the population has failed.
- **B50 life** (median life): the time by which 50% has failed.
- **Reliability function** $R(t)$: the probability of survival past time $t$.
- **Hazard function** $h(t)$: the instantaneous failure rate at time $t$.

This vignette demonstrates how `iQualityR.reliability` fits the three most
common lifetime distributions --- Weibull, exponential, and lognormal ---
using the [ReliabilityAnalyzer] class. All fitting is delegated to
`iQualityR.stat::fit_distribution()`, ensuring consistency with the rest
of the `iQualityR` ecosystem.

## 2. The Three Lifetime Distributions

### 2.1 Weibull Distribution

The Weibull is the **workhorse of reliability engineering** because its
shape parameter $\beta$ (often denoted $k$) lets it model all three phases
of the bathtub curve:

$$R(t) = \exp\left[-\left(\frac{t}{\eta}\right)^{\beta}\right]$$

- $\beta < 1$: decreasing failure rate (early-life / infant mortality).
- $\beta = 1$: constant failure rate (random failures); reduces to exponential.
- $\beta > 1$: increasing failure rate (wear-out failures).

The scale parameter $\eta$ is the **characteristic life**: the time by
which 63.2% of the population has failed.

### 2.2 Exponential Distribution

The exponential is the simplest lifetime model, with a constant hazard
rate $h(t) = \lambda$:

$$R(t) = e^{-\lambda t}, \qquad \mathrm{MTTF} = \frac{1}{\lambda}$$

It is the **memoryless** distribution: the residual lifetime of a surviving
unit is independent of its current age. This makes it suitable for
electronic components that fail randomly, but inappropriate for mechanical
components that wear out.

### 2.3 Lognormal Distribution

The lognormal models lifetimes whose logarithm is normally distributed. It
is common in fatigue, crack-growth, and repair-time studies:

$$R(t) = 1 - \Phi\left(\frac{\ln t - \mu}{\sigma}\right)$$

The hazard rises to a peak and then slowly declines, which can model
processes with a "burn-in" of weak subpopulations.

| Distribution | Parameters | Hazard shape | Typical use |
|:------------|:-----------|:--------------|:------------|
| Weibull | shape $\beta$, scale $\eta$ | flexible (beta < 1, = 1, > 1) | Mechanical, electronic, mixed |
| Exponential | rate $\lambda$ | constant | Electronic, random failures |
| Lognormal | meanlog $\mu$, sdlog $\sigma$ | unimodal (rises then falls) | Fatigue, repair times |

## 3. Fitting a Weibull Distribution

### 3.1 Simulated Failure Data

We simulate 80 failure times from a Weibull with $\beta = 2$ and $\eta = 100$
hours, representing a wear-out failure mode.

```{r simulate-weibull}
set.seed(42)
times <- rweibull(80, shape = 2, scale = 100)
data <- data.frame(time = times)
head(data, 5)
```

### 3.2 Build a Plan and Analyze

A `ReliabilityPlan` holds the analysis configuration; the
`ReliabilityAnalyzer` performs the computation.

```{r weibull-analyze}
plan <- ReliabilityPlan$new(
  time_var     = "time",
  distribution = "weibull",
  method       = "parametric"
)

analyzer <- ReliabilityAnalyzer$new()
result  <- analyzer$analyze(data, plan)

# Estimated parameters
result$distribution_fit$parameters

# Goodness of fit
result$distribution_fit$goodness_of_fit
```

### 3.3 Reliability Metrics

```{r weibull-metrics}
metrics <- result$reliability_metrics
metrics
```

Compare the estimated characteristic life with the true value of 100 hours.

### 3.4 Diagnostic Recommendations

The analyzer returns recommendations based on the shape parameter and
goodness-of-fit.

```{r weibull-recs}
result$diagnostics$recommendations
```

## 4. Visualizing the Weibull Fit

The `ReliabilityPlotter` class produces publication-quality plots. The
`"full"` layout combines a Weibull probability plot, the survival curve,
the hazard curve, and a metrics summary.

```{r weibull-full-panel, fig.cap = "Full reliability panel for the Weibull fit"}
plotter <- ReliabilityPlotter$new()
theme   <- iQualityR.core::IqrTheme$new("academic")
p <- plotter$render(result, theme, type = "full")
print(p)
```

### 4.1 Weibull Probability Plot

The Weibull probability plot linearizes the survival function on
$\ln(t)$ vs $\ln(-\ln(R(t)))$ axes. A straight line indicates a good
Weibull fit.

```{r weibull-prob, fig.cap = "Weibull probability plot"}
plotter$render(result, theme, type = "probability")
```

### 4.2 Survival Curve with Confidence Band

```{r weibull-surv, fig.cap = "Reliability (survival) curve with 95% confidence band"}
plotter$render(result, theme, type = "survival")
```

### 4.3 Hazard Rate Curve

```{r weibull-hazard, fig.cap = "Hazard rate curve"}
plotter$render(result, theme, type = "hazard")
```

## 5. Comparing Distributions

When the Weibull fit is poor, the analyst should try alternative
distributions. Here we fit all three and compare.

```{r compare-fits}
fit_results <- list(
  weibull     = reliability_analysis(data, "time", distribution = "weibull"),
  exponential = reliability_analysis(data, "time", distribution = "exponential"),
  lognormal   = reliability_analysis(data, "time", distribution = "lognormal")
)

# Compare AIC (lower is better)
sapply(fit_results, function(r) r$distribution_fit$aic)
```

The Weibull should have the lowest AIC because the data were simulated
from a Weibull.

## 6. Exponential Distribution: The Constant-Hazard Case

```{r exponential-fit}
set.seed(7)
exp_times <- rexp(100, rate = 0.01)
exp_data  <- data.frame(time = exp_times)

exp_result <- reliability_analysis(exp_data, "time",
                                   distribution = "exponential")
exp_result$distribution_fit$parameters
exp_result$reliability_metrics
```

The MTTF should be close to 100 (= 1 / 0.01).

## 7. Lognormal Distribution

```{r lognormal-fit}
set.seed(11)
lnorm_times <- rlnorm(100, meanlog = log(80), sdlog = 0.4)
lnorm_data  <- data.frame(time = lnorm_times)

lnorm_result <- reliability_analysis(lnorm_data, "time",
                                     distribution = "lognormal")
lnorm_result$distribution_fit$parameters
lnorm_result$reliability_metrics
```

## 8. Using the Task Coordinator

For a complete workflow (analysis + plot + report), use
[IqrReliabilityTask]:

```{r task-workflow}
set.seed(13)
data <- data.frame(time = rweibull(60, shape = 2, scale = 100))
plan <- ReliabilityPlan$new(time_var = "time", distribution = "weibull")

task <- IqrReliabilityTask$new(data, plan)
task$compute()
task$summary()
```

```{r task-plot, fig.cap = "Task-level survival curve"}
task$plot(type = "survival")
```

### 8.1 Excel Report Export

```{r task-excel, eval = FALSE}
# Export to Excel (requires the 'openxlsx' package)
tmp <- tempfile(fileext = ".xlsx")
task$report(format = "excel", path = tmp)
```

## 9. Theoretical Summary

| Distribution | $R(t)$ | $h(t)$ | MTTF |
|:------------|:-------|:-------|:-----|
| Weibull | $\exp[-(t/\eta)^\beta]$ | $(\beta/\eta)(t/\eta)^{\beta-1}$ | $\eta\,\Gamma(1+1/\beta)$ |
| Exponential | $e^{-\lambda t}$ | $\lambda$ | $1/\lambda$ |
| Lognormal | $1-\Phi((\ln t - \mu)/\sigma)$ | $f(t)/R(t)$ | $\exp(\mu + \sigma^2/2)$ |

## 10. References

- Lawless, J. F. (2003). *Statistical Models and Methods for Lifetime
  Data*, 2nd ed. Wiley.
- Meeker, W. Q., & Escobar, L. A. (1998). *Statistical Methods for
  Reliability Data*. Wiley.
- Nelson, W. (2004). *Applied Life Data Analysis*. Wiley-Interscience.
