---
title: "Survival Analysis: Kaplan-Meier Estimation and Cox Proportional Hazards"
author: "iQualityR Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Survival Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(iQualityR.reliability)
library(ggplot2)
```

## 1. Introduction

Survival analysis deals with **time-to-event** data in which some
observations may be **censored** --- that is, the event of interest (death,
failure, relapse) has not been observed for that subject by the end of the
study. Unlike parametric reliability (see the
`reliability_parametric` vignette), survival analysis makes weaker
distributional assumptions:

- **Kaplan-Meier** estimation is fully nonparametric: it estimates the
  survival curve $S(t)$ directly from observed event and censoring times.
- The **Cox proportional hazards** model is semiparametric: it estimates
  the effect of covariates on the hazard without specifying the baseline
  hazard.

This vignette demonstrates both methods using `iQualityR.reliability`. The
underlying computations rely on the `survival` package, which is loaded
on demand via `requireNamespace()`.

## 2. Censoring: Why It Matters

In a reliability or clinical study, an observation is **right-censored**
if the study ends before the subject fails. Treating censored
observations as if they had failed would underestimate survival; treating
them as if they had not been observed would waste information. Survival
methods use the censoring indicator to incorporate partial information:

| Status | Meaning |
|:-------|:--------|
| 1 | Event observed (failure, death, etc.) |
| 0 | Right-censored (study ended before event) |

Let $T_i$ be the true failure time and $C_i$ the censoring time. We
observe $X_i = \min(T_i, C_i)$ and the indicator $\delta_i = I(T_i \le
C_i)$. The Kaplan-Meier estimator uses only the observed event times and
the at-risk set at each event time.

## 3. Kaplan-Meier Estimation

### 3.1 Simulated Survival Data

We simulate 100 subjects from an exponential failure distribution with
rate 0.01 (MTTF = 100) and 80% event probability.

```{r simulate-km}
set.seed(42)
n <- 100
data <- data.frame(
  time   = rexp(n, rate = 0.01),
  status = rbinom(n, 1, 0.8)
)
head(data, 5)
table(data$status)
```

### 3.2 Estimating the Survival Curve

```{r km-fit}
result <- kaplan_meier_estimate(data, "time", "status")

# Number of events and censored
result$n
result$n_events
result$n_censored

# Survival curve (first 6 rows)
head(result$survival_curve)
```

### 3.3 Visualizing the Kaplan-Meier Curve

```{r km-plot, fig.cap = "Kaplan-Meier survival curve with confidence band"}
plan <- ReliabilityPlan$new(
  time_var       = "time",
  status_var     = "status",
  method         = "kaplan_meier",
  censoring_type = "right"
)
analyzer <- ReliabilityAnalyzer$new()
km_result <- analyzer$analyze(data, plan)

plotter <- ReliabilityPlotter$new()
theme <- iQualityR.core::IqrTheme$new("academic")
plotter$render(km_result, theme, type = "kaplan_meier")
```

The step function drops at each observed event time and is flat during
censoring periods.

### 3.4 Median Survival

The median survival time is the time at which the survival curve crosses
0.5. It is the most commonly reported summary statistic in survival
analysis.

```{r median-survival}
result$reliability_metrics$median_survival
```

## 4. The Cox Proportional Hazards Model

### 4.1 The Proportional Hazards Assumption

The Cox model assumes that the hazard for subject $i$ is:

$$h_i(t) = h_0(t) \exp(\beta_1 x_{i1} + \beta_2 x_{i2} + \dots + \beta_p x_{ip})$$

where $h_0(t)$ is an unspecified baseline hazard and $\beta_j$ are the
covariate effects. The key assumption is that the **hazard ratio**
$\exp(\beta_j)$ is constant over time --- hence "proportional hazards".

| Quantity | Interpretation |
|:---------|:----------------|
| $\beta_j > 0$ | Covariate $j$ increases the hazard (risk factor) |
| $\beta_j < 0$ | Covariate $j$ decreases the hazard (protective factor) |
| $\exp(\beta_j) = 2$ | Doubles the hazard at all times |
| $\exp(\beta_j) = 0.5$ | Halves the hazard at all times |

### 4.2 Simulated Covariate Data

```{r cox-data}
set.seed(7)
n <- 120
data <- data.frame(
  time   = rexp(n, rate = 0.01),
  status = rbinom(n, 1, 0.7),
  age    = rnorm(n, mean = 50, sd = 10),
  treat  = rbinom(n, 1, 0.5)
)
head(data, 5)
```

### 4.3 Fitting the Cox Model

```{r cox-fit}
result <- cox_model(data, "time", "status",
                     factors = c("age", "treat"))

# Coefficient table
result$cox_model$coefficients

# Model-level summary
result$cox_model$concordance
result$cox_model$likelihood_ratio_test
```

The **concordance** (C-index) measures the model's discriminative power:
0.5 is random, 1.0 is perfect. The likelihood-ratio test compares the
fitted model to the null model (no covariates).

### 4.4 Proportional Hazards Test

The `cox.zph()` test from the `survival` package checks whether the
proportional hazards assumption is violated. A p-value below 0.05 suggests
a time-varying effect.

```{r ph-test}
result$cox_model$proportional_hazards_test
```

### 4.5 Forest Plot

The forest plot visualizes hazard ratios with their 95% confidence
intervals.

```{r forest-plot, fig.cap = "Cox model forest plot of hazard ratios"}
plotter <- ReliabilityPlotter$new()
plotter$render(result, theme, type = "forest")
```

Points to the right of 1.0 indicate risk-increasing factors; points to
the left indicate protective factors.

## 5. Interpreting the Coefficients

For our simulated data:

- **age**: Each additional year of age multiplies the hazard by
  $\exp(\beta_{\text{age}})$. If the 95% CI includes 1.0, the effect is
  not statistically significant at the 5% level.
- **treat**: A binary treatment indicator. $\exp(\beta_{\text{treat}}) < 1$
  suggests the treatment is protective.

The recommendations returned by the analyzer summarize the significant
factors:

```{r recs}
result$diagnostics$recommendations
```

## 6. Full Workflow with IqrReliabilityTask

For a complete analysis pipeline (compute + plot + report), use the
[IqrReliabilityTask] coordinator:

```{r task-workflow}
plan <- ReliabilityPlan$new(
  time_var       = "time",
  status_var     = "status",
  method         = "cox",
  censoring_type = "right",
  factors        = c("age", "treat")
)

task <- IqrReliabilityTask$new(data, plan)
task$compute()
task$summary()
```

```{r task-forest, fig.cap = "Task-level forest plot"}
task$plot(type = "forest")
```

## 7. Choosing Between Methods

| Method | When to use | Strengths | Limitations |
|:-------|:------------|:----------|:------------|
| Kaplan-Meier | No covariates; purely nonparametric | No distribution assumption; visual | Cannot adjust for covariates |
| Cox PH | Covariates; semiparametric | Adjustable; widely validated | Requires proportional hazards |
| Parametric (Weibull) | Distribution known; small samples | Efficient; extrapolates | Sensitive to distribution misspecification |

## 8. Common Pitfalls

1. **Treating censored observations as failures**: This underestimates
   survival. Always include the status variable (1 = event, 0 = censored).
2. **Violated proportional hazards**: If the PH test fails, consider
   stratified Cox or time-dependent covariates.
3. **Small sample size**: With fewer than 10-15 events per covariate,
   Cox coefficient estimates become unstable.
4. **Tied event times**: The default Efron approximation handles ties
   well; Breslow is faster but less accurate.

## 9. References

- Cox, D. R. (1972). Regression Models and Life-Tables. *Journal of the
  Royal Statistical Society. Series B*, 34(2), 187-220.
- Kaplan, E. L., & Meier, P. (1958). Nonparametric Estimation from
  Incomplete Observations. *Journal of the American Statistical
  Association*, 53(282), 457-481.
- Therneau, T. M., & Grambsch, P. M. (2000). *Modeling Survival Data:
  Extending the Cox Model*. Springer.
