---
title: "Attribute Agreement Analysis Report"
subtitle: "`r params$study_type %||% 'Attribute agreement study'`"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 9
    fig_height: 5.5
params:
  results: NULL
  plan: NULL
  theme_obj: NULL
  timestamp: NULL
  statistics: NULL
  diagnostics: NULL
  data_tables: NULL
  raw_output: NULL
  plan_name: NULL
  study_type: NULL
  kappa_results: NULL
  detection_results: NULL
  plots: NULL
  report_date: NULL
---

```{r attr-report-setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
`%||%` <- function(x, y) if (is.null(x)) y else x
scalar <- function(x, default = NA_real_) {
  if (is.null(x) || length(x) == 0) return(default)
  x[[1]]
}
fmt_num <- function(x, digits = 4) {
  x <- suppressWarnings(as.numeric(scalar(x)))
  if (is.na(x)) return("NA")
  sprintf(paste0("%.", digits, "f"), x)
}
fmt_pct <- function(x, digits = 2) {
  x <- suppressWarnings(as.numeric(scalar(x)))
  if (is.na(x)) return("NA")
  sprintf(paste0("%.", digits, "f%%"), x * 100)
}
fmt_pct100 <- function(x, digits = 2) {
  x <- suppressWarnings(as.numeric(scalar(x)))
  if (is.na(x)) return("NA")
  sprintf(paste0("%.", digits, "f%%"), x)
}
fmt_p <- function(x, digits = 4) {
  x <- suppressWarnings(as.numeric(scalar(x)))
  if (is.na(x)) return("NA")
  if (x < 10^-digits) return(paste0("<", sprintf(paste0("%.", digits, "f"), 10^-digits)))
  sprintf(paste0("%.", digits, "f"), x)
}
fmt_ci <- function(x, digits = 4) {
  if (is.null(x) || length(x) < 2 || any(is.na(x[1:2]))) return("NA")
  paste0("[", fmt_num(x[1], digits), ", ", fmt_num(x[2], digits), "]")
}
fmt_ci_pct <- function(x, digits = 2) {
  if (is.null(x) || length(x) < 2 || any(is.na(x[1:2]))) return("NA")
  paste0("[", fmt_pct(x[1], digits), ", ", fmt_pct(x[2], digits), "]")
}
fmt_chr <- function(x, default = "NA") {
  if (is.null(x) || length(x) == 0 || is.na(x[[1]])) return(default)
  as.character(x[[1]])
}
round_df <- function(x, digits = 4) {
  if (is.null(x) || !is.data.frame(x) || nrow(x) == 0) return(x)
  for (nm in names(x)) if (is.numeric(x[[nm]])) x[[nm]] <- round(x[[nm]], digits)
  x
}
kappa_raw <- params$kappa_results$raw_output %||% NULL
detection_raw <- params$detection_results$raw_output %||% NULL
plots <- params$plots %||% list()
```

<style>
:root {
  --iqr-ink: #1f2933;
  --iqr-muted: #5f6b7a;
  --iqr-line: #d9e2ec;
  --iqr-band: #f5f8fb;
  --iqr-header: #102a43;
  --iqr-accent: #2f80ed;
}

body {
  color: var(--iqr-ink);
  font-size: 15px;
  line-height: 1.55;
}

.main-container {
  max-width: 1160px;
}

h1, h2, h3 {
  color: var(--iqr-header);
  font-weight: 650;
  letter-spacing: 0;
}

h1 {
  border-bottom: 2px solid var(--iqr-line);
  padding-bottom: 0.35rem;
  margin-top: 2.2rem;
}

h2 {
  border-left: 4px solid var(--iqr-accent);
  padding-left: 0.65rem;
  margin-top: 1.8rem;
}

table {
  width: 100%;
  border-collapse: collapse;
  margin: 0.85rem 0 1.35rem 0;
  font-size: 0.94rem;
  border: 1px solid var(--iqr-line);
}

caption {
  caption-side: top;
  text-align: left;
  color: var(--iqr-header);
  font-weight: 650;
  padding: 0 0 0.35rem 0;
}

thead th {
  background: var(--iqr-header);
  color: #ffffff;
  border: 1px solid var(--iqr-header);
  padding: 0.58rem 0.7rem;
  font-weight: 650;
}

tbody td {
  border: 1px solid var(--iqr-line);
  padding: 0.5rem 0.7rem;
  vertical-align: middle;
}

tbody tr:nth-child(even) {
  background: var(--iqr-band);
}

tbody tr:hover {
  background: #eef6ff;
}

td:last-child,
th:last-child {
  text-align: right;
  font-variant-numeric: tabular-nums;
}

pre, code {
  background-color: #f7fafc;
}

.conclusion-panel {
  border: 1px solid var(--iqr-line);
  border-left: 4px solid var(--iqr-accent);
  background: #f8fbff;
  padding: 1rem 1.1rem;
  margin: 1rem 0 1.4rem 0;
}

.conclusion-panel h3 {
  margin-top: 0;
  color: var(--iqr-header);
}

.conclusion-panel ol {
  margin-bottom: 0;
}
</style>

# Study Overview

```{r attr-report-overview-table}
overview_tbl <- data.frame(
  Item = c("Plan", "Study mode", "Scale type", "Report date"),
  Value = c(
    params$plan_name %||% "Attr_Gage_Study",
    params$study_type %||% "attribute agreement",
    fmt_chr(kappa_raw$scale_type %||% "nominal", "nominal"),
    format(Sys.time(), "%Y-%m-%d")
  )
)
knitr::kable(overview_tbl, align = c("l", "r"), caption = "Study overview")
```

Attribute agreement analysis evaluates whether appraisers classify parts consistently, whether appraisers agree with each other, and, when a known reference is available, whether each appraiser agrees with the standard. Kappa corrects the observed agreement for agreement expected by chance, so it is more informative than percent agreement alone.

When repeated trials and a known standard are supplied, appraiser-vs-standard Kappa is calculated by comparing each appraiser-trial stream with the standard and then combining the trial-level Kappa estimates. This preserves the repeated-trial structure of the study instead of treating every rating event as an independent pooled comparison. Response-level Kappa is also reported by collapsing each response into "this response" versus "all other responses"; this helps identify categories that are harder to classify consistently.

# Kappa Analysis

```{r attr-report-kappa-summary}
if (!is.null(kappa_raw)) {
  ci <- kappa_raw$ci %||% c(NA_real_, NA_real_)
  summary_tbl <- data.frame(
    Metric = c("Method", "Kappa", "SE Kappa", "Z", "P(vs > 0)", "95% CI", "Observed agreement", "Expected agreement", "Interpretation"),
    Value = c(
      kappa_raw$method %||% "N/A",
      fmt_num(kappa_raw$kappa %||% kappa_raw$V %||% NA_real_, 4),
      fmt_num(kappa_raw$se %||% kappa_raw$se_V %||% NA_real_, 4),
      fmt_num(kappa_raw$z %||% NA_real_, 4),
      fmt_p(kappa_raw$p_value %||% NA_real_),
      fmt_ci(ci, 4),
      fmt_pct(kappa_raw$Po %||% NA_real_, 2),
      fmt_pct(kappa_raw$Pe %||% NA_real_, 2),
      fmt_chr(kappa_raw$interpretation %||% kappa_raw$grade, "N/A")
    )
  )
  knitr::kable(summary_tbl, caption = "Overall kappa statistics")
} else {
  cat("Kappa analysis was not requested or no kappa result is available.")
}
```

```{r attr-report-kappa-benchmark, fig.height=3.4}
if (!is.null(plots$kappa_benchmark)) print(plots$kappa_benchmark)
```

## Between Appraisers

```{r attr-report-pairwise-table}
if (!is.null(kappa_raw$pairwise_appraisers) && nrow(kappa_raw$pairwise_appraisers) > 0) {
  knitr::kable(round_df(kappa_raw$pairwise_appraisers, 4), caption = "Pairwise appraiser agreement")
} else {
  cat("Pairwise appraiser agreement is not available for this data structure.")
}
```

```{r attr-report-pairwise-plot, fig.height=4.5}
if (!is.null(plots$pairwise_agreement)) print(plots$pairwise_agreement)
```

## Each Appraiser vs Standard

```{r attr-report-standard-table}
if (!is.null(kappa_raw$appraiser_vs_standard) && nrow(kappa_raw$appraiser_vs_standard) > 0) {
  knitr::kable(round_df(kappa_raw$appraiser_vs_standard, 4), caption = "Each appraiser compared with standard")
} else {
  cat("No standard/reference column was supplied, so appraiser-vs-standard agreement is not available.")
}
```

```{r attr-report-standard-plot, fig.height=4.5}
if (!is.null(plots$appraiser_standard)) print(plots$appraiser_standard)
```

## Within Appraiser Repeatability

```{r attr-report-within-table}
if (!is.null(kappa_raw$within_appraiser) && nrow(kappa_raw$within_appraiser) > 0) {
  knitr::kable(round_df(kappa_raw$within_appraiser, 4), caption = "Within-appraiser repeatability by repeated trials")
} else {
  cat("Repeated trial data was not supplied, so within-appraiser repeatability is not available.")
}
```

## Ordinal Kendall Statistics

```{r attr-report-kendall-intro, results='asis'}
if (!is.null(kappa_raw$ordinal)) {
  cat("<p>Ordinal ratings were supplied, so Kendall statistics are reported in addition to Kappa. Kappa evaluates exact agreement, while Kendall coefficients account for the order of rating levels.</p>")
} else {
  cat("Ordinal Kendall statistics are shown only when scale_type is set to 'ordinal' and ordered categories are supplied or inferred.")
}
```

```{r attr-report-kendall-tables}
if (!is.null(kappa_raw$ordinal)) {
  if (!is.null(kappa_raw$ordinal$within_appraiser) && nrow(kappa_raw$ordinal$within_appraiser) > 0) {
    knitr::kable(round_df(kappa_raw$ordinal$within_appraiser, 4), caption = "Kendall's coefficient of concordance within appraiser")
  }
  if (!is.null(kappa_raw$ordinal$between_appraisers) && nrow(kappa_raw$ordinal$between_appraisers) > 0) {
    knitr::kable(round_df(kappa_raw$ordinal$between_appraisers, 4), caption = "Kendall's coefficient of concordance between appraisers")
  }
  if (!is.null(kappa_raw$ordinal$appraiser_vs_standard) && nrow(kappa_raw$ordinal$appraiser_vs_standard) > 0) {
    knitr::kable(round_df(kappa_raw$ordinal$appraiser_vs_standard, 4), caption = "Kendall's correlation coefficient: each appraiser vs standard")
  }
  if (!is.null(kappa_raw$ordinal$all_appraisers_vs_standard) && nrow(kappa_raw$ordinal$all_appraisers_vs_standard) > 0) {
    knitr::kable(round_df(kappa_raw$ordinal$all_appraisers_vs_standard, 4), caption = "Kendall's correlation coefficient: all appraisers vs standard")
  }
}
```

```{r attr-report-kendall-plot, fig.height=4.8}
if (!is.null(plots$kendall_ordinal)) print(plots$kendall_ordinal)
```

## Response-Level Agreement

```{r attr-report-response-table}
if (!is.null(kappa_raw$response_table) && nrow(kappa_raw$response_table) > 0) {
  knitr::kable(round_df(kappa_raw$response_table, 4), caption = "Response-level agreement summary")
}
```

```{r attr-report-response-kappa-table}
if (!is.null(kappa_raw$response_kappa) && nrow(kappa_raw$response_kappa) > 0) {
  knitr::kable(
    round_df(kappa_raw$response_kappa, 4),
    caption = "Response-level Fleiss Kappa"
  )
}
```

```{r attr-report-response-kappa-standard-table}
if (!is.null(kappa_raw$response_kappa_vs_standard) && nrow(kappa_raw$response_kappa_vs_standard) > 0) {
  knitr::kable(
    round_df(kappa_raw$response_kappa_vs_standard, 4),
    caption = "Response-level Kappa compared with standard"
  )
}
```

```{r attr-report-response-plot, fig.height=4.4}
if (!is.null(plots$response_agreement)) print(plots$response_agreement)
```

## Sample-Level Disagreement Diagnostics

Sample-level heatmaps are intentionally not used as the main report graphic for large studies. When many samples are present, a full rating matrix compresses too many sample labels into a small vertical space and becomes unreadable. The table and plot below focus on the samples with the highest disagreement.

```{r attr-report-sample-disagreement-table}
if (!is.null(kappa_raw$sample_disagreement) && nrow(kappa_raw$sample_disagreement) > 0) {
  knitr::kable(
    round_df(utils::head(kappa_raw$sample_disagreement, 12), 4),
    caption = "Top samples by disagreement"
  )
}
```

```{r attr-report-sample-disagreement-plot, fig.height=4.8}
if (!is.null(plots$sample_disagreement)) print(plots$sample_disagreement)
```

# Detection Performance

```{r attr-report-detection-summary}
if (!is.null(detection_raw)) {
  det_tbl <- data.frame(
    Metric = c("Sensitivity", "False negative rate", "Specificity", "False positive rate", "Youden index", "LR+", "LR-", "Prevalence"),
    Estimate = c(
      fmt_pct(detection_raw$detection_rate, 2),
      fmt_pct(detection_raw$false_negative_rate, 2),
      fmt_pct(detection_raw$specificity, 2),
      fmt_pct(detection_raw$false_positive_rate, 2),
      fmt_num(detection_raw$youden_index, 4),
      fmt_num(detection_raw$LR_positive, 4),
      fmt_num(detection_raw$LR_negative, 4),
      fmt_pct(detection_raw$prevalence, 2)
    ),
    CI = c(
      fmt_ci_pct(c(detection_raw$detection_ci$lower, detection_raw$detection_ci$upper), 2),
      "",
      fmt_ci_pct(c(detection_raw$specificity_ci$lower, detection_raw$specificity_ci$upper), 2),
      "", "", "", "", ""
    )
  )
  knitr::kable(det_tbl, caption = "Detection performance")
} else {
  cat("Detection analysis was not requested or no detection result is available.")
}
```

```{r attr-report-detection-tables}
if (!is.null(detection_raw$confusion_matrix_table)) {
  knitr::kable(detection_raw$confusion_matrix_table, caption = "Detection confusion matrix")
}
if (!is.null(detection_raw$risk_table)) {
  knitr::kable(round_df(detection_raw$risk_table, 4), caption = "Classification risk summary")
}
```

```{r attr-report-detection-plots, fig.height=4.5}
if (!is.null(plots$detection_ci)) print(plots$detection_ci)
if (!is.null(plots$detection_metrics)) print(plots$detection_metrics)
```

# Conclusion

```{r attr-report-conclusion, results='asis'}
if (!is.null(kappa_raw)) {
  kappa_value <- kappa_raw$kappa %||% kappa_raw$V %||% NA_real_
  kappa_interpretation <- kappa_raw$interpretation %||% kappa_raw$grade %||% "N/A"
  kappa_action <- if (!is.na(kappa_value) && kappa_value < 0.75) {
    "Review appraiser training, classification criteria, and borderline samples before relying on these classifications for final release decisions."
  } else {
    "The appraiser system shows strong agreement for the supplied study data."
  }
  cat("<div class='conclusion-panel'>\n")
  cat("<h3>Conclusion</h3>\n")
  cat("<ol>\n")
  cat(sprintf("<li><strong>Overall agreement:</strong> Kappa is <strong>%s</strong>, interpreted as <strong>%s</strong>.</li>\n", fmt_num(kappa_value, 4), kappa_interpretation))
  if (!is.null(kappa_raw$ordinal)) {
    kendall_between <- if (!is.null(kappa_raw$ordinal$between_appraisers) && nrow(kappa_raw$ordinal$between_appraisers) > 0) kappa_raw$ordinal$between_appraisers$Coef[[1]] else NA_real_
    kendall_standard <- if (!is.null(kappa_raw$ordinal$all_appraisers_vs_standard) && nrow(kappa_raw$ordinal$all_appraisers_vs_standard) > 0) kappa_raw$ordinal$all_appraisers_vs_standard$Coef[[1]] else NA_real_
    cat(sprintf(
      "<li><strong>Ordinal agreement:</strong> Kendall between-appraiser coefficient is <strong>%s</strong>%s.</li>\n",
      fmt_num(kendall_between, 4),
      if (!is.na(kendall_standard)) sprintf(", and Kendall vs standard is <strong>%s</strong>", fmt_num(kendall_standard, 4)) else ""
    ))
  }
  cat(sprintf("<li><strong>Appraiser decision:</strong> %s</li>\n", kappa_action))
  if (!is.null(detection_raw)) {
    cat(sprintf(
      "<li><strong>Detection performance:</strong> Sensitivity is <strong>%s</strong>, specificity is <strong>%s</strong>, and Youden index is <strong>%s</strong>. False negatives and false positives should be reviewed against business risk and inspection cost.</li>\n",
      fmt_pct(detection_raw$detection_rate, 2),
      fmt_pct(detection_raw$specificity, 2),
      fmt_num(detection_raw$youden_index, 4)
    ))
  }
  cat("</ol>\n")
  cat("</div>\n")
} else if (!is.null(detection_raw)) {
  cat("<div class='conclusion-panel'>\n")
  cat("<h3>Conclusion</h3>\n")
  cat("<ol>\n")
  cat(sprintf(
    "<li><strong>Detection performance:</strong> Sensitivity is <strong>%s</strong>, specificity is <strong>%s</strong>, and Youden index is <strong>%s</strong>.</li>\n",
    fmt_pct(detection_raw$detection_rate, 2),
    fmt_pct(detection_raw$specificity, 2),
    fmt_num(detection_raw$youden_index, 4)
  ))
  cat("<li><strong>Risk review:</strong> False negatives and false positives should be reviewed against business risk and inspection cost.</li>\n")
  cat("</ol>\n")
  cat("</div>\n")
} else {
  cat("<div class='conclusion-panel'>\n")
  cat("<h3>Conclusion</h3>\n")
  cat("<p>No Kappa or detection result was available for conclusion generation.</p>\n")
  cat("</div>\n")
}
```

---

Report generated by iQualityR.msa.
