---
title: "Gage R&R Measurement System Analysis Report"
subtitle: "`r params$study_type %||% 'Gage R&R Study'`"
date: "`r params$report_date %||% Sys.Date()`"
output:
  html_document:
    theme: readable
    highlight: tango
    toc: true
    toc_depth: 3
    number_sections: true
params:
  results: NULL
  plan: NULL
  theme_obj: NULL
  timestamp: NULL
  statistics: NULL
  study_type: NULL
  chart_paths: NULL
  report_date: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6, fig.align = "center")
`%||%` <- function(x, y) if (is.null(x)) y else x
fmt_num <- function(x, digits = 4) ifelse(is.na(x), "NA", sprintf(paste0("%.", digits, "f"), x))
fmt_pct100 <- function(x, digits = 2) ifelse(is.na(x), "NA", sprintf(paste0("%.", digits, "f%%"), x))
round_df <- function(x, digits = 4) {
  if (is.null(x) || !is.data.frame(x) || nrow(x) == 0) return(x)
  for (nm in names(x)) if (is.numeric(x[[nm]])) x[[nm]] <- round(x[[nm]], digits)
  x
}
results <- params$results
chart_paths <- params$chart_paths %||% list()
show_chart <- function(name) {
  path <- chart_paths[[name]]
  if (!is.null(path) && file.exists(path)) knitr::include_graphics(path) else cat("Chart is not available for this report.")
}
```

# Study Overview

| Item | Value |
|---|---|
| Study | `r params$study_type %||% 'Gage R&R Study'` |
| Method | `r results$analysis_method` |
| Operators | `r results$num_operators %||% length(unique(results$data$Operator))` |
| Parts | `r results$num_parts %||% length(unique(results$data$Part))` |
| Repeated measurements per operator-part cell | `r results$num_repeats %||% results$num_measurements` |
| Total measurements | `r results$total_measurements %||% nrow(results$data)` |

Gage R&R separates measurement variation into repeatability, reproducibility, and part-to-part variation. A useful system should contribute little variation relative to the process spread or tolerance and should have enough distinct categories to separate real part differences.

# ANOVA and Variation Components

## ANOVA Table

```{r anova-table}
if (!is.null(results$anova_table) && nrow(results$anova_table) > 0) {
  anova_display <- round_df(results$anova_table, 4)
  # 格式化 P 值为科学计数法或小数
  if ("P" %in% names(anova_display)) {
    anova_display$P <- sapply(anova_display$P, function(x) {
      if (is.na(x)) return(" ")
      if (x < 0.001) return("<0.001")
      sprintf("%.4f", x)
    })
  }
  knitr::kable(anova_display, caption = "ANOVA Table", align = c("l", "r", "r", "r", "r", "r"))
}
```

## Variance Components

```{r variance-table}
vc <- results$variance_components
if (!is.null(vc) && nrow(vc) > 0) {
  vc_display <- vc
  names(vc_display) <- gsub("Variance", "VarComp", names(vc_display), fixed = TRUE)
  names(vc_display) <- gsub("PercentContribution", "%Contribution", names(vc_display), fixed = TRUE)
  names(vc_display) <- gsub("PercentStudyVar", "%StudyVar", names(vc_display), fixed = TRUE)
  names(vc_display) <- gsub("PercentTolerance", "%Tolerance", names(vc_display), fixed = TRUE)
  # 清理 Source 列的前导空格
  if ("Source" %in% names(vc_display)) {
    vc_display$Source <- trimws(vc_display$Source)
  }
  knitr::kable(round_df(vc_display, 4), caption = "Variance Components", align = c("l", "r", "r", "r", "r", "r", "r"))
}
```

## Gage Evaluation

```{r gage-evaluation}
ge <- results$gage_evaluation
if (!is.null(ge) && nrow(ge) > 0) {
  ge_display <- round_df(ge, 4)
  # 清理 Source 列的前导空格
  if ("Source" %in% names(ge_display)) {
    ge_display$Source <- trimws(ge_display$Source)
  }
  knitr::kable(ge_display, caption = "Gage Evaluation Summary", align = c("l", "r", "r", "r", "r", "r"))
}
```

```{r variation-plot}
show_chart("variation")
```

# Key Statistics

```{r key-stats}
stat_tbl <- data.frame(
  Metric = c("Total Gage R&R %Contribution", "Total Gage R&R %Tolerance", "Number of distinct categories"),
  Value = c(
    fmt_pct100(results$statistics$total_gage_rr_percent, 2),
    fmt_pct100(results$statistics$tolerance_percent, 2),
    as.character(results$statistics$ndc)
  ),
  Decision = c(
    ifelse(results$statistics$total_gage_rr_percent < 10, "Accept", ifelse(results$statistics$total_gage_rr_percent < 30, "Marginal", "Reject")),
    ifelse(is.na(results$statistics$tolerance_percent), "Not evaluated", ifelse(results$statistics$tolerance_percent < 10, "Accept", ifelse(results$statistics$tolerance_percent < 30, "Marginal", "Reject"))),
    ifelse(results$statistics$ndc >= 5, "Accept", ifelse(results$statistics$ndc >= 4, "Marginal", "Reject"))
  )
)
knitr::kable(stat_tbl, caption = "Decision statistics")
```

# Graphical Diagnostics

## Measurements by Part

```{r by-part}
show_chart("by_part")
```

## Measurements by Operator

```{r by-operator}
show_chart("by_operator")
```

## Part by Operator Interaction

```{r interaction}
show_chart("interaction")
```

## Control Charts

```{r xbar-chart}
show_chart("xbar_chart")
```

```{r r-chart}
show_chart("r_chart")
```

# Interpretation

```{r conclusion}
interpretation <- results$interpretation
cat(sprintf("Overall decision: **%s**. %s\n\n", toupper(interpretation$status), interpretation$message))
cat(sprintf("Total Gage R&R contributes %s of total variance contribution and the number of distinct categories is %d. ", fmt_pct100(interpretation$gage_rr_percent, 2), interpretation$ndc))
if (interpretation$status == "accept") {
  cat("The measurement system is suitable for the supplied study conditions.")
} else if (interpretation$status == "marginal") {
  cat("Use may be acceptable depending on application risk, but repeatability, reproducibility, and operator method should be reviewed.")
} else {
  cat("The measurement system should not be released for routine decisions without improvement and revalidation.")
}
```

---

Report generated by iQualityR.msa.
