---
title: "Type 1 Gage Study Report"
subtitle: "`r ifelse(params$study_type == 'bias', 'Single reference bias and capability', 'Gage linearity and bias')`"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output:
  html_document:
    theme: cosmo
    highlight: tango
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 9
    fig_height: 5.5
params:
  results: NULL
  plan: NULL
  theme_obj: NULL
  timestamp: NULL
  study_type: "bias"
  statistics: NULL
  diagnostics: NULL
  data_tables: NULL
  raw_output: NULL
  plots: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
`%||%` <- function(x, y) if (is.null(x)) y else x
fmt_num <- function(x, digits = 4) ifelse(is.na(x), "NA", sprintf(paste0("%.", digits, "f"), x))
fmt_pct100 <- function(x, digits = 2) ifelse(is.na(x), "NA", sprintf(paste0("%.", digits, "f%%"), x))
fmt_p <- function(x) ifelse(is.na(x), "NA", ifelse(x < 0.0001, "<0.0001", sprintf("%.4f", x)))
round_df <- function(x, digits = 4) {
  if (is.null(x) || !is.data.frame(x) || nrow(x) == 0) return(x)
  for (nm in names(x)) if (is.numeric(x[[nm]])) x[[nm]] <- round(x[[nm]], digits)
  x
}
s <- params$statistics
d <- params$diagnostics
plots <- params$plots %||% list()
```

# Executive Summary

This report follows a Type 1 gage study structure: repeated measurements are compared with a known reference value to estimate bias, repeatability, and gage capability. Capability is judged with Cg and Cgk, while the VDA 5 uncertainty view separates repeatability, resolution, bias, calibration, and other uncertainty components.

```{r summary-table}
if (params$study_type == "bias") {
  summary_df <- data.frame(
    Metric = c("Number of measurements", "Reference value", "Mean measurement", "StDev", "6 * StDev", "Tolerance", "k factor", "Bias", "T", "P-value", "Cg", "Cgk", "%Var (Repeatability)", "%Var (Repeatability and Bias)", "VDA5 u_EVR (=s)", "VDA5 u_BI (=|Bias|)", "VDA5 u_MS", "VDA5 Expanded U (k=2)", "VDA5 %QMS"),
    Value = c(
      s$n,
      fmt_num(s$reference_value, 4),
      fmt_num(s$mean_meas, 4),
      fmt_num(s$sd_meas, 4),
      fmt_num(6 * s$sd_meas, 4),
      fmt_num(s$tolerance, 4),
      fmt_num(s$k_factor, 2),
      fmt_num(s$bias, 4),
      fmt_num(s$t_stat, 4),
      fmt_p(s$p_value),
      fmt_num(s$Cg, 3),
      fmt_num(s$Cgk, 3),
      fmt_pct100(s$percent_repeatability, 2),
      fmt_pct100(s$percent_repeatability_bias, 2),
      fmt_num(s$vda5_u_evr, 6),
      fmt_num(s$vda5_u_bi, 6),
      fmt_num(s$vda5_u_ms, 6),
      fmt_num(s$vda5_expanded_u, 6),
      fmt_pct100(s$vda5_qms_percent, 2)
    ),
    Decision = c("", "", "", "", "", "", "", "", "", "",
                 ifelse(d$Cg_ok, "PASS", "FAIL"),
                 ifelse(d$Cgk_ok, "PASS", "FAIL"),
                 "", "",
                 "", "", "", "",
                 ifelse(d$vda5_qms_ok, "PASS", "FAIL"))
  )
  knitr::kable(summary_df, caption = "Type 1 gage study summary")
} else {
  summary_df <- data.frame(
    Metric = c("Total measurements", "Reference points", "Tolerance", "Process variation (PV)", "Average bias", "%Avg bias of PV", "Max absolute bias", "%Max bias of PV", "Linearity (|slope|*PV)", "%Linearity of PV", "Linearity 95% CI", "Slope", "Slope 95% CI", "R-squared"),
    Value = c(
      s$n_total,
      s$n_ref_points,
      fmt_num(s$tolerance, 4),
      fmt_num(s$process_variation, 4),
      fmt_num(s$avg_bias, 4),
      fmt_pct100(s$percent_avg_bias, 2),
      fmt_num(s$max_bias, 4),
      fmt_pct100(s$percent_max_bias, 2),
      fmt_num(s$linearity, 4),
      fmt_pct100(s$percent_linearity, 2),
      paste0("[", fmt_num(s$ci_linearity[1], 4), ", ", fmt_num(s$ci_linearity[2], 4), "]"),
      fmt_num(s$slope, 4),
      paste0("[", fmt_num(s$ci_slope[1], 4), ", ", fmt_num(s$ci_slope[2], 4), "]"),
      fmt_num(s$r_squared, 4)
    ),
    Decision = c("", "", "", "", "", "", ifelse(d$percent_max_bias_ok, "PASS", "FAIL"), "", "", "", "", ifelse(d$slope_ok, "PASS", "FAIL"), "", "info")
  )
  knitr::kable(summary_df, caption = "Linearity and bias study summary")
}
```

# Statistical Results

```{r detailed-stats}
if (params$study_type == "bias") {
  detail <- data.frame(
    Item = c("Bias", "95% CI lower", "95% CI upper", "Bias contains zero", "Percent bias of tolerance", "Cg criterion (k*T/(6s))", "Cgk criterion ((k*T/2-|Bias|)/(3s))", "VDA5 u_EVR (single-meas repeatability)", "VDA5 u_BI (uncorrected |Bias|)", "VDA5 capability (<15% capable)"),
    Value = c(
      fmt_num(s$bias, 4),
      fmt_num(s$ci_bias[1], 4),
      fmt_num(s$ci_bias[2], 4),
      ifelse(0 >= s$ci_bias[1] && 0 <= s$ci_bias[2], "YES", "NO"),
      fmt_pct100(s$percent_bias, 2),
      ">= 1.33",
      ">= 1.33",
      fmt_num(s$vda5_u_evr, 6),
      fmt_num(s$vda5_u_bi, 6),
      s$vda5_qms_capability
    )
  )
  knitr::kable(detail, caption = "Bias test, capability criteria, and VDA5 uncertainty")
} else {
  detail <- data.frame(
    Item = c("Process variation (PV)", "Intercept", "Slope", "Slope SE", "Slope t-stat", "Slope p-value", "Slope CI lower", "Slope CI upper", "Linearity (|slope|*PV)", "Linearity CI lower", "Linearity CI upper", "R-squared", "S (regression)", "Average StDev"),
    Value = c(
      fmt_num(s$process_variation, 4),
      fmt_num(s$intercept, 4),
      fmt_num(s$slope, 6),
      fmt_num(s$se_slope, 6),
      fmt_num(s$t_slope, 4),
      fmt_p(s$p_slope),
      fmt_num(s$ci_slope[1], 6),
      fmt_num(s$ci_slope[2], 6),
      fmt_num(s$linearity, 4),
      fmt_num(s$ci_linearity[1], 4),
      fmt_num(s$ci_linearity[2], 4),
      fmt_num(s$r_squared, 4),
      fmt_num(s$s_regression, 6),
      fmt_num(s$avg_sd, 4)
    )
  )
  knitr::kable(detail, caption = "Regression of individual biases on reference value (Minitab Gage Linearity and Bias Study)")
}
```

```{r vda5, eval=params$study_type == "bias"}
if (!is.null(params$data_tables$vda5_uncertainty)) {
  knitr::kable(round_df(params$data_tables$vda5_uncertainty, 4),
               caption = "VDA 5 uncertainty budget (Type A = statistical, Type B = other means; per GUM/VDA 5 5.3)")
}
```

```{r stability, eval=params$study_type == "bias"}
stab <- s$stability_ok
if (!is.null(stab)) {
  stab_df <- data.frame(
    Rule = c("Beyond +-3 sigma (Nelson R1)",
             "9-pt same side (Nelson R2, shift)",
             "6-pt trend (Nelson R3)",
             "2 of 3 beyond +-2 sigma (Nelson R5)",
             "Overall stability"),
    Triggered = c(
      ifelse(isTRUE(s$stability_beyond_3s), "YES", "no"),
      ifelse(isTRUE(s$stability_shift_9), "YES", "no"),
      ifelse(isTRUE(s$stability_trend_6), "YES", "no"),
      ifelse(isTRUE(s$stability_2of3_beyond_2s), "YES", "no"),
      ifelse(is.na(stab), "N/A", ifelse(stab, "no", "YES"))
    )
  )
  knitr::kable(stab_df, caption = "Run-chart stability diagnostics (MSA 4th ed. requires in-control process)")
}
```

```{r ref-summary, eval=params$study_type == "linearity"}
if (!is.null(params$data_tables$ref_summary)) {
  knitr::kable(round_df(params$data_tables$ref_summary, 4),
               caption = "Per-reference bias summary (with stability flags)")
}
```

```{r vda5-linearity, eval=params$study_type == "linearity"}
if (!is.null(params$data_tables$vda5_uncertainty)) {
  knitr::kable(round_df(params$data_tables$vda5_uncertainty, 4),
               caption = "VDA 5 uncertainty budget (Type A = statistical, Type B = other means; u_LIN from linearity per VDA 5 5.3.4)")
}
```

# Graphics

```{r plots, fig.height=5.5}
for (nm in names(plots)) {
  if (inherits(plots[[nm]], "ggplot") || inherits(plots[[nm]], "patchwork")) print(plots[[nm]])
}
```

# Interpretation

```{r conclusion}
if (params$study_type == "bias") {
  ok <- isTRUE(d$Cg_ok) && isTRUE(d$Cgk_ok) && isTRUE(d$vda5_qms_ok)
  cat(sprintf("Overall decision: **%s**.\n\n", ifelse(ok, "PASS", "FAIL")))
  cat(sprintf("The measured bias is %s, representing %s of tolerance. Cg is %s and Cgk is %s.\n\n", fmt_num(s$bias, 4), fmt_pct100(s$percent_bias, 2), fmt_num(s$Cg, 3), fmt_num(s$Cgk, 3)))
  cat(sprintf("VDA 5: u_EVR = s = %s (single-measurement repeatability), u_BI = |Bias| = %s (uncorrected systematic), u_MS = %s, %%QMS = %s (%s).\n\n", fmt_num(s$vda5_u_evr, 6), fmt_num(s$vda5_u_bi, 6), fmt_num(s$vda5_u_ms, 6), fmt_pct100(s$vda5_qms_percent, 2), s$vda5_qms_capability))
  if (!is.null(s$stability_ok) && !is.na(s$stability_ok)) {
    cat(sprintf("Run-chart stability (MSA 4th ed.): %s. Statistical control of the measurement process is a prerequisite for interpreting Cg/Cgk and the bias test.\n\n",
               ifelse(s$stability_ok, "PASS (in control)", "FAIL (out of control)")))
  }
  if (!ok) cat("The gage should be reviewed for repeatability, calibration bias, resolution, and operator procedure before release for production decisions.")
} else {
  ok <- isTRUE(d$slope_ok) && isTRUE(d$r2_ok) && isTRUE(d$percent_linearity_ok) &&
         isTRUE(d$percent_avg_bias_ok) && isTRUE(d$per_ref_bias_ok) &&
         isTRUE(d$vda5_qms_ok) && isTRUE(d$per_ref_stability_ok)
  cat(sprintf("Overall decision: **%s**.\n\n", ifelse(ok, "PASS", "FAIL")))
  cat(sprintf("Linearity = |slope|*PV = %s (%s of process variation); the maximum absolute bias is %s (%s of process variation). Slope 95%% CI = [%s, %s].",
             fmt_num(s$linearity, 4), fmt_pct100(s$percent_linearity, 2),
             fmt_num(s$max_bias, 4), fmt_pct100(s$percent_max_bias, 2),
             fmt_num(s$ci_slope[1], 4), fmt_num(s$ci_slope[2], 4)))
  cat(sprintf("\n\nVDA 5 (linearity study): u_EVR = s_pooled = %s (pooled within-reference repeatability), u_LIN = |Linearity|/sqrt(3) = %s (rectangular, uncorrected), u_MS = %s, %%QMS = %s (%s).",
             fmt_num(s$vda5_u_evr, 6), fmt_num(s$vda5_u_lin, 6),
             fmt_num(s$vda5_u_ms, 6), fmt_pct100(s$vda5_qms_percent, 2),
             s$vda5_qms_capability))
  if (!is.null(d$per_ref_stability_ok)) {
    cat(sprintf("\n\nPer-reference stability (MSA 4th ed.): %s. Each reference's run chart must be in statistical control for the per-reference bias t-tests to be valid.",
               ifelse(d$per_ref_stability_ok, "PASS (all references in control)",
                      "FAIL (some references out of control)")))
  }
  if (!ok) cat("\n\nThe gage should be reviewed for linearity, per-reference bias, and repeatability before release for production decisions.")
}
```

# Appendix

```{r protocol}
cat(sprintf("Specification limits: LSL = %s, USL = %s\n", fmt_num(s$spec_limits$lsl, 4), fmt_num(s$spec_limits$usl, 4)))
cat(sprintf("Tolerance: %s\n", fmt_num(s$tolerance, 4)))
```

---

Report generated by iQualityR.msa.
