---
title: "Analysis of Variance for Process Comparison"
author: "iQualityR.stat Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Analysis of Variance for Process Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

library(iQualityR.stat)
library(iQualityR.plot)
library(ggplot2)

# Helpers
fmt_num <- function(x, digits = 4) formatC(x, format = "f", digits = digits)
fmt_pct <- function(x, digits = 2) paste0(formatC(x * 100, format = "f", digits = digits), "%")
```

## 1. Introduction

When a quality engineer needs to compare **more than two** groups at once -
three suppliers, four machines, five catalyst formulations - pairwise t-tests
are the wrong tool. With $k$ groups there are $\binom{k}{2}$ pairwise
comparisons, and the family-wise Type I error inflates rapidly: five groups
produce ten comparisons, and at $\alpha = 0.05$ the chance of at least one false
alarm is $1 - (1-0.05)^{10} \approx 40\%$.

**Analysis of variance (ANOVA)** solves this by testing the single global null
$H_0: \mu_1 = \mu_2 = \dots = \mu_k$ with one F-test at level $\alpha$, and only
then - if the global test rejects - drilling into pairwise comparisons with a
procedure that controls the family-wise error rate. **Analysis of means (ANOM)**
is a complementary graphical technique that compares each group mean to the
grand mean with decision limits derived from the studentized range; it answers
a slightly different question ("which groups differ from the average?") and is
often more intuitive for shop-floor audiences.

This vignette covers the `iqr_anova` module, which provides:

- **Seven analysis types**, auto-detected from the formula: one-way, two-way,
  multi-factor, repeated-measures, mixed-effects, MANOVA, and ANOM.
- **Automatic Tukey HSD** pairwise comparisons for every aov-based fit.
- **Effect-size reporting** (eta-squared and partial eta-squared).
- **Eight plot types** including the ANOM decision chart, interaction plots,
  F-curve rejection-region plots, and a residual diagnostic panel.

## 2. Theory: The Analysis of Variance

### 2.1 Partitioning total variation

The fundamental identity of ANOVA partitions the total sum of squares into
between-group and within-group components:

$$
\underbrace{\sum_{i=1}^{k}\sum_{j=1}^{n_i}(y_{ij} - \bar{y}_{\cdot\cdot})^2}_{SS_T \text{ (total)}}
= \underbrace{\sum_{i=1}^{k} n_i (\bar{y}_{i\cdot} - \bar{y}_{\cdot\cdot})^2}_{SS_B \text{ (between / treatment)}}
+ \underbrace{\sum_{i=1}^{k}\sum_{j=1}^{n_i}(y_{ij} - \bar{y}_{i\cdot})^2}_{SS_W \text{ (within / error)}}
$$

where $\bar{y}_{i\cdot}$ is the mean of group $i$, $\bar{y}_{\cdot\cdot}$ is the
grand mean, $k$ is the number of groups, and $n_i$ is the size of group $i$.

### 2.2 The F-statistic and its null distribution

Under the null hypothesis $H_0: \mu_1 = \dots = \mu_k$ and the assumptions of
independence, normality, and equal variance, the ratio

$$
F = \frac{SS_B / (k-1)}{SS_W / (N-k)} = \frac{MS_B}{MS_W}
$$

follows an F-distribution with $(k-1, N-k)$ degrees of freedom, where
$N = \sum n_i$. We reject $H_0$ when $F > F_{1-\alpha,\, k-1,\, N-k}$. A large
F means the between-group variation is large relative to the within-group
("noise") variation - i.e., the group means differ more than chance would
produce.

### 2.3 The one-way ANOVA model

$$
y_{ij} = \mu + \alpha_i + \varepsilon_{ij}, \qquad
\varepsilon_{ij} \stackrel{iid}{\sim} N(0, \sigma^2),
\quad \sum n_i \alpha_i = 0
$$

where $\alpha_i = \mu_i - \mu$ is the treatment effect of group $i$. The
constraint $\sum n_i \alpha_i = 0$ makes the model identifiable. The ANOVA
table has one treatment row (Df $= k-1$, SS $= SS_B$) and one error row
(Df $= N-k$, SS $= SS_W$).

### 2.4 Two-way ANOVA and interactions

With two factors $A$ (levels $a$) and $B$ (levels $b$), the model becomes:

$$
y_{ijk} = \mu + \alpha_i + \beta_j + (\alpha\beta)_{ij} + \varepsilon_{ijk}
$$

The ANOVA table now has **three treatment rows**: factor $A$ (Df $= a-1$),
factor $B$ (Df $= b-1$), and the **interaction** $A \times B$
(Df $= (a-1)(b-1)$). The interaction term answers: *"Does the effect of $A$
depend on the level of $B$?"* If the interaction is significant, the main
effects must be interpreted conditionally - an interaction plot (parallel lines
$\Rightarrow$ no interaction; crossing or fanning lines $\Rightarrow$ interaction)
is the standard visual diagnostic.

### 2.5 Multiple comparisons: Tukey's HSD

When the global F-test rejects, Tukey's Honestly Significant Difference
controls the family-wise error rate at $\alpha$ for all $\binom{k}{2}$ pairwise
comparisons. The procedure compares each pair $| \bar{y}_{i\cdot} -
\bar{y}_{j\cdot} |$ to the critical value:

$$
HSD = q_{1-\alpha,\, k,\, N-k} \cdot \sqrt{\frac{MS_W}{2}\left(\frac{1}{n_i} + \frac{1}{n_j}\right)}
$$

where $q_{1-\alpha,\, k,\, N-k}$ is the studentized range quantile. The
`iqr_anova` module computes Tukey HSD automatically via `emmeans` and stores
the result in `$multiple_comparisons`.

### 2.6 Analysis of Means (ANOM)

ANOM, introduced by Ott (1967) and standardized in ISO 22518-2, takes a
different angle: instead of comparing groups to each other, it compares each
group mean $\bar{y}_{i\cdot}$ to the **grand mean** $\bar{y}_{\cdot\cdot}$ using
decision limits derived from the studentized range:

$$
\bar{y}_{\cdot\cdot} \pm h_\alpha \cdot \sqrt{MS_W \left(\frac{1}{n_i} - \frac{1}{N}\right)},
\qquad h_\alpha = \frac{q_{1-\alpha,\, k,\, N-k}}{\sqrt{2}}
$$

Any group whose mean falls outside the limits is flagged as significantly
different from the grand average. ANOM is more intuitive than ANOVA for
shop-floor audiences because it produces a single chart with explicit
"acceptance limits" - it looks like a control chart for group means.

### 2.7 Effect sizes

The **eta-squared** $\eta^2 = SS_B / SS_T$ is the proportion of total variation
explained by the treatment. The **partial eta-squared**
$\eta^2_p = SS_B / (SS_B + SS_W)$ is the proportion of variation explained by
the treatment *after* accounting for other terms in the model. Both are returned
by `iqr_anova` in the `$effect_size` field.

| Convention | $\eta^2$ (Cohen, 1988) |
|------------|------------------------|
| Small      | 0.01                   |
| Medium     | 0.06                   |
| Large      | 0.14                   |

### 2.8 Assumptions and their checks

| Assumption | How `iqr_anova` surfaces it |
|------------|------------------------------|
| Independence of observations | Design responsibility (randomization) |
| Normality of residuals | Residual panel (`plot_type = "residual"`) |
| Homogeneity of variance | Levene's test (see Section 8) |
| Additivity (two-way: no interaction) | Interaction plot (`plot_type = "interaction"`) |

## 3. The iqr_anova Class — Unified Entry Point

All seven ANOVA types share one R6 entry point. The formula **auto-detects**
the analysis type; `method = "anom"` overrides to force Analysis of Means.

```r
obj <- iqr_anova$new(theme = "academic")
obj$run(yield ~ supplier, data = df)               # one-way (auto)
obj$run(yield ~ temp * catalyst, data = df)         # two-way (auto)
obj$run(y ~ A + Error(subject), data = df)          # repeated measures (auto)
obj$run(y ~ supplier, data = df, method = "anom")   # force ANOM
obj$plot(plot_type = "auto")                        # effects / interaction / anom / ...
obj$report(format = "excel")                        # multi-sheet workbook
```

### 3.1 Method reference

| Method | Signature | Returns |
|--------|-----------|---------|
| `$new()` | `theme = "academic"` | `iqr_anova` object |
| `$run()` | `formula, data, ..., method = NULL, plot = FALSE, plot_type = "auto", interpret = FALSE, audience = "manager", data_name = NULL` | invisible `self` |
| `$plot()` | `plot_type = "auto"` | ggplot / patchwork |
| `$report()` | `format = "excel", path = NULL` | file path |

### 3.2 Auto-detection rules

| Formula feature | Analysis type |
|-----------------|---------------|
| `Error(...)` term | Repeated-measures ANOVA |
| `(1\|...)` random effect | Linear mixed model (`lmerMod`) |
| `cbind(...)` response | MANOVA |
| One factor | One-way ANOVA |
| Two factors | Two-way ANOVA (with interaction) |
| Three or more factors | Multi-factor ANOVA |
| `method = "anom"` | Analysis of Means (one-way only) |

### 3.3 Plot-type reference

| `plot_type` | What it shows | Applicable to |
|-------------|---------------|---------------|
| `"auto"` | Smart selection (see below) | All |
| `"effects"` | Main-effects plot (mean $\pm$ 1 SE per level) | One-way / two-way / multi-factor |
| `"interaction"` | Two-way interaction plot (parallel-lines check) | Two-way / multi-factor |
| `"comparison"` | Forest plot of Tukey HSD pairwise CIs | One-way / two-way / multi-factor |
| `"residual"` | 4-panel diagnostic (residuals vs fitted, Q-Q, scale-location, vs factor) | All aov-based |
| `"f_curve"` | F-distribution rejection-region plot with observed F | One-way / two-way / multi-factor / mixed |
| `"anom"` | ANOM decision chart with grand mean and limits | ANOM only |
| `"variance"` | Variance-components bar chart | Mixed models only |
| `"summary"` | 2x2 patchwork dashboard (effects + residuals + interaction + table) | One-way / two-way |

The `"auto"` selection logic:
- ANOM result $\to$ `"anom"`
- Mixed model (`lmerMod`) $\to$ `"variance"`
- Two or more factors $\to$ `"interaction"`
- Tukey comparisons present $\to$ `"comparison"`
- Otherwise $\to$ `"effects"`

## 4. Quality Scenario 1 — Three Suppliers' Tensile Strength (One-Way ANOVA + Tukey)

### 4.1 Business context

A composite-materials manufacturer must qualify a third carbon-fiber supplier.
The critical-to-quality (CTQ) attribute is tensile strength (MPa); the
specification calls for a minimum mean of 350 MPa. Three suppliers submit
sample panels; the question is whether the three suppliers differ in mean
tensile strength at $\alpha = 0.05$, and if so, which pairs differ.

The data below is consistent with the cotton-fiber tensile study reported in
Montgomery (2019, Ch. 3) - three suppliers, unbalanced sample sizes (8/8/7
panels), within-supplier SD ~ 8-10 MPa, with supplier C visibly lower.

```{r tensile_setup}
# Tensile strength (MPa) of carbon-fiber panels from three suppliers.
# Realistic values: supplier A and B comparable (~ 360 MPa), C lower (~ 345 MPa).
tensile_data <- data.frame(
  strength = c(
    # Supplier A (n = 8)
    365.2, 358.7, 361.4, 367.0, 354.8, 363.5, 359.1, 362.3,
    # Supplier B (n = 8)
    358.0, 362.5, 355.3, 360.8, 364.1, 357.6, 361.9, 359.4,
    # Supplier C (n = 7) - visibly lower
    348.1, 342.6, 345.9, 351.3, 339.8, 347.2, 343.5
  ),
  supplier = factor(rep(c("A", "B", "C"), c(8, 8, 7)))
)

# Quick group summaries
aggregate(strength ~ supplier, data = tensile_data,
          FUN = function(x) c(mean = mean(x), sd = sd(x), n = length(x)))
```

### 4.2 Plan -> Task -> Analysis -> Summary/Plot -> Report

**Plan**. Response: tensile strength (MPa). Factor: supplier (3 levels).
Hypotheses: $H_0: \mu_A = \mu_B = \mu_C$ vs $H_1$: at least one differs.
Design: one-way, independent samples, $\alpha = 0.05$. Sample sizes are
unbalanced but close, so the F-test is robust.

**Task**. Fit the one-way ANOVA.

```{r tensile_anova}
ao <- iqr_anova$new(theme = "academic")
ao$run(strength ~ supplier, data = tensile_data)
r <- ao$last_results

# ANOVA table (Df, Sum Sq, Mean Sq, F value, Pr(>F))
print(r$anova_table)

cat(sprintf("\nR-squared = %.4f, Adj R-squared = %.4f\n",
            r$r_squared, r$adj_r_squared))
cat(sprintf("Eta-squared = %.4f, Partial eta-squared = %.4f\n",
            r$effect_size$eta, r$effect_size$partial_eta))
```

**Analysis**. The F-test rejects $H_0$ at $\alpha = 0.05$: the three suppliers
do not all share the same mean tensile strength. The effect size
($\eta^2 \approx 0.90$) is very large - supplier identity explains about 90 %
of the variation in the sample, which is consistent with supplier C being
visibly lower.

```{r tensile_effects_plot, fig.cap="Main-effects plot: mean tensile strength per supplier with 1 SE bars."}
ao$plot(plot_type = "effects")
```

### 4.3 Tukey HSD pairwise comparisons

The global F-test tells us "at least one differs"; Tukey HSD tells us *which*
pairs differ, controlling the family-wise error rate across all three
comparisons.

```{r tensile_tukey}
# The Tukey HSD result is cached automatically
print(r$multiple_comparisons)
```

```{r tensile_comparison_plot, fig.cap="Tukey HSD pairwise comparisons: forest plot of mean differences with 95% family-wise CIs."}
ao$plot(plot_type = "comparison")
```

### 4.4 F-curve visualization

```{r tensile_fcurve, fig.cap="F-distribution rejection region with the observed F statistic marked."}
ao$plot(plot_type = "f_curve")
```

### 4.5 Residual diagnostics

```{r tensile_residuals, fig.cap="ANOVA residual diagnostics: residuals vs fitted, Q-Q, scale-location, and by-factor."}
ao$plot(plot_type = "residual")
```

The Q-Q plot checks the normality-of-residuals assumption; the
scale-location plot checks homogeneity of variance. Both should show no
systematic pattern. A formal Levene's test (Section 8) confirms the
equal-variance assumption.

**Report**. Export a multi-sheet workbook for the quality-engineering file.

```{r tensile_report, eval=FALSE}
ao$report(format = "excel", path = "tensile_anova.xlsx")
```

## 5. Quality Scenario 2 — Two-Way ANOVA with Interaction (Injection Molding)

### 5.1 Business context

An injection-molding process for polypropylene casings has two controllable
factors: **melt temperature** (three levels: 200, 220, 240 °C) and **holding
pressure** (two levels: 60, 80 bar). The CTQ is part shrinkage (%). The process
engineer wants to know:

1. Does melt temperature affect shrinkage?
2. Does holding pressure affect shrinkage?
3. Does the effect of temperature *depend on* pressure (interaction)?

This is a $3 \times 2$ factorial with $n = 4$ replicates per cell (24 runs
total). The data below mirrors a real injection-molding study where higher
temperature increases shrinkage and the pressure effect is larger at high
temperature.

```{r mold_setup}
set.seed(13)
temps <- c(200, 220, 240)
pressures <- c(60, 80)
grid <- expand.grid(replicate = 1:4, pressure = pressures, temp = temps)

# Mean shrinkage model: 0.8% baseline, +0.004%/C above 200, -0.05% per bar
# above 60, plus a positive interaction (high T amplifies pressure effect).
shrink <- with(grid,
  0.80 + 0.004 * (temp - 200) - 0.05 * (pressure - 60) +
  0.002 * (temp - 200) * (pressure - 60) +
  rnorm(nrow(grid), 0, 0.05)
)
mold_data <- data.frame(
  shrinkage = round(shrink, 4),
  temp      = factor(grid$temp),
  pressure  = factor(grid$pressure)
)
head(mold_data, 8)
```

### 5.2 Two-way ANOVA with interaction

```{r mold_twoway}
ao <- iqr_anova$new(theme = "academic")
ao$run(shrinkage ~ temp * pressure, data = mold_data)
r2 <- ao$last_results
print(r2$anova_table)
```

The `temp:pressure` interaction row tests whether the temperature effect is the
same at both pressures. A small p-value means the temperature effect depends on
pressure.

```{r mold_interaction_plot, fig.cap="Two-way interaction plot: shrinkage vs temperature, split by pressure. Non-parallel lines indicate an interaction."}
ao$plot(plot_type = "interaction")
```

```{r mold_summary, fig.cap="ANOVA summary dashboard: effects, residuals, interaction, and the ANOVA table."}
ao$plot(plot_type = "summary")
```

### 5.3 Interpreting a significant interaction

If the interaction is significant, the main-effect p-values must **not** be read
in isolation: a significant main effect does not guarantee that the effect
holds at every level of the other factor. The correct follow-up is to examine
simple effects (the effect of one factor at each level of the other) using the
Tukey comparisons cached in `$multiple_comparisons`, or to switch to
`vignette("regression-modeling")` and fit a response-surface model.

## 6. Quality Scenario 3 — Analysis of Means for Multi-Machine Comparison

### 6.1 Why ANOM instead of ANOVA?

A printed-circuit-board fabricator drills via five CNC stations. The QA manager
must present the per-station hole-diameter offset (relative to nominal 1.00 mm)
to shop-floor supervisors. ANOVA answers "are the stations *different from
each other*"; ANOM answers "which stations differ *from the overall average*"
and produces a single chart with explicit accept/reject limits - a much clearer
message for the shop floor.

The data below is consistent with a multi-station drilling study (Montgomery,
2019, Ch. 3 exercise data): five stations, $n = 6$ parts per station, hole
diameter offset in micrometres, with station 3 visibly high.

```{r anom_setup}
# Hole-diameter offset (micrometres) from nominal, 5 CNC stations, n = 6 each.
anom_data <- data.frame(
  offset = c(
    # Station 1: ~ +2.0 um (near grand mean)
    1.8, 2.3, 1.5, 2.1, 2.6, 1.9,
    # Station 2: ~ +2.2 um (near grand mean)
    2.0, 2.4, 1.7, 2.5, 2.1, 2.3,
    # Station 3: ~ +5.0 um (high - the outlier)
    5.1, 4.7, 5.4, 4.9, 5.3, 4.6,
    # Station 4: ~ +2.0 um (near grand mean)
    2.0, 1.7, 2.4, 1.9, 2.2, 1.6,
    # Station 5: ~ +2.3 um (near grand mean)
    2.2, 2.5, 1.9, 2.6, 2.0, 2.4
  ),
  station = factor(rep(c("S1","S2","S3","S4","S5"), each = 6))
)
```

### 6.2 Running ANOM

```{r anom_run}
ao <- iqr_anova$new(theme = "academic")
ao$run(offset ~ station, data = anom_data, method = "anom",
       conf_level = 0.95)
r3 <- ao$last_results

cat(sprintf("Grand mean = %.4f um\n", r3$grand_mean))
cat(sprintf("Decision limits (a = %.4f): L = %.4f, U = %.4f\n",
            r3$h_alpha,
            r3$grand_mean - r3$h_alpha * sqrt(r3$mse * (1/6 - 1/r3$N)),
            r3$grand_mean + r3$h_alpha * sqrt(r3$mse * (1/6 - 1/r3$N))))
print(r3$anom_table)
```

The `anom_table` data frame lists, per station: the level, sample size, mean,
deviation from the grand mean, standard error, lower/upper decision limits, and
a boolean `out_of_limits` flag. Any station with `out_of_limits = TRUE` is
significantly different from the grand mean.

### 6.3 The ANOM decision chart

```{r anom_plot, fig.cap="ANOM decision chart: per-station mean offset vs grand mean, with 95% decision limits. Stations outside the limits differ significantly from the overall average."}
ao$plot(plot_type = "anom")
```

Station 3 falls above the upper decision limit - its mean offset is
significantly higher than the overall average. The shop-floor action is to
investigate station 3's tooling and fixturing; no formal pairwise test is
needed because the ANOM chart already identifies the outlier.

### 6.4 ANOM vs ANOVA on the same data

For completeness, run a standard one-way ANOVA on the same data and compare:

```{r anom_vs_anova}
ao$run(offset ~ station, data = anom_data)  # auto-detects one-way ANOVA
cat(sprintf("ANOVA F = %.4f, p = %.4f\n",
            ao$last_results$anova_table[["F value"]][1],
            ao$last_results$anova_table[["Pr(>F)"]][1]))
```

ANOVA and ANOM agree: the global F-test rejects (the stations differ) and ANOM
identifies *which* station is the outlier. The two procedures are
complementary, not contradictory.

## 7. Quality Scenario 4 — Repeated Measures in a Gage R&R Context

### 7.1 Business context

In a measurement-system analysis (MSA), each of 10 parts is measured once by
each of 4 operators, in random order. The part is the **block**; the operator
is the **treatment**. Because the same part is measured by every operator, the
design is a randomized complete block - the operator effect must be tested
against the part-by-operator interaction, not against pure error. This is the
**repeated-measures ANOVA** setup.

```{r rmm_setup}
set.seed(17)
n_parts <- 10
n_ops   <- 4
part_eff <- rnorm(n_parts, 0, 1.5)  # parts vary
op_eff   <- c(0, 0.05, -0.03, 0.08)  # operators slightly biased
rmm_data <- data.frame(
  measurement = rep(25, n_parts * n_ops) +
                rep(part_eff, each = n_ops) +
                rep(op_eff, n_parts) +
                rnorm(n_parts * n_ops, 0, 0.15),
  part     = factor(rep(paste0("P", 1:n_parts), each = n_ops)),
  operator = factor(rep(paste0("OP", 1:n_ops), n_parts))
)
```

### 7.2 Repeated-measures ANOVA

Use the `Error()` formula to declare the blocking structure:

```{r rmm_run}
ao <- iqr_anova$new(theme = "academic")
ao$run(measurement ~ operator + Error(part), data = rmm_data)
r4 <- ao$last_results

# Repeated-measures returns a LIST of ANOVA tables, one per stratum.
cat("Strata:\n")
print(names(r4$anova_tables))
for (s in names(r4$anova_tables)) {
  cat(sprintf("\n--- Stratum: %s ---\n", s))
  print(r4$anova_tables[[s]])
}
```

**Pitfall.** Repeated-measures stores its tables in `$anova_tables` (plural -
a named list), not `$anova_table`. Always access with double brackets,
`result[["anova_table"]]`, to avoid R's partial-matching on `$`.

## 8. Assumption Checking for ANOVA

ANOVA assumes independence, normality of residuals, and homogeneity of
variance. The first is a design responsibility (randomization). The latter two
are checked formally:

```{r anova_assumptions}
# Re-use the tensile ANOVA model
ao <- iqr_anova$new(theme = "academic")
ao$run(strength ~ supplier, data = tensile_data)
r <- ao$last_results

# Levene's test for equal variance (Brown-Forsythe, robust)
ht <- iqr_htest$new(theme = "academic")
ht$run("levene_test", x = tensile_data$strength, g = tensile_data$supplier,
       center = "median")
cat(sprintf("Levene F = %.4f, p = %.4f\n",
            ht$last_results$statistic, ht$last_results$p.value))

# Normality of residuals
nm <- iqr_normality$new()
nm$test(r$residuals, method = "auto")
```

If Levene's test rejects, switch to Welch's one-way ANOVA (`oneway.test` in
base R) or transform the response (see
`vignette("descriptive-and-distributions")` for Box-Cox / Yeo-Johnson). If the
residuals are non-normal and the group sizes are small, switch to the
Kruskal-Wallis test (`vignette("hypothesis-testing-and-intervals")`).

## 9. Interpreting ANOVA — Decision Guide

| Situation | Recommended type | Follow-up |
|-----------|------------------|-----------|
| One factor, $k \ge 3$ groups, normal residuals, equal variance | One-way ANOVA | Tukey HSD (`$multiple_comparisons`) |
| One factor, non-normal residuals | Kruskal-Wallis (see hypothesis vignette) | Dunn test |
| One factor, unequal variances | `oneway.test` (base R, Welch) | Games-Howell |
| Two factors, no interaction expected | Two-way ANOVA (additive) | Main-effects plots |
| Two factors, interaction suspected | Two-way ANOVA (full model) | Interaction plot + simple effects |
| Three or more factors | Multi-factor ANOVA | Reduced model via stepwise |
| Same subjects measured under all treatments | Repeated-measures ANOVA | `$anova_tables` per stratum |
| Fixed + random effects (e.g., operators random) | Mixed model (`lmerMod`) | `plot_type = "variance"` |
| Multiple correlated responses | MANOVA | Per-response univariate follow-up |
| Shop-floor communication, "which differs from average?" | ANOM | `plot_type = "anom"` |
| Pairwise comparisons after significant F | Tukey HSD (automatic) | `plot_type = "comparison"` |

### Reading the ANOVA table

- **F value, Pr(>F)**: the global test. Small p $\Rightarrow$ at least one group
  differs.
- **Df**: $k - 1$ for treatment, $N - k$ for error. Verify these match the
  design - an off-by-one suggests a data-entry or factor-level mistake.
- **Sum Sq, Mean Sq**: $MS = SS / Df$. The ratio $MS_B / MS_W$ is the F.
- **R-squared**: proportion of total variation explained by the model. High
  $R^2$ with a non-significant F is unusual and worth investigating.

### Choosing between ANOVA and ANOM

- Use **ANOVA** when the question is "do *any* groups differ?" and you will
  follow up with pairwise comparisons.
- Use **ANOM** when the question is "which groups differ from the *average*?"
  and you want a single intuitive chart for a non-statistical audience.
- The two agree on the global decision but flag different contrasts: ANOVA's
  Tukey compares pairs, ANOM compares each group to the grand mean.

## 10. Parameter Reference

### 10.1 `iqr_anova$run()` arguments

| Argument | Type | Default | Meaning |
|----------|------|---------|---------|
| `formula` | formula | (required) | R formula, e.g. `y ~ A * B` or `y ~ A + Error(block)` |
| `data` | data.frame | (required) | Data frame containing the variables |
| `method` | character | `NULL` | `"anom"` forces ANOM; `NULL` auto-detects |
| `plot` | logical | `FALSE` | Whether to plot immediately |
| `plot_type` | character | `"auto"` | See Section 3.3 |
| `interpret` | logical | `FALSE` | Print a short interpretation |
| `audience` | character | `"manager"` | Not deeply wired for ANOVA; reserved |
| `data_name` | character | `NULL` | Optional label for the dataset |

### 10.2 `anova_mixed()` extra arguments

| Argument | Default | Meaning |
|----------|---------|---------|
| `method` | `"REML"` | `lme4` fitting method: `"REML"` or `"ML"` |

### 10.3 `manova()` extra arguments

| Argument | Default | Meaning |
|----------|---------|---------|
| `test` | `"Wilks"` | Multivariate test statistic: `"Wilks"`, `"Pillai"`, `"Hotelling-Lawley"`, `"Roy"` |

### 10.4 Result fields by analysis type

| Field | One-way | Two-way | Multi | Repeated | Mixed | MANOVA | ANOM |
|-------|:-------:|:-------:|:-----:|:--------:|:-----:|:------:|:----:|
| `anova_table` | Y | Y | Y | - | Y | - | Y |
| `anova_tables` (list) | - | - | - | Y | - | - | - |
| `anom_table` | - | - | - | - | - | - | Y |
| `coefficients` | Y | Y | Y | Y | Y | Y | - |
| `r_squared`, `adj_r_squared` | Y | Y | Y | - | - | - | - |
| `effect_size` (eta, partial_eta) | Y | Y | Y | - | - | - | - |
| `multiple_comparisons` (Tukey) | Y | Y | Y | - | Y | - | - |
| `fitted`, `residuals` | Y | Y | Y | Y | Y | Y | - |
| `grand_mean`, `h_alpha`, `mse`, `k` | - | - | - | - | - | - | Y |
| `fixed_effects`, `random_effects` | - | - | - | - | Y | - | - |
| `summary` | Y | Y | Y | Y | - | Y | - |
| `model` (raw fit object) | Y | Y | Y | Y | Y | Y | Y |

## 11. References

1. Montgomery, D. C. (2019). *Introduction to Statistical Quality Control*
   (8th ed.). Wiley. — Foundational reference for ANOVA in quality, including
   the supplier-comparison and injection-molding scenarios.

2. Ott, E. R. (1967). "Analysis of Means - A Graphical Procedure."
   *Industrial Quality Control*, 24(2), 101-109. — The original ANOM paper.

3. Nelson, P. R. (1983). "A Comparison of Sample Sizes for the Analysis of
   Means." *Technometrics*, 25(3), 249-257. — ANOM decision-limit theory.

4. Tukey, J. W. (1949). "Comparing Individual Means in the Analysis of
   Variance." *Biometrics*, 5(2), 99-114. — The HSD procedure.

5. ISO 22518-2:2020. *Statistical Methods for Quality Control - Analysis of
   Means*. — International standard for ANOM.

6. Cohen, J. (1988). *Statistical Power Analysis for the Behavioral Sciences*
   (2nd ed.). Lawrence Erlbaum. — Eta-squared effect-size conventions.

7. Searle, S. R., Casella, G. & McCulloch, C. E. (1992). *Variance Components*.
   Wiley. — Mixed-models theory underlying `anova_mixed`.

8. Pinheiro, J. & Bates, D. M. (2000). *Mixed-Effects Models in S and S-PLUS*.
   Springer. — `lme4` and REML estimation.

## Further Reading

- For the formal hypothesis-test counterparts of the global F-test (Welch
  one-way test, Kruskal-Wallis, Friedman), see
  `vignette("hypothesis-testing-and-intervals", package = "iQualityR.stat")`.
- For the sample-size calculations that size an ANOVA study to a target power,
  see `vignette("sample-size-and-power", package = "iQualityR.stat")`.
- For regression-based alternatives (response-surface methodology, polynomial
  modelling of factor effects), see
  `vignette("regression-modeling", package = "iQualityR.stat")`.
- For the ANOVA visualization infrastructure (F-curve, interaction,
  comparison forest plot, ANOM chart), see
  `vignette("anova-plots", package = "iQualityR.plot")`.
- For process-capability statements on the residuals or per-group variation,
  see `vignette("process-control-and-capability", package = "iQualityR.stat")`.
