---
title: "SPC Control Chart Constants: Statistical Theory and Application Guide"
author: "iQualityR Package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SPC Control Chart Constants: Statistical Theory and Application Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(iQualityR.stat)
```

## Overview

Control chart constants in Statistical Process Control (SPC) are the bridge between sample statistics and process parameter estimates. This document provides a detailed exposition of the definition, derivation, and practical application scenarios of each constant from a mathematical statistics perspective.

## 1. Basic Constants: d2, d3, c4, c4'

### 1.1 d2 Constant: Unbiased Expectation Factor for Range

#### Mathematical Definition

Let $X_1, X_2, ..., X_n$ be independent and identically distributed samples from the normal distribution $N(\mu, \sigma^2)$. Define the sample range:

$$R = \max(X_i) - \min(X_i)$$

Then the relative range is:

$$W = \frac{R}{\sigma}$$

**Mathematical definition of d2**:

$$d_2(n) = E[W] = E\left[\frac{R}{\sigma}\right]$$

#### Theoretical Derivation

According to range distribution theory, the distribution function of $W$ is:

$$F_W(w) = n \int_{-\infty}^{\infty} [\Phi(x+w) - \Phi(x)]^{n-1} \phi(x) dx$$

Where $\Phi$ and $\phi$ are the CDF and PDF of the standard normal distribution, respectively.

Therefore:

$$d_2(n) = \int_0^{\infty} [1 - F_W(w)] dw$$

#### Application Scenarios

**Core Purpose: Estimating Process Standard Deviation**

$$\hat{\sigma} = \frac{\bar{R}}{d_2}$$

Where $\bar{R}$ is the average of multiple subgroup ranges.

**Why is d2 needed?**

- Range $R$ is a biased estimator: $E[R] = d_2 \cdot \sigma$
- Dividing by $d_2$ gives an unbiased estimate
- $d_2$ increases with sample size $n$ (expected range increases with sample size)

#### Function Usage

```{r warning=FALSE}
# Calculate d2 for a single sample size
get_d2(5)
 

# Vectorized calculation for multiple sample sizes
get_d2(c(2, 3, 4, 5))
 

# Practical application: estimating process standard deviation
R_bar <- 4.5  # Average subgroup range
n <- 5        # Subgroup size
sigma_hat <- R_bar / get_d2(n)
sigma_hat
 
```

**Typical Applications**:
- Xbar-R control chart (mean-range chart)
- Short-term process capability studies
- Preferred when subgroup sample size $n \leq 10$

### 1.2 d3 Constant: Standard Deviation Factor for Range

#### Mathematical Definition

$$d_3(n) = \sqrt{\text{Var}(W)} = \sqrt{\text{Var}\left(\frac{R}{\sigma}\right)}$$

#### Theoretical Derivation

First, compute the second moment:

$$E[W^2] = \int_0^{\infty} 2w[1 - F_W(w)] dw$$

Then:

$$d_3(n) = \sqrt{E[W^2] - (E[W])^2} = \sqrt{E[W^2] - d_2^2}$$

#### Application Scenarios

**Core Purpose: Calculating R Chart Control Limits**

$$\sigma_R = d_3 \cdot \sigma$$

Since $\sigma$ is unknown, substitute with $\hat{\sigma} = \bar{R}/d_2$:

$$\hat{\sigma}_R = d_3 \cdot \frac{\bar{R}}{d_2}$$

Therefore, the control limits for the R chart are:

$$UCL_R = \bar{R} + 3\hat{\sigma}_R = \bar{R} + 3\frac{d_3}{d_2}\bar{R} = D_4 \bar{R}$$
$$LCL_R = \bar{R} - 3\frac{d_3}{d_2}\bar{R} = D_3 \bar{R}$$

Where:
$$D_4 = 1 + 3\frac{d_3}{d_2}, \quad D_3 = \max(0, 1 - 3\frac{d_3}{d_2})$$

#### Function Usage

```{r warning=FALSE}
# Calculate d3 value
get_d3(5)
 

# Vectorized calculation
get_d3(c(2, 3, 4, 5))
 

# Practical application: calculating R chart control limits
R_bar <- 4.5
n <- 5
d2 <- get_d2(n)
d3 <- get_d3(n)
D4 <- 1 + 3 * (d3 / d2)
D3 <- max(0, 1 - 3 * (d3 / d2))

UCL_R <- D4 * R_bar
LCL_R <- D3 * R_bar

cat("UCL_R =", UCL_R, "\nLCL_R =", LCL_R)
 
```

**Why is d3 needed?**
- The variability of range depends on sample size $n$
- $d_3$ quantifies this dependency
- The $d_3/d_2$ ratio determines the sensitivity of the control chart

### 1.3 c4 Constant: Unbiased Correction Factor for Standard Deviation

#### Mathematical Definition

Let the sample standard deviation be:

$$s = \sqrt{\frac{1}{n-1}\sum_{i=1}^n (X_i - \bar{X})^2}$$

**Mathematical definition of c4**:

$$c_4(n) = \frac{E[s]}{\sigma}$$

#### Theoretical Derivation

According to chi-square distribution theory:

$$\frac{(n-1)s^2}{\sigma^2} \sim \chi^2_{n-1}$$

Therefore:

$$E[s] = \sigma \cdot \sqrt{\frac{2}{n-1}} \cdot \frac{\Gamma(n/2)}{\Gamma((n-1)/2)}$$

So:

$$c_4(n) = \sqrt{\frac{2}{n-1}} \cdot \frac{\Gamma(n/2)}{\Gamma((n-1)/2)}$$

**Important Properties**:
- $c_4 < 1$ for all $n > 1$
- $\lim_{n \to \infty} c_4 = 1$
- $c_4$ monotonically increases toward 1

#### Function Usage

```{r warning=FALSE}
# Calculate c4 value
get_c4(5)
# [1] 0.94

# Vectorized calculation
get_c4(c(2, 3, 4, 5, 10, 20))
 

# Practical application: estimating process standard deviation based on sample standard deviation
s_bar <- 2.3  # Average subgroup standard deviation
n <- 5
sigma_hat <- s_bar / get_c4(n)
sigma_hat
 

# Verify c4 property: approaches 1 as n increases
n_values <- seq(2, 50, by = 2)
c4_values <- get_c4(n_values)
plot(n_values, c4_values, type = "b", 
     xlab = "Sample size (n)", ylab = "c4",
     main = "c4 constant changes with sample size")
abline(h = 1, lty = 2, col = "red")
```

**Application Scenarios**

**Core Purpose: Estimating Process Standard Deviation Based on Sample Standard Deviation**

$$\hat{\sigma} = \frac{\bar{s}}{c_4}$$

Where $\bar{s}$ is the average of multiple subgroup standard deviations.

**Why is c4 needed?**

- Sample standard deviation $s$ is a biased estimator: $E[s] = c_4 \cdot \sigma < \sigma$
- This is caused by Jensen's inequality: $E[\sqrt{\cdot}] \neq \sqrt{E[\cdot]}$
- $c_4$ corrects this bias

**Typical Applications**:
- Xbar-S control chart (mean-standard deviation chart)
- Preferred when subgroup sample size $n > 10$
- Preferred method in modern SPC software
- Process capability analysis (Cp, Cpk)

**Comparison d2 vs c4**:

| Property | d2 (Range Method) | c4 (Standard Deviation Method) |
|----------|------------------|-------------------------------|
| Efficiency | Lower (usable for n<=10) | Higher (all n) |
| Robustness | Lower (sensitive to outliers) | Higher |
| Computation Complexity | Simple | Slightly more complex |
| Applicable Sample Size | n <= 10 | n >= 2 |

### 1.4 c4' Constant: Bias Correction Factor for MSSD Estimation

#### Mathematical Definition

**Mathematical definition of c4'**:

$$c_4'(n) = E\left[\sqrt{\frac{\text{MSSD}}{2}}\right]$$

Where MSSD (Mean Successive Squared Difference) is defined as:

$$\text{MSSD} = \frac{\sum_{i=2}^n (X_i - X_{i-1})^2}{n-1}$$

#### Theoretical Derivation

For independent and identically distributed normal random variables, the expectation of successive differences satisfies:

$$E[(X_i - X_{i-1})^2] = 2\sigma^2$$

Therefore, MSSD is an unbiased estimator of $2\sigma^2$:

$$E[\text{MSSD}] = 2\sigma^2$$

However, due to the Jensen effect of square root transformation:

$$E\left[\sqrt{\frac{\text{MSSD}}{2}}\right] = c_4' \cdot \sigma$$

Where $c_4' < 1$ is the bias correction factor.

#### Monte Carlo Implementation

c4' is computed via Monte Carlo simulation:

1. Generate standard normal random numbers (sigma=1)
2. Calculate MSSD statistic
3. Calculate $\sqrt{\text{MSSD}/2}$
4. Repeat B times and average

```{r warning=FALSE}
# Internal implementation (C++ accelerated)
# Parameters:
#   n: sample size
#   B: number of simulation iterations (default 1,000,000)
```

#### Function Usage

```{r warning=FALSE}
# Calculate c4_prime value (using default 1 million simulations)
get_c4_prime(5)
 

# Calculate c4_prime for multiple sample sizes
get_c4_prime(c(3, 5, 10, 20))
 

# Use fewer iterations (quick test)
get_c4_prime(5, B = 10000)
 

# Practical application: estimating process standard deviation based on MSSD
data <- c(10.2, 10.5, 10.3, 10.7, 10.4, 10.6, 10.5, 10.8)
n <- length(data)
mssd <- sum(diff(data)^2) / (n - 1)
sigma_hat <- sqrt(mssd / 2) / get_c4_prime(n)
sigma_hat
 
```

#### Application Scenarios

**Core Purpose: Estimating Process Standard Deviation Based on MSSD**

$$\hat{\sigma} = \frac{\sqrt{\text{MSSD}/2}}{c_4'}$$

**Why is c4' needed?**

- MSSD is an estimator of process variance, but the square root introduces bias
- Applicable to **destructive testing** (only one sample per batch)
- Used for variability estimation in **time series data**
- Used in **Measurement System Analysis (MSA)** to evaluate stability

**Typical Application Scenarios**:

1. **Destructive Testing**: Products cannot be measured again after testing, so only single-value sequences can be obtained
2. **Continuous Production Process**: One sample extracted per hour, estimating variation through successive differences
3. **I-MR Control Chart**: Provides more precise sigma estimation for moving range method
4. **Trend Analysis**: Eliminates the effect of serial autocorrelation

**Comparison c4 vs c4'**:

| Property | c4 (Standard Deviation Method) | c4' (MSSD Method) |
|----------|-------------------------------|-------------------|
| Data Requirements | Requires subgroups | Single value sequence |
| Independence Assumption | Within-group independence | Successive differences independent |
| Efficiency | High | Slightly lower |
| Applicable Scenarios | Batch production | Continuous flow, destructive testing |

## 2. Control Chart Constants: A2, A3, D3, D4, B3, B4

### 2.1 A2 Constant: Mean Chart Control Limit Factor Based on Range

#### Mathematical Definition

$$A_2(n, k) = \frac{k}{d_2\sqrt{n}}$$

In standard SPC, $k=3$, corresponding to 3-sigma control limits.

#### Theoretical Derivation

Theoretical formula for Xbar chart control limits:

$$UCL/LCL = \mu \pm k\frac{\sigma}{\sqrt{n}}$$

Substituting $\hat{\sigma} = \bar{R}/d_2$ for $\sigma$:

$$UCL/LCL = \bar{\bar{X}} \pm k\frac{\bar{R}}{d_2\sqrt{n}} = \bar{\bar{X}} \pm A_2 \bar{R}$$

#### Function Usage

```{r warning=FALSE}
# Calculate A2 constant (default k=3)
get_A2(5)
 

# Custom sigma level
get_A2(5, k = 2)  # 2-sigma control limits
 

# Vectorized calculation
get_A2(c(2, 3, 4, 5))
 

# Practical application: calculating X-bar chart control limits
X_double_bar <- 50.2
R_bar <- 4.5
n <- 5
A2 <- get_A2(n)

UCL <- X_double_bar + A2 * R_bar
LCL <- X_double_bar - A2 * R_bar

cat("UCL =", UCL, "\nLCL =", LCL)
 
```

**Purpose**: When using range to estimate standard deviation, calculate control limits for the Xbar chart

**Typical Situations**:
- Subgroup sample size $n \leq 10$
- Traditional manufacturing
- Scenarios friendly to manual calculation

### 2.2 A3 Constant: Mean Chart Control Limit Factor Based on Standard Deviation

#### Mathematical Definition

$$A_3(n, k) = \frac{k}{c_4\sqrt{n}}$$

#### Theoretical Derivation

Substituting $\hat{\sigma} = \bar{s}/c_4$ for $\sigma$:

$$UCL/LCL = \bar{\bar{X}} \pm k\frac{\bar{s}}{c_4\sqrt{n}} = \bar{\bar{X}} \pm A_3 \bar{s}$$

#### Function Usage

```{r warning=FALSE}
# Calculate A3 constant
get_A3(5)
 

# Compare A2 and A3 (at n=5)
A2 <- get_A2(5)
A3 <- get_A3(5)
cat("A2 =", A2, "\nA3 =", A3)
 

# Practical application: X-bar chart based on standard deviation
X_double_bar <- 50.2
s_bar <- 1.8
n <- 5
A3 <- get_A3(n)

UCL <- X_double_bar + A3 * s_bar
LCL <- X_double_bar - A3 * s_bar

cat("UCL =", UCL, "\nLCL =", LCL)
 
```

**Purpose**: When using standard deviation to estimate process variation, calculate control limits for the Xbar chart

**Advantages**:
- More precise than A2 (especially for large samples)
- Compatible with modern calculation methods
- Recommended for automated SPC systems

### 2.3 D3, D4 Constants: R Chart Control Limit Factors

#### Mathematical Definition

$$D_4(n, k) = 1 + k\frac{d_3}{d_2}$$
$$D_3(n, k) = \max\left(0, 1 - k\frac{d_3}{d_2}\right)$$

#### Theoretical Derivation

Statistical model for R chart:

$$E[R] = d_2\sigma$$
$$\sigma_R = d_3\sigma$$

Therefore, control limits for R chart:

$$UCL_R = E[R] + k\sigma_R = d_2\sigma + k d_3\sigma$$

Substituting $\hat{\sigma} = \bar{R}/d_2$:

$$UCL_R = \bar{R} + k\frac{d_3}{d_2}\bar{R} = D_4\bar{R}$$

Similarly:

$$LCL_R = \bar{R} - k\frac{d_3}{d_2}\bar{R} = D_3\bar{R}$$

**Important Properties**:
- When $n \leq 6$, $D_3 = 0$ (lower control limit is 0)
- This is because when sample size is small, the range cannot be small enough to produce a lower bound

#### Function Usage

```{r warning=FALSE}
# Calculate D4 and D3 constants
get_D4(5)
 
get_D3(5)
 

# Vectorized calculation
n_values <- 2:8
D4_vals <- get_D4(n_values)
D3_vals <- get_D3(n_values)

# See when D3 starts being non-zero
data.frame(n = n_values, D3 = D3_vals, D4 = D4_vals)
 

# Practical application: calculating R chart control limits
R_bar <- 4.5
n <- 5
D4 <- get_D4(n)
D3 <- get_D3(n)

UCL_R <- D4 * R_bar
LCL_R <- D3 * R_bar

cat("R chart control limits: UCL =", UCL_R, ", LCL =", LCL_R)
 
```

**Purpose**: Control limit calculation for R chart

**Interpretation**:
- $D_4 > 1$: Upper control limit
- $D_3 < 1$: Lower control limit (may be 0)

### 2.4 B3, B4 Constants: S Chart Control Limit Factors

#### Mathematical Definition

Let $\tau = \sqrt{1 - c_4^2}$, then:

$$B_4(n, k) = 1 + k\frac{\tau}{c_4}$$
$$B_3(n, k) = \max\left(0, 1 - k\frac{\tau}{c_4}\right)$$

#### Theoretical Derivation

Theoretical basis for S chart:

$$E[s] = c_4\sigma$$
$$\sigma_s = \sigma\sqrt{1 - c_4^2}$$

Therefore, control limits are:

$$UCL_s = c_4\sigma + k\sigma\sqrt{1 - c_4^2}$$

Substituting $\hat{\sigma} = \bar{s}/c_4$:

$$UCL_s = \bar{s} + k\frac{\sqrt{1 - c_4^2}}{c_4}\bar{s} = B_4\bar{s}$$

**Note**: $B_3$ and $B_4$ are more precise than $D_3$ and $D_4$ because standard deviation utilizes sample information more efficiently than range.

#### Function Usage

```{r warning=FALSE}
# Calculate B4 and B3 constants
get_B4(5)
 
get_B3(5)
 

# Compare D constants and B constants
n <- 5
D4 <- get_D4(n)
B4 <- get_B4(n)
cat("D4 =", D4, ", B4 =", B4)
 

# When n is large, B constants are more precise
n <- 15
D4 <- get_D4(n)
B4 <- get_B4(n)
cat("n=15: D4 =", D4, ", B4 =", B4)
 

# Practical application: calculating S chart control limits
s_bar <- 2.3
n <- 5
B4 <- get_B4(n)
B3 <- get_B3(n)

UCL_s <- B4 * s_bar
LCL_s <- B3 * s_bar

cat("S chart control limits: UCL =", UCL_s, ", LCL =", LCL_s)
 
```

**Purpose**: Control limit calculation for S chart

**Advantages**:
- Applicable to all sample sizes
- More sensitive than R chart
- Preferred in modern SPC

## 3. Individual Control Chart Constants: E2, d4

### 3.1 E2 Constant: Control Limit Factor for Individual Chart

#### Mathematical Definition

$$E_2(n, k) = \frac{k}{d_2}$$

Note: Here $n$ is typically taken as 2 because moving range is based on successive two points.

#### Theoretical Derivation

For individual control chart (I-MR chart), using moving range:

$$MR_i = |X_i - X_{i-1}|$$
$$\overline{MR} = \frac{\sum_{i=2}^m MR_i}{m-1}$$

Process standard deviation estimate:

$$\hat{\sigma} = \frac{\overline{MR}}{d_2}$$

Therefore, control limits for individual chart:

$$UCL/LCL = \bar{X} \pm k\hat{\sigma} = \bar{X} \pm k\frac{\overline{MR}}{d_2} = \bar{X} \pm E_2 \overline{MR}$$

**Standard Practice**: Usually use $n=2$, so $d_2(2) \approx 1.128$, $E_2 = 3/1.128 \approx 2.660$.

#### Function Usage

```{r warning=FALSE}
# Calculate E2 constant (default n=2)
get_E2(2)
 

# Custom sigma level
get_E2(2, k = 2)
 

# Practical application: calculating I-MR chart individual chart control limits
X_bar <- 50.2
MR_bar <- 3.8
E2 <- get_E2(2)

UCL <- X_bar + E2 * MR_bar
LCL <- X_bar - E2 * MR_bar

cat("I chart control limits: UCL =", UCL, ", LCL =", LCL)
 

# Practical data example
set.seed(123)
individual_values <- rnorm(20, mean = 50, sd = 2)
MR <- abs(diff(individual_values))
MR_bar <- mean(MR)
X_bar <- mean(individual_values)

UCL <- X_bar + get_E2(2) * MR_bar
LCL <- X_bar - get_E2(2) * MR_bar

cat("Process mean:", round(X_bar, 2), 
    "\nControl limits: [", round(LCL, 2), ",", round(UCL, 2), "]")
 
```

### 3.2 d4 Constant: Median Moving Range Factor

#### Mathematical Definition

d4 is the median of the relative range distribution:

$$P(W \leq d_4) = 0.5$$

That is:

$$\int_0^{d_4} f_W(w) dw = 0.5$$

#### Function Usage

```{r warning=FALSE}
# Calculate d4 constant (median range factor)
get_d4(2)
 
get_d4(5)
 

# Vectorized calculation
get_d4(c(2, 3, 4, 5, 6))
 

# Compare d2 and d4 (mean vs median)
n <- 5
d2 <- get_d2(n)
d4 <- get_d4(n)
cat("d2 (mean) =", d2, "\nd4 (median) =", d4)
 

# Practical application: estimating standard deviation based on median moving range
MR_values <- c(2.5, 3.1, 2.8, 4.2, 3.5, 2.9, 3.3, 2.7)
MR_median <- median(MR_values)
n <- 2  # Moving range based on 2 points
sigma_hat <- MR_median / get_d4(n)
sigma_hat
 

# Compare mean method
MR_mean <- mean(MR_values)
sigma_hat_mean <- MR_mean / get_d2(n)
sigma_hat_mean
 
```

#### Application Scenarios

**Purpose**: Estimating standard deviation based on median moving range

$$\hat{\sigma} = \frac{\widetilde{MR}}{d_4}$$

Where $\widetilde{MR}$ is the median of moving ranges.

**Why use median?**

- Median is not sensitive to outliers
- More robust when data contains outliers
- Applicable to non-normal or contaminated distributions

**Comparison**:

| Estimator | Advantages | Disadvantages |
|-----------|-----------|---------------|
| Mean MR (d2) | Efficient | Affected by outliers |
| Median MR (d4) | Robust | Slightly less efficient |

## 4. Quick Reference Tables

### 4.1 Standard Constants Table (k=3)

| n | d2 | d3 | c4 | A2 | A3 | D3 | D4 | B3 | B4 | E2 | d4 |
|---|---|---|---|---|---|---|---|---|---|---|---|
| 2 | 1.128 | 0.853 | 0.798 | 1.880 | 2.659 | 0 | 3.267 | 0 | 3.267 | 2.660 | 0.954 |
| 3 | 1.693 | 0.888 | 0.886 | 1.023 | 1.954 | 0 | 2.575 | 0 | 2.568 | 1.772 | 1.592 |
| 4 | 2.059 | 0.880 | 0.921 | 0.729 | 1.628 | 0 | 2.282 | 0 | 2.266 | 1.457 | 2.060 |
| 5 | 2.326 | 0.864 | 0.940 | 0.577 | 1.427 | 0 | 2.115 | 0 | 2.089 | 1.290 | 2.257 |
| 6 | 2.534 | 0.848 | 0.952 | 0.483 | 1.287 | 0 | 2.004 | 0.030 | 1.970 | 1.184 | 2.534 |
| 7 | 2.704 | 0.833 | 0.959 | 0.419 | 1.182 | 0.076 | 1.924 | 0.118 | 1.882 | 1.109 | 2.704 |
| 8 | 2.847 | 0.820 | 0.965 | 0.373 | 1.099 | 0.136 | 1.864 | 0.185 | 1.815 | 1.054 | 2.847 |
| 9 | 2.970 | 0.808 | 0.969 | 0.337 | 1.032 | 0.184 | 1.816 | 0.239 | 1.761 | 1.010 | 2.970 |
| 10| 3.078 | 0.797 | 0.973 | 0.308 | 0.975 | 0.223 | 1.777 | 0.284 | 1.716 | 0.975 | 3.078 |

### 4.2 Quick Generation of Constants Table

```{r warning=FALSE}
# Generate constants table for common n values
n_values <- 2:10
constants_table <- data.frame(
  n = n_values,
  d2 = round(get_d2(n_values), 3),
  d3 = round(get_d3(n_values), 3),
  c4 = round(get_c4(n_values), 3),
  A2 = round(get_A2(n_values), 3),
  A3 = round(get_A3(n_values), 3),
  D3 = round(get_D3(n_values), 3),
  D4 = round(get_D4(n_values), 3),
  B3 = round(get_B3(n_values), 3),
  B4 = round(get_B4(n_values), 3),
  E2 = round(get_E2(n_values), 3),
  d4 = round(get_d4(n_values), 3)
)

print(constants_table)
```

## 5. Practical Recommendations and Common Mistakes

### 5.1 Choosing Which Constant?

Choose based on actual situations:

1. **Subgroup sample size n <= 10**: Can use d2/d3 (range method)
2. **Subgroup sample size n > 10**: Recommended to use c4 (standard deviation method)
3. **Data contains outliers**: Consider d4 (median method)
4. **Automated SPC systems**: Prefer c4/B3/B4
5. **Manual calculation**: Use d2/D3/D4 (constants are more regular)
6. **Destructive testing/continuous flow**: Use c4' (MSSD method)

### 5.2 Common Mistakes and Precautions

1. **Confusing d2 and c4**
   ```r
   # Wrong: using c4 to estimate sigma for range method
   sigma_hat <- R_bar / get_c4(5)  # X
   
   # Correct: range method uses d2
   sigma_hat <- R_bar / get_d2(5)  # ✓
   
   # Correct: standard deviation method uses c4
   sigma_hat <- s_bar / get_c4(5)  # ✓
   ```

2. **Ignoring sample size effect**
   ```r
   # Wrong: using same constant for all n
   D4 <- 2.115  # This is only valid for n=5
   
   # Correct: calculate based on actual n
   D4 <- get_D4(actual_n)
   ```

3. **Misusing E2**
   ```r
   # Correct: E2 is only for n=2
   E2 <- get_E2(2)  # ✓
   
   # Wrong: using other n values
   E2 <- get_E2(5)  # ⚠️ Technically feasible but meaningless
   ```

4. **Forgetting D3 lower bound**
   ```r
   # D3 is automatically set to 0 when calculation is negative
   D3 <- get_D3(3)  # Returns 0, not negative
   ```

5. **c4' iteration count**
   ```r
   # Default 1 million simulations are precise enough
   c4_prime <- get_c4_prime(5)  # High precision
   
   # Can reduce iterations for quick testing
   c4_prime_fast <- get_c4_prime(5, B = 10000)  # Slightly larger error
   ```

### 5.3 Complete SPC Workflow Example

```{r warning=FALSE}
# Load package
library(iQualityR.stat)

# Example data: 5 subgroups, 5 samples each
set.seed(123)
subgroups <- matrix(rnorm(25, mean = 50, sd = 2), nrow = 5, ncol = 5)

# Calculate subgroup statistics
subgroup_means <- apply(subgroups, 1, mean)
subgroup_ranges <- apply(subgroups, 1, function(x) diff(range(x)))
subgroup_sds <- apply(subgroups, 1, sd)

# Calculate overall statistics
X_double_bar <- mean(subgroup_means)
R_bar <- mean(subgroup_ranges)
s_bar <- mean(subgroup_sds)

# Sample size
n <- ncol(subgroups)

# Method 1: Range method (traditional)
sigma_hat_R <- R_bar / get_d2(n)
A2 <- get_A2(n)
UCL_R_method1 <- X_double_bar + A2 * R_bar
LCL_R_method1 <- X_double_bar - A2 * R_bar

# Method 2: Standard deviation method (modern)
sigma_hat_s <- s_bar / get_c4(n)
A3 <- get_A3(n)
UCL_s_method2 <- X_double_bar + A3 * s_bar
LCL_s_method2 <- X_double_bar - A3 * s_bar

# Compare results
cat("Range method: sigma_hat =", round(sigma_hat_R, 3), 
    ", control limits: [", round(LCL_R_method1, 2), ",", round(UCL_R_method1, 2), "]\n")
cat("Standard deviation method: sigma_hat =", round(sigma_hat_s, 3), 
    ", control limits: [", round(LCL_s_method2, 2), ",", round(UCL_s_method2, 2), "]")
```

## References

1. Montgomery, D. C. (2019). *Introduction to Statistical Quality Control* (8th ed.). Wiley.
2. Shewhart, W. A. (1931). *Economic Control of Quality of Manufactured Product*. Van Nostrand.
3. ASTM E2587-16: Standard Practice for Use of Control Charts in Statistical Process Control.
4. Wheeler, D. J. (1995). *Advanced Topics in Statistical Process Control*. SPC Press.
5. Ryan, T. P. (2011). *Statistical Methods for Quality Improvement* (3rd ed.). Wiley.
6. ISO 7870-2:2013: Control charts - Part 2: Shewhart control charts.
