---
title: "Data Transformation and Distribution Fitting"
author: "iQualityR Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Transformation and Distribution Fitting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)


```

```{r setup, echo = FALSE,include=FALSE}
library(iQualityR.stat)
library(R6)
library(patchwork)
library(iQualityR.core)
library(iQualityR.plot)

```


## Overview

When data does not follow a normal distribution, data transformation and distribution fitting are two commonly used strategies. This module provides 6 transformation methods and automatic distribution fitting.

## Data Transformation

### Supported Transformation Methods

| Method | Description | Data Requirements |
|--------|-------------|-------------------|
| Box-Cox | Optimal lambda transformation | All positive values |
| Yeo-Johnson | Supports positive and negative values | Any real number |
| Johnson | Johnson distribution family | Any real number |
| Log | Logarithmic transformation | All positive values |
| Square Root | Square root transformation | Non-negative values |
| Reciprocal | Reciprocal transformation | Non-zero values |

### Box-Cox Transformation

```{r box_cox}
library(iQualityR.stat)

set.seed(123)
x <- rexp(100, rate = 0.5)

result <- box_cox_transform(x)
result$lambda
result$normality_before
result$normality_after
```

### Yeo-Johnson Transformation

```{r yeo_johnson}
set.seed(123)
x <- rnorm(100, mean = -1, sd = 2)

result <- yeo_johnson_transform(x)
result$lambda
result$normality_after
```

### Logarithmic Transformation

```{r log_transform}
set.seed(123)
x <- rexp(100, rate = 0.5)

result <- log_transform(x, base = "natural")
result$normality_before
result$normality_after
```

### Auto-Select Best Transformation

```{r auto_transform}
set.seed(123)
x <- rexp(100, rate = 0.5)

result <- auto_transform(x)
result$best_method
result$scores
```

### Inverse Transformation

```{r inverse}
set.seed(123)
x <- rexp(100, rate = 0.5)

result <- box_cox_transform(x)
original <- inverse_transform(result$transformed, result$method,
                               list(lambda = result$lambda))

head(original)
```

## Distribution Fitting

### Single Distribution Fitting

```{r fit_single}
set.seed(123)
x <- rnorm(100)

result <- fit_distribution(x, "norm")
result$params
result$AIC
result$BIC
result$ks_test
```

### Auto Distribution Fitting

```{r auto_fit}
set.seed(123)
x <- rexp(100, rate = 0.5)

result <- auto_fit_distribution(x)
result$best_dist
result$ranking
```

### Compare Fits

```{r compare}
set.seed(123)
x <- rnorm(100)

f1 <- fit_distribution(x, "norm")
f2 <- fit_distribution(x, "logis")

result <- compare_fits(list(norm = f1, logis = f2))
result$best_by_ks
result$test_results
```

### Empirical Distribution

```{r empirical}
set.seed(123)
x <- rnorm(100)

result <- empirical_distribution(x)
head(result$points)
```

### QQ Data Calculation

```{r qq_data}
set.seed(123)
x <- rnorm(100)

result <- calc_qq_data(x, "norm", list(mean = 0, sd = 1))
head(result)
```
