---
title: "Descriptive Statistics, Probability Distributions, and Data Transformation"
author: "iQualityR.stat Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Descriptive Statistics, Probability Distributions, and Data Transformation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

library(iQualityR.stat)
library(iQualityR.plot)
library(ggplot2)

# Helpers
fmt_num <- function(x, digits = 4) formatC(x, format = "f", digits = digits)
fmt_pct <- function(x, digits = 2) paste0(formatC(x * 100, format = "f", digits = digits), "%")
```

## 1. Introduction

Before we can test, compare, or control a process, we must first **understand**
it. This vignette covers three foundational tools for characterizing quality
data:

- **Descriptive statistics** — summarize the central tendency, dispersion, and
  shape of a sample.
- **Probability distributions** — model the theoretical population from which
  the sample is drawn.
- **Distribution fitting and data transformation** — match data to a theoretical
  distribution and, when necessary, transform non-normal data to meet the
  normality assumptions required by SPC, capability analysis, and parametric
  tests.

These three topics form a natural workflow: *describe → model → transform*.

## 2. Theory: Measures and Distributions

### 2.1 Measures of Central Tendency

| Measure | Formula | When to Use |
|---------|---------|-------------|
| Mean | $\bar{x} = \frac{1}{n}\sum_{i=1}^{n} x_i$ | Symmetric data; sensitive to outliers |
| Median | $\tilde{x} = x_{((n+1)/2)}$ | Skewed data; robust to outliers |
| Mode | Most frequent value | Categorical or discrete data |

In quality engineering, the **mean** is the primary measure because it directly
relates to process targeting ($\bar{x}$ vs. target) and capability indices
($C_{pk}$ uses $\hat{\mu}$). The **median** is valuable for skewed processes
(e.g., cycle time, time-to-failure).

### 2.2 Measures of Dispersion

| Measure | Formula | Interpretation |
|---------|---------|---------------|
| Standard deviation | $s = \sqrt{\frac{1}{n-1}\sum(x_i - \bar{x})^2}$ | Absolute spread; same units as data |
| Variance | $s^2$ | Squared spread |
| Range | $R = x_{max} - x_{min}$ | Quick estimate; used in R charts |
| IQR | $Q_3 - Q_1$ | Robust spread; used in box plots |
| CV | $CV = s / \bar{x}$ | Relative spread; cross-process comparison |

The **standard error** of the mean, $SE = s / \sqrt{n}$, is critical for
confidence intervals and hypothesis tests.

### 2.3 Measures of Shape

**Skewness** (Fisher-Pearson coefficient):

$$
g_1 = \frac{\frac{1}{n}\sum_{i=1}^{n}(x_i - \bar{x})^3}{\left(\frac{1}{n}\sum_{i=1}^{n}(x_i - \bar{x})^2\right)^{3/2}}
$$

- $g_1 \approx 0$: symmetric (normal-like)
- $g_1 > 0$: right-skewed (common in cycle time, fatigue life)
- $g_1 < 0$: left-skewed

**Excess kurtosis**:

$$
g_2 = \frac{\frac{1}{n}\sum(x_i - \bar{x})^4}{\left(\frac{1}{n}\sum(x_i - \bar{x})^2\right)^2} - 3
$$

- $g_2 \approx 0$: mesokurtic (normal-like)
- $g_2 > 0$: leptokurtic (heavy tails; more extreme values)
- $g_2 < 0$: platykurtic (light tails)

### 2.4 Confidence Interval for the Mean

For a sample from a normal distribution, the $100(1-\alpha)\%$ confidence
interval for the mean is:

$$
\bar{x} \pm t_{\alpha/2, \, n-1} \cdot \frac{s}{\sqrt{n}}
$$

where $t_{\alpha/2, \, n-1}$ is the critical value from Student's t-distribution
with $n-1$ degrees of freedom.

### 2.5 Probability Distributions in Quality Engineering

The table below maps common quality-engineering scenarios to the appropriate
distribution:

| Distribution | PDF/PMF | Quality Application |
|-------------|---------|-------------------|
| Normal | $\frac{1}{\sigma\sqrt{2\pi}} e^{-\frac{(x-\mu)^2}{2\sigma^2}}$ | Dimensional measurements; natural variation |
| Exponential | $\lambda e^{-\lambda x}$ | Time between failures (constant failure rate) |
| Weibull | $\frac{\beta}{\eta}\left(\frac{x}{\eta}\right)^{\beta-1} e^{-(x/\eta)^\beta}$ | Time-to-failure (flexible failure rate) |
| Lognormal | $\frac{1}{x\sigma\sqrt{2\pi}} e^{-\frac{(\ln x - \mu)^2}{2\sigma^2}}$ | Cycle time; particle size |
| Gamma | $\frac{\beta^\alpha}{\Gamma(\alpha)} x^{\alpha-1} e^{-\beta x}$ | Time to $\alpha$ events; queueing |
| Beta | $\frac{x^{\alpha-1}(1-x)^{\beta-1}}{B(\alpha,\beta)}$ | Proportions; Bayesian priors |
| Binomial | $\binom{n}{p} p^x (1-p)^{n-x}$ | Defective counts in $n$ items |
| Poisson | $\frac{\lambda^x e^{-\lambda}}{x!}$ | Defects per unit; rare events |
| t | — | Small-sample inference on means |
| F | — | Variance comparison; ANOVA |
| Chi-square | — | Goodness-of-fit; variance CI |

### 2.6 The Need for Data Transformation

Many quality-engineering tools (t-tests, control charts, capability indices)
assume normality. When the underlying process is inherently non-normal (e.g.,
cycle time is lognormal), applying these tools directly leads to incorrect
conclusions. The **Box-Cox transformation** is the most widely used remedy:

$$
y^{(\lambda)} = \begin{cases} \frac{y^\lambda - 1}{\lambda} & \lambda \neq 0 \\ \ln(y) & \lambda = 0 \end{cases}
$$

The optimal $\lambda$ is found by maximizing the normality of the transformed
data (typically via MLE). The **Yeo-Johnson** transformation extends Box-Cox to
handle negative values, and the **Johnson system** (SU/SB/SL) provides a
three-parameter family for extreme non-normality.

## 3. Descriptive Statistics

### 3.1 The `desc_calc()` Function

`desc_calc()` provides a one-call summary of a numeric vector:

```{r}
# CNC turned shaft diameter (mm)
# Process: target = 25.00 mm, sigma ~ 0.015 mm
# Reference: Montgomery (2013), ISQC 7th Ed., turning operation case
set.seed(2026)
shaft_dia <- rnorm(60, mean = 25.002, sd = 0.015)

d <- desc_calc(shaft_dia)
cat(sprintf("n         = %d\n",   d$n))
cat(sprintf("Mean      = %.4f\n", d$mean))
cat(sprintf("Std dev   = %.4f\n", d$stdev))
cat(sprintf("Median    = %.4f\n", d$median))
cat(sprintf("Min/Max   = %.4f / %.4f\n", d$min, d$max))
cat(sprintf("Q1/Q3     = %.4f / %.4f\n", d$q1, d$q3))
cat(sprintf("IQR       = %.4f\n", d$iqr))
cat(sprintf("Skewness  = %.4f\n", d$skew))
cat(sprintf("Kurtosis  = %.4f (excess)\n", d$kurt))
cat(sprintf("CV        = %.4f%%\n", d$cv * 100))
cat(sprintf("SE        = %.4f\n", d$se))
cat(sprintf("95%% CI    = [%.4f, %.4f]\n", d$ci_mean[1], d$ci_mean[2]))
cat(sprintf("AD stat   = %.4f (p=%.4f)\n", d$ad_stat, d$p_value))
```

The function also returns an Anderson-Darling normality statistic (`ad_stat`)
and its p-value (`p_value`) as a quick normality screen.

### 3.2 Multi-Variable Analysis with `desc_analyze()`

For data frames with multiple measurement columns, `desc_analyze()` computes
descriptive statistics for every numeric variable:

```{r}
# Multi-characteristic measurement: shaft diameter, surface roughness, hardness
set.seed(2026)
measure_df <- data.frame(
  diameter_mm   = rnorm(60, 25.002, 0.015),
  roughness_um  = rnorm(60, 1.60, 0.20),
  hardness_hrc  = rnorm(60, 45.0, 1.5)
)

multi <- desc_analyze(measure_df)

# Summary table
cat(sprintf("%-16s %8s %8s %8s %8s %8s\n",
            "Variable", "Mean", "SD", "CV%", "Skew", "Kurt"))
cat(rep("-", 64), "\n", sep = "")
for (v in names(multi)) {
  r <- multi[[v]]
  cat(sprintf("%-16s %8.4f %8.4f %8.2f %8.4f %8.4f\n",
              v, r$mean, r$stdev, r$cv * 100, r$skew, r$kurt))
}
```

### 3.3 Visualization with `desc_hist()` and `desc_box()`

```{r, fig.width=8, fig.height=4}
# Histogram with density overlay and normal curve
p_hist <- base_plot(data.frame(x = shaft_dia), aes(x)) +
  layers_histogram_density(fill = "steelblue", alpha = 0.6) +
  labs(title = "CNC Shaft Diameter Distribution",
       x = "Diameter (mm)", y = "Density")
p_hist
```

```{r, fig.width=8, fig.height=4}
# Box plot for multi-variable comparison (scale-standardized for visibility)
plot_df <- data.frame(
  Value = c(scale(measure_df$diameter_mm),
            scale(measure_df$roughness_um),
            scale(measure_df$hardness_hrc)),
  Characteristic = rep(names(measure_df), each = nrow(measure_df))
)

p_box <- base_plot(plot_df, aes(Characteristic, Value, fill = Characteristic)) +
  layers_boxplot(add_jitter = TRUE) +
  labs(title = "Multi-Characteristic Comparison (Standardized)",
       x = "Characteristic", y = "Standardized Value") +
  theme(legend.position = "none")
p_box
```

### 3.4 Decision Guide: Interpreting Descriptive Statistics

| Observation | Possible Cause | Recommended Action |
|-------------|---------------|-------------------|
| Mean far from target | Process drift; tool wear | Check control chart; recalibrate |
| High CV (>10%) | Excessive variation | Investigate 4M1E sources |
| |Skew| > 1 | Non-normal process | Transform data; use nonparametric methods |
| Kurtosis > 3 (excess) | Heavy tails; mixed populations | Stratify data; check for mixtures |
| AD p-value < 0.05 | Non-normal | Formal normality test; consider transformation |

## 4. Probability Distributions

### 4.1 The `iqr_prob` Entry Class

The `iqr_prob` R6 class provides an interactive interface for probability
calculations. You register one or more distributions, then compute probabilities
or quantiles.

```{r}
# Create a probability engine
prob <- iqr_prob$new()

# Register a normal distribution: mean = 25.00, sd = 0.015
prob$add_dist(type = "norm", params = list(mean = 25.00, sd = 0.015), reset = TRUE)

# P(X <= 25.03) — probability of a part within upper spec
result <- prob$calc(values = 25.03, mode = "prob", calc_type = "lower")
p_val <- result$all_res[[1]]$result_p
cat(sprintf("P(X <= 25.03) = %.4f\n", p_val))
```

### 4.2 The 12 Registered Distributions

The package supports the following distribution types (using short codes):

| Code | Distribution | Key Parameters | Quality Application |
|------|-------------|----------------|-------------------|
| `norm` | Normal | mean, sd | Dimensional measurements |
| `exp` | Exponential | rate | Constant failure rate |
| `weibull` | Weibull | shape, scale | Time-to-failure |
| `lnorm` | Lognormal | meanlog, sdlog | Cycle time |
| `gamma` | Gamma | shape, rate | Time to $\alpha$ events |
| `beta` | Beta | shape1, shape2 | Proportions |
| `unif` | Uniform | min, max | Random sampling |
| `cauchy` | Cauchy | location, scale | Robust statistics |
| `t` | Student t | df | Small-sample inference |
| `f` | F | df1, df2 | Variance comparison |
| `chisq` | Chi-square | df | Goodness-of-fit |
| `binom` | Binomial | size, prob | Defective counts |
| `pois` | Poisson | lambda | Defects per unit |
| `nbinom` | Negative binomial | size, prob | Overdispersed counts |
| `hyper` | Hypergeometric | m, n, k | Sampling without replacement |
| `geom` | Geometric | prob | Trials until first success |

```{r}
# List all available distributions
list_prob_distributions()
```

### 4.3 Quality Scenario: LED Luminaire Failure Time

**Background.** A manufacturer of LED street luminaires needs to estimate the
probability that a unit survives 50,000 hours of operation. Accelerated life
testing (ALT) at elevated temperature has yielded a Weibull model with shape
parameter $\beta = 2.5$ (indicating wear-out failure mode) and scale parameter
$\eta = 60{,}000$ hours. These parameters are consistent with the L70 lumen
maintenance data reported in IES TM-21-11 for high-quality LED packages
operated at 25°C case temperature.

```{r}
# Weibull model for LED luminaire lifetime
# shape (beta) = 2.5, scale (eta) = 60000 hours
# Reference: IES TM-21-11, extrapolated L70 data for mid-power LEDs

# P(T <= 50000) — probability of failure before 50,000 hours
result <- prob_calc(type = "weibull", x = 50000,
                    params = list(shape = 2.5, scale = 60000))
p_fail <- result$all_res[[1]]$result_p
cat(sprintf("P(failure by 50,000h) = %.4f = %.2f%%\n", p_fail, p_fail * 100))
cat(sprintf("Reliability R(50,000h) = %.4f = %.2f%%\n", 1 - p_fail, (1 - p_fail) * 100))
```

```{r}
# What lifetime corresponds to 1% failure probability (B1 life)?
result_b1 <- prob_calc(type = "weibull", x = 0.01,
                       params = list(shape = 2.5, scale = 60000),
                       mode = "quantile")
b1_life <- result_b1$all_res[[1]]$result_x
cat(sprintf("B1 life (1%% failure) = %.0f hours\n", b1_life))
```

### 4.4 Probability Calculation Modes

The `mode` parameter controls the calculation direction:

| Mode | Input | Output | Question Answered |
|------|-------|--------|------------------|
| `"prob"` | x value | P(X ≤ x) | What is the probability of observing this value? |
| `"quantile"` | probability p | x = F⁻¹(p) | What value corresponds to this probability? |
| `"interval"` | x1, x2 | P(x1 ≤ X ≤ x2) | What is the probability of falling in this range? |

```{r, eval=FALSE}
# Interval probability: P(24.985 <= X <= 25.025) for shaft diameter
prob_calc(type = "norm", x = c(24.985, 25.025),
          params = list(mean = 25.00, sd = 0.015), mode = "interval")
```

## 5. Distribution Fitting

### 5.1 Fitting a Single Distribution

`fit_distribution()` fits a specified distribution to data using maximum
likelihood estimation (MLE) and provides goodness-of-fit statistics:

```{r}
# Fit Weibull to simulated LED failure data
set.seed(2026)
fail_times <- rweibull(80, shape = 2.5, scale = 60000)

fit <- fit_distribution(fail_times, dist = "weibull")
cat(sprintf("Distribution:  %s\n", fit$dist))
cat(sprintf("Shape (beta):  %.4f\n", fit$params$shape))
cat(sprintf("Scale (eta):   %.4f\n", fit$params$scale))
cat(sprintf("Log-likelihood: %.2f\n", fit$logLik))
cat(sprintf("AIC:           %.2f\n", fit$AIC))
cat(sprintf("BIC:           %.2f\n", fit$BIC))
cat(sprintf("KS p-value:    %.4f\n", fit$ks_test$p.value))
```

### 5.2 Auto-Fitting Multiple Candidates

When the underlying distribution is unknown, `auto_fit_distribution()` fits
multiple candidates and ranks them by information criterion:

```{r}
# Auto-fit: try all applicable distributions and rank
afit <- auto_fit_distribution(fail_times)

cat("=== Distribution Ranking ===\n")
for (i in 1:nrow(afit$ranking)) {
  cat(sprintf("  %d. %-12s  AIC = %.2f  KS p = %.4f\n", i,
              afit$ranking$dist[i], afit$ranking$AIC[i], afit$ranking$ks_p[i]))
}
cat(sprintf("\nBest fit: %s\n", afit$best_dist))
```

### 5.3 Comparing Fits

```{r}
# Compare Weibull vs Lognormal vs Gamma for the same data
fits <- list(
  weibull  = fit_distribution(fail_times, dist = "weibull"),
  lognorm  = fit_distribution(fail_times, dist = "lnorm"),
  gamma    = fit_distribution(fail_times, dist = "gamma")
)

cat(sprintf("%-12s %10s %10s %10s %10s\n", "Distribution", "LogLik", "AIC", "BIC", "KS p"))
cat(rep("-", 56), "\n", sep = "")
for (nm in names(fits)) {
  f <- fits[[nm]]
  cat(sprintf("%-12s %10.2f %10.2f %10.2f %10.4f\n",
              nm, f$logLik, f$AIC, f$BIC, f$ks_test$p.value))
}
```

The Weibull fit should have the lowest AIC and the highest KS p-value, since
the data was generated from a Weibull distribution.

### 5.4 Q-Q Data with `calc_qq_data()`

The `calc_qq_data()` function computes the quantile-quantile coordinates needed
to assess distributional fit visually:

```{r, eval=FALSE}
# Compute Q-Q data for the fitted Weibull
qq <- calc_qq_data(fail_times, dist = "weibull",
                   params = list(shape = fit$params$shape, scale = fit$params$scale))
head(qq)
```

## 6. Data Transformation

### 6.1 Box-Cox Transformation

`box_cox_transform()` finds the optimal $\lambda$ and applies the
transformation:

```{r}
# Electroplating cycle time data (seconds)
# Cycle time typically follows a lognormal distribution
# Reference: lognormal cycle time is documented in:
#   John, P. (1971). Statistical Methods in Engineering and Quality Assurance
set.seed(2026)
cycle_time <- rlnorm(100, meanlog = log(180), sdlog = 0.25)

bc <- box_cox_transform(cycle_time)
cat(sprintf("Optimal lambda = %.4f\n", bc$lambda))
cat(sprintf("Method         = %s\n", bc$method))
cat(sprintf("Normality before: Shapiro p = %.4f\n", bc$normality_before))
cat(sprintf("Normality after:  Shapiro p = %.4f\n", bc$normality_after))
```

A $\lambda$ close to 0 confirms the lognormal nature of the data (since
$\lambda = 0$ corresponds to a log transformation). The p-value should improve
substantially after transformation.

### 6.2 Automatic Transformation Selection

`auto_transform()` tries multiple transformation families and selects the best:

```{r}
at <- auto_transform(cycle_time)
cat(sprintf("Best method: %s\n", at$best_method))
cat(sprintf("Criterion:   %s\n", at$criterion))

cat("\n=== Transformation Ranking ===\n")
for (i in seq_along(at$all_results)) {
  r <- at$all_results[[i]]
  cat(sprintf("  %d. %-16s  score = %.4f\n", i, r$method, at$scores[i]))
}
```

### 6.3 Available Transformation Methods

| Function | Formula | When to Use |
|----------|---------|-------------|
| `box_cox_transform()` | $(y^\lambda - 1)/\lambda$ | Positive data; flexible |
| `yeo_johnson_transform()` | Piecewise | Data with zeros or negatives |
| `johnson_transform()` | SU/SB/SL system | Extreme non-normality |
| `log_transform()` | $\ln(y)$ | Multiplicative processes |
| `sqrt_transform()` | $\sqrt{y}$ | Count data; Poisson-like |
| `reciprocal_transform()` | $1/y$ | Rate data |
| `auto_transform()` | Best of above | When unsure |

### 6.4 Inverse Transformation

After computing statistics on transformed data, use `inverse_transform()` to
back-transform results to the original scale:

```{r, eval=FALSE}
# Transform, compute CI on transformed scale, back-transform
bc_data <- bc$transformed
ci_trans <- desc_calc(bc_data)$ci_mean
ci_orig <- inverse_transform(ci_trans, lambda = bc$lambda, method = "box_cox")
```

## 7. Quality Scenario: Plating Line Cycle Time

**Background.** An electroplating line produces gold-plated connectors. The
quality engineer needs to establish process capability for cycle time (the time
to complete one plating rack). Cycle time is critical because it determines
throughput and affects coating thickness uniformity. The spec requires cycle
time between 120 and 240 seconds.

The engineer collects 100 consecutive cycle-time measurements. Because cycle
time is bounded below by zero and right-skewed, it is expected to follow a
lognormal distribution. The goal is to:

1. Describe the data
2. Fit a distribution
3. Transform to normality for capability analysis

### Step 1: Describe

```{r}
set.seed(2026)
plating_cycle <- rlnorm(100, meanlog = log(175), sdlog = 0.22)

desc <- desc_calc(plating_cycle)
cat(sprintf("Mean      = %.2f s\n", desc$mean))
cat(sprintf("Median    = %.2f s\n", desc$median))
cat(sprintf("Std dev   = %.2f s\n", desc$stdev))
cat(sprintf("Skewness  = %.4f\n", desc$skew))
cat(sprintf("AD p-value= %.4f\n", desc$p_value))
cat(sprintf("95%% CI mean = [%.2f, %.2f]\n", desc$ci_mean[1], desc$ci_mean[2]))
```

The positive skewness and low AD p-value confirm the data is right-skewed and
non-normal.

### Step 2: Fit

```{r}
# Auto-fit multiple distributions
afit <- auto_fit_distribution(plating_cycle)
cat(sprintf("Best distribution: %s\n", afit$best_dist))

# Fit lognormal explicitly
fit_ln <- fit_distribution(plating_cycle, dist = "lnorm")
cat(sprintf("Lognormal: meanlog = %.4f, sdlog = %.4f\n",
            fit_ln$params$meanlog, fit_ln$params$sdlog))
cat(sprintf("KS p-value = %.4f\n", fit_ln$ks_test$p.value))
```

### Step 3: Transform

```{r}
bc <- box_cox_transform(plating_cycle)
cat(sprintf("Optimal lambda = %.4f (close to 0 → log transform)\n", bc$lambda))
cat(sprintf("Normality before: p = %.4f\n", bc$normality_before))
cat(sprintf("Normality after:  p = %.4f\n", bc$normality_after))

# Descriptive statistics on transformed data
desc_trans <- desc_calc(bc$transformed)
cat(sprintf("\nTransformed mean = %.4f\n", desc_trans$mean))
cat(sprintf("Transformed SD   = %.4f\n", desc_trans$stdev))
cat(sprintf("Transformed skew = %.4f (near zero ✓)\n", desc_trans$skew))
```

### Step 4: Visualize the Transformation

```{r, fig.width=8, fig.height=4}
# Before vs after transformation: side-by-side histograms
df_compare <- data.frame(
  value = c(plating_cycle, bc$transformed),
  group = rep(c("Original", "Transformed"), each = length(plating_cycle))
)

p <- base_plot(df_compare, aes(value, fill = group)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, alpha = 0.6,
                 position = "identity") +
  geom_density(alpha = 0.2, linewidth = 0.8) +
  facet_wrap(~ group, scales = "free") +
  labs(title = "Plating Cycle Time: Before vs After Box-Cox Transformation",
       x = "Value", y = "Density") +
  theme(legend.position = "none")
p
```

The transformed data is visibly more symmetric and bell-shaped, confirming that
the Box-Cox transformation has successfully normalized the data. The engineer
can now proceed with capability analysis on the transformed scale.

## 8. Parameter Reference

### 8.1 Descriptive Statistics Functions

| Function | Signature | Returns |
|----------|-----------|---------|
| `desc_calc()` | `desc_calc(x, conf = 0.95)` | List: n, mean, stdev, median, min, max, range, q1, q3, iqr, skew, kurt, cv, ad_stat, p_value, ci_mean, se |
| `desc_analyze()` | `desc_analyze(data, vars = NULL, conf = 0.95)` | List keyed by variable name |
| `iqr_desc()` | `iqr_desc(data, vars = NULL, conf = 0.95, plot = FALSE, report = FALSE)` | List or report |
| `desc_hist()` | `desc_hist(x, ...)` | ggplot object |
| `desc_box()` | `desc_box(data, vars = NULL, ...)` | ggplot object |

### 8.2 Probability Distribution Functions

| Function | Signature | Returns |
|----------|-----------|---------|
| `iqr_prob$new()` | — | R6 object |
| `$add_dist()` | `add_dist(type, params, loc = 0, reset = FALSE)` | self |
| `$calc()` | `calc(values, mode = "prob", calc_type = "lower")` | List with `all_res` |
| `prob_calc()` | `prob_calc(type, x, params, mode = "prob")` | List with `all_res` |
| `list_prob_distributions()` | — | Character vector |

### 8.3 Distribution Fitting Functions

| Function | Signature | Returns |
|----------|-----------|---------|
| `fit_distribution()` | `fit_distribution(x, dist)` | List: dist, params, logLik, AIC, BIC, ks_test, n, data |
| `auto_fit_distribution()` | `auto_fit_distribution(x)` | List: best_dist, best_result, all_results, ranking, criterion |
| `compare_fits()` | `compare_fits(x, distributions)` | Comparison list |
| `calc_qq_data()` | `calc_qq_data(x, dist, params)` | Data frame of Q-Q coordinates |

### 8.4 Data Transformation Functions

| Function | Signature | Returns |
|----------|-----------|---------|
| `box_cox_transform()` | `box_cox_transform(x, lambda = NULL)` | List: transformed, lambda, original, method, normality_before, normality_after |
| `yeo_johnson_transform()` | `yeo_johnson_transform(x, lambda = NULL)` | Similar structure |
| `johnson_transform()` | `johnson_transform(x)` | Similar structure |
| `auto_transform()` | `auto_transform(x)` | List: best_method, best_result, all_results, scores, criterion |
| `inverse_transform()` | `inverse_transform(x, lambda, method)` | Numeric vector |

## 9. References

1. Montgomery, D.C. (2013). *Introduction to Statistical Quality Control*, 7th
   Edition. John Wiley & Sons. — Turning operation and dimensional measurement
   case studies (Ch. 5–6).

2. IES TM-21-11 (2011). *Projecting Long Term Lumen Maintenance of LED Light
   Sources*. Illuminating Engineering Society. — Basis for the LED Weibull
   lifetime model parameters.

3. John, P.W.M. (1971). *Statistical Methods in Engineering and Quality
   Assurance*. John Wiley & Sons. — Lognormal distribution in cycle-time
   analysis.

4. Box, G.E.P. & Cox, D.R. (1964). "An Analysis of Transformations."
   *Journal of the Royal Statistical Society, Series B*, 26(2), 211–252. —
   The original Box-Cox transformation paper.

5. Yeo, I.K. & Johnson, R.A. (2000). "A New Family of Power Transformations to
   Improve Normality or Symmetry." *Biometrika*, 87(4), 954–959.

6. Johnson, N.L., Kotz, S. & Balakrishnan, N. (1994). *Continuous Univariate
   Distributions*, Vol. 1, 2nd Edition. John Wiley & Sons.

## Further Reading

- For normality testing (the formal next step after descriptive screening),
  see `vignette("hypothesis-testing-and-intervals", package = "iQualityR.stat")`.
- For capability analysis on transformed data, see
  `vignette("process-control-and-capability", package = "iQualityR.stat")`.
- For the visualization infrastructure used by `desc_hist()` and `desc_box()`,
  see `vignette("plotting-overview", package = "iQualityR.plot")`.
