---
title: "Hypothesis Testing and Normality Testing"
author: "iQualityR Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hypothesis Testing and Normality Testing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r include=FALSE}
library(iQualityR.stat)
library(R6)

library(patchwork)
library(iQualityR.core)
library(iQualityR.plot)

```

## Overview

Hypothesis testing is a core tool in quality management for determining whether differences are statistically significant. Normality testing is a prerequisite for many statistical methods. This module provides complete calculation, visualization, interpretation, and reporting functions.

## Hypothesis Testing

### Quick Functions

```{r setup}
library(iQualityR.stat)
```

#### One-Sample t Test

```{r one_sample}
set.seed(123)
x <- rnorm(30, mean = 5.2, sd = 1)

result <- htest_run(test_type = "t_test_1s", x = x, mu = 5)
result$p.value
```

#### Two-Sample t Test

```{r two_sample}
set.seed(123)
x <- rnorm(30, mean = 5, sd = 1)
y <- rnorm(30, mean = 5.5, sd = 1)

result <- htest_run(test_type = "t_test_2s", x = x, y = y)
result$p.value
```

#### Paired t Test

```{r paired}
set.seed(123)
before <- rnorm(20, mean = 100, sd = 10)
after <- before + rnorm(20, mean = 3, sd = 5)

result <- htest_run(test_type = "t_test_paired", x = before, y = after)
result$p.value
```

#### Variance Test

```{r var_test}
set.seed(123)
x <- rnorm(30, mean = 5, sd = 1)
y <- rnorm(30, mean = 5, sd = 1.5)

result <- htest_run(test_type = "f_test", x = x, y = y)
result$p.value
```

## Normality Testing

### Supported Test Methods

| Method | Description | Applicable Sample Size |
|--------|-------------|------------------------|
| `sw` | Shapiro-Wilk | 3-5000 |
| `ad` | Anderson-Darling | 8+ |
| `lillie` | Lilliefors (Kolmogorov-Smirnov) | 4+ |
| `sf` | Shapiro-Francia | 5-5000 |
| `cvm` | Cramer-von Mises | 8+ |

### Quick Functions

```{r normality_test}
set.seed(123)
x <- rnorm(50)

result <- normality_test(x, method = "sw")
result$p.value
result$is_normal
```

### Auto-Select Method

```{r auto_method}
set.seed(123)
x <- rnorm(100)

result <- normality_test(x, method = "auto")
result$method
```

## Unified Interpreter

`StatInterpreter` provides unified statistical result interpretation:

```{r stat_interpreter}
interpreter <- StatInterpreter$new()

interpreter$interpret(result, audience = "manager")
```
