---
title: "Hypothesis Testing for Quality Decisions"
author: "iQualityR.stat Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Hypothesis Testing for Quality Decisions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

library(iQualityR.stat)
library(ggplot2)

# Helpers
fmt_num <- function(x, digits = 4) formatC(x, format = "f", digits = digits)
fmt_pct <- function(x, digits = 2) paste0(formatC(x, format = "f", digits = digits), "%")
```

## 1. Introduction

Hypothesis testing is the decision backbone of quality engineering. From
verifying that a process mean has not drifted, to comparing two suppliers, to
checking whether a defect rate exceeds a threshold, every quality decision
eventually reduces to a formal test of hypotheses.

This vignette covers **all 8 parametric test types** supported by `iqr_htest`:

| # | Test code | Family | Decision question |
|---|-----------|--------|-------------------|
| 1 | `t_test_1s`     | Mean, one sample    | Is the process mean on target? |
| 2 | `z_test_1s`     | Mean, one sample    | Is the mean shifted (sigma known)? |
| 3 | `t_test_2s`     | Mean, two samples   | Do two processes share a mean? |
| 4 | `t_test_paired` | Mean, paired        | Did a treatment change the mean? |
| 5 | `prop_test_1s`  | Proportion, one     | Is the defect rate below spec? |
| 6 | `prop_test_2s`  | Proportion, two     | Did a change reduce defect rate? |
| 7 | `f_test`        | Variance            | Do two processes share variance? |
| 8 | `chisq_test`    | Distribution        | Does the output match a target distribution? |

### 1.1 The iQualityR methodology

Every analysis in iQualityR follows the unified workflow
**Plan -> Task -> Analysis -> Summary/Plot -> Report** (see the meta-package
vignette `vignette("iQualityR", package = "iQualityR")` for the full
framework). For hypothesis testing:

- **Plan**: capture 4M1E context (Man, Machine, Material, Method, Environment)
- **Task**: create `iqr_htest$new(theme = ...)`
- **Analysis**: call `$run(test_type, ...)` to obtain a `stat_result`
- **Summary/Plot**: call `$plot()` (delegates to `iQualityR.plot`) and `$interpret()`
- **Report**: call `$report(format = "excel")` to export

### 1.2 Who should read this vignette

Quality engineers, SPC practitioners, validation engineers, and anyone who
needs to make data-driven decisions about process means, proportions,
variances, or distributions.

## 2. Theory: The Logic of Hypothesis Testing

### 2.1 Null and alternative hypotheses

A hypothesis test pits two statements against each other:

- **Null hypothesis H0**: the "no change" / "no difference" baseline, e.g.
  `H0: mu = mu0` or `H0: p1 = p2`.
- **Alternative hypothesis Ha**: the claim the engineer wants to detect, e.g.
  `Ha: mu != mu0` (two-sided), `Ha: mu > mu0` (greater), or `Ha: mu < mu0` (less).

The test assumes H0 is true and asks: "How surprising is the observed data
under H0?" If the data are sufficiently surprising, we reject H0 in favour of
Ha.

### 2.2 Type I and Type II errors, significance level, power

| Decision reality | H0 true | H0 false |
|---|---|---|
| Fail to reject H0 | Correct | Type II error (beta) |
| Reject H0 | Type I error (alpha) | Correct (power = 1 - beta) |

- **Significance level alpha** = P(Type I error) = P(reject H0 | H0 true).
  Typical values: 0.05, 0.01, 0.10. Set by the engineer before collecting data.
- **Power 1 - beta** = P(correctly reject H0 | H0 false). Depends on effect
  size, sample size, and alpha. Use `sample_size_*` functions to design for
  adequate power (see `vignette("sample-size-and-power", package = "iQualityR.stat")`).

### 2.3 Test statistics and sampling distributions

Each test reduces the data to a single number, the **test statistic**, whose
distribution under H0 is known:

| Test | Statistic | Null distribution |
|------|-----------|-------------------|
| Z-test (mean) | `Z = (x_bar - mu0) / (sigma / sqrt(n))` | Standard normal N(0,1) |
| t-test (mean) | `t = (x_bar - mu0) / (s / sqrt(n))` | Student's t with `df = n - 1` |
| Two-sample t  | `t = (m1 - m2) / SE` | t with Welch or pooled df |
| Proportion Z  | `Z = (p_hat - p0) / sqrt(p0(1-p0)/n)` | Standard normal |
| F-test        | `F = s1^2 / s2^2` | F with `(n1-1, n2-1)` df |
| Chi-square    | `X^2 = sum((O - E)^2 / E)` | Chi-square with `k - 1` or `(r-1)(c-1)` df |

### 2.4 p-value interpretation

The **p-value** is the probability of observing a test statistic at least as
extreme as the one computed, assuming H0 is true:

- `p < alpha` -> reject H0; data are inconsistent with H0.
- `p >= alpha` -> fail to reject H0; data are consistent with H0.

**Caveats**:

- A small p-value does **not** prove Ha is true; it only rules out H0 as a
  plausible explanation for the data.
- A large p-value does **not** prove H0 is true; the test may simply be
  underpowered.
- Statistical significance is not practical significance; always report effect
  size and confidence interval alongside p-value.

### 2.5 One-sided vs two-sided alternatives

The choice of alternative determines the rejection region and the p-value
calculation:

- **Two-sided** (`alternative = "two.sided"`): Ha contains `!=`. Rejection
  region split across both tails. `p = 2 * P(|T| >= |stat|)`.
- **Greater** (`alternative = "greater"`): Ha contains `>`. Rejection region
  in the right tail. `p = P(T >= stat)`.
- **Less** (`alternative = "less"`): Ha contains `<`. Rejection region in the
  left tail. `p = P(T <= stat)`.

Choose the alternative **before** collecting data, based on the engineering
question. Switching alternatives after seeing the data inflates Type I error.

## 3. Test Types and Their Statistics

### 3.1 Z-test and t-test (one-sample, two-sample)

**One-sample Z-test** (population sigma known):

`Z = (x_bar - mu0) / (sigma / sqrt(n))`

Use when sigma is known from long-term capability data. Critical value is the
fixed `qnorm(1 - alpha/2)` (does not depend on the sample).

**One-sample t-test** (sigma unknown):

`t = (x_bar - mu0) / (s / sqrt(n))`, with `df = n - 1`.

Use when sigma must be estimated from the sample. Critical value
`qt(1 - alpha/2, df)` depends on n; t converges to Z as n grows.

**Two-sample t-test** (independent samples):

`t = (m1 - m2) / SE`, where the standard error depends on whether variances
are pooled:

- Pooled (equal variances): `SE = sqrt(((n1-1)s1^2 + (n2-1)s2^2) / (n1+n2-2)) * sqrt(1/n1 + 1/n2)`,
  `df = n1 + n2 - 2`.
- Welch (unequal variances, default): `SE = sqrt(s1^2/n1 + s2^2/n2)`,
  `df = (s1^2/n1 + s2^2/n2)^2 / ((s1^2/n1)^2/(n1-1) + (s2^2/n2)^2/(n2-1))`.

Confidence interval is on the **mean difference** `m1 - m2`.

**Paired t-test**: reduce to one-sample t on the differences
`d_i = x_i - y_i`, with `t = d_bar / (s_d / sqrt(n))`, `df = n - 1`.

### 3.2 Proportion tests (one-sample, two-sample)

**One-sample proportion Z-test**:

`Z = (p_hat - p0) / sqrt(p0(1-p0)/n)`, where `p_hat = x/n`.

Use the null standard error `sqrt(p0(1-p0)/n)` for the test statistic; use the
estimated standard error `sqrt(p_hat(1-p_hat)/n)` for the confidence interval.

**Two-sample proportion Z-test**:

`Z = (p1 - p2) / sqrt(p_pool(1-p_pool)(1/n1 + 1/n2))`, where
`p_pool = (x1 + x2) / (n1 + n2)`.

Confidence interval for `p1 - p2` uses the unpooled standard error
`sqrt(p1(1-p1)/n1 + p2(1-p2)/n2)`.

### 3.3 Chi-square test

**Goodness-of-fit**: `X^2 = sum((O_i - E_i)^2 / E_i)`, with
`E_i = n * p_i` and `df = k - 1` (k = number of categories).

**Contingency table** (test of independence): same formula with
`E_{ij} = (row_total * col_total) / grand_total` and
`df = (r - 1)(c - 1)`.

Pearson residuals `(O - E) / sqrt(E)` identify which cells drive significance.

### 3.4 Variance test (F-test)

`F = s1^2 / s2^2`, with `df1 = n1 - 1`, `df2 = n2 - 1`.

- Two-sided: reject if `F < qf(alpha/2, df1, df2)` or `F > qf(1-alpha/2, df1, df2)`.
- Greater: reject if `F > qf(1-alpha, df1, df2)`.
- Less: reject if `F < qf(alpha, df1, df2)`.

**Caution**: the F-test is sensitive to non-normality. For non-normal data,
prefer Levene's or Bartlett's test (planned for a future release).

## 4. The iqr_htest Class

`iqr_htest` is the R6 integrator that orchestrates L1 (Analyzer) + L2
(Plotter/Reporter/Interpreter). It is the recommended entry point for
interactive use.

### 4.1 Lifecycle

```{r iqr-htest-lifecycle, eval = FALSE}
library(iQualityR.stat)

htest <- iqr_htest$new(theme = "academic")        # 1. create
htest$run("t_test_1s", x = my_data, mu = 25)      # 2. analyze -> stat_result
htest$plot(plot_type = "combined")                # 3. visualize -> ggplot/patchwork
htest$interpret(audience = "manager")             # 4. interpret -> text
htest$report(format = "excel", path = "out.xlsx") # 5. export
htest$summary()                                   # 6. console summary
```

### 4.2 Key fields

| Field | Type | Description |
|-------|------|-------------|
| `last_results` | `stat_result` S3 | Cached result of the last `$run()` |
| `theme` | character | Active theme name |
| `analyzer` | `HTestAnalyzer` | L1 engine (R6) |
| `plotter` | `HTestPlotter` | L2 plotter (R6) |
| `reporter` | `HTestReporter` | L2 reporter (R6) |

### 4.3 Supported test_type codes

`"z_test_1s"`, `"t_test_1s"`, `"t_test_2s"`, `"t_test_paired"`,
`"prop_test_1s"`, `"prop_test_2s"`, `"f_test"`, `"chisq_test"`.

## 5. Functional API: htest_run(), htest_plot(), htest_interpret()

For scripting / pipelines, prefer the stateless convenience functions:

```{r functional-api, eval = FALSE}
# Single-shot: run + return stat_result
result <- htest_run("t_test_1s", x = my_data, mu = 25, conf_level = 0.95)

# Plot from a stat_result
p <- htest_plot(result, plot_type = "combined")

# Interpret for different audiences
htest_interpret(result, audience = "manager")    # business impact
htest_interpret(result, audience = "technical")  # statistical terms
htest_interpret(result, audience = "client")     # audit-ready

# Export
htest_report(result, format = "excel", path = "out.xlsx")
```

The functional API re-runs the Analyzer + Plotter on every call; use the R6
class when you want to cache `last_results` for repeated `$plot()` /
`$interpret()` calls.

## 6. Workflow: test type -> data -> parameters

The workflow is **always**:

1. **Choose test type** based on the engineering question and data shape.
2. **Provide data** in the format the test type expects.
3. **Set parameters** including the alternative (one-sided vs two-sided) whose
   semantics vary by test.

### 6.1 Data input by test type

| Test type | Data format |
|-----------|-------------|
| `t_test_1s`, `z_test_1s` | Single numeric vector `x` (or list `list(mean=, n=, sd=)` for summary-only Z) |
| `t_test_2s`, `f_test` | Two numeric vectors `x`, `y` |
| `t_test_paired` | Two equal-length numeric vectors `x`, `y` (paired) |
| `prop_test_1s` | `x` = successes (scalar), `n` = trials (scalar) |
| `prop_test_2s` | `x1, n1, x2, n2` (four scalars) |
| `chisq_test` | Vector of observed frequencies, or matrix/table for contingency; optional `p` for expected proportions |

### 6.2 Alternative semantics by test type

The `alternative` argument is universal, but the symbols in Ha differ:

| Test type | two.sided | greater | less |
|-----------|-----------|---------|------|
| `t_test_1s`, `z_test_1s` | `mu != mu0` | `mu > mu0` | `mu < mu0` |
| `t_test_2s` | `mu1 != mu2` | `mu1 > mu2` | `mu1 < mu2` |
| `t_test_paired` | `mu_d != 0` | `mu_d > 0` | `mu_d < 0` |
| `prop_test_1s` | `p != p0` | `p > p0` | `p < p0` |
| `prop_test_2s` | `p1 != p2` | `p1 > p2` | `p1 < p2` |
| `f_test` | `sigma1 != sigma2` | `sigma1 > sigma2` | `sigma1 < sigma2` |
| `chisq_test` | (always two-sided) | -- | -- |

### 6.3 Test-specific parameters

| Test type | Required | Optional |
|-----------|----------|----------|
| `t_test_1s` | `x`, `mu` | `alternative`, `conf_level` |
| `z_test_1s` | `x`, `mu`, `sigma` | `alternative`, `conf_level` |
| `t_test_2s` | `x`, `y` | `alternative`, `conf_level`, `var.equal` |
| `t_test_paired` | `x`, `y` (equal length) | `alternative`, `conf_level` |
| `prop_test_1s` | `x`, `n`, `p0` | `alternative`, `conf_level` |
| `prop_test_2s` | `x1`, `n1`, `x2`, `n2` | `alternative`, `conf_level` |
| `f_test` | `x`, `y` | `alternative`, `conf_level` |
| `chisq_test` | `x` | `p` (expected proportions; default uniform) |

## 7. Quality Scenario: 8 Cross-Industry Cases

Each case follows the engineering workflow
**Problem -> Data -> Analysis -> Visualization -> Confirmation**.

### 7.1 Case 1: Automotive Machining -- One-Sample t Test (`t_test_1s`)

**Problem.** A CNC turning line machines transmission shafts with nominal
diameter 25.000 mm (tolerance 24.97 / 25.03). The quality engineer pulls a
random sample of n = 30 parts to verify the process mean has not drifted.

```{r case1-data}
shaft_diameter <- c(
  25.008, 25.012, 24.995, 25.021, 25.003, 24.998, 25.015, 25.006,
  25.010, 24.992, 25.018, 25.004, 25.007, 24.999, 25.013, 25.005,
  25.020, 24.996, 25.011, 25.002, 25.016, 24.994, 25.009, 25.001,
  25.014, 25.019, 24.997, 25.006, 25.022, 25.000
)
cat(sprintf("n = %d, mean = %.4f mm, sd = %.4f mm\n",
            length(shaft_diameter), mean(shaft_diameter), sd(shaft_diameter)))
```

```{r case1-run}
result <- htest_run("t_test_1s", x = shaft_diameter, mu = 25.000, conf_level = 0.95)
cat(sprintf("t = %.4f, df = %.0f, p-value = %.4f\n",
            result$statistic, result$parameter, result$p.value))
cat(sprintf("95%% CI: [%.4f, %.4f] mm\n", result$conf.int[1], result$conf.int[2]))
cat(sprintf("se = %.4f, critical t = %.4f\n", result$se, result$critical_value))
```

```{r case1-plot}
htest_plot(result, plot_type = "combined")
```

**Confirmation.** If the 95% CI excludes 25.000 mm, investigate tool offset;
otherwise continue production. Increase to n = 60 if marginal.

### 7.2 Case 2: Steel Manufacturing -- One-Sample Z Test (`z_test_1s`)

**Problem.** A steel mill produces rebar with specified tensile strength
mu = 420 MPa, sigma = 15 MPa (known from years of capability data). A
metallurgist tests whether a new heat-treatment schedule has shifted the mean
strength, using n = 36 specimens.

```{r case2-data}
set.seed(42)
tensile_strength <- rnorm(36, mean = 424, sd = 15)
cat(sprintf("n = %d, mean = %.2f MPa, known sigma = 15 MPa\n",
            length(tensile_strength), mean(tensile_strength)))
```

```{r case2-run}
result <- htest_run("z_test_1s", x = tensile_strength,
                    mu = 420, sigma = 15, conf_level = 0.95)
cat(sprintf("Z = %.4f, p-value = %.4f\n", result$statistic, result$p.value))
cat(sprintf("se = %.4f, critical Z = %.4f\n", result$se, result$critical_value))
cat(sprintf("95%% CI: [%.2f, %.2f] MPa\n", result$conf.int[1], result$conf.int[2]))
```

```{r case2-plot}
htest_plot(result, plot_type = "combined")
```

**Confirmation.** The Z-test uses known population sigma, so the critical
value is the fixed Z = 1.96. If p < 0.05, investigate heat-treatment
parameters (temperature, dwell time).

### 7.3 Case 3: Pharmaceutical -- Two-Sample t Test (`t_test_2s`)

**Problem.** A tablet compression room has two presses (A and B) producing
500 mg acetaminophen tablets. After PM on Press B, QA confirms the mean
tablet weight has not shifted relative to reference Press A.

```{r case3-data}
press_a <- c(
  502, 499, 501, 500, 498, 503, 501, 499, 502, 500,
  501, 498, 502, 500, 503, 499, 501, 500, 502, 499,
  501, 500, 502, 498, 501
)
press_b <- c(
  500, 503, 502, 501, 500, 504, 502, 501, 503, 500,
  502, 504, 501, 502, 503, 501, 500, 502, 504, 501,
  503, 502, 501, 500, 502
)
cat(sprintf("Press A: mean = %.2f mg, sd = %.2f mg\n", mean(press_a), sd(press_a)))
cat(sprintf("Press B: mean = %.2f mg, sd = %.2f mg\n", mean(press_b), sd(press_b)))
```

```{r case3-run}
result <- htest_run("t_test_2s", x = press_a, y = press_b, conf_level = 0.95)
cat(sprintf("t = %.4f, df = %.1f, p-value = %.4f\n",
            result$statistic, result$parameter, result$p.value))
cat(sprintf("mean A = %.2f, mean B = %.2f, diff = %.2f mg\n",
            result$mean1, result$mean2, result$diff_mean))
cat(sprintf("se = %.4f, 95%% CI for diff: [%.2f, %.2f] mg\n",
            result$se, result$conf.int[1], result$conf.int[2]))
```

```{r case3-plot}
htest_plot(result, plot_type = "combined")
```

**Confirmation.** Per ICH Q2(R2), a significant shift triggers 100%
tablet-weight verification on Press B. Recalibrate die-fill cam if verified
mean difference exceeds 1.0 mg.

### 7.4 Case 4: Semiconductor -- Paired t Test (`t_test_paired`)

**Problem.** A fab measures wafer thickness before and after a wet-etch
process on n = 20 wafers. The process engineer wants to confirm the etch
removes a consistent amount of material (target removal = 0.50 um).

```{r case4-data}
before <- c(725.2, 725.5, 724.8, 725.1, 725.3, 724.9, 725.0, 725.4,
            725.1, 724.7, 725.3, 725.2, 724.9, 725.0, 725.5, 725.1,
            724.8, 725.2, 725.3, 725.0)
after  <- c(724.6, 725.0, 724.3, 724.6, 724.8, 724.4, 724.5, 724.9,
            724.6, 724.2, 724.8, 724.7, 724.4, 724.5, 725.0, 724.6,
            724.3, 724.7, 724.8, 724.5)
cat(sprintf("n = %d pairs, mean removal = %.4f um, sd of diffs = %.4f um\n",
            length(before), mean(before - after), sd(before - after)))
```

```{r case4-run}
result <- htest_run("t_test_paired", x = before, y = after,
                    alternative = "greater", conf_level = 0.95)
cat(sprintf("t = %.4f, df = %.0f, p-value = %.4f\n",
            result$statistic, result$parameter, result$p.value))
cat(sprintf("mean diff = %.4f um, sd diff = %.4f um, se = %.4f\n",
            result$mean_diff, result$sd_diff, result$se))
cat(sprintf("95%% CI for mean diff: [%.4f, Inf)\n", result$conf.int[1]))
```

```{r case4-plot}
htest_plot(result, plot_type = "combined")
```

**Confirmation.** The one-sided test confirms etch removal is significantly
greater than zero. Verify etch uniformity across the wafer via 9-point
thickness mapping.

### 7.5 Case 5: Textile -- One-Sample Proportion Test (`prop_test_1s`)

**Problem.** A textile mill's internal spec requires defect rate below 3% for
premium grade fabric. QA inspected n = 800 meters and found x = 35 defective
meters. Test whether the defect rate exceeds the 3% threshold.

```{r case5-data}
x_defects <- 35
n_inspected <- 800
p0 <- 0.03
cat(sprintf("Defects: %d / %d = %.2f%% (spec limit: %.1f%%)\n",
            x_defects, n_inspected, 100 * x_defects / n_inspected, 100 * p0))
```

```{r case5-run}
result <- htest_run("prop_test_1s", x = x_defects, n = n_inspected,
                    p0 = p0, alternative = "greater", conf_level = 0.95)
cat(sprintf("Z = %.4f, p-value = %.4f\n", result$statistic, result$p.value))
cat(sprintf("p_hat = %.4f, p0 = %.4f, se = %.4f\n",
            result$p_hat, result$p0, result$se))
cat(sprintf("95%% CI: [%.4f, %.4f]\n", result$conf.int[1], result$conf.int[2]))
```

```{r case5-plot}
htest_plot(result, plot_type = "curve")
```

**Confirmation.** If p < 0.05, the defect rate exceeds 3% and the lot must be
downgraded to standard grade. Investigate loom tension and yarn quality.

### 7.6 Case 6: Electronics -- Two-Sample Proportion Test (`prop_test_2s`)

**Problem.** An SMT line tracked solder-bridge defect rate after switching
from Flux X to Flux Y. The engineer tests whether Flux Y reduced the defect
proportion.

```{r case6-data}
flux_x_defects <- 48
flux_x_total   <- 2400
flux_y_defects <- 27
flux_y_total   <- 2400
cat(sprintf("Flux X: %d/%d = %.2f%%\n",
            flux_x_defects, flux_x_total, 100 * flux_x_defects / flux_x_total))
cat(sprintf("Flux Y: %d/%d = %.2f%%\n",
            flux_y_defects, flux_y_total, 100 * flux_y_defects / flux_y_total))
```

```{r case6-run}
result <- htest_run("prop_test_2s",
                    x1 = flux_y_defects, n1 = flux_y_total,
                    x2 = flux_x_defects, n2 = flux_x_total,
                    alternative = "less", conf_level = 0.95)
cat(sprintf("Z = %.4f, p-value = %.4f\n", result$statistic, result$p.value))
cat(sprintf("p1 = %.4f, p2 = %.4f, diff = %.4f\n",
            result$p1, result$p2, result$diff))
cat(sprintf("95%% CI for p_Y - p_X: [%.4f, %.4f]\n",
            result$conf.int[1], result$conf.int[2]))
```

```{r case6-plot}
htest_plot(result, plot_type = "curve")
```

**Confirmation.** Roll out Flux Y to all SMT lanes and monitor via p-chart
for 4 production days. Lock in the improvement only if the run-chart stays
below the Flux X baseline.

### 7.7 Case 7: Chemical -- F Test for Variance Equality (`f_test`)

**Problem.** A chemical plant operates two reactors (R1 and R2) producing
the same polymer. The process engineer verifies that reactor-to-reactor
variance is equivalent before pooling their output into a single storage tank.

```{r case7-data}
reactor_1 <- c(
  98.2, 101.5, 99.8, 100.3, 97.9, 102.1, 99.5, 100.8,
  98.7, 101.2, 99.9, 100.5, 98.3, 101.8, 99.6, 100.1,
  98.9, 101.0, 99.7, 100.4
)
reactor_2 <- c(
  99.1, 100.5, 99.8, 100.2, 99.5, 100.8, 99.3, 100.6,
  99.0, 100.3, 99.6, 100.1, 99.2, 100.7, 99.4, 100.0,
  99.3, 100.4, 99.7, 100.2, 99.1, 100.5, 99.6, 100.3
)
cat(sprintf("R1: var = %.4f (n = %d)\n", var(reactor_1), length(reactor_1)))
cat(sprintf("R2: var = %.4f (n = %d)\n", var(reactor_2), length(reactor_2)))
```

```{r case7-run}
result <- htest_run("f_test", x = reactor_1, y = reactor_2, conf_level = 0.95)
cat(sprintf("F = %.4f, df1 = %d, df2 = %d, p-value = %.4f\n",
            result$statistic, result$parameter[1], result$parameter[2],
            result$p.value))
cat(sprintf("var1 = %.4f, var2 = %.4f, ratio = %.4f\n",
            result$var1, result$var2, result$var1 / result$var2))
```

```{r case7-plot}
htest_plot(result, plot_type = "combined")
```

**Confirmation.** If p > 0.05, the variances are equivalent and the reactors'
output may be pooled. Otherwise, investigate reactor temperature control
stability before pooling.

### 7.8 Case 8: Food Packaging -- Chi-Square Goodness-of-Fit (`chisq_test`)

**Problem.** A 250 g coffee pouch line is designed so fill weights follow a
uniform distribution across five 2-gram bands. QA collected 500 consecutive
pouches and binned them to test whether the filler head distributes product
uniformly.

```{r case8-data}
observed_counts <- c(112, 96, 88, 110, 94)
expected_probs  <- rep(0.2, 5)
cat("Observed:", observed_counts, "(total =", sum(observed_counts), ")\n")
cat("Expected proportion:", expected_probs, "\n")
```

```{r case8-run}
result <- htest_run("chisq_test", x = observed_counts, p = expected_probs)
cat(sprintf("X-squared = %.4f, df = %.0f, p-value = %.4f\n",
            result$statistic, result$parameter, result$p.value))
cat("\nObserved vs Expected:\n")
print(data.frame(
  Band     = paste0("B", seq_along(observed_counts)),
  Observed = result$observed,
  Expected = round(result$expected, 1),
  Residual = round(result$residuals, 2)
))
```

```{r case8-plot}
htest_plot(result, plot_type = "curve")
```

**Confirmation.** p > 0.05 means the fill-weight distribution is consistent
with the uniform target. Document in the monthly HACCP review; re-check at
next filler-head PM.

## 8. Interpreting Test Results -- Decision Guide

### 8.1 The four-field decision matrix

| p-value | CI excludes H0? | Decision | Action |
|---------|-----------------|----------|--------|
| < alpha | Yes | Reject H0 | Implement change; investigate root cause |
| < alpha | No (boundary) | Reject H0 (marginal) | Increase sample size; re-test |
| >= alpha | No | Fail to reject H0 | Continue monitoring; do NOT claim "H0 true" |
| >= alpha | Yes (rare) | Contradiction | Check data quality; consult statistician |

### 8.2 Audience-specific interpretation

`htest_interpret(result, audience = ...)` returns tailored language:

- `"manager"`: business impact, dollar risk, go/no-go recommendation.
- `"technical"`: statistical terms, effect size, CI, assumptions.
- `"client"`: audit-ready phrasing, citation of standard (ICH, ASTM, ISO).

### 8.3 Common pitfalls

- **P-hacking**: do not switch alternatives or drop outliers after seeing p.
- **Significance != importance**: a 0.01 mg shift can be statistically
  significant but practically irrelevant; always pair p with effect size and CI.
- **Underpowered test**: failing to reject H0 with n = 5 is not evidence for H0.
- **F-test on non-normal data**: the F-test is sensitive to non-normality;
  verify normality first (see
  `vignette("probability-distributions", package = "iQualityR.stat")`).

## 9. Parameter Reference

### 9.1 Analyzer parameters by test type

Each test returns a `stat_result` S3 object (class
`c("stat_result", "htest_result")`).

| Test type | Statistic | df | SE | Critical value | CI for | Extra fields |
|-----------|-----------|----|----|----------------|--------|--------------|
| `t_test_1s` | t | n-1 | s/sqrt(n) | qt(alpha/2, df) | mean | n, mean, sd |
| `z_test_1s` | Z | -- | sigma/sqrt(n) | qnorm(alpha/2) | mean | n, mean, sd, sigma |
| `t_test_2s` | t | Welch or n1+n2-2 | sqrt(s1^2/n1 + s2^2/n2) | qt(alpha/2, df) | diff_mean | mean1/2, sd1/2, diff_mean |
| `t_test_paired` | t | n-1 | s_d/sqrt(n) | qt(alpha/2, df) | mean_diff | mean_diff, sd_diff |
| `prop_test_1s` | Z | -- | sqrt(p0(1-p0)/n) | qnorm(alpha/2) | p_hat | p_hat, p0, x_success |
| `prop_test_2s` | Z | -- | sqrt(p_pool(1-p_pool)(1/n1+1/n2)) | qnorm(alpha/2) | diff | p1, p2, p_pool, diff |
| `f_test` | F | (n1-1, n2-1) | -- | qf(alpha/2, df1, df2) | -- | var1, var2, critical_value_low/upp |
| `chisq_test` | X^2 | k-1 or (r-1)(c-1) | -- | -- | -- | observed, expected, residuals |

### 9.2 Visualization dispatch by test type

| Test type | auto plot | curve method | box method | .plot function called |
|-----------|-----------|-------------|------------|-----------------------|
| `t_test_1s` | combined | t-curve | single box | `plot_hypothesis_curve` + `plot_hypothesis_box` |
| `z_test_1s` | combined (raw) / curve (summary) | Z-curve | single box | `plot_hypothesis_curve` + `plot_hypothesis_box` |
| `t_test_2s` | combined | t-curve (diff) | two-group box | `plot_hypothesis_curve` + `plot_hypothesis_box_two_group` |
| `t_test_paired` | combined | t-curve (diff) | paired plot | `plot_hypothesis_curve` + `plot_paired_before_after` |
| `prop_test_1s` | curve | Z-curve | -- | `plot_hypothesis_curve` |
| `prop_test_2s` | curve | Z-curve | -- | `plot_hypothesis_curve` |
| `f_test` | combined | F-curve | two-group box | `plot_hypothesis_curve_f` + `plot_hypothesis_box_two_group` |
| `chisq_test` | curve | obs-vs-exp bar | -- | `plot_chi_square_observed_expected` |

Per the Plotter coding standard (see `docs/plotter_coding_standards.md`),
`HTestPlotter` is a thin delegation layer -- all visualization logic lives in
`iQualityR.plot`.

### 9.3 Common stat_result fields

Every test returns these common fields: `test_type`, `method`, `statistic`,
`p.value`, `conf.int`, `alternative`, `conf.level`, `dist_type`, `data`.

Test-specific fields are listed in Section 9.1's "Extra fields" column.

## 10. References

1. Montgomery, D. C. (2020). *Introduction to Statistical Quality Control*,
   8th ed. Wiley.
2. NIST/SEMATECH (2012). *e-Handbook of Statistical Methods*,
   https://www.itl.nist.gov/div898/handbook/
3. ICH Q2(R2) (2023). *Validation of Analytical Procedures*.
4. Tukey, J. W. (1977). *Exploratory Data Analysis*. Addison-Wesley.
5. Welch, B. L. (1947). The generalization of "Student's" problem when
   several different population variances are involved. *Biometrika*, 34(1-2),
   28-35.
6. Pearson, K. (1900). On the criterion that a given system of deviations
   from the probable ... *Phil. Mag.*, Series 5, 50(302), 157-175.

## Further Reading

- For the iQualityR unified workflow:
  `vignette("iQualityR", package = "iQualityR")`
- For SPC constants and sigma estimation:
  `vignette("sigma-estimation", package = "iQualityR.stat")`
- For sample size and power design:
  `vignette("sample-size-and-power", package = "iQualityR.stat")`
- For distribution fitting and normality assessment:
  `vignette("probability-distributions", package = "iQualityR.stat")`
- For hypothesis-test visualization internals:
  `vignette("hypothesis-plots", package = "iQualityR.plot")`
- For the Plotter coding standard:
  `docs/plotter_coding_standards.md` in the source tree

## Session Information

```{r session-info}
sessionInfo()
```
