---
title: "Model Diagnostics and Outlier Detection"
author: "iQualityR Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Diagnostics and Outlier Detection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

Regression model diagnostics ensure model assumptions are met, and outlier detection identifies abnormal observations in data. Both are fundamental tools for data analysis in quality management.

## Model Diagnostics

### Complete Diagnostics

```{r setup}
library(iQualityR.stat)

set.seed(123)
x <- rnorm(100)
y <- 2 * x + rnorm(100, sd = 0.5)
model <- lm(y ~ x)

result <- diagnose_lm(model)
names(result)
```

### Residual Statistics

```{r residuals}
result$residuals
```

### Normality Test

```{r normality}
result$normality
```

### Heteroscedasticity Test

```{r hetero}
result$heteroscedasticity
```

### Multicollinearity

```{r vif}
result$multicollinearity
```

### Influence Point Analysis

```{r influence}
result$influence
```

### Hypothesis Test Summary

```{r assumptions}
summarize_assumptions(model)
```

## Outlier Detection

### IQR Method

```{r iqr}
set.seed(123)
x <- c(rnorm(100), 10, -10)

result <- detect_outliers_iqr(x)
result$n_outliers
result$outliers
```

### Z-Score Method

```{r zscore}
result <- detect_outliers_zscore(x, threshold = 3)
result$n_outliers
```

### Grubbs Method

```{r grubbs}
set.seed(123)
x <- c(rnorm(100), 10)

result <- detect_outliers_grubbs(x)
result$p.value
result$outliers
```

### Dixon Method

```{r dixon}
set.seed(123)
x <- c(rnorm(20), 10)

result <- detect_outliers_dixon(x)
result$outliers
```

### Consensus Method

```{r consensus}
set.seed(123)
x <- c(rnorm(100), 10, -10)

result <- detect_outliers_all(x)
result$consensus
```

## SPC Control Rule Violations

### 8 Western Electric Rules

```{r spc_rules}
rules <- list_spc_rules()
rules
```

### Violation Detection

```{r spc_detect}
set.seed(123)
x <- c(rnorm(20, mean = 10, sd = 1), 13.5, rnorm(10, mean = 10, sd = 1))

result <- detect_spc_violations(x, center = 10, sigma = 1)
result$is_in_control
result$rules_triggered
```

### In-Control Data

```{r spc_in_control}
set.seed(123)
x <- rnorm(30, mean = 10, sd = 1)

result <- detect_spc_violations(x, center = 10, sigma = 1)
result$is_in_control
```
