---
title: "iQualityR Probability and Statistics Module Complete Guide"
author: "Chen Zhiming"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
vignette: >
  %\VignetteIndexEntry{iQualityR Probability Statistics Complete Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# 1. Core Features and Characteristics Introduction

The probability module (`iqr_prob`) of `iQualityR` is an advanced statistical analysis tool designed for quality management personnel, data scientists, and industrial engineers. Its core design philosophy is **"deep decoupling of computation and visualization"**.

### Main Features:

-   **Bidirectional Calculation Engine**: Supports "given observation value, find probability (Prob Mode)" and "given probability, find quantile (Quantile Mode)".
-   **Minitab Logic Alignment**: Calculation constants and discrete distribution handling logic strictly follow industrial standards (such as interval endpoint inclusion determination).
-   **Multi-Distribution Comparison (Overlay & Facet)**: Supports comparing different parameters and even different types of distributions in the same coordinate system or faceted plots.
-   **Intelligent Visual Management**: Achieves full-chain style unification from `ggplot2` plotting to Excel export through `IqrTheme`.
-   **Automated Annotation**: Automatically generates PDF probability density curve annotations and CDF cumulative distribution wires that comply with statistical reporting standards.

------------------------------------------------------------------------

# Complete List of Supported Distributions and Application Scenarios

This module is driven by `ProbNode` at the bottom layer, providing 12 distribution supports for different data types in quality management.

## Continuous Distributions

| Distribution Type | Corresponding R Function | Core Parameters | Industrial/Quality Management Application Scenarios |
|:-----------------|:-----------------|:-----------------|:-----------------|
| **Normal Distribution** | `norm` | `mean, sd` | **Core of measurement data**. Used for CPK analysis, control chart limit setting. |
| **T Distribution** | `t` | `df` | Small sample mean testing, confidence interval estimation. |
| **F Distribution** | `f` | `df1, df2` | Analysis of variance (ANOVA), evaluating batch-to-batch fluctuation differences. |
| **Chi-Square Distribution** | `chisq` | `df` | Used for variation evaluation in measurement system analysis (MSA). |
| **Weibull** | `weibull` | `shape, scale` | **Life analysis**. Used for product reliability, fatigue strength evaluation. |
| **Exponential Distribution** | `exp` | `rate` | Analyzing equipment mean time between failures (MTBF). |
| **Lognormal Distribution** | `lnorm` | `meanlog, sdlog` | Describes asymmetric part life due to wear or corrosion. |
| **Gamma Distribution** | `gamma` | `shape, rate` | Describes skewed process time distributions. |
| **Extreme Value Distribution** | `gumbel` | `loc, scale` | Extreme loads, maximum fracture strength analysis. |

## Discrete Distributions

| Distribution Type | Corresponding R Function | Core Parameters | Industrial/Quality Management Application Scenarios |
|:-----------------|:-----------------|:-----------------|:-----------------|
| **Binomial Distribution** | `binom` | `size, prob` | **Counting sampling**. Calculating batch non-conforming rate risk (PQC/OQC). |
| **Poisson Distribution** | `pois` | `lambda` | **Counting sampling**. Unit defect count (DPU) monitoring. |
| **Hypergeometric Distribution** | `hyper` | `m, n, k` | Sampling inspection without replacement from finite populations, more precise than binomial distribution. |

------------------------------------------------------------------------

```{r include=FALSE}
library(iQualityR.core)
library(iQualityR.plot)
library(iQualityR.stat)
```



# Function Feature Introduction: 8 Standard Stress Test Cases

The following cases demonstrate the performance of `iqr_prob` in terms of calculation logic, faceting, wires, and robustness.



```{r warning=FALSE}
# Initialization
iq <- iqr_prob$new()

# 1. Basic Quantile (Quantile, Lower)
iq$add_dist("Norm01", "norm", list(mean=0, sd=1))$calc(0.95, mode="quant")$plot(show_cdf=T)

# 2. Two-Sided Quantile (95% Confidence Interval)
iq$calc(c(0.025, 0.975), mode="quant", calc_type="between")$plot()

# 3. Discrete Quantile Stress Test (Binomial Upper)
iq$add_dist("B20", "binom", list(size=20, prob=0.5))$calc(0.1, mode="quant", calc_type="upper")$plot()

# 4. Multi-Distribution Facet Comparison (Facet Mode)
iq$add_dist("A", "norm", list(mean=0, sd=1))$add_dist("B", "t", list(df=5))$
   calc(c(-2, 2), calc_type="outside")$plot(facet=T)

# 5. Overlay Display (Overlay Mode, Verify Label Displacement)
iq$nodes <- list() 
iq$add_dist("D1", "norm", list(mean=10, sd=2))$add_dist("D2", "norm", list(mean=12, sd=2))$
   calc(11, calc_type="lower")$plot(facet=F)

# 6. Poisson Interval Probability (Discrete Between)
iq$nodes <- list()
iq$add_dist("P10", "pois", list(lambda=10))$calc(c(8, 12), calc_type="between")$plot()

# 7. Automatic Parameter Correction (Input c(12, 8) Auto-Sorted)
iq$calc(c(12, 8), calc_type="between")$plot()

# 8. Forced CDF Puzzle Mode
iq$calc(10)$plot(show_cdf=T)
```

------------------------------------------------------------------------

# Theme System (IqrTheme) Usage Examples

This section demonstrates how to inject visual styles at different lifecycle stages.

## Scenario A: Set Built-in Style at Instantiation (Tech)

Applicable for dashboards requiring modern, high-contrast UI displays.

``` {r warning=FALSE}
# Direct injection of preset at instantiation
iq_a <- iqr_prob$new(theme = "tech")
iq_a$add_dist("T1", "norm", list(mean=100, sd=10))$calc(115)$plot()
```

## Scenario B: Switch Mature Commercial Theme During Runtime (Economist)

Applicable for generating business reports for management.

```{r warning=FALSE}
# Switch to Economist style
iq_a$set_theme("economist")
iq_a$plot()
```

## Scenario C: Apply Custom Function Theme

Using the academic paper theme provided by the package.

```{r warning=FALSE}
iq_a$plot(theme_obj = "academic")
```

------------------------------------------------------------------------

# Practical Application Case Analysis

## Case 1: Manufacturing - Bearing Dimensional Out-of-Tolerance Risk Assessment

A precision bearing inner diameter requirement is $20 \pm 0.02 mm$, the actual production process mean is $20.005$, and the standard deviation is $0.008$. Evaluate the single-sided upper limit out-of-tolerance risk.

```{r warning=FALSE}
bearing_iq <- iqr_prob$new()
bearing_iq$add_dist("Bearing_ID", "norm", list(mean=20.005, sd=0.008))
bearing_iq$calc(20.02, calc_type="upper")
bearing_iq$plot()
```

## Case 2: Healthcare - Blood Sample Test Value Confidence Assessment

A certain biomarker follows a Weibull distribution (shape=1.5, scale=50) in healthy populations. A doctor needs to determine the critical concentration where 99% of the population falls below.

```{r warning=FALSE}
med_iq <- iqr_prob$new(theme = "tech")
med_iq$add_dist("Marker_Level", "weibull", list(shape=1.5, scale=50))
med_iq$calc(0.99, mode="quant")
med_iq$plot(show_cdf=T)
```

## Case 3: Service Industry - Call Center Peak Stress Test Analysis

Call arrivals per minute at a call center follow a Poisson distribution with $\lambda=8$. Evaluate the steady-state probability that the number of calls per minute is between 5 and 12.

```{r warning=FALSE}
service_iq <- iqr_prob$new()
service_iq$add_dist("Calls", "pois", list(lambda=8))
service_iq$calc(c(5, 12), calc_type="between")
service_iq$plot()
```

## Case 4: Old vs. New Process Comparison (F Distribution Variance Test)

Compare the stability of two production lines A and B, and determine the significance interval through F statistic.

```{r warning=FALSE}
compare_iq <- iqr_prob$new(theme = "tech")
compare_iq$add_dist("Process_Contrast", "f", list(df1=10, df2=10))
# Find the critical value for the two-sided 5% significance level
compare_iq$calc(c(0.025, 0.975), mode="quant")
compare_iq$plot()
```
