---
title: "iQualityR: Probability Distribution Calculation and Visualization Engine Guide"
author: "Chen Zhiming"
date: "2026-03-30"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{iQualityR: Probability Distribution Calculation and Visualization Engine Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(iQualityR.stat)
knitr::opts_chunk$set(fig.width = 6, fig.height = 4)
```

# 1. Overview: Why iQualityR?

In quality management, medical research, or economic forecasting, we often need to answer: "What is the probability of a certain value occurring?" Or "Where should specification limits be set to ensure a 95% pass rate?"

The `iQualityR` probability engine, encapsulated through R6 classes, solves pain points of the native `stats` package such as inconsistent parameter naming, repetitive visualization code, and labels that are difficult to use directly in reports.

------------------------------------------------------------------------

# 2. 12 Supported Distributions and Application Scenarios

This engine strictly aligns with the 12 most commonly used probability models in industry and research:

| Classification | Distribution Type (`type`) | Typical Application Scenarios |
|:-----------------------|:-----------------------|:-----------------------|
| **Symmetric Distributions** | `norm`, `t` | Measurement data analysis (such as part length), small sample mean testing. |
| **Skewed/Life Distributions** | `lnorm`, `weibull`, `gamma`, `exp` | Product reliability, failure time, income distribution, medical duration. |
| **Ratio/Count Distributions** | `binom`, `pois`, `nbinom`, `hyper` | Sampling inspection pass rate, defects per unit, rare disease incidence. |
| **Sampling Distributions** | `f`, `chisq` | Analysis of variance (ANOVA), chi-square independence testing. |

------------------------------------------------------------------------

# 3. Real-World Cases: From Symbols to Decisions

### 3.1 Quality Management: HNB Tobacco Process Optimization (Weibull)

In the R&D of heat-not-burn (HNB) tobacco, the heating element's lifespan typically follows a **Weibull distribution**. * **Scenario**: Known component `shape = 2`, `scale = 1000` (hours). * **Task**: Calculate the probability that this component fails within 800 hours.

```{r warning=FALSE}
# Instantiate and calculate cumulative failure probability
iq_life <- iqr_prob$new(type = "weibull", params = list(shape = 2, scale = 1000))
iq_life$calc(values = 800, calc_type = "lower")

# Visualization output
iq_life$plot(show_cdf = TRUE, facet = TRUE)
```

### 3.2 Biomedical: Rare Disease Screening (Poisson)

-   **Scenario**: The incidence rate of a certain rare disease in a specific population is `lambda = 2` (persons/year).
-   **Task**: Calculate the probability of more than 5 cases occurring in one year.

```{r warning=FALSE}
# Instantiate Poisson distribution
iq_med <- iqr_prob$new(type = "pois", params = list(lambda = 2),theme = "economist")
iq_med$calc(values = 5, calc_type = "upper")

# Presentation: Discrete distribution automatically switches to bar chart
iq_med$plot()
```

------------------------------------------------------------------------

# 4. Feature Combinations: DOE Cross-Validation Case

To demonstrate the system's flexibility, complex presentation requirements can be achieved through different "factor" combinations.

### Case A: Interval Probability + Facet View + Academic Theme

**Factor Combination**: `norm` + `between` + `facet=TRUE` + `theme="academic"`

```{r warning=FALSE}
# Calculate the coverage rate of standard normal distribution [-1.96, 1.96]
iq_norm <- iqr_prob$new(type = "norm",theme = "academic")
iq_norm$calc(values = c(-1.96, 1.96), calc_type = "between")

# Display: PDF shows P(-1.96 <= X <= 1.96), CDF shows deltaP height difference
iq_norm$plot(facet = TRUE,theme_obj = "economist")
```

### Case B: Two-Sided Tail (Outside) + Single View + Industrial Theme

**Factor Combination**: `t` + `outside` + `facet=FALSE` + `theme="mtr"`

```{r warning=FALSE}
# Verify the significance region outside +/-2 for t distribution (df=10)
iq_t <- iqr_prob$new(type = "t", params = list(df = 10))
iq_t$calc(values = c(-2, 2), calc_type = "outside")

# Present using Minitab style (mtr) theme
iq_t$plot(show_cdf = FALSE,theme_obj = "tech")
```

------------------------------------------------------------------------

# 5. Theme Settings and Personalized Usage

`iQualityR` supports highly customizable theme transmission. Users can switch different visual styles through the `theme_obj` parameter:

-   **`academic`**: Classic white background with black lines, suitable for inserting into academic papers or formal technical reports.
-   **`mtr`**: Simulates Minitab style with deep blue tones, conforming to industrial aesthetic habits.
-   **`dark`**: Dark mode, suitable for integration into large-screen monitoring or Dashboard systems.

**Advanced Usage: Custom Layer Overlay** Since `plot()` returns a standard `ggplot2` object, you can use ggplot2 functions directly:

```{r warning=FALSE}
p <- iq_norm$plot()
p + ggplot2::labs(title = "Custom Title")
```

------------------------------------------------------------------------

# 6. Precautions and Defensive Programming

1.  **Parameter Validation**: The system has built-in `validate_dist_params`. For example, if you try to pass `sd = -1` to `norm`, the engine will immediately error and provide a Chinese explanation.
2.  **Hard-Coded Labels**: All graph labels use hard-coded text mode with `parse = FALSE`. This means you can safely generate graphs containing `≤` or `ΔP` on any operating system without encountering `unexpected symbol` errors.
3.  **Calculation Precision**: Numerical results default to 4 decimal places, meeting most industrial standards.

------------------------------------------------------------------------

*End of Vignette*
