---
title: "Probability Distribution Analysis Guide"
author: "iQualityR Development Team"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Probability Distribution Analysis Guide}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

 
```{r setup, echo = FALSE}
library(iQualityR.stat)
library(R6)

library(patchwork)
library(iQualityR.core)
library(iQualityR.plot)

```

## 1. What is Probability Distribution Analysis?

In quality management, we often need to answer questions like:

-   **Conformity Assessment**: What is the probability that the number of non-conforming items in a batch does not exceed 3?
-   **Specification Compliance**: What is the probability that the product dimension falls within specification limits?
-   **Risk Assessment**: What is the likelihood that the time between failures is less than 100 hours?
-   **Sampling Decision**: If we inspect 50 items, what is the maximum number of non-conforming items allowed to accept the batch as acceptable?

**iQualityR's Probability Distribution Module**
is designed to help you answer these questions quickly and accurately. It supports **16
probability distributions**, covering the vast majority of statistical inference scenarios in quality management.

### Module Features

| Feature | Description |
|---------|-------------|
| **Dual Interface Design** | R6 class interface (chained calls) + convenience functions (one-time use) |
| **16 Distributions** | 12 continuous + 4 discrete |
| **4 Calculation Modes** | Left tail, right tail, within interval, outside interval |
| **Visualization** | PDF/CDF plots, shaded regions, wire annotations |
| **Result Interpretation** | Translates statistical terminology into language understandable by quality personnel |
| **Report Export** | Console printing, data frame export, Excel report |

------------------------------------------------------------------------

## 2. Quick Start

### Scenario 1: Normal Distribution - Product Dimension Compliance Probability

Assuming a product's critical dimension follows a normal distribution N(100, 5), with an upper specification limit of 108, find the probability that the dimension exceeds the upper limit.

```{r norm-example, warning=FALSE}
# library(iQualityR.stat)

# Create normal distribution object
prob <- iqr_prob$new(type = "norm", params = list(mean = 100, sd = 5),theme = "academic")

# Calculate P(X > 108)
prob$calc(values = 108, mode = "prob", calc_type = "upper")

# Plot (with CDF)
prob$plot(show_cdf = TRUE)
```

**Interpretation**: P(X \> 108) = 0.0548, meaning approximately 5.5%
of products will exceed the upper specification limit, corresponding to approximately 55,000 DPMO.

### Scenario 2: Binomial Distribution - Sampling Compliance Decision

A production process has an average non-conforming rate of 5%. If we inspect 50 items, find the probability that the number of non-conforming items does not exceed 3.

```{r binom-example}
# Create binomial distribution object
prob_binom <- iqr_prob$new(type = "binom",  params = list(size = 50, prob = 0.05))

# Calculate P(X <= 3)
prob_binom$calc(values = 3, mode = "prob", calc_type = "lower")

# Generate management-level interpretation
prob_binom$interpret(audience = "manager")
```

**Common Misconception**: Many people directly calculate 3/50 =
6%, but this differs from the meaning of the binomial distribution. The 87.3% from the binomial distribution represents the likelihood of the event "no more than 3 items" occurring, taking into account sampling variability.

### Scenario 3: Exponential Distribution - Equipment Time Between Failures

The time between failures for a piece of equipment follows an exponential distribution (average 500 hours). Find the probability that the equipment fails within 100 hours.

```{r exp-example, warning=FALSE}
prob_exp <- iqr_prob$new(type = "exp", params = list(rate = 1/500))
prob_exp$calc(values = 100, mode = "prob", calc_type = "lower")
prob_exp$plot(show_cdf = TRUE)
```

------------------------------------------------------------------------

## 3. Available Distributions Overview

### 3.1 View All Distributions

```{r list-dists}
# List all available distributions
list_prob_distributions()
```

### 3.2 Continuous Distributions (12 types)

| Distribution | Parameters | Typical Applications |
|--------------|------------|---------------------|
| **norm** | mean, sd | Measurement error, product quality characteristics |
| **weibull** | shape, scale | Life analysis, reliability |
| **lnorm** | meanlog, sdlog | Particle size, reaction time |
| **gamma** | shape, scale | Waiting time, cumulative quantities |
| **exp** | rate | Time between failures |
| **t** | df | Small sample inference |
| **f** | df1, df2 | Variance comparison, ANOVA |
| **chisq** | df | Goodness of fit, variance test |
| **beta** | shape1, shape2 | Proportion/rate (yield, defect rate) |
| **unif** | min, max | Uniform distribution (MSA) |
| **logis** | location, scale | Robust analysis |
| **cauchy** | location, scale | Heavy-tailed distribution |

### 3.3 Discrete Distributions (5 types)

| Distribution | Parameters | Typical Applications |
|--------------|------------|---------------------|
| **binom** | size, prob | Number of non-conforming items |
| **pois** | lambda | Number of defects, event occurrence count |
| **nbinom** | size, prob | Number of successes before a given number of failures |
| **hyper** | m, n, k | Finite population sampling |
| **geom** | prob | Number of trials before first failure |

------------------------------------------------------------------------

## 4. Calculation Mode Details

### 4.1 Probability Calculation (Given X, Find P)

```{r prob-modes}
# Standard normal distribution
prob <- iqr_prob$new(type = "norm", params = list(mean = 0, sd = 1))

# Left tail: P(X <= 1.96)
prob$calc(values = 1.96, calc_type = "lower")

# Right tail: P(X > 1.96)
prob$calc(values = 1.96, calc_type = "upper")

# Within interval: P(-1.96 <= X <= 1.96)
prob$calc(values = c(-1.96, 1.96), calc_type = "between")

# Outside interval: P(X < -1.96 or X > 1.96)
prob$calc(values = c(-1.96, 1.96), calc_type = "outside")
```

### 4.2 Quantile Calculation (Given P, Find X)

```{r quant-mode}
# Given probability, find quantile
prob$calc(values = 0.975, mode = "quant")  # 97.5% quantile
prob$calc(values = c(0.025, 0.975), mode = "quant")  # Two-sided quantiles
```

------------------------------------------------------------------------

## 5. Visualization Options

### 5.1 Basic PDF Plot

```{r basic-plot, warning=FALSE}
prob <- iqr_prob$new(type = "norm", params = list(mean = 0, sd = 1))
prob$calc(values = 1.96, calc_type = "upper")
prob$plot()
```

### 5.2 PDF + CDF Combined Plot

```{r pdf-cdf-plot,warning=FALSE}
prob$plot(show_cdf = TRUE)
```

### 5.3 Multi-Distribution Comparison (Facet)

```{r multi-dist-plot, warning=FALSE}
prob_multi <- iqr_prob$new(
  dist_list = list(
    list(id = "Standard Normal", type = "norm", params = list(mean = 0, sd = 1)),
    list(id = "Wide Distribution", type = "norm", params = list(mean = 0, sd = 2)),
    list(id = "Narrow Distribution", type = "norm", params = list(mean = 0, sd = 0.5))
  )
)

prob_multi$calc(values = 1.96, calc_type = "upper")
prob_multi$plot(facet = TRUE)
```

### 5.4 Discrete Distribution Visualization

```{r discrete-plot, warning=FALSE}
prob_disc <- iqr_prob$new(type = "pois", params = list(lambda = 5))
prob_disc$calc(values = 7, calc_type = "lower")
prob_disc$plot(show_cdf = TRUE)
```

------------------------------------------------------------------------

## 6. Result Interpretation Features

### 6.1 For Management

```{r interp-manager, eval = TRUE}
prob <- iqr_prob$new(type = "binom", params = list(size = 50, prob = 0.05))
prob$calc(values = 3, calc_type = "lower")
prob$interpret(audience = "manager")
```

### 6.2 For Technical Personnel

```{r interp-technical}
prob$interpret(audience = "technical")
```

### 6.3 For Customers

```{r interp-client}
prob$interpret(audience = "client")
```

### 6.4 Hypothesis Test Interpretation

```{r htest-interp}
# Hypothesis test results can also be interpreted directly
set.seed(123)
group1 <- rnorm(30, mean = 100, sd = 10)
group2 <- rnorm(30, mean = 105, sd = 10)

t_result <- t.test(group1, group2)
interp <- StatInterpreter$new()
interp$interpret(t_result, audience = "manager")
```

------------------------------------------------------------------------

## 7. Convenience Functions (No R6 Object Required)

If you only need quick queries and don't want to create R6 objects, you can use convenience functions:

### 7.1 prob_calc() - Quick Calculation

```{r prob-calc}
# Normal distribution: P(X > 1.96)
prob_calc(type = "norm", params = list(mean = 0, sd = 1),
          x = 1.96, calc_type = "upper")

# Binomial distribution: P(X <= 3), with interpretation
prob_calc(type = "binom", params = list(size = 50, prob = 0.05),
          x = 3, calc_type = "lower", interpret = TRUE)
```

### 7.2 prob_plot() - Quick Plotting

```{r prob-plot-func, warning=FALSE}
prob_plot(type = "norm", params = list(mean = 0, sd = 1),
          x = 1.96, calc_type = "upper", show_cdf = TRUE)
```

### 7.3 list_prob_distributions() - View Distribution List

```{r list-prob-func}
list_prob_distributions()
```

------------------------------------------------------------------------

## 8. Report Export

### 8.1 Data Frame Export

```{r report-df}
prob <- iqr_prob$new(type = "norm", params = list(mean = 100, sd = 5))
prob$calc(values = 108, calc_type = "upper")

# Export as data frame
df <- prob$report()
print(df)
```

### 8.2 Excel Report

```{r report-excel, eval = FALSE}
# Export as Excel report (multiple worksheets)
prob$report_excel(path = "Probability_Analysis_Report.xlsx")
```

Excel report contains: - **Calculation Results**: Input values, calculation results, formula descriptions -
**Distribution Parameters**: Distribution type, parameters, support set, description - **Usage Instructions**: Meaning of mode and
calc_type

------------------------------------------------------------------------

## 9. Custom Distributions

If built-in distributions don't meet your needs, you can add custom distributions through the registry:

```{r custom-dist, eval = FALSE}
# Register custom distribution
register_dist(
  dist_type = "my_custom",
  d = function(x, p) dnorm(x, p$mean, p$sd) * 1.1,  # Custom density function
  p = function(q, p, lower.tail) pnorm(q, p$mean, p$sd, lower.tail),
  q = function(p_val, p, lower.tail) qnorm(p_val, p$mean, p$sd, lower.tail),
  defaults = list(mean = 0, sd = 1),
  is_discrete = FALSE,
  support = "(-inf, +inf)",
  description = "My custom distribution"
)

# Use custom distribution
prob <- iqr_prob$new(type = "my_custom", params = list(mean = 0, sd = 1))
prob$calc(values = 1.96, calc_type = "upper")
```

------------------------------------------------------------------------

## 10. Common Scenario Quick Reference

| Scenario | Distribution | Code Example |
|---------------------|---------------------|------------------------------|
| Product dimension yield | norm | `prob_calc("norm", list(mean=100, sd=5), x=108, "upper")` |
| Sampling non-conforming count | binom | `prob_calc("binom", list(size=50, prob=0.05), x=3, "lower")` |
| Defects per unit time | pois | `prob_calc("pois", list(lambda=2), x=4, "upper")` |
| Equipment life | weibull | `prob_calc("weibull", list(shape=2, scale=1000), x=500, "lower")` |
| Reaction time | lnorm | `prob_calc("lnorm", list(meanlog=4, sdlog=0.5), x=100, "upper")` |
| Variance ratio test | f | `prob_calc("f", list(df1=9, df2=9), x=3.18, "upper")` |
| Finite population sampling | hyper | `prob_calc("hyper", list(m=10, n=90, k=5), x=1, "upper")` |

------------------------------------------------------------------------

## 11. Frequently Asked Questions

### Q1: What's the difference between "lower" and "upper" for calc_type?

-   **lower**: Calculate left-tail probability P(X <= x)
-   **upper**: Calculate right-tail probability P(X > x)
-   **between**: Calculate interval probability P(a <= X <= b)
-   **outside**: Calculate probability outside interval P(X < a or X > b)

### Q2: When to use probability calculation vs quantile calculation?

-   **Probability Calculation** (mode = "prob"): Given X value, find corresponding probability. For example: "probability that dimension exceeds 108"
-   **Quantile Calculation** (mode = "quant"): Given probability, find corresponding X value. For example: "what dimension do 95% of products not exceed"

### Q3: How to choose the appropriate distribution?

| Data Type | Recommended Distribution |
|-----------|--------------------------|
| Continuous measurement data (symmetric) | Normal distribution (norm) |
| Continuous measurement data (right-skewed) | Lognormal (lnorm) or Gamma (gamma) |
| Life/reliability | Weibull (weibull) or Exponential (exp) |
| Non-conforming count (fixed sample) | Binomial (binom) |
| Defects per unit area/time | Poisson (pois) |
| Proportion/rate | Beta distribution (beta) |

### Q4: How are "business recommendations" in result interpretation generated?

The system automatically provides recommendations based on calculated probability values: - **>= 95%**: Process is good, maintain status quo - **90-95%**: Basically in control, pay appropriate attention - **80-90%**: Acceptable but with risk, strengthen monitoring - **50-80%**: Risk exists, consider process adjustment - **< 50%**: High risk, recommend investigation of causes

------------------------------------------------------------------------

*This vignette was written by the iQualityR Team | Version
`r packageVersion("iQualityR.stat")`*
