---
title: "Probability Distribution and Sigma Estimation"
author: "iQualityR Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Probability Distribution and Sigma Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r include=FALSE}
library(iQualityR.stat)
library(R6)

library(patchwork)
library(iQualityR.core)
library(iQualityR.plot)

```

## Overview

Probability distributions are the mathematical foundation of quality management. `iQualityR.stat` includes 16+ commonly used probability distributions and provides 7 Sigma estimation methods (following ASTM E2587 standard).

## Probability Distribution

### View Available Distributions

```{r list_dists}
list_available_dists()
```

### Get Distribution Information

```{r dist_info}
get_dist_info("norm")
get_dist_info("weibull")
```

### Distribution Calculation

```{r prob_calc}
prob_calc("norm", x = 1.96, mean = 0, sd = 1)
prob_calc("binom", x = 5, size = 10, prob = 0.5)
```

### Distribution Visualization

```{r prob_plot, fig.width=7, fig.height=5}
prob_plot("norm", mean = 0, sd = 1)
```

## Sigma Estimation

### 7 Estimation Methods

| Method | Description |
|--------|-------------|
| `overall` | Overall standard deviation |
| `within` | Within-group standard deviation |
| `rbar` | Average range method |
| `sbar` | Average standard deviation method |
| `mr` | Moving range method |
| `pooled` | Pooled standard deviation method |
| `median` | Median method |

### Example

```{r sigma}
set.seed(123)
x <- rnorm(100, mean = 10, sd = 2)

sigma_estimate(x, method = "total")
sigma_estimate(x, method = "mr_bar")
```

### Sigma Decomposition

```{r sigma_decomp}
set.seed(123)
x <- rnorm(100, mean = 10, sd = 2)
subgroup <- rep(1:5, each = 20)

sigma_decomposition(x, subgroup = subgroup)
```

## SPC Constants

```{r spc_constants}
get_d2(2)
get_A2(5)
get_c4(3)
```

## Quality Metrics

### Process Capability and PPM Conversion

```{r capability}
capability_to_ppm(cpk = 1.33, usl = 10, lsl = 0)
capability_to_ppm(cpk = 1.67, ppk = 1.50, usl = 10, lsl = 0)
```

### Sigma Level and PPM

```{r sigma_ppm}
sigma_to_ppm(sigma = 3)
sigma_to_ppm(sigma = 6)
ppm_to_sigma(ppm = 2700)
```

### Throughput Yield

```{r ytp}
throughput_yield(yield = c(0.95, 0.98, 0.99))
```

### Reliability Metrics

```{r reliability}
reliability(lambda = 0.001, t = 100)
availability(mtbf = 1000, mttr = 10)
```
