---
title: "Sample Size and Power Analysis"
author: "iQualityR.stat Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Sample Size and Power Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

library(iQualityR.stat)
library(iQualityR.plot)
library(ggplot2)

# Helpers
fmt_num <- function(x, digits = 4) formatC(x, format = "f", digits = digits)
fmt_pct <- function(x, digits = 2) paste0(formatC(x * 100, format = "f", digits = digits), "%")
```

## 1. Introduction

Sample size and power analysis sit at the planning stage of every quality
study. Too few samples and a real process shift goes undetected (Type II
error); too many and resources are wasted. The `iQualityR.stat` package
provides a unified family of functions for sizing studies across the quality
engineering spectrum — from simple mean comparisons to reliability
demonstration tests.

**Module map**

| Function family | Key functions |
|-----------------|---------------|
| Means | `sample_size_mean()`, `sample_size_two_means()`, `sample_size_paired()` |
| Proportions | `sample_size_proportion()`, `sample_size_two_proportions()` |
| ANOVA | `sample_size_anova()` |
| Regression | `sample_size_regression()` |
| Confidence-interval width | `sample_size_ci()` |
| Tolerance-interval width | `sample_size_tolerance()` |
| Reliability / success-run | `sample_size_reliability()` |
| Power & effect size | `calc_power()`, `effect_size()`, `power_table()`, `odds_ratio()` |

## 2. Theory: Power and Sample Size

### 2.1 Hypothesis testing framework

Every sample-size calculation derives from the same logical scaffold. Given a
null hypothesis $H_0$ and an alternative $H_1$, we control two error rates:

| Truth | Reject $H_0$ | Fail to reject $H_0$ |
|-------|-------------|---------------------|
| $H_0$ true | Type I error ($\alpha$) | Correct ($1-\alpha$) |
| $H_1$ true | Correct (power $= 1-\beta$) | Type II error ($\beta$) |

- **Significance level** $\alpha$: probability of a false alarm.
- **Power** $1-\beta$: probability of detecting a real effect of size
  $\delta$.
- **Effect size** $d$: the magnitude of the effect in standardised units,
  e.g. Cohen's $d = \delta / \sigma$.

### 2.2 The normal-approximation formula

For a one-sample / two-sided test the required sample size is:

$$
n = \left( \frac{z_{1-\alpha/2} + z_{1-\beta}}{d} \right)^2, \qquad d = \frac{\mu_1 - \mu_0}{\sigma}
$$

where $z_{1-\alpha/2}$ and $z_{1-\beta}$ are standard-normal quantiles. For
small $n$ the $t$-distribution replaces $z$, and the formula is solved
iteratively.

### 2.3 The power curve

Power grows with sample size $n$ and effect size $d$. A **power curve** plots
$1-\beta$ against $n$ for a fixed $d$, answering the planner's question:
*"How many units must I test to be 80 % sure of catching a 0.5$\sigma$
shift?"*

### 2.4 Effect-size conventions

| Effect size | Cohen's $d$ | Cohen's $f^2$ (regression) |
|-------------|-------------|---------------------------|
| Small       | 0.20        | 0.02                      |
| Medium      | 0.50        | 0.15                      |
| Large       | 0.80        | 0.35                      |

For ANOVA, $f^2 = \eta^2 / (1 - \eta^2)$; for regression,
$f^2 = R^2 / (1 - R^2)$.

## 3. Sample Size for Means

### 3.1 One-sample mean — `sample_size_mean()`

```{r mean_one}
# Detect a 0.5-sigma shift in a turning operation (mu0 = 10, mu1 = 10.5, sigma = 1)
res <- sample_size_mean(mu0 = 10, mu1 = 10.5, sigma = 1, power = 0.80)
res
```

The returned list carries the planned $n$, the **actual power** achieved
(after $t$-correction), and the standardised effect size:

```{r mean_one_extras}
c(n = res$n, actual_power = res$actual_power, effect_size = res$effect_size)
```

### 3.2 Two-sample mean — `sample_size_two_means()`

```{r mean_two}
# Compare two suppliers: detect a 0.3-unit difference with sigma = 1
sample_size_two_means(mu1 = 10.0, mu2 = 10.3, sigma = 1, power = 0.80)
```

### 3.3 Paired t-test — `sample_size_paired()` {.newest}

The paired test reduces to a one-sample test on the within-pair differences
$d = y_2 - y_1$. The sample size is driven by the **standard deviation of the
differences** $\sigma_d$, not the raw measurement SD:

$$
n = \left( \frac{z_{1-\alpha/2} + z_{1-\beta}}{\delta / \sigma_d} \right)^2
$$

```{r paired}
# Before/after gauge calibration: expect 0.4-unit shift, SD of diffs = 1.0
sample_size_paired(mu0 = 0, mu1 = 0.4, sigma_d = 1.0, power = 0.80)
```

Because the paired test is a one-sample $t$-test on differences, when
$\sigma_d = \sigma$ the result matches `sample_size_mean()`:

```{r paired_matches}
paired_res   <- sample_size_paired(mu0 = 0, mu1 = 0.5, sigma_d = 1, power = 0.80)
onesample_res <- sample_size_mean(mu0 = 0, mu1 = 0.5, sigma = 1, power = 0.80)
c(paired_n = paired_res$n, one_sample_n = onesample_res$n)
```

## 4. Sample Size for Proportions

### 4.1 One-sample proportion — `sample_size_proportion()`

```{r prop_one}
# Detect improvement from 95% to 98% yield
sample_size_proportion(p0 = 0.95, p1 = 0.98, power = 0.90)
```

### 4.2 Two-sample proportion — `sample_size_two_proportions()`

```{r prop_two}
# Compare two lines: 92% vs 96% first-pass yield
sample_size_two_proportions(p1 = 0.92, p2 = 0.96, power = 0.80)
```

## 5. Sample Size for ANOVA — `sample_size_anova()`

For a one-way ANOVA with $k$ groups, the effect size is
$f = \sigma_{\text{means}} / \sigma_{\text{error}}$, and the sample size is
found via the noncentral $F$ distribution.

```{r anova_ss}
# Compare 3 suppliers with means 10, 10.5, 11 and common sigma = 1.5
sample_size_anova(k = 3, means = c(10, 10.5, 11), sigma = 1.5, power = 0.80)
```

## 6. Sample Size for Regression — `sample_size_regression()` {.newest}

The global $F$-test of a regression with $p$ predictors tests
$H_0\colon R^2 = 0$. The effect size is Cohen's
$f^2 = R^2 / (1 - R^2)$, and the required $n$ is found by searching the
noncentral $F$ distribution until the achieved power meets the target.

```{r reg_ss}
# 3 predictors, anticipated R^2 = 0.30 (medium effect)
sample_size_regression(p = 3, r_squared = 0.30, power = 0.80)
```

Adding predictors raises the required $n$ (more degrees of freedom in the
numerator, more parameters to estimate):

```{r reg_compare}
sample_size_regression(p = 2,  r_squared = 0.30, power = 0.80)$n
sample_size_regression(p = 10, r_squared = 0.30, power = 0.80)$n
```

## 7. CI-Width-Driven Sample Size — `sample_size_ci()` {.newest}

Sometimes the goal is not hypothesis testing but **precision**: *"How many
samples do I need so that the 95 % CI half-width is at most $h$?"*

**Mean** (known $\sigma$):
$$
n = \left( \frac{z_{1-\alpha/2}\, \sigma}{h} \right)^2
$$

**Proportion**:
$$
n = \frac{z_{1-\alpha/2}^2 \; p(1-p)}{h^2}
$$

```{r ci_mean}
# 95% CI half-width <= 0.3 for a measurement with sigma = 1
sample_size_ci(type = "mean", h = 0.3, sigma = 1, conf_level = 0.95)
```

```{r ci_prop}
# 95% CI half-width <= 5% for a proportion near 50%
sample_size_ci(type = "proportion", h = 0.05, p = 0.5, conf_level = 0.95)
```

## 8. Tolerance-Interval-Driven Sample Size — `sample_size_tolerance()` {.newest}

A normal-theory tolerance interval $\bar{x} \pm k \cdot s$ must cover at least
proportion $p$ of the population with confidence $1-\gamma$. The
**k-factor** (Howe 1969 / exact chi-square) is:

$$
k = \sqrt{ \frac{\nu \, (1 + 1/n) \, z_p^2}{\chi^2_{\gamma,\,\nu}} }, \qquad \nu = n - 1
$$

`sample_size_tolerance()` finds the smallest $n$ such that the expected
half-width $k \cdot E[s] = k \cdot c_4(n)\,\sigma$ does not exceed
`max_half_width` $\times\; \sigma$.

```{r tol}
# 95% content / 95% confidence; half-width must stay within 3 sigma
sample_size_tolerance(p = 0.95, conf_level = 0.95, max_half_width = 3.0)
```

A tighter width budget demands more data:

```{r tol_compare}
sample_size_tolerance(p = 0.95, conf_level = 0.95, max_half_width = 5.0)$n
sample_size_tolerance(p = 0.95, conf_level = 0.95, max_half_width = 2.5)$n
```

## 9. Reliability / Success-Run — `sample_size_reliability()` {.newest}

For a reliability demonstration test with zero allowed failures, the
**success-run formula** gives:

$$
n = \frac{\ln(1 - \text{conf})}{\ln(R)}
$$

where $R$ is the target reliability and `conf` the desired confidence. When
$n_{\text{failures}} > 0$ failures are allowed, $n$ is found by solving the
binomial CDF constraint $P(X \le r \mid n, 1-R) \le 1 - \text{conf}$.

```{r rel_zero}
# Demonstrate 95% reliability with 90% confidence, 0 failures
sample_size_reliability(reliability = 0.95, conf_level = 0.90)
```

Allowing failures increases the required $n$ but permits a more forgiving
test:

```{r rel_failures}
sample_size_reliability(reliability = 0.95, conf_level = 0.90, n_failures = 0)$n
sample_size_reliability(reliability = 0.95, conf_level = 0.90, n_failures = 2)$n
```

## 10. Power, Effect Size, and Odds Ratios

### 10.1 Post-hoc power — `calc_power()`

Given a fixed $n$ and effect size, compute the achieved power:

```{r calc_power}
# n = 50, effect size d = 0.5, two-sided t-test
calc_power(n = 50, effect_size = 0.5, alpha = 0.05)
```

### 10.2 Effect size — `effect_size()`

```{r effect_size}
# Cohen's d: two groups with means 10 and 11, pooled SD = 1.5
effect_size(type = "cohens_d", mean1 = 10, mean2 = 11, sd_pooled = 1.5)

# Eta-squared from an F-test
effect_size(type = "eta_squared", F = 4.5, df_between = 3, df_within = 96)
```

### 10.3 Power curve — `power_table()`

```{r power_table, fig.width = 8, fig.height = 4}
# Power vs n for effect size d = 0.5
 
ptab <- power_table(effect_size = 0.5, n_range = seq(10, 120, by = 5))

p <- base_plot(ptab, aes(n, power), theme = "academic") +
    layers_trend_line(add_points = TRUE)
 
thm <- attr(p, "iqr_theme")
ui  <- thm$get_ui_colors()
ref_color <- ui$border    

p <- p +
    geom_hline(yintercept = 0.80, linetype = "dashed", color = ref_color) +
    annotate("text", x = 110, y = 0.82, label = "80% target",
             size = 3, color = ref_color) +
    labs(title = "Power Curve: effect size d = 0.5",
         x = "Sample size n", y = "Power")
p
```

### 10.4 Odds ratio — `odds_ratio()`

For a 2×2 contingency table (cells $a, b, c, d$):

```{r odds_ratio}
# Defective counts: a=5 (A defective), b=45 (A good), c=15 (B defective), d=35 (B good)
odds_ratio(a = 5, b = 45, c = 15, d = 35, conf_level = 0.95)
```

## 11. Quality Scenario: Sizing a Capability Study

**Context.** A plating line manager must verify that a process mean has not
drifted more than $0.5\sigma$ from target. The study must achieve 80 % power
at $\alpha = 0.05$. After data collection the team also needs to report the
95 % CI for the mean with a half-width no larger than $0.3\sigma$.

### Step 1 — Size for detection

```{r scenario_detect}
s_detect <- sample_size_mean(mu0 = 100, mu1 = 100.5, sigma = 1, power = 0.80)
s_detect$n
```

### Step 2 — Size for precision

```{r scenario_precision}
s_ci <- sample_size_ci(type = "mean", h = 0.3, sigma = 1, conf_level = 0.95)
s_ci$n
```

### Step 3 — Take the larger sample size

```{r scenario_final}
n_final <- max(s_detect$n, s_ci$n)
n_final
```

### Step 4 — Verify achieved power

```{r scenario_power}
calc_power(n = n_final, effect_size = 0.5, alpha = 0.05)
```

## 12. Parameter Reference

| Function | Key parameters | Returns |
|----------|---------------|---------|
| `sample_size_mean()` | `mu0, mu1, sigma, alpha, power, alternative` | `n, actual_power, effect_size` |
| `sample_size_two_means()` | `mu1, mu2, sigma, alpha, power, alternative, ratio` | `n1, n2, total_n, actual_power, effect_size` |
| `sample_size_paired()` | `mu0, mu1, sigma_d, alpha, power, alternative` | `n, actual_power, effect_size` |
| `sample_size_proportion()` | `p0, p1, alpha, power, alternative` | `n, actual_power` |
| `sample_size_two_proportions()` | `p1, p2, alpha, power, alternative, ratio` | `n1, n2, total_n, actual_power, effect_size` |
| `sample_size_anova()` | `k, means, sigma, alpha, power` | `n, actual_power, effect_size` |
| `sample_size_regression()` | `p, r_squared, alpha, power` | `n, actual_power, f_squared` |
| `sample_size_ci()` | `type, h, sigma, p, conf_level` | `n, half_width, full_width` |
| `sample_size_tolerance()` | `p, conf_level, max_half_width, sigma` | `n, k_factor, half_width` |
| `sample_size_reliability()` | `reliability, conf_level, n_failures` | `n, achieved_conf_level` |
| `calc_power()` | `n, effect_size, alpha, alternative, test_type` | numeric power |
| `effect_size()` | `type` (`cohens_d`, `hedges_g`, `eta_squared`, `omega_squared`, `cohens_h`, `r`) + args | numeric |
| `power_table()` | `effect_size, alpha, n_range, alternative` | data.frame |
| `odds_ratio()` | `a, b, c, d, conf_level` | `odds_ratio, conf.int` |

## 13. Decision Guide

| Planning question | Use |
|-------------------|-----|
| Detect a mean shift of $\delta$ | `sample_size_mean()` or `sample_size_two_means()` |
| Before/after paired measurement | `sample_size_paired()` |
| Detect a yield / proportion change | `sample_size_proportion()` or `sample_size_two_proportions()` |
| Compare $k$ groups (suppliers, machines) | `sample_size_anova()` |
| Regression model significance | `sample_size_regression()` |
| CI must be narrower than $2h$ | `sample_size_ci()` |
| Tolerance interval width budget | `sample_size_tolerance()` |
| Reliability demonstration (zero failures) | `sample_size_reliability()` |
| Post-hoc: did my study have enough power? | `calc_power()` |
| Visualise power vs $n$ | `power_table()` + `ggplot()` |

## 14. References

- Cohen, J. (1988). *Statistical Power Analysis for the Behavioral Sciences* (2nd ed.). Lawrence Erlbaum.
- Howe, W. G. (1969). Two-sided tolerance limits for normal populations — some improvements. *Journal of the American Statistical Association*, 64(326), 610–620.
- Montgomery, D. C. (2019). *Introduction to Statistical Quality Control* (8th ed.). Wiley.
- Meeker, W. Q., Hahn, G. J., & Escobar, L. A. (2017). *Statistical Intervals* (2nd ed.). Wiley.

## Further Reading

- See the **Hypothesis Testing and Intervals** vignette for the test procedures these sample sizes feed into:
  `vignette("hypothesis-testing-and-intervals", package = "iQualityR.stat")`
- See the **ANOVA** vignette for multi-group comparisons:
  `vignette("anova", package = "iQualityR.stat")`
- See the **Descriptive Statistics and Distributions** vignette for tolerance intervals in context:
  `vignette("descriptive-and-distributions", package = "iQualityR.stat")`
