---
title: "Sample Size and Power Analysis"
author: "iQualityR Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sample Size and Power Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r include=FALSE}
library(iQualityR.stat)
library(R6)

library(patchwork)
library(iQualityR.core)
library(iQualityR.plot)

```

## Overview

Sample size calculation ensures statistical analysis has sufficient power to avoid false negative results due to insufficient sample size. This module supports sample size calculation for means, proportions, and ANOVA.

## Sample Size Calculation

### Mean Comparison

```{r setup}
sample_size_mean(mu0 = 10, mu1 = 10.5, sigma = 1, power = 0.80)
```

### Proportion Comparison

```{r prop}
sample_size_proportion(p0 = 0.5, p1 = 0.6, power = 0.80)
```

### ANOVA

```{r anova}
sample_size_anova(k = 3, means = c(10, 10.5, 11), sigma = 1.5, power = 0.80)
```

## Power Calculation

### Mean Power

```{r power_mean}
# Calculate power given sample size and effect size
calc_power(n = 50, effect_size = 0.5, alpha = 0.05)
```

### Proportion Power

```{r power_prop}
# Calculate power for proportion test
calc_power(n = 100, effect_size = 0.1, alpha = 0.05)
```

### ANOVA Power

```{r power_anova}
# Calculate sample size for ANOVA
sample_size_anova(k = 3, means = c(10, 10.5, 11), sigma = 1.5, power = 0.80)
```

## Effect Size

### Cohen's d

```{r cohen_d}
effect_size(type = "cohens_d", mu1 = 10, mu2 = 11, sd_pooled = 1)
```

### Cohen's f

```{r cohen_f}
# Cohen's f for ANOVA effect size
# f = sigma_m / sigma where sigma_m is SD of group means and sigma is error SD
effect_size(type = "eta_squared", F = 4.5, df_between = 3, df_within = 96)
```

## Application Scenarios

### Scenario 1: New Product Validation

Assuming a new product has an average lifespan of 1000 hours with a standard deviation of 50 hours, and we want to detect a 10-hour difference with 80% power:

```{r scenario1}
sample_size_mean(mu0 = 1000, mu1 = 1010, sigma = 50, power = 0.80)
```

### Scenario 2: Yield Comparison

Current yield is 95%, and we want to detect a 3% improvement with 90% power:

```{r scenario2}
sample_size_proportion(p0 = 0.95, p1 = 0.98, power = 0.90)
```

### Scenario 3: Multi-group Comparison

Comparing the mean differences of 4 processes, with effect size f = 0.3 and 85% power:

```{r scenario3}
# For f = 0.3, sigma = 1, the means should have SD = 0.3
# With k=4 groups and equal spacing: means = c(0, 0.3, 0.6, 0.9) has mean = 0.45, SD ~ 0.37
# Let's use a simpler setup with known means
sample_size_anova(k = 4, means = c(10, 10.3, 10.6, 10.9), sigma = 1, power = 0.85)
```
