---
title: "Standard Deviation Estimation Methods: Statistical Theory and Application Guide"
author: "iQualityR Package"
date: "2026-03-29"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Standard Deviation Estimation Methods: Statistical Theory and Application Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

```{r setup, echo = FALSE}
library(iQualityR.stat)
library(R6)

library(patchwork)
library(iQualityR.core)
library(iQualityR.plot)

```


## Overview

In Statistical Process Control (SPC) and process capability analysis, **accurate estimation of standard deviation ($\sigma$)** is central to measuring process variation. Incorrect estimation methods lead to control limits that are too wide or too narrow, causing false alarms or missed detections.

This document details the various standard deviation estimation methods implemented in the `iQualityR` package. These methods are mainly divided into two categories:
1.  **Within-Subgroup Variation Estimation**: Uses fluctuation within subgroups to estimate short-term process variation.
2.  **Between-Subgroup Variation Estimation**: Uses fluctuation between subgroup means to estimate long-term or systematic variation.

---

## 1. Basic Statistics and Unbiased Correction

Before diving into specific estimation methods, it is essential to understand the concept of unbiased estimation. Sample statistics (such as range $R$ and sample standard deviation $s$) are typically biased estimators of population parameters. We need to correct them using specific constants.

### 1.1 Unbiased Correction Factors for Range and Standard Deviation

Let $X_1, X_2, ..., X_n$ be independent and identically distributed samples from the normal distribution $N(\mu, \sigma^2)$.

*   **$d_2$ constant**: Expectation factor for range.
    $$E[R] = d_2(n) \cdot \sigma$$
    Where $R = \max(X_i) - \min(X_i)$.

*   **$c_4$ constant**: Expectation factor for sample standard deviation.
    $$E[s] = c_4(n) \cdot \sigma$$
    Where $s$ is the sample standard deviation.

### 1.2 Estimation Strategy

To obtain an unbiased standard deviation estimate $\hat{\sigma}$, we use the following general formulas:

*   **Based on range**:
    $$\hat{\sigma} = \frac{\bar{R}}{d_2(n)}$$

*   **Based on sample standard deviation**:
    $$\hat{\sigma} = \frac{\bar{s}}{c_4(n)}$$

If we do not apply the above corrections (i.e., default $d_2=1$ or $c_4=1$), for small samples, the estimates will systematically underestimate the true value.

---

## 2. Within-Subgroup Standard Deviation Estimation Methods

Within-group estimation methods aim to capture the "short-term" or "inherent" variation (Within Variation) of a process, i.e., the random fluctuations within subgroups.

### 2.1 Range Method ("r_bar")

This is the most classical method, applicable when subgroup sample size is small (typically $n \leq 10$).

*   **Mathematical Logic**:
    Calculate the range $R_i$ for each subgroup, then calculate the average range $\bar{R}$.
    $$\hat{\sigma}_{r\_bar} = \frac{\bar{R}}{d_2(n)}$$

*   **R Code Implementation**:
```r
# Assume stats is a summary dataframe containing sub_range (subgroup range)
r_bar <- mean(stats$sub_range, na.rm = TRUE)
sigma_hat <- r_bar / get_d2(n_size)
```

### 2.2 Standard Deviation Method ("s_bar")

When subgroup sample size is large, the range method loses information. At this point, the standard deviation method should be used.

*   **Mathematical Logic**:
    Calculate the standard deviation $s_i$ for each subgroup, then calculate the weighted average.
    $$\hat{\sigma}_{s\_bar} = \frac{\sum s_i \cdot c_4(n_i)}{\sum c_4(n_i)}$$
    If all subgroups are the same size, this simplifies to $\bar{s} / c_4(n)$.

*   **R Code Implementation**:
```r
# Assume stats is a summary dataframe containing sub_sd (subgroup standard deviation)
s_bar <- mean(stats$sub_sd, na.rm = TRUE)
if (use_unbiased) {
  sigma_hat <- s_bar / get_c4(n_size)
} else {
  sigma_hat <- s_bar
}
```

### 2.3 Pooled Standard Deviation Method ("pooled_s")

Used when it is necessary to comprehensively estimate from all subgroup data at once, and subgroup sizes may differ.

*   **Mathematical Logic**:
    Based on the weighted average of variances of all observations.
    $$s_p = \sqrt{\frac{\sum_{i=1}^m (n_i-1)s_i^2}{\sum_{i=1}^m (n_i-1)}}$$

*   **R Code Implementation**:
```r
# Assume stats contains sub_var (subgroup variance) and sub_n (subgroup size)
numerator <- sum((stats$sub_n - 1) * stats$sub_var, na.rm = TRUE)
denominator <- sum(stats$sub_n - 1, na.rm = TRUE)
pooled_s <- sqrt(numerator / denominator)
```

---

## 3. Between-Subgroup and Moving Variation Estimation Methods

When subgroup data cannot be obtained (e.g., destructive testing) or long-term variation needs to be evaluated, estimation methods based on sequential data are required.

### 3.1 Moving Range Method: Mean ("mr_bar")

Applicable when variation between subgroups is large, estimates through the moving range of subgroup means.

*   **Mathematical Logic**:
    Calculate the moving range $MR_i$ of consecutive $m$ subgroup means, and take its average.
    $$\hat{\sigma}_{mr\_bar} = \frac{\overline{MR}}{d_2(m)}$$
    Where $m$ is the moving span.

*   **R Code Implementation**:
```r
means <- stats$sub_mean
if (length(means) <= m_span) return(NA_real_)
mr <- abs(diff(means, lag = (m_span - 1))) 
mean_mr <- mean(mr, na.rm = TRUE)
sigma_hat <- mean_mr / get_d2(m_span)
```

### 3.2 Moving Range Method: Median ("mr_median")

This is a robust version of `mr_bar`, applicable when data contains outliers.

*   **Mathematical Logic**:
    Similar to `mr_bar`, but uses median instead of mean, and uses the median correction factor $ d_4 $.

$$
\hat{\sigma}_{mr\_median} = \frac{MR_{median}}{d_4(m)}
$$

*   **R Code Implementation**:
```r
means <- stats$sub_mean
if (length(means) <= m_span) return(NA_real_)
mr <- abs(diff(means, lag = (m_span - 1)))
median_mr <- stats::median(mr, na.rm = TRUE)
sigma_hat <- median_mr / get_d4(m_span) # Note: use get_d4 here
```

### 3.3 Mean Squared Successive Difference (MSSD)

MSSD (Mean Squared Successive Difference) is a statistic used to estimate variation in autocorrelated processes or single-value sequences.

*   **Mathematical Logic**:
    Based on the mean of squared differences between consecutive observations.

$$
MSSD = \frac{\sum_{i=2}^n (x_i - x_{i-1})^2}{n-1}
$$

    MSSD is an unbiased estimator of $2\sigma^2 $. Therefore:

$$
\hat{\sigma}_{mssd} = \sqrt{\frac{MSSD}{2}}
$$

    To apply unbiased correction to the square root, introduce factor $ c_4' $:

$$
\hat{\sigma}_{mssd} = \frac{\sqrt{MSSD/2}}{c_4'(n)}
$$

*   **R Code Implementation**:
```r
# Calculate MSSD
diff_sq <- diff(data_vector)^2
mssd_val <- sqrt(sum(diff_sq) / (2 * (length(data_vector) - 1)))

# Apply unbiased correction
if (use_unbiased) {
  mssd_val <- mssd_val / get_c4_prime(length(data_vector))
}
```

---

## 4. Total Standard Deviation Estimation

### 4.1 Total Standard Deviation ("total")

This is the most direct method, directly calculating the sample standard deviation of all data, reflecting the "long-term" total variation (Total Variation) of the process.

*   **Mathematical Logic**:

$$
s_{total} = \sqrt{\frac{1}{N-1}\sum_{i=1}^N (x_i - \bar{x})^2}
$$

    If an unbiased estimate of population $ \sigma $ is needed, divide by $ c_4(N) $.

*   **Application Scenarios**:
    Used to calculate long-term process performance indices such as $ P_p $ and $ P_{pk} $.

*   **R Code Implementation**:
```r
total_sd <- sd(all_data, na.rm = TRUE)
if (use_unbiased) {
  total_sd <- total_sd / get_c4(length(all_data))
}
```

---

## 5. Standard Deviation Decomposition

In advanced SPC analysis, we often decompose the total standard deviation into within-group and between-group components to identify the main sources of variation.

### 5.1 Decomposition Principle

Based on the additivity of variance:

$$
\sigma_{total}^2 = \sigma_{within}^2 + \sigma_{between}^2
$$

The specific formula is:

$$
\hat{\sigma}_{total} = \sqrt{\hat{\sigma}_{within}^2 + \frac{\hat{\sigma}_{between}^2}{n}}
$$

Where:
*    $ \hat{\sigma}_{within} $: Usually estimated by "r_bar" or "s_bar".
*    $ \hat{\sigma}_{between} $: Usually estimates variation in subgroup means by "mr_bar" or "mssd", then converts to single-value variation by multiplying by $ \sqrt{n} $.

### 5.2 Application Scenarios

*   **Large within-group variation ( $ \sigma_{within} $ )**: Indicates the process is greatly affected by random noise; equipment or environment needs improvement.
*   **Large between-group variation ( $ \sigma_{between} $ )**: Indicates serious drift in process mean (Drift); need to check tool wear, operator differences, or material batches.

---

## 6. Method Selection Guide

| Method Name | Applicable Scenarios | Sample Requirements | Advantages | Disadvantages |
| :--- | :--- | :--- | :--- | :--- |
| **r_bar** | Traditional manufacturing, $ n \le 10 $ | Subgroup data | Simple calculation, historical standard | Low information utilization |
| **s_bar** | Precision manufacturing, $ n > 10 $ | Subgroup data | More accurate than range method | Slightly more complex calculation |
| **pooled_s** | Summary analysis | Subgroup data | Integrates all sample information | Sensitive to non-normality |
| **mr_bar** | Between-batch variation monitoring | Subgroup mean sequence | Applicable to large subgroups or single values | Sensitive to outliers |
| **mr_median** | Data containing outliers | Subgroup mean sequence | **Robust** | Slightly less efficient |
| **mssd** | Destructive testing, time series | Single value sequence | Can handle autocorrelated data | Requires large sample correction |

---

## 7. Practical Recommendations and Common Mistakes

### 7.1 Common Mistakes

1.  **Confusing $ d_2 $ and $ c_4 $**:
    *   Wrong: Using the expectation $ d_2 $ of $ R/\sigma $ to correct sample standard deviation $ s $.
    *   Correct: Use $ d_2 $ for range, use $ c_4 $ for standard deviation.

2.  **Ignoring $ c_4' $ correction**:
    *   When using MSSD to estimate $ \sigma $, the square root operation introduces bias. Although it can be ignored for large samples, it is recommended to use $ c_4' $ for correction when $ n < 20 $.

3.  **Improper selection of moving span ( $ m $ )**:
    *   Usually $ m=2 $ (adjacent two points). If the process has periodicity, adjust $ m $ according to the cycle.

### 7.2 Complete Example

```{r warning=FALSE}
# Assume we have a set of simulated data
set.seed(123)
data <- rnorm(50, mean = 100, sd = 10)
subgroup <- rep(1:10, each = 5)

# 1. Estimate within-group standard deviation (using range method)
sigma_within <- sigma_estimate(data, subgroup, method = "r_bar")

# 2. Estimate between-group standard deviation (using MSSD)
sigma_between <- sigma_estimate(data, subgroup, method = "mssd")

# 3. Decompose total standard deviation
decomposition_result <- sigma_decomposition(data, subgroup, 
                                            within_method = "r_bar", 
                                            between_method = "mssd")

print(decomposition_result)
```

Through the systematic methods introduced in this document, users can select the most appropriate estimation strategy based on actual data characteristics, thereby ensuring the accuracy of process control and capability analysis.
