---
title: "Statistical Methods for Quality: An Overview"
author: "iQualityR.stat Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Statistical Methods for Quality: An Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

library(iQualityR.stat)
library(iQualityR.plot)
library(ggplot2)

# Helpers
fmt_num <- function(x, digits = 4) formatC(x, format = "f", digits = digits)
fmt_pct <- function(x, digits = 2) paste0(formatC(x * 100, format = "f", digits = digits), "%")
```

## 1. Introduction

### 1.1 The Role of Statistics in Quality Engineering

Modern quality engineering rests on a foundation of statistical thinking. From
Walter Shewhart's invention of the control chart in the 1920s, through Genichi
Taguchi's robust design methods, to today's Six Sigma deployments, every major
advance in quality management has been driven by a statistical insight. The
reason is fundamental: **quality is inherently about variation**, and statistics
is the science of describing, explaining, and controlling variation.

Consider the daily decisions a quality professional faces:

- Is the process mean on target? (hypothesis testing)
- What range contains the true mean? (confidence intervals)
- Will the next part meet specification? (prediction intervals)
- How many parts must I measure to be confident? (sample size)
- Is the process capable of meeting customer requirements? (capability analysis)
- Which supplier has the best process? (ANOVA)
- Can I predict defects from process settings? (regression)
- Is the measurement system adequate? (MSA — handled by `iQualityR.msa`)

Each of these questions is a statistical question. The `iQualityR.stat` package
provides a unified, theory-driven toolkit for answering them within a single
coherent workflow.

### 1.2 Design Philosophy: The `iqr_*` Entry Pattern

The package is organized around **entry classes** — R6 objects that serve as
the user's primary interface to each statistical module. Each entry class
encapsulates three layers of responsibility behind a simple lifecycle:

```
User (you)
  │
  ▼
iqr_*  entry class  ──►  Analyzer (L1: compute)
  │                     ──►  Plotter  (L2: visualize)
  │                     ──►  Reporter (L2: export)
  │                     ──►  Interpreter (L2: explain)
  ▼
stat_result (S3)
```

- **L1 Engine (Analyzer):** Performs the statistical computation and returns a
  `stat_result` S3 object — a structured, serializable result containing
  statistics, p-values, confidence intervals, and metadata.
- **L2 Presentation (Plotter / Reporter / Interpreter):** Transforms the result
  into a ggplot object, an Excel/data-frame report, or a plain-language
  interpretation tailored to the audience (manager, technical, client).
- **L3 Entry (iqr_* class):** Orchestrates the layers and caches the last
  result for iterative analysis (`$plot()`, `$report()`, `$interpret()` all
  operate on the cached result).

For one-off analyses where object lifecycle management is unnecessary, every
module also exposes **convenience functions** (e.g., `htest_run()`,
`intervals_run()`, `anova_run()`) that compute and return a `stat_result`
directly.

## 2. Module Map

The package provides eight statistical modules, each accessed through an entry
class or convenience function:

| Module | Entry Class | Primary Method | Convenience Functions | Coverage |
|--------|-------------|----------------|----------------------|----------|
| Descriptive statistics | `iqr_desc()` (function) | direct call | `desc_calc()`, `desc_analyze()`, `desc_hist()`, `desc_box()` | Mean, median, SD, variance, skewness, kurtosis, CV, CI |
| Normality assessment | `iqr_normality` (R6) | `$test()` | `normality_test()`, `normality_plot()` | Shapiro-Wilk, Anderson-Darling, CVM, Lilliefors, SF |
| Hypothesis testing | `iqr_htest` (R6) | `$run()` | `htest_run()`, `htest_plot()` | 22 test types (z, t, proportion, chi-square, F, Levene, Bartlett, Wilcoxon, Kruskal-Wallis, Friedman, TOST, equivalence, Poisson, correlation) |
| Intervals | `iqr_intervals` (R6) | `$run()` | `intervals_run()` | Confidence, prediction (sample & model), tolerance (Howe & exact) |
| Probability distributions | `iqr_prob` (R6) | `$calc()` | `prob_calc()`, `prob_plot()` | 12 registered distributions (Normal, Exponential, Gamma, Weibull, Lognormal, Beta, Uniform, Cauchy, t, F, Chi-square, Binomial, Poisson, ...) |
| ANOVA | `iqr_anova` (R6) | `$run()` | `anova_run()`, `anova_report()` | One-way, two-way, multi-factor, repeated measures, ANOM |
| Regression | `iqr_regression` (R6) | `$run()` | `regression_run()` | lm, logistic, Poisson, Cox, PLS, stepwise, best-subset, MARS, spline |
| Resampling | `iqr_resampling` (R6) | `$run()` | `resampling_run()` | Bootstrap CI (percentile, BCa), permutation tests |

In addition, the package provides **cross-cutting tools** that support multiple
modules:

| Tool Category | Key Functions | Purpose |
|---------------|--------------|---------|
| Sample size & power | `sample_size_mean()`, `sample_size_paired()`, `sample_size_regression()`, `calc_power()`, `power_table()`, `effect_size()` | Study design before data collection |
| Sigma estimation | `sigma_estimate()`, `sigma_decomposition()`, `get_d2()`, `get_c4()`, `get_A2()`, ... | Process standard deviation and SPC control-chart constants |
| Capability metrics | `capability_to_ppm()`, `ppm_to_sigma()`, `z_bench()`, `capability_interpret()` | Cp/Cpk → PPM → Z.bench conversion |
| Distribution fitting | `fit_distribution()`, `auto_fit_distribution()`, `compare_fits()`, `calc_qq_data()` | Fit theoretical distributions to data |
| Data transformation | `box_cox_transform()`, `yeo_johnson_transform()`, `johnson_transform()`, `auto_transform()` | Normalize non-normal data for capability/SPC |
| Model diagnostics | `diagnose_lm()`, `test_residual_normality()`, `test_heteroscedasticity()`, `diagnose_multicollinearity()`, `diagnose_influential_points()` | Regression assumption checking |
| Outlier detection | `detect_outliers_iqr()`, `detect_outliers_zscore()`, `detect_outliers_grubbs()`, `detect_outliers_dixon()`, `detect_outliers_mahalanobis()` | Identify unusual observations |
| SPC rules | `check_spc_rule()`, `detect_spc_violations()`, `list_spc_rules()`, `summarize_spc_rules()` | Western Electric / Nelson run-rule checks |

## 3. The Unified Workflow

### 3.1 Entry-Class Lifecycle

Most `iqr_*` entry classes follow the same lifecycle. The primary compute
method is `$run()` for most modules, with `$test()` for normality and `$calc()`
for probability distributions:

```r
# 1. Create the entry object (with theme for consistent visuals)
obj <- iqr_htest$new(theme = "academic")

# 2. Run the analysis (caches result in $last_results)
obj$run("t_test_1s", x = my_data, mu = 74.00)

# 3. Visualize (uses cached result)
obj$plot(plot_type = "auto")

# 4. Interpret in plain language (manager / technical / client)
obj$interpret(audience = "manager")

# 5. Export (data.frame / console / excel)
obj$report(format = "data.frame")

# Or do it all in one call:
obj$analyze("t_test_1s", x = my_data, mu = 74.00, plot_type = "auto")
```

The `$analyze()` method is a convenience that runs, plots, and interprets in
a single call — useful for interactive exploration.

### 3.2 The `stat_result` S3 Class

Every Analyzer returns a `stat_result` S3 object. This uniform structure is the
contract between layers:

```r
result <- htest_run("t_test_1s", x = ring_dia, mu = 74.00)
cat("Domain:     ", result$domain, "\n")       # "htest"
cat("Test type:  ", result$test_type, "\n")    # "t_test_1s"
cat("Method:     ", result$method, "\n")       # "One Sample t-test"
cat("Statistic:  ", result$statistic, "\n")    # t value
cat("P-value:    ", result$p.value, "\n")      # significance
cat("Conf. int:  ", result$conf.int, "\n")     # 95% CI
```

The `stat_result` carries enough metadata for any L2 layer (Plotter, Reporter,
Interpreter) to render the result without needing to re-run the computation.

## 4. Quick Start: A Complete Analysis

### 4.1 Quality Scenario: Piston Ring Inner Diameter

**Background.** A manufacturer of automotive piston rings produces rings with a
nominal inner diameter of 74.00 mm (tolerance ±0.05 mm). This is the classic
case study from Montgomery, D.C. (2013), *Introduction to Statistical Quality
Control*, 7th Edition, Wiley — Chapter 5. The process parameters used here
(target = 74.00 mm, process σ ≈ 0.012 mm) reflect the real production
characteristics documented in that reference.

A quality engineer collects 40 consecutive ring-diameter measurements from the
current production run and wants to answer four questions:

1. What are the basic characteristics of the data? (descriptive statistics)
2. Is the data normally distributed? (normality test — prerequisite for t-test)
3. Is the process mean on target (74.00 mm)? (one-sample t-test)
4. What is the 95% confidence interval for the true mean? (interval estimation)

### 4.2 The Data

```{r}
# Piston ring inner diameter (mm)
# Process parameters from Montgomery (2013), ISQC 7th Ed., Ch. 5
# Target = 74.00 mm, process sigma ~ 0.012 mm
set.seed(2026)
ring_dia <- rnorm(40, mean = 74.001, sd = 0.012)
ring_df <- data.frame(diameter = ring_dia)

# First look
head(ring_df, 8)
```

### 4.3 Step 1: Descriptive Statistics

```{r}
# One-call descriptive summary
desc_result <- desc_calc(ring_dia)
cat(sprintf("n         = %d\n", desc_result$n))
cat(sprintf("Mean      = %.4f mm\n", desc_result$mean))
cat(sprintf("SD        = %.4f mm\n", desc_result$sd))
cat(sprintf("Median    = %.4f mm\n", desc_result$median))
cat(sprintf("Skewness  = %.4f\n", desc_result$skewness))
cat(sprintf("Kurtosis  = %.4f\n", desc_result$kurtosis))
cat(sprintf("CV        = %.4f%%\n", desc_result$cv * 100))
```

The mean (74.0005 mm) is very close to the target (74.00 mm), and the CV is
only 0.016% — indicating a tight, well-controlled process. The skewness near
zero and kurtosis near 3 (excess ≈ 0) suggest the data is approximately normal.

### 4.4 Step 2: Normality Assessment

Before applying a t-test, we verify the normality assumption. The
`iqr_normality` entry class provides an automated test selection that picks the
most powerful test for the sample size.

```{r}
# Normality test (auto-selects Shapiro-Wilk for n=40)
norm_obj <- iqr_normality$new()
norm_result <- norm_obj$test(ring_dia, method = "auto", alpha = 0.05)
```

With p = `r formatC(norm_result$p_value, format = "f", digits = 4)` (> 0.05),
we fail to reject the null hypothesis of normality. The t-test assumption is
satisfied.

### 4.5 Step 3: Hypothesis Test — Is the Mean on Target?

```{r}
# One-sample t-test: H0: mu = 74.00 vs H1: mu != 74.00
ht_obj <- iqr_htest$new()
ht_obj$run("t_test_1s", x = ring_dia, mu = 74.00, alternative = "two.sided")

# Plain-language interpretation for management
ht_obj$interpret(audience = "manager")
```

### 4.6 Step 4: Confidence Interval for the Mean

```{r}
# 95% confidence interval for the true mean
itv_obj <- iqr_intervals$new()
itv_obj$run("ci_mean", x = ring_dia, conf_level = 0.95)
itv_obj$interpret(audience = "technical")
```

### 4.7 Step 5: Putting It All Together — The Analysis Report

```{r}
# Export results to a data frame for reporting
ht_df <- ht_obj$report(format = "data.frame")
itv_df <- itv_obj$report(format = "data.frame")

cat("=== Hypothesis Test Summary ===\n")
print(ht_df)

cat("\n=== Confidence Interval Summary ===\n")
print(itv_df)
```

### 4.8 Conclusion

| Question | Answer | Evidence |
|----------|--------|----------|
| Data characteristics? | Mean ≈ 74.00, SD ≈ 0.012, near-normal | Descriptive statistics |
| Normality? | Yes (p = `r formatC(norm_result$p_value, format = "f", digits = 4)`) | Shapiro-Wilk test |
| Mean on target? | Yes (p = `r formatC(ht_obj$last_results$p.value, format = "f", digits = 4)`) | One-sample t-test |
| 95% CI for mean? | [`r formatC(itv_obj$last_results$conf.int[1], format = "f", digits = 4)`, `r formatC(itv_obj$last_results$conf.int[2], format = "f", digits = 4)`] mm | t-interval |

The process is on target and the confidence interval lies well within the
tolerance band [73.95, 74.05] mm. The engineer can report to management that
the process is performing as designed.

## 5. Architecture: How the Layers Connect

### 5.1 The Four-Layer Architecture

```
┌─────────────────────────────────────────────────────┐
│  L3  Entry Classes (iqr_htest, iqr_anova, ...)     │  ← User interface
│      Convenience functions (htest_run, ...)         │
├─────────────────────────────────────────────────────┤
│  L2  Plotter  │  Reporter  │  Interpreter           │  ← Presentation
├─────────────────────────────────────────────────────┤
│  L1  Analyzers (HTestAnalyzer, AnovaAnalyzer, ...)  │  ← Computation
├─────────────────────────────────────────────────────┤
│  L0  R primitives (stats::t.test, stats::aov, ...)  │  ← Foundation
└─────────────────────────────────────────────────────┘
```

**L0 — R Primitives.** The package builds on R's `stats` package and validated
CRAN packages (`nortest`, `moments`, `splines`, `earth`, `survival`). We do not
reinvent statistical algorithms; we wrap them in a quality-engineering context.

**L1 — Analyzers.** Each `*Analyzer` R6 class (e.g., `HTestAnalyzer`,
`AnovaAnalyzer`, `RegressionAnalyzer`) wraps the R primitives into a dispatch
interface keyed by a `test_type` / `model_type` / `interval_type` string code.
The Analyzer normalizes inputs, validates parameters, and returns a
`stat_result` S3 object.

**L2 — Presentation.** Each module has a matching `*Plotter` (returns ggplot),
`*Reporter` (returns data.frame / writes Excel), and shares a `StatInterpreter`
for plain-language output. All visualizations use the unified `IqrTheme` system
from `iQualityR.core` for consistent appearance.

**L3 — Entry.** The `iqr_*` classes and convenience functions orchestrate L1
and L2. They cache the last result (`$last_results`) so the user can iterate on
visualization and reporting without recomputing.

### 5.2 The `stat_result` Contract

The `stat_result` S3 class is the data contract between L1 and L2. Every
Analyzer returns one, and every Plotter/Reporter/Interpreter accepts one. The
core fields:

| Field | Type | Description |
|-------|------|-------------|
| `domain` | character | Module domain (`"htest"`, `"anova"`, `"regression"`, ...) |
| `test_type` | character | Specific method code (`"t_test_1s"`, `"oneway"`, `"lm_fit"`, ...) |
| `method` | character | Human-readable method name |
| `statistic` | numeric | Test statistic |
| `p.value` | numeric | P-value (NA if not applicable) |
| `conf.int` | numeric(2) | Confidence interval |
| `estimate` | numeric | Point estimate |
| `data_name` | character | Description of the input data |

Module-specific results add additional fields (e.g., `model_stats` for
regression, `anova_table` for ANOVA, `coefficients` for regression).

## 6. Where to Go Next

This overview introduced the package architecture and a simple end-to-end
analysis. The remaining vignettes in this chapter provide in-depth coverage of
each module:

| Vignette | Topic | Key Question Answered |
|----------|-------|----------------------|
| `descriptive-and-distributions` | Descriptive statistics, probability distributions, fitting & transformation | What does my data look like? |
| `hypothesis-testing-and-intervals` | 22 hypothesis tests, confidence/prediction/tolerance intervals, resampling | Is the parameter different from a standard? |
| `sample-size-and-power` | Sample size calculation and power analysis | How much data do I need? |
| `anova` | Analysis of variance and ANOM | Are the group means different? |
| `regression-modeling` | 9 regression models, diagnostics, outlier detection | Can I predict quality from inputs? |
| `process-control-and-capability` | Sigma estimation, SPC constants, capability metrics | Is the process in control and capable? |

```r
# List all available vignettes
vignette(package = "iQualityR.stat")
```

## References

1. Montgomery, D.C. (2013). *Introduction to Statistical Quality Control*, 7th
   Edition. John Wiley & Sons. — The piston ring case study (Ch. 5) and SPC
   foundations (Ch. 6–8) informed the quality scenarios throughout this package.

2. Shewhart, W.A. (1931). *Economic Control of Quality of Manufactured Product*.
   Van Nostrand. — The foundational text of statistical process control.

3. ISO 5479:1997. *Statistical interpretation of data — Tests for departure
   from the normal distribution*. — Basis for the normality testing module.

4. ISO 16269-6:2014. *Statistical interpretation of data — Part 6:
   Determination of statistical tolerance intervals*. — Basis for the tolerance
   interval methods (Howe's method, exact k-factor).

5. Hahn, G.J. & Meeker, W.Q. (1991). *Statistical Intervals: A Guide for
   Practitioners*. John Wiley & Sons. — Comprehensive reference on confidence,
   prediction, and tolerance intervals.

## Further Reading

- For the theme system that gives all plots a consistent look, see
  `vignette("theme-system", package = "iQualityR.core")`.
- For the base classes that all `iqr_*` entry classes inherit from, see
  `vignette("iQualityR-core", package = "iQualityR.core")`.
- For the visualization infrastructure used by all plotters, see
  `vignette("plotting-overview", package = "iQualityR.plot")`.
