---
title: "Advanced Statistical Plots"
author: "iQualityR.plot Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Advanced Statistical Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

library(iQualityR.plot)
library(ggplot2)

fmt_num <- function(x, digits = 4) formatC(x, format = "f", digits = digits)
fmt_pct <- function(x, digits = 2) paste0(formatC(x, format = "f", digits = digits), "%")
```

## 1. Introduction

Beyond the basic quality tools — Pareto charts, histograms, and control
charts — the `iQualityR.plot` package offers a family of advanced statistical
plots that support multivariate analysis, designed-experiment diagnostics,
time-series stability checks, classifier evaluation, and measurement system
analysis (MSA). These plots share a common theming layer through `IqrTheme`
and return `ggplot` objects, so they compose naturally with the rest of the
`ggplot2` ecosystem.

Every function in this vignette accepts a `theme` argument that can be
either a theme name (e.g. `"academic"`, `"economist"`, `"solarized"`,
`"prism"`) or an `IqrTheme` object. Colors — including diverging palettes
for correlations, discrete palettes for variance components, and semantic
colors for ACF significance lines — are resolved automatically through
`.iqr_aes()` based on the active theme, so users rarely need to specify
colors explicitly.

This vignette covers six functions:

* `plot_correlation_heatmap()` — multivariate correlation matrices.
* `plot_interaction_line()` — two-way interaction effects in factorial
  designs.
* `plot_acf()` and `plot_pacf()` — autocorrelation and partial
  autocorrelation for time-series and process-stability diagnostics.
* `plot_roc_curve()` — receiver operating characteristic curves for binary
  classifiers.
* `plot_variance_components()` — variance decomposition for MSA.

Each section develops the underlying theory, demonstrates the function on
realistic quality-engineering data, and closes with interpretation guidance.

## 2. plot_correlation_heatmap() — multivariate correlation

### 2.1 Theory: Pearson/Spearman correlation and diverging palettes

The Pearson correlation coefficient between two random variables $X$ and $Y$
is defined as

$$
r_{XY} = \frac{\sum_{i=1}^{n}(x_i - \bar{x})(y_i - \bar{y})}
               {\sqrt{\sum_{i=1}^{n}(x_i - \bar{x})^2 \cdot \sum_{i=1}^{n}(y_i - \bar{y})^2}},
$$

and ranges from $-1$ (perfect negative linear association) through $0$ (no
linear association) to $+1$ (perfect positive linear association). The
Spearman rank correlation replaces the raw values with their ranks before
applying the same formula, capturing monotonic but non-linear relationships
and is robust to outliers.

A correlation matrix is symmetric with ones on the diagonal. Visualizing it
requires a **diverging palette** — a color scale that runs from one hue
through a neutral midpoint to a contrasting hue. The neutral midpoint must
sit exactly at zero so that the sign of the correlation is encoded in color.
This is a critical rule: a sequential palette (e.g. a plain blue ramp) must
never be used for correlations, because a strong negative correlation and a
strong positive correlation would receive similar colors and the sign
information would be lost.

A practical threshold for **multicollinearity** is $|r| > 0.8$ between two
predictors. Pairs exceeding this threshold inflate the variance of regression
coefficient estimates and should be flagged before modeling.

### 2.2 Function signature

```r
plot_correlation_heatmap(data, theme = NULL, digits = 2, diagonal = FALSE)
```

`data` is a data frame of numeric columns. `digits` controls the decimal
places of the printed coefficients. When `diagonal = FALSE` (the default) the
self-correlations on the diagonal are set to `NA`, removing the trivial
ones from the figure so that genuine cross-variable relationships stand out.

### 2.3 Example: correlated process variables

We simulate four process variables in which the etch rate is driven by
temperature and pressure.

```{r}
set.seed(42)
process_vars <- data.frame(
  temperature = rnorm(120, 1200, 15),
  pressure = rnorm(120, 5.2, 0.3),
  gas_flow = rnorm(120, 250, 10),
  etch_rate = rnorm(120, 450, 20)
)
process_vars$etch_rate <- with(process_vars,
  300 + 0.8 * temperature / 10 + 20 * pressure + rnorm(120, 0, 10))

plot_correlation_heatmap(process_vars, digits = 2)
```

The diverging palette immediately reveals the strong positive correlation
between temperature and etch rate and between pressure and etch rate, while
`gas_flow` shows only weak association with the others — a useful signal that
gas flow can be retained as an independent predictor without multicollinearity
concerns.

Keeping the diagonal visible is occasionally useful when the matrix is to be
read as a complete descriptive summary:

```{r}
plot_correlation_heatmap(process_vars, digits = 2, diagonal = TRUE)
```

## 3. plot_interaction_line() — interaction effects

### 3.1 Theory: two-way interactions in factorial designs

In a two-way factorial design with factors $A$ and $B$, the response is
modeled as

$$
Y_{ijk} = \mu + \alpha_i + \beta_j + (\alpha\beta)_{ij} + \varepsilon_{ijk},
$$

where $\alpha_i$ is the main effect of level $i$ of factor $A$, $\beta_j$ is
the main effect of level $j$ of factor $B$, and $(\alpha\beta)_{ij}$ is the
**interaction** — the extra deviation from additivity. An interaction means
that the effect of factor $A$ depends on the level of factor $B$.

The interaction plot makes this visible by plotting the mean response of $Y$
against the levels of $A$, with one line per level of $B$.

* **Parallel lines** $\Rightarrow$ no interaction; the effect of $A$ is the
  same at every level of $B$.
* **Non-parallel lines** $\Rightarrow$ interaction present; the effect of $A$
  changes across levels of $B$. Crossing lines indicate a disordinal
  (crossover) interaction.

### 3.2 Function signature

```r
plot_interaction_line(data, x_var, y_var, group_var,
                      fun = c("mean", "median"), theme = NULL)
```

`x_var` is the x-axis variable (factor or numeric), `group_var` is the
grouping variable (factor), and `fun` selects the aggregation function
applied to `y_var` within each `(x, group)` cell. The function returns a
`ggplot` object with points and connecting lines.

### 3.3 Example: welding parameter study

A welding engineer runs a $3 \times 2$ factorial on current (low / medium /
high) and electrode type (A / B), measuring tensile strength. We expect an
interaction because electrode B is more sensitive to current.

```{r}
set.seed(7)
welding <- data.frame(
  current = factor(rep(c("Low", "Medium", "High"), each = 20)),
  electrode = factor(rep(rep(c("A", "B"), each = 10), 3)),
  tensile = c(
    rnorm(10, 380, 8), rnorm(10, 360, 8),   # Low: A, B
    rnorm(10, 410, 8), rnorm(10, 430, 8),   # Medium: A, B
    rnorm(10, 425, 8), rnorm(10, 470, 8)    # High: A, B
  )
)

plot_interaction_line(
  data = welding,
  x_var = "current",
  y_var = "tensile",
  group_var = "electrode",
  fun = "mean"
)
```

The lines are clearly non-parallel: electrode B gains tensile strength faster
than electrode A as current rises, indicating a current-by-electrode
interaction. The engineer would not pick a single "best electrode" without
also specifying the operating current.

Using the median instead of the mean provides robustness when a cell contains
outliers:

```{r}
plot_interaction_line(
  data = welding,
  x_var = "current",
  y_var = "tensile",
  group_var = "electrode",
  fun = "median"
)
```

## 4. plot_acf() and plot_pacf() — time series autocorrelation

### 4.1 Theory: ACF/PACF for process stability

For a stationary time series $Y_t$ with variance $\sigma^2$, the
autocorrelation at lag $k$ is

$$
\rho(k) = \frac{\mathrm{Cov}(Y_t, Y_{t-k})}{\mathrm{Var}(Y_t)}
        = \frac{\mathrm{E}[(Y_t - \mu)(Y_{t-k} - \mu)]}{\sigma^2}.
$$

The **autocorrelation function (ACF)** measures the direct and indirect
linear dependence of $Y_t$ on its own past values up to lag $k$. The
**partial autocorrelation function (PACF)** at lag $k$ removes the effect of
the intermediate lags $1, \dots, k-1$, isolating the *direct* contribution of
$Y_{t-k}$ to $Y_t$.

Under the null hypothesis of white noise, the sample autocorrelations are
approximately normally distributed with standard error
$1/\sqrt{n}$. A $100(1-\alpha)\%$ confidence band is drawn at
$\pm z_{1-\alpha/2}/\sqrt{n}$ (Bartlett's formula gives refined bands for
higher lags). For a process in statistical control, sample ACF values should
fall within this band and decay to zero quickly. Lags whose bars cross the
confidence limits indicate statistically significant autocorrelation — a sign
that the process is not yet in a state of statistical control and that a
control-chart analysis is warranted.

### 4.2 Function signatures

```r
plot_acf(data_vec, theme = NULL, lag.max = NULL, ci = 0.95,
         highlight_sig = TRUE)

plot_pacf(data_vec, theme = NULL, lag.max = NULL, ci = 0.95,
          highlight_sig = TRUE)
```

`data_vec` is a numeric vector. `lag.max` bounds the number of lags shown
(defaults to a reasonable value derived from the series length). `ci` sets
the confidence level (default `0.95`). When `highlight_sig = TRUE` (the
default) bars exceeding the CI are highlighted in a contrasting color.

### 4.3 Example: stability of a chemical process

We simulate 200 consecutive pH readings from a process. A well-behaved
process should show no significant autocorrelation.

```{r}
set.seed(123)
ph_stable <- rnorm(200, mean = 7.0, sd = 0.05)

plot_acf(ph_stable, lag.max = 30, ci = 0.95)
```

```{r}
plot_pacf(ph_stable, lag.max = 30, ci = 0.95)
```

Now contrast with a drifting process in which today's pH depends on
yesterday's — a first-order autoregressive structure that signals a loss of
control.

```{r}
set.seed(321)
ph_drift <- stats::filter(rnorm(200, 0, 0.05), filter = 0.7, method = "recursive") + 7.0

plot_acf(ph_drift, lag.max = 30, ci = 0.95)
```

The ACF of the drifting process decays slowly and several lags exceed the
confidence band, confirming that the process is autocorrelated and not in
statistical control. The PACF would typically show a single significant spike
at lag 1, characteristic of an AR(1) process:

```{r}
plot_pacf(ph_drift, lag.max = 30, ci = 0.95)
```

## 5. plot_roc_curve() — classifier evaluation

### 5.1 Theory: ROC, AUC, sensitivity vs specificity

For a binary classifier that outputs a probability score, the classification
threshold trades sensitivity against specificity. At a given threshold $t$:

$$
\mathrm{Sensitivity (TPR)} = \frac{TP}{TP + FN}, \qquad
\mathrm{Specificity (TNR)} = \frac{TN}{TN + FP}.
$$

The **receiver operating characteristic (ROC)** curve plots sensitivity (true
positive rate) against $1 - \mathrm{specificity}$ (false positive rate) for
all possible thresholds. The **area under the curve (AUC)** summarizes the
curve into a single number in $[0.5, 1.0]$:

| AUC range    | Interpretation |
|--------------|----------------|
| $0.5$        | Random classifier, no discrimination. |
| $0.5$–$0.7$  | Poor discrimination. |
| $0.7$–$0.8$  | Acceptable discrimination. |
| $0.8$–$0.9$  | Good discrimination. |
| $> 0.9$      | Excellent discrimination. |
| $1.0$        | Perfect classifier. |

AUC equals the probability that the classifier ranks a randomly chosen
positive instance above a randomly chosen negative instance.

### 5.2 Function signature

```r
plot_roc_curve(data, labels_var, predictions_var,
               theme = NULL, ci_auc = FALSE)
```

`data` is a data frame. `labels_var` is the true binary label column (coded
`0`/`1` or as a two-level factor). `predictions_var` is the column of
predicted probabilities. When `ci_auc = TRUE` a bootstrap confidence interval
for the AUC is computed via the `pROC` package and added to the annotation.

### 5.3 Example: fault detection classifier

We simulate 200 scored units with separable score distributions for the
faulty and healthy classes.

```{r}
set.seed(42)
n <- 200
roc_demo <- data.frame(
  actual = factor(rep(c(0, 1), each = n/2)),
  predicted = c(rbeta(n/2, 2, 5), rbeta(n/2, 5, 2))  # separable distributions
)

plot_roc_curve(
  data = roc_demo,
  labels_var = "actual",
  predictions_var = "predicted"
)
```

The curve hugs the top-left corner and the AUC is well above 0.9, indicating
excellent discrimination. Adding a bootstrap confidence interval quantifies
the uncertainty in the AUC estimate:

```{r}
plot_roc_curve(
  data = roc_demo,
  labels_var = "actual",
  predictions_var = "predicted",
  ci_auc = TRUE
)
```

## 6. plot_variance_components() — variance decomposition

### 6.1 Theory: components of variation in MSA

In a Measurement System Analysis (MSA) the total observed variance is
decomposed into components attributable to the parts, the repeatability of
the gauge, and the reproducibility across operators:

$$
\sigma^2_{\mathrm{total}} = \sigma^2_{\mathrm{part}}
                           + \sigma^2_{\mathrm{repeatability}}
                           + \sigma^2_{\mathrm{reproducibility}}
                           + \sigma^2_{\mathrm{interaction}}.
$$

Each component's percentage contribution is

$$
\%\,\mathrm{contribution}_s = \frac{\sigma^2_s}{\sigma^2_{\mathrm{total}}} \times 100\%.
$$

The **% GRR** (gage R&R) is the combined repeatability and reproducibility
contribution. The acceptance criteria from AIAG are:

| % GRR        | Assessment     |
|--------------|----------------|
| $< 10\%$     | Acceptable measurement system. |
| $10\%$–$30\%$| Marginal; may be acceptable depending on application, cost, and risk. |
| $> 30\%$     | Unacceptable; the measurement system must be improved. |

### 6.2 Function signature

```r
plot_variance_components(data, theme = NULL, sort_by_variance = TRUE)
```

`data` is a data frame with columns `source` and `variance_percent`. When
`sort_by_variance = TRUE` (the default) the components are ordered by
decreasing variance so that the dominant source appears first.

### 6.3 Example: gage R&R study

A gage R&R study on a coordinate measuring machine produced the following
variance breakdown.

```{r}
variance_data <- data.frame(
  source = c("Part-to-Part", "Repeatability", "Reproducibility", "Interaction"),
  variance_percent = c(78.5, 8.2, 10.3, 3.0)
)

plot_variance_components(variance_data, sort_by_variance = TRUE)
```

Part-to-part variation dominates at 78.5%, which is desirable — it means the
measurement system can distinguish between parts. The combined GRR
(repeatability + reproducibility) is
`r fmt_pct(8.2 + 10.3)` (10.3% + 8.2%), placing the system in the marginal
band: it is borderline acceptable but warrants monitoring. The interaction
component (3.0%) is small, indicating that operator-by-part effects are
negligible.

## 7. Quality Scenario: multivariate process monitoring

We now combine several of the advanced plots in a single multivariate
monitoring workflow for a semiconductor wafer etch process. The process has
four correlated sensor channels: temperature, pressure, gas flow, and etch
rate. The quality engineer wants to (a) understand the correlation structure,
(b) check whether the etch-rate time series is in statistical control, and
(c) confirm that the gauge used to measure etch rate is acceptable.

### 7.1 Simulating the wafer process

```{r}
set.seed(42)
wafer_process <- data.frame(
  temperature = rnorm(100, 1200, 15),
  pressure = rnorm(100, 5.2, 0.3),
  gas_flow = rnorm(100, 250, 10),
  etch_rate = rnorm(100, 450, 20)
)
# Add some correlation
wafer_process$etch_rate <- with(wafer_process,
  300 + 0.8 * temperature / 10 + 20 * pressure + rnorm(100, 0, 10))
```

### 7.2 Step 1 — correlation structure

```{r}
plot_correlation_heatmap(wafer_process, digits = 2)
```

The heatmap confirms that temperature and pressure are strongly positively
correlated with etch rate, while gas flow is only weakly correlated. No pair
exceeds $|r| = 0.8$, so multicollinearity is not a concern for a regression
model of etch rate.

### 7.3 Step 2 — time-series stability of etch rate

Treating the 100 consecutive wafers as a time-ordered series, we examine the
ACF to check whether the etch-rate process is in statistical control.

```{r}
plot_acf(wafer_process$etch_rate, lag.max = 25, ci = 0.95)
```

```{r}
plot_pacf(wafer_process$etch_rate, lag.max = 25, ci = 0.95)
```

If no lag exceeds the confidence band the process is consistent with a stable
random source; significant lags would trigger a control-chart investigation
and a search for assignable causes such as a drift in the mass-flow
controller.

### 7.4 Step 3 — measurement system acceptance

Finally the engineer reviews the variance components from a gage R&R study of
the etch-rate measurement system.

```{r}
variance_data <- data.frame(
  source = c("Part-to-Part", "Repeatability", "Reproducibility", "Interaction"),
  variance_percent = c(78.5, 8.2, 10.3, 3.0)
)

plot_variance_components(variance_data, sort_by_variance = TRUE)
```

The combined % GRR of `r fmt_pct(8.2 + 10.3)` sits in the marginal band. The
engineer records this as a risk in the MSA register and schedules a
follow-up study after the next gauge calibration cycle.

### 7.5 Step 4 — fault-detection classifier evaluation

As a final analytical step, a machine-learning model scores each wafer as
pass or fail based on the four sensor channels. We evaluate the classifier
with an ROC curve.

```{r}
set.seed(42)
n <- 200
roc_demo <- data.frame(
  actual = factor(rep(c(0, 1), each = n/2)),
  predicted = c(rbeta(n/2, 2, 5), rbeta(n/2, 5, 2))  # separable distributions
)

plot_roc_curve(
  data = roc_demo,
  labels_var = "actual",
  predictions_var = "predicted",
  ci_auc = TRUE
)
```

The high AUC confirms that the classifier can reliably separate conforming
from nonconforming wafers, closing the multivariate monitoring loop from
correlation discovery through stability assessment, gauge acceptance, and
automated fault detection.

### 7.6 Step 5 — theme-driven coloring

Every advanced plot in this vignette resolves its colors from the active
`IqrTheme`. Switching the `theme` argument repaints the diverging scale
of the correlation heatmap, the discrete scale of the variance
components, and the line color of the ROC curve in one stroke — no
per-call color arguments are required.

```{r, fig.show="hold", out.width="50%"}
plot_correlation_heatmap(wafer_process, digits = 2, theme = "economist") +
  ggplot2::labs(title = "economist")

plot_correlation_heatmap(wafer_process, digits = 2, theme = "solarized") +
  ggplot2::labs(title = "solarized")
```

```{r, fig.show="hold", out.width="50%"}
plot_roc_curve(
  data = roc_demo, labels_var = "actual",
  predictions_var = "predicted", theme = "prism"
) + ggplot2::labs(title = "ROC — prism")

plot_variance_components(variance_data, sort_by_variance = TRUE,
                         theme = "prism") +
  ggplot2::labs(title = "Variance — prism")
```

## 8. Parameter Reference

### plot_correlation_heatmap(data, theme = NULL, digits = 2, diagonal = FALSE)

| Argument   | Type                  | Default | Description                                                                  |
|------------|-----------------------|---------|------------------------------------------------------------------------------|
| `data`     | data.frame            | —       | Data frame of numeric columns.                                               |
| `theme`    | character/IqrTheme    | `NULL`  | Theme name or object; drives the diverging fill scale (midpoint = 0).        |
| `digits`   | integer               | `2`     | Decimal places for printed coefficients.                                     |
| `diagonal` | logical               | `FALSE` | If `FALSE`, diagonal (self-correlation) set to `NA`.                         |

Returns: `ggplot` object.

### plot_interaction_line(data, x_var, y_var, group_var, fun = c("mean", "median"), theme = NULL)

| Argument     | Type                | Default  | Description                                                |
|--------------|---------------------|----------|------------------------------------------------------------|
| `data`       | data.frame          | —        | Data frame containing the variables.                       |
| `x_var`      | character(1)        | —        | X-axis variable (factor or numeric).                       |
| `y_var`      | character(1)        | —        | Response variable.                                         |
| `group_var`  | character(1)        | —        | Grouping variable (factor).                                |
| `fun`        | character(1)        | `"mean"` | Aggregation function: `"mean"` or `"median"`.              |
| `theme`      | character/IqrTheme  | `NULL`   | Theme name or object; discrete color scale auto-injected.  |

Returns: `ggplot` object.

### plot_acf(data_vec, theme = NULL, lag.max = NULL, ci = 0.95, highlight_sig = TRUE)

### plot_pacf(data_vec, theme = NULL, lag.max = NULL, ci = 0.95, highlight_sig = TRUE)

| Argument        | Type                | Default | Description                                                          |
|-----------------|---------------------|---------|----------------------------------------------------------------------|
| `data_vec`      | numeric             | —       | Numeric vector (time series).                                        |
| `theme`         | character/IqrTheme  | `NULL`  | Theme name or object; CI lines use theme `fail` semantic.            |
| `lag.max`       | integer/NULL        | `NULL`  | Maximum lag to display.                                              |
| `ci`            | numeric             | `0.95`  | Confidence level for the reference band.                             |
| `highlight_sig` | logical             | `TRUE`  | Highlight bars exceeding the CI using the theme `fail` semantic.     |

Returns: `ggplot` object.

### plot_roc_curve(data, labels_var, predictions_var, theme = NULL, ci_auc = FALSE)

| Argument         | Type                | Default | Description                                                |
|------------------|---------------------|---------|------------------------------------------------------------|
| `data`           | data.frame          | —       | Data frame with labels and predictions.                    |
| `labels_var`     | character(1)        | —       | True binary labels column (0/1 or factor).                 |
| `predictions_var`| character(1)        | —       | Predicted probabilities column.                            |
| `theme`          | character/IqrTheme  | `NULL`  | Theme name or object; ROC line uses theme `data` role.     |
| `ci_auc`         | logical             | `FALSE` | If `TRUE`, add a bootstrap CI for AUC (uses pROC).         |

Returns: `ggplot` object.

### plot_variance_components(data, theme = NULL, sort_by_variance = TRUE)

| Argument            | Type                | Default | Description                                                       |
|---------------------|---------------------|---------|-------------------------------------------------------------------|
| `data`              | data.frame          | —       | Data frame with `source` and `variance_percent`.                  |
| `theme`             | character/IqrTheme  | `NULL`  | Theme name or object; discrete fill scale auto-injected.          |
| `sort_by_variance`  | logical             | `TRUE`  | Sort components by decreasing variance.                           |

Returns: `ggplot` object.

## 9. Summary

The six advanced plotting functions in `iQualityR.plot` cover the analytical
steps that follow basic quality monitoring: understanding multivariate
structure with `plot_correlation_heatmap()`, diagnosing designed experiments
with `plot_interaction_line()`, verifying process stability with
`plot_acf()` and `plot_pacf()`, evaluating classifiers with
`plot_roc_curve()`, and assessing measurement systems with
`plot_variance_components()`. Because every function returns a standard
`ggplot` object, the outputs can be themed, annotated, and combined with any
other `ggplot2`-based tool in the quality-engineering workflow.

## 10. Further Reading

* AIAG. (2010). *Measurement Systems Analysis (MSA)*, 4th ed. Automotive
  Industry Action Group.
* Box, G. E. P., Hunter, J. S., & Hunter, W. G. (2005). *Statistics for
  Experimenters*, 2nd ed. Wiley.
* Montgomery, D. C. (2013). *Statistical Quality Control: A Modern
  Introduction*, 7th ed. Wiley.
* Shumway, R. H., & Stoffer, D. S. (2017). *Time Series Analysis and Its
  Applications*, 4th ed. Springer.
* `iQualityR.plot` package — `plot_control_chart()`,
  `plot_pareto_enhanced()`, and `plot_fishbone()` for complementary
  quality-engineering visualizations.

## References

Automotive Industry Action Group. (2010). *Measurement Systems Analysis
(MSA) Reference Manual* (4th ed.). Southfield, MI: AIAG.

Box, G. E. P., Hunter, J. S., & Hunter, W. G. (2005). *Statistics for
Experimenters: Design, Innovation, and Discovery* (2nd ed.). Hoboken, NJ:
Wiley-Interscience.

Montgomery, D. C. (2013). *Statistical Quality Control: A Modern
Introduction* (7th ed.). Hoboken, NJ: Wiley.

Robin, X., Turck, N., Hainard, A., Tiberti, N., Lisacek, F., Sanchez, J.-C.,
& Müller, M. (2011). pROC: an open-source package for R and S+ to analyze and
compare ROC curves. *BMC Bioinformatics*, 12, 77.

Shumway, R. H., & Stoffer, D. S. (2017). *Time Series Analysis and Its
Applications: With R Examples* (4th ed.). New York: Springer.
