---
title: "Advanced Analysis Plots"
author: "iQualityR Team"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Advanced Analysis Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

# Advanced Analysis Plots

## Overview

In the in-depth analysis phase of quality management, it is necessary to explore complex relationships between variables, time series characteristics, and more. `iQualityR.plot` provides multiple advanced analysis plots:

- `plot_correlation_heatmap()` - Correlation heatmap (variable correlation visualization)
- `plot_interaction_line()` - Interaction plot (factor interaction effects)
- `plot_acf()` - Autocorrelation function plot (time series analysis)
- `plot_pacf()` - Partial autocorrelation function plot (time series analysis)

## Load Data

```{r setup}
library(ggplot2)
library(datasets)
library(iQualityR.plot)

data("mtcars")
data("airquality")
data("Nile")
```

## 1. Correlation Heatmap: plot_correlation_heatmap()

The correlation heatmap intuitively displays the correlation coefficient matrix between multiple variables through color encoding, and is an important tool for exploratory data analysis.

### Basic Usage

```{r heatmap_basic}
plot_correlation_heatmap(
  data = mtcars
)
```

### Custom Display Options

```{r heatmap_custom}
plot_correlation_heatmap(
  data = mtcars[, c("mpg", "cyl", "disp", "hp", "wt", "qsec")],
  digits = 2,
  diagonal = TRUE
)
```

### Interpreting the Correlation Heatmap

- **Color intensity**: Indicates the magnitude of the correlation coefficient
- **Red**: Positive correlation (one variable increases, the other also increases)
- **Blue**: Negative correlation (one variable increases, the other decreases)
- **White/light color**: Weak or no correlation
- **Numerical labels**: Exact correlation coefficient values

### Practical Application: Multicollinearity Detection

```{r heatmap_multicollinearity}
cor_vars <- mtcars[, c("mpg", "cyl", "disp", "hp", "wt")]

plot_correlation_heatmap(
  data = cor_vars,
  digits = 2
)
```

## 2. Interaction Plot: plot_interaction_line()

The interaction plot is used to display the interaction effect of two factor variables on a response variable, and is a core tool in DOE (Design of Experiments) analysis.

### Basic Usage

```{r interaction_basic, eval=TRUE}
data("mtcars")

mtcars$cyl_factor <- factor(mtcars$cyl)
mtcars$am_factor <- factor(mtcars$am, labels = c("Auto", "Manual"))

plot_interaction_line(
  data = mtcars,
  x_var = "cyl_factor",
  y_var = "mpg",
  group_var = "am_factor",
  fun = "mean"
)
```

### Using Median Aggregation

```{r interaction_median, eval=TRUE}
plot_interaction_line(
  data = mtcars,
  x_var = "cyl_factor",
  y_var = "mpg",
  group_var = "am_factor",
  fun = "median"
)
```

### Interpreting the Interaction Plot

- **Parallel lines**: No interaction (two factors independently affect the response variable)
- **Crossing lines**: Interaction exists (the effect of one factor depends on the level of the other factor)
- **Line slope**: Indicates the magnitude of the factor effect

### Practical Application: Factorial Experiment Analysis

```{r interaction_doe, eval=TRUE}
set.seed(123)
doe_data <- expand.grid(
  Temperature = c("Low", "High"),
  Pressure = c("Low", "High")
)
doe_data$Yield <- c(85, 92, 78, 95) + rnorm(4, 0, 2)

doe_data <- doe_data[rep(1:4, each = 3), ]
doe_data$Yield <- doe_data$Yield + rnorm(12, 0, 1)

plot_interaction_line(
  data = doe_data,
  x_var = "Temperature",
  y_var = "Yield",
  group_var = "Pressure",
  fun = "mean"
)
```

## 3. Autocorrelation Function Plot: plot_acf()

The Autocorrelation Function (ACF) plot is used to detect autocorrelation in time series data, and is an important tool in SPC and time series analysis.

### Basic Usage

```{r acf_basic}
plot_acf(
  data_vec = Nile
)
```

### Custom Options

```{r acf_custom}
plot_acf(
  data_vec = Nile,
  lag.max = 20,
  ci = 0.95,
  highlight_sig = TRUE
)
```

### Interpreting the ACF Plot

- **Vertical line height**: Indicates the autocorrelation coefficient at each lag order
- **Blue dashed lines**: Confidence interval boundaries (typically 95%)
- **Lines exceeding the dashed lines**: Statistically significant autocorrelation
- **Slow decay**: A trend may exist
- **Periodic fluctuation**: Seasonality may exist

## 4. Partial Autocorrelation Function Plot: plot_pacf()

The Partial Autocorrelation Function (PACF) plot is used to identify the autoregressive order of a time series, and is used in conjunction with the ACF plot.

### Basic Usage

```{r pacf_basic}
plot_pacf(
  data_vec = Nile
)
```

### Custom Options

```{r pacf_custom}
plot_pacf(
  data_vec = Nile,
  lag.max = 15,
  ci = 0.95,
  highlight_sig = TRUE
)
```

### Interpreting the PACF Plot

- **Cutoff**: The autocorrelation coefficient suddenly becomes zero after a certain order, indicating the order of an AR process
- **Tailing**: The autocorrelation coefficient gradually decays, indicating an MA process
- **Combined ACF judgment for model type**:
  - ACF tailing, PACF cutoff -> AR model
  - ACF cutoff, PACF tailing -> MA model
  - Both tailing -> ARMA model

## Practical Case Studies

### Case 1: Automobile Performance Correlation Analysis

```{r case_correlation}
car_vars <- mtcars[, c("mpg", "cyl", "disp", "hp", "drat", "wt", "qsec")]

plot_correlation_heatmap(
  data = car_vars,
  digits = 2,
  diagonal = FALSE
)
```

### Case 2: Air Quality Time Series Analysis

```{r case_time_series}
ozone_data <- na.omit(airquality$Ozone)

plot_acf(
  data_vec = ozone_data,
  lag.max = 30,
  ci = 0.95,
  highlight_sig = TRUE
)
```

### Case 3: Multi-Variable Interaction Analysis

```{r case_interaction, eval=FALSE}
data("mtcars")

mtcars$vs_factor <- factor(mtcars$vs, labels = c("V-shaped", "Straight"))
mtcars$am_factor <- factor(mtcars$am, labels = c("Auto", "Manual"))

plot_interaction_line(
  data = mtcars,
  x_var = "vs_factor",
  y_var = "mpg",
  group_var = "am_factor",
  fun = "mean"
)
```

## Parameter Details

### plot_correlation_heatmap() Parameters

| Parameter | Type | Description | Default |
|------|------|------|--------|
| `data` | data.frame | Data frame | Required |
| `theme` | theme | Theme specification | NULL |
| `digits` | numeric | Number of decimal places for correlation coefficients | 2 |
| `diagonal` | logical | Whether to show diagonal | FALSE |

### plot_interaction_line() Parameters

| Parameter | Type | Description | Default |
|------|------|------|--------|
| `data` | data.frame | Data frame | Required |
| `x_var` | character | X-axis variable | Required |
| `y_var` | character | Y-axis variable | Required |
| `group_var` | character | Grouping variable | Required |
| `fun` | character | Aggregation function | "mean" |
| `theme` | theme | Theme specification | NULL |

### plot_acf() Parameters

| Parameter | Type | Description | Default |
|------|------|------|--------|
| `data_vec` | numeric | Time series vector | Required |
| `theme` | theme | Theme specification | NULL |
| `lag.max` | numeric | Maximum lag order | NULL (auto) |
| `ci` | numeric | Confidence level | 0.95 |
| `highlight_sig` | logical | Highlight significant correlations | TRUE |

### plot_pacf() Parameters

| Parameter | Type | Description | Default |
|------|------|------|--------|
| `data_vec` | numeric | Time series vector | Required |
| `theme` | theme | Theme specification | NULL |
| `lag.max` | numeric | Maximum lag order | NULL (auto) |
| `ci` | numeric | Confidence level | 0.95 |
| `highlight_sig` | logical | Highlight significant correlations | TRUE |

## Best Practices

### 1. Correlation Analysis Workflow

```{r best_practice_correlation, eval=FALSE}
plot_correlation_heatmap(data)

cor_matrix <- cor(data)
high_cor <- which(abs(cor_matrix) > 0.8 & abs(cor_matrix) < 1, arr.ind = TRUE)
```

### 2. Interaction Analysis Considerations

```{r best_practice_interaction, eval=FALSE}
# - Ensure grouping variables are factor types
# - Each cell should have sufficient observations
# - Using means is sensitive to outliers; consider medians
# - When interaction is significant, interpret main effects at each level separately
```

### 3. Time Series Analysis Recommendations

```{r best_practice_timeseries, eval=FALSE}
# - First plot the time series to observe trends and seasonality
# - Use ACF and PACF together to determine model type
# - Pay attention to data stationarity; non-stationary data requires differencing
# - Sample size should be sufficient (at least 50 observations)
```

## FAQ

### Q: The correlation heatmap colors are not intuitive?

You can adjust the color scheme through theme parameters:

```{r faq_heatmap_color, eval=FALSE}
plot_correlation_heatmap(data, theme = "tech")
```

### Q: What does crossing lines in the interaction plot mean?

Crossing lines indicate an interaction, meaning the effect of one factor on the response variable depends on the level of the other factor. In this case, an analysis of variance (ANOVA) should be performed to test whether the interaction is statistically significant.

### Q: What if all lags in the ACF plot are significant?

This may indicate a trend in the time series, requiring differencing to make it stationary:

```{r faq_acf_trend, eval=FALSE}
diff_data <- diff(time_series)
plot_acf(diff_data)
```

## Summary

- `plot_correlation_heatmap()` intuitively displays the correlation structure between variables
- `plot_interaction_line()` is a core tool for DOE analysis, used to identify factor interaction effects
- `plot_acf()` and `plot_pacf()` are fundamental tools for time series analysis
- These advanced analysis plots are suitable for in-depth quality data analysis scenarios

## Further Reading

- [Quality Tools Charts](quality_tools.html) - Learn about PP/QQ/Turtle diagrams
- [Scatter Plots](scatter_plots.html) - Learn about basic scatter plots
- [Layer Builders](layer_builders.html) - Learn about low-level layer functions
