---
title: "ANOVA Visualization"
author: "iQualityR.plot Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{ANOVA Visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

library(iQualityR.plot)
library(ggplot2)

fmt_num <- function(x, digits = 4) formatC(x, format = "f", digits = digits)
fmt_pct <- function(x, digits = 2) paste0(formatC(x, format = "f", digits = digits), "%")
```

## 1. Introduction

Analysis of variance (ANOVA) is the workhorse for comparing the means of two or
more groups. In quality engineering it shows up whenever we ask: *do these
suppliers, machines, operators, or batches differ?* The global F-test answers
the overall question; main-effects plots, pairwise comparisons, and residual
diagnostics answer the follow-up questions that the F-test cannot.

The `iQualityR.plot` package provides a family of plotters that turn an ANOVA
fit into a complete visual report:

- `plot_f_curve()` — the global F-test on the F distribution.
- `plot_anova_effects()` — factor level means with standard error bars.
- `plot_anova_comparison()` — forest plot of pairwise comparisons.
- `plot_anova_residuals()` — residual diagnostic panel.
- `plot_anova_summary()` — a dashboard combining effects, residuals, and table.
- `plot_anova_diagnostic()` — assumption checks for an ANOVA object.
- `create_anova_table()` — a table grob for embedding in reports.

This vignette walks through the theory behind each plot, demonstrates the
function signatures on a real three-supplier tensile-strength comparison, and
finishes with a parameter reference.

## 2. Theory: Visualizing ANOVA Results

### 2.1 The F-distribution and rejection region

Under $H_0: \mu_1 = \mu_2 = \dots = \mu_k$, the ratio of between-group to
within-group mean squares follows an F distribution:

$$
F = \frac{\text{MSB}}{\text{MSW}} \sim F_{k-1,\,N-k}.
$$

The F distribution is **right-skewed** and only takes non-negative values, so
the rejection region lies entirely in the **right tail**:

$$
\mathcal{R} = \left\{ F : F > F_{\alpha,\,k-1,\,N-k} \right\}.
$$

Here $k$ is the number of groups, $N$ is the total sample size, $\text{df}_1 =
k-1$ is the numerator degrees of freedom, and $\text{df}_2 = N-k$ is the
denominator degrees of freedom. A large F means the between-group variation
dominates the within-group variation, which is evidence that the group means
differ.

### 2.2 Main effects and interactions

A **main effect** is the effect of a single factor on the response, averaging
over the levels of all other factors. Visually it is the spread of the group
means around the grand mean.

An **interaction** occurs when the effect of one factor depends on the level of
another. In a two-way ANOVA the interaction is visualized as **non-parallel
lines** in an interaction plot: if the lines cross or fan out, the factors
interact and the main effects alone are misleading.

### 2.3 Residual diagnostics for ANOVA assumptions

ANOVA rests on three assumptions:

1. **Normality of residuals.** The residuals should be approximately normal.
   Check with a Normal Q-Q plot; points should hug the reference line.
2. **Homogeneity of variance.** The residual spread should be similar across
   fitted values. Check with a Scale-Location plot; the loess trend should be
   roughly flat.
3. **Independence.** Observations should be independent. Check with the study
   design; residuals vs order can reveal time trends.

The residual diagnostic panel plots Residuals vs Fitted, Normal Q-Q,
Scale-Location, and Residuals vs Factor so all three assumptions can be audited
at a glance.

### Tukey HSD for pairwise comparisons

When the global F-test rejects $H_0$, the natural follow-up is *which* groups
differ. Tukey's Honestly Significant Difference controls the family-wise error
rate across all pairwise comparisons:

$$
\text{HSD} = q_{\alpha,\,k,\,N-k} \sqrt{\frac{\text{MSW}}{n}},
$$

where $q_{\alpha,\,k,\,N-k}$ is the studentized range critical value and $n$ is
the common group size. The forest-style comparison plot shows each contrast's
estimate and confidence interval; intervals crossing zero indicate no
significant difference.

## 3. plot_f_curve() — global F-test visualization

`plot_f_curve()` draws the F distribution with `df1` and `df2` degrees of
freedom, shades the right-tail rejection region in red, and drops a blue
vertical line at the observed F-statistic. As with the hypothesis curve,
**blue in red — reject $H_0$**.

### Function signature

```{r, eval = FALSE}
plot_f_curve(
  f_stat,
  df1,
  df2,
  alpha = 0.05,
  p_value = NULL,
  title = NULL,
  theme = NULL,
  ...
)
```

### Example

Suppose an ANOVA produced $F = 4.87$ with $\text{df}_1 = 2$ and $\text{df}_2 =
42$ at $\alpha = 0.05$.

```{r}
plot_f_curve(
  f_stat = 4.87,
  df1 = 2,
  df2 = 42,
  alpha = 0.05,
  p_value = 0.0123
)
```

The blue line at $F = 4.87$ falls inside the red-shaded right tail, so we
reject $H_0$: at least one group mean differs. The p-value annotation
corroborates the graphical verdict.

## 4. plot_anova_effects() — main effects with SE bars

`plot_anova_effects()` plots the factor level means with standard error bars.
It can optionally add a compact letter display (groups sharing a letter are not
significantly different) and a summary table below the plot.

### Function signature

```{r, eval = FALSE}
plot_anova_effects(
  data = NULL,
  response_col = NULL,
  factor_col = NULL,
  means = NULL,
  show_letters = FALSE,
  letters = NULL,
  show_table = FALSE,
  point_color = NULL,
  error_color = NULL,
  theme = NULL,
  title = NULL,
  ...
)
```

### Example from data

```{r}
set.seed(42)
tensile <- data.frame(
  supplier = factor(rep(c("A", "B", "C"), each = 15)),
  strength = c(rnorm(15, 250, 8), rnorm(15, 265, 8), rnorm(15, 255, 8))
)

plot_anova_effects(
  data = tensile,
  response_col = "strength",
  factor_col = "supplier",
  show_table = TRUE
)
```

### With compact letter display

```{r}
# Pre-computed compact letters (in practice from multcomp::cld or similar)
cld_df <- data.frame(
  group = c("A", "B", "C"),
  letter = c("a", "b", "ab")
)

plot_anova_effects(
  data = tensile,
  response_col = "strength",
  factor_col = "supplier",
  show_letters = TRUE,
  letters = cld_df
)
```

### From a pre-computed means table

You can also supply a pre-computed means data frame with columns `group`,
`mean`, `se`, and `n`:

```{r}
means_df <- data.frame(
  group = c("A", "B", "C"),
  mean = c(250.2, 265.1, 255.4),
  se = c(2.1, 2.3, 2.0),
  n = c(15, 15, 15)
)

plot_anova_effects(means = means_df, show_table = TRUE)
```

## 5. plot_anova_comparison() — multiple comparison forest plot

`plot_anova_comparison()` turns a set of pairwise comparisons into a forest
plot. Each row is a contrast; the point is the estimated mean difference and
the bar is the confidence interval. Intervals crossing the reference line at
zero indicate no significant difference. Significant comparisons are
highlighted by default.

### Function signature

```{r, eval = FALSE}
plot_anova_comparison(
  comparison_data,
  estimate_col = "estimate",
  lower_col = "lower",
  upper_col = "upper",
  p_col = NULL,
  alpha = 0.05,
  sort_by_estimate = TRUE,
  reference_line = 0,
  point_color = NULL,
  error_color = NULL,
  highlight_sig = TRUE,
  theme = NULL,
  title = NULL,
  ...
)
```

### Example

```{r}
comparisons <- data.frame(
  contrast = c("B - A", "C - A", "C - B"),
  estimate = c(14.9, 5.2, -9.7),
  lower = c(8.1, -1.6, -16.5),
  upper = c(21.7, 12.0, -2.9),
  p_value = c(0.0003, 0.184, 0.007)
)

plot_anova_comparison(
  comparison_data = comparisons,
  p_col = "p_value",
  highlight_sig = TRUE
)
```

Supplier B is significantly higher than A (interval entirely above zero) and C
is significantly lower than B (interval entirely below zero), while C vs A is
not significant (interval crosses zero).

## 6. plot_anova_residuals() — residual diagnostic panel

`plot_anova_residuals()` produces a four-panel patchwork of standard residual
diagnostics: Residuals vs Fitted, Normal Q-Q, Scale-Location, and Residuals vs
Factor. These underpin the three ANOVA assumptions.

### Function signature

```{r, eval = FALSE}
plot_anova_residuals(
  model,
  factor_col = NULL,
  add_qq = TRUE,
  add_scale_location = TRUE,
  theme = NULL,
  title = NULL,
  ...
)
```

### Example

```{r, fig.height = 7}
set.seed(42)
tensile <- data.frame(
  supplier = factor(rep(c("A", "B", "C"), each = 15)),
  strength = c(rnorm(15, 250, 8), rnorm(15, 265, 8), rnorm(15, 255, 8))
)
model <- aov(strength ~ supplier, data = tensile)

plot_anova_residuals(model, factor_col = "supplier")
```

Read the panels as follows:

- **Residuals vs Fitted** — random scatter around zero with no pattern indicates
  linearity and roughly constant variance.
- **Normal Q-Q** — points following the reference line indicate normality.
- **Scale-Location** — a flat loess trend indicates homoscedasticity.
- **Residuals vs Factor** — similar spread across factor levels supports the
  equal-variance assumption.

## 7. plot_anova_summary() — ANOVA dashboard

`plot_anova_summary()` assembles a single dashboard that combines main effects,
an interaction panel (for multi-factor models), residual diagnostics, and the
ANOVA table. It is the one-stop view for a report or a slide.

### Function signature

```{r, eval = FALSE}
plot_anova_summary(
  model,
  factor_cols = NULL,
  show_interaction = TRUE,
  show_residuals = TRUE,
  show_table = TRUE,
  layout = "2x2",
  theme = NULL,
  title = NULL,
  ...
)
```

### Example

```{r, fig.height = 7}
set.seed(42)
tensile <- data.frame(
  supplier = factor(rep(c("A", "B", "C"), each = 15)),
  strength = c(rnorm(15, 250, 8), rnorm(15, 265, 8), rnorm(15, 255, 8))
)
model <- aov(strength ~ supplier, data = tensile)

plot_anova_summary(
  model,
  factor_cols = "supplier",
  show_interaction = FALSE,
  layout = "2x2"
)
```

For a one-factor model, set `show_interaction = FALSE` since there is no
interaction to display. The dashboard then shows main effects, residuals, and
the ANOVA table in a compact 2x2 layout.

## 8. plot_anova_diagnostic() — assumption checks

`plot_anova_diagnostic()` is a thin wrapper that routes an `iqr_anova` object
or a raw model object to the appropriate residual diagnostics. It is the
recommended entry point when you already have an iQualityR ANOVA object and
just want the assumption-check panel.

### Function signature

```{r, eval = FALSE}
plot_anova_diagnostic(x, ...)
```

### Example

```{r, fig.height = 7}
set.seed(42)
tensile <- data.frame(
  supplier = factor(rep(c("A", "B", "C"), each = 15)),
  strength = c(rnorm(15, 250, 8), rnorm(15, 265, 8), rnorm(15, 255, 8))
)
model <- aov(strength ~ supplier, data = tensile)

plot_anova_diagnostic(model)
```

## 9. create_anova_table() — table grob for reports

`create_anova_table()` converts an ANOVA table data frame (as produced by
`summary(aov(...))`) into a `tableGrob` from `gridExtra` so it can be embedded
in a patchwork composition or a knitr report. Significant p-values are
highlighted by default.

### Function signature

```{r, eval = FALSE}
create_anova_table(
  anova_table,
  theme = NULL,
  digits = 4,
  font_size = 9,
  highlight_sig = TRUE,
  sig_level = 0.05
)
```

### Example

```{r}
set.seed(42)
tensile <- data.frame(
  supplier = factor(rep(c("A", "B", "C"), each = 15)),
  strength = c(rnorm(15, 250, 8), rnorm(15, 265, 8), rnorm(15, 255, 8))
)
model <- aov(strength ~ supplier, data = tensile)

# Capture the ANOVA table as a data frame
anova_df <- as.data.frame(summary(model)[[1]])
anova_df$`Pr(>F)` <- fmt_num(anova_df$`Pr(>F)`, 4)

tbl_grob <- create_anova_table(
  anova_table = anova_df,
  digits = 4,
  highlight_sig = TRUE,
  sig_level = 0.05
)

library(grid)
grid.newpage()
grid.draw(tbl_grob)
```

## 10. Quality Scenario: comparing three suppliers' tensile strength

A quality engineer is qualifying three potential suppliers of a polymer resin.
The key response is tensile strength (MPa). Fifteen specimens are pulled from
each supplier. The engineering question: *do the three suppliers differ in mean
tensile strength, and if so, which pair differs?*

### Step 1 — prepare the data

```{r}
set.seed(42)
tensile <- data.frame(
  supplier = factor(rep(c("A", "B", "C"), each = 15)),
  strength = c(rnorm(15, 250, 8), rnorm(15, 265, 8), rnorm(15, 255, 8))
)

# Group summaries
agg <- aggregate(strength ~ supplier, data = tensile,
                 FUN = function(x) c(mean = mean(x), sd = sd(x), n = length(x)))
agg <- do.call(data.frame, agg)
agg$se <- agg$strength.sd / sqrt(agg$strength.n)
print(agg)
```

### Step 2 — fit the ANOVA model

```{r}
model <- aov(strength ~ supplier, data = tensile)
summary(model)
```

### Step 3 — visualize the global F-test

```{r}
f_val <- summary(model)[[1]]$`F value`[1]
df1 <- summary(model)[[1]]$Df[1]
df2 <- summary(model)[[1]]$Df[2]
p_val <- summary(model)[[1]]$`Pr(>F)`[1]

plot_f_curve(
  f_stat = f_val,
  df1 = df1,
  df2 = df2,
  alpha = 0.05,
  p_value = p_val
)
```

### Step 4 — visualize the main effects

```{r}
plot_anova_effects(
  data = tensile,
  response_col = "strength",
  factor_col = "supplier",
  show_table = TRUE
)
```

### Step 5 — pairwise comparisons

```{r}
tukey <- TukeyHSD(model)
tukey_df <- as.data.frame(tukey$supplier)
tukey_df$contrast <- rownames(tukey_df)
colnames(tukey_df) <- c("estimate", "lower", "upper", "p_value", "contrast")

plot_anova_comparison(
  comparison_data = tukey_df,
  p_col = "p_value",
  highlight_sig = TRUE
)
```

### Step 6 — residual diagnostics

```{r, fig.height = 7}
plot_anova_residuals(model, factor_col = "supplier")
```

### Step 7 — full dashboard

```{r, fig.height = 7}
plot_anova_summary(
  model,
  factor_cols = "supplier",
  show_interaction = FALSE,
  layout = "2x2"
)
```

### Step 8 — theme-driven coloring

Every ANOVA plot in this vignette resolves its colors from the active
`IqrTheme`. The rejection ribbon of `plot_f_curve()` uses the theme
`fail` semantic; the F-statistic line uses the theme `primary` semantic;
the mean points and error bars of `plot_anova_effects()` use the theme
`data` and `muted` roles. Switching `theme` therefore repaints the whole
report without per-call color arguments:

```{r, fig.show="hold", out.width="50%"}
plot_f_curve(
  f_stat = f_val, df1 = df1, df2 = df2,
  alpha = 0.05, p_value = p_val, theme = "economist"
) + ggplot2::labs(title = "F-curve — economist")

plot_anova_effects(
  data = tensile, response_col = "strength",
  factor_col = "supplier", theme = "prism"
) + ggplot2::labs(title = "Effects — prism")
```

To override an individual role, pull the relevant color from the
`IqrTheme` object retrieved via `attr(p, "iqr_theme")`:

```{r}
p_seed <- base_plot(tensile, ggplot2::aes(x = .data$supplier,
                                          y = .data$strength),
                    theme = "solarized")
c_sol <- attr(p_seed, "iqr_theme")

plot_anova_effects(
  data        = tensile,
  response_col = "strength",
  factor_col   = "supplier",
  theme        = "solarized",
  point_color  = c_sol$get_pal("semantic", name = "pass"),
  error_color  = c_sol$get_pal("semantic", name = "watch")
) + ggplot2::labs(title = "Effects — solarized with semantic overrides")
```

### Conclusion

The global F-test rejects $H_0$ ($p < 0.05$), so at least one supplier differs
in mean tensile strength. The main-effects plot shows supplier B is highest,
followed by C, then A. The Tukey comparison plot confirms that B is
significantly higher than both A and C, while C vs A is not significant at the
0.05 level. Residual diagnostics show no serious violations of normality or
equal variance, so the ANOVA conclusions are trustworthy. **Recommendation:
prefer supplier B for tensile-strength-critical applications; suppliers A and C
are statistically comparable on this metric.**

## 11. Parameter Reference

### plot_f_curve()

| Argument | Type | Description |
|----------|------|-------------|
| `f_stat` | numeric | Observed F-statistic. |
| `df1` | numeric | Numerator degrees of freedom ($k-1$). |
| `df2` | numeric | Denominator degrees of freedom ($N-k$). |
| `alpha` | numeric | Significance level (default 0.05). |
| `p_value` | numeric | Optional p-value, displayed on the plot. |
| `title` | character | Optional plot title. |
| `theme` | character/IqrTheme | Theme name or object; rejection ribbon uses `fail` semantic, F-line uses `primary`. |
| `...` |  | Additional arguments passed to ggplot layers. |

### plot_anova_effects()

| Argument | Type | Description |
|----------|------|-------------|
| `data` | data.frame | Long-format data with response and factor columns. |
| `response_col` | character | Name of the response column. |
| `factor_col` | character | Name of the factor column. |
| `means` | data.frame | Optional pre-computed means (`group`, `mean`, `se`, `n`). |
| `show_letters` | logical | If TRUE, add compact letter display. |
| `letters` | data.frame | Letters data frame (`group`, `letter`). |
| `show_table` | logical | If TRUE, add summary table below plot. |
| `point_color` | character | Point color; `NULL` resolves to theme `data` role. |
| `error_color` | character | Error bar color; `NULL` resolves to theme `muted` role. |
| `theme` | character/IqrTheme | Theme name or object; propagates to `base_plot()` and `.iqr_aes()`. |
| `title` | character | Optional plot title. |

### plot_anova_comparison()

| Argument | Type | Description |
|----------|------|-------------|
| `comparison_data` | data.frame | Pairwise comparisons with `contrast`, `estimate`, `lower`, `upper`. |
| `estimate_col` | character | Column name for estimates (default `"estimate"`). |
| `lower_col` | character | Column name for lower CI bound. |
| `upper_col` | character | Column name for upper CI bound. |
| `p_col` | character | Optional column name for p-values. |
| `alpha` | numeric | Significance level (default 0.05). |
| `sort_by_estimate` | logical | If TRUE, sort rows by estimate. |
| `reference_line` | numeric | x-axis reference line (default 0). |
| `highlight_sig` | logical | If TRUE, color significant comparisons using the theme `fail` semantic. |
| `point_color` | character | Point color; `NULL` resolves to theme `data` role. |
| `error_color` | character | Error bar color; `NULL` resolves to theme `muted` role. |
| `theme` | character/IqrTheme | Theme name or object for color resolution. |
| `title` | character | Optional plot title. |

### plot_anova_residuals()

| Argument | Type | Description |
|----------|------|-------------|
| `model` | lm/aov | Fitted model object. |
| `factor_col` | character | Optional factor column for the vs-factor panel. |
| `add_qq` | logical | If TRUE, include the Normal Q-Q panel. |
| `add_scale_location` | logical | If TRUE, include the Scale-Location panel. |
| `theme` | character/IqrTheme | Theme name or object applied to every diagnostic panel. |
| `title` | character | Optional plot title. |

### plot_anova_summary()

| Argument | Type | Description |
|----------|------|-------------|
| `model` | lm/aov | Fitted model object. |
| `factor_cols` | character vector | Factor names to display. |
| `show_interaction` | logical | If TRUE, include interaction panel. |
| `show_residuals` | logical | If TRUE, include residual diagnostics. |
| `show_table` | logical | If TRUE, include the ANOVA table. |
| `layout` | character | `"2x2"` or `"1x3"`. |
| `theme` | character/IqrTheme | Theme name or object propagated to every sub-plot. |
| `title` | character | Optional plot title. |

### plot_anova_diagnostic()

| Argument | Type | Description |
|----------|------|-------------|
| `x` | object | An `iqr_anova` object or an `lm`/`aov` model object. |
| `...` |  | Additional arguments passed to the underlying diagnostics. |

### create_anova_table()

| Argument | Type | Description |
|----------|------|-------------|
| `anova_table` | data.frame | ANOVA table from `summary(aov(...))`. |
| `theme` | character/IqrTheme | Theme name or object for table styling. |
| `digits` | integer | Number of digits to display. |
| `font_size` | numeric | Font size for the table. |
| `highlight_sig` | logical | If TRUE, highlight significant p-values. |
| `sig_level` | numeric | Significance threshold for highlighting. |

## 12. Summary

The ANOVA visualization family in `iQualityR.plot` covers the full lifecycle of
an ANOVA analysis:

- **`plot_f_curve()`** visualizes the global F-test against the F distribution.
- **`plot_anova_effects()`** shows factor level means with SE bars and optional
  compact letter display.
- **`plot_anova_comparison()`** renders pairwise comparisons as a forest plot.
- **`plot_anova_residuals()`** and **`plot_anova_diagnostic()`** audit the
  normality, equal-variance, and independence assumptions.
- **`plot_anova_summary()`** assembles everything into a single dashboard.
- **`create_anova_table()`** produces a report-ready table grob.

Used together, they let a quality engineer move from raw data to a defensible,
communicable conclusion about supplier or process differences in a single
analysis pipeline.

## 13. Further Reading

- The `iQualityR.plot` package vignettes on **Hypothesis Testing Visualization**
  and **MSA (Measurement System Analysis)** for related visualization tools.
- Montgomery, D. C. (2019). *Introduction to Statistical Quality Control*,
  8th ed. Wiley. — Chapters on ANOVA, random effects, and Gauge R&R.
- NIST/SEMATECH e-Handbook of Statistical Methods,
  <https://www.itl.nist.gov/div898/handbook/> — sections on ANOVA and Tukey's
  method.
- `?stats::aov`, `?stats::TukeyHSD`, and `?stats::lm` for the underlying R
  machinery that produces the fits visualized here.

## References

Montgomery, D. C. (2019). *Introduction to Statistical Quality Control* (8th
ed.). John Wiley & Sons.

NIST/SEMATECH. (2024). *e-Handbook of Statistical Methods*.
<https://www.itl.nist.gov/div898/handbook/>

R Core Team. (2024). *R: A Language and Environment for Statistical
Computing*. R Foundation for Statistical Computing, Vienna, Austria.
<https://www.R-project.org/>

Tukey, J. W. (1949). Comparing individual means in the analysis of variance.
*Biometrics*, 5(2), 99–114.

Wickham, H. (2016). *ggplot2: Elegant Graphics for Data Analysis*.
Springer-Verlag New York. <https://ggplot2-book.org/>
