---
title: "Basic Plotting and Theme System"
author: "iQualityR Team"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Basic Plotting and Theme System}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

# Basic Plotting and Theme System

## Overview

The `iQualityR.plot` sub-package provides a unified, extensible plotting infrastructure on which all graphical functions are built. This chapter introduces:

- `as_iqr_theme()` - Theme conversion function
- `base_plot()` - Base plotting function
- The flexibility and extensibility of the theme system

## Load Data and Packages

```{r setup}
library(ggplot2)
library(iQualityR.core)
library(iQualityR.plot)
data("mtcars")
```

## Theme System: as_iqr_theme()

`as_iqr_theme()` is the core function of the iQualityR plotting system, responsible for converting various theme specifications into ggplot2-recognizable theme objects.

### Supported Input Types

This function supports four input types:

```{r theme_types, eval=FALSE}
# 1. NULL - Use default theme
as_iqr_theme(NULL)

# 2. Character string - Use built-in theme name
as_iqr_theme("academic")
 
as_iqr_theme("tech")

# 3. Function - Theme generation function
as_iqr_theme(function() theme_minimal())

# 4. IqrTheme object - Theme object instance
theme_obj <- IqrTheme$new("academic")
as_iqr_theme(theme_obj)
```

### Default Theme Behavior

When no theme is specified, the system attempts to retrieve the default theme from global options:

```{r default_theme}
# Check current default theme setting
getOption("iqr.default_theme", "Not set")

# Set default theme to academic
options(iqr.default_theme = "academic")
```

### Theme Fallback Mechanism

If the specified theme is unavailable, the system automatically falls back to `theme_minimal()`:

```{r fallback_demo, eval=FALSE}
# If IqrTheme is unavailable, it will fall back to theme_minimal()
p <- ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point() +
  as_iqr_theme("nonexistent_theme")  # Will issue a warning and use theme_minimal()

# throw out a error message,we need give a right style, can not fallback to theme_minimal()
```

## Base Plotting: base_plot()

`base_plot()` is the entry function of the iQualityR plotting system, creating a base ggplot object with a unified theme.

### Basic Usage

```{r base_plot_basic}
# Create a basic scatter plot using base_plot
p <- base_plot(
  data = mtcars,
  mapping = aes(x = hp, y = mpg),
  theme = NULL  # Use default theme
) +
  geom_point(color = "#1f77b4", alpha = 0.7) +
  labs(
    title = "Horsepower vs Fuel Efficiency",
    x = "Horsepower (hp)",
    y = "Miles per Gallon (mpg)"
  )

print(p)
```

### Custom Theme

```{r base_plot_custom_theme}
# Use a custom theme
p_custom <- base_plot(
  data = mtcars,
  mapping = aes(x = wt, y = mpg),
  theme = function() theme_bw() + theme(plot.title = element_text(face = "bold"))
) +
  geom_point(color = "#e74c3c", alpha = 0.6, size = 2) +
  labs(
    title = "Weight vs Fuel Efficiency",
    x = "Weight (wt)",
    y = "Miles per Gallon (mpg)"
  )

print(p_custom)
```

### Comparison with Native ggplot2 Syntax

```{r comparison, eval=FALSE}
# Traditional ggplot2 approach
ggplot(mtcars, aes(hp, mpg)) +
  geom_point() +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text = element_text(size = 11)
  )

# iQualityR approach - more concise, better theme consistency
base_plot(mtcars, aes(hp, mpg), theme = "academic") +
  geom_point()
```

## Theme Consistency Example

The following example demonstrates how to maintain theme consistency across different plots:

```{r theme_consistency}
# Set a unified theme
my_theme <- "academic"

# Plot 1: Scatter plot
p1 <- base_plot(mtcars, aes(hp, mpg), theme = my_theme) +
  geom_point(color = "#2ecc71", alpha = 0.6) +
  labs(title = "Scatter Plot")

# Plot 2: Histogram
p2 <- base_plot(mtcars, aes(mpg), theme = my_theme) +
  geom_histogram(fill = "#3498db", color = "white", bins = 20) +
  labs(title = "Histogram")

# Plot 3: Box plot
p3 <- base_plot(mtcars, aes(factor(cyl), mpg), theme = my_theme) +
  geom_boxplot(fill = "#e67e22", alpha = 0.7) +
  labs(title = "Box Plot")

# Print plots
print(p1)
```

## Advanced Theme Customization

### Creating a Custom Theme Function

```{r custom_theme_function}
# Create a corporate-style theme
corporate_theme <- function() {
  theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold", color = "#2c3e50"),
      plot.subtitle = element_text(hjust = 0.5, size = 11, color = "#7f8c8d"),
      axis.title = element_text(size = 11, face = "bold"),
      axis.text = element_text(size = 10),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "#ecf0f1", linewidth = 0.5),
      plot.background = element_rect(fill = "#fafafa", color = NA),
      panel.background = element_rect(fill = "#fafafa", color = NA)
    )
}

# Apply custom theme
p_corp <- base_plot(
  data = mtcars,
  mapping = aes(x = disp, y = mpg),
  theme = corporate_theme
) +
  geom_point(color = "#8e44ad", alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", color = "#e74c3c", fill = "#e74c3c", alpha = 0.1) +
  labs(
    title = "Engine Displacement vs Fuel Efficiency",
    x = "Displacement (disp)",
    y = "Miles per Gallon (mpg)"
  )

print(p_corp)
```

## Best Practices

### 1. Unified Theme Management

```{r best_practice_1, eval=FALSE}
# Recommended: Set default theme at the start of the project
options(iqr.default_theme = "academic")

# All subsequent plots automatically use this theme
p1 <- base_plot(data, aes(x, y)) + geom_point()
p2 <- base_plot(data, aes(x, z)) + geom_line()
```

### 2. Separate Theme from Plot

```{r best_practice_2, eval=FALSE}
# Recommended: Separate theme definition from plot creation
my_theme <- function() theme_bw() + theme(plot.title = element_text(hjust = 0.5))

# Reuse across different plots
plot1 <- base_plot(data1, aes(x, y), theme = my_theme) + geom_point()
plot2 <- base_plot(data2, aes(a, b), theme = my_theme) + geom_bar()
```

### 3. Theme Inheritance

```{r best_practice_3, eval=FALSE}
# Recommended: Fine-tune on top of a base theme
base_theme <- as_iqr_theme("academic")

# Override specific elements in a particular plot
p <- base_plot(data, aes(x, y), theme = "academic") +
  geom_point() +
  theme(
    plot.title = element_text(size = 16, color = "darkblue"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## FAQ

### Q: Why is my theme not taking effect?

Make sure the theme object is correctly passed to `base_plot()`:

```{r faq_1, eval=FALSE}
# Wrong: Theme not passed to base_plot
ggplot(data, aes(x, y)) +
  geom_point() +
  as_iqr_theme("academic")  # This will not take effect

# Correct: Pass through base_plot
base_plot(data, aes(x, y), theme = "academic") +
  geom_point()
```

### Q: How to maintain consistent style across all plots in a report?

```{r faq_2, eval=FALSE}
# Method 1: Set global default theme
options(iqr.default_theme = "academic")

# Method 2: Create a theme variable and use it in all plots
report_theme <- "academic"
p1 <- base_plot(data1, aes(x, y), theme = report_theme) + ...
p2 <- base_plot(data2, aes(a, b), theme = report_theme) + ...
```

## Summary

- `as_iqr_theme()` provides a flexible theme conversion mechanism, supporting multiple input types
- `base_plot()` is the unified entry point of the iQualityR plotting system, ensuring theme consistency
- Through theme functions and objects, you can easily create and reuse custom themes
- It is recommended to set a default theme at the start of a project to ensure consistent style across all plots

## Further Reading

- [Scatter Plots](scatter_plots.html) - Learn how to create scatter plots using the base plotting system
- [Pareto Charts](pareto_charts.html) - Learn about Pareto chart plotting methods
- [Layer Builders](layer_builders.html) - Dive deeper into layer function usage
