---
title: "Distribution and Normality Visualization"
author: "iQualityR.plot Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Distribution and Normality Visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

library(iQualityR.plot)
library(ggplot2)

fmt_num <- function(x, digits = 4) formatC(x, format = "f", digits = digits)
fmt_pct <- function(x, digits = 2) paste0(formatC(x, format = "f", digits = digits), "%")
```

## 1. Introduction

Many statistical methods used in quality engineering — process capability
analysis, control chart design, hypothesis tests on means, and gauge R&R
studies — rely on the assumption that the underlying process output follows
a specified probability distribution. The most common assumption is
**normality**, but Weibull, exponential, and uniform distributions also
appear in reliability, waiting-time, and randomness studies.

Visual diagnostics are the first line of defense against distributional
misspecification. A well-constructed probability plot can reveal skewness,
heavy tails, multimodality, and outliers long before a formal hypothesis
test would flag them. The `iQualityR.plot` package provides three building
blocks for this purpose:

* `plot_pp()` — Probability-Probability (P-P) plots
* `plot_qq()` — Quantile-Quantile (Q-Q) plots with confidence bands and
  goodness-of-fit tests
* `layers_qq()` — Composable `ggplot2` layers for custom Q-Q constructions

This vignette walks through the theory behind each plot, demonstrates every
function on realistic quality-engineering data, and provides a decision
guide for interpreting the results.

## 2. Theory: Assessing Distributional Assumptions

### 2.1 P-P plots vs Q-Q plots — what each diagnoses

Both P-P and Q-Q plots compare an empirical distribution to a theoretical
one, but they emphasize different regions of the data.

**P-P plot.** For an ordered sample $x_{(1)} \le x_{(2)} \le \dots \le x_{(n)}$,
the P-P plot pairs the empirical cumulative probability

$$
p_i^{emp} = \frac{i - 0.5}{n}
$$

against the theoretical cumulative probability

$$
p_i^{theo} = F\!\left(x_{(i)};\,\theta\right)
$$

where $F$ is the candidate CDF with parameters $\theta$. If the data follow
$F$, the points fall on the $y = x$ diagonal. Because both axes are
cumulative *probabilities* (not raw data values), the P-P plot is
**most sensitive to deviations in the center** of the distribution, where
most observations accumulate.

**Q-Q plot.** The Q-Q plot pairs the ordered sample values (sample
quantiles) against the theoretical quantiles

$$
q_i = F^{-1}\!\left(\frac{i - 0.5}{n};\,\theta\right)
$$

so the horizontal axis is in the **units of the theoretical distribution**
while the vertical axis is in the units of the data. The Q-Q plot is
**most sensitive to deviations in the tails**, because tail quantiles are
spread far apart along the theoretical axis.

| Feature | P-P plot | Q-Q plot |
|---|---|---|
| Horizontal axis | Empirical CDF | Theoretical quantile |
| Vertical axis | Theoretical CDF | Sample quantile |
| Sensitive region | Center | Tails |
| Units | Probabilities (0–1) | Data units |
| Outlier detection | Weak | Strong |

### 2.2 Expected quantile calculation

For a normal distribution, the plotting position most commonly used in Q-Q
plots is the **Blom approximation**,

$$
z_i = \Phi^{-1}\!\left(\frac{i - 3/8}{n + 1/4}\right),
$$

although the simpler form $\Phi^{-1}((i - 0.5)/n)$ is adequate for most
practical work and is the convention used by `stats::qqnorm()`. For
non-normal families the inverse CDF $F^{-1}$ is substituted directly; for
example, the Weibull quantile function is

$$
Q(p) = \eta\,[-\ln(1 - p)]^{1/\beta},
$$

with shape $\beta$ and scale $\eta$.

### 2.3 Common distribution families in quality engineering

| Family | CDF | Typical use in quality |
|---|---|---|
| Normal | $\Phi((x-\mu)/\sigma)$ | Measurement error, dimensional data |
| Uniform | $(x - a)/(b - a)$ | Randomness checks, lottery, balance |
| Exponential | $1 - e^{-\lambda x}$ | Waiting times between events |
| Weibull | $1 - e^{-(x/\eta)^\beta}$ | Time-to-failure, reliability |

The `plot_pp()` and `plot_qq()` functions accept a `dist_family` shortcut
that selects one of `"none"`, `"norm"`, `"unif"`, `"exp"`, or `"weibull"`,
or a custom `distribution`/`dist_params` pair for any `p*`/`q*` function.

## 3. plot_pp() — Probability-Probability Plots

`plot_pp()` produces a P-P plot of the empirical cumulative probability
against the theoretical cumulative probability. Set `add_test = TRUE` to
also print a Kolmogorov–Smirnov goodness-of-fit test to the console.

```{r}
# Basic normal P-P plot on simulated measurement data
set.seed(1)
meas <- data.frame(length = rnorm(120, mean = 50, sd = 0.3))

p1 <- plot_pp(
  data        = meas,
  sample_col  = "length",
  dist_family = "norm",
  add_test    = TRUE
)
p1
```

The points hug the diagonal, consistent with normality. The Kolmogorov–
Smirnov statistic reported to the console quantifies the maximum vertical
distance between the empirical and theoretical CDFs.

**Custom distribution.** Any valid `p*` function can be supplied. Here the
theoretical distribution is forced to be uniform, and the departure from
the diagonal is immediately visible:

```{r}
plot_pp(
  data        = meas,
  sample_col  = "length",
  dist_family = "unif",
  add_test    = TRUE
)
```

**Exponential family** for simulated inter-arrival times:

```{r}
set.seed(7)
waits <- data.frame(t = rexp(150, rate = 0.4))

plot_pp(
  data        = waits,
  sample_col  = "t",
  dist_family = "exp",
  add_test    = TRUE
)
```

## 4. plot_qq() — Quantile-Quantile Plots

`plot_qq()` is the workhorse for distributional assessment. Beyond the
base Q-Q point cloud, it supports a reference line, a simulated confidence
band, and a choice of goodness-of-fit test.

### 4.1 Normal Q-Q

By default `plot_qq()` produces a normal Q-Q plot with a quartile reference
line (the line passes through the first and third sample quartiles) and an
Anderson–Darling test.

```{r}
plot_qq(
  data           = meas,
  sample_col     = "length",
  dist_family    = "norm",
  qq_line        = "quartiles",
  add_confidence = TRUE,
  conf_level     = 0.95,
  n_sim          = 1000,
  add_test       = TRUE,
  test_method    = "auto"
)
```

The grey band represents the region in which 95% of the points are expected
to fall if the data truly follow the fitted normal distribution. Points
that leave the band signal departures from normality.

**Identity line.** When the theoretical and sample scales already match
(for example, standardized residuals), use `qq_line = "identity"` to draw
the $y = x$ line:

```{r}
std_resid <- data.frame(r = scale(meas$length)[, 1])

plot_qq(
  data        = std_resid,
  sample_col  = "r",
  dist_family = "norm",
  qq_line     = "identity",
  add_test    = FALSE
)
```

### 4.2 Fitting other distributions

`dist_family = "weibull"` fits a two-parameter Weibull distribution to
simulated time-to-failure data. Note that the Anderson–Darling test is
preferred for Weibull because it weights tail discrepancies more heavily
than the K-S test.

```{r}
set.seed(21)
fail_times <- data.frame(t = rweibull(100, shape = 1.7, scale = 500))

plot_qq(
  data           = fail_times,
  sample_col     = "t",
  dist_family    = "weibull",
  qq_line        = "quartiles",
  add_confidence = TRUE,
  conf_level     = 0.95,
  add_test       = TRUE,
  test_method    = "ad"
)
```

The `test_method` argument controls which goodness-of-fit statistic is
reported:

* `"auto"` — Anderson–Darling for normal, Kolmogorov–Smirnov otherwise
* `"ad"` — Anderson–Darling (tail-weighted)
* `"ks"` — Kolmogorov–Smirnov (center-weighted)

## 5. layers_qq() — Composable Q-Q layers

`layers_qq()` returns a list of `ggplot2` layers (a `stat_qq` layer and a
`stat_qq_line` layer) that can be added to any `ggplot()` call. This is
useful when you need fine control over aesthetics or want to combine the
Q-Q construction with other geoms. The function accepts a `theme` argument
so the layer colors follow the active theme, just like the other `layers_*`
builders.

```{r}
# Build a custom Q-Q plot by hand using layers_qq().
# Point and line colors are pulled from the resolved theme via attr(p, "iqr_theme").
p_qq <- base_plot(meas, aes(sample = .data$length), theme = "academic")
c_qq <- attr(p_qq, "iqr_theme")

p_qq <- p_qq +
  layers_qq(
    distribution = "norm",
    dparams      = list(mean = 50, sd = 0.3),
    theme        = "academic",
    qq_args      = list(color = c_qq$get_pal("discrete")[1], alpha = 0.7),
    line_args    = list(color = c_qq$get_pal("semantic", name = "fail"),
                        linewidth = 1.1)
  ) +
  labs(
    title    = "Custom Q-Q plot using layers_qq()",
    subtitle = "Normal reference, parameters fixed by the analyst",
    x        = "Theoretical quantiles",
    y        = "Sample quantiles"
  )

print(p_qq)
```

The layers accept any arguments that `ggplot2::stat_qq()` and
`ggplot2::stat_qq_line()` accept, passed through `qq_args` and `line_args`.

## 6. Quality Scenario: Normality Assessment of Injection-Molded Parts

A Six Sigma team is verifying the normality assumption for a wall-thickness
capability study on injection-molded plastic housings. One hundred parts
were sampled from a single cavity. Past experience suggests the process is
"mostly normal" but occasionally contaminated by a slightly off-center
short shot. We simulate this structure and apply every diagnostic in turn.

```{r}
set.seed(42)
molded <- data.frame(
  thickness = c(rnorm(80, 2.5, 0.08), rnorm(20, 2.6, 0.12))  # mostly normal with some contamination
)

# Quick look at the data
summary(molded$thickness)
#>    Min. 1st Qu.  Median    Mean 3rd Qu.    Max.
#>   2.282   2.446   2.501   2.518   2.579   2.812
```

**Step 1 — P-P plot.** The center of the distribution looks healthy, but
recall that P-P plots are insensitive to tail contamination.

```{r}
plot_pp(
  data        = molded,
  sample_col  = "thickness",
  dist_family = "norm",
  add_test    = TRUE
)
```

**Step 2 — Q-Q plot with confidence band.** The tails reveal the
contamination: the upper-tail points drift outside the 95% simulated band.

```{r}
plot_qq(
  data           = molded,
  sample_col     = "thickness",
  dist_family    = "norm",
  qq_line        = "quartiles",
  add_confidence = TRUE,
  conf_level     = 0.95,
  n_sim          = 1000,
  add_test       = TRUE,
  test_method    = "ad"
)
```

**Step 3 — Compare Anderson–Darling vs Kolmogorov–Smirnov.** The
Anderson–Darling statistic is more likely to flag the tail contamination
because it weights squared discrepancies in the tails.

```{r}
plot_qq(
  data        = molded,
  sample_col  = "thickness",
  dist_family = "norm",
  add_test    = TRUE,
  test_method = "ks"
)
```

**Step 4 — Custom Q-Q with layers_qq().** The team wants to overlay the
Q-Q construction on a plot with their own theme and annotation. Colors are
resolved from the active theme via `attr(p, "iqr_theme")`, so no hardcoded
values are needed.

```{r}
p_molded <- base_plot(molded, aes(sample = .data$thickness), theme = "academic")
c_molded <- attr(p_molded, "iqr_theme")

p_molded +
  layers_qq(
    distribution = "norm",
    theme        = "academic",
    qq_args      = list(color = c_molded$get_pal("discrete")[1], alpha = 0.7),
    line_args    = list(color = c_molded$get_pal("semantic", name = "fail"),
                        linetype = "dashed")
  ) +
  annotate(
    "text", x = 2, y = 2.75,
    label = "Upper-tail contamination\nfrom short shots",
    color = c_molded$get_ui_colors()$warning, hjust = 0
  ) +
  labs(
    title    = "Wall thickness — Q-Q diagnostic",
    subtitle = "Injection-molded housings, cavity A",
    x        = "Theoretical quantiles",
    y        = "Sample thickness (mm)"
  )
```

**Step 5 — Decision.** The combined evidence (visible tail departure on the
Q-Q plot, elevated Anderson–Darling statistic) tells the team that the
cavity produces a contaminated mixture. Before computing $C_p$/$C_{pk}$ they
should investigate the short-shot root cause or use a robust capability
method rather than assuming normality.

## 7. Interpreting Distribution Plots — Decision Guide

| Pattern on Q-Q plot | Likely cause | Recommended action |
|---|---|---|
| Points follow the line, inside band | Data consistent with distribution | Proceed with capability / SPC analysis |
| Both tails bend upward | Right-skewed (positive skew) | Try Weibull or lognormal; transform data |
| Both tails bend downward | Left-skewed (negative skew) | Try reflected-lognormal; check for floor effect |
| Heavy tails (S-shape) | Leptokurtic; mixture or contamination | Investigate subgroups; consider robust methods |
| Light tails (inverted S) | Platykurtic; truncated data | Check for inspection screening |
| Single off-diagonal point | Outlier | Verify measurement; consider retention |
| Curved overall pattern | Wrong distribution family | Re-fit with alternative `dist_family` |

**Reading the confidence band.** The band shows the envelope within which
*every* point is expected to fall, with probability `conf_level`, when the
data truly follow the fitted distribution. One or two points leaving the
band is not alarming; systematic excursions in the tails or center are.

**Test vs plot.** A non-significant goodness-of-fit test does **not**
prove the distribution fits — large samples will reject nearly any
distribution, and small samples rarely reject. Always read the test
alongside the plot.

## 8. Parameter Reference

### plot_pp()

| Argument | Type | Default | Description |
|---|---|---|---|
| `data` | data.frame | — | Data frame containing the sample |
| `sample_col` | bare name | — | Column with sample values |
| `theme` | ggplot theme | `NULL` | Optional `ggplot2` theme |
| `distribution` | function | `pnorm` | CDF function of candidate distribution |
| `dist_params` | list | `list()` | Named parameters passed to `distribution` |
| `dist_family` | character | `"none"` | Shortcut: `"norm"`, `"unif"`, `"exp"`, `"weibull"` |
| `add_test` | logical | `TRUE` | Print Kolmogorov–Smirnov test to console |

### plot_qq()

| Argument | Type | Default | Description |
|---|---|---|---|
| `data` | data.frame | — | Data frame containing the sample |
| `sample_col` | bare name | — | Column with sample values |
| `theme` | ggplot theme | `NULL` | Optional `ggplot2` theme |
| `distribution` | function | `qnorm` | Quantile function of candidate distribution |
| `dist_params` | list | `list()` | Named parameters passed to `distribution` |
| `dist_family` | character | `"none"` | Shortcut family selector |
| `qq_line` | character | `"quartiles"` | `"quartiles"` or `"identity"` |
| `add_confidence` | logical | `FALSE` | Add simulated confidence band |
| `conf_level` | numeric | `0.95` | Confidence level for the band |
| `n_sim` | integer | `1000` | Number of simulations for the band |
| `add_test` | logical | `TRUE` | Print goodness-of-fit test |
| `test_method` | character | `"auto"` | `"auto"`, `"ad"`, or `"ks"` |

### layers_qq()

| Argument | Type | Default | Description |
|---|---|---|---|
| `distribution` | character | `"norm"` | Distribution name understood by `stat_qq` |
| `dparams` | list | `list()` | Distribution parameters |
| `theme` | character/IqrTheme | `NULL` | Theme name or object; defaults resolved via `as_iqr_theme_object()` |
| `qq_args` | list | `list()` | Extra arguments forwarded to `stat_qq()` |
| `line_args` | list | `list()` | Extra arguments forwarded to `stat_qq_line()` |

## 9. Summary

* **P-P plots** are best for diagnosing the *center* of a distribution.
* **Q-Q plots** are best for diagnosing the *tails* and detecting outliers.
* Use **Anderson–Darling** when tail behavior matters (reliability,
  capability); use **Kolmogorov–Smirnov** for a general-purpose center
  check.
* The **simulated confidence band** provides a visual acceptance region
  without relying on asymptotic approximations.
* **layers_qq()** lets you embed Q-Q diagnostics inside bespoke
  `ggplot2` constructions for reports and dashboards.
* Always pair a visual diagnostic with a goodness-of-fit test, and never
  treat a non-rejection as proof of fit.

## 10. Further Reading

* Montgomery, D. C. (2013). *Introduction to Statistical Quality Control*,
  7th ed. Wiley. — Chapters on process capability and measurement systems.
* D'Agostino, R. B., & Stephens, M. A. (1986). *Goodness-of-Fit
  Techniques*. Marcel Dekker. — Reference for Anderson–Darling and related
  statistics.
* Wickham, H. (2016). *ggplot2: Elegant Graphics for Data Analysis*.
  Springer. — Background on `stat_qq()` and layer composition.
* Meeker, W. Q., & Escobar, L. A. (1998). *Statistical Methods for
  Reliability Data*. Wiley. — Weibull and lognormal Q-Q methodology.

## References

D'Agostino, R. B., & Stephens, M. A. (1986). *Goodness-of-Fit Techniques*.
New York: Marcel Dekker.

Meeker, W. Q., & Escobar, L. A. (1998). *Statistical Methods for
Reliability Data*. New York: Wiley.

Montgomery, D. C. (2013). *Introduction to Statistical Quality Control*
(7th ed.). Hoboken, NJ: Wiley.

Wickham, H. (2016). *ggplot2: Elegant Graphics for Data Analysis*.
New York: Springer.
