---
title: "Fishbone Diagrams"
author: "iQualityR Team"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Fishbone Diagrams}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 12,
  fig.height = 8,
  warning = FALSE,
  message = FALSE
)
```

# Fishbone Diagrams

## Overview

The Fishbone Diagram, also known as the Ishikawa Diagram or Cause-and-Effect Diagram, is an essential tool in quality management for root cause analysis. `iQualityR.plot` provides three fishbone diagram functions:

- `plot_fishbone_basic()` - Basic fishbone diagram (without weight and importance)
- `plot_fishbone_weighted()` - Weighted fishbone diagram (line thickness indicates weight, color indicates importance)
- `plot_fishbone_faceted()` - Faceted fishbone diagram (multi-group comparison)

## Fishbone Diagram Structure

A standard fishbone diagram contains the following hierarchical structure:

- **Spine**: Central horizontal line, with an arrow pointing to the problem statement
- **Primary categories (main bones)**: At a 45-degree angle to the spine, alternating above and below (e.g., 5M1E)
- **Secondary categories (sub-bones)**: Horizontal lines, parallel to the spine
- **Tertiary categories (small bones)**: Parallel to the main bones (at a 45-degree angle)

## Load Package

```{r setup}
library(ggplot2)
 
library(iQualityR.plot)
library(iQualityR.core)

 
```

## 1. Basic Fishbone Diagram: plot_fishbone_basic()

The basic fishbone diagram is suitable for simple root cause analysis scenarios, without weight and importance encoding.

### Manufacturing 5M1E Template

```{r basic_manufacturing, warning=FALSE}
# Define cause categories
categories <- list(
  Manpower = c("Lack of training", "Insufficient skills", "Low motivation"),
  Machine = c("Improper maintenance", "Equipment aging", "Missing calibration"),
  Material = c("Raw material defects", "Supplier changes"),
  Method = c("Unclear SOP", "Unreasonable process"),
  Measurement = c("Measurement error", "Inconsistent standards"),
  Environment = c("Temperature fluctuation", "Excessive humidity")
)

plot_fishbone_basic(
  problem = "Product dimension deviation",
  categories = categories,
  industry = "manufacturing",
  title = "Product Dimension Deviation Fishbone Diagram",
  subtitle = "Manufacturing 5M1E Analysis"
)
```

### Service 5P Template

```{r basic_service,warning=FALSE}
# Service categories
service_categories <- list(
  People = c("Poor service attitude", "Insufficient expertise", "Poor communication"),
  Process = c("Complex process", "Long approval time", "Lack of standardization"),
  Place = c("Inconvenient location", "Noisy environment", "Outdated facilities"),
  Product = c("Incomplete features", "Unstable quality", "Slow updates"),
  Promotion = c("Inadequate promotion", "Non-transparent pricing", "Few promotions")
)

plot_fishbone_basic(
  problem = "Low customer satisfaction",
  categories = service_categories,
  industry = "service",
  title = "Low Customer Satisfaction Fishbone Diagram",
  subtitle = "Service 5P Analysis"
)
```

### Custom Categories

```{r basic_custom,warning=FALSE}
# Custom categories (not relying on industry templates)
custom_categories <- list(
  Manpower = c("Lack of experience", "Weak responsibility", "Insufficient training"),
  Machine = c("Insufficient precision", "Delayed maintenance", "Severe aging"),
  Material = c("Batch variation", "Supplier changes", "Improper storage"),
  Method = c("Unreasonable process", "Incorrect parameter settings", "Unclear standards")
)

plot_fishbone_basic(
  problem = "Welding quality defects",
  categories = custom_categories,
  title = "Welding Quality Defect Analysis",
  subtitle = "Custom Categories"
)
```

## 2. Weighted Fishbone Diagram: plot_fishbone_weighted()

The weighted fishbone diagram uses line thickness to indicate the weight of causes and color to indicate importance level.

### Basic Usage

```{r weighted_basic,warning=FALSE}
# Define categories with weight and importance
weighted_categories <- list(
  Manpower = list(
    list(name = "Lack of training", weight = 8, importance = "critical"),
    list(name = "Insufficient skills", weight = 6, importance = "important"),
    list(name = "Low motivation", weight = 3, importance = "normal")
  ),
  Machine = list(
    list(name = "Improper maintenance", weight = 7, importance = "important"),
    list(name = "Equipment aging", weight = 9, importance = "critical"),
    list(name = "Missing calibration", weight = 4, importance = "normal")
  ),
  Material = list(
    list(name = "Raw material defects", weight = 8, importance = "critical"),
    list(name = "Supplier changes", weight = 5, importance = "important")
  ),
  Method = list(
    list(name = "Unclear SOP", weight = 6, importance = "important"),
    list(name = "Unreasonable process", weight = 7, importance = "important")
  ),
  Measurement = list(
    list(name = "Measurement error", weight = 5, importance = "important"),
    list(name = "Inconsistent standards", weight = 4, importance = "normal")
  ),
  Environment = list(
    list(name = "Temperature fluctuation", weight = 3, importance = "normal"),
    list(name = "Excessive humidity", weight = 4, importance = "normal")
  )
)

plot_fishbone_weighted(
  problem = "Product quality issues",
  categories = weighted_categories,
  industry = "manufacturing",
  title = "Product Quality Issues Fishbone Diagram",
  subtitle = "With Weight and Importance Encoding"
)
```

### Custom Importance Colors

```{r weighted_colors,warning=FALSE}
# Custom importance colors
custom_colors <- c(
  "critical" = "#8e44ad",  # Purple for critical
  "important" = "#e67e22", # Orange for important
  "normal" = "#95a5a6"     # Gray for normal
)

plot_fishbone_weighted(
  problem = "Delivery delay",
  categories = weighted_categories,
  importance_colors = custom_colors,
  title = "Custom Importance Colors"
)
```

### Control Display Elements

```{r weighted_display, eval=FALSE}
# Show weights only, hide importance labels
plot_fishbone_weighted(
  problem = "Problem",
  categories = weighted_categories,
  show_weights = TRUE,
  show_importance = FALSE
)

# Show importance only, hide weight values
plot_fishbone_weighted(
  problem = "Problem",
  categories = weighted_categories,
  show_weights = FALSE,
  show_importance = TRUE
)
```

### Tertiary Category Support

```{r weighted_tertiary,warning=FALSE}
# Complete structure with tertiary categories
tertiary_categories <- list(
  Manpower = list(
    list(
      name = "Lack of training",
      weight = 8,
      importance = "critical",
      sub_categories = list(
        list(name = "Insufficient new employee training", weight = 7, importance = "important"),
        list(name = "Missing on-the-job training", weight = 6, importance = "important"),
        list(name = "Training effectiveness not evaluated", weight = 5, importance = "normal")
      )
    ),
    list(
      name = "Insufficient skills",
      weight = 6,
      importance = "important",
      sub_categories = list(
        list(name = "Poor operational skills", weight = 5, importance = "important"),
        list(name = "Weak problem-solving ability", weight = 6, importance = "important")
      )
    )
  ),
  Machine = list(
    list(name = "Improper maintenance", weight = 7, importance = "important"),
    list(name = "Equipment aging", weight = 9, importance = "critical")
  )
)

plot_fishbone_weighted(
  problem = "High product defect rate",
  categories = tertiary_categories,
  title = "Tertiary Category Fishbone Diagram",
  subtitle = "Including main bones, sub-bones, and small bones"
)
```

## 3. Faceted Fishbone Diagram: plot_fishbone_faceted()

The faceted fishbone diagram is used to compare root cause analysis results across different strata (e.g., shifts, departments, time periods).

### Basic Usage

```{r faceted_basic,warning=FALSE}
# Prepare stratified data
strata_data <- data.frame(
  stratum = rep(c("Day shift", "Night shift"), each = 12),
  category = rep(c("Manpower", "Machine", "Material", "Method", "Measurement", "Environment"), 2),
  sub_category = c(
    "Lack of training", "Improper maintenance", "Raw material defects", "Unclear SOP", "Measurement error", "Temperature fluctuation",
    "Insufficient skills", "Equipment aging", "Supplier changes", "Unreasonable process", "Inconsistent standards", "Excessive humidity",
    "Lack of training", "Improper maintenance", "Raw material defects", "Unclear SOP", "Measurement error", "Temperature fluctuation",
    "Insufficient skills", "Equipment aging", "Supplier changes", "Unreasonable process", "Inconsistent standards", "Excessive humidity"
  ),
  weight = c(8, 7, 8, 6, 5, 3, 6, 9, 5, 7, 4, 4,
             7, 6, 7, 5, 4, 2, 5, 8, 4, 6, 3, 3),
  importance = c("critical", "important", "critical", "important", "important", "normal",
                 "important", "critical", "important", "important", "normal", "normal",
                 "important", "important", "important", "important", "normal", "normal",
                 "important", "critical", "important", "important", "normal", "normal")
)

plot_fishbone_faceted(
  problem = "Product quality issues",
  strata_data = strata_data,
  title = "Day Shift vs Night Shift Comparison Fishbone Diagram"
)
```

## Parameter Details

### plot_fishbone_basic() Parameters

| Parameter | Type | Description | Default |
|------|------|------|--------|
| `problem` | character | Problem statement | Required |
| `categories` | list | Cause categories | NULL |
| `industry` | character | Industry template | "manufacturing" |
| `font_size` | numeric | Base font size | 10 |
| `title` | character | Chart title | NULL |
| `subtitle` | character | Subtitle | NULL |

### plot_fishbone_weighted() Parameters

| Parameter | Type | Description | Default |
|------|------|------|--------|
| `problem` | character | Problem statement | Required |
| `categories` | list | Weighted cause categories | NULL |
| `industry` | character | Industry template | "manufacturing" |
| `weight_range` | numeric | Line thickness range | c(0.8, 3.5) |
| `importance_colors` | named vector | Importance colors | c(critical="red", ...) |
| `show_weights` | logical | Whether to show weights | TRUE |
| `show_importance` | logical | Whether to show importance | TRUE |
| `font_size` | numeric | Base font size | 12 |

### Data Structure Description

**Basic fishbone diagram category format:**
```r
list(
  Category1 = c("Cause1", "Cause2", "Cause3"),
  Category2 = c("Cause1", "Cause2")
)
```

**Weighted category format:**
```r
list(
  Category1 = list(
    list(name = "Cause1", weight = 8, importance = "critical"),
    list(name = "Cause2", weight = 6, importance = "important")
  ),
  Category2 = list(...)
)
```

**Tertiary category format:**
```r
list(
  Category1 = list(
    list(
      name = "Cause1",
      weight = 8,
      importance = "critical",
      sub_categories = list(
        list(name = "Sub-cause1", weight = 7, importance = "important"),
        list(name = "Sub-cause2", weight = 6, importance = "normal")
      )
    )
  )
)
```

## Practical Case Studies

### Case 1: Manufacturing Product Quality Issues

```{r case_manufacturing,warning=FALSE}
# Product quality issues at an auto parts factory
auto_parts_categories <- list(
  Manpower = list(
    list(name = "Insufficient operational skills", weight = 7, importance = "important"),
    list(name = "Weak quality awareness", weight = 6, importance = "important"),
    list(name = "High proportion of new employees", weight = 5, importance = "normal")
  ),
  Machine = list(
    list(name = "Declined machine precision", weight = 9, importance = "critical"),
    list(name = "Fixture wear", weight = 7, importance = "important"),
    list(name = "Improper tool life management", weight = 6, importance = "important")
  ),
  Material = list(
    list(name = "Uneven raw material hardness", weight = 8, importance = "critical"),
    list(name = "Large batch variation", weight = 5, importance = "important")
  ),
  Method = list(
    list(name = "Unreasonable cutting parameters", weight = 7, importance = "important"),
    list(name = "Insufficient inspection frequency", weight = 6, importance = "important")
  ),
  Measurement = list(
    list(name = "Expired gauge calibration", weight = 5, importance = "important"),
    list(name = "Inconsistent measurement methods", weight = 4, importance = "normal")
  ),
  Environment = list(
    list(name = "Workshop temperature fluctuation", weight = 3, importance = "normal"),
    list(name = "Delayed chip cleaning", weight = 4, importance = "normal")
  )
)

plot_fishbone_weighted(
  problem = "Parts dimension out of tolerance",
  categories = auto_parts_categories,
  title = "Auto Parts Dimension Out of Tolerance Fishbone Diagram",
  subtitle = "Root Cause Analysis at a Manufacturing Plant"
)
```

### Case 2: Software Project Delay Analysis

```{r case_software,warning=FALSE}
# Root cause analysis for software project delays
software_delay <- list(
  Manpower = list(
    list(name = "Requirements misunderstanding", weight = 8, importance = "critical"),
    list(name = "Insufficient technical capability", weight = 6, importance = "important"),
    list(name = "Poor team collaboration", weight = 5, importance = "important")
  ),
  Process = list(
    list(name = "Frequent requirement changes", weight = 9, importance = "critical"),
    list(name = "Insufficient reviews", weight = 7, importance = "important"),
    list(name = "Missing progress tracking", weight = 6, importance = "important")
  ),
  Technology = list(
    list(name = "Inappropriate technology selection", weight = 7, importance = "important"),
    list(name = "Architecture design flaws", weight = 8, importance = "critical"),
    list(name = "Insufficient test coverage", weight = 6, importance = "important")
  ),
  Tools = list(
    list(name = "Unstable development environment", weight = 4, importance = "normal"),
    list(name = "Chaotic version management", weight = 5, importance = "important")
  )
)

plot_fishbone_weighted(
  problem = "Project delivery delay",
  categories = software_delay,
  title = "Software Project Delay Fishbone Diagram",
  subtitle = "IT Project Root Cause Analysis"
)
```

## Best Practices

### 1. Choose the Appropriate Number of Categories

```{r best_practice_categories, eval=FALSE}
# Manufacturing: Use 5M1E (6 categories)
industry = "manufacturing"

# Service: Use 5P (5 categories)
industry = "service"

# Custom: Recommend 4-8 categories; too many will make the chart crowded
```

### 2. Weight Setting Recommendations

```{r best_practice_weights, eval=FALSE}
# Weight range: 1-10
# 1-3: Minor impact
# 4-6: Moderate impact
# 7-8: Significant impact
# 9-10: Critical impact

# Importance levels:
# "critical": Must be resolved immediately
# "important": Requires focused attention
# "normal": General attention
```

### 3. Team Collaboration

```{r best_practice_team, eval=FALSE}
# Recommend using brainstorming to determine causes and weights
# 1. Collect team member opinions
# 2. Vote to determine weights
# 3. Reach consensus on importance levels
# 4. Use fishbone diagram to visualize results
```

## FAQ

### Q: How to adjust font size?

```{r faq_font, eval=FALSE}
plot_fishbone_basic(problem, categories, font_size = 12)  # Increase font size
plot_fishbone_basic(problem, categories, font_size = 8)   # Decrease font size
```

### Q: How to customize the problem box color?

The problem box color is fixed to red (#E74C3C). For customization, you can modify the source code or use `plot_fishbone_weighted()` and adjust the theme.

### Q: How many levels of categories does the fishbone diagram support?

Three levels of categories are supported:
- Primary: Main bones (at a 45-degree angle to the spine)
- Secondary: Sub-bones (parallel to the spine, horizontal lines)
- Tertiary: Small bones (parallel to the main bones, at a 45-degree angle)

## Summary

- `plot_fishbone_basic()` is suitable for simple root cause analysis
- `plot_fishbone_weighted()` intuitively displays cause weight and importance through line thickness and color encoding
- `plot_fishbone_faceted()` supports multi-group comparison analysis
- Supports three-level category structure to meet complex analysis needs
- Provides manufacturing 5M1E and service 5P industry templates

## Further Reading

- [Pareto Charts](pareto_charts.html) - Learn about Pareto chart plotting
- [Hypothesis Testing Visualization](hypothesis_visualization.html) - Learn about statistical test visualization
- [Quality Tools Charts](quality_tools.html) - Learn about PP/QQ/Turtle diagrams
