---
title: "Ishikawa Fishbone Diagrams for Root Cause Analysis"
author: "iQualityR.plot Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Ishikawa Fishbone Diagrams for Root Cause Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

library(iQualityR.plot)
library(ggplot2)

fmt_num <- function(x, digits = 4) formatC(x, format = "f", digits = digits)
fmt_pct <- function(x, digits = 2) paste0(formatC(x, format = "f", digits = digits), "%")
```

## 1. Introduction

When a process goes out of control, the first question is *what* changed; the
second, harder question is *why*. **Root cause analysis (RCA)** is the structured
discipline of moving from the observed symptom to the underlying cause. Among the
graphical tools available to RCA, the **Ishikawa diagram** — also called the
**fishbone diagram** — is the most widely used. It was developed by Kaoru
Ishikawa in the 1950s at Kawasaki Steel and popularized in the postwar Japanese
quality movement.

The diagram's strength is that it forces a team to brainstorm causes within
predefined major categories, ensuring broad coverage rather than a fixation on the
first plausible explanation. The `iQualityR.plot` package provides three
functions that cover the common variants:

* `plot_fishbone_basic()` — the classic unweighted diagram.
* `plot_fishbone_weighted()` — line width encodes a numeric weight, color encodes
  importance, so the team can triage causes visually.
* `plot_fishbone_faceted()` — small-multiples layout for comparing cause
  structures across strata (departments, shifts, plants).

This vignette covers the theory, the functions, and a worked example from a
plating line.

## 2. Theory: Cause-and-Effect Analysis

### 2.1 The 5M1E framework (Man/Machine/Material/Method/Measurement/Environment)

Ishikawa's original template organized causes into six categories, all beginning
with the letter **M** in English (or with the corresponding Japanese terms):

| Category     | Typical causes                                              |
|--------------|------------------------------------------------------------|
| Man          | Skills, training, fatigue, motivation, turnover.           |
| Machine      | Equipment age, calibration, tooling wear, maintenance.     |
| Material     | Raw-material variability, supplier changes, storage.       |
| Method       | Procedures, work instructions, setup parameters.           |
| Measurement  | Inspection method, gage R&R, sampling plan.                |
| Environment  | Temperature, humidity, lighting, cleanliness, vibration.   |

In practice, **Measurement** is frequently folded into **Method** because the two
are tightly coupled: a measurement procedure *is* a method. The `iQualityR.plot`
package follows this convention. When `industry = "manufacturing"` is selected,
the template exposes five bones — Man, Machine, Material, Method, Environment —
which is the classic 5M1E layout used in automotive and electronics quality
manuals.

### 2.2 The 5P framework for service industries

Service processes do not have raw materials or machines in the manufacturing
sense. The **5P** framework adapts the categories:

| Category         | Typical causes                                          |
|------------------|---------------------------------------------------------|
| People           | Staff skills, training, scheduling, morale.             |
| Policies         | Rules, authorization limits, pricing.                   |
| Procedures       | Workflows, scripts, checklists.                         |
| Place            | Branch layout, signage, queue design.                   |
| Product/Service  | Service catalogue design, IT systems, physical product. |

Select `industry = "service"` in any of the fishbone functions to load the 5P
template. The two templates are otherwise structurally identical, which keeps
training overhead low when an organization spans both manufacturing and service
operations.

### 2.3 Weighting and importance coding

A fishbone diagram produced in a one-hour brainstorming session may list twenty
candidate causes. Acting on all twenty is rarely feasible, so the team needs a
triage mechanism. `plot_fishbone_weighted()` and `plot_fishbone_faceted()`
support two complementary encodings:

**Weight (numeric).** Each sub-cause carries a numeric weight, typically on a 1-3
or 1-5 scale, that reflects the team's collective belief about how likely the
cause is to be a true driver of the problem. The weight is mapped to **line
thickness** through

$$
\ell_i = \ell_{\min} + \frac{w_i - w_{\min}}{w_{\max} - w_{\min}} \cdot (\ell_{\max} - \ell_{\min}),
$$

where $[\ell_{\min}, \ell_{\max}]$ is the `weight_range` argument. Higher weight
$\Rightarrow$ thicker bone $\Rightarrow$ more probable cause.

**Importance (categorical).** Each sub-cause is tagged as `"critical"`,
`"important"`, or `"normal"`. The mapping is resolved from the active
`IqrTheme` semantic palette:

| Importance  | Theme semantic | Default academic look | Meaning                                  |
|-------------|----------------|-----------------------|------------------------------------------|
| `critical`  | `fail`         | red                   | Must be investigated; high confirmed impact. |
| `important` | `watch`        | orange/amber          | Likely contributor; needs verification.      |
| `normal`    | `neutral`      | gray                  | Plausible but lower priority.                 |

Switching the `theme` argument repaints every bone in one stroke, so
the diagram always matches the surrounding report's color system.

The two encodings are orthogonal: a cause can be thick-lined (high weight) and
gray (low importance) if the team suspects it strongly but has not yet gathered
evidence. Conversely, a thin red line marks a cause with confirmed impact but
narrow scope. The combination lets a facilitator steer the team's attention
without discarding hypotheses prematurely.

### Ishikawa's principle

Ishikawa's central insight was that *structure improves brainstorming*. An
unstructured brainstorm gravitates toward the loudest voice in the room; a
fishbone forces the team to consider, in turn, whether each major category could
harbor a cause. The diagram is therefore as much a *process* artifact — a record
of disciplined thinking — as it is a *communication* artifact.

## 3. plot_fishbone_basic() — standard Ishikawa diagram

`plot_fishbone_basic()` draws the classic diagram without weight or importance
encoding. It is the right starting point for a first-pass brainstorm.

```{r}
str(plot_fishbone_basic)
```

Key arguments:

| Argument      | Purpose                                                       |
|---------------|---------------------------------------------------------------|
| `problem`     | Problem statement; rendered at the fish head.                 |
| `categories`  | Named list of sub-causes per main bone. `NULL` loads template.|
| `industry`    | `"manufacturing"` (5M1E) or `"service"` (5P).                 |
| `font_size`   | Base font size for labels.                                    |
| `title`       | Optional plot title.                                          |
| `subtitle`    | Optional plot subtitle.                                       |
| `theme`       | Optional theme override.                                      |

### Default manufacturing template

With `categories = NULL`, the function emits a blank 5M1E skeleton that the team
can use as a worksheet.

```{r, fig.height = 5.5}
plot_fishbone_basic(
  problem   = "Excessive variation in plating thickness",
  industry  = "manufacturing",
  font_size = 10,
  title     = "5M1E Skeleton for Plating Variation"
)
```

### Custom categories

Once the brainstorm is complete, supply the named list to render the populated
diagram.

```{r, fig.height = 5.5}
plating_basic <- list(
  Man = c("Insufficient training", "Fatigue on night shift"),
  Machine = c("Anode wear", "Bath temperature drift"),
  Material = c("Contaminated electrolyte", "Inconsistent substrate"),
  Method = c("Outdated plating procedure", "Inadequate rinsing step"),
  Environment = c("High humidity in plating room")
)

plot_fishbone_basic(
  problem    = "Plating defects on Line 3",
  categories = plating_basic,
  industry   = "manufacturing",
  font_size  = 10,
  title      = "Plating Defects — Initial Brainstorm",
  subtitle   = "Five-bone 5M1E layout"
)
```

### Service template

Switching to `industry = "service"` loads the 5P skeleton.

```{r, fig.height = 5.5}
service_basic <- list(
  People = c("New teller training gaps", "High turnover"),
  Policies = c("Authorization threshold too low"),
  Procedures = c("Manual reconciliation step", "No dual-control check"),
  Place = c("Counter layout forces multitasking"),
  Product = c("Core banking system latency")
)

plot_fishbone_basic(
  problem    = "Deposit posting errors at Branch 12",
  categories = service_basic,
  industry   = "service",
  font_size  = 10,
  title      = "Service 5P Fishbone for Posting Errors"
)
```

## 4. plot_fishbone_weighted() — line width = weight, color = importance

`plot_fishbone_weighted()` adds the two visual encodings described in Section
2.3. The `categories` argument changes shape: each sub-cause is now a list with
elements `name`, `weight`, and `importance`, plus an optional
`sub_categories` for a third level of detail.

```{r}
str(plot_fishbone_weighted)
```

| Argument             | Purpose                                                        |
|----------------------|----------------------------------------------------------------|
| `problem`            | Problem statement.                                             |
| `categories`         | Named list of bones; each sub-cause is a list.                 |
| `industry`           | `"manufacturing"` or `"service"`.                              |
| `weight_range`       | Numeric range mapped to line thickness.                        |
| `importance_colors`  | Named vector of colors; `NULL` resolves to theme `fail`/`watch`/`neutral` semantics. |
| `show_weights`       | Print weight numbers next to bones.                            |
| `show_importance`    | Print importance label next to bones.                          |
| `font_size`          | Base font size.                                                |
| `theme`              | Theme name or `IqrTheme` object; drives the default `importance_colors`. |
| `title`, `subtitle`  | Optional plot annotations.                                     |

### Worked example with weights

We use the plating scenario with weights on a 0.5-3.5 scale.

```{r, fig.height = 6, fig.width = 9}
plating_causes <- list(
  Man = list(
    list(name = "Insufficient training", weight = 2.5, importance = "critical"),
    list(name = "Fatigue on night shift", weight = 1.5, importance = "important")
  ),
  Machine = list(
    list(name = "Anode wear", weight = 3.0, importance = "critical"),
    list(name = "Bath temperature drift", weight = 2.0, importance = "important")
  ),
  Material = list(
    list(name = "Contaminated electrolyte", weight = 2.8, importance = "critical"),
    list(name = "Inconsistent substrate quality", weight = 1.2, importance = "normal")
  ),
  Method = list(
    list(name = "Outdated plating procedure", weight = 1.8, importance = "important"),
    list(name = "Inadequate rinsing step", weight = 1.0, importance = "normal")
  ),
  Environment = list(
    list(name = "High humidity in plating room", weight = 0.9, importance = "normal")
  )
)

plot_fishbone_weighted(
  problem         = "Plating defects on Line 3",
  categories      = plating_causes,
  industry        = "manufacturing",
  weight_range    = c(0.8, 3.5),
  show_weights    = TRUE,
  show_importance = TRUE,
  font_size       = 11,
  title           = "Weighted Fishbone: Plating Defects",
  subtitle        = "Line width = weight, color = importance"
)
```

### Reading the weighted chart

* The thickest bones are `Anode wear` (weight 3.0) and `Contaminated electrolyte`
  (2.8) — both also tagged `critical`. These are the team's first-priority
  hypotheses.
* `Bath temperature drift` (2.0, important) and `Outdated plating procedure`
  (1.8, important) form a second tier.
* The thinnest bones (`High humidity`, `Inadequate rinsing`) are kept on the
  diagram for completeness but visually de-emphasized.

### Custom importance colors

The default color scheme resolves from the active theme's semantic
palette: `critical` uses the `fail` semantic, `important` uses the
`watch` semantic, and `normal` uses the `neutral` semantic. Switching
`theme` therefore repaints every bone in one stroke — no per-call color
arguments are required:

```{r, fig.height = 6, fig.width = 9}
plot_fishbone_weighted(
  problem         = "Plating defects on Line 3",
  categories      = plating_causes,
  industry        = "manufacturing",
  weight_range    = c(0.8, 3.5),
  theme           = "economist",
  show_weights    = TRUE,
  show_importance = TRUE,
  font_size       = 11,
  title           = "Theme-driven importance colors (economist)",
  subtitle        = "critical = fail semantic, important = watch, normal = neutral"
)
```

For accessibility (e.g., color-blind-safe palettes) or for matching
corporate brand guidelines, `importance_colors` can override the
theme-derived defaults. To keep the override consistent with the active
theme, build the vector from the `IqrTheme` object retrieved via
`attr()` on a seed plot:

```{r, fig.height = 6, fig.width = 9}
p_seed <- base_plot(data.frame(x = 1), ggplot2::aes(x = .data$x),
                    theme = "solarized")
c_sol <- attr(p_seed, "iqr_theme")
sem <- c_sol$get_pal("semantic")

plot_fishbone_weighted(
  problem            = "Plating defects on Line 3",
  categories         = plating_causes,
  industry           = "manufacturing",
  weight_range       = c(0.8, 3.5),
  theme              = "solarized",
  importance_colors  = c(critical = sem[["fail"]],
                         important = sem[["watch"]],
                         normal    = sem[["neutral"]]),
  show_weights       = TRUE,
  show_importance    = FALSE,
  font_size          = 11,
  title              = "Override built from theme semantics (solarized)"
)
```

## 5. plot_fishbone_faceted() — comparing across departments/shifts

When the same problem manifests differently across strata — different shifts,
lines, or plants — a single fishbone hides the contrast. `plot_fishbone_faceted()`
produces one small fishbone per stratum and arranges them in a grid via
`patchwork`.

```{r}
str(plot_fishbone_faceted)
```

The `strata_data` argument is a data frame with columns `stratum`, `category`,
`sub_category`, `weight`, and `importance`. Each row is one sub-cause within one
stratum.

### Example: two-shift comparison

Suppose the plating line runs a day shift and a night shift, and the quality team
suspects that the cause profile differs.

```{r, fig.height = 7, fig.width = 10}
strata_df <- data.frame(
  stratum = rep(c("Day shift", "Night shift"), each = 8),
  category = rep(c("Man", "Machine", "Material", "Method"), times = 4),
  sub_category = rep(
    c("Training", "Anode wear", "Electrolyte", "Procedure",
      "Fatigue", "Temp drift", "Substrate", "Rinsing"),
    times = 2
  ),
  weight = c(1.0, 2.5, 2.0, 1.2,
             2.8, 1.8, 1.5, 2.2),
  importance = rep(
    c("normal", "critical", "important", "normal",
      "critical", "important", "important", "important"),
    times = 2
  )
)

# Make the night shift slightly different: bump fatigue and reduce training
strata_df$weight[strata_df$stratum == "Night shift" &
                   strata_df$sub_category == "Fatigue"] <- 3.2
strata_df$weight[strata_df$stratum == "Night shift" &
                   strata_df$sub_category == "Training"] <- 0.6
strata_df$weight[strata_df$stratum == "Day shift" &
                   strata_df$sub_category == "Fatigue"] <- 1.2

plot_fishbone_faceted(
  problem     = "Plating defects across shifts",
  strata_data = strata_df,
  industry    = "manufacturing",
  weight_range = c(0.5, 3),
  ncol        = 2,
  show_weights = TRUE,
  font_size   = 8,
  title       = "Faceted Fishbone: Day vs Night Shift",
  subtitle    = "Fatigue dominates the night-shift diagram"
)
```

The faceted view makes the contrast obvious: on the day shift, `Anode wear` and
`Electrolyte` are the leading causes; on the night shift, `Fatigue` becomes the
thickest bone. The team can then design shift-specific countermeasures rather
than a single one-size-fits-all action plan.

## 6. Quality Scenario: root cause analysis of plating defects

We now assemble the full workflow on the plating scenario.

### Step 1 — Frame the problem

The line has reported an increase in plating defects (nodules, thin spots,
adhesion failures) over the past two weeks. The problem statement should be
specific and measurable:

> "Plating defect rate on Line 3 rose from 1.2% to 3.8% between 2026-07-01 and
> 2026-07-15, primarily as nodules and adhesion failures."

### Step 2 — Initial brainstorm (basic fishbone)

The cross-functional team meets for 30 minutes and populates a 5M1E skeleton.

```{r, fig.height = 5.5}
plot_fishbone_basic(
  problem    = "Plating defect rate 1.2% -> 3.8% on Line 3",
  categories = plating_basic,
  industry   = "manufacturing",
  font_size  = 10,
  title      = "Step 2: Initial Brainstorm"
)
```

### Step 3 — Weighted triage

The team then votes on each cause using a 1-3 scale and tags importance based on
historical evidence.

```{r, fig.height = 6, fig.width = 9}
plot_fishbone_weighted(
  problem         = "Plating defect rate 1.2% -> 3.8% on Line 3",
  categories      = plating_causes,
  industry        = "manufacturing",
  weight_range    = c(0.8, 3.5),
  show_weights    = TRUE,
  show_importance = TRUE,
  font_size       = 11,
  title           = "Step 3: Weighted Triage",
  subtitle        = "Top-priority causes in red"
)
```

### Step 4 — Stratum comparison

Maintenance records reveal that the defect spike is concentrated on the night
shift. The team builds a faceted fishbone to confirm the shift effect.

```{r, fig.height = 7, fig.width = 10}
plot_fishbone_faceted(
  problem      = "Plating defect rate spike",
  strata_data  = strata_df,
  industry     = "manufacturing",
  weight_range = c(0.5, 3),
  ncol         = 2,
  show_weights = TRUE,
  font_size    = 8,
  title        = "Step 4: Shift-stratified Comparison",
  subtitle     = "Night-shift fatigue emerges as the leading cause"
)
```

### Step 5 — Action plan

Based on the weighted and faceted diagrams, the team defines three
countermeasures, prioritized by weight and importance:

| Priority | Cause                       | Action                                          | Owner        | Due       |
|----------|-----------------------------|-------------------------------------------------|--------------|-----------|
| 1        | Anode wear (Machine)        | Replace anode; tighten PM interval to 2 weeks.  | Maintenance  | 2026-08-05|
| 2        | Contaminated electrolyte    | Drain and refill bath; add filtration step.     | Process Eng. | 2026-08-01|
| 3        | Night-shift fatigue (Man)   | Add second break; rotate tasks every 2 hours.   | HR / Superv. | 2026-08-10|

After implementation, the team re-runs the data collection and rebuilds the
weighted fishbone to confirm that the previously critical bones have shrunk.

## 7. Parameter Reference

| Function                  | Argument             | Type / Default                | Description                                      |
|---------------------------|----------------------|-------------------------------|--------------------------------------------------|
| `plot_fishbone_basic`     | `problem`            | character                     | Problem statement at fish head.                  |
| `plot_fishbone_basic`     | `categories`         | `NULL`                        | Named list of sub-causes; `NULL` loads template. |
| `plot_fishbone_basic`     | `industry`           | `"manufacturing"`             | `"manufacturing"` (5M1E) or `"service"` (5P).    |
| `plot_fishbone_basic`     | `font_size`          | `10`                          | Base font size.                                  |
| `plot_fishbone_basic`     | `title` / `subtitle` | `NULL`                        | Optional annotations.                            |
| `plot_fishbone_basic`     | `theme`              | `NULL`                        | Theme name or `IqrTheme` object; forwarded to `plot_fishbone_weighted()`. |
| `plot_fishbone_weighted`  | `categories`         | `NULL`                        | Named list; sub-causes are lists.                |
| `plot_fishbone_weighted`  | `weight_range`       | `c(0.8, 3.5)`                 | Line-thickness range.                            |
| `plot_fishbone_weighted`  | `importance_colors`  | `NULL`                        | Custom color map; `NULL` resolves `critical`/`important`/`normal` to theme `fail`/`watch`/`neutral` semantics. |
| `plot_fishbone_weighted`  | `show_weights`       | `TRUE`                        | Print weight values next to bones.               |
| `plot_fishbone_weighted`  | `show_importance`    | `TRUE`                        | Print importance tags next to bones.             |
| `plot_fishbone_weighted`  | `theme`              | `NULL`                        | Theme name or `IqrTheme` object; drives semantic colors via `.iqr_aes()`. |
| `plot_fishbone_faceted`   | `strata_data`        | data.frame                    | Columns: stratum, category, sub_category, weight, importance. |
| `plot_fishbone_faceted`   | `ncol`               | `2`                           | Number of facet columns.                         |
| `plot_fishbone_faceted`   | `show_weights`       | `FALSE`                       | Print weight values.                             |
| `plot_fishbone_faceted`   | `show_importance`    | `FALSE`                       | Print importance tags.                           |
| `plot_fishbone_faceted`   | `theme`              | `NULL`                        | Theme name or `IqrTheme` object propagated to every facet. |

### Return values

* `plot_fishbone_basic()` and `plot_fishbone_weighted()` return `ggplot` objects.
* `plot_fishbone_faceted()` returns a `patchwork` composition.

Because the basic and weighted variants return `ggplot` objects, additional
layers (annotations, highlighted rectangles) can be added with the `+` operator
familiar from `ggplot2`.

```{r, eval = FALSE}
# Example: save the weighted fishbone to PDF for the CAPA report
ggplot2::ggsave(
  filename = "plating_fishbone_weighted.pdf",
  plot     = plot_fishbone_weighted(
    problem    = "Plating defects on Line 3",
    categories = plating_causes,
    industry   = "manufacturing"
  ),
  width    = 10,
  height   = 6
)
```

## 8. Summary

The fishbone diagram remains one of the most effective RCA tools because it
combines structure with brevity. The `iQualityR.plot` fishbone family supports
the full RCA workflow:

1. **Frame** the problem with a measurable, specific problem statement.
2. **Brainstorm** with `plot_fishbone_basic()` and the 5M1E or 5P template.
3. **Triage** with `plot_fishbone_weighted()` using weight and importance
   encodings.
4. **Stratify** with `plot_fishbone_faceted()` when the problem profile differs
   across subgroups.

The discipline of producing the diagram — not the diagram itself — is what drives
improvement. A team that debates each weight and importance tag has already begun
to converge on the true root cause.

## 9. Further Reading

* Ishikawa, K. (1968). *Guide to Quality Control*. Tokyo: JUSE. — The original
  textbook treatment of the cause-and-effect diagram.
* ISO 9001:2015, Clause 10.2 (Nonconformity and corrective action). — Positions
  RCA within the broader QMS.
* Custer, J. W., et al. (2015). "Root cause analysis using the Ishikawa fishbone
  diagram in healthcare." *BMJ Quality & Safety*. — Service-industry adaptation.
* Wilson, P. F., Dell, L. D., & Anderson, G. F. (1993). *Root Cause Analysis: A
  Tool for Total Quality Management*. ASQC Quality Press. — Compares fishbone,
  FMEA, and fault-tree analysis.
* `patchwork` package documentation for composing faceted fishbone layouts.

## References

Ishikawa, K. (1968). *Guide to Quality Control*. Tokyo: Japanese Standards
Association / JUSE.

Ishikawa, K. (1985). *What Is Total Quality Control? The Japanese Way*
(D. J. Lu, Trans.). Englewood Cliffs, NJ: Prentice-Hall.

International Organization for Standardization. (2015). *ISO 9001:2015 — Quality
management systems — Requirements*. Geneva: ISO.

Wilson, P. F., Dell, L. D., & Anderson, G. F. (1993). *Root Cause Analysis: A
Tool for Total Quality Management*. Milwaukee, WI: ASQC Quality Press.

Custer, J. W., et al. (2015). Root cause analysis using the Ishikawa fishbone
diagram in healthcare. *BMJ Quality & Safety*, 24(suppl 1), i37-i44.

Pedersen, T. L. (2024). *patchwork: The Composer of Plots*. R package version
1.2.0.
