---
title: "Hypothesis Testing Visualization"
author: "iQualityR Team"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Hypothesis Testing Visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

# Hypothesis Testing Visualization

## Overview

Hypothesis testing is a core tool of statistical inference, used to determine whether sample data supports a particular statistical hypothesis. `iQualityR.plot` provides three hypothesis testing visualization functions that transform abstract statistical results into intuitive graphics:

- `plot_hypothesis_curve()` - Rejection region diagram (normal/t distribution curve + rejection region shading)
- `plot_hypothesis_box()` - Box plot (sample data + confidence interval + statistics table)
- `plot_hypothesis_combined()` - Combined diagram (rejection region + box plot)

## Load Data

```{r setup}
library(ggplot2)
library(datasets)
library(iQualityR.plot)
library(iQualityR.core)

data("sleep")
data("mtcars")
```

## 1. Rejection Region Diagram: plot_hypothesis_curve()

The rejection region diagram intuitively shows whether the test statistic falls within the rejection region through the distribution curve, helping understand the decision-making process of hypothesis testing.

### Two-sided Z Test

```{r curve_two_sided_z, warning=FALSE}
plot_hypothesis_curve(
  dist = "norm",
  stat_value = 1.53,
  critical_value = 1.96,
  alternative = "two.sided",
  alpha = 0.05,
  stat_label = "Z",
  p_value = 0.126,
  theme = "tech"
)
```

### Right-tailed t Test

```{r curve_right_t, warning=FALSE}
plot_hypothesis_curve(
  dist = "t",
  stat_value = 2.46,
  critical_value = 1.645,
  alternative = "greater",
  alpha = 0.05,
  df = 399,
  stat_label = "t",
  p_value = 0.007,
  theme = "academic"
)
```

### Left-tailed Test

```{r curve_left, warning=FALSE}
plot_hypothesis_curve(
  dist = "norm",
  stat_value = -2.15,
  critical_value = 1.645,
  alternative = "less",
  alpha = 0.05,
  stat_label = "Z",
  p_value = 0.016
)
```

### Graphical Element Description

- **Blue curve**: Probability density function of the normal or t distribution
- **Red shaded area**: Rejection region (if the statistic falls in this region, reject H0)
- **Blue vertical line**: Calculated test statistic value
- **Gray dashed line**: Critical value position
- **Top annotation**: Statistic value and p-value

## 2. Box Plot: plot_hypothesis_box()

The box plot displays the sample data distribution while annotating the H0 hypothesized value, sample mean, and confidence interval.

### One-sample t Test

```{r box_one_sample, warning=FALSE}
set.seed(123)
sample_data <- rnorm(50, mean = 102, sd = 5)

plot_hypothesis_box(
  x = sample_data,
  mu = 100,
  alternative = "two.sided",
  conf_level = 0.95,
  show_table = TRUE,
  table_position = "right"
)
```

### Z Test with Known Standard Deviation

```{r box_z_test, warning=FALSE}
set.seed(456)
sample_data_z <- rnorm(100, mean = 178, sd = 68)

plot_hypothesis_box(
  x = sample_data_z,
  mu = 170,
  sigma = 65,
  alternative = "greater",
  conf_level = 0.95,
  show_table = TRUE
)
```

### Graphical Element Description

- **Blue box**: Box plot of sample data (median, quartiles)
- **Gray scatter points**: Raw data points (with jitter)
- **Red cross**: H0 hypothesized value (mu0)
- **Blue dot**: Sample mean (x-bar)
- **Blue error bar**: Confidence interval
- **Right table**: Summary of key statistics

## 3. Combined Diagram: plot_hypothesis_combined()

The combined diagram integrates the rejection region diagram and the box plot into one figure, suitable for reports and presentations.

### Basic Usage

```{r combined_basic, warning=FALSE,fig.width=10}
set.seed(789)
sample_data_combined <- rnorm(60, mean = 103, sd = 6)

plot_hypothesis_combined(
  x = sample_data_combined,
  mu = 100,
  sigma = NULL,
  alternative = "two.sided",
  conf_level = 0.95
)
```

### Combined Diagram with Known Standard Deviation

```{r combined_z, warning=FALSE,fig.width=10}
set.seed(101)
sample_data_z_combined <- rnorm(80, mean = 52, sd = 10)

plot_hypothesis_combined(
  x = sample_data_z_combined,
  mu = 50,
  sigma = 10,
  alternative = "greater",
  conf_level = 0.95
)
```

## Practical Case Studies

### Case 1: Sleep Medication Effect Test (sleep dataset)

```{r case_sleep, warning=FALSE,fig.width=10}
data("sleep")

diff_values <- sleep$extra[sleep$group == 2] - sleep$extra[sleep$group == 1]

plot_hypothesis_combined(
  x = diff_values,
  mu = 0,
  alternative = "two.sided",
  conf_level = 0.95
)
```

### Case 2: Automobile Fuel Efficiency Test

```{r case_mtcars,warning=FALSE,fig.width=8}
mtcars_mpg <- mtcars$mpg

plot_hypothesis_box(
  x = mtcars_mpg,
  mu = 20,
  alternative = "two.sided",
  conf_level = 0.95,
  show_table = TRUE
)
```

### Case 3: Production Process Mean Test

```{r case_process,warning=FALSE,fig.width=8}
set.seed(202)
process_data <- rnorm(30, mean = 502, sd = 3)

plot_hypothesis_curve(
  dist = "t",
  stat_value = (mean(process_data) - 500) / (sd(process_data) / sqrt(30)),
  critical_value = qt(0.975, df = 29),
  alternative = "two.sided",
  alpha = 0.05,
  df = 29,
  stat_label = "t",
  p_value = t.test(process_data, mu = 500)$p.value
)
```

## Parameter Details

### plot_hypothesis_curve() Parameters

| Parameter | Type | Description | Default |
|------|------|------|--------|
| `dist` | character | Distribution type ("norm" or "t") | "norm" |
| `stat_value` | numeric | Test statistic value | Required |
| `critical_value` | numeric | Critical value (positive) | Required |
| `alternative` | character | Test type | "two.sided" |
| `alpha` | numeric | Significance level | 0.05 |
| `df` | numeric | Degrees of freedom (required for t distribution) | NULL |
| `stat_label` | character | Statistic label | NULL (auto) |
| `p_value` | numeric | p-value (optional) | NULL |
| `theme` | theme | Theme specification | NULL |

### plot_hypothesis_box() Parameters

| Parameter | Type | Description | Default |
|------|------|------|--------|
| `x` | numeric | Sample data vector | Required |
| `mu` | numeric | H0 hypothesized value | Required |
| `sigma` | numeric | Known population standard deviation | NULL |
| `alternative` | character | Test type | "two.sided" |
| `conf_level` | numeric | Confidence level | 0.95 |
| `show_table` | logical | Whether to show statistics table | TRUE |
| `table_position` | character | Table position | "right" |
| `theme` | theme | Theme specification | NULL |

### plot_hypothesis_combined() Parameters

| Parameter | Type | Description | Default |
|------|------|------|--------|
| `x` | numeric | Sample data vector | Required |
| `mu` | numeric | H0 hypothesized value | Required |
| `sigma` | numeric | Known population standard deviation | NULL |
| `alternative` | character | Test type | "two.sided" |
| `conf_level` | numeric | Confidence level | 0.95 |
| `theme` | theme | Theme specification | NULL |

## Test Type Description

### Two-sided Test (two.sided)

```
H0: mu = mu0
Ha: mu != mu0
```

The rejection region is distributed on both sides, suitable for testing whether the mean equals a specific value.

### Right-tailed Test (greater)

```
H0: mu = mu0
Ha: mu > mu0
```

The rejection region is on the right side, suitable for testing whether the mean is greater than a specific value.

### Left-tailed Test (less)

```
H0: mu = mu0
Ha: mu < mu0
```

The rejection region is on the left side, suitable for testing whether the mean is less than a specific value.

## Z Test vs t Test

| Feature | Z Test | t Test |
|------|--------|--------|
| Population standard deviation | Known (sigma) | Unknown (estimated with sample standard deviation s) |
| Sample size | Usually > 30 | Any size |
| Distribution | Standard normal distribution | t distribution (depends on degrees of freedom) |
| Parameter setting | `sigma` is not NULL | `sigma = NULL` |

## Best Practices

### 1. Choose the Appropriate Test Type

```{r best_practice_test_type, eval=FALSE}
plot_hypothesis_curve(alternative = "greater")

plot_hypothesis_curve(alternative = "two.sided")
```

### 2. Report Complete Information

```{r best_practice_report, eval=FALSE}
plot_hypothesis_combined(x, mu, sigma, alternative)

p1 <- plot_hypothesis_curve(dist, stat_value, critical_value, alternative)
p2 <- plot_hypothesis_box(x, mu, sigma, alternative)
```

### 3. Understanding p-values

```{r best_practice_pvalue, eval=FALSE}
# p < 0.001: Very strong evidence to reject H0
# 0.001 <= p < 0.01: Strong evidence
# 0.01 <= p < 0.05: Moderate evidence
# 0.05 <= p < 0.1: Weak evidence
# p >= 0.1: Insufficient evidence to reject H0
```

## FAQ

### Q: How to choose between Z test and t test?

```{r faq_z_vs_t, eval=FALSE}
plot_hypothesis_box(x, mu, sigma = 10)

plot_hypothesis_box(x, mu, sigma = NULL)
```

### Q: The statistics table is not displayed?

Make sure `show_table = TRUE`:

```{r faq_table, eval=FALSE}
plot_hypothesis_box(x, mu, show_table = TRUE)
```

### Q: How to customize colors?

```{r faq_color, eval=FALSE}
plot_hypothesis_curve(dist, stat_value, critical_value, alternative, theme = "prism")
```

## Summary

- `plot_hypothesis_curve()` intuitively displays the rejection region and statistic position
- `plot_hypothesis_box()` displays sample data distribution and confidence interval
- `plot_hypothesis_combined()` combines both views, suitable for report presentations
- Supports Z test and t test, two-sided and one-sided tests
- Automatically calculates and displays key statistics

## Further Reading

- [Fishbone Diagrams](fishbone_diagrams.html) - Learn about root cause analysis tools
- [Quality Tools Charts](quality_tools.html) - Learn about PP/QQ/Turtle diagrams
- [Scatter Plots](scatter_plots.html) - Learn about scatter plot plotting
