---
title: "Hypothesis Testing Visualization"
author: "iQualityR.plot Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Hypothesis Testing Visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

library(iQualityR.plot)
library(ggplot2)

fmt_num <- function(x, digits = 4) formatC(x, format = "f", digits = digits)
fmt_pct <- function(x, digits = 2) paste0(formatC(x, format = "f", digits = digits), "%")
```

## 1. Introduction

Hypothesis testing is the backbone of statistical decision making in quality
engineering. Whether you are verifying a supplier's mean fill weight, auditing a
process center against a specification, or validating a measurement target, you
ultimately translate sample evidence into one of two decisions: **reject** or
**fail to reject** the null hypothesis ($H_0$).

Numbers alone are easy to misread. A test statistic of $z = 2.13$ tells a story,
but seeing where that statistic sits relative to the rejection region makes the
conclusion immediate and defensible to non-statisticians on the shop floor. The
`iQualityR.plot` package provides three visualization helpers that turn a
hypothesis test into a clear graphical verdict:

- `plot_hypothesis_curve()` — draws the sampling distribution and shades the
  rejection region, with the test statistic marked as a vertical line.
- `plot_hypothesis_box()` — draws the sample data as a boxplot with the
  hypothesized mean, sample mean, and confidence interval annotated.
- `plot_hypothesis_combined()` — stacks the two views in a single patchwork
  composition so the distribution logic and the raw data evidence can be read
  together.

This vignette walks through the theory behind each plot, demonstrates the
function signatures on a real fill-weight verification scenario, and finishes
with a decision guide and parameter reference. Every function accepts a
`theme` argument (a theme name such as `"academic"` or an `IqrTheme`
object) and resolves its colors — rejection ribbon, statistic line,
critical-value dashes, boxplot fill — from the active theme through
`.iqr_aes()`, so users rarely need to specify colors explicitly.

## 2. Theory: Visualizing the Logic of Hypothesis Testing

A hypothesis test partitions the sample space of a test statistic into two
regions: a region that is consistent with $H_0$ (the *non-rejection region*) and
a region that is so unlikely under $H_0$ that observing the statistic there
forces us to reject $H_0$ (the *rejection region*). Visualization makes that
partition tangible.

### 2.1 The rejection region

Let $T$ be a test statistic whose distribution under $H_0$ is known. The
rejection region $\mathcal{R}$ is the set of statistic values for which we
reject $H_0$:

$$
\mathcal{R} = \{ t : P_{H_0}(T \geq |t|) \leq \alpha \}
$$

For a two-sided z-test at significance level $\alpha = 0.05$, the rejection
region is

$$
\mathcal{R} = \{ z : |z| > 1.96 \}.
$$

On a plot of the standard normal density, this region is the red-shaded area in
both tails, whose combined area equals $\alpha$. The blue vertical line marking
the observed statistic tells you at a glance whether you are inside or outside
the red zone.

### 2.2 Critical values vs test statistics

The **critical value** $c_\alpha$ is the boundary of the rejection region. The
**test statistic** $t_{\text{obs}}$ is the value computed from the sample. The
decision rule is simply

$$
\text{Reject } H_0 \iff |t_{\text{obs}}| > |c_\alpha|.
$$

Visually, the critical value is where the red shading starts; the test
statistic is where the blue line lands. If the blue line sits in the red
shading, reject $H_0$; if it sits in the white area, fail to reject $H_0$. The
plot removes any ambiguity about whether a "borderline" statistic is on the
correct side of the threshold.

### 2.3 One-sided vs two-sided alternatives

The shape of the rejection region depends on the alternative hypothesis
$H_1$:

| Alternative | Rejection region | Tail(s) |
|-------------|------------------|---------|
| $H_1: \mu \neq \mu_0$ (two-sided) | $\|t\| > c_{\alpha/2}$ | both |
| $H_1: \mu > \mu_0$ (greater) | $t > c_{\alpha}$ | right |
| $H_1: \mu < \mu_0$ (less) | $t < -c_{\alpha}$ | left |

The **p-value** is the probability of observing a statistic at least as extreme
as the one observed, assuming $H_0$ is true:

$$
p = P_{H_0}\bigl(|T| \geq |t_{\text{obs}}|\bigr).
$$

When $p < \alpha$ the observed statistic falls inside the rejection region and
we reject $H_0$. The functions in this package let you annotate the p-value
directly on the curve so the graphical verdict and the numeric verdict always
agree.

## 3. plot_hypothesis_curve() — rejection region plot

`plot_hypothesis_curve()` draws a normal or Student's t sampling distribution,
shades the rejection region in red, and drops a blue vertical line at the
observed test statistic. The rule is simple: **blue in red — reject; blue in
white — fail to reject**.

### Function signature

```{r, eval = FALSE}
plot_hypothesis_curve(
  dist = c("norm", "t"),
  stat_value,
  critical_value,
  alternative = c("two.sided", "less", "greater"),
  alpha = 0.05,
  df = NULL,
  stat_label = NULL,
  p_value = NULL,
  theme = NULL
)
```

### Two-sided z-test example

Suppose a z-statistic of 2.34 is observed against a critical value of 1.96
under a two-sided alternative at $\alpha = 0.05$.

```{r}
p <- plot_hypothesis_curve(
  dist = "norm",
  stat_value = 2.34,
  critical_value = 1.96,
  alternative = "two.sided",
  alpha = 0.05,
  stat_label = "Z",
  p_value = 0.0192
)
p
```

The blue line at $Z = 2.34$ falls inside the right-tail red shading, so we
reject $H_0$. The p-value annotation confirms the decision numerically.

### One-sided greater alternative

For a one-sided test, only one tail is shaded. With a statistic of 1.72 against
a one-sided critical value of 1.645:

```{r}
plot_hypothesis_curve(
  dist = "norm",
  stat_value = 1.72,
  critical_value = 1.645,
  alternative = "greater",
  alpha = 0.05,
  stat_label = "Z",
  p_value = 0.0427
)
```

### t-distribution example

When the population standard deviation is unknown and estimated from the
sample, the test statistic follows a t distribution with $n-1$ degrees of
freedom. With $\text{df} = 24$, a statistic of 2.20, and a two-sided critical
value of about 2.064:

```{r}
plot_hypothesis_curve(
  dist = "t",
  stat_value = 2.20,
  critical_value = 2.064,
  alternative = "two.sided",
  alpha = 0.05,
  df = 24,
  stat_label = "t",
  p_value = 0.0382
)
```

The heavier tails of the t distribution are visible: the rejection region is
slightly wider than under the normal curve for the same $\alpha$.

### One-sided less alternative

When the alternative is $H_1: \mu < \mu_0$, the rejection region is in the left
tail. A statistic of $-2.05$ against $-1.96$:

```{r}
plot_hypothesis_curve(
  dist = "norm",
  stat_value = -2.05,
  critical_value = 1.96,
  alternative = "less",
  alpha = 0.05,
  stat_label = "Z",
  p_value = 0.0202
)
```

## 4. plot_hypothesis_box() — H0 line with confidence interval

While `plot_hypothesis_curve()` shows the *distribution* of the test statistic,
`plot_hypothesis_box()` grounds the decision in the *raw sample data*. It draws
a boxplot of the sample, marks the hypothesized mean $\mu_0$ with a red cross,
marks the sample mean $\bar{x}$ with a blue point, and draws the confidence
interval as blue error bars. An embedded statistics table reports the test
statistic, p-value, and decision.

### Function signature

```{r, eval = FALSE}
plot_hypothesis_box(
  x,
  mu,
  sigma = NULL,
  alternative = c("two.sided", "less", "greater"),
  conf_level = 0.95,
  show_table = TRUE,
  table_position = "right",
  theme = NULL
)
```

### Example with simulated data

```{r}
set.seed(42)
fill_weights <- rnorm(30, mean = 498, sd = 4)

p_box <- plot_hypothesis_box(
  x = fill_weights,
  mu = 500,
  alternative = "two.sided",
  conf_level = 0.95
)
p_box
```

The blue point (sample mean) is clearly below the red cross ($\mu_0 = 500$), and
the confidence interval does not overlap $\mu_0$, signaling rejection of $H_0$.

### Z-test mode with known sigma

When the population standard deviation is known, pass `sigma` to perform a
Z-test rather than a t-test:

```{r}
plot_hypothesis_box(
  x = fill_weights,
  mu = 500,
  sigma = 4,
  alternative = "two.sided",
  conf_level = 0.95
)
```

### Table below the plot

Set `table_position = "bottom"` to stack the statistics table under the
boxplot:

```{r}
plot_hypothesis_box(
  x = fill_weights,
  mu = 500,
  alternative = "less",
  conf_level = 0.95,
  table_position = "bottom"
)
```

## 5. plot_hypothesis_combined() — curve + box

`plot_hypothesis_combined()` stacks the rejection-region curve and the sample
boxplot into a single patchwork composition. The top panel shows the
distribution logic; the bottom panel shows the data evidence. Reading the two
together is the most convincing way to communicate a hypothesis test result to
a mixed audience.

### Function signature

```{r, eval = FALSE}
plot_hypothesis_combined(
  x,
  mu,
  sigma = NULL,
  alternative = c("two.sided", "less", "greater"),
  conf_level = 0.95,
  theme = NULL
)
```

### Combined view

```{r, fig.height = 7}
set.seed(42)
fill_weights <- rnorm(30, mean = 498, sd = 4)

p_combined <- plot_hypothesis_combined(
  x = fill_weights,
  mu = 500,
  alternative = "two.sided",
  conf_level = 0.95
)
p_combined
```

The top panel shows the test statistic sitting in the rejection region; the
bottom panel shows the sample mean and its confidence interval clearly below
$\mu_0 = 500$. Both panels tell the same story, which is exactly what you want
in a regulatory or audit setting.

## 6. Quality Scenario: verifying a supplier's mean fill weight claim

A supplier of bottled beverage claims the mean fill weight is 500 g. As the
quality engineer, you draw a random sample of 30 bottles from the latest
shipment. Unbeknownst to the supplier, the process has drifted and the true
mean is 498 g with a standard deviation of 4 g.

```{r}
set.seed(42)
fill_weights <- rnorm(30, mean = 498, sd = 4)

# Sample summary
sample_mean <- mean(fill_weights)
sample_sd <- sd(fill_weights)
sample_se <- sample_sd / sqrt(length(fill_weights))

cat("Sample mean      :", fmt_num(sample_mean), "g\n")
cat("Sample SD         :", fmt_num(sample_sd), "g\n")
cat("Standard error    :", fmt_num(sample_se), "g\n")
```

### Step 1 — state the hypotheses

$$
H_0: \mu = 500 \quad \text{vs} \quad H_1: \mu \neq 500
$$

### Step 2 — compute the test statistic

Because the population SD is unknown, we use a one-sample t-test with
$\text{df} = 29$.

```{r}
t_stat <- (sample_mean - 500) / sample_se
critical_t <- qt(0.975, df = 29)
p_val <- 2 * pt(-abs(t_stat), df = 29)

cat("t statistic       :", fmt_num(t_stat), "\n")
cat("Critical t (0.025):", fmt_num(critical_t), "\n")
cat("p-value           :", fmt_num(p_val), "\n")
```

### Step 3 — visualize the rejection region

```{r}
plot_hypothesis_curve(
  dist = "t",
  stat_value = t_stat,
  critical_value = critical_t,
  alternative = "two.sided",
  alpha = 0.05,
  df = 29,
  stat_label = "t",
  p_value = p_val
)
```

The blue line falls inside the left-tail rejection region: reject $H_0$.

### Step 4 — visualize the sample evidence

```{r}
plot_hypothesis_box(
  x = fill_weights,
  mu = 500,
  alternative = "two.sided",
  conf_level = 0.95
)
```

### Step 5 — combined view for the audit report

```{r, fig.height = 7}
plot_hypothesis_combined(
  x = fill_weights,
  mu = 500,
  alternative = "two.sided",
  conf_level = 0.95
)
```

### Step 6 — theme-driven coloring

Every hypothesis plot resolves its colors from the active `IqrTheme`.
The rejection-region ribbon uses the theme `fail` semantic, the
statistic line uses the theme `primary` semantic, and the critical-value
dashes use the theme `muted` role. Switching `theme` repaints the entire
report — no per-call color arguments are required:

```{r, fig.show="hold", out.width="50%"}
plot_hypothesis_curve(
  dist = "t", stat_value = t_stat, critical_value = critical_t,
  alternative = "two.sided", alpha = 0.05, df = 29,
  stat_label = "t", p_value = p_val, theme = "economist"
) + ggplot2::labs(title = "t-curve — economist")

plot_hypothesis_box(
  x = fill_weights, mu = 500, alternative = "two.sided",
  conf_level = 0.95, theme = "prism"
) + ggplot2::labs(title = "Box — prism")
```

### Conclusion

Because $|t| > c_{0.025, 29}$ and $p < 0.05$, we reject $H_0$. There is
statistically significant evidence that the supplier's mean fill weight differs
from 500 g — specifically, it is below the claimed target. The lot should be
flagged for review.

## 7. Decision Guide — reading the plots

Use the following table to translate what you see on the plot into a decision.

| What you see on the curve plot | What you see on the box plot | Decision |
|--------------------------------|------------------------------|----------|
| Blue line in red shading | CI does not overlap $\mu_0$ | Reject $H_0$ |
| Blue line in white area | CI overlaps $\mu_0$ | Fail to reject $H_0$ |
| Blue line exactly on the critical value | CI edge touches $\mu_0$ | Borderline — collect more data |

Practical tips:

- **Always check both panels.** The curve plot confirms the statistical
  decision; the box plot confirms the practical magnitude. A statistically
  significant shift of 0.1 g may not matter operationally.
- **Mind the alternative.** A one-sided test has more power in its favored
  direction but no power in the opposite direction. Match the alternative to
  the engineering question, not to the desired p-value.
- **Report the p-value, do not worship it.** The plot conveys the same
  information as $p < 0.05$ but also shows *how far* into the rejection region
  the statistic falls, which a bare p-value hides.
- **Use the t distribution when sigma is unknown.** With small samples the
  heavier tails of the t distribution make a real difference to the decision.
- **Check assumptions.** These plots assume the sampling distribution of the
  mean is approximately normal (CLT) or that the underlying data are normal.
  For small samples from skewed data, consider a non-parametric alternative.

## 8. Parameter Reference

### plot_hypothesis_curve()

| Argument | Type | Description |
|----------|------|-------------|
| `dist` | character | `"norm"` for normal, `"t"` for Student's t. |
| `stat_value` | numeric | Observed test statistic (z or t). |
| `critical_value` | numeric | Critical value (positive scalar). |
| `alternative` | character | `"two.sided"`, `"less"`, or `"greater"`. |
| `alpha` | numeric | Significance level (default 0.05). |
| `df` | numeric | Degrees of freedom for the t distribution. |
| `stat_label` | character | Label for the statistic axis (e.g. `"Z"`, `"t"`). |
| `p_value` | numeric | Optional p-value, displayed on the plot. |
| `theme` | character/IqrTheme | Theme name or object; ribbon uses `fail` semantic, statistic line uses `primary`, critical dashes use `muted`. |

### plot_hypothesis_box()

| Argument | Type | Description |
|----------|------|-------------|
| `x` | numeric | Sample data vector. |
| `mu` | numeric | Hypothesized population mean under $H_0$. |
| `sigma` | numeric | Known population SD (Z-test); `NULL` uses sample SD (t-test). |
| `alternative` | character | `"two.sided"`, `"less"`, or `"greater"`. |
| `conf_level` | numeric | Confidence level for the interval (default 0.95). |
| `show_table` | logical | Whether to embed the statistics table. |
| `table_position` | character | `"right"` or `"bottom"`. |
| `theme` | character/IqrTheme | Theme name or object; H0 marker uses `fail` semantic, sample mean uses `primary`. |

### plot_hypothesis_combined()

| Argument | Type | Description |
|----------|------|-------------|
| `x` | numeric | Sample data vector. |
| `mu` | numeric | Hypothesized population mean under $H_0$. |
| `sigma` | numeric | Known population SD (Z-test); `NULL` uses sample SD (t-test). |
| `alternative` | character | `"two.sided"`, `"less"`, or `"greater"`. |
| `conf_level` | numeric | Confidence level (default 0.95). |
| `theme` | character/IqrTheme | Theme name or object propagated to both panels. |

## 9. Summary

The three hypothesis-testing plotters in `iQualityR.plot` turn the abstract
logic of rejection regions, critical values, and p-values into a graphical
verdict that engineers, auditors, and managers can read at a glance:

- **`plot_hypothesis_curve()`** shows whether the test statistic falls inside
  the rejection region of the sampling distribution.
- **`plot_hypothesis_box()`** grounds the decision in the raw sample data with
  the hypothesized mean, sample mean, and confidence interval annotated.
- **`plot_hypothesis_combined()`** stacks both views so the distribution logic
  and the data evidence reinforce each other.

Used together, they make hypothesis test communication defensible,
reproducible, and accessible to non-statisticians.

## 10. Further Reading

- The `iQualityR.plot` package vignettes on **MSA (Measurement System
  Analysis)** and **ANOVA Visualization** for related visualization tools.
- Montgomery, D. C. (2019). *Introduction to Statistical Quality Control*,
  8th ed. Wiley. — Chapters on hypothesis testing and process capability.
- NIST/SEMATECH e-Handbook of Statistical Methods,
  <https://www.itl.nist.gov/div898/handbook/> — sections on test statistics
  and critical values.
- `?stats::t.test` and `?stats::pnorm` for the underlying R machinery that
  computes the statistics visualized here.

## References

Montgomery, D. C. (2019). *Introduction to Statistical Quality Control* (8th
ed.). John Wiley & Sons.

NIST/SEMATECH. (2024). *e-Handbook of Statistical Methods*.
<https://www.itl.nist.gov/div898/handbook/>

R Core Team. (2024). *R: A Language and Environment for Statistical
Computing*. R Foundation for Statistical Computing, Vienna, Austria.
<https://www.R-project.org/>

Wickham, H. (2016). *ggplot2: Elegant Graphics for Data Analysis*.
Springer-Verlag New York. <https://ggplot2-book.org/>
