---
title: "Composable Layer Builders for Quality Charts"
author: "iQualityR.plot Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Composable Layer Builders for Quality Charts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

library(iQualityR.plot)
library(ggplot2)

fmt_num <- function(x, digits = 4) formatC(x, format = "f", digits = digits)
fmt_pct <- function(x, digits = 2) paste0(formatC(x, format = "f", digits = digits), "%")
```

## 1. Introduction

Every chart in `iQualityR.plot` is assembled from small, reusable pieces.
This vignette documents the `layers_*` family of builder functions, which
return ready-to-add lists of ggplot2 layers encapsulating best-practice
defaults for common quality-engineering visualizations: distribution
diagnostics, group comparisons, trend bands, and the reference lines that
make a chart a *quality* chart -- specification limits and control limits.

The running scenario throughout is a pharmaceutical tablet-press process:
individual tablet weights (nominal 250 mg) sampled over 30 consecutive
subgroups, with an engineering tolerance of 242-258 mg. We will use the
layer builders to answer two distinct questions about the same data:

- *Is the process stable?* -- answered by control limits (from the data).
- *Is the stable process capable?* -- answered by specification limits
  (from the customer).

### 1.1 Why composable layers? Reuse and consistency

A quality engineer builds dozens of charts per week that differ only in
details: a histogram today with a density overlay, a control chart tomorrow
with the same axes, a capability panel next week that overlays both. Writing
each chart from scratch with raw `geom_*` calls invites two problems:

1. **Inconsistency.** Slightly different defaults creep in -- bin width,
   jitter width, line type for the UCL -- so charts are not comparable.
2. **Repetition.** The same five-line idiom is copied into every script,
   so a change in house style must be propagated by hand.

The `layers_*` functions solve both problems by encapsulating a
best-practice layer set behind a single call that returns a *list* of
ggplot2 layers. Because the return value is a list, it can be added to any
`base_plot()` with `+`, mixed freely with raw `geom_*` calls, or composed
into a dashboard with `combine_plots()`. The defaults are centralised in one
place, so updating a house style updates every chart at once.

### 1.2 Theme-aware color resolution

Every `layers_*` function accepts a `theme` argument with the same
semantics as `base_plot()`: `NULL` (use the global default), a theme name
string (`"academic"`, `"economist"`, ...), a theme function, or an
`IqrTheme` object. The function resolves the theme once via
`as_iqr_theme_object()` and pulls every color it needs from the theme's
palettes -- so switching the theme recolors the layers automatically.

This means that, in the overwhelming majority of cases, **you do not need
to pass any color argument to a `layers_*` function.** The conventional
pattern is:

```r
base_plot(data, aes(...), theme = "economist") +
  layers_control_chart(data, theme = "economist") +
  layers_spec_limits(lsl = 242, usl = 258, theme = "economist")
```

All three calls resolve the same `IqrTheme` object, and the layers pull
their colors from it: control-limit lines use the semantic `"fail"` color,
the center line uses `"neutral"`, the data line uses the first discrete
palette entry, and the specification lines use `"fail"` as well. Switch the
theme string on all three calls and the whole chart recolors consistently.

The `theme` argument is the *primary* color knob. The legacy
`ucl_color` / `cl_color` / `data_color` / `spec_color` arguments are still
accepted, default to `NULL`, and are only meant for one-off overrides; when
left `NULL`, they fall back to the theme palette.

## 2. Theory: The Grammar of Graphics and Layer Composition

### 2.1 ggplot2 layer model

A ggplot2 plot is a stack of independent layers. Each layer combines five
components:

| Component | Role | Example |
|---|---|---|
| **Data** | The data frame the layer draws from | `cc_data` |
| **Mapping** | Which columns map to which aesthetics | `aes(x = .data$x, y = .data$y)` |
| **Geometry (geom)** | The mark drawn (points, lines, bars) | `geom_point` |
| **Statistic (stat)** | Optional transformation of the data | `stat_qq`, binning |
| **Position** | Adjustments for overlapping marks | `jitter`, `dodge` |

When you write `p + geom_boxplot(...)`, ggplot2 wraps your arguments into a
layer object and appends it to the plot's layer list. Crucially, **a list of
layers can be added with a single `+`**: `p + list(layer1, layer2)` is valid
and appends every layer in the list. This is the mechanism every `layers_*`
function relies on.

### 2.2 How layers_* functions encapsulate best-practice defaults

Each `layers_*` function follows the same pattern:

1. Collect the user's customisation arguments (e.g. `hist_args`,
   `density_args`).
2. Build a default argument list for each underlying geom.
3. Merge defaults with user overrides via `utils::modifyList()` -- user
   wins, defaults fill the gaps.
4. Construct each layer with `do.call(geom_*, final_args)`.
5. Return the layers as an R `list`.

Because `modifyList()` merges recursively, you can override a single nested
field (say, the histogram fill) without re-specifying every other default.
This keeps call sites short while leaving the full geom API available when
you need it.

## 3. Distribution Layers

We first prepare the process data: 30 tablet weights with their
individuals-chart statistics.

```{r process-data}
set.seed(7)
n_points <- 30
tablet_mean <- 250   # mg nominal
tablet_sd   <- 2.5   # mg short-term sigma
x <- 1:n_points
y <- rnorm(n_points, tablet_mean, tablet_sd)
cl  <- mean(y)
sig <- sd(y)
ucl <- cl + 3 * sig
lcl <- cl - 3 * sig
cc_data <- data.frame(x = x, y = y, cl = cl, lcl = lcl, ucl = ucl)
head(cc_data)
#>   x        y       cl      lcl      ucl
#> 1 1 255.7181 250.9892 242.5404 259.4379
#> 2 2 247.0081 250.9892 242.5404 259.4379
#> 3 3 248.2643 250.9892 242.5404 259.4379
#> 4 4 248.9693 250.9892 242.5404 259.4379
#> 5 5 247.5733 250.9892 242.5404 259.4379
#> 6 6 247.6318 250.9892 242.5404 259.4379
```

### 3.1 layers_histogram_density() — histogram + density overlay

`layers_histogram_density(bins = NULL, theme = NULL, ..., hist_args = list(), density_args = list())`
returns a list of two layers: a `geom_histogram` (with `aes(y = after_stat(density))`
so the bar heights are on the density scale) and a `geom_density`. The `...`
arguments are passed to both geoms, while `hist_args` / `density_args` target
a single geom.

The example below resolves the histogram fill and the density line color
from the active theme: the fill is taken from the first discrete palette
entry and the density line from the semantic `"fail"` slot, so both colors
follow the theme automatically. No `scale_*` call and no hardcoded color
appear in the code.

```{r hist-density}
# Resolve theme-driven colors once, then pass them to the layer builders.
p_hd <- base_plot(cc_data, aes(x = .data$y), theme = "academic")
c <- attr(p_hd, "iqr_theme")
hist_fill <- c$get_pal("discrete")[1]
dens_color <- c$get_pal("semantic", name = "fail")

p_hd <- p_hd +
  layers_histogram_density(
    bins         = 12,
    theme        = "academic",
    hist_args    = list(fill = hist_fill, alpha = 0.6),
    density_args = list(color = dens_color, linewidth = 1)
  ) +
  labs(title = "Tablet Weight Distribution",
       x = "Weight (mg)", y = "Density")

print(p_hd)
```

To overlay a theoretical normal curve, add a `geom_line()` with a pre-computed
density -- the layer set is just a starting point and accepts any further
ggplot2 layers.

### 3.2 layers_qq() — quantile-quantile layers

`layers_qq(distribution = "norm", dparams = list(), theme = NULL, ..., qq_args = list(), line_args = list())`
returns `stat_qq` (the points) and `stat_qq_line` (the reference line). The
`distribution` argument is the suffix appended to `q` to fetch the quantile
function (e.g. `"norm"` -> `qnorm`), and `dparams` supplies its parameters.
Use the `sample` aesthetic in the base mapping.

The points and the reference line pull their colors from the resolved theme
so the Q-Q plot matches the rest of the chart.

```{r qq-layer}
p_qq <- base_plot(cc_data, aes(sample = .data$y), theme = "academic")
c_qq <- attr(p_qq, "iqr_theme")

p_qq <- p_qq +
  layers_qq(
    distribution = "norm",
    theme        = "academic",
    qq_args      = list(color = c_qq$get_pal("discrete")[1],
                        size = 2.5, alpha = 0.7),
    line_args    = list(color = c_qq$get_pal("semantic", name = "fail"),
                        linetype = "dashed", linewidth = 1)
  ) +
  labs(title = "Normal Q-Q Plot of Tablet Weight",
       x = "Theoretical Quantiles", y = "Sample Quantiles")

print(p_qq)
```

A linear point cloud hugging the reference line supports the normality
assumption that underpins the ±3sigma control limits used later.

## 4. Comparison Layers

### 4.1 layers_boxplot() — with jitter option

`layers_boxplot(add_jitter = TRUE, theme = NULL, ..., boxplot_args = list(), jitter_args = list())`
returns a `geom_boxplot` and, when `add_jitter = TRUE`, a `geom_jitter` so
individual observations are visible alongside the five-number summary. This is
the recommended default for small-to-moderate quality samples, where hiding
the raw points behind a box can mask outliers and bimodality.

```{r boxplot-layer}
p_box <- base_plot(cc_data, aes(x = factor(1), y = .data$y), theme = "academic")
c_box <- attr(p_box, "iqr_theme")

p_box <- p_box +
  layers_boxplot(
    add_jitter   = TRUE,
    theme        = "academic",
    boxplot_args = list(fill = c_box$get_pal("discrete")[1], alpha = 0.6,
                        outlier.color = c_box$get_pal("semantic", name = "fail"),
                        outlier.size = 2),
    jitter_args  = list(width = 0.12, alpha = 0.6, size = 2,
                        color = c_box$get_ui_colors()$muted)
  ) +
  labs(title = "Tablet Weight Summary", x = NULL, y = "Weight (mg)")

print(p_box)
```

### 4.2 layers_violin() — with optional boxplot overlay

`layers_violin(add_boxplot = TRUE, theme = NULL, ..., violin_args = list(), boxplot_args = list())`
returns a `geom_violin` and, when `add_boxplot = TRUE`, a narrow
`geom_boxplot` drawn on top. The violin shows the full distribution shape
(useful for detecting bimodality that a boxplot hides), while the inner box
preserves the median and quartiles. The inner boxplot fill is resolved from
the theme's `"surface"` UI slot so it contrasts with the violin fill
regardless of which theme is active.

```{r violin-layer}
# Add a second 'shift' to compare two process states
cc2 <- cc_data
cc2$state <- "Before"
cc2$y[16:30] <- cc2$y[16:30] + 1.2   # simulate a small mean shift
cc2$state[16:30] <- "After"

# base_plot() auto-injects the discrete fill scale for the mapped `fill`
# aesthetic, so the violin fills follow the theme without any scale_* call.
# layers_violin(theme = ...) resolves the inner boxplot fill from the theme.
p_vio <- base_plot(cc2,
                   aes(x = .data$state, y = .data$y, fill = .data$state),
                   theme = "academic") +
  layers_violin(
    add_boxplot  = TRUE,
    theme        = "academic",
    violin_args  = list(alpha = 0.55),
    boxplot_args = list(width = 0.12, alpha = 0.9, outlier.shape = NA)
  ) +
  labs(title = "Weight Distribution: Before vs After Shift",
       x = "Process State", y = "Weight (mg)") +
  theme(legend.position = "none")

print(p_vio)
```

## 5. Trend and Band Layers

### 5.1 layers_trend_line() — regression smoothing

`layers_trend_line(mapping = NULL, data = NULL, add_points = TRUE, smoothing = NULL, theme = NULL, line_args = list(), point_args = list(), smooth_args = list())`
returns a `geom_line` plus, optionally, a `geom_point` and a `geom_smooth`.
The optional `mapping`/`data` arguments let the trend line read from a
different data frame than the base plot (e.g. aggregated means over raw
points). `smoothing` accepts any method supported by `geom_smooth`, such as
`"lm"`, `"loess"`, or `"glm"`.

```{r trend-line}
p_tr <- base_plot(cc_data, aes(x = .data$x, y = .data$y), theme = "academic")
c_tr <- attr(p_tr, "iqr_theme")

p_tr <- p_tr +
  layers_trend_line(
    add_points  = TRUE,
    smoothing   = "lm",
    theme       = "academic",
    line_args   = list(color = c_tr$get_pal("discrete")[1], linewidth = 0.6),
    point_args  = list(color = c_tr$get_ui_colors()$muted,
                       size = 1.8, alpha = 0.7),
    smooth_args = list(color = c_tr$get_pal("semantic", name = "fail"),
                       se = FALSE, linetype = "dashed")
  ) +
  labs(title = "Tablet Weight Trend Over Subgroups",
       x = "Subgroup", y = "Weight (mg)")

print(p_tr)
```

A flat, non-significant regression slope is consistent with a process free
of drift -- a first screen for stability before drawing the control limits.

### 5.2 layers_percentile_band() — reference bands

`layers_percentile_band(theme = NULL, ..., errorbar_args = list(), point_args = list())`
returns a `geom_errorbar` plus a `geom_point`, intended for forest-plot
styles where each row is a group estimate with an uncertainty band. The
default `orientation = "y"` places the groups on the y-axis, suitable for
comparing several production lines. The point center fill is resolved from
the theme's `"surface"` UI slot so the point contrasts with its stroke
regardless of which theme is active.

```{r percentile-band}
line_est <- data.frame(
  line  = c("Line 1", "Line 2", "Line 3"),
  est   = c(249.8, 250.6, 250.1),
  lower = c(249.1, 249.9, 249.6),
  upper = c(250.5, 251.3, 250.6)
)

p_band <- base_plot(line_est,
                    aes(x = .data$est, y = .data$line,
                        xmin = .data$lower, xmax = .data$upper),
                    theme = "academic")
c_band <- attr(p_band, "iqr_theme")

p_band <- p_band +
  layers_percentile_band(
    theme         = "academic",
    errorbar_args = list(color = c_band$get_pal("discrete")[1],
                         width = 0.2, linewidth = 1),
    point_args    = list(size = 3,
                         color = c_band$get_pal("semantic", name = "fail"),
                         stroke = 1.5)
  ) +
  geom_vline(xintercept = 250, linetype = "dotted",
             color = c_band$get_ui_colors()$muted) +
  labs(title = "Tablet Weight by Production Line",
       x = "Mean Weight (mg)", y = NULL)

print(p_band)
```

## 6. Quality-Reference Layers

### 6.1 layers_spec_limits() — LSL/USL annotation

`layers_spec_limits(lsl = NULL, usl = NULL, theme = NULL, spec_color = NULL, lsl_label = paste("LSL =", lsl), usl_label = paste("USL =", usl), vline_args = list(), annotate_args = list())`
returns a list of `geom_vline` + `annotate` layers for each supplied limit.
The lines are *vertical*, so this builder is designed for charts whose
measurement axis is the x-axis (histograms, dot plots, capability panels).
Limits left `NULL` are simply not drawn.

When `spec_color = NULL` (the default), the color is resolved from the
theme's semantic `"fail"` slot, so the specification lines follow the
active theme automatically. Pass an explicit color only for one-off
overrides.

```{r spec-limits}
# The histogram fill is resolved from the theme; the spec lines default to
# the theme's semantic "fail" color (no spec_color argument needed).
p_spec <- base_plot(cc_data, aes(x = .data$y), theme = "academic")
c_spec <- attr(p_spec, "iqr_theme")

p_spec <- p_spec +
  layers_histogram_density(bins = 12,
    theme        = "academic",
    hist_args    = list(fill = c_spec$get_pal("discrete")[1], alpha = 0.5),
    density_args = list(color = c_spec$get_ui_colors()$muted,
                        linewidth = 0.8)) +
  layers_spec_limits(lsl = 242, usl = 258, theme = "academic") +
  labs(title = "Capability: Distribution vs Specification",
       x = "Weight (mg)", y = "Density")

print(p_spec)
```

#### Theory: specification vs control limits

Specification limits and control limits are often confused but answer
fundamentally different questions:

- **Specification limits (LSL / USL)** originate from *engineering or
  customer requirements*. They define the range of acceptable product. They
  are fixed before production starts and are independent of what the process
  actually does. A part outside the spec limits is non-conforming.
- **Control limits (UCL / LCL)** originate from *the process data itself*.
  They describe the natural voice of the process -- the range within which a
  stable process is expected to vary. They are computed from the process
  mean and sigma, typically at mean +/- 3sigma, and have no knowledge of
  what the customer wants.

A process can be *in control* (all points inside the control limits) yet
*not capable* (the control limits lie partly outside the spec limits), or
*capable* yet *out of control* (a special cause has shifted the mean). The
two panels built in Section 7 make this distinction visible at a glance.

### 6.2 layers_control_chart() — center line + control limits + warning limits

`layers_control_chart(data, theme = NULL, ucl_color = NULL, cl_color = NULL, data_color = NULL)`
returns a list of layers that draw the full individuals control chart: the
upper and lower control limits (dashed), the center line (solid), the
connecting data line, the data points, text labels for UCL/LCL/CL, and a
`coord_cartesian` that pads the y-axis. The `data` argument must be a data
frame with columns `x`, `y`, `cl`, `lcl`, `ucl`; the geom layers inherit the
base plot's data and mapping, so pass the same data frame to `base_plot()`.

The color arguments all default to `NULL` and are resolved from the theme
when omitted:

- `ucl_color` / `lcl_color` lines <- theme semantic `"fail"` (signals a limit breach)
- `cl_color` line            <- theme semantic `"neutral"` (reference baseline)
- `data_color` line/points   <- first discrete palette entry

Pass explicit colors only when you need to override the theme for a specific
chart.

```{r control-chart}
# All three color arguments are left NULL, so the colors are pulled from
# the "academic" theme. Switching theme = "economist" would recolor every
# line in one step.
p_cc <- base_plot(cc_data, aes(x = .data$x, y = .data$y), theme = "academic") +
  layers_control_chart(cc_data, theme = "academic") +
  labs(title = "Tablet Weight Individuals Control Chart",
       x = "Subgroup", y = "Weight (mg)")

print(p_cc)
```

#### Theory: ±3σ limits, Western Electric zones

The control limits are placed at the process mean plus and minus three
standard deviations:

UCL = mu + 3*sigma,   CL = mu,   LCL = mu - 3*sigma

where `mu` and `sigma` are estimated from the process data (here, the sample
mean and sample standard deviation of the 30 tablet weights). Under the
normality assumption, a point falling outside the +/- 3sigma band is rare
(approximately 0.27% probability) and is treated as a signal of a special
cause.

Between the center line and each control limit, the chart is divided into
three **Western Electric zones**, each one sigma wide:

- **Zone A** (2sigma to 3sigma from the mean) -- the outer band.
- **Zone B** (1sigma to 2sigma from the mean) -- the middle band.
- **Zone C** (0 to 1sigma from the mean) -- the inner band, around the mean.

The Western Electric run rules use these zones to detect non-random
patterns *before* a point actually breaches the control limit. The most
common rules are:

1. One point beyond Zone A (i.e. beyond +/- 3sigma) -- a single special
   cause.
2. Two of three consecutive points in Zone A or beyond, on the same side --
   a small sustained shift.
3. Four of five consecutive points in Zone B or beyond, on the same side --
   a moderate drift.
4. Eight consecutive points on one side of the center line -- a sustained
   shift or stratification.

These rules turn the control chart from a binary "in / out" gate into an
early-warning system. The `layers_control_chart()` builder draws the
center line and +/- 3sigma limits; the sigma-zone shading required by the
run rules can be added on top with `geom_hline()` or `annotate("rect")` when
a full zone chart is needed.

## 7. Combining Layers with base_plot

The clearest way to answer "stable and capable?" for the tablet process is
a two-panel dashboard. The left panel uses `layers_control_chart()` to show
*stability* (control limits, drawn from the data, on the measurement y-axis).
The right panel uses `layers_histogram_density()` plus
`layers_spec_limits()` to show *capability* (specification limits, drawn
from the customer, as vertical lines bracketing the distribution on the
x-axis). `combine_plots()` assembles both panels with `patchwork`.

```{r combined-dashboard}
# Stability panel: control chart (measurement on y-axis).
# layers_control_chart() pulls UCL/CL/data colors from the "academic" theme.
p_stability <- base_plot(cc_data, aes(x = .data$x, y = .data$y),
                         theme = "academic") +
  layers_control_chart(cc_data, theme = "academic") +
  labs(title = "Stability: In Control?",
       subtitle = "Control limits from process data (+/- 3 sigma)",
       x = "Subgroup", y = "Weight (mg)")

# Capability panel: distribution vs spec limits (measurement on x-axis).
# The histogram fill and the spec-line color are both resolved from the theme.
p_capability <- base_plot(cc_data, aes(x = .data$y), theme = "academic")
c_cap <- attr(p_capability, "iqr_theme")
p_capability <- p_capability +
  layers_histogram_density(bins = 12,
    theme        = "academic",
    hist_args    = list(fill = c_cap$get_pal("discrete")[1], alpha = 0.55),
    density_args = list(color = c_cap$get_ui_colors()$muted,
                        linewidth = 0.8)) +
  layers_spec_limits(lsl = 242, usl = 258, theme = "academic") +
  labs(title = "Capability: Meets Spec?",
       subtitle = "Spec limits from engineering tolerance",
       x = "Weight (mg)", y = "Density")

combine_plots(p_stability, p_capability, ncol = 2)
```

Read together, the two panels tell a single story: all 30 points fall
inside the control limits, so the process is stable; and the distribution
sits comfortably within the 242-258 mg specification band, so the stable
process is also capable. Layer builders make this kind of combined
narrative a few lines to assemble.

A custom chart can also mix layer builders with raw ggplot2 layers. For
example, adding a normal-curve overlay to the capability panel:

```{r mix-layers}
x_grid <- seq(min(cc_data$y), max(cc_data$y), length.out = 100)
normal_df <- data.frame(
  x = x_grid,
  y = dnorm(x_grid, mean = mean(cc_data$y), sd = sd(cc_data$y))
)

p_mix <- base_plot(cc_data, aes(x = .data$y), theme = "academic")
c_mix <- attr(p_mix, "iqr_theme")

p_mix <- p_mix +
  layers_histogram_density(bins = 12,
    theme     = "academic",
    hist_args = list(fill = c_mix$get_pal("discrete")[1], alpha = 0.5)) +
  layers_spec_limits(lsl = 242, usl = 258, theme = "academic") +
  geom_line(data = normal_df, aes(x = .data$x, y = .data$y),
            color = c_mix$get_pal("semantic", name = "fail"),
            linewidth = 1, linetype = "dashed") +
  labs(title = "Capability with Normal Overlay",
       x = "Weight (mg)", y = "Density")

print(p_mix)
```

## 8. Parameter Reference Table

| Function | Key parameters | Returns |
|---|---|---|
| `layers_histogram_density()` | `bins`, `theme`, `hist_args`, `density_args` | `geom_histogram` + `geom_density` |
| `layers_qq()` | `distribution`, `dparams`, `theme`, `qq_args`, `line_args` | `stat_qq` + `stat_qq_line` |
| `layers_boxplot()` | `add_jitter`, `theme`, `boxplot_args`, `jitter_args` | `geom_boxplot` (+ `geom_jitter`) |
| `layers_violin()` | `add_boxplot`, `theme`, `violin_args`, `boxplot_args` | `geom_violin` (+ `geom_boxplot`, inner fill from theme `"surface"`) |
| `layers_trend_line()` | `add_points`, `smoothing`, `theme`, `line_args`, `point_args`, `smooth_args` | `geom_line` (+ `geom_point`, `geom_smooth`) |
| `layers_percentile_band()` | `theme`, `errorbar_args`, `point_args` | `geom_errorbar` + `geom_point` (point fill from theme `"surface"`) |
| `layers_spec_limits()` | `lsl`, `usl`, `theme`, `spec_color` (default: theme `"fail"`), `lsl_label`, `usl_label`, `vline_args`, `annotate_args` | `geom_vline` + `annotate` per limit |
| `layers_control_chart()` | `data` (cols x, y, cl, lcl, ucl), `theme`, `ucl_color` (default: theme `"fail"`), `cl_color` (default: theme `"neutral"`), `data_color` (default: discrete[1]) | UCL/LCL/CL lines + data line/points + labels + coord |

All `*_args` lists are merged over sensible defaults with `modifyList()`, so
you only need to specify the arguments you want to change. The `theme`
argument is the primary color knob: every color argument that defaults to
`NULL` falls back to a theme palette slot, so switching the theme recolors
the layers automatically.

## 9. Summary

- Every `layers_*` function returns a *list* of ggplot2 layers that can be
  added to a `base_plot()` with a single `+`.
- Defaults are centralised and merged with user overrides via
  `modifyList()`, so call sites stay short while the full ggplot2 API
  remains available.
- Distribution layers (`layers_histogram_density`, `layers_qq`) diagnose the
  shape and normality of the process data.
- Comparison layers (`layers_boxplot`, `layers_violin`) contrast groups
  while keeping individual points visible.
- Trend and band layers (`layers_trend_line`, `layers_percentile_band`)
  expose drift and cross-group uncertainty.
- Quality-reference layers separate the two questions a quality engineer
  must always keep distinct: `layers_spec_limits()` answers *is it
  capable?* (limits from the customer), while `layers_control_chart()`
  answers *is it stable?* (limits from the data at +/- 3sigma).
- Combining layer builders with `combine_plots()` turns these pieces into a
  coherent stability-and-capability dashboard in a handful of lines.

## 10. Further Reading

- **Visualization Infrastructure: base_plot and the Theme Pipeline** --
  `vignette("plotting-overview", package = "iQualityR.plot")` covers the
  `base_plot()` foundation, theme resolution, and the unified color
  pipeline that the layer builders in this vignette sit on top of.
- **Basic Plotting and Theme System** --
  `vignette("basic_plotting_theme", package = "iQualityR.plot")`.
- **Hypothesis Visualization**, **Scatter Plots**, and **Pareto Charts**
  vignettes show the higher-level `plot_*` functions that are themselves
  assembled from these layer builders.
- The `iQualityR.core` package documents the `IqrTheme` and
  `IqrPlotterBase` classes that supply the colors and scales.

## References

Wickham, H. (2016). *ggplot2: Elegant Graphics for Data Analysis* (2nd ed.).
Springer. <https://ggplot2-book.org>

Wilkinson, L. (2005). *The Grammar of Graphics* (2nd ed.). Springer, New York.

Montgomery, D. C. (2013). *Introduction to Statistical Quality Control*
(7th ed.). John Wiley & Sons, Hoboken, NJ. -- control limits, Western
Electric run rules, and capability analysis.

Western Electric Company. (1956). *Statistical Quality Control Handbook*
(1st ed.). Western Electric, Indianapolis, IN. -- the original source of
the zone-based run rules.

Juran, J. M., & Godfrey, A. B. (1999). *Juran's Quality Handbook*
(5th ed.). McGraw-Hill, New York. -- specification versus control limits
and the distinction between stability and capability.
