---
title: "Layer Builders"
author: "iQualityR Team"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Layer Builders}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

# Layer Builders

## Overview

The underlying architecture of `iQualityR.plot` is based on the ggplot2 layer system, providing a series of reusable layer builder functions. These functions are the foundation of high-level plotting functions and can also be called directly by users to customize plots.

This chapter introduces:

- `layers_histogram_density()` - Histogram + density curve layer
- `layers_qq()` - QQ plot layer
- `layers_boxplot()` - Box plot layer
- `layers_trend_line()` - Trend line layer
- `layers_spec_limits()` - Specification limits layer
- The architectural philosophy of the layer system

## Load Data

```{r setup}
library(ggplot2)
library(datasets)
library(iQualityR.core)

library(iQualityR.plot)



data("mtcars")
data("iris")
```

## 1. Layer System Architecture

### What is a Layer?

The core idea of ggplot2 is the "Grammar of Graphics", which decomposes a plot into multiple independent layers:

- **Data layer**: Defines data and aesthetic mappings
- **Geometry layer**: Defines plot types (points, lines, areas, etc.)
- **Statistics layer**: Defines statistical transformations (smoothing, binning, etc.)
- **Coordinate layer**: Defines the coordinate system
- **Facet layer**: Defines facet layout
- **Theme layer**: Defines visual style

### iQualityR Layer Functions

`iQualityR.plot` encapsulates common plot combinations as layer functions that can be directly added to base plots:

```{r layer_concept, eval=FALSE}
# Conceptual example
base_plot(data, aes(x, y)) +
  layers_histogram_density() +  # Add histogram + density curve
  layers_qq() +                  # Add QQ plot elements
  theme_iqr()                    # Apply theme
```

## 2. Histogram + Density Curve Layer

`layers_histogram_density()` is a commonly used distribution visualization layer combination.

### Basic Usage

```{r histogram_density, warning=FALSE,dev="png"}
p <- base_plot(mtcars, aes(x = mpg)) +
  layers_histogram_density(bins = 15,
                           hist_args = list(fill = "#3498db", color = "white", alpha = 0.7),
                           density_args = list(color = "#e74c3c", linewidth = 1.2)) +
  labs(
    title = "Car Fuel Efficiency Distribution",
    x = "MPG",
    y = "Density"
  )

print(p)
```

### Adding a Normal Curve

```{r histogram_normal, warning=FALSE}
x_vals <- seq(min(mtcars$mpg), max(mtcars$mpg), length.out = 100)
normal_df <- data.frame(
  x = x_vals,
  y = dnorm(x_vals, mean = mean(mtcars$mpg), sd = sd(mtcars$mpg))
)

p_normal <- base_plot(mtcars, aes(x = mpg)) +
  layers_histogram_density(bins = 12,
                           hist_args = list(fill = "#2ecc71", color = "white", alpha = 0.6),
                           density_args = list(color = "#e74c3c", linewidth = 1.2)) +
  geom_line(data = normal_df, aes(x = x, y = y),
            color = "#9b59b6", linewidth = 1.2, linetype = "dashed") +
  labs(
    title = "MPG Distribution with Normal Fit",
    x = "MPG",
    y = "Density"
  )

print(p_normal)
```

## 3. QQ Plot Layer

The QQ plot layer is used to quickly create quantile-quantile plots.

### Basic Usage

```{r qq_layer, warning=FALSE}
p_qq <- base_plot(mtcars, aes(sample = mpg)) +
  layers_qq(distribution = "norm",
            qq_args = list(color = "#3498db", size = 3, alpha = 0.7),
            line_args = list(color = "#e74c3c", linetype = "dashed", linewidth = 1)) +
  labs(
    title = "MPG Normal QQ Plot",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  )

print(p_qq)
```

## 4. Box Plot Layer

The box plot layer is used to display the five-number summary of data distribution.

### Basic Usage

```{r boxplot_layer, warning=FALSE}
p_box <- base_plot(mtcars, aes(x = factor(cyl), y = mpg)) +
  layers_boxplot(add_jitter = TRUE,
                 boxplot_args = list(fill = "#e67e22", alpha = 0.7,
                                     outlier.color = "#e74c3c", outlier.size = 3),
                 jitter_args = list(width = 0.15, alpha = 0.5, size = 2,
                                    color = "#2c3e50")) +
  labs(
    title = "MPG Box Plot by Cylinder Count",
    x = "Cylinders",
    y = "MPG"
  )

print(p_box)
```

### Adding Statistical Annotations

```{r boxplot_stats,   warning=FALSE}
# Add statistical annotations (mean, median, etc.)
library(dplyr)

stats <- mtcars %>%
  group_by(cyl) %>%
  summarise(
    mean_mpg = mean(mpg),
    median_mpg = median(mpg),
    .groups = "drop"
  )

p_box +
  geom_point(data = stats, 
             aes(x = factor(cyl), y = mean_mpg),
             color = "red", size = 4, shape = 17) +
  geom_text(data = stats,
            aes(x = factor(cyl), y = mean_mpg, 
                label = sprintf("Mean: %.1f", mean_mpg)),
            vjust = -1, color = "blue", size = 4)
```

## 5. Custom Layer Combinations

The greatest advantage of the layer system is the ability to freely combine layers to create plots that meet specific needs.

### Combination Example: Distribution Diagnostic Plot

```{r custom_combo, warning=FALSE}
x_vals <- seq(min(mtcars$mpg), max(mtcars$mpg), length.out = 100)
normal_df <- data.frame(
  x = x_vals,
  y = dnorm(x_vals, mean = mean(mtcars$mpg), sd = sd(mtcars$mpg))
)

p_diagnostic <- base_plot(mtcars, aes(x = mpg)) +
  layers_histogram_density(bins = 10,
                           hist_args = list(fill = "#3498db", color = "white", alpha = 0.5),
                           density_args = list(fill = "#e74c3c", alpha = 0.2,
                                               color = "#e74c3c", linewidth = 1.2)) +
  geom_line(data = normal_df, 
            aes(x = x, y = y),
            color = "#2c3e50",
            linewidth = 1,
            linetype = "dashed") +
  geom_vline(xintercept = mean(mtcars$mpg),
             color = "#e67e22",
             linewidth = 1,
             linetype = "dotted") +
  labs(
    title = "MPG Distribution Diagnostic Plot",
    subtitle = sprintf("Mean = %.1f, SD = %.1f, n = %d",
                       mean(mtcars$mpg), sd(mtcars$mpg), nrow(mtcars)),
    x = "MPG",
    y = "Density"
  )

print(p_diagnostic)
```

## 6. Relationship Between Layers and High-Level Functions

The high-level plotting functions of `iQualityR.plot` are all built on these layers:

### Layer Structure of plot_qq()

```{r qq_structure, eval=FALSE}
# plot_qq() internally uses the following layers:
# 1. geom_point() - Sample quantile points
# 2. geom_abline() - Reference line
# 3. geom_ribbon() - Confidence band (optional)
# 4. annotate() - Statistical test annotation
```

### Layer Structure of plot_scatter_basic()

```{r scatter_structure, eval=FALSE}
# plot_scatter_basic() internally uses the following layers:
# 1. geom_point() - Scatter points
# 2. geom_smooth(method = "lm") - Regression line
# 3. annotate() - Correlation coefficient and equation annotation
```

## Practical Case Studies

### Case 1: Multi-Variable Distribution Comparison

```{r case_multivariate, warning=FALSE}
# Compare MPG distribution by cylinder count
p_multi <- base_plot(mtcars, aes(x = mpg, fill = factor(cyl))) +
  geom_density(alpha = 0.5) +
  labs(
    title = "MPG Density Distribution by Cylinder Count",
    x = "MPG",
    y = "Density",
    fill = "Cylinders"
  ) +
  scale_fill_brewer(palette = "Set1")

print(p_multi)
```

### Case 2: Process Capability Visualization

```{r case_capability, warning=FALSE}
set.seed(123)
process_data <- rnorm(100, mean = 50, sd = 2)
process_df <- data.frame(value = process_data)

USL <- 56
LSL <- 44

p_capa <- base_plot(process_df, aes(x = value)) +
  layers_histogram_density(bins = 15,
                           hist_args = list(fill = "#3498db", color = "white", alpha = 0.6),
                           density_args = list(color = "#e74c3c", linewidth = 1.2)) +
  layers_spec_limits(lsl = LSL, usl = USL) +
  geom_vline(xintercept = mean(process_data), 
             color = "#2c3e50", linewidth = 1) +
  labs(
    title = "Process Capability Visualization",
    subtitle = sprintf("Cp = %.2f, Cpk = %.2f",
                       (USL - LSL) / (6 * sd(process_data)),
                       min((USL - mean(process_data)), 
                           (mean(process_data) - LSL)) / (3 * sd(process_data))),
    x = "Measurement Value",
    y = "Density"
  )

print(p_capa)
```

## Best Practices

### 1. Layer Order

```{r best_practice_order, eval=FALSE}
# Recommended layer addition order:
# 1. Geometry layer (points, lines, areas)
# 2. Statistics layer (smoothing, binning)
# 3. Annotation layer (text, labels)
# 4. Theme layer (style)

base_plot(data, aes(x, y)) +
  geom_point() +           # Geometry layer
  geom_smooth() +          # Statistics layer
  annotate() +             # Annotation layer
  theme_iqr()              # Theme layer
```

### 2. Layer Reuse

```{r best_practice_reuse, eval=FALSE}
# Create reusable layer combinations
my_layers <- list(
  geom_point(color = "#3498db", alpha = 0.7),
  geom_smooth(method = "lm", color = "#e74c3c"),
  theme_minimal()
)

# Reuse across different plots
p1 <- base_plot(data1, aes(x, y)) + my_layers
p2 <- base_plot(data2, aes(a, b)) + my_layers
```

### 3. Performance Optimization

```{r best_practice_performance, eval=FALSE}
# For large datasets, use stat_summary instead of geom_point
base_plot(large_data, aes(x, y)) +
  stat_summary(fun = mean, geom = "point", size = 2) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2)
```

## FAQ

### Q: How to create a custom layer function?

```{r faq_custom_layer, eval=FALSE}
# Create a custom layer function
my_custom_layer <- function(color = "blue", size = 1) {
  list(
    geom_point(color = color, size = size),
    geom_smooth(method = "lm", color = color)
  )
}

# Usage
base_plot(data, aes(x, y)) + my_custom_layer("red", 2)
```

### Q: What to do about layer conflicts?

If multiple layers use the same aesthetic mapping, the later-added layer will override the previous one. Use `inherit.aes = FALSE` to avoid inheritance conflicts:

```{r faq_conflict, eval=FALSE}
geom_line(data = other_df, aes(x = x, y = y), inherit.aes = FALSE)
```

## Summary

- The layer system is the core architecture of ggplot2 and `iQualityR.plot`
- Layer functions can be used individually or combined to create complex plots
- High-level plotting functions are all built on layers
- Understanding the layer system helps in creating custom visualizations

## Further Reading

- [Basic Plotting and Theme System](basic_plotting_theme.html) - Learn about the theme system
- [Scatter Plots](scatter_plots.html) - Learn about scatter plot plotting
- [Advanced Analysis Plots](advanced_plots.html) - Learn about correlation heatmaps and interaction plots
