---
title: "Pareto Charts"
author: "iQualityR Team"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Pareto Charts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

# Pareto Charts

## Overview

The Pareto Chart is one of the most important tools in quality management. Based on the "80/20 rule" (Pareto principle), it helps identify the few critical causes that lead to the majority of problems. `iQualityR.plot` provides multiple Pareto chart functions:

- `plot_pareto_enhanced()` - Enhanced Pareto chart (supports grouping, weighting, summary tables)
- `quick_pareto()` - Quick Pareto chart
- `plot_pareto_image_labels()` - Pareto chart with image labels
- `plot_pareto_emoji_labels()` - Pareto chart with Emoji labels

## Load Data

```{r setup}
knitr::opts_chunk$set(warning = FALSE)
library(ggplot2)
library(iQualityR.plot)
library(iQualityR.core)

```

## 1. Enhanced Pareto Chart: plot_pareto_enhanced()

This is the most feature-rich Pareto chart function, supporting multiple input formats and advanced features.

### Create from Data Frame

```{r pareto_dataframe}
# Defect type statistics
defect_data <- data.frame(
  defect = c("Scratch", "Dent", "Missing part", "Color issue", "Dimension error", "Weight issue"),
  count = c(45, 23, 15, 8, 5, 3)
)

plot_pareto_enhanced(
  data = defect_data,
  category_col = "defect",
  count_col = "count",
  title = "Product Defect Pareto Analysis",
  subtitle = "Identifying Key Defect Types"
)
```

### Create from Named Vector

```{r pareto_vector}
# Using a named vector (more concise)
defect_counts <- c(
  "Scratch" = 45,
  "Dent" = 23,
  "Missing part" = 15,
  "Color issue" = 8,
  "Dimension error" = 5,
  "Weight issue" = 3
)

plot_pareto_enhanced(
  data = defect_counts,
  title = "Product Defect Pareto Chart",
  subtitle = "Data source: Quality inspection records"
)
```

### Display Summary Table

```{r pareto_with_table, eval=FALSE}
# Display a statistics table below the chart
plot_pareto_enhanced(
  data = defect_counts,
  show_table = TRUE,
  title = "Pareto Chart with Statistics Table"
)
```

### Grouped Pareto Chart

```{r pareto_grouped}
# Compare defect distribution across different periods
grouped_data <- data.frame(
  defect = rep(c("Scratch", "Dent", "Missing part", "Color issue"), 2),
  count = c(45, 23, 15, 8, 30, 18, 10, 5),
  period = rep(c("Before improvement", "After improvement"), each = 4)
)

plot_pareto_enhanced(
  data = grouped_data,
  category_col = "defect",
  count_col = "count",
  group_col = "period",
  title = "Defect Comparison Before and After Improvement",
  subtitle = "Grouped Pareto Analysis"
)
```

### "Other" Category Aggregation

```{r pareto_other}
# Aggregate small categories into "Other"
small_defects <- c(
  "Scratch" = 45,
  "Dent" = 23,
  "Missing part" = 15,
  "Color issue" = 8,
  "Dimension error" = 5,
  "Weight issue" = 3,
  "Packaging issue" = 2,
  "Label error" = 1
)

plot_pareto_enhanced(
  data = small_defects,
  threshold = 0.05,  # Categories after cumulative 95% are grouped as "Other"
  title = "Pareto Chart with 'Other' Aggregation",
  subtitle = "threshold = 0.05"
)
```

### Weighted Pareto Analysis

```{r pareto_weighted}
# Weighted analysis considering defect severity
weighted_data <- data.frame(
  defect = c("Scratch", "Dent", "Missing part", "Color issue"),
  count = c(45, 23, 15, 8),
  severity = c(2, 3, 5, 1)  # Severity weight
)

plot_pareto_enhanced(
  data = weighted_data,
  category_col = "defect",
  count_col = "count",
  weight_col = "severity",
  title = "Weighted Pareto Analysis",
  subtitle = "Considering Defect Severity"
)
```

## 2. Quick Pareto Chart: quick_pareto()

`quick_pareto()` is a shortcut version of `plot_pareto_enhanced()`, suitable for quick exploratory analysis.

### Basic Usage

```{r quick_pareto_basic}
# Quick creation from a named vector
quick_pareto(
  data = c("A" = 30, "B" = 25, "C" = 20, "D" = 15, "E" = 10),
  title = "Quick Pareto Chart"
)
```

### Quick Creation from Data Frame

```{r quick_pareto_df}
# Quick creation from a data frame
df <- data.frame(
  category = c("Process issue", "Equipment issue", "Material issue", "Personnel issue", "Environment issue"),
  count = c(35, 28, 20, 12, 5)
)

quick_pareto(
  data = df,
  category_col = "category",
  count_col = "count",
  title = "Quality Issue Root Cause Analysis"
)
```

## 3. Pareto Chart with Image Labels: plot_pareto_image_labels()

In quality management, using actual defect images can enhance the persuasiveness of reports.

### Basic Usage

```{r pareto_image, eval=FALSE}
# Prepare defect image paths
defect_data <- data.frame(
  defect = c("Scratch", "Dent", "Missing part"),
  count = c(45, 23, 15)
)

image_paths <- c(
  "Scratch" = "path/to/scratch.png",
  "Dent" = "path/to/dent.png",
  "Missing part" = "path/to/missing.png"
)

plot_pareto_image_labels(
  data = defect_data,
  category_col = "defect",
  count_col = "count",
  image_paths = image_paths,
  image_size = 30,
  title = "Pareto Chart with Defect Images"
)
```

## 4. Pareto Chart with Emoji Labels: plot_pareto_emoji_labels()

Use Emoji icons to make charts more vivid and intuitive.

### Basic Usage

```{r pareto_emoji, eval=FALSE, warning=FALSE}
# Use text labels to represent defect types
emoji_data <- data.frame(
  defect = c("Scratch", "Dent", "Missing part", "Color issue"),
  count = c(45, 23, 15, 8)
)

label_map <- c(
  "Scratch" = "[Scratch]",
  "Dent" = "[Dent]",
  "Missing part" = "[Missing]",
  "Color issue" = "[Color]"
)

plot_pareto_emoji_labels(
  data = emoji_data,
  category_col = "defect",
  count_col = "count",
  emojis = label_map,
  emoji_size = 5,
  title = "Pareto Chart with Text Labels"
)
```

## Practical Case Studies

### Case 1: Manufacturing Defect Analysis

```{r case_manufacturing, warning=FALSE}
# Defect statistics from a manufacturing plant
manufacturing_defects <- c(
  "Surface scratch" = 52,
  "Dimension deviation" = 31,
  "Welding defect" = 24,
  "Uneven coating" = 15,
  "Assembly error" = 12,
  "Material defect" = 8,
  "Other" = 5
)

plot_pareto_enhanced(
  data = manufacturing_defects,
  threshold = 0.05,
  title = "Manufacturing Defect Pareto Analysis",
  subtitle = "Identifying Key Quality Improvement Areas"
)
```

### Case 2: Service Industry Customer Complaint Analysis

```{r case_service, warning=FALSE}
# Customer complaint type statistics
complaints <- data.frame(
  type = c("Slow response", "Poor attitude", "Unresolved", "Complex process", "Inaccurate info"),
  count = c(120, 85, 65, 45, 30),
  department = c("Customer Service", "Customer Service", "Technical", "Operations", "Operations")
)

plot_pareto_enhanced(
  data = complaints,
  category_col = "type",
  count_col = "count",
  group_col = "department",
  title = "Customer Complaint Pareto Analysis",
  subtitle = "Grouped by Department"
)
```

### Case 3: Software Defect Priority Analysis

```{r case_software, warning=FALSE}
# Software bug statistics (considering severity weight)
software_bugs <- data.frame(
  bug_type = c("Crash", "Functional error", "Performance issue", "UI issue", "Compatibility issue"),
  count = c(15, 45, 28, 60, 20),
  severity = c(5, 4, 3, 1, 2)  # 1-5 severity
)

plot_pareto_enhanced(
  data = software_bugs,
  category_col = "bug_type",
  count_col = "count",
  weight_col = "severity",
  title = "Software Defect Weighted Pareto Analysis",
  subtitle = "Weight = Severity (1-5)"
)
```

## Parameter Details

### plot_pareto_enhanced() Parameters

| Parameter | Type | Description | Default |
|------|------|------|--------|
| `data` | data.frame/list/vector | Input data | Required |
| `category_col` | character | Category column name | NULL |
| `count_col` | character | Count column name | NULL |
| `group_col` | character | Grouping column name (optional) | NULL |
| `weight_col` | character | Weight column name (optional) | NULL |
| `threshold` | numeric | "Other" aggregation threshold | 0.05 |
| `show_table` | logical | Whether to show table | FALSE |
| `theme` | character | Theme name | "academic" |
| `title` | character | Chart title | NULL |
| `subtitle` | character | Subtitle | NULL |

### threshold Parameter Description

```{r threshold_explain, eval=FALSE}
# threshold = 0.05: Categories after cumulative percentage reaches 95% are grouped as "Other"
# threshold = 0.10: Categories after cumulative percentage reaches 90% are grouped as "Other"
# threshold = 0: No aggregation, show all categories

plot_pareto_enhanced(data, threshold = 0.10)  # More aggressive aggregation
plot_pareto_enhanced(data, threshold = 0)     # No aggregation
```

## Best Practices

### 1. Choose the Appropriate Threshold

```{r best_practice_threshold, eval=FALSE}
# When there are many categories (> 10), use threshold aggregation
plot_pareto_enhanced(data, threshold = 0.05)

# When there are few categories (< 8), no aggregation
plot_pareto_enhanced(data, threshold = 0)
```

### 2. Use Weighted Analysis

```{r best_practice_weight, eval=FALSE}
# When different categories have different impact levels, use weighting
weighted_data <- data.frame(
  category = c("A", "B", "C"),
  count = c(100, 50, 30),
  impact = c(1, 3, 5)  # Impact level
)

plot_pareto_enhanced(weighted_data, weight_col = "impact")
```

### 3. Grouped Comparison

```{r best_practice_group, eval=FALSE}
# Compare changes before and after improvement
before <- data.frame(defect = defects, count = before_counts, period = "Before improvement")
after <- data.frame(defect = defects, count = after_counts, period = "After improvement")

combined <- rbind(before, after)
plot_pareto_enhanced(combined, group_col = "period")
```

## FAQ

### Q: How to customize colors?

Use the theme parameter:

```{r faq_color, eval=FALSE}
plot_pareto_enhanced(data, theme = "tech")  # Tech style
plot_pareto_enhanced(data, theme = "prism") # Prism style
```

### Q: What if the data contains negative values?

Pareto charts require positive values; filter first:

```{r faq_negative, eval=FALSE}
valid_data <- data[data$count > 0, ]
plot_pareto_enhanced(valid_data, ...)
```

### Q: How to export high-quality images?

```{r faq_export, eval=FALSE}
p <- plot_pareto_enhanced(data)
ggsave("pareto.png", p, width = 10, height = 6, dpi = 300)
```

## Summary

- `plot_pareto_enhanced()` is the most feature-rich Pareto chart function
- Supports multiple input formats: data frame, named vector, list
- Grouping feature supports comparison analysis across different groups
- Weighted analysis can consider the impact level of different categories
- "Other" aggregation simplifies the chart and highlights key categories

## Further Reading

- [Scatter Plots](scatter_plots.html) - Learn about scatter plot plotting
- [Fishbone Diagrams](fishbone_diagrams.html) - Learn about root cause analysis tools
- [Basic Plotting and Theme System](basic_plotting_theme.html) - Learn about the theme system
