---
title: "Pareto Charts for Defect Prioritization"
author: "iQualityR.plot Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Pareto Charts for Defect Prioritization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

library(iQualityR.plot)

library(iQualityR.core)

library(ggplot2)

fmt_num <- function(x, digits = 4) formatC(x, format = "f", digits = digits)
fmt_pct <- function(x, digits = 2) paste0(formatC(x, format = "f", digits = digits), "%")
```

## 1. Introduction

In industrial quality management, defect data is rarely homogeneous: a handful of
defect modes typically account for the bulk of nonconformities, while a long tail of
rare modes consumes disproportionate engineering attention. The **Pareto chart** is
the canonical tool for exposing this imbalance. It couples a sorted bar chart of
defect frequencies with a cumulative-percentage line, so that engineers can see at a
glance which categories to attack first.

The `iQualityR.plot` package provides a family of Pareto functions that go beyond the
textbook recipe:

* `plot_pareto_enhanced()` supports grouping, weighting, "Other" aggregation, and an
  optional summary table.
* `quick_pareto()` accepts data frames, lists, or named vectors for rapid analysis.
* `plot_pareto_image_labels()` and `plot_pareto_emoji_labels()` replace plain axis
  text with images or emoji, which is useful for shop-floor dashboards where defect
  photographs are more recognizable than text labels.

This vignette develops the theory, walks through each function, and finishes with a
worked example from printed circuit board (PCB) manufacturing. Every function in
the Pareto family accepts a `theme` argument (a theme name such as
`"academic"` or an `IqrTheme` object) and resolves its bar fill, cumulative
line, and table shading from the active theme through `.iqr_aes()`, so users
rarely need to specify colors explicitly.

## 2. Theory: The 80/20 Rule and Pareto Analysis

### 2.1 The Pareto principle in quality management

The Pareto principle, popularly known as the *80/20 rule*, states that roughly 80%
of observed effects originate from 20% of the causes. The principle is named after
the Italian economist Vilfredo Pareto, who observed that 80% of Italian land was
owned by 20% of the population. Joseph M. Juran adapted the idea to quality
improvement, coining the phrases *vital few* and *useful many*.

For a defect dataset with $n$ categories, let $c_i$ be the count of the $i$-th
category sorted in descending order. The total count is

$$
C = \sum_{i=1}^{n} c_i
$$

and the cumulative share of the top $k$ categories is

$$
P_k = \frac{1}{C}\sum_{i=1}^{k} c_i \times 100\%.
$$

The "vital few" are the smallest $k$ such that $P_k \ge 80\%$. The Pareto chart
makes this cutoff visually obvious: the cumulative line crosses the 80% reference
near the top of the sorted bars.

### 2.2 Constructing a Pareto chart (cumulative percentage)

A standard Pareto chart is built in four steps:

1. **Tally** the count (or cost) of each defect category.
2. **Sort** categories in descending order of count.
3. **Plot** the sorted counts as a vertical bar chart with the left axis showing
   the count.
4. **Overlay** the cumulative percentage $P_k$ as a line on a secondary right axis
   scaled from 0% to 100%.

The dual-axis construction means that the engineer reads *frequency* from the bars
and *coverage* from the line. A typical reading is: "the top three categories cover
78% of all defects — fixing those three will eliminate almost four out of every
five nonconformities."

#### Aggregating the long tail

When many small categories exist, the chart becomes cluttered and the cumulative
line flattens prematurely. A common remedy is to collapse categories whose
individual share is below a threshold $\tau$ into a single "Other" bar. In
`iQualityR.plot` the threshold is controlled by the `threshold` argument (default
`0.05`, i.e. 5%). Formally, a category $i$ is folded into "Other" when

$$
\frac{c_i}{C} < \tau.
$$

#### Weighted Pareto

When defects carry different severity or cost, raw frequency can be misleading. A
short-run defect that scraps a $500 board should not be ranked below a cosmetic
scratch that affects thousands of cheap parts. A **weighted Pareto** replaces $c_i$
with the impact score

$$
w_i = c_i \cdot s_i,
$$

where $s_i$ is the severity, cost, or any per-unit weight supplied by the engineer.
The cumulative line is then computed on $w_i$ rather than $c_i$, prioritizing
categories by *total impact* rather than by *count*.

## 3. plot_pareto_enhanced() — with grouping and weights

`plot_pareto_enhanced()` is the flagship Pareto function. It accepts a data frame,
a list, or a named vector, and returns a `ggplot` object (or a `patchwork`
composition when `show_table = TRUE`).

```{r}
str(plot_pareto_enhanced)
```

The key arguments are:

| Argument       | Purpose                                                    |
|----------------|------------------------------------------------------------|
| `data`         | A data.frame, list, or named vector of counts.             |
| `category_col` | Name of the category column.                               |
| `count_col`    | Name of the count / value column.                          |
| `group_col`    | Optional grouping column for facetting.                    |
| `weight_col`   | Optional weight column for weighted Pareto.                |
| `threshold`    | Share below which categories are folded into "Other".      |
| `show_table`   | If `TRUE`, append a summary table below the chart.         |
| `theme`        | Theme name (e.g. `"academic"`, `"economist"`, `"prism"`) or `IqrTheme` object; drives bar fill, cumulative line, and table shading. |
| `title`        | Optional plot title.                                       |
| `subtitle`     | Optional plot subtitle.                                    |

### A minimal example

We start with a small synthetic dataset of surface defects on a metal-stamping
line.

```{r}
stamp_defects <- data.frame(
  defect_type = c("Scratch", "Dent", "Burr", "Rust", "Crack", "Warpage"),
  count = c(120, 86, 54, 30, 12, 8)
)

p <- plot_pareto_enhanced(
  data         = stamp_defects,
  category_col = "defect_type",
  count_col    = "count",
  title        = "Surface Defects on Stamping Line A",
  subtitle     = "Single shift, n = 310 units inspected"
)
print(p)
```

The chart shows that the top two categories — `Scratch` and `Dent` — together
account for the majority of defects, while `Warpage` and `Crack` fall below the
5% threshold and would be folded into "Other" on a larger dataset.

### Weighted Pareto with severity

When severity scores are available, the `weight_col` argument re-ranks the bars by
total impact.

```{r}
stamp_defects$severity <- c(3, 5, 4, 7, 9, 8)

p_w <- plot_pareto_enhanced(
  data         = stamp_defects,
  category_col = "defect_type",
  count_col    = "count",
  weight_col   = "severity",
  title        = "Weighted Pareto: Impact = Count x Severity",
  subtitle     = "Severity on a 1-10 scale"
)
print(p_w)
```

Note how `Crack` (high severity, low count) climbs in the ranking once weights are
applied — this is exactly the insight that a frequency-only chart hides.

### Grouping with facets

When the same defect taxonomy is observed across multiple lines, shifts, or
products, the `group_col` argument produces a small-multiples layout.

```{r}
stamp_grouped <- data.frame(
  defect_type = rep(c("Scratch", "Dent", "Burr", "Rust"), times = 2),
  count       = c(120, 86, 54, 30, 70, 95, 40, 18),
  line        = rep(c("Line A", "Line B"), each = 4)
)

p_g <- plot_pareto_enhanced(
  data         = stamp_grouped,
  category_col = "defect_type",
  count_col    = "count",
  group_col    = "line",
  title        = "Pareto by Production Line",
  subtitle     = "Top defect mode differs between lines"
)
print(p_g)
```

### Summary table

Setting `show_table = TRUE` appends a formatted summary table — sorted counts,
percentages, and cumulative percentages — beneath the chart. The return value
becomes a `patchwork` object rather than a bare `ggplot`.

```{r}
pw <- plot_pareto_enhanced(
  data         = stamp_defects,
  category_col = "defect_type",
  count_col    = "count",
  show_table   = TRUE,
  title        = "Pareto with Summary Table"
)
print(pw)
```

## 4. quick_pareto() — rapid input variants

`quick_pareto()` is a thin wrapper that infers the input shape and forwards
arguments to `plot_pareto_enhanced()`. It is useful for ad-hoc analysis at the
console.

```{r}
str(quick_pareto)
```

### From a named vector

```{r}
defect_vec <- c(
  Solder = 145,
  Missing = 98,
  Align = 72,
  Short = 56,
  Open = 38
)

quick_pareto(defect_vec, title = "Quick Pareto from Named Vector")
```

### From a list

A list input must carry two parallel vectors: `category` (the defect names)
and `count` (the corresponding frequencies).

```{r}
defect_list <- list(
  category = c("Solder", "Missing", "Align", "Short", "Open"),
  count    = c(145, 98, 72, 56, 38)
)

quick_pareto(defect_list, title = "Quick Pareto from List")
```

### From a data frame

```{r}
quick_pareto(
  data         = stamp_defects,
  category_col = "defect_type",
  count_col    = "count",
  title        = "Quick Pareto from Data Frame"
)
```

## 5. plot_pareto_image_labels() / plot_pareto_emoji_labels() — visual labels

On the shop floor, operators recognize defect photographs faster than they read
text labels. Two convenience functions replace the X-axis text with images or
emoji.

### Emoji labels

`plot_pareto_emoji_labels()` takes a named vector `emojis` that maps each category
to an emoji glyph.

```{r}
defects_emoji <- data.frame(
  defect_type = c("Solder Bridge", "Component Missing", "Misalignment", "Short Circuit"),
  count = c(145, 98, 72, 56)
)

emoji_map <- c(
  "Solder Bridge"     = "\U0001F527",
  "Component Missing" = "\U0001F504",
  "Misalignment"      = "\U0001F4D0",
  "Short Circuit"     = "\U000026A1"
)

plot_pareto_emoji_labels(
  data         = defects_emoji,
  category_col = "defect_type",
  count_col    = "count",
  emojis       = emoji_map,
  emoji_size   = 6,
  title        = "Pareto with Emoji Labels",
  subtitle     = "Useful for operator-facing dashboards"
)
```

### Image labels

`plot_pareto_image_labels()` accepts a named vector of file paths. In this vignette
we point at placeholder paths; in production these would be photographs of the
actual defect modes.

```{r, eval = FALSE}
image_map <- c(
  "Solder Bridge"     = "images/solder_bridge.png",
  "Component Missing" = "images/component_missing.png",
  "Misalignment"      = "images/misalignment.png",
  "Short Circuit"     = "images/short_circuit.png"
)

plot_pareto_image_labels(
  data         = defects_emoji,
  category_col = "defect_type",
  count_col    = "count",
  image_paths  = image_map,
  image_size   = 22,
  title        = "Pareto with Image Labels"
)
```

The `image_size` and `emoji_size` arguments control the glyph / image size in
points; tune them together with `fig.width` so the labels do not overlap.

## 6. Quality Scenario: defect types in PCB manufacturing

We now bring the tools together on a realistic dataset from a PCB assembly line.
The line records eight defect modes and a 1-10 severity score per mode.

```{r}
pcb_defects <- data.frame(
  defect_type = c("Solder Bridge", "Component Missing", "Misalignment",
                   "Short Circuit", "Open Circuit", "Wrong Component",
                   "Cold Solder", "Solder Ball"),
  count = c(145, 98, 72, 56, 38, 22, 15, 8),
  severity = c(8, 7, 5, 9, 8, 6, 4, 3)  # 1-10 severity scale
)

# Total defect count and basic summary
total_defects <- sum(pcb_defects$count)
total_impact  <- sum(pcb_defects$count * pcb_defects$severity)

cat("Total defects inspected :", total_defects, "\n")
cat("Total weighted impact   :", total_impact, "\n")
```

### Step 1 — Frequency Pareto

The frequency Pareto reveals which defect modes occur most often, regardless of
severity.

```{r}
p_freq <- plot_pareto_enhanced(
  data         = pcb_defects,
  category_col = "defect_type",
  count_col    = "count",
  threshold    = 0.05,
  title        = "PCB Defect Frequency Pareto",
  subtitle     = paste0("n = ", total_defects, " defects inspected")
)
print(p_freq)
```

Reading the chart: `Solder Bridge` and `Component Missing` together cover roughly
half of all defects. The bottom three modes (`Cold Solder`, `Solder Ball`, and
`Wrong Component`) each contribute less than 5% and would be aggregated into
"Other" by the default threshold.

### Step 2 — Weighted Pareto by severity

Severity reshuffles the priorities. `Short Circuit` carries severity 9, so even
though it is only the fourth-most-frequent mode, its total impact climbs near the
top.

```{r}
p_impact <- plot_pareto_enhanced(
  data         = pcb_defects,
  category_col = "defect_type",
  count_col    = "count",
  weight_col   = "severity",
  threshold    = 0.05,
  title        = "PCB Defect Impact Pareto (Count x Severity)",
  subtitle     = "Severity reorders the vital few"
)
print(p_impact)
```

### Step 3 — Pareto with summary table

For the quality team's weekly report, we attach the summary table.

```{r}
p_tbl <- plot_pareto_enhanced(
  data         = pcb_defects,
  category_col = "defect_type",
  count_col    = "count",
  weight_col   = "severity",
  show_table   = TRUE,
  title        = "PCB Defect Pareto with Summary Table"
)
print(p_tbl)
```

### Step 4 — Operator-facing emoji variant

For the line-side display board we switch to emoji labels.

```{r}
pcb_emoji <- c(
  "Solder Bridge"     = "\U0001F527",
  "Component Missing" = "\U0001F504",
  "Misalignment"      = "\U0001F4D0",
  "Short Circuit"     = "\U000026A1",
  "Open Circuit"      = "\U0001F50C",
  "Wrong Component"   = "\U00002757",
  "Cold Solder"       = "\U00002744",
  "Solder Ball"       = "\U0001F7E2"
)

plot_pareto_emoji_labels(
  data         = pcb_defects,
  category_col = "defect_type",
  count_col    = "count",
  emojis       = pcb_emoji,
  emoji_size   = 5,
  title        = "PCB Defect Pareto for Operator Display",
  subtitle     = "Top 8 modes with severity-coded emoji"
)
```

### Step 5 — Theme-driven coloring

The Pareto family resolves its bar fill, cumulative line, and table
shading from the active `IqrTheme`. Switching `theme` repaints the
chart and the optional summary table in one stroke — no per-call color
arguments are required:

```{r, fig.show="hold", out.width="50%"}
plot_pareto_enhanced(
  data         = pcb_defects,
  category_col = "defect_type",
  count_col    = "count",
  weight_col   = "severity",
  theme        = "economist",
  title        = "Impact Pareto — economist"
)

plot_pareto_enhanced(
  data         = pcb_defects,
  category_col = "defect_type",
  count_col    = "count",
  weight_col   = "severity",
  show_table   = TRUE,
  theme        = "prism",
  title        = "Impact Pareto + table — prism"
)
```

### Interpretation

The frequency Pareto would lead the team to attack `Solder Bridge` first. The
weighted Pareto tells a more nuanced story: `Short Circuit` is rarer but more
severe, and its impact rivals `Component Missing`. A balanced improvement plan
should target the top three modes of *each* ranking — `Solder Bridge`,
`Component Missing`, and `Short Circuit` — because together they dominate both
frequency and impact.

## 7. Parameter Reference

The table below consolidates the arguments used across the Pareto family.

| Function                       | Argument        | Type / Default     | Description                                          |
|--------------------------------|-----------------|--------------------|------------------------------------------------------|
| `plot_pareto_enhanced`         | `data`          | data.frame / list  | Input data.                                          |
| `plot_pareto_enhanced`         | `category_col`  | `NULL`             | Category column name.                                |
| `plot_pareto_enhanced`         | `count_col`     | `NULL`             | Count / value column name.                           |
| `plot_pareto_enhanced`         | `group_col`     | `NULL`             | Grouping column for facetting.                       |
| `plot_pareto_enhanced`         | `weight_col`    | `NULL`             | Weight column for weighted Pareto.                   |
| `plot_pareto_enhanced`         | `threshold`     | `0.05`             | Share below which categories fold into "Other".      |
| `plot_pareto_enhanced`         | `show_table`    | `FALSE`            | Append summary table.                                |
| `plot_pareto_enhanced`         | `theme`         | `"academic"`       | Theme name or `IqrTheme` object; drives bar fill, cumulative line, and table shading via `.iqr_aes()`. |
| `plot_pareto_enhanced`         | `title`         | `NULL`             | Plot title.                                          |
| `plot_pareto_enhanced`         | `subtitle`      | `NULL`             | Plot subtitle.                                       |
| `quick_pareto`                 | `data`          | data.frame / list  | Input data.                                          |
| `quick_pareto`                 | `category_col`  | `NULL`             | Category column name.                                |
| `quick_pareto`                 | `count_col`     | `NULL`             | Count / value column name.                           |
| `quick_pareto`                 | `...`           |                    | Forwarded to `plot_pareto_enhanced` (including `theme`). |
| `plot_pareto_image_labels`     | `image_paths`   | named vector       | Category -> image file path.                         |
| `plot_pareto_image_labels`     | `image_size`    | `20`               | Image size in points.                                |
| `plot_pareto_image_labels`     | `theme`         | `"academic"`       | Theme name or `IqrTheme` object.                     |
| `plot_pareto_emoji_labels`     | `emojis`        | named vector       | Category -> emoji.                                   |
| `plot_pareto_emoji_labels`     | `emoji_size`    | `5`                | Emoji size in points.                                |
| `plot_pareto_emoji_labels`     | `theme`         | `"academic"`       | Theme name or `IqrTheme` object.                     |

### Return values

* `plot_pareto_enhanced()` returns a `ggplot` object by default, or a `patchwork`
  object when `show_table = TRUE`.
* `quick_pareto()` inherits the return type of `plot_pareto_enhanced()`.
* `plot_pareto_image_labels()` and `plot_pareto_emoji_labels()` return `ggplot`
  objects.

Because the return values are ggplot / patchwork objects, they can be composed
with other layers or saved with `ggplot2::ggsave()`.

```{r, eval = FALSE}
# Example: saving a Pareto chart to PNG
ggplot2::ggsave(
  filename = "pcb_pareto.png",
  plot     = p_freq,
  width    = 8,
  height   = 5,
  dpi      = 300
)
```

## 8. Summary

A Pareto chart is the first diagnostic a quality engineer should reach for when
faced with a stream of defect data. The `iQualityR.plot` Pareto family supports
the full workflow:

1. **Triage** with `quick_pareto()` from any input shape.
2. **Deep-dive** with `plot_pareto_enhanced()` for grouping, weighting, and the
   summary table.
3. **Communicate** with `plot_pareto_image_labels()` or
   `plot_pareto_emoji_labels()` for operator-facing displays.

The weighted variant is especially important: severity-aware prioritization
prevents the team from over-investing in frequent-but-cheap defects while ignoring
rare-but-expensive ones.

## 9. Further Reading

* Juran, J. M. (1951). *Quality Control Handbook*. McGraw-Hill. — Original
  articulation of the *vital few* / *useful many* concepts.
* Pareto, V. (1896). *Cours d'Economie Politique*. — The economic observation
  that inspired the principle.
* ISO 2859-1:1999. *Sampling procedures for inspection by attributes*. — Acceptance
  sampling context in which Pareto analysis is often applied.
* Pyzdek, T., & Keller, P. A. (2018). *The Six Sigma Handbook* (5th ed.).
  McGraw-Hill. — Modern treatment of Pareto analysis within DMAIC.
* `ggplot2` documentation on secondary axes and `patchwork` for composing charts
  with tables.

## References

Pareto, V. (1896). *Cours d'Economie Politique*. Lausanne: F. Rouge.

Juran, J. M. (1951). *Quality Control Handbook*. New York: McGraw-Hill.

Juran, J. M., & Godfrey, A. B. (1999). *Juran's Quality Handbook* (5th ed.).
New York: McGraw-Hill.

International Organization for Standardization. (1999). *ISO 2859-1:1999 —
Sampling procedures for inspection by attributes — Part 1*. Geneva: ISO.

Pyzdek, T., & Keller, P. A. (2018). *The Six Sigma Handbook* (5th ed.). New York:
McGraw-Hill.

Wickham, H. (2016). *ggplot2: Elegant Graphics for Data Analysis*. New York:
Springer.
