---
title: "Visualization Infrastructure: base_plot and the Theme Pipeline"
author: "iQualityR.plot Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Visualization Infrastructure: base_plot and the Theme Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

library(iQualityR.plot)
library(ggplot2)

fmt_num <- function(x, digits = 4) formatC(x, format = "f", digits = digits)
fmt_pct <- function(x, digits = 2) paste0(formatC(x, format = "f", digits = digits), "%")
```

## 1. Introduction

The `iQualityR.plot` sub-package provides the graphical foundation on which
every quality-engineering chart in the iQualityR ecosystem is built. Rather
than scattering theme and color logic across dozens of plotting functions, the
package centralizes it in a small, predictable pipeline:

- `base_plot()` creates a `ggplot` object with the iQualityR theme already
  attached.
- A small family of resolvers (`as_iqr_theme()`, `as_iqr_theme_object()`,
  `set_default_theme()`) convert any reasonable theme specification into a
  concrete `IqrTheme` object.
- A package-level singleton, `.iqr_plotter`, exposes the unified color
  toolbox that every `plot_*` function consults for palettes and scales.

This vignette walks through that pipeline end to end, using a steel-plate
thickness measurement scenario that recurs throughout the quality lab.

### 1.1 The grammar of iQualityR graphics

`iQualityR.plot` is built on top of the *grammar of graphics* implemented by
`ggplot2`. In that grammar a plot is not a single function call but the
composition of independent layers:

1. **Data layer** -- a data frame and a set of aesthetic mappings (`aes()`).
2. **Geometry layer** -- the geometric objects that represent the data
   (points, lines, bars, ribbons).
3. **Statistics layer** -- optional transformations (binning, smoothing,
   quantile computation).
4. **Scale layer** -- how data values are mapped to visual properties such as
   color, size, or position.
5. **Coordinate layer** -- the coordinate system (cartesian, fixed, flipped).
6. **Theme layer** -- the non-data ink: fonts, grid lines, backgrounds.

The iQualityR layer model adds two conventions on top of this grammar:

- A single entry point, `base_plot()`, attaches the iQualityR theme to the
  data and mapping so every downstream chart starts from the same baseline.
- A shared color toolbox (`.iqr_plotter`) keeps every `plot_*` function
  consistent: switching the active theme recolors every chart uniformly
  without touching individual function calls.

### 1.2 How plot functions pick colors (the IqrPlotterBase toolbox)

Every higher-level `plot_*` function resolves its colors through the same set
of accessor methods exposed by the `IqrPlotterBase` class:

| Accessor | Returns | Typical visual job |
|---|---|---|
| `.pal_discrete(theme_obj, n)` | discrete color vector | categorical groups |
| `.pal_sequential(theme_obj)` | 3-stop sequential gradient | ordered, low-to-high |
| `.pal_diverging(theme_obj)` | 3-stop diverging gradient | deviations from a midpoint |
| `.pal_semantic(theme_obj, name)` | named pass/fail/watch colors | acceptance regions |
| `.pal_ui(theme_obj, name, default)` | UI color (primary, muted, danger) | annotation ink |
| `.contrast_text(bg)` | black or white text | readable labels on fills |
| `.scale_fill_discrete(theme_obj)` | ggplot2 fill scale | bars, violins, boxes |
| `.scale_color_discrete(theme_obj)` | ggplot2 color scale | lines, points |

Because these accessors always consult the *active* `IqrTheme` object, a single
theme switch propagates consistently to histograms, control charts, Pareto
charts, and interaction plots alike.

## 2. Quick Start

We begin with a realistic data set: thickness measurements (in millimetres)
taken on 100 steel plates across four production batches.

```{r steel-data}
set.seed(42)
steel <- data.frame(
  thickness = rnorm(100, mean = 10, sd = 0.3),
  batch     = rep(c("A", "B", "C", "D"), each = 25)
)
head(steel)
#>   thickness batch
#> 1 10.411288     A
#> 2  9.830591     A
#> 3 10.108939     A
#> 4 10.189859     A
#> 5 10.121280     A
#> 6  9.968163     A
```

### 2.1 Your first base_plot

`base_plot(data, mapping, theme)` returns a standard `ggplot` object with the
iQualityR theme already attached. Layer geometries on top with `+`, exactly
as you would with `ggplot()`.

The key convention is: **map an aesthetic to a column, and the theme palette
follows automatically.** In the example below `fill = .data$batch` is mapped
to a discrete column, so `base_plot()` auto-injects the theme's discrete fill
scale -- no `scale_fill_*()` call is required. Switch the `theme` argument and
the fill colors change with it.

```{r first-base-plot}
p <- base_plot(steel,
               aes(x = .data$batch, y = .data$thickness, fill = .data$batch)) +
  geom_boxplot(alpha = 0.6) +
  labs(
    title = "Steel Plate Thickness by Batch",
    x = "Batch",
    y = "Thickness (mm)"
  ) +
  theme(legend.position = "none")

print(p)
```

### 2.2 Switching themes with one line

The same data, rendered with a different theme, is a one-token change. Pass
the theme name as a string and `base_plot()` resolves it for you. Because the
fill scale is auto-injected from the resolved theme, the bar colors change
automatically -- note the absence of any `scale_fill_*()` or hardcoded color
in the call.

```{r switch-theme}
base_plot(steel, aes(x = .data$thickness, fill = .data$batch), theme = "prism") +
  geom_histogram(bins = 20, alpha = 0.8) +
  labs(title = "Thickness Distribution (prism theme)", x = "Thickness (mm)")
```

## 3. Theory: The Unified Color Pipeline

### 3.1 The .iqr_plotter singleton

At package load time, `iQualityR.plot` creates a single shared instance of
`IqrPlotterBase`:

```r
.iqr_plotter <- iQualityR.core::IqrPlotterBase$new()
```

This object holds **no state of its own**. It is a stateless toolbox of palette
and scale accessors that always read from the `IqrTheme` object handed to
them. Keeping the toolbox stateless has two consequences:

1. There is never any risk of one plot mutating global color state for
   another.
2. The function-style public API (`plot_qq()`, `plot_acf()`, etc.) is
   preserved -- the toolbox is consulted internally without forcing the user
   to thread a plotter object through every call.

### 3.2 How a plot_* function resolves colors (flowchart as text description)

The resolution chain is the same for every plotting function in the package.
Written as a text flowchart:

```
User calls plot_xxx(data, ..., theme = NULL)
        |
        v
as_iqr_theme_object(theme)
   |-- theme is NULL?  --> read options("iqr.default_theme")  (default "academic")
   |-- theme is string? --> IqrTheme$new(theme_style = theme)
   |-- theme is function? -> IqrTheme$new(theme_style = <fn>)
   |-- theme is IqrTheme? -> returned as-is
   |-- otherwise        --> stop("theme must be NULL, a character string, ...")
        |
        v
   theme_obj  (a concrete IqrTheme object)
        |
        +-- as_iqr_theme(theme)  --> theme_obj$theme_iqr()  [the ggplot2 theme]
        |
        +-- .iqr_plotter$.pal_discrete(theme_obj, n)   [colors]
        +-- .iqr_plotter$.scale_fill_discrete(theme_obj)  [scales]
        +-- .iqr_plotter$.pal_semantic(theme_obj, "fail") [regions]
        |
        v
   ggplot(data, mapping) + theme_iqr + scales + geoms
```

The key insight is that **all color decisions are deferred to the resolved
`IqrTheme` object**, never hard-coded. This is what makes "switch the theme,
recolor everything" possible with a single argument.

### 3.3 The four palette classes matched to visual jobs

The color toolbox exposes four palette classes, each matched to a distinct
visual job:

- **Discrete palette** (`.pal_discrete`) -- for categorical groups such as
  machine ID, operator, or batch. Auto-extended via `colorRampPalette` when
  the number of groups exceeds the base palette length.
- **Sequential palette** (`.pal_sequential`) -- for ordered, single-direction
  data such as counts, durations, or defect severity ranks.
- **Diverging palette** (`.pal_diverging`) -- for deviations from a neutral
  midpoint, for example standardized residuals or correlation matrices
  centred on zero.
- **Semantic palette** (`.pal_semantic`) -- named colors such as `pass`,
  `fail`, and `watch` that encode *meaning* rather than magnitude, used for
  acceptance regions on capability and control charts.

Choosing the correct class is a design decision baked into each `plot_*`
function, so users normally never need to think about it. When composing
custom charts with `base_plot()`, however, the same accessors are available
to keep hand-built layers consistent.

## 4. base_plot() — The Foundation

`base_plot()` is the single entry point every chart starts from. It is a
thin wrapper around `ggplot2::ggplot()` that resolves the theme, attaches it,
and -- crucially -- **auto-injects theme-aware scales for any `fill` /
`color` (or `colour`) aesthetic present in `mapping`**.

### 4.1 Signature and parameters

```{r base-plot-signature, eval = FALSE}
base_plot(data, mapping, theme = NULL)
```

- `data` -- a `data.frame` for plotting. Verified with `stopifnot(is.data.frame(data))`.
- `mapping` -- an aesthetic mapping created by `aes()`. Use the `.data$`
  pronoun to reference columns, e.g. `aes(x = .data$thickness)`.
- `theme` -- the theme specification; see Section 4.3.

The return value is a `ggplot` object, so any standard ggplot2 layer can be
added with `+`. The resolved `IqrTheme` object is stashed as attribute
`"iqr_theme"` on the returned object, so downstream code can retrieve it via
`attr(p, "iqr_theme")` and use the full `theme_obj$plot$*` /
`theme_obj$get_pal(...)` API directly.

### 4.2 Auto-injected scales: map it, and the theme colors it

When `mapping` contains a `fill` or `color` aesthetic that refers to a column
in `data`, `base_plot()` inspects the column type and attaches the matching
theme scale for you:

| Mapped aesthetic | Column type | Auto-injected scale |
|---|---|---|
| `fill = .data$group` | discrete (factor/character) | `theme_obj$plot$scale_fill_iqr(discrete = TRUE)` |
| `fill = .data$score` | continuous (numeric) | `theme_obj$plot$scale_fill_sequential()` |
| `color = .data$group` | discrete | `theme_obj$plot$scale_color_iqr(discrete = TRUE)` |
| `color = .data$score` | continuous | `theme_obj$plot$scale_color_sequential()` |

This means the overwhelming majority of charts need **zero color code**: map
the aesthetic, pick a theme, and the palette follows.

```{r auto-injection-demo}
# Discrete fill: batch is a factor -> discrete palette auto-injected
base_plot(steel, aes(x = .data$batch, y = .data$thickness, fill = .data$batch),
          theme = "economist") +
  geom_boxplot(alpha = 0.6) +
  labs(title = "Discrete fill, economist theme (no scale_* call)") +
  theme(legend.position = "none")

# Continuous fill: thickness is numeric -> sequential gradient auto-injected
base_plot(steel, aes(x = .data$batch, y = .data$thickness, fill = .data$thickness),
          theme = "solarized") +
  geom_col(alpha = 0.8) +
  labs(title = "Continuous fill, solarized theme (no scale_* call)")
```

### 4.3 Advanced scales via attr(p, "iqr_theme")

Three scale families **cannot** be auto-injected because they encode user
intent rather than column type. For these, retrieve the `IqrTheme` object from
the returned plot and attach the scale explicitly using the `.core` API:

- **Paired mode** (fill + darkened outline, for boxplots/bars) --
  `theme_obj$plot$scale_fill_iqr(discrete = TRUE, style = "paired")` plus
  `theme_obj$plot$scale_color_iqr(discrete = TRUE, style = "paired")`.
- **Diverging gradient** (for residuals, correlations centred on zero) --
  `theme_obj$plot$scale_fill_diverging(midpoint = 0)`.
- **Semantic palette** (pass / fail / watch) --
  `theme_obj$plot$scale_fill_semantic(labels = c("pass", "fail", "watch"))`.

```{r paired-mode}
# Paired mode: fill and outline both follow the theme, outline darkened 25%.
# Requires mapping BOTH fill and color to the same discrete column.
p <- base_plot(steel,
               aes(x = .data$batch, y = .data$thickness,
                   fill = .data$batch, color = .data$batch),
               theme = "academic")
theme_obj <- attr(p, "iqr_theme")

p +
  geom_violin(alpha = 0.55) +
  theme_obj$plot$scale_fill_iqr(discrete = TRUE, style = "paired") +
  theme_obj$plot$scale_color_iqr(discrete = TRUE, style = "paired") +
  labs(title = "Paired mode: fill + darkened outline from the theme") +
  theme(legend.position = "none")
```

When a chart needs a **single solid color** that is not mapped to any column
(for example, filling a one-group histogram), retrieve the first discrete
palette color from the theme object so the color still tracks the active
theme:

```{r unmapped-fill}
p <- base_plot(steel, aes(x = .data$thickness), theme = "prism")
theme_obj <- attr(p, "iqr_theme")
fill_color <- theme_obj$get_pal("discrete")[1]

p +
  geom_histogram(bins = 20, fill = fill_color) +
  labs(title = "Unmapped fill resolved from the theme (prism)")
```

### 4.4 Theme input types: NULL / string / function / IqrTheme

The `theme` argument accepts four mutually compatible input types. All four
produce the same kind of object internally:

```{r theme-input-types}
# 1. NULL -- falls back to the global default (academic by default)
p1 <- base_plot(steel, aes(x = .data$thickness), theme = NULL) +
  geom_histogram(bins = 20)

# 2. Character string -- a built-in or external style name
p2 <- base_plot(steel, aes(x = .data$thickness), theme = "economist") +
  geom_histogram(bins = 20)

# 3. Function -- any function returning a ggplot2 theme
p3 <- base_plot(steel, aes(x = .data$thickness), theme = theme_minimal) +
  geom_histogram(bins = 20)

# 4. IqrTheme object -- resolved and returned as-is
theme_obj <- iQualityR.core::IqrTheme$new(theme_style = "solarized")
p4 <- base_plot(steel, aes(x = .data$thickness), theme = theme_obj) +
  geom_histogram(bins = 20)

class(p1)
#> [1] "ggplot2::ggplot" "ggplot"          "ggplot2::gg"     "S7_object"
#> [5] "gg"
```

### 4.5 Theme error behavior (no silent fallback)

When a theme specification cannot be resolved, `as_iqr_theme_object()` raises
an explicit error rather than silently substituting a default. This is a
deliberate design choice: silent fallback hides typos and configuration
mistakes that would otherwise surface as mysteriously unstyled charts.

```{r theme-error, error = TRUE}
base_plot(steel, aes(x = .data$thickness), theme = 123)
#> Error: theme must be NULL, a character string, a function, or an IqrTheme object
```

For cases where you genuinely want a fallback (for example, accepting
user-supplied theme strings in a Shiny app), wrap the call in `tryCatch()` and
re-issue with a known-good theme.

## 5. as_iqr_theme() and as_iqr_theme_object()

Two resolvers sit underneath `base_plot()` and the higher-level `plot_*`
functions:

- `as_iqr_theme_object(theme = NULL)` resolves any theme specification into a
  concrete `IqrTheme` object. This is what you call when you need access to
  the *palette accessors* (e.g. to fetch discrete colors for a custom legend).
- `as_iqr_theme(theme = NULL)` is a thin wrapper that calls
  `as_iqr_theme_object(theme)$theme_iqr()`, returning a ggplot2 theme object
  that can be added directly with `+`. Use this when you only need the
  *non-data ink* (fonts, grid lines) for an existing ggplot.

```{r resolvers}
# Object form -- exposes the palette accessors
obj <- as_iqr_theme_object("prism")
plotter <- iQualityR.core::IqrPlotterBase$new()
discrete_colors <- plotter$.pal_discrete(obj)
length(discrete_colors)
#> [1] 10

# Theme form -- ready to add to a ggplot
ggplot(steel, aes(x = .data$thickness)) +
  geom_histogram(bins = 20, fill = "grey50") +
  as_iqr_theme("economist") +
  labs(title = "Economist theme via as_iqr_theme()")
```

Both resolvers share the same NULL-handling rule: when `theme` is `NULL`, the
value of `options("iqr.default_theme")` is used (default `"academic"`).

## 6. set_default_theme() and the global option

`set_default_theme(theme = "academic")` writes the
`options("iqr.default_theme")` global option. Every subsequent call to
`base_plot()` or any `plot_*` function that receives `theme = NULL` will pick
up the new default.

```{r set-default-theme}
# Inspect the current default
getOption("iqr.default_theme", "academic")
#> [1] "academic"

# Switch the session-wide default to "prism"
set_default_theme("prism")
getOption("iqr.default_theme")
#> [1] "prism"

# A plot created with theme = NULL now inherits "prism"
p_default <- base_plot(steel, aes(x = .data$thickness), theme = NULL)
theme_obj <- attr(p_default, "iqr_theme")
fill_color <- theme_obj$get_pal("discrete")[1]

p_default +
  geom_histogram(bins = 20, fill = fill_color) +
  labs(title = "Inherits the session default (prism)")

# Restore the default for the rest of the vignette
set_default_theme("academic")
```

`set_default_theme()` validates the theme spec by attempting to build it; if
the build fails it emits a warning but still records the option, so a broken
theme never silently breaks downstream plots.

## 7. combine_plots() with patchwork

`combine_plots()` arranges multiple `ggplot` objects into a single dashboard
layout via `patchwork::wrap_plots()`. It accepts plots either as separate
arguments or as a single `plotlist`, and optionally applies a shared theme to
every plot.

```{r combine-plots}
# Each plot resolves its own fill color from the active theme via attr(p, "iqr_theme").
p_hist <- base_plot(steel, aes(x = .data$thickness), theme = "academic")
hist_fill <- attr(p_hist, "iqr_theme")$get_pal("discrete")[1]
p_hist <- p_hist +
  geom_histogram(bins = 20, fill = hist_fill) +
  labs(title = "Distribution")

p_box <- base_plot(steel,
                   aes(x = .data$batch, y = .data$thickness, fill = .data$batch),
                   theme = "academic") +
  geom_boxplot(alpha = 0.7) +
  labs(title = "By Batch") +
  theme(legend.position = "none")

p_qq <- base_plot(steel, aes(sample = .data$thickness), theme = "academic") +
  layers_qq() +
  labs(title = "Normal QQ")

dashboard <- combine_plots(p_hist, p_box, p_qq, ncol = 3)
print(dashboard)
```

The same layout can be produced from a list:

```{r combine-plots-list}
combine_plots(plotlist = list(p_hist, p_box, p_qq), nrow = 1)
```

## 8. save_diagram() — exporting to PNG/SVG/PDF

`save_diagram(obj, filename, type, width, height, dpi, ...)` is a unified
export helper. It inspects the class of `obj` and dispatches to the correct
backend:

- **ggplot2 objects** and **patchwork** objects are saved with `ggplot2::ggsave()`.
- **DiagrammeR** objects (`grViz`, `htmlwidget`, `dgr_graph`) are exported via
  `DiagrammeRsvg` / `rsvg`.

When the file name carries a recognized extension (`.png`, `.svg`, `.pdf`)
the `type` argument is inferred; otherwise the `type` argument is used and
appended to the file name.

```{r save-diagram, eval = FALSE}
# Export the dashboard as a high-resolution PNG
save_diagram(dashboard, "thickness_dashboard.png",
             width = 12, height = 4, dpi = 300)

# Export as vector PDF for a print-quality report
save_diagram(p_box, "thickness_by_batch", type = "pdf",
             width = 6, height = 4)

# Export an SVG for embedding in a web dashboard
save_diagram(p_hist, "thickness_hist.svg")
```

For `ggplot` and `patchwork` objects the `width`/`height` units are inches;
for DiagrammeR `grViz` widgets the units are pixels.

## 9. Theme Switching Demonstration

### 9.1 Same plot, four themes (academic / economist / prism / solarized)

A recurring quality-engineering task is the same chart rendered for
different audiences: a peer-reviewed report (academic), an executive brief
(economist), a clean slide deck (prism), and a high-contrast screen
(solarized). The unified color pipeline makes this a one-token change per
chart.

```{r four-themes}
themes_to_show <- c("academic", "economist", "prism", "solarized")
plots <- lapply(themes_to_show, function(th) {
  # base_plot() auto-injects the discrete fill scale for the mapped `fill`
  # aesthetic, so switching `th` recolors both the theme chrome and the fill
  # palette in one step -- no manual scale_* call needed.
  base_plot(steel,
            aes(x = .data$batch, y = .data$thickness, fill = .data$batch),
            theme = th) +
    geom_violin(alpha = 0.6) +
    labs(title = paste0("Theme: ", th), x = "Batch", y = "Thickness (mm)") +
    theme(legend.position = "none")
})

combine_plots(plotlist = plots, ncol = 2, nrow = 2)
```

Note that switching the theme changed not only the grid lines and fonts but
also the fill palette -- because `base_plot()` resolves the theme into an
`IqrTheme` object and auto-injects the matching discrete fill scale through
the unified color toolbox. Every `plot_*` function in the package does this
automatically.

## 10. Summary

- `base_plot()` is the single foundation: it returns a themed `ggplot`
  object that any standard ggplot2 layer can be added to.
- Theme resolution is centralized in `as_iqr_theme_object()` /
  `as_iqr_theme()`. NULL falls back to `options("iqr.default_theme")`, which
  `set_default_theme()` controls.
- All color decisions are deferred to the resolved `IqrTheme` object through
  the stateless `.iqr_plotter` singleton, so a single theme switch recolors
  every chart consistently.
- `combine_plots()` lays out multiple plots with `patchwork`, and
  `save_diagram()` exports the result to PNG, SVG, or PDF.
- Invalid theme specifications raise explicit errors rather than silently
  falling back -- a deliberate choice that keeps configuration mistakes
  visible.

## 11. Further Reading

- **Layer Builders for Quality Charts** -- the companion vignette
  (`vignette("layer-builders", package = "iQualityR.plot")`) shows how to
  compose reusable layer functions such as `layers_histogram_density()` and
  `layers_control_chart()` on top of `base_plot()`.
- **Basic Plotting and Theme System** -- `vignette("basic_plotting_theme", package = "iQualityR.plot")`.
- **Scatter Plots**, **Pareto Charts**, **Hypothesis Visualization**, and
  **Fishbone Diagrams** vignettes cover the higher-level `plot_*` functions
  that are all built on this same pipeline.
- The `iQualityR.core` package documents the `IqrTheme` and
  `IqrPlotterBase` classes that underpin the color toolbox.

## References

Wickham, H. (2016). *ggplot2: Elegant Graphics for Data Analysis* (2nd ed.).
Springer. <https://ggplot2-book.org>

Wilkinson, L. (2005). *The Grammar of Graphics* (2nd ed.). Springer, New York.

Wickham, H., & Sievert, C. (2016). *ggplot2: Elegant Graphics for Data Analysis*
(online documentation). <https://ggplot2.tidyverse.org>

Pedersen, T. L. (2024). *patchwork: The Composer of Plots*. R package.
<https://patchwork.data-imaginist.com>

Arnold, J. B. (2024). *ggthemes: Extra Themes, Scales and Geoms for ggplot2*.
R package. <https://jrnold.github.io/ggthemes>
