---
title: "Quality Tools Charts"
author: "iQualityR Team"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Quality Tools Charts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

# Quality Tools Charts

## Overview

Quality management has various specialized graphical tools for data analysis and process evaluation. This chapter introduces three core quality tool charts:

- `plot_pp()` - PP Plot (Probability-Probability Plot, for distribution fitting tests)
- `plot_qq()` - QQ Plot (Quantile-Quantile Plot, for normality testing)
- `plot_turtle_diagram()` - Turtle Diagram (process analysis tool, IATF 16949 standard)

## Load Data

```{r setup}
library(ggplot2)
library(datasets)
library(iQualityR.plot)
library(iQualityR.core)

data("mtcars")
data("faithful")

```

## 1. PP Plot: plot_pp()

The PP Plot (Probability-Probability Plot) evaluates whether data follows a specific distribution by comparing the empirical cumulative distribution function with the theoretical cumulative distribution function.

### Normal Distribution PP Plot

```{r pp_norm}
plot_pp(
  data = mtcars,
  sample_col = "mpg",
  dist_family = "norm",
  add_test = TRUE
)
```

### Exponential Distribution PP Plot

```{r pp_exp}
plot_pp(
  data = faithful,
  sample_col = "eruptions",
  dist_family = "exp",
  add_test = TRUE
)
```

### Uniform Distribution PP Plot

```{r pp_unif}
plot_pp(
  data = mtcars,
  sample_col = "wt",
  dist_family = "unif",
  add_test = TRUE
)
```

### Custom Distribution Parameters

```{r pp_custom_params, eval=FALSE}
plot_pp(
  data = mtcars,
  sample_col = "mpg",
  distribution = pnorm,
  dist_params = list(mean = 20, sd = 6),
  add_test = TRUE
)
```

### PP Plot Interpretation

- **Ideal case**: All points fall on the red diagonal line
- **Deviating from the diagonal**: Data does not follow the specified distribution
- **KS test**: p > 0.05 indicates failure to reject the null hypothesis (data follows the distribution)

## 2. QQ Plot: plot_qq()

The QQ Plot (Quantile-Quantile Plot) evaluates data distribution characteristics by comparing sample quantiles with theoretical quantiles. It is more sensitive than the PP Plot, especially for tail analysis.

### Normal QQ Plot

```{r qq_norm}
plot_qq(
  data = mtcars,
  sample_col = "mpg",
  dist_family = "norm",
  qq_line = "quartiles",
  add_test = TRUE
)
```

### QQ Plot with Confidence Band

```{r qq_confidence}
plot_qq(
  data = mtcars,
  sample_col = "hp",
  dist_family = "norm",
  add_confidence = TRUE,
  conf_level = 0.95,
  n_sim = 1000,
  add_test = TRUE
)
```

### Exponential Distribution QQ Plot

```{r qq_exp}
plot_qq(
  data = faithful,
  sample_col = "waiting",
  dist_family = "exp",
  qq_line = "quartiles",
  add_test = TRUE
)
```

### Weibull Distribution QQ Plot

```{r qq_weibull}
plot_qq(
  data = mtcars,
  sample_col = "qsec",
  dist_family = "weibull",
  qq_line = "quartiles",
  add_test = TRUE
)
```

### Using Different Reference Lines

```{r qq_lines, eval=FALSE}
plot_qq(data, "mpg", dist_family = "norm", qq_line = "quartiles")

plot_qq(data, "mpg", dist_family = "norm", 
        dist_params = list(mean = 20, sd = 6),
        qq_line = "identity")
```

### QQ Plot Interpretation

- **Points near the reference line**: Data approximately follows the specified distribution
- **Tail separation**: Data tails are heavier or lighter than the theoretical distribution
- **S-shaped curve**: Data distribution is skewed
- **Points outside confidence band**: May be outliers

## 3. Turtle Diagram: plot_turtle_diagram()

The Turtle Diagram is a process analysis tool required by the IATF 16949 standard, used to identify the key elements of a process.

### Basic Usage

```{r turtle_basic }
plot_turtle_diagram(
  process_name = "Parts Machining Process",
  process_owner = "Production Manager",
  activities = c("Receive order", "Prepare materials", "Machining", "Inspection", "Packaging"),
  inputs = c("Raw materials", "Process drawings", "Semi-finished products from previous process"),
  outputs = c("Finished parts", "Inspection records", "Non-conforming product records"),
  resources = c("CNC machine", "Gauges", "Cutting tools"),
  procedures = c("Work instruction", "Inspection specification", "Equipment operation procedure"),
  metrics = c("Pass rate >= 98%", "On-time delivery rate >= 95%", "Equipment OEE >= 85%"),
  responsibilities = c("Operator", "Inspector", "Team leader"),
  direction = "T",
  theme_style = "tech",
  title = "Parts Machining Process Turtle Diagram"
)
```

### Service Industry Turtle Diagram

```{r turtle_service }
plot_turtle_diagram(
  process_name = "Customer Complaint Handling",
  process_owner = "Customer Service Supervisor",
  activities = c("Receive complaint", "Record and classify", "Investigate cause", "Develop solution", "Feedback to customer", "Follow-up verification"),
  inputs = c("Customer complaint information", "Product records", "Historical complaint data"),
  outputs = c("Handling report", "Improvement measures", "Customer satisfaction feedback"),
  resources = c("CRM system", "Customer service personnel", "Communication channels"),
  procedures = c("Complaint handling process", "Escalation mechanism", "Follow-up system"),
  metrics = c("Response time < 2 hours", "Resolution rate >= 90%", "Customer satisfaction >= 85%"),
  responsibilities = c("Customer service representative", "Technical support", "Quality engineer"),
  direction = "R",
  theme_style = "academic",
  title = "Customer Complaint Handling Process Turtle Diagram"
)
```

### Different Directions

```{r turtle_directions }
plot_turtle_diagram(process_name = "Process A", direction = "T")

plot_turtle_diagram(process_name = "Process B", direction = "B")

plot_turtle_diagram(process_name = "Process C", direction = "L")

plot_turtle_diagram(process_name = "Process D", direction = "R")
```

### Custom Colors

```{r turtle_colors }
plot_turtle_diagram(
  process_name = "Custom Process",
  user_colors = list(
    process = "#3498db",
    inputs = "#2ecc71",
    outputs = "#e74c3c",
    resources = "#f39c12",
    procedures = "#9b59b6",
    metrics = "#1abc9c",
    responsibilities = "#e67e22"
  )
)
```

### Export Images

```{r turtle_export }
plot_turtle_diagram(
  process_name = "Export Example",
  output_file = "turtle_diagram.png",
  output_type = "png"
)

plot_turtle_diagram(
  process_name = "Export Example",
  output_file = "turtle_diagram.svg",
  output_type = "svg"
)

plot_turtle_diagram(
  process_name = "Export Example",
  output_file = "turtle_diagram.pdf",
  output_type = "pdf"
)
```

### Turtle Diagram Element Description

| Element | Position | Description |
|------|------|------|
| Process | Center circle | Process name, owner, activity steps |
| Inputs | Tail | Materials, information, upstream process |
| Outputs | Head | Products, records, downstream process |
| Resources | Left leg | Equipment, personnel, tools |
| Procedures | Right leg | Procedures, standards, specifications |
| Metrics | Left front leg | KPIs, target values |
| Responsibilities | Right front leg | Roles, authorities, competency requirements |

## Practical Case Studies

### Case 1: Normality Testing Complete Workflow

```{r case_normality, warning=FALSE}


p_pp <- plot_pp(
  data = mtcars,
  sample_col = "hp",
  dist_family = "norm",
  add_test = TRUE
)

p_qq <- plot_qq(
  data = mtcars,
  sample_col = "hp",
  dist_family = "norm",
  add_confidence = TRUE,
  add_test = TRUE,title = "qq plot",theme = "tech"
)

print(p_qq)
```

### Case 2: Process Analysis Turtle Diagram

```{r case_process_turtle }
plot_turtle_diagram(
  process_name = "SMT Placement Process",
  process_owner = "SMT Supervisor",
  activities = c("Solder paste printing", "Component placement", "Reflow soldering", "AOI inspection", "Rework"),
  inputs = c("PCB board", "Electronic components", "Solder paste", "Placement program"),
  outputs = c("Completed PCBA", "Inspection records", "Defective products"),
  resources = c("Placement machine", "Reflow oven", "AOI equipment", "Stencil"),
  procedures = c("SOP", "Solder paste control specification", "Equipment checklist", "First article inspection specification"),
  metrics = c("First-pass yield >= 99.5%", "Throw rate < 0.1%", "Equipment utilization >= 90%"),
  responsibilities = c("Operator", "Technician", "Quality inspector"),
  direction = "T",
  theme_style = "tech",
  title = "SMT Placement Process Turtle Diagram",
  subtitle = "IATF 16949 Process Analysis",
  annotation = "Note: This diagram is based on IATF 16949:2016 standard"
)
```

### Case 3: Distribution Fitting Comparison

```{r case_distribution_fit}
data("faithful")

p_norm <- plot_qq(
  data = faithful,
  sample_col = "eruptions",
  dist_family = "norm",
  add_test = TRUE
)

p_exp <- plot_qq(
  data = faithful,
  sample_col = "eruptions",
  dist_family = "exp",
  add_test = TRUE
)

print(p_norm)
```

## Parameter Details

### plot_pp() Parameters

| Parameter | Type | Description | Default |
|------|------|------|--------|
| `data` | data.frame | Data frame | Required |
| `sample_col` | character | Sample column name | Required |
| `theme` | theme | Theme specification | NULL |
| `distribution` | function | Custom CDF function | pnorm |
| `dist_params` | list | Distribution parameters | list() |
| `dist_family` | character | Distribution type | "none" |
| `add_test` | logical | Whether to add KS test | TRUE |
| `...` |  | Extra arguments passed to ggplot2::labs() |  |

### plot_qq() Parameters

| Parameter | Type | Description | Default |
|------|------|------|--------|
| `data` | data.frame | Data frame | Required |
| `sample_col` | character | Sample column name | Required |
| `theme` | theme | Theme specification | NULL |
| `distribution` | function | Custom quantile function | qnorm |
| `dist_params` | list | Distribution parameters | list() |
| `dist_family` | character | Distribution type | "none" |
| `qq_line` | character | Reference line type | "quartiles" |
| `add_confidence` | logical | Whether to add confidence band | FALSE |
| `conf_level` | numeric | Confidence level | 0.95 |
| `n_sim` | numeric | Number of simulations | 1000 |
| `add_test` | logical | Whether to add test | TRUE |
| `test_method` | character | Test method | "auto" |
| `...` |  | Extra arguments passed to ggplot2::labs() |  |

### plot_turtle_diagram() Parameters

| Parameter | Type | Description | Default |
|------|------|------|--------|
| `process_name` | character | Process name | Required |
| `process_owner` | character | Process owner | NULL |
| `activities` | character | Activity steps | NULL |
| `inputs` | character/list | Inputs | NULL |
| `outputs` | character/list | Outputs | NULL |
| `resources` | character | Resources | NULL |
| `procedures` | character | Procedures | NULL |
| `metrics` | character | Metrics | NULL |
| `responsibilities` | character | Responsibilities | NULL |
| `direction` | character | Direction | "T" |
| `theme_style` | character | Theme style | "tech" |
| `user_colors` | list | Custom colors | NULL |
| `title` | character | Main title | NULL |
| `subtitle` | character | Subtitle | NULL |
| `subtitle_fontsize` | numeric | Subtitle font size | 14 |
| `annotation` | character | Footnote/annotation | NULL |
| `annotation_fontsize` | numeric | Footnote font size | 10 |
| `output_file` | character | Output file | NULL |
| `output_type` | character | Output type | "png" |

## Best Practices

### 1. Normality Testing Workflow

```{r best_practice_normality, eval=FALSE}
hist(data$x)

plot_qq(data, "x", dist_family = "norm", add_confidence = TRUE)

shapiro.test(data$x)
```

### 2. Turtle Diagram Use Cases

```{r best_practice_turtle, eval=FALSE}
# Applicable scenarios:
# - IATF 16949 process analysis
# - ISO 9001 process approach implementation
# - Process risk assessment
# - Process audit preparation
# - Process documentation development
```

### 3. Distribution Selection Guide

| Data Type | Recommended Distribution | Application Scenario |
|---------|---------|---------|
| Measurement data | Normal distribution | Process capability, SPC |
| Life data | Weibull distribution | Reliability analysis |
| Waiting time | Exponential distribution | Queueing theory |
| Count data | Poisson distribution | Defect analysis |
| Proportion data | Binomial distribution | Pass rate analysis |

## FAQ

### Q: What is the difference between PP Plot and QQ Plot?

- **PP Plot**: Compares cumulative probabilities, sensitive to the middle portion
- **QQ Plot**: Compares quantiles, sensitive to tails, more commonly used

### Q: How to interpret the confidence band in QQ Plot?

Points within the confidence band indicate no significant difference between the data and the theoretical distribution at the specified confidence level. Points outside the band may be outliers.

### Q: Can Turtle Diagrams be exported to editable formats?

Currently supports PNG, SVG, and PDF formats. SVG and PDF are vector formats that can be edited in software such as Illustrator.

## Summary

- `plot_pp()` is used to evaluate whether data follows a specific distribution
- `plot_qq()` is the standard tool for normality testing, supporting multiple distributions and confidence bands
- `plot_turtle_diagram()` is a process analysis tool required by IATF 16949
- The three tools are applicable to statistical analysis, quality assessment, and process management scenarios respectively

## Further Reading

- [Hypothesis Testing Visualization](hypothesis_visualization.html) - Learn about statistical test visualization
- [Advanced Analysis Plots](advanced_plots.html) - Learn about correlation heatmaps and interaction plots
- [Basic Plotting and Theme System](basic_plotting_theme.html) - Learn about the theme system
