---
title: "Scatter Plots"
author: "iQualityR Team"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Scatter Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

# Scatter Plots

## Overview

Scatter plots are one of the most fundamental and important analytical tools in quality management, used to study the relationship between two variables. `iQualityR.plot` provides four scatter plot functions:

- `plot_scatter_basic()` - Basic scatter plot (with regression line)
- `plot_scatter_grouped()` - Grouped scatter plot
- `plot_scatter_bubble()` - Bubble scatter plot
- `plot_scatter_density()` - Density scatter plot (handling overlapping points)

## Load Data

```{r setup}
library(ggplot2)
library(datasets)
library(iQualityR.core)
library(iQualityR.plot)

data("mtcars")
data("iris")
```

## 1. Basic Scatter Plot: plot_scatter_basic()

The basic scatter plot is used to explore the linear relationship between two continuous variables, automatically adding a regression line and correlation coefficient.

### Basic Usage

```{r basic_scatter}
# Analyze the relationship between horsepower and fuel efficiency
plot_scatter_basic(
  data = mtcars,
  x_var = "hp",
  y_var = "mpg",
  title = "Horsepower vs Fuel Efficiency"
)
```

### Custom Options

```{r basic_custom}
# Custom scatter plot style
plot_scatter_basic(
  data = mtcars,
  x_var = "wt",
  y_var = "mpg",
  add_regression = TRUE,
  add_ci = TRUE,
  add_correlation = TRUE,
  correlation_method = "pearson",
  point_color = "#e74c3c",
  point_size = 2.5,
  point_alpha = 0.7,
  line_color = "#2c3e50",
  line_size = 1.5,
  ci_fill = "#3498db",
  ci_alpha = 0.15,
  title = "Weight vs Fuel Efficiency"
)
```

### Without Regression Line

```{r basic_no_regression}
# Show scatter points only, without regression line
plot_scatter_basic(
  data = mtcars,
  x_var = "disp",
  y_var = "qsec",
  add_regression = FALSE,
  add_correlation = TRUE,
  title = "Engine Displacement vs Acceleration Time"
)
```

### Using Different Correlation Methods

```{r correlation_methods, eval=FALSE}
# Pearson correlation coefficient (default, suitable for linear relationships)
plot_scatter_basic(mtcars, "hp", "mpg", correlation_method = "pearson")

# Spearman rank correlation (suitable for monotonic relationships, robust to outliers)
plot_scatter_basic(mtcars, "hp", "mpg", correlation_method = "spearman")

# Kendall rank correlation (suitable for small samples)
plot_scatter_basic(mtcars, "hp", "mpg", correlation_method = "kendall")
```

## 2. Grouped Scatter Plot: plot_scatter_grouped()

When data contains categorical variables, the grouped scatter plot can use different colors for different groups and fit regression lines separately.

### Basic Usage

```{r grouped_basic}
# Analyze by number of cylinders
plot_scatter_grouped(
  data = mtcars,
  x_var = "hp",
  y_var = "mpg",
  group_var = "cyl",
  title = "Horsepower vs Fuel Efficiency by Cylinder Count"
)
```

### Custom Grouped Style

```{r grouped_custom}
# Custom grouped scatter plot
plot_scatter_grouped(
  data = mtcars,
  x_var = "wt",
  y_var = "mpg",
  group_var = "am",
  add_regression = TRUE,
  add_ci = TRUE,
  add_correlation = TRUE,
  point_size = 2,
  point_alpha = 0.7,
  line_size = 1.2,
  palette = c("0" = "#e74c3c", "1" = "#3498db"),
  title = "Weight vs Fuel Efficiency by Transmission Type"
)
```

### Without Regression Line

```{r grouped_no_regression}
# Show grouped scatter points only
plot_scatter_grouped(
  data = iris,
  x_var = "Sepal.Length",
  y_var = "Petal.Length",
  group_var = "Species",
  add_regression = FALSE,
  title = "Iris Sepal Length vs Petal Length"
)
```

## 3. Bubble Scatter Plot: plot_scatter_bubble()

The bubble scatter plot uses point size to display a third variable, enabling three-dimensional data visualization.

### Basic Usage

```{r bubble_basic}
# Use bubble size to represent car weight
plot_scatter_bubble(
  data = mtcars,
  x_var = "hp",
  y_var = "mpg",
  size_var = "wt",
  title = "Horsepower, Fuel Efficiency, and Weight"
)
```

### Bubble Plot with Grouping

```{r bubble_grouped}
# Group by transmission type, bubble size represents weight
plot_scatter_bubble(
  data = mtcars,
  x_var = "hp",
  y_var = "mpg",
  size_var = "disp",
  group_var = "am",
  size_range = c(2, 12),
  point_alpha = 0.7,
  title = "Three-Dimensional Car Performance Comparison"
)
```

### Adding Labels

```{r bubble_labels, eval=FALSE}
# Add labels to each bubble
plot_scatter_bubble(
  data = mtcars,
  x_var = "wt",
  y_var = "mpg",
  size_var = "hp",
  add_labels = TRUE,
  label_var = "row_names",  # Use row names as labels
  label_size = 3,
  title = "Weight, Fuel Efficiency, and Horsepower"
)
```

## 4. Density Scatter Plot: plot_scatter_density()

When data points are densely overlapping, the density scatter plot provides multiple methods to address overplotting.

### Alpha Transparency Method

```{r density_alpha}
# Use transparency to distinguish density
set.seed(123)
large_data <- data.frame(
  x = rnorm(1000),
  y = rnorm(1000) + 0.5 * rnorm(1000)
)

plot_scatter_density(
  data = large_data,
  x_var = "x",
  y_var = "y",
  method = "alpha",
  point_alpha = 0.4,
  point_size = 2,
  title = "High-Density Scatter Plot (Alpha Transparency)"
)
```

### Jitter Method

```{r density_jitter}
# Use jitter to avoid overlap
plot_scatter_density(
  data = large_data,
  x_var = "x",
  y_var = "y",
  method = "jitter",
  jitter_width = 0.1,
  jitter_height = 0.1,
  point_alpha = 0.6,
  title = "High-Density Scatter Plot (Jitter Method)"
)
```

### Two-Dimensional Binning Method

```{r density_bins}
# Use two-dimensional binning
plot_scatter_density(
  data = large_data,
  x_var = "x",
  y_var = "y",
  method = "bins",
  bins = 30,
  color_low = "yellow",
  color_high = "red",
  title = "High-Density Scatter Plot (2D Binning)"
)
```

### Density Contour Method

```{r density_contour}
# Use density contours
plot_scatter_density(
  data = large_data,
  x_var = "x",
  y_var = "y",
  method = "density",
  bins = 20,
  color_low = "#3498db",
  color_high = "#e74c3c",
  title = "High-Density Scatter Plot (Density Contour)"
)
```

### Adding Smooth Line

```{r density_smooth, eval=FALSE}
# Add a smooth line to the density plot
plot_scatter_density(
  data = mtcars,
  x_var = "hp",
  y_var = "mpg",
  method = "alpha",
  add_smooth = TRUE,
  smooth_method = "loess",
  title = "Density Scatter Plot with Smooth Line"
)
```

## Practical Case Studies

### Case 1: Automobile Performance Analysis

```{r case_study_auto}
# Comprehensive analysis of the mtcars dataset using multiple scatter plots

# 1. Basic scatter plot: horsepower vs fuel efficiency
p1 <- plot_scatter_basic(
  mtcars, "hp", "mpg",
  title = "Horsepower vs Fuel Efficiency"
)

# 2. Grouped scatter plot: grouped by cylinder count
p2 <- plot_scatter_grouped(
  mtcars, "hp", "mpg", "cyl",
  title = "Grouped by Cylinder Count"
)

# 3. Bubble plot: adding weight dimension
p3 <- plot_scatter_bubble(
  mtcars, "hp", "mpg", "wt",
  title = "Adding Weight Dimension"
)

print(p1)
```

### Case 2: Iris Data Analysis

```{r case_study_iris}
# Use the iris dataset to demonstrate grouped and bubble plots

# Grouped scatter plot
p_iris <- plot_scatter_grouped(
  data = iris,
  x_var = "Sepal.Length",
  y_var = "Petal.Width",
  group_var = "Species",
  add_regression = TRUE,
  add_ci = FALSE,
  palette = c("setosa" = "#e74c3c", 
              "versicolor" = "#3498db", 
              "virginica" = "#2ecc71"),
  title = "Iris Sepal Length vs Petal Width",
  subtitle = "Grouped by Species"
)

print(p_iris)
```

## Method Selection Guide

| Scenario | Recommended Function | Key Parameters |
|------|---------|---------|
| Explore relationship between two variables | `plot_scatter_basic()` | `add_regression=TRUE` |
| Compare different groups | `plot_scatter_grouped()` | `group_var` |
| Display three variables | `plot_scatter_bubble()` | `size_var` |
| Severe data point overlap | `plot_scatter_density()` | `method="alpha/bins/density"` |
| Need correlation coefficient | `plot_scatter_basic()` | `add_correlation=TRUE` |
| Nonlinear relationship | `plot_scatter_density()` | `add_smooth=TRUE, smooth_method="loess"` |

## Best Practices

### 1. Choose the Appropriate Sample Size

```{r best_practice_sample, eval=FALSE}
# Small sample (< 100): Use basic scatter plot
plot_scatter_basic(data, "x", "y")

# Medium sample (100-500): Can use grouped or bubble plot
plot_scatter_grouped(data, "x", "y", "group")

# Large sample (> 500): Use density scatter plot
plot_scatter_density(data, "x", "y", method = "alpha")
```

### 2. Handling Outliers

```{r best_practice_outliers, eval=FALSE}
# Detect outliers
outliers <- which(abs(scale(data$x)) > 3 | abs(scale(data$y)) > 3)

# Annotate outliers
plot_scatter_basic(data, "x", "y") +
  geom_point(data = data[outliers, ], color = "red", size = 3)
```

### 3. Multi-Plot Comparison

```{r best_practice_comparison, eval=FALSE}
# Use patchwork to combine multiple scatter plots
library(patchwork)

p1 <- plot_scatter_basic(data, "x1", "y")
p2 <- plot_scatter_basic(data, "x2", "y")
p3 <- plot_scatter_basic(data, "x3", "y")

(p1 + p2) / p3
```

## FAQ

### Q: Regression line not showing?

Make sure `add_regression = TRUE` and the data has a linear relationship:

```{r faq_regression, eval=FALSE}
plot_scatter_basic(data, "x", "y", add_regression = TRUE)
```

### Q: How to change the regression line color?

Use the `line_color` parameter:

```{r faq_line_color, eval=FALSE}
plot_scatter_basic(data, "x", "y", line_color = "red")
```

### Q: What to do when data points overlap severely?

Use a density scatter plot:

```{r faq_overlap, eval=FALSE}
plot_scatter_density(data, "x", "y", method = "alpha", point_alpha = 0.3)
```

## Summary

- `plot_scatter_basic()` is suitable for basic two-dimensional relationship exploration
- `plot_scatter_grouped()` is suitable for grouped comparison analysis
- `plot_scatter_bubble()` is suitable for three-dimensional data visualization
- `plot_scatter_density()` is suitable for addressing overplotting in high-density data
- All functions support the iQualityR theme system, ensuring consistent plot style

## Further Reading

- [Basic Plotting and Theme System](basic_plotting_theme.html) - Learn about the theme system
- [Pareto Charts](pareto_charts.html) - Learn about Pareto chart plotting
- [Advanced Analysis Plots](advanced_plots.html) - Learn about correlation heatmaps and interaction plots
