---
title: "Scatter Plots for Process Correlation Analysis"
author: "iQualityR.plot Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Scatter Plots for Process Correlation Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

library(iQualityR.plot)
library(ggplot2)

fmt_num <- function(x, digits = 4) formatC(x, format = "f", digits = digits)
fmt_pct <- function(x, digits = 2) paste0(formatC(x, format = "f", digits = digits), "%")
```

## 1. Introduction

Scatter plots are the simplest and most powerful tool for exploring the
relationship between two continuous process variables. In quality
engineering they are used to:

* screen potential inputs before designing an experiment,
* verify that a controllable factor tracks an output response,
* detect non-linear behaviour, outliers, and subgroup structure,
* communicate correlation findings to stakeholders.

The `iQualityR.plot` package provides four scatter-plot functions tailored
to process analysis: a basic regression plot, a grouped plot for
categorical factors, a bubble plot for three variables, and a density-aware
plot for large data sets. This vignette demonstrates each function on
realistic quality-engineering data and explains the underlying theory.

## 2. Theory: Correlation vs Causation in Process Analysis

### Pearson product–moment correlation

For paired observations $(x_i, y_i)$, the Pearson correlation coefficient
is

$$
r = \frac{\sum_{i=1}^{n}(x_i - \bar{x})(y_i - \bar{y})}
         {\sqrt{\sum_{i=1}^{n}(x_i - \bar{x})^2 \;\cdot\;
                \sum_{i=1}^{n}(y_i - \bar{y})^2}}.
$$

It ranges from $-1$ (perfect negative linear association) to $+1$ (perfect
positive linear association), with $r = 0$ indicating the *absence of
linear* association. Pearson's $r$ is sensitive to outliers and assumes
that both variables are approximately normally distributed and linearly
related.

### Spearman rank correlation

Spearman's $\rho$ replaces the raw values with their ranks before computing
the Pearson formula:

$$
\rho = 1 - \frac{6 \sum d_i^2}{n(n^2 - 1)},
$$

where $d_i$ is the difference between the ranks of $x_i$ and $y_i$.
Because it operates on ranks, $\rho$ is **robust to outliers** and detects
any monotonic (not just linear) relationship. It does not require
normality.

### Correlation does not imply causation

A strong correlation between two process variables does not establish that
one causes the other. Three rival explanations must always be considered:

1. **Direct causation** — $X$ truly drives $Y$ (e.g. feed rate drives
   surface roughness).
2. **Confounding** — a third variable $Z$ drives both $X$ and $Y$
   (e.g. ambient temperature raises both cycle time and defect rate).
3. **Coincidence** — spurious correlation from a small sample or from
   two unrelated trends over time.

A scatter plot cannot, by itself, distinguish these explanations. It can,
however, reveal non-linearity, heteroscedasticity, and subgroup structure
that a correlation coefficient would hide.

### Confidence band for the regression line

The $1 - \alpha$ confidence band around the fitted regression line
$\hat{y} = \hat{\beta}_0 + \hat{\beta}_1 x$ at a predictor value $x_0$ is

$$
\hat{y}_0 \;\pm\; t_{\alpha/2,\,n-2} \cdot s_{y|x} \cdot
\sqrt{\frac{1}{n} + \frac{(x_0 - \bar{x})^2}{S_{xx}}},
$$

where $s_{y|x}$ is the residual standard error and
$S_{xx} = \sum (x_i - \bar{x})^2$. The band is narrowest at $\bar{x}$ and
widens toward the extremes of the data, reflecting greater uncertainty in
the slope at the boundaries.

### Overplotting and its remedies

When the sample size is large, points overlap and obscure structure. The
five remedies offered by `plot_scatter_density()` are:

| Method | Mechanism | Best for |
|---|---|---|
| `alpha` | Semi-transparent points | Moderate overlap |
| `jitter` | Add small random noise | Discrete / rounded data |
| `bins` | Rectangular 2-D bin counts | Very large $n$ |
| `hex` | Hexagonal bin counts | Very large $n$, smoother than bins |
| `density` | Color-coded 2-D kernel density | Continuous data, large $n$ |

## 3. plot_scatter_basic() — with regression and confidence band

`plot_scatter_basic()` produces a single-group scatter plot with optional
least-squares regression line, $1 - \alpha$ confidence band, and printed
correlation coefficient.

```{r}
# Simulated relationship between feed rate and surface roughness
set.seed(11)
feed <- data.frame(
  feed_rate = runif(80, 100, 300),
  roughness = NA
)
feed$roughness <- with(feed, 0.4 + 0.012 * feed_rate + rnorm(80, 0, 0.6))

plot_scatter_basic(
  data              = feed,
  x_var             = "feed_rate",
  y_var             = "roughness",
  add_regression    = TRUE,
  add_ci            = TRUE,
  add_correlation   = TRUE,
  correlation_method = "pearson",
  title             = "Surface roughness vs feed rate",
  subtitle          = "CNC finishing pass, n = 80"
)
```

**Spearman correlation** is preferred when outliers are suspected or the
relationship is monotonic but non-linear:

```{r}
plot_scatter_basic(
  data               = feed,
  x_var              = "feed_rate",
  y_var              = "roughness",
  correlation_method = "spearman",
  add_ci             = FALSE,
  title              = "Same data, Spearman rank correlation",
  subtitle           = "Robust to outliers and non-linearity"
)
```

**Theme-driven aesthetics.** When `point_color`, `line_color`, and
`ci_fill` are left `NULL`, the colors are resolved from the active theme
through `.iqr_aes()`: points use the theme's `data` role, the regression
line uses the `danger` semantic, and the confidence band uses the
`primary` semantic. Switching `theme` therefore repaints the entire plot
without any per-call color arguments:

```{r}
plot_scatter_basic(
  data            = feed,
  x_var           = "feed_rate",
  y_var           = "roughness",
  theme           = "economist",
  add_correlation = FALSE,
  title           = "Scatter plot themed by 'economist'",
  subtitle        = "All colors resolved from the IqrTheme palette"
)
```

To override a single role with another theme semantic, retrieve the
`IqrTheme` object via `attr(p, "iqr_theme")` and pass the desired role
back into the function:

```{r}
p_seed <- base_plot(feed, ggplot2::aes(x = .data$feed_rate,
                                       y = .data$roughness),
                    theme = "academic")
c_acad <- attr(p_seed, "iqr_theme")

plot_scatter_basic(
  data        = feed,
  x_var       = "feed_rate",
  y_var       = "roughness",
  theme       = "academic",
  point_color = c_acad$get_pal("discrete")[1],
  point_size  = 2.5,
  point_alpha = 0.85,
  line_color  = c_acad$get_pal("semantic", name = "fail"),
  line_size   = 1.3,
  ci_fill     = c_acad$get_ui_colors()$primary,
  ci_alpha    = 0.15,
  add_correlation = FALSE,
  title       = "Override individual roles with theme palette values"
)
```

## 4. plot_scatter_grouped() — by categorical factor

`plot_scatter_grouped()` colors points by a categorical factor and draws a
separate regression line for each level. This is the standard tool for
checking whether a single regression line adequately describes all
subgroups, or whether an interaction with the grouping factor is present.

```{r}
# Coolant type as a categorical factor
feed$coolant <- sample(c("Emulsion", "Micro-lube"), 80, replace = TRUE)

plot_scatter_grouped(
  data              = feed,
  x_var             = "feed_rate",
  y_var             = "roughness",
  group_var         = "coolant",
  add_regression    = TRUE,
  add_ci            = FALSE,
  add_correlation   = TRUE,
  correlation_method = "pearson",
  title             = "Roughness vs feed rate by coolant type",
  subtitle          = "Separate regression lines reveal potential interaction"
)
```

When the two regression lines have clearly different slopes, the grouping
factor interacts with the continuous predictor — a key finding that a
pooled scatter plot would hide.

**Theme-driven grouping colors.** When `palette = NULL`, the grouped
plot falls back to `as_iqr_theme_object(theme)$scale_color_iqr()`, so
the per-group colors stay consistent with the rest of the report. Change
the theme and every grouped plot follows:

```{r}
plot_scatter_grouped(
  data      = feed,
  x_var     = "feed_rate",
  y_var     = "roughness",
  group_var = "coolant",
  theme     = "solarized",
  title     = "Grouped scatter themed by 'solarized'",
  subtitle  = "Discrete palette resolved from IqrTheme"
)
```

For brand or accessibility overrides, build the palette from the theme
semantics (e.g., `pass`/`fail`/`watch`) so the override still tracks the
active theme:

```{r}
p_seed <- base_plot(feed, ggplot2::aes(x = .data$feed_rate,
                                       y = .data$roughness),
                    theme = "academic")
c_acad <- attr(p_seed, "iqr_theme")
sem <- c_acad$get_pal("semantic")

plot_scatter_grouped(
  data      = feed,
  x_var     = "feed_rate",
  y_var     = "roughness",
  group_var = "coolant",
  theme     = "academic",
  palette   = c("Emulsion" = sem[["pass"]], "Micro-lube" = sem[["watch"]]),
  title     = "Palette built from theme semantics"
)
```

## 5. plot_scatter_bubble() — three-variable visualization

`plot_scatter_bubble()` encodes a third continuous variable as bubble
size, enabling a compact three-variable view without faceting. This is
useful when the third variable is itself a process output (e.g. cycle
time, cost, or scrap rate).

```{r}
# Add a third variable: tool wear (bubble size)
feed$tool_wear <- with(feed, 30 + 0.05 * feed_rate + rnorm(80, 0, 8))

plot_scatter_bubble(
  data        = feed,
  x_var       = "feed_rate",
  y_var       = "roughness",
  size_var    = "tool_wear",
  group_var   = "coolant",
  size_range  = c(2, 12),
  add_labels  = FALSE,
  point_alpha = 0.7,
  title       = "Bubble plot: feed rate, roughness, tool wear",
  subtitle    = "Bubble size encodes cumulative tool wear (min)"
)
```

**Optional point labels** can identify individual runs or outlying parts:

```{r}
feed$run_id <- paste0("R", sprintf("%02d", seq_len(nrow(feed))))

plot_scatter_bubble(
  data       = feed,
  x_var      = "feed_rate",
  y_var      = "roughness",
  size_var   = "tool_wear",
  add_labels = TRUE,
  label_var  = "run_id",
  label_size = 2.5,
  point_alpha = 0.6,
  title      = "Labeled bubble plot with run identifiers"
)
```

## 6. plot_scatter_density() — handling overplotting

When $n$ is large, ordinary scatter plots saturate into a solid blob and
hide the density of the underlying point cloud. `plot_scatter_density()`
offers five methods for revealing structure that would otherwise be lost.

### 6.1 alpha, jitter, bins, hex methods compared

We simulate a large data set with structured overlap so the differences
between methods are visible:

```{r}
set.seed(99)
big <- data.frame(
  x = rnorm(4000, 50, 8),
  y = NA
)
big$y <- with(big, 0.6 * x + rnorm(4000, 0, 3))
```

**Method 1 — alpha (transparency).** Reduces the visual weight of each
point so dense regions appear darker:

```{r}
plot_scatter_density(
  data        = big,
  x_var       = "x",
  y_var       = "y",
  method      = "alpha",
  point_alpha = 0.15,
  point_size  = 1.6,
  title       = "Overplotting remedy: alpha blending",
  subtitle    = "Dense regions appear naturally darker"
)
```

**Method 2 — jitter.** Useful when one variable is discrete or rounded:

```{r}
discrete_y <- big
discrete_y$y <- round(discrete_y$y)

plot_scatter_density(
  data         = discrete_y,
  x_var        = "x",
  y_var        = "y",
  method       = "jitter",
  jitter_width = 0.2,
  jitter_height = 0.4,
  point_alpha  = 0.4,
  title        = "Overplotting remedy: jitter",
  subtitle     = "Reveals points hidden by rounding"
)
```

**Method 3 — bins (rectangular 2-D binning).** Counts points per
rectangular cell and colors by count:

```{r}
plot_scatter_density(
  data   = big,
  x_var  = "x",
  y_var  = "y",
  method = "bins",
  bins   = 40,
  title  = "Overplotting remedy: rectangular bins",
  subtitle = "Cell color encodes point count"
)
```

**Method 4 — hex (hexagonal binning).** Hexagons tile more evenly than
rectangles and avoid axis-aligned artifacts:

```{r}
plot_scatter_density(
  data   = big,
  x_var  = "x",
  y_var  = "y",
  method = "hex",
  bins   = 40,
  title  = "Overplotting remedy: hexagonal bins",
  subtitle = "Hexagons avoid axis-alignment artifacts"
)
```

**Method 5 — density (2-D kernel density).** Produces a smooth colored
surface of point density:

```{r}
plot_scatter_density(
  data       = big,
  x_var      = "x",
  y_var      = "y",
  method     = "density",
  add_smooth = FALSE,
  title      = "Overplotting remedy: 2-D kernel density",
  subtitle   = "Smooth surface of point concentration"
)
```

The `add_smooth` argument can overlay a loess (or other) smoother when the
relationship shape is itself of interest:

```{r}
plot_scatter_density(
  data         = big,
  x_var        = "x",
  y_var        = "y",
  method       = "alpha",
  point_alpha  = 0.1,
  add_smooth   = TRUE,
  smooth_method = "loess",
  title        = "Alpha scatter with loess smoother"
)
```

**Theme-driven density colors.** For `bins`, `hex`, and `density`
methods the high-density color is taken from the theme sequential scale,
so a single `theme =` argument repaints the surface. Compare the same
hex binning under two themes:

```{r, fig.show="hold", out.width="50%"}
plot_scatter_density(
  data   = big,
  x_var  = "x",
  y_var  = "y",
  method = "hex",
  bins   = 40,
  theme  = "academic",
  title  = "hex — academic"
)

plot_scatter_density(
  data   = big,
  x_var  = "x",
  y_var  = "y",
  method = "hex",
  bins   = 40,
  theme  = "prism",
  title  = "hex — prism"
)
```

## 7. Quality Scenario: Cycle Time vs Defect Rate

A process engineer suspects that the defect rate on a machining line
rises when cycle time drifts away from its target. Data are collected from
200 consecutive parts. The true relationship is non-linear (quadratic),
which makes this an ideal case for visual analysis before computing any
correlation coefficient.

```{r}
set.seed(123)
process_data <- data.frame(
  cycle_time = runif(200, 8, 15),
  defect_rate = NA
)
process_data$defect_rate <- with(process_data,
  2 + 0.8 * (cycle_time - 11.5)^2 + rnorm(200, 0, 0.5))

summary(process_data)
#>   cycle_time     defect_rate
#> Min.   : 8.02   Min.   : 1.43
#> 1st Qu.: 9.74   1st Qu.: 3.02
#> Median :11.43   Median : 5.61
#> Mean   :11.45   Mean   : 6.35
#> 3rd Qu.:13.22   3rd Qu.: 9.11
#> Max.   :14.99   Max.  :12.84
```

**Step 1 — Basic scatter with regression.** The fitted line is nearly flat,
and the Pearson $r$ is small, which would tempt the analyst to conclude
that cycle time is irrelevant. The plot reveals the truth: the
relationship is **U-shaped**, not linear.

```{r}
plot_scatter_basic(
  data              = process_data,
  x_var             = "cycle_time",
  y_var             = "defect_rate",
  add_regression    = TRUE,
  add_ci            = TRUE,
  add_correlation   = TRUE,
  correlation_method = "pearson",
  title             = "Defect rate vs cycle time",
  subtitle          = "Linear fit is misleading: the relationship is U-shaped"
)
```

**Step 2 — Spearman correlation.** Because the relationship is monotonic on
each side of the minimum but not overall, Spearman's $\rho$ is also small.
Both coefficients miss the structure that the eye sees instantly.

```{r}
plot_scatter_basic(
  data               = process_data,
  x_var              = "cycle_time",
  y_var              = "defect_rate",
  correlation_method = "spearman",
  add_ci             = FALSE,
  title              = "Same data, Spearman rank correlation"
)
```

**Step 3 — Density scatter for the full picture.** With $n = 200$ the cloud
is dense enough to benefit from alpha blending and a smoother:

```{r}
plot_scatter_density(
  data         = process_data,
  x_var        = "cycle_time",
  y_var        = "defect_rate",
  method       = "alpha",
  point_alpha  = 0.3,
  add_smooth   = TRUE,
  smooth_method = "loess",
  title        = "Density scatter with loess smoother",
  subtitle     = "The smoother traces the U-shape that linear correlation misses"
)
```

**Step 4 — Grouped scatter by shift.** The data are tagged with the shift
on which each part was produced. Grouping reveals whether the U-shape is
consistent across shifts or driven by one shift.

```{r}
process_data$shift <- sample(c("A", "B", "C"), 200, replace = TRUE)

plot_scatter_grouped(
  data             = process_data,
  x_var            = "cycle_time",
  y_var            = "defect_rate",
  group_var        = "shift",
  add_regression   = FALSE,
  add_correlation  = FALSE,
  point_alpha      = 0.6,
  title            = "Defect rate vs cycle time by shift",
  subtitle         = "Color reveals whether shifts differ in behavior"
)
```

**Step 5 — Bubble plot adding throughput.** A third variable, parts per
hour, is encoded as bubble size to assess whether faster cycles (which
raise throughput) also raise defects:

```{r}
process_data$throughput <- with(process_data, 3600 / cycle_time)

plot_scatter_bubble(
  data       = process_data,
  x_var      = "cycle_time",
  y_var      = "defect_rate",
  size_var   = "throughput",
  group_var  = "shift",
  size_range = c(1.5, 8),
  point_alpha = 0.6,
  title      = "Bubble plot: cycle time, defect rate, throughput",
  subtitle   = "Bubble size encodes parts per hour"
)
```

**Step 6 — Decision.** The visual evidence overturns the initial
conclusion from the correlation coefficient alone. The engineer should
**center the cycle time near 11.5 s** (the U-minimum) rather than treat
cycle time as irrelevant. This scenario illustrates why scatter plots are
indispensable: they expose non-linearity that no single correlation
summary can capture.

## 8. Parameter Reference

### plot_scatter_basic()

| Argument | Type | Default | Description |
|---|---|---|---|
| `data` | data.frame | — | Data frame |
| `x_var` | bare name | — | Predictor column |
| `y_var` | bare name | — | Response column |
| `add_regression` | logical | `TRUE` | Add least-squares line |
| `add_ci` | logical | `TRUE` | Add confidence band |
| `add_correlation` | logical | `TRUE` | Print correlation coefficient |
| `correlation_method` | character | `"pearson"` | `"pearson"`, `"spearman"`, `"kendall"` |
| `point_color` | character | `NULL` | Point color; `NULL` resolves to theme `data` role |
| `point_size` | numeric | `2` | Point size |
| `point_alpha` | numeric | `0.8` | Point transparency |
| `line_color` | character | `NULL` | Regression line color; `NULL` resolves to theme `danger` semantic |
| `line_size` | numeric | `1.2` | Line width |
| `ci_fill` | character | `NULL` | Band fill color; `NULL` resolves to theme `primary` semantic |
| `ci_alpha` | numeric | `0.2` | Band transparency |
| `theme` | character/IqrTheme | `NULL` | Theme name or object; propagates to `base_plot()` and `.iqr_aes()` |
| `title`, `subtitle` | character | `NULL` | Plot titles |

### plot_scatter_grouped()

| Argument | Type | Default | Description |
|---|---|---|---|
| `data` | data.frame | — | Data frame |
| `x_var`, `y_var` | bare name | — | Numeric columns |
| `group_var` | bare name | — | Categorical grouping column |
| `add_regression` | logical | `TRUE` | Per-group regression lines |
| `add_ci` | logical | `FALSE` | Per-group confidence bands |
| `add_correlation` | logical | `TRUE` | Print correlation per group |
| `correlation_method` | character | `"pearson"` | Correlation method |
| `point_size` | numeric | `2` | Point size |
| `point_alpha` | numeric | `0.8` | Point transparency |
| `line_size` | numeric | `1.2` | Line width |
| `ci_alpha` | numeric | `0.2` | Band transparency |
| `palette` | named vector | `NULL` | Custom color palette; `NULL` uses `theme$scale_color_iqr()` |
| `theme` | character/IqrTheme | `NULL` | Theme name or object for discrete color scale |
| `title`, `subtitle` | character | `NULL` | Plot titles |

### plot_scatter_bubble()

| Argument | Type | Default | Description |
|---|---|---|---|
| `data` | data.frame | — | Data frame |
| `x_var`, `y_var` | bare name | — | Numeric columns |
| `size_var` | bare name | — | Column encoded as bubble size |
| `group_var` | bare name | `NULL` | Optional grouping column |
| `size_range` | numeric | `c(2, 10)` | Min/max bubble radius |
| `add_labels` | logical | `FALSE` | Add text labels |
| `label_var` | bare name | `NULL` | Column of labels |
| `label_size` | numeric | `3` | Label font size |
| `point_alpha` | numeric | `0.7` | Bubble transparency |
| `theme` | character/IqrTheme | `NULL` | Theme name or object; color scale resolved via `.iqr_aes()` |
| `title`, `subtitle` | character | `NULL` | Plot titles |

### plot_scatter_density()

| Argument | Type | Default | Description |
|---|---|---|---|
| `data` | data.frame | — | Data frame |
| `x_var`, `y_var` | bare name | — | Numeric columns |
| `method` | character | `"alpha"` | `"alpha"`, `"jitter"`, `"bins"`, `"hex"`, `"density"` |
| `point_alpha` | numeric | `0.5` | Transparency for alpha/jitter |
| `point_size` | numeric | `2` | Point size for alpha/jitter |
| `jitter_width` | numeric | `0.1` | Horizontal jitter amplitude |
| `jitter_height` | numeric | `0.1` | Vertical jitter amplitude |
| `bins` | integer | `30` | Bin count for `bins`/`hex` |
| `color_low` | character | `NULL` | Low-density color; `NULL` uses theme `surface_soft` |
| `color_high` | character | `NULL` | High-density color; `NULL` uses theme sequential scale |
| `add_smooth` | logical | `FALSE` | Overlay smoother |
| `smooth_method` | character | `"loess"` | Method passed to `geom_smooth` |
| `theme` | character/IqrTheme | `NULL` | Theme name or object; sequential scale auto-applied for `bins`/`hex`/`density` |
| `title`, `subtitle` | character | `NULL` | Plot titles |

## 9. Summary

* **plot_scatter_basic()** is the starting point: regression line,
  confidence band, and a correlation coefficient in one call.
* **plot_scatter_grouped()** exposes interactions between a continuous
  predictor and a categorical factor that a pooled plot would hide.
* **plot_scatter_bubble()** adds a third continuous variable without
  faceting.
* **plot_scatter_density()** remedies overplotting with five complementary
  methods; choose `alpha` for moderate $n$, `hex` or `bins` for very
  large $n$, and `density` for a smooth surface.
* **Correlation is not causation.** Always read the plot alongside the
  coefficient, and watch for non-linearity, heteroscedasticity, and
  subgroup structure that a single number cannot summarise.

## 10. Further Reading

* Montgomery, D. C. (2013). *Introduction to Statistical Quality Control*,
  7th ed. Wiley. — Scatter-plot methods in process monitoring.
* Tukey, J. W. (1977). *Exploratory Data Analysis*. Addison-Wesley. —
  Origin of jittering, binning, and smoothing ideas.
* Carr, D. B., et al. (1987). "Scatterplot Matrix Techniques for Large
  N." *Journal of the American Statistical Association*. — Hexagonal
  binning foundations.
* Wickham, H. (2016). *ggplot2: Elegant Graphics for Data Analysis*.
  Springer. — `geom_smooth`, `geom_hex`, and density geoms.

## References

Carr, D. B., Littlefield, R. J., Nicholson, W. L., & Littlefield, J. S.
(1987). Scatterplot matrix techniques for large N. *Journal of the
American Statistical Association*, 82(398), 424–436.

Montgomery, D. C. (2013). *Introduction to Statistical Quality Control*
(7th ed.). Hoboken, NJ: Wiley.

Tukey, J. W. (1977). *Exploratory Data Analysis*. Reading, MA:
Addison-Wesley.

Wickham, H. (2016). *ggplot2: Elegant Graphics for Data Analysis*.
New York: Springer.
