---
title: "Turtle Diagram in Quality Management"
author: "iQualityR Team"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Turtle Diagram in Quality Management}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

The Turtle Diagram is a quality management tool for systematically describing and analyzing processes. It originated from the automotive industry standard IATF 16949 (formerly ISO/TS 16949). It visualizes all relevant dimensions of a process through seven key elements (body, head, tail, and four legs), forming a turtle-like shape. The Turtle Diagram helps teams:

- Reach a shared understanding of the process
- Identify potential risks in inputs, resources, methods, etc.
- Meet the process approach requirements of IATF 16949, ISO 9001, and other standards
- Serve as the basis for FMEA, control plans, and work instructions

This package provides the `plot_turtle_diagram()` function, allowing users to quickly generate high-quality, customizable Turtle Diagrams with simple R code, and supports export to PNG, SVG, PDF, and other formats.

## Theoretical Framework

The seven core questions of the Turtle Diagram correspond as follows:

| Body Part | Question | Corresponding Element |
|--------|----------------------------|----------------|
| Body | What is the process? Who is responsible? | Process name, owner, activity steps |
| Head | What is produced? To whom? | Outputs (products, records, downstream) |
| Tail | What input is needed? From where? | Inputs (materials, information, upstream) |
| Left front leg | What resources are needed? | Equipment, tools, software |
| Right front leg | What method is followed? | Procedures, standards, work instructions |
| Left rear leg | How is performance measured? | Indicators (KPIs), target values |
| Right rear leg | Who executes? What capabilities are needed? | Roles, responsibilities, competency requirements |

When filling in each element, follow the **SMART** principle (Specific, Measurable, Achievable, Relevant, Time-bound) and ensure alignment with upstream and downstream processes.

## Installation and Loading

```{r}
library(iQualityR.core)
library(iQualityR.plot)
library(DiagrammeR)
```

## Basic Usage

The simplest invocation — specify only the process name, with other parameters using default values:

```{r}
plot_turtle_diagram(process_name = "Order Processing")
```

The default direction is head-up (`direction = "T"`), the six support boxes are filled with the same light gray-blue, and the center circle is a soft blue.

## Complete Example: Assembly Process

The following example includes all seven elements and uses the `"academic"` theme (white background, gray borders).

```{r}
g <- plot_turtle_diagram(
  process_name = "Assembly Process",
  process_owner = "Production Manager",
  activities = c("Receive components", "Position fixture", "Tighten screws", "Check torque", "Apply label"),
  inputs = c("Aluminum profiles", "Screws"),
  outputs = c("Assembly", "Self-inspection form"),
  resources = c("Torque wrench", "Pneumatic screwdriver", "Custom fixture", "MES terminal"),
  procedures = c("SOP-ASSY-01", "Torque standard 5.0+/-0.5 N*m", "First article inspection process"),
  metrics = c("First-pass yield >=98%", "Cycle time <=3 min/unit", "Torque CPk>=1.33"),
  responsibilities = c("Assembler (certified)", "Quality patrol inspector", "Production team leader"),
  theme_style = "tech",
  title = "Turtle Diagram Example: Assembly Process",
  subtitle = "Version 1.0",
  annotation = "Note: Based on IATF 16949 process approach"
)
g
```

After executing the above code, the diagram will automatically appear in the RStudio Viewer pane.

## Custom Fill Colors

Using the `user_colors` parameter, you can set the fill colors for the center circle and the six support boxes separately (it is recommended to use the same color for all six boxes to avoid visual clutter).

```{r}
my_colors <- list(
  process = "#FFB6C1",   # Center circle (pink)
  inputs  = "#C0E0FF",
  outputs = "#C0E0FF",
  resources = "#C0E0FF",
  procedures = "#C0E0FF",
  metrics = "#C0E0FF",
  responsibilities = "#C0E0FF"
)

plot_turtle_diagram(
  process_name = "Order Processing",
  user_colors = my_colors,
  title = "Custom Color Example"
)
```

## Adjusting Direction

The Turtle Diagram supports head-up (`"T"`), head-down (`"B"`), head-left (`"L"`), and head-right (`"R"`), with limb positions automatically adapting.

```{r}
plot_turtle_diagram(
  process_name = "Logistics Shipping",
  direction = "L",
  title = "Head-Left Turtle Diagram"
)
```

## Saving the Diagram

The diagram object (class `grViz`) generated by the function can be exported to PNG, SVG, or PDF format via the `save_diagram()` utility function.

```{r}
g <- plot_turtle_diagram(process_name = "Assembly Process")
save_diagram(g, "assembly_turtle.png", type = "png")
```

Batch generation of Turtle Diagrams for different processes:

```{r}
processes <- c("Injection Molding", "Assembly", "Testing")
for (proc in processes) {
  g <- plot_turtle_diagram(process_name = proc)
  save_diagram(g, filename = paste0(proc, "_turtle"), type = "png")
}
```

## Practical Case: Injection Molding Process

**Background**: An automotive parts supplier needs to conduct a standardized analysis of its "Injection Molding Process" to meet IATF 16949 requirements.

**Process Information**:

| Element | Content |
|--------|--------------------------------------------------------------|
| Process Name | Injection Molding |
| Process Owner | Injection Molding Workshop Director |
| Core Activities | Material preparation -> Drying -> Barrel heating -> Mold closing & injection -> Pressure holding & cooling -> Removal & inspection |
| Inputs | Plastic pellets (warehouse), molds (mold storage), process cards (engineering) |
| Outputs | Injection molded parts (quality), molding parameter records, scrap labels |
| Resources | Injection molding machine, robot arm, mold temperature controller, dryer, vernier caliper |
| Methods | WI-IM-01 Injection Molding Work Instruction, Process parameter table (temperature/pressure/speed) |
| Metrics | Dimension CPk >=1.33, Appearance pass rate >=99.5%, Mold change time <=30 min |
| Responsibilities | Operator (certified), Technician (parameter adjustment), Inspector (first article confirmation) |

**Generation Code**:

```{r}
plot_turtle_diagram(
  process_name = "Injection Molding",
  process_owner = "Injection Molding Workshop Director",
  activities = c("Material preparation", "Drying", "Barrel heating", "Mold closing & injection", "Pressure holding & cooling", "Removal & inspection"),
  inputs = c("Plastic pellets", "Molds", "Process cards"),
  outputs = c("Injection molded parts", "Molding parameter records", "Scrap labels"),
  resources = c("Injection molding machine", "Robot arm", "Mold temperature controller", "Dryer", "Vernier caliper"),
  procedures = c("WI-IM-01 Injection Molding Work Instruction", "Process parameter table (temperature/pressure/speed)"),
  metrics = c("Dimension CPk >=1.33", "Appearance pass rate >=99.5%", "Mold change time <=30 min"),
  responsibilities = c("Operator (certified)", "Technician", "Inspector"),
  theme_style = "tech",
  title = "Injection Molding Process Turtle Diagram",
  subtitle = "Compliant with IATF 16949 Process Analysis Requirements"
)
```

This Turtle Diagram can be directly used in quality manuals, process risk analyses, or internal training materials.

## FAQ and Best Practices

**Q1: How to determine the granularity of a Turtle Diagram?**  
Principle: One Turtle Diagram corresponds to one independent process (with clear inputs, outputs, and boundaries). Avoid merging multiple large processes (e.g., "Production Management" should be split into "Scheduling", "Material Preparation", "Production", "Inspection", etc.).

**Q2: How to align inputs/outputs with upstream and downstream processes?**  
The "downstream process" of the output should match the name of another Turtle Diagram, ensuring continuity of the process map. For example, the output "Injection molded parts" from injection molding should be exactly the input of the "Assembly Process".

**Q3: How should metrics be defined?**  
Use the SMART principle: **S**pecific, **M**easurable, **A**chievable, **R**elevant, **T**ime-bound. For example, "First-pass yield >=98%" is more effective than "Good quality".

**Q4: How to perform risk analysis based on the Turtle Diagram?**  
Ask "What if missing or failing?" for each element. For example:
- Missing input -> Process cannot start
- Resource failure -> Process interruption
- Metrics not met -> Improvement needed
These risks can be fed into FMEA.

**Q5: Why do texts overlap or positions appear disordered in my diagram?**  
Adjust parameters such as `node_width`, `node_height`, `process_width`, `process_height` (passed through the `theme` list). Also, ensure input text is not too long (split or abbreviate if necessary).

## Conclusion

The Turtle Diagram is a simple yet powerful process visualization tool. Through the `plot_turtle_diagram()` function, quality engineers, process owners, and R users can quickly generate Turtle Diagrams that meet industry standards and easily integrate them into documents or reports. The function supports extensive customization (colors, direction, title, footnotes) and provides export functionality for use in batch production environments.

Future version plans include:
- More built-in themes (e.g., corporate color schemes)
- Interactive editing (via `shiny` or `htmlwidgets`)
- Direct export to Word or PowerPoint formats

Feedback and suggestions are welcome!

## References

1. IATF 16949:2016 - Automotive Quality Management System Requirements
2. ISO 9001:2015 - Quality Management Systems - Requirements
3. AIAG Process Approach Guide (2008)
