---
title: "Turtle Diagrams for Process Mapping"
author: "iQualityR.plot Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Turtle Diagrams for Process Mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

library(iQualityR.plot)
library(iQualityR.core)
library(ggplot2)

fmt_num <- function(x, digits = 4) formatC(x, format = "f", digits = digits)
fmt_pct <- function(x, digits = 2) paste0(formatC(x, format = "f", digits = digits), "%")
```

## 1. Introduction

A quality management system lives or dies by how well its processes are
understood. The **turtle diagram** is a one-page visual artifact that captures
every essential facet of a single process on a single canvas, making it a
favorite of ISO 9001 auditors and process owners alike. Where a flowchart shows
*what happens next*, a turtle diagram answers the broader question *what does
this process need to operate correctly, and how do we know it worked?*

The `iQualityR.plot` package provides `plot_turtle_diagram()`, a function that
renders a turtle diagram as a Graphviz graph integrated with the package's
`IqrTheme` system. The output is an interactive `DiagrammeR::grViz` object that
can be displayed in the RStudio viewer, embedded in an R Markdown document, or
exported to PNG, SVG, or PDF for inclusion in quality manuals and audit
evidence packs.

A defining feature of `plot_turtle_diagram()` is that **every box color is
resolved automatically from the active `IqrTheme`**. The process body pulls
the theme's `primary` UI color, while the six surrounding element boxes
(inputs, outputs, resources, procedures, metrics, responsibilities) pull the
first six entries of the theme's discrete palette. Switching the theme -- by
name (`"academic"`, `"economist"`, `"solarized"`, ...) or by passing a custom
`IqrTheme` object -- recolors the entire diagram in one step, so the turtle
diagram stays visually consistent with the rest of the `iQualityR.plot`
family (control charts, Pareto charts, fishbone diagrams) without any
per-element color arguments.

This vignette develops the theory of the process approach, walks through the
seven elements of a turtle diagram, demonstrates the function with a worked
calibration-process example, and closes with a parameter reference.

## 2. Theory: The Process Approach (ISO 9001)

### 2.1 What a turtle diagram represents

ISO 9001:2015 promotes the **process approach** in Clause 0.3, which asks
organizations to understand and manage their interrelated processes as a
coherent system rather than as isolated procedures. The standard defines a
process as a set of interrelated activities that transform inputs into outputs
to deliver a planned result. A turtle diagram is the visual expression of that
definition: a central body of activities flanked by the resources,
responsibilities, methods, and measures that make the transformation
consistent and controllable.

The metaphor is anatomical. The *shell* (the central body) carries the
process activities. The *head* points toward the methods and procedures that
govern the work. The *flippers* hold the resources above and the
responsibilities below. The *tail* carries the measures that tell the
organization whether the process is behaving. Inputs flow in from the left and
outputs flow out to the right, completing the transformation picture.

ISO 9001 connects turtle diagrams to several operative clauses:

* **Clause 8.2** — Requirements for products and services: inputs and outputs
  make explicit what the customer needs and what the process delivers.
* **Clause 8.5** — Production and service provision: activities, resources,
  and procedures are the controlled conditions of provision.
* **Clause 9.1** — Monitoring, measurement, analysis, and evaluation: the
  measures element is the direct evidence base for Clause 9.1.

### 2.2 The seven elements: Inputs, Outputs, Process, Resources, Responsibility, Measures, Methods

A complete turtle diagram always carries seven elements. Leaving any of them
blank is a signal that the process is not yet fully specified.

1. **Inputs** — materials, information, or requirements entering the process.
  An input is something the process consumes or reacts to; without it the
  process cannot start.
2. **Outputs** — products, services, or information leaving the process.
  Outputs must be verifiable against input requirements, closing the
  transformation loop.
3. **Process (body)** — the core activities and transformations that turn
  inputs into outputs. This is the *what we do* of the process.
4. **Resources (top flipper)** — equipment, people, facilities, software, and
  materials required to perform the activities. Resources answer *what do we
  need to do it well?*
5. **Responsibility (bottom flipper)** — who owns and runs the process.
  ISO 9001 Clause 5.3 requires roles and responsibilities to be assigned; the
  responsibility element makes this explicit for a single process.
6. **Measures (tail)** — key performance indicators and metrics used to
  monitor the process. These feed Clause 9.1 monitoring and are the basis for
  data-driven improvement.
7. **Methods / Procedures (head direction)** — standard operating procedures,
  work instructions, and standards that define *how* the work is done. These
  are the controlled documents referenced in Clause 7.5.

The head direction of the turtle indicates where the methods element is drawn.
The default `"T"` (top) places methods above the body alongside resources, but
the diagram can be reoriented to `"B"`, `"L"`, or `"R"` to fit the available
page layout.

## 3. plot_turtle_diagram() — single process

The `plot_turtle_diagram()` function accepts the seven elements as character
vectors and renders a Graphviz-based turtle diagram. The minimal call
specifies only the process name and the activities; the remaining elements
default to placeholders that the user is expected to fill in.

Let us begin with a minimal diagram showing only the process body and the
seven-element skeleton.

```{r}
minimal_turtle <- plot_turtle_diagram(
  process_name = "Raw Material Inspection",
  activities = c(
    "Receive material lot",
    "Verify certificate of analysis",
    "Sample per AQL plan",
    "Perform incoming tests",
    "Record results in ERP"
  )
)

minimal_turtle
```

The function returns a `DiagrammeR::grViz` object invisibly, which renders
inline in the viewer. The default `theme_style = "workbench"` palette uses
muted engineering tones suitable for documentation: the process body is
filled with the theme's `primary` UI color and the six element boxes take
the first six entries of the theme's discrete palette, so the entire diagram
is themed without any color argument. Each activity is listed in the central
body, while the surrounding boxes prompt the user to supply the remaining six
elements.

A more complete call fills in every element. Here we map a simple document
control process.

```{r}
doc_control <- plot_turtle_diagram(
  process_name = "Document Control Process",
  process_owner = "Document Control Coordinator",
  activities = c(
    "Receive document for review",
    "Check format against template",
    "Route for approval signatures",
    "Publish to controlled repository",
    "Distribute change notification"
  ),
  inputs = c(
    "Draft document from author",
    "Document change request form",
    "Current revision history"
  ),
  outputs = c(
    "Approved controlled document",
    "Updated master list",
    "Obsolete document withdrawn"
  ),
  resources = c(
    "Document management system",
    "Approval workflow software",
    "Trained document coordinators"
  ),
  procedures = c(
    "SOP-DOC-001: Document Control",
    "ISO 9001:2015 Clause 7.5",
    "Template library"
  ),
  metrics = c(
    "Average approval cycle time",
    "Document revision rate",
    "Outdated-document incidents"
  ),
  responsibilities = c(
    "Document Control Coordinator - owns process",
    "Process Owner - approves content",
    "QA Manager - audits compliance"
  )
)

doc_control
```

## 4. Customizing the turtle diagram

### 4.1 Theme style and IqrTheme objects

The `theme_style` argument accepts `"tech"`, `"academic"`, or `"workbench"`.
Each preset applies a distinct color palette to the seven element boxes. For
audit submissions and academic reports the `"academic"` style is often
preferred because it uses high-contrast monochrome tones that print cleanly in
black and white.

Internally, every preset is resolved into an `IqrTheme` object via
`as_iqr_theme_object()`, and the seven box fills are pulled from two theme
slots:

| Diagram element              | Theme slot                  |
|------------------------------|-----------------------------|
| Process body                 | `ui$primary`                |
| Inputs box                   | `discrete` palette entry 1  |
| Outputs box                  | `discrete` palette entry 2  |
| Resources box                | `discrete` palette entry 3  |
| Procedures / methods box     | `discrete` palette entry 4  |
| Metrics box                  | `discrete` palette entry 5  |
| Responsibilities box         | `discrete` palette entry 6  |

Because the mapping is fixed, switching the theme recolors the whole diagram
deterministically. The chunk below renders the same calibration skeleton under
three preset themes so the difference is visible at a glance.

```{r}
calib_activities <- c(
  "Receive instrument",
  "Verify environmental conditions",
  "Perform calibration against reference standard",
  "Record results",
  "Issue calibration certificate"
)

plot_turtle_diagram(
  process_name = "Calibration - workbench",
  activities = calib_activities, theme_style = "workbench"
)

plot_turtle_diagram(
  process_name = "Calibration - academic",
  activities = calib_activities, theme_style = "academic"
)

plot_turtle_diagram(
  process_name = "Calibration - tech",
  activities = calib_activities, theme_style = "tech"
)
```

```{r}
plot_turtle_diagram(
  process_name = "Internal Audit Process",
  activities = c("Plan audit", "Conduct opening meeting", "Gather evidence"),
  theme_style = "academic",
  title = "Internal Audit Process Map",
  subtitle = "Prepared for ISO 9001 surveillance audit"
)
```

When an organization has codified its own colors in an `IqrTheme` object, the
`iqr_theme` argument takes precedence over `theme_style`. This is the
recommended route for enterprise consistency: build the theme once with
`IqrTheme$new()` and pass it to every plot function in the package. The
process body then inherits the corporate `primary` color and the six element
boxes inherit the corporate discrete palette, so the turtle diagram matches
the rest of the quality dashboard automatically.

`IqrTheme$new(style, ...)` starts from a built-in preset and accepts nested
overrides via `...`. The `merge_config` step does a deep merge, so you can
override individual slots without re-specifying the whole configuration --
for example, replacing only the `primary` UI color and the `discrete` data
palette while keeping the rest of the preset.

```{r}
# A corporate IqrTheme: start from "academic", override the primary UI color
# and the discrete data palette so all seven turtle boxes track the brand.
custom_theme <- IqrTheme$new(
  "academic",
  ui   = list(primary = "#1F3A5F"),
  data = list(discrete = c("#1F3A5F", "#C9A227", "#4C8C2B", "#6A4C93",
                           "#A8431F", "#0F766E", "#7C3AED", "#0891B2"))
)

plot_turtle_diagram(
  process_name = "Supplier Qualification Process",
  activities = c(
    "Receive supplier questionnaire",
    "Perform risk assessment",
    "Conduct on-site audit",
    "Approve or reject supplier"
  ),
  iqr_theme = custom_theme,
  theme_style = "tech"
)
```

### 4.2 User-defined colors (one-off overrides)

The theme is the primary color mechanism and should cover the overwhelming
majority of use cases. The `user_colors` argument is an escape hatch for the
rare situation where a single diagram must highlight one element outside the
theme system -- for example, flagging a broken process step in a corrective
action report. It accepts a named vector; recognized names map to the seven
elements: `body`, `inputs`, `outputs`, `resources`, `responsibilities`,
`measures`, and `methods`. Any name not supplied falls back to the theme
palette, so you can override a single box without re-specifying the other six.

The recommended pattern is to derive even one-off overrides from the active
theme's semantic palette (e.g. `fail` for a broken step, `watch` for a
degraded one) rather than hard-coding raw hex values, so the diagram still
tracks the corporate palette.

```{r}
# Highlight the "resources" box with the theme's "fail" semantic color to
# signal a resourcing gap; the other six boxes keep their theme fills.
theme_obj <- as_iqr_theme_object("workbench")
fail_col  <- theme_obj$get_pal("semantic", name = "fail")

plot_turtle_diagram(
  process_name = "Training Management Process",
  activities = c(
    "Identify training needs",
    "Schedule sessions",
    "Deliver training",
    "Evaluate effectiveness"
  ),
  theme_style  = "workbench",
  user_colors  = c(resources = fail_col)
)
```

For ad-hoc diagrams that are not part of a themed report, raw hex values are
also accepted:

```{r, eval = FALSE}
plot_turtle_diagram(
  process_name = "Training Management Process",
  activities = c("Identify training needs", "Deliver training"),
  user_colors = c(body = "#2E5C8A", resources = "#E8A33D", measures = "#4C8C2B")
)
```

### 4.3 Head direction

The `direction` argument controls where the methods/procedures box is drawn
relative to the body. The default `"T"` places it on top. On a portrait page,
`"R"` (right) often produces a more balanced layout.

```{r}
plot_turtle_diagram(
  process_name = "Corrective Action Process",
  activities = c(
    "Identify nonconformity",
    "Determine root cause",
    "Implement corrective action",
    "Verify effectiveness"
  ),
  direction = "R",
  theme_style = "tech"
)
```

### 4.4 Title, subtitle, and annotation

A title and subtitle appear above the diagram, while an annotation line is
drawn at the bottom — useful for recording the document number, revision, and
approval date required by controlled-document systems.

```{r}
plot_turtle_diagram(
  process_name = "Risk Management Process",
  activities = c(
    "Identify risks",
    "Assess severity and likelihood",
    "Define mitigation actions",
    "Monitor residual risk"
  ),
  title = "Risk Management Process Map",
  subtitle = "Aligned with ISO 31000 and ISO 9001:2015 Clause 6.1",
  subtitle_fontsize = 12,
  annotation = "Document No. QM-PR-007  |  Rev. 3  |  Approved: 2025-11-04",
  annotation_fontsize = 9
)
```

### 4.5 Exporting to file

Setting `output_file` writes the diagram to disk in the format specified by
`output_type`. PNG is appropriate for slide decks, SVG for editable graphics
in technical documentation, and PDF for inclusion in printed quality manuals.

```{r, eval=FALSE}
plot_turtle_diagram(
  process_name = "Calibration Process",
  activities = c("Receive instrument", "Perform calibration", "Issue certificate"),
  output_file = "calibration_turtle",
  output_type = "png"
)
```

## 5. Quality Scenario: mapping the calibration process

To bring the elements together we now map a realistic metrology process: the
**Instrument Calibration Process** operated by a manufacturing site's
Metrology Laboratory. Calibration is a high-risk process because every
downstream measurement depends on the integrity of the calibrated instrument,
so a complete turtle diagram is essential audit evidence under both ISO 9001
and ISO/IEC 17025.

### 5.1 Process context

Instruments used for product verification must be calibrated against
traceable reference standards at defined intervals. The metrology lab receives
an uncalibrated instrument, verifies that environmental conditions support a
valid calibration, performs the calibration against a reference standard,
records the results, issues a certificate, and applies a calibration label.
The process owner is the Metrology Lab Manager, and the work is governed by
ISO/IEC 17025:2017 and a family of site standard operating procedures.

### 5.2 Building the diagram

We pass the seven elements as character vectors. Note that long text strings
wrap automatically inside each box, so there is no need to pre-format the
content.

```{r}
calibration_turtle <- plot_turtle_diagram(
  process_name = "Instrument Calibration Process",
  process_owner = "Metrology Lab Manager",
  activities = c(
    "Receive instrument",
    "Verify environmental conditions",
    "Perform calibration against reference standard",
    "Record results",
    "Issue calibration certificate",
    "Apply calibration label"
  ),
  inputs = c(
    "Uncalibrated instrument",
    "Calibration procedure document",
    "Previous calibration record"
  ),
  outputs = c(
    "Calibrated instrument",
    "Calibration certificate",
    "Updated calibration database"
  ),
  resources = c(
    "Reference standards (traceable to NIST)",
    "Calibration software",
    "Climate-controlled lab",
    "Trained calibration technician"
  ),
  procedures = c(
    "ISO 17025:2017",
    "SOP-CAL-001: General Calibration Procedure",
    "SOP-CAL-012: Temperature Calibration"
  ),
  metrics = c(
    "Calibration interval adherence rate",
    "Out-of-tolerance events per month",
    "Average calibration turnaround time",
    "Repeatability of reference standards"
  ),
  responsibilities = c(
    "Metrology Lab Manager - overall process",
    "Calibration Technician - execution",
    "QA Manager - periodic review"
  ),
  direction = "T",
  theme_style = "workbench",
  title = "Instrument Calibration Process",
  subtitle = "Metrology Laboratory  |  ISO/IEC 17025:2017",
  subtitle_fontsize = 12,
  annotation = "Document No. MET-PR-014  |  Rev. 2  |  Owner: Metrology Lab Manager",
  annotation_fontsize = 9
)

calibration_turtle
```

### 5.3 Interpreting the diagram for an audit

An auditor reading this turtle diagram can verify each ISO 9001 clause
efficiently:

* **Inputs and outputs** map to Clause 8.2 requirements — the uncalibrated
  instrument and procedure document enter, and a calibrated instrument plus a
  certificate leave.
* **Activities and procedures** map to Clause 8.5 controlled conditions — the
  six activity steps are governed by SOP-CAL-001 and ISO/IEC 17025.
* **Resources** map to Clause 7.1.4 environment for the operation of
  processes — the climate-controlled lab and traceable reference standards are
  explicit.
* **Responsibilities** map to Clause 5.3 organizational roles — three roles
  are named with their scope.
* **Measures** map to Clause 9.1 monitoring — four KPIs are listed, including
  the out-of-tolerance event count that drives corrective action.

### 5.4 Exporting the calibration diagram

Because the calibration turtle is part of the controlled quality manual, we
export it as a PDF for the document control repository.

```{r, eval=FALSE}
plot_turtle_diagram(
  process_name = "Instrument Calibration Process",
  process_owner = "Metrology Lab Manager",
  activities = c(
    "Receive instrument",
    "Verify environmental conditions",
    "Perform calibration against reference standard",
    "Record results",
    "Issue calibration certificate",
    "Apply calibration label"
  ),
  inputs = c(
    "Uncalibrated instrument",
    "Calibration procedure document",
    "Previous calibration record"
  ),
  outputs = c(
    "Calibrated instrument",
    "Calibration certificate",
    "Updated calibration database"
  ),
  resources = c(
    "Reference standards (traceable to NIST)",
    "Calibration software",
    "Climate-controlled lab",
    "Trained calibration technician"
  ),
  procedures = c(
    "ISO 17025:2017",
    "SOP-CAL-001: General Calibration Procedure",
    "SOP-CAL-012: Temperature Calibration"
  ),
  metrics = c(
    "Calibration interval adherence rate",
    "Out-of-tolerance events per month",
    "Average calibration turnaround time",
    "Repeatability of reference standards"
  ),
  responsibilities = c(
    "Metrology Lab Manager - overall process",
    "Calibration Technician - execution",
    "QA Manager - periodic review"
  ),
  title = "Instrument Calibration Process",
  subtitle = "Metrology Laboratory  |  ISO/IEC 17025:2017",
  output_file = "calibration_process_map",
  output_type = "pdf"
)
```

## 6. Parameter Reference

The table below summarizes every argument of `plot_turtle_diagram()`.

| Argument              | Type                | Default      | Description                                                       |
|-----------------------|---------------------|--------------|-------------------------------------------------------------------|
| `process_name`        | character(1)        | —            | Name of the process, drawn at the top of the body.                |
| `process_owner`       | character(1)        | `NULL`       | Process owner; appears in the responsibility box header.          |
| `activities`          | character           | `NULL`       | Core activity steps, listed in the central body.                  |
| `inputs`              | character/list      | `NULL`       | Inputs entering the process (left side).                         |
| `outputs`             | character/list      | `NULL`       | Outputs leaving the process (right side).                        |
| `resources`           | character           | `NULL`       | Resources required (top flipper).                                |
| `procedures`          | character           | `NULL`       | Methods, SOPs, and standards (head direction).                   |
| `metrics`             | character           | `NULL`       | KPIs and measures (tail).                                        |
| `responsibilities`    | character           | `NULL`       | Roles and responsibility assignments (bottom flipper).           |
| `direction`           | character(1)        | `"T"`        | Head direction: `"T"`, `"B"`, `"L"`, or `"R"`.                   |
| `theme_style`         | character(1)        | `"workbench"`| Preset name resolved into an `IqrTheme`: `"tech"`, `"academic"`, or `"workbench"`. Drives all seven box fills (body <- `ui$primary`; elements <- `discrete` 1-6). |
| `iqr_theme`           | IqrTheme/NULL       | `NULL`       | Custom `IqrTheme` object (built via `IqrTheme$new()`); takes precedence over `theme_style`. Recommended for enterprise consistency. |
| `user_colors`         | named vector/NULL   | `NULL`       | Per-element one-off overrides; names: `body`, `inputs`, `outputs`, `resources`, `responsibilities`, `measures`, `methods`. Unspecified names fall back to the theme palette. |
| `title`               | character(1)/NULL   | `NULL`       | Diagram title above the figure.                                  |
| `subtitle`            | character(1)/NULL   | `NULL`       | Subtitle line below the title.                                   |
| `subtitle_fontsize`   | numeric             | `14`         | Font size for the subtitle.                                      |
| `annotation`          | character(1)/NULL   | `NULL`       | Annotation line at the bottom (e.g. document number).            |
| `annotation_fontsize` | numeric             | `10`         | Font size for the annotation.                                    |
| `output_file`         | character(1)/NULL   | `NULL`       | If set, exports the diagram to this file path.                   |
| `output_type`         | character(1)        | `"png"`      | Export format: `"png"`, `"svg"`, or `"pdf"`.                     |
| `...`                 | any                 | —            | Additional arguments forwarded to the Graphviz renderer.         |

**Return value:** A `DiagrammeR::grViz` object, returned invisibly. When
`output_file` is set, the file is written to disk in addition to the return
value.

**Color resolution:** The theme is the single source of truth for color.
Internally, `as_iqr_theme_object(iqr_theme %||% theme_style)` resolves the
active `IqrTheme`; the process body fill is pulled from `ui$primary` and the
six element boxes from `discrete` palette entries 1-6. `user_colors` overrides
individual boxes after the theme has been applied. No color argument is
required for the default themed look.

## 7. Summary

The turtle diagram is a compact, audit-ready representation of a single
process. By forcing the author to articulate all seven elements — inputs,
outputs, activities, resources, responsibilities, measures, and methods — it
exposes gaps in process definition that a flowchart alone would hide.
`plot_turtle_diagram()` brings this artifact into the reproducible R workflow:
the diagram is generated from data, themed consistently with the rest of the
`iQualityR.plot` family, and exported to the formats required by controlled
document systems. Used alongside the package's other process-mapping and
statistical tools, it closes the loop between process definition, monitoring,
and improvement that ISO 9001 calls the process approach.

## 8. Further Reading

* ISO 9001:2015 — *Quality management systems — Requirements*, Clauses 0.3,
  5.3, 7.5, 8.2, 8.5, and 9.1.
* ISO/IEC 17025:2017 — *General requirements for the competence of testing
  and calibration laboratories*.
* ISO 31000:2018 — *Risk management — Guidelines*.
* Juran, J. M. and De Feo, J. A. (2010). *Juran's Quality Handbook*, 6th ed.
  McGraw-Hill.
* `iQualityR.plot` package — `plot_fishbone()`, `plot_pareto_enhanced()`, and
  `IqrTheme$new()` for complementary quality-engineering visualizations.

## References

International Organization for Standardization. (2015). *ISO 9001:2015
Quality management systems — Requirements*. Geneva: ISO.

International Organization for Standardization. (2017). *ISO/IEC 17025:2017
General requirements for the competence of testing and calibration
laboratories*. Geneva: ISO.

International Organization for Standardization. (2018). *ISO 31000:2018 Risk
management — Guidelines*. Geneva: ISO.

Juran, J. M., & De Feo, J. A. (2010). *Juran's Quality Handbook: The Complete
Guide to Performance Excellence* (6th ed.). New York: McGraw-Hill.

Socrates, J. A. (2015). *Process Mapping and Management*. Quality Press.
