---
title: "Professional Predictive Modeling Configuration Guide"
author: "iQualityR Development Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Professional Predictive Modeling Configuration Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

```{r setup, echo = FALSE}
library(iQualityR.predict)
```

## 1. Expert Mode Overview

Expert mode is designed for **data scientists, statistical experts, and advanced quality engineers**, providing complete control over the modeling process.

### Business vs Expert Mode Comparison

| Dimension | Business Mode | Expert Mode |
|-----------|---------------|-------------|
| **Algorithm Selection** | Auto | Manual specification |
| **Parameter Tuning** | Default | Fully customizable |
| **Validation Strategy** | 5-fold CV | Any resampling strategy |
| **Feature Engineering** | None | Full recipes package functionality |
| **Model Comparison** | None | Multi-model parallel evaluation |
| **Low-level Access** | None | Direct access to tidymodels objects |

---

## 2. Expert Mode Basic Configuration

### 2.1 Passing Configuration via expert_config

The core of expert mode is the `expert_config` parameter. Once this parameter is passed, the system automatically switches to expert mode.

```{r expert-basic-setup, eval = FALSE}
# Expert mode example
task <- regression_analysis(
  data = your_data,
  target = "your_target",
  expert_config = list(
    # Pass tidymodels/mlr3 configuration here
  )
)
```

### 2.2 Defining Model Specifications

Define models using the `parsnip` package:

```{r model-spec, eval = FALSE}
library(parsnip)

# Linear regression
model_spec <- linear_reg() %>%
  set_engine("lm")

# Ridge regression (with regularization parameter)
model_spec <- linear_reg(penalty = 0.1, mixture = 0) %>%
  set_engine("glmnet")

# Random forest (with hyperparameters)
model_spec <- rand_forest(trees = 1000, min_n = 5) %>%
  set_engine("randomForest") %>%
  set_mode("regression")

# XGBoost
model_spec <- boost_tree(
  trees = 500,
  tree_depth = 6,
  min_n = 10,
  loss_reduction = 0.01,
  sample_size = 0.8,
  learn_rate = 0.01
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")
```

### 2.3 Defining Preprocessing Recipes

Define feature engineering workflows using the `recipes` package:

```{r recipe-def, eval = FALSE}
library(recipes)

rec <- recipe(target ~ ., data = training_data) %>%
  step_normalize(all_numeric_predictors()) %>%      # Standardize
  step_dummy(all_nominal_predictors()) %>%           # One-hot encode categorical variables
  step_nzv(all_predictors()) %>%                     # Remove near-zero variance variables
  step_corr(all_predictors(), threshold = 0.9) %>%   # Remove highly correlated variables
  step_pca(all_numeric_predictors(), num_comp = 5)   # PCA dimensionality reduction
```

---

## 3. Expert Mode Complete Examples

### 3.1 Complete Workflow with Parameter Tuning

```{r full-tuning-example, eval = FALSE}
library(rsample)
library(tune)
library(yardstick)
library(workflows)

# 1. Define resampling strategy
cv_folds <- vfold_cv(solder_data, v = 5, repeats = 3)

# 2. Define tuning grid
rf_grid <- expand.grid(
  trees = c(200, 500, 1000),
  min_n = c(2, 5, 10)
)

# 3. Define evaluation metrics
metrics <- metric_set(rmse, rsq, mae)

# 4. Build expert config
expert_config <- list(
  model_spec = rand_forest(trees = tune(), min_n = tune()) %>%
    set_engine("randomForest", importance = "permutation") %>%
    set_mode("regression"),

  recipe = recipe(strength ~ ., data = solder_data) %>%
    step_normalize(all_numeric_predictors()),

  resampling = cv_folds,
  tune_grid = rf_grid,
  metric = metrics,

  control = control_grid(
    verbose = TRUE,
    save_pred = TRUE,
    extract = function(x) extract_model(x)
  )
)

# 5. Execute modeling
task <- regression_analysis(
  data = solder_data,
  target = "strength",
  expert_config = expert_config
)

task$compute()
```

### 3.2 Accessing Internal Objects

In expert mode, you can access all intermediate objects:

```{r access-internal, eval = FALSE}
# Access internal objects after training

# Access final model
task$results$raw_model

# Access tidymodels workflow object
task$results$workflow

# Access tuning results
task$results$tuning_results

# View tuning metrics
tuning_results <- task$executor$analyzer$tuning_results
show_best(tuning_results, metric = "rmse")

# Visualize tuning results
autoplot(tuning_results, metric = "rmse")
```

---

## 4. Advanced Features

### 4.1 Multi-Model Comparison

```{r model-comparison, eval = FALSE}
# Train and compare multiple models separately
models_list <- list(
  linear = regression_analysis(data, "strength", c("temp", "pressure"),
                               model_type = "linear"),
  rf = regression_analysis(data, "strength", c("temp", "pressure"),
                           model_type = "random_forest"),
  xgb = regression_analysis(data, "strength", c("temp", "pressure"),
                            model_type = "xgboost")
)

# Execute all models
models_list <- lapply(models_list, function(m) m$compute())

# Compare R-squared
sapply(models_list, function(m) m$results$metrics$r_squared)

# Compare RMSE
sapply(models_list, function(m) m$results$metrics$rmse)
```

### 4.2 Custom Workflows

Use the full tidymodels workflow:

```{r custom-workflow, eval = FALSE}
library(tidymodels)

# 1. Data split
set.seed(123)
split <- initial_split(solder_data, prop = 0.8)
train_data <- training(split)
test_data <- testing(split)

# 2. Cross-validation
cv <- vfold_cv(train_data, v = 5)

# 3. Define recipe
rec <- recipe(strength ~ ., data = train_data) %>%
  step_normalize(all_numeric_predictors())

# 4. Define model
mod <- rand_forest(trees = 500, min_n = 5) %>%
  set_engine("randomForest") %>%
  set_mode("regression")

# 5. Workflow
wf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(mod)

# 6. Train
fit_model <- wf %>% fit(data = train_data)

# 7. Test set evaluation
predictions <- predict(fit_model, test_data)
test_data$predicted <- predictions$.pred

# Evaluation metrics
test_data %>%
  rmse(strength, predicted)
```

### 4.3 Using mlr3 Backend

If you prefer the mlr3 ecosystem:

```{r mlr3-backend, eval = FALSE}
library(mlr3)
library(mlr3learners)

# Define mlr3 config
mlr3_config <- list(
  learner = lrn("regr.ranger", num.trees = 500, importance = "permutation"),
  resampling = rsmp("cv", folds = 5)
)

task <- regression_analysis(
  data = solder_data,
  target = "strength",
  backend = "mlr3",
  expert_config = mlr3_config
)

task$compute()
```

---

## 5. Advanced Model Diagnostics and Interpretation

### 5.1 Complete Diagnostic Plots

```{r full-diagnostics, eval = FALSE}
# Generate all diagnostic plots
task$plot(type = "diagnosis")
```

### 5.2 SHAP Value Analysis

```{r shap-analysis, eval = FALSE}
# Enable SHAP values
task_shap <- regression_analysis(
  data = solder_data,
  target = "strength",
  factors = c("temperature", "pressure", "time"),
  explanation = list(
    shap_values = TRUE
  )
)

task_shap$compute()
task_shap$plot(type = "explanation")
```

### 5.3 Partial Dependence Plots (PDP)

```{r pdp-analysis, eval = FALSE}
# Enable partial dependence analysis
task_pdp <- regression_analysis(
  data = solder_data,
  target = "strength",
  factors = c("temperature", "pressure", "time"),
  explanation = list(
    partial_dependence = TRUE
  )
)

task_pdp$compute()
task_pdp$plot(type = "explanation")
```

---

## 6. Performance Optimization

### 6.1 Large Data Processing

```{r big-data, eval = FALSE}
# For large datasets, recommended approaches:
# 1. Reduce cross-validation folds
# 2. Use more efficient algorithms
# 3. Parallel computing

# Use parallel computing
library(future)
plan(multisession, workers = 4)

# Reduce tuning grid size
small_grid <- grid_regular(
  trees(range = c(100, 500)),
  min_n(range = c(5, 15)),
  levels = 3  # 3 values per parameter
)
```

### 6.2 Memory Optimization

```{r memory-optimization, eval = FALSE}
# Reduce saved intermediate results
task <- regression_analysis(
  data = large_data,
  target = "quality",
  expert_config = list(
    control = control_grid(
      verbose = FALSE,
      save_pred = FALSE,  # Do not save predictions
      extract = NULL      # Do not extract intermediate objects
    )
  )
)
```

---

## 7. Best Practices

### 7.1 Modeling Workflow Recommendations

1. **Start Simple**: First run through the full workflow in business mode
2. **Check Diagnostics**: Review model rating and diagnostic conclusions
3. **Optimize Gradually**: If rating < four stars, try:
   - Increase sample size
   - Add missing factors
   - Switch to more complex algorithms
4. **Verify Robustness**: Use multi-fold cross-validation to confirm results
5. **Interpret Results**: Use factor influence and SHAP values to understand the model

### 7.2 Common Pitfalls

| Pitfall | Symptoms | Solution |
|---------|---------|---------|
| **Overfitting** | High training R2, low test R2 | Use cross-validation, regularization |
| **Collinearity** | VIF > 10, unstable coefficients | Use ridge regression, PCA |
| **Insufficient Samples** | Unstable results | Collect more data |
| **Data Leakage** | Abnormally high accuracy | Check if features contain future information |

---

## 8. Reference Resources

### 8.1 tidymodels Ecosystem

- [tidymodels.org](https://www.tidymodels.org/) - Official documentation
- [Tidy Modeling with R](https://www.tmwr.org/) - Free e-book
- [parsnip](https://parsnip.tidymodels.org/) - Model interface
- [recipes](https://recipes.tidymodels.org/) - Feature engineering
- [tune](https://tune.tidymodels.org/) - Parameter tuning

### 8.2 mlr3 Ecosystem

- [mlr3 book](https://mlr3book.mlr-org.com/) - Official documentation
- [mlr3learners](https://mlr3learners.mlr-org.com/) - Algorithm list

---

*This vignette is written by the iQualityR Team | Version 0.1.0*
