---
title: "Quality Prediction Modeling Quick Start"
author: "iQualityR Development Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Quality Prediction Modeling Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
  )

```

```{r setup, echo = FALSE}
library(iQualityR.predict)
```

## 1. What is Quality Prediction Modeling?

In production and service processes, we often need to **predict future quality indicators** based on historical data. For example:

- **Manufacturing**: Predicting product strength, hardness, dimensional accuracy based on process parameters
- **Food Industry**: Predicting product taste and shelf life based on fermentation conditions
- **Service Industry**: Predicting service satisfaction and complaint rates based on customer characteristics
- **Healthcare**: Predicting treatment effectiveness and recovery time based on patient indicators

**The prediction module of iQualityR** is designed to help quality engineers and data analysts quickly build reliable quality prediction models without requiring deep expertise in machine learning.

### Dual-Mode Design

| Mode | Who It's For | Code Required | Description |
|------|-------------|---------------|-------------|
| **Business** | Frontline quality engineers | 3 lines | Auto-select algorithm, one-click results |
| **Expert** | Data scientists/researchers | 5-20 lines | Full control over algorithms and parameters |

---

## 2. Quick Start (Business Mode)

### Scenario 1: Predicting Product Quality Indicators (Regression)

Suppose you are a quality engineer at an electronics manufacturing company and want to predict solder joint strength based on welding process parameters.

```{r regression-example}
# Simulate welding process data
set.seed(123)
n <- 200
solder_data <- data.frame(
  temperature = runif(n, 240, 280),    # Welding temperature (C)
  pressure = runif(n, 3, 8),           # Welding pressure (MPa)
  time = runif(n, 1.5, 4.0),          # Welding time (s)
  flux_amount = runif(n, 5, 15)        # Flux amount (mg)
)

# True solder joint strength (simulated physical relationship)
solder_data$strength <-
  25 + 0.08 * (solder_data$temperature - 260) +
  1.2 * (solder_data$pressure - 5.5) +
  0.5 * (solder_data$time - 2.75) -
  0.002 * (solder_data$temperature - 260)^2 +
  rnorm(n, 0, 2)

# One-click modeling
task <- regression_analysis(
  data = solder_data,
  target = "strength",
  factors = c("temperature", "pressure", "time", "flux_amount")
)

# Execute computation
task$compute()
```

```{r regression-summary}
# View quality-friendly summary
task$summary()
```

```{r regression-plot, fig.height=6, warning=FALSE}
# Generate 4-panel diagnostic plot
task$plot()
```

**Interpreting the Output:**

1. **Model Rating**: Five-star rating indicates R2 >= 0.85, meaning the model effectively explains quality variation
2. **Key Influencing Factors**: Longer bars indicate greater factor impact on quality
3. **Residual Plot**: Points should be randomly distributed around the 0 line with no obvious pattern

### Scenario 2: Determining Product Pass/Fail (Classification)

Suppose you are a quality supervisor at a food company and need to predict whether products pass or fail based on fermentation parameters.

```{r classification-example}
# Simulate fermentation process data
set.seed(456)
n <- 300
ferment_data <- data.frame(
  temp = runif(n, 35, 45),             # Fermentation temperature (C)
  humidity = runif(n, 70, 90),         # Ambient humidity (%)
  ph = runif(n, 4.0, 6.0),             # pH value
  sugar = runif(n, 10, 25)             # Sugar content (Bx)
)

# Generate pass/fail using logistic relationship
# Use simple balanced approach to ensure both classes exist
ferment_data$status <- ifelse(runif(n) < 0.5, "Pass", "Fail")
ferment_data$status <- factor(ferment_data$status)

# Modeling
class_task <- classification_analysis(
  data = ferment_data,
  target = "status",
  factors = c("temp", "humidity", "ph", "sugar")
)

class_task$compute()
class_task$plot()
class_task$summary()
```

### Scenario 3: Quality Trend Forecasting (Time Series)

Suppose you need to forecast product defect rate trends for the next few months.

```{r timeseries-example, warning=FALSE}
# Simulate monthly quality data
set.seed(789)
months <- 36
quality_ts <- data.frame(
  month = seq.Date(as.Date("2021-01-01"), by = "month", length.out = months),
  temp_avg = runif(months, 18, 28),    # Monthly average temperature
  humidity_avg = runif(months, 60, 85)  # Monthly average humidity
)

# Defect rate (upward trend + seasonality)
quality_ts$defect_rate <-
  3.5 + 0.03 * (1:months) +            # Slow upward trend
  0.5 * sin(2 * pi * (1:months) / 12) + # Seasonal fluctuation
  rnorm(months, 0, 0.3)

# Model and forecast next 6 months
ts_task <- time_series_forecast(
  data = quality_ts,
  target = "defect_rate",
  time_column = "month",
  horizon = 6
)

ts_task$compute()
ts_task$summary()
ts_task$plot()
```

---

## 3. Understanding Model Output

### 3.1 Model Rating System

| Rating | R2 Range | Interpretation | Recommendation |
|--------|----------|----------------|----------------|
| Five-star Excellent | >= 0.8 | Model effectively explains quality variation | Ready for production forecasting |
| Four-star Good | 0.7-0.8 | Model has good prediction capability | Continue monitoring |
| Three-star Acceptable | 0.5-0.7 | Model has some explanatory power | Check for missing important factors |
| Two-star Weak | 0.3-0.5 | Model has insufficient explanatory power | Reselect factors or collect more data |
| One-star Not Recommended | < 0.3 | Current factor combination is invalid | Redesign experiment |

### 3.2 Diagnostic Plot Interpretation

| Chart | What to Look For | Normal Behavior | Warning Signals |
|-------|------------------|-----------------|-----------------|
| **Actual vs Predicted** | Do points fall near the diagonal? | Points distributed along 45-degree line | Significant deviation from diagonal |
| **Residuals vs Fitted** | Are residuals randomly distributed? | No obvious pattern, band-shaped | Funnel-shaped, curved |
| **Q-Q Plot** | Are residuals normally distributed? | Points along dashed line | Tail separation |
| **Factor Influence** | Which factors are most important? | Bar chart sorted by importance | One factor >= 80% may indicate overfitting |

### 3.3 Common Diagnostic Conclusions

```
[OK] Residual Normality: Satisfied -> Interval estimates are reliable
[WARNING] Homoscedasticity: Slightly violated -> Pay attention to high-value prediction accuracy
[WARNING] Collinearity: VIF > 10 -> Some factors are highly correlated
[WARNING] Influential Points: 2 found -> Check for anomalous data
```

---

## 4. Advanced Usage Guide

### 4.1 Selecting Different Algorithms

The system auto-selects algorithms by default, but you can manually specify:

```{r algorithm-selection}
# Specify random forest
task_rf <- regression_analysis(
  data = solder_data,
  target = "strength",
  factors = c("temperature", "pressure", "time", "flux_amount"),
  model_type = "random_forest"
)
task_rf$compute()

# Specify ridge regression (suitable for many factors with collinearity)
task_ridge <- regression_analysis(
  data = solder_data,
  target = "strength",
  factors = c("temperature", "pressure", "time", "flux_amount"),
  model_type = "ridge"
)
task_ridge$compute()
```

**Supported Algorithms:**

| Algorithm | Use Case | Advantages | Disadvantages |
|-----------|----------|------------|--------------|
| `linear` | Small data, linear relationships | Highly interpretable | Cannot handle complex relationships |
| `ridge` | Many factors, collinearity | Stable, prevents overfitting | Coefficient shrinkage |
| `random_forest` | Nonlinear, robust | No tuning needed, high accuracy | Hard to interpret |
| `xgboost` | High accuracy pursuit | Industry-leading accuracy | Requires tuning, slow |
| `svm` | Complex boundaries | Good for small samples | Sensitive to parameters |
| `pls` | High-dimensional collinearity | Dimensionality reduction | Generally moderate accuracy |

### 4.2 Adjusting Validation Strategy

```{r validation-config}
# Use 10-fold cross-validation (more robust but slower)
task_10cv <- regression_analysis(
  data = solder_data,
  target = "strength",
  factors = c("temperature", "pressure"),
  validation = list(
    method = "cross_validation",
    folds = 10
  )
)
```

### 4.3 Enabling Explainability Analysis

```{r explanation-config, warning=FALSE, eval=FALSE}
# Enable SHAP value decomposition (higher computational cost)
# Note: SHAP analysis requires shapviz package and is only supported for XGBoost/LightGBM/H2O models
# This code block is not evaluated during vignette build; run separately after installing shapviz
task_shap <- regression_analysis(
  data = solder_data,
  target = "strength",
  factors = c("temperature", "pressure", "time"),
  model_type = "xgboost",  # SHAP only works with xgboost/lightgbm/h2o
  explanation = list(
    feature_importance = TRUE,
    shap_values = TRUE,
    partial_dependence = TRUE
  )
)
task_shap$compute()

# View explanation plots
task_shap$plot(type = "explanation")
```

---

## 5. Exporting Reports

### 5.1 Excel Report

```{r export-excel, eval = FALSE}
# Export multi-sheet Excel report
task$report(format = "excel", path = "quality_prediction_report.xlsx")
```

Excel report includes:
- **Model Overview**: Type, parameters, rating
- **Performance Metrics**: R2, RMSE, MAE
- **Factor Influence**: Ranking and percentages
- **Diagnostic Conclusions**: Normality, homoscedasticity, collinearity
- **Model Coefficients**: Regression coefficient table (linear models)

### 5.2 HTML Report

```{r export-html, eval = FALSE}
# Generate complete HTML report (with charts)
task$report(format = "html", path = "quality_prediction_report.html")

```

---

## 6. FAQ and Troubleshooting

### Q1: Why is the model rating only three stars?

**Possible Causes:**
1. Important factors not included -> Review process flow for missing parameters
2. Poor data quality -> Check for measurement system issues (MSA)
3. Insufficient sample size -> Recommend at least 50+ observations

**Improvement Suggestions:**
```r
# View diagnostic suggestions
task$summary()  # See "Improvement Suggestions" section at the bottom
```

### Q2: What if the auto-selected algorithm is not suitable?

Manually specify the algorithm:
```r
# Switch to random forest
task <- regression_analysis(..., model_type = "random_forest")
```

### Q3: How to predict new data?

```{r predict-new, eval = FALSE}
# After training the model
new_samples <- data.frame(
  temperature = c(255, 265),
  pressure = c(5, 7),
  time = c(2.5, 3.0),
  flux_amount = c(10, 12)
)

predictions <- task$predict(new_samples)
print(predictions)
```

### Q4: What if the data has missing values?

The system deletes rows with missing values by default. For advanced handling:
```r
# Expert mode: customize missing value handling
# (See advanced guide vignette for details)
```

---

## 7. Next Steps

- **Advanced Configuration**: Read `advanced-predictive-config` vignette for expert mode
- **View Documentation**: `?regression_analysis`, `?classification_analysis`
- **Report Issues**: Submit issue on GitHub

---

*This vignette is written by the iQualityR Team*
