---
title: "DOE Experiment Design Report"
subtitle: "`r params$plan$design_type` design"
date: "`r params$report_date`"
output:
  html_document:
    theme: readable
    highlight: tango
    toc: true
    toc_depth: 3
    number_sections: true
params:
  results: NULL
  plan: NULL
  report_date: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  fig.align = "center"
)

results <- params$results
plan <- params$plan
pkg_version <- tryCatch(
  as.character(utils::packageVersion("iQualityR.doe")),
  error = function(e) "unknown"
)
```

---

## 1. Experiment Overview

### 1.1 Basic Information

| Item | Value |
|------|-------|
| **Design type** | `r plan$design_type` |
| **Number of factors** | `r length(plan$factors)` |
| **Replications** | `r plan$replication` |
| **Center points** | `r plan$center_points` |
| **Total runs** | `r if (!is.null(results$design_info)) nrow(results$design_info) else "N/A"` |
| **Resolution** | `r if (!is.null(plan$resolution)) plan$resolution else "N/A"` |
| **Randomized** | `r if (plan$randomize) "Yes" else "No"` |

### 1.2 Factor Configuration

```{r factor-config}
factor_df <- do.call(rbind, lapply(plan$factors, function(f) {
  data.frame(
    Factor = f$name,
    Type = if (f$type == "continuous") "Continuous" else "Categorical",
    N_Levels = length(f$levels),
    Levels = paste(f$levels, collapse = ", "),
    stringsAsFactors = FALSE
  )
}))

knitr::kable(factor_df, caption = "Factor configuration", align = "c")
```

---

## 2. Design Table

### 2.1 Experiment Plan

```{r design-table}
if (!is.null(results$design_info)) {
  design_display <- results$design_info
  # Show at most the first 20 rows to avoid an overly long table.
  if (nrow(design_display) > 20) {
    knitr::kable(head(design_display, 20), caption = "Design table (first 20 rows)")
    cat(sprintf("\n... %d runs in total", nrow(design_display)))
  } else {
    knitr::kable(design_display, caption = "Design table")
  }
}
```

---

## 3. Analysis of Variance

### 3.1 ANOVA Table

```{r anova-results}
if (!is.null(results$anova_results) && length(results$anova_results) > 0) {
  for (response_name in names(results$anova_results)) {
    cat(sprintf("\n### Response variable: %s\n\n", response_name))

    anova_res <- results$anova_results[[response_name]]

    if (!is.null(anova_res$anova)) {
      anova_df <- as.data.frame(anova_res$anova)
      anova_df$Term <- rownames(anova_df)
      anova_df <- anova_df[, c("Term", "Df", "Sum Sq", "Mean Sq", "F value", "Pr(>F)")]

      knitr::kable(anova_df,
                   caption = sprintf("ANOVA table for %s", response_name),
                   digits = 4, align = "c")

      # Significance assessment.
      significant <- anova_df[!is.na(anova_df$`Pr(>F)`) & anova_df$`Pr(>F)` < 0.05, ]
      if (nrow(significant) > 0) {
        cat(sprintf("\n**Significant factors (alpha = 0.05)**: %s\n",
                    paste(significant$Term, collapse = ", ")))
      } else {
        cat("\n**No significant factors** (alpha = 0.05)\n")
      }
    }
  }
} else {
  cat("No response data provided; analysis of variance skipped.")
}
```

### 3.2 Model Fit Statistics

```{r model-fit}
if (!is.null(results$anova_results) && length(results$anova_results) > 0) {
  for (response_name in names(results$anova_results)) {
    fit <- results$anova_results[[response_name]]$model_fit

    if (!is.null(fit)) {
      cat(sprintf("\n#### %s\n\n", response_name))

      fit_df <- data.frame(
        Statistic = c("R-squared", "Adjusted R-squared", "Lack-of-fit p-value"),
        Value = c(
          sprintf("%.4f", fit$r_squared),
          sprintf("%.4f", fit$adj_r_squared),
          if (is.na(fit$lack_of_fit_p)) "N/A" else sprintf("%.4f", fit$lack_of_fit_p)
        ),
        stringsAsFactors = FALSE
      )

      knitr::kable(fit_df, align = "c")
    }
  }
}
```

---

## 4. Effect Estimates

### 4.1 Main Effects

```{r main-effects}
if (!is.null(results$anova_results)) {
  for (response_name in names(results$anova_results)) {
    effects <- results$anova_results[[response_name]]$effects

    if (!is.null(effects$main) && length(effects$main) > 0) {
      cat(sprintf("\n#### %s\n\n", response_name))

      effects_df <- data.frame(
        Factor = names(effects$main),
        Effect = unlist(effects$main),
        stringsAsFactors = FALSE
      )

      # Sort by absolute effect size.
      effects_df <- effects_df[order(abs(effects_df$Effect), decreasing = TRUE), ]

      knitr::kable(effects_df, digits = 4, align = "c",
                   caption = sprintf("Main effect estimates for %s", response_name))
    }
  }
}
```

### 4.2 Interaction Effects

```{r interaction-effects}
if (!is.null(results$anova_results)) {
  for (response_name in names(results$anova_results)) {
    effects <- results$anova_results[[response_name]]$effects

    if (!is.null(effects$interaction) && length(effects$interaction) > 0) {
      cat(sprintf("\n#### %s\n\n", response_name))

      int_df <- data.frame(
        Interaction = names(effects$interaction),
        Effect = unlist(effects$interaction),
        stringsAsFactors = FALSE
      )

      int_df <- int_df[order(abs(int_df$Effect), decreasing = TRUE), ]

      knitr::kable(int_df, digits = 4, align = "c",
                   caption = sprintf("Interaction effect estimates for %s", response_name))
    }
  }
}
```

---

## 5. Graphical Analysis

### 5.1 Main Effects Plot

*The main effects plot shows the average influence of each factor on the response variable.*

```{r main-effects-plot, fig.width=9, fig.height=5}
if (!is.null(results$effects) && !is.null(results$effects$main) &&
    length(results$effects$main) > 0) {
  effects_data <- data.frame(
    Factor = names(results$effects$main),
    Effect = unlist(results$effects$main)
  )
  effects_data <- effects_data[order(abs(effects_data$Effect), decreasing = TRUE), ]
  effects_data$Factor <- factor(effects_data$Factor, levels = effects_data$Factor)

  p <- ggplot2::ggplot(effects_data, ggplot2::aes(x = Factor, y = Effect)) +
    ggplot2::geom_bar(stat = "identity", fill = "#2C7BB6") +
    ggplot2::geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    ggplot2::labs(
      title = "Main Effects Plot",
      x = "Factor",
      y = "Effect Estimate"
    ) +
    ggplot2::theme_minimal() +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))

  print(p)
} else {
  cat("No effect estimates available for plotting.")
}
```

### 5.2 Residual Diagnostic Plots

*Residual plots are used to validate model assumptions (normality, homoscedasticity, independence).*

```{r residual-plot, fig.width=9, fig.height=5}
if (!is.null(results$model)) {
  model <- results$model
  diag_df <- data.frame(
    Fitted    = fitted(model),
    Residuals = resid(model)
  )

  p_res <- ggplot2::ggplot(diag_df, ggplot2::aes(x = Fitted, y = Residuals)) +
    ggplot2::geom_point() +
    ggplot2::geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
    ggplot2::labs(
      title = "Residuals vs Fitted",
      x = "Fitted Values",
      y = "Residuals"
    ) +
    ggplot2::theme_minimal()

  print(p_res)
} else {
  cat("No fitted model available for residual diagnostics.")
}
```

---

## 6. Conclusions and Recommendations

### 6.1 Key Findings

```{r key-findings}
if (!is.null(results$anova_results) && length(results$anova_results) > 0) {
  for (response_name in names(results$anova_results)) {
    cat(sprintf("\n**%s**:\n\n", response_name))

    effects <- results$anova_results[[response_name]]$effects
    fit <- results$anova_results[[response_name]]$model_fit

    if (!is.null(effects$main)) {
      # Identify the most important factor.
      main_effects_sorted <- sort(abs(unlist(effects$main)), decreasing = TRUE)
      if (length(main_effects_sorted) > 0) {
        most_important <- names(main_effects_sorted)[1]
        cat(sprintf("- **Most critical factor**: %s (effect: %.4f)\n",
                    most_important, main_effects_sorted[1]))
      }
    }

    if (!is.null(fit)) {
      cat(sprintf("- **Model explanatory power**: R^2 = %.2f%%\n",
                  fit$r_squared * 100))
    }
  }
}
```

### 6.2 Optimization Recommendations

- **Factor screening**: For non-significant factors, fix them at any level to reduce cost.
- **Interactions**: For significant interactions, investigate further in subsequent experiments.
- **Response optimization**: Consider using response surface methodology (RSM) to further optimize the key factors.
- **Confirmation trials**: The optimal condition should be validated with 3-5 confirmation runs.

---

## 7. Appendix

### 7.1 Statistical Methods

**Analysis of Variance (ANOVA)**:
- Used to test the significance of each factor's main effect and interaction effects.
- F-test: $F_0 = MS_{\text{Effect}} / MS_{\text{Error}}$.
- Significance level: alpha = 0.05.

**R-squared**:
- Represents the proportion of total variation explained by the model.
- Adjusted R-squared applies a degrees-of-freedom correction.
- Values closer to 1 indicate a better fit.

### 7.2 References

- Montgomery, D. C. (2019). *Design and Analysis of Experiments* (10th ed.).
- Box, G. E. P., Hunter, W. G., & Hunter, J. S. (2005). *Statistics for Experimenters* (2nd ed.).

### 7.3 Report Generation Information

- **Generated on**: `r params$report_date`
- **Software version**: iQualityR.doe `r pkg_version`

---

*This report was generated automatically by iQualityR.doe.*
