---
title: "Single Sampling Plans and OC Curves"
author: "iQualityR Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Single Sampling Plans and OC Curves}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```

## 1. Introduction

Acceptance sampling is a fundamental technique in statistical quality control.
Rather than inspecting every unit in a lot, we inspect a random sample of size
`n` and decide whether to accept or reject the lot based on the number of
defectives `d` found. If `d <= c` (the acceptance number), the lot is accepted;
otherwise it is rejected.

This vignette walks through the **single sampling plan** workflow provided by
`iQualityR.sampling`, including:

- Building a sampling plan
- Computing the **Operating Characteristic (OC) curve**
- Evaluating **producer's risk** and **consumer's risk**
- Performing **power analysis**
- Visualizing the plan
- Exporting results to Excel

## 2. Theoretical background

### 2.1 Binomial model

For a lot with true fraction defective `p`, the number of defectives `D` in a
random sample of size `n` follows (approximately, for large lots) a binomial
distribution:

```
D ~ Binomial(n, p)
```

The probability of acceptance is therefore:

```
Pa(p) = P(D <= c) = sum_{k=0}^{c} C(n, k) * p^k * (1 - p)^(n - k)
        = F_Binom(c; n, p)
```

where `F_Binom` is the binomial CDF. In R this is `pbinom(c, n, p)`.

### 2.2 Key quality levels

- **AQL** (Acceptable Quality Level): the maximum fraction defective that is
  still considered acceptable. The producer wants Pa(AQL) to be high
  (typically >= 1 - alpha, with alpha = 0.05).
- **RQL** (Rejectable Quality Level, also called LTPD): the fraction
  defective that is clearly unacceptable. The consumer wants Pa(RQL) to be
  low (typically <= beta, with beta = 0.10).

### 2.3 Risks

- **Producer's risk** (`alpha`): the probability of rejecting a good lot
  (i.e., a lot at AQL). Computed as `1 - Pa(AQL)`.
- **Consumer's risk** (`beta`): the probability of accepting a bad lot
  (i.e., a lot at RQL). Computed as `Pa(RQL)`.

## 3. Loading the package

```{r setup}
library(iQualityR.sampling)
library(ggplot2)
```

## 4. Building a single sampling plan

We use `sampling_single()` to create a plan with `n = 50` and `c = 1`:

```{r build-plan}
task <- sampling_single(
  sample_size      = 50,
  acceptance_number = 1,
  aql              = 0.01,
  rql              = 0.05,
  alpha            = 0.05,
  beta             = 0.10
)
```

The returned object is an `IqrSamplingTask` that already contains the computed
results. We can inspect the plan configuration:

```{r plan-summary}
plan <- task$plan
cat(sprintf("n = %d, c = %d, AQL = %.3f, RQL = %.3f\n",
            plan$sample_size, plan$acceptance_number,
            plan$aql, plan$rql))
```

## 5. OC curve

The OC curve plots `Pa(p)` against `p`. It is stored in `task$results$oc_curve`:

```{r oc-curve-plot}
task$plot(type = "oc")
```

Key points on the OC curve:

```{r oc-key-points}
kp <- task$results$oc_curve$key_points
cat(sprintf("Pa(AQL = %.3f) = %.4f\n",
            kp$aql_point$p, kp$aql_point$prob))
cat(sprintf("Pa(RQL = %.3f) = %.4f\n",
            kp$rql_point$p, kp$rql_point$prob))
```

A good plan should have `Pa(AQL)` close to `1 - alpha = 0.95` and `Pa(RQL)`
close to `beta = 0.10`.

## 6. Risk analysis

```{r risk}
rk <- task$results$risk_analysis
cat(sprintf("Producer's risk (actual): %.4f (target alpha = %.4f)\n",
            rk$producer_risk, rk$risk_profile$target_alpha))
cat(sprintf("Consumer's risk (actual): %.4f (target beta  = %.4f)\n",
            rk$consumer_risk, rk$risk_profile$target_beta))
```

We can also visualize the risk profile as a bar chart, with the targets
shown as diamonds:

```{r risk-plot}
task$plot(type = "risk")
```

## 7. Power analysis

The **power** of a sampling plan at a given `p` is `1 - Pa(p)`, i.e., the
probability of correctly rejecting a lot with fraction defective `p`. At RQL,
the power should be at least `1 - beta = 0.90`.

```{r power}
pw <- task$results$power_analysis
cat(sprintf("Achieved power at RQL: %.4f\n", pw$achieved_power))
cat(sprintf("Required n for 80%% power at RQL: %s\n",
            if (is.null(pw$required_sample_size)) "N/A"
            else as.character(pw$required_sample_size)))
```

```{r power-plot}
task$plot(type = "power")
```

## 8. Full panel

A combined view of all available plots:

```{r full-panel, fig.width=9, fig.height=7}
task$plot(type = "full")
```

## 9. Console report

```{r console-report, results = "hold"}
task$summary()
```

## 10. Excel export

If the `openxlsx` package is installed, you can export all results to an Excel
workbook with separate sheets for the summary, OC curve, and power curve:

```{r excel-export, eval = FALSE}
tmp <- tempfile(fileext = ".xlsx")
task$report(path = tmp)
# file.path(tmp) is an .xlsx workbook
```

## 11. Choosing `n` and `c`

The trade-off in designing a sampling plan is between inspection cost (small
`n`) and protection (low `alpha` and `beta`). Larger `n` produces a steeper
OC curve, which means better discrimination between good and bad lots, at
the cost of more inspection.

| Plan | n  | c | Pa(AQL) | Pa(RQL) | Comment |
|------|----|---|---------|---------|---------|
| A    | 20 | 0 | high    | high    | Cheap but permissive |
| B    | 50 | 1 | high    | lower   | Balanced |
| C    | 125| 3 | ~0.95   | ~0.10   | Tight protection |

You can compare plans by building multiple `sampling_single()` tasks and
plotting their OC curves side by side.

## 12. Common pitfalls

1. **AQL >= RQL**: The plan requires AQL strictly less than RQL. Otherwise,
   the producer's and consumer's risk targets cannot both be met.
2. **c >= n**: An acceptance number larger than the sample size is
   nonsensical; a warning is issued.
3. **Small `n`, large `c`**: A permissive plan with `c` close to `n` accepts
   almost any lot, including bad ones. Always check `Pa(RQL)`.
4. **Ignoring lot size**: For small lots, the hypergeometric distribution
   is more accurate than the binomial approximation. The current
   implementation uses the binomial model (large-lot approximation).

## 13. References

- Schilling, E. G., & Neubauer, D. V. (2017). *Acceptance Sampling in Quality
  Control* (3rd ed.). CRC Press.
- Montgomery, D. C. (2020). *Introduction to Statistical Quality Control*
  (9th ed.). Wiley.
- ISO 2859-1:1999. *Sampling procedures for inspection by attributes -- Part 1*.

## 14. Session info

```{r session-info}
sessionInfo()
```
