---
title: "Double and Multiple Sampling Plans"
author: "iQualityR Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Double and Multiple Sampling Plans}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```

## 1. Introduction

A **single sampling plan** inspects one sample and makes a single accept/reject
decision. A **double sampling plan** offers a second chance: if the first
sample is inconclusive, a second sample is drawn and a combined decision is
made. **Multiple sampling plans** extend this idea to three or more stages.

The advantage of multi-stage plans is a lower **Average Sample Number (ASN)**:
when quality is very good or very bad, the decision is usually reached after
the first stage, saving inspection effort.

This vignette demonstrates:

- Building double and multiple sampling plans
- OC curves for multi-stage plans
- ASN curves
- Comparing single vs. double sampling
- Full workflow with `IqrSamplingTask`

## 2. Double sampling: theory

A double sampling plan is defined by two stages:

- Stage 1: sample `n1`, accept if `d1 <= c1`, reject if `d1 >= r1`,
  otherwise continue to stage 2.
- Stage 2: sample `n2`, accept if `d1 + d2 <= c2`, otherwise reject.

Here `r1` is the rejection number of stage 1, with `r1 > c1 + 1` typically.

The probability of acceptance is:

```
Pa(p) = P(D1 <= c1) + P(c1 < D1 < r1) * P(D1 + D2 <= c2 | c1 < D1 < r1)
```

Under the simplified independent-sample approximation, this becomes:

```
Pa(p) = F_Binom(c1; n1, p)
        + [F_Binom(r1-1; n1, p) - F_Binom(c1; n1, p)]
          * F_Binom(c2 - c1; n2, p)
```

The expected sample size (ASN) is:

```
ASN(p) = n1 + n2 * P(continue to stage 2)
       = n1 + n2 * [F_Binom(r1-1; n1, p) - F_Binom(c1; n1, p)]
```

## 3. Loading the package

```{r setup}
library(iQualityR.sampling)
library(ggplot2)
```

## 4. Building a double sampling plan

We use the `sampling_double()` convenience function. Each stage is a list
with `n` (sample size), `c` (acceptance number), and optionally `r`
(rejection number). If `r` is omitted, it defaults to `c + 1`.

```{r build-double}
task_d <- sampling_double(
  stage1 = list(n = 32, c = 0, r = 2),
  stage2 = list(n = 32, c = 1),
  aql    = 0.01,
  rql    = 0.10
)
```

## 5. OC curve for double sampling

```{r oc-double}
task_d$plot(type = "oc")
```

The OC curve shape is similar to single sampling but is the result of
combining two binomial probabilities.

## 6. ASN curve

The **Average Sample Number** curve shows how many units we expect to inspect
on average for each lot fraction defective `p`:

```{r asn-plot}
task_d$plot(type = "asn")
```

Key observations:

- When `p` is very small (good quality), we usually accept at stage 1, so
  ASN approaches `n1 = 32`.
- When `p` is very large (bad quality), we usually reject at stage 1, so
  ASN also approaches `n1 = 32`.
- For intermediate `p`, we often continue to stage 2, so ASN peaks near
  `n1 + n2 = 64`.

```{r asn-summary}
asn <- task_d$results$asn_curve
cat(sprintf("Single-stage n: %d\n", asn$single_sample_n))
cat(sprintf("Mean ASN over curve: %.2f\n", mean(asn$asn_values, na.rm = TRUE)))
cat(sprintf("Max ASN over curve: %.2f\n", max(asn$asn_values, na.rm = TRUE)))
```

## 7. Risk analysis

```{r risk-double}
rk <- task_d$results$risk_analysis
cat(sprintf("Producer's risk: %.4f\n", rk$producer_risk))
cat(sprintf("Consumer's risk: %.4f\n", rk$consumer_risk))
```

```{r risk-plot}
task_d$plot(type = "risk")
```

## 8. Multiple sampling plans

A **multiple sampling plan** extends double sampling to three or more
stages. The implementation in `iQualityR.sampling` uses a simplified
independence-based approximation:

```
Pa(p) = Pa_1(p) + (1 - Pa_1(p)) * Pa_2(p) + ...
```

This is an upper bound on the true acceptance probability; the exact
computation requires convolution over all possible cumulative paths.

```{r build-multiple}
task_m <- sampling_multiple(
  stages = list(
    list(n = 20, c = 0),
    list(n = 20, c = 1),
    list(n = 20, c = 2)
  ),
  aql = 0.01,
  rql = 0.10
)
```

```{r multiple-oc}
task_m$plot(type = "oc")
```

```{r multiple-asn}
task_m$plot(type = "asn")
```

> **Note**: The ASN curve for multiple sampling is computed with a simplified
> reach-probability approximation in the current implementation. For
> production use against ANSI/ASQ Z1.4 tables, please consult the standard
> directly.

## 9. Comparing single vs. double sampling

Let's build a single sampling plan with comparable discrimination and
compare OC and ASN curves:

```{r compare}
task_s <- sampling_single(
  sample_size       = 50,
  acceptance_number = 1,
  aql = 0.01,
  rql = 0.10
)

# Compare risk and ASN
cat(sprintf(
  "Single: producer's risk = %.4f, consumer's risk = %.4f\n",
  task_s$results$risk_analysis$producer_risk,
  task_s$results$risk_analysis$consumer_risk
))
cat(sprintf(
  "Double: producer's risk = %.4f, consumer's risk = %.4f\n",
  task_d$results$risk_analysis$producer_risk,
  task_d$results$risk_analysis$consumer_risk
))

# Single-stage n vs. mean ASN
cat(sprintf(
  "Single-stage n = %d; Double plan mean ASN = %.2f\n",
  task_s$plan$sample_size,
  mean(task_d$results$asn_curve$asn_values, na.rm = TRUE)
))
```

A well-designed double sampling plan can achieve similar protection (similar
OC curve) with a lower average sample number than a single plan.

## 10. Full panel

```{r full-panel, fig.width=9, fig.height=7}
task_d$plot(type = "full")
```

## 11. Console report

```{r console, results = "hold"}
task_d$summary()
```

## 12. Excel export

```{r excel, eval = FALSE}
tmp <- tempfile(fileext = ".xlsx")
task_d$report(path = tmp)
```

The workbook contains separate sheets for the Summary, OC Curve, Power Curve,
and ASN Curve (when applicable).

## 13. Designing your own plan

When designing a double sampling plan, consider:

1. **Stage 1 sample size `n1`**: typically half of an equivalent single
   plan's `n`.
2. **Acceptance number `c1`**: usually 0 or 1, to provide early acceptance
   for very good lots.
3. **Rejection number `r1`**: should satisfy `r1 > c1 + 1` to give the
   plan a "continue" zone.
4. **Stage 2 sample size `n2`**: typically equal to `n1`.
5. **Acceptance number `c2`**: should be larger than `c1` but not too large,
   to keep consumer's risk under control.

You can iterate on these values and use `sampling_double()` to evaluate
producer's risk, consumer's risk, and ASN.

## 14. Common pitfalls

- **Setting `r1 <= c1`**: This is invalid because stage 1 would have no
  "continue" zone. `SamplingPlan$validate()` will reject this.
- **Forgetting to specify `r`**: When `r` is omitted, the plan defaults to
  `c + 1`, which creates a minimal "continue" zone of just one value.
  This is often too narrow; consider `c + 2`.
- **Comparing plans only by OC curves**: Two plans with similar OC curves
  may have very different ASN. Always check ASN.

## 15. References

- Schilling, E. G., & Neubauer, D. V. (2017). *Acceptance Sampling in Quality
  Control* (3rd ed.). CRC Press.
- Montgomery, D. C. (2020). *Introduction to Statistical Quality Control*
  (9th ed.). Wiley.
- ANSI/ASQ Z1.4-2003. *American National Standard: Sampling Procedures and
  Tables for Inspection by Attributes*.

## 16. Session info

```{r session-info}
sessionInfo()
```
