---
title: "iQualityR: An Integrated Quality Engineering Framework"
author: "Chen Zhiming"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{iQualityR: An Integrated Quality Engineering Framework}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(iQualityR)
knitr::opts_chunk$set(fig.width = 6, fig.height = 4)
```

# Overview

`iQualityR` is a meta-package that installs and loads a coordinated set of R
packages for quality engineering. It follows the `tidyverse` convention: one
call to `library(iQualityR)` attaches all member packages so their APIs are
immediately available in the current session.

The framework is organized in four dependency layers:

| Layer | Package(s) | Responsibility |
|:------|:-----------|:---------------|
| Foundation | `iQualityR.core` | R6 abstract bases, themes, Excel export, i18n |
| Generic | `iQualityR.plot`, `iQualityR.doe`, `iQualityR.sampling`, `iQualityR.predict` | Visualization, design of experiments, acceptance sampling, predictive modeling |
| Statistical | `iQualityR.stat` | Distributions, SPC constants, hypothesis tests, normality, outliers |
| Business | `iQualityR.msa`, `iQualityR.capa`, `iQualityR.reliability` | Measurement system analysis, process capability, reliability/survival |

# Quick Start

```{r}
# Attach the entire framework
library(iQualityR)

# Inspect member packages
iQualityR_packages()
```

Once attached, every member API is callable without a namespace prefix:

```{r, eval=FALSE}
# Statistical foundations
iqr_prob$new(type = "norm", params = list(mean = 100, sd = 5))$calc(values = 108)

# Process capability
capability_normal(x = rnorm(100, 50, 1), lsl = 47, usl = 53)

# Measurement system analysis
iqr_msa(data = my_msa_data, part = "part", operator = "op", measurement = "y")

# Design of experiments
factorial_design(factors = list(A = c(-1, 1), B = c(-1, 1)))
```

# Member Packages

Use `iQualityR_packages()` to enumerate the currently attached members:

```{r}
iQualityR_packages()
```

To detach a specific member package without unloading the framework, use
`detach()` as usual.

# Design Rationale

The meta-package intentionally does **not** re-export member functions.
Keeping the public API in each member package preserves clear ownership of
documentation and behaviour, avoids name collisions, and ensures that
`iQualityR.msa::iqr_msa` and `iQualityR::iqr_msa` (after attach) refer to the
same object without duplication.

*End of Vignette*
```

