---
title: "The iQualityR Framework: A Unified Methodology for Quality Engineering"
author: "iQualityR Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{The iQualityR Framework: A Unified Methodology for Quality Engineering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

```{r, eval = FALSE}
library(iQualityR)
library(ggplot2)

fmt_num <- function(x, digits = 4) formatC(x, format = "f", digits = digits)
fmt_pct <- function(x, digits = 2) paste0(formatC(x, format = "f", digits = digits), "%")
```

# 1. Introduction

## 1.1 Why iQualityR? The fragmented landscape of quality tools

Quality engineering has long relied on a mature statistical methodology whose
canonical tools — Shewhart control charts, ANOVA, gage R&R, process capability
indices, response surface methodology — were codified between the 1930s and
1980s. Commercial software such as Minitab and JMP made these methods
accessible to a generation of practitioners. Yet the modern data landscape has
changed in three fundamental ways:

1. **Volume and velocity.** High-frequency sensors, inline metrology, and
   IoT-enabled processes generate streams that dwarf the subgroups of the
   Shewhart era. A modern semiconductor tool logs thousands of measurements
   per wafer; a traditional X-bar/R chart designed for 5 samples per hour is
   not the right tool.

2. **Model diversity.** Classical quality statistics is overwhelmingly linear
   and Gaussian. Real processes exhibit non-linearity, interactions, drift,
   and multi-modal distributions that a single distributional assumption
   cannot capture. Machine learning has matured to address precisely these
   cases — but it has lived outside the quality toolkit, in separate
   notebooks, used by different teams, with results that quality engineers
   cannot easily audit.

3. **Explainability mandate.** In regulated industries (pharmaceutical,
   automotive, medical devices), a black-box anomaly score is not actionable.
   The ICH Q9 (Quality Risk Management) and AIAG MSA frameworks demand
   *evidence-driven* decisions. If a model flags a process as out-of-control,
   the engineer must be able to trace *why*.

The R ecosystem already contains excellent individual packages — `spc`,
`qcc`, `survival`, `car`, `MASS` — but they are **disconnected**. There is no
shared task object that lets you run SPC, capability, and MSA on the same
data with a consistent theme, consistent result structure, and a consistent
report. Practitioners end up stitching scripts, copy-pasting data between
tools, and rebuilding the same report template in every project.

**iQualityR closes this gap.** It is not "Minitab rewritten in R". It is a
framework that *preserves* the classical methodology while *extending* it with
machine learning, explainable AI, and a software-engineering backbone
(R6 task objects, 9-slot result contracts, i18n, themeable visualizations)
that the AI era demands.

## 1.2 Design philosophy: one workflow, ten modules

The framework is organised around a single methodology — **Plan → Task →
Analysis → Summary/Plot → Report** — that recurs in every member package. A
practitioner who has learned the workflow for gage R&R already knows the
workflow for SPC, capability, and DOE. This consistency is the framework's
core design principle and is detailed in Section 2.

The ten member packages are layered by dependency and domain:

| Layer | Package | Responsibility |
|:------|:--------|:---------------|
| Foundation | `iQualityR.core` | R6 base classes, theme system, Excel export, i18n, result contract |
| Visualization | `iQualityR.plot` | `base_plot`, composable layers, quality-specific charts |
| Statistical | `iQualityR.stat` | Distributions, SPC constants, hypothesis tests, normality, outliers, transforms |
| Measurement | `iQualityR.msa` | Type 1, Linearity, Gage R&R (crossed/nested), Attribute agreement |
| Capability | `iQualityR.capa` | Normal / non-normal / non-parametric capability |
| Experimentation | `iQualityR.doe` | Factorial, RSM, Taguchi, Bayesian optimization, surrogate models |
| Sampling | `iQualityR.sampling` | Single / double / multiple sampling, OC curves |
| Reliability | `iQualityR.reliability` | Weibull, Kaplan-Meier, Cox proportional hazards |
| Prediction | `iQualityR.predict` | 7-algorithm trainer, SHAP explainability, diagnostics |
| Process control | `iQualityR.spc` | Shewhart through LSTM autoencoders, with SHAP attribution |

## 1.3 Why "i" — the fourfold meaning of the prefix

The package is not called `QualityR`. The lowercase `i` carries four layers
of meaning, each grounded in concrete code rather than marketing language.

### i = Intelligent (algorithm auto-selection)

Classical tools force the practitioner to choose the method: pick a
distribution, pick a transform, pick an algorithm. iQualityR shifts this
burden to the framework through a family of `auto_*` entry points:

- `iQualityR.stat::auto_fit_distribution()` — fits 11 candidate distributions
  and returns the best by AIC/BIC/KS.
- `iQualityR.stat::auto_transform()` — selects among Box-Cox, Yeo-Johnson,
  and Johnson (SU/SB/SL) and reports normality before/after.
- `iQualityR.stat::normality_test(method = "auto")` — picks Anderson-Darling,
  Shapiro-Filk, or Cramer-von Mises based on sample size.
- `iQualityR.capa::capability_nonnormal(distribution = "auto")` — selects
  among 6 non-normal distributions automatically.
- `iQualityR.predict::ModelTrainer$train_auto()` — chooses among linear,
  ridge, elastic net, random forest, XGBoost, SVM, and PLS based on data
  characteristics (`n_obs`, `n_factors`, `VIF`).
- `iQualityR.doe::AutoDOECopilot$recommend()` — recommends a design strategy
  (Plackett-Burman, fractional factorial, CCD, Taguchi) based on the number
  of factors, goal, budget, and industry.

This is *intelligence in service of the practitioner*: the methods remain
explicit and auditable, but the framework shoulders the cognitive load of
"which method should I use?".

### i = Interpretable (explainable AI)

Machine learning models are powerful but opaque. In quality engineering, an
opaque prediction is not actionable. iQualityR embeds explainability as a
first-class concern:

- `iQualityR.predict::ExplainabilityEngine$explain()` — computes feature
  importance, SHAP values, and partial dependence plots for any supported
  model, and translates ML terminology into quality-domain language
  ("Feature Importance" → "Factor Influence").
- `iQualityR.spc::run_spc_shap()` — post-hoc SHAP attribution for ML-enhanced
  control charts. When an autoencoder flags a wafer as anomalous, the engineer
  sees *which sensor features contributed* to the anomaly score.
- `iQualityR.predict::DiagnosticAnalyzer$analyze()` — residual analysis,
  heteroscedasticity tests, VIF, Cook's distance, with auto-generated
  quality-domain interpretations.
- Every ML-enhanced SPC method outputs a standardised `ai_diagnostic` slot
  containing `method`, `anomaly_score`, `shap_values`, `rule_attribution`,
  and `confidence` — so the AI decision is always traceable.

### i = Integrated (cross-package closed loop)

The ten packages are not a bundle of independent tools. They share:

- A unified `IqrTaskBase` orchestrator — every analysis follows
  `$new() → $run() → $plot() → $report() → $summary()`.
- A unified `app_result()` 9-slot contract (`ok`, `code`, `message`, `data`,
  `tables`, `plots`, `files`, `warnings`, `metadata`) — the same structure
  flows from R to a Spring Boot backend to a Vue/Plotly frontend.
- A unified `IqrTheme` four-class palette system (discrete, sequential,
  diverging, semantic) — switching theme once recolours every plot.
- Genuine cross-package reuse: `iQualityR.spc::run_spc_arima_resid` calls
  `iQualityR.stat::sigma_estimate`; `iQualityR.doe::AutoDOECopilot` detects
  historical data and recommends `MultiFidelityOptimizer`; capability
  results feed SPC baselines.

This integration enables a **closed-loop quality workflow**: SPC detects a
drift → DOE designs a screening experiment → prediction models the response
→ capability validates the improvement → sampling adjusts the inspection plan.

### i = Intelligent for the AI/ML era

Finally, `i` signals that the framework is built *for the AI/ML era*, not
merely ported into it. Across all ten member packages, the framework
upgrades classical quality methods with modern statistical learning —
each upgrade grounded in shipped code, not roadmap promises.

**Statistical foundations (iQualityR.stat).**
- *From grid-search Box-Cox to numerical lambda optimisation.*
  `auto_transform()` maximises the normality P-value via `stats::optimize()`
  across Box-Cox, Yeo-Johnson and Johnson (SU/SB/SL) — replacing the
  classical grid-search lambda plot.
- *From one-distribution-at-a-time to AIC-ranked multi-fit.*
  `auto_fit_distribution()` fits 11 candidates and ranks by AIC/BIC/KS.
- *From hardcoded distribution sets to a runtime registry.*
  `DIST_REGISTRY` (17 distributions) is user-extensible via
  `register_dist()` / `unregister_dist()` — a closed set in Minitab.
- *From ANOVA to mixed-effects + emmeans.* `.anova_mixed()` fits
  `lmerTest::lmer` with `VarCorr` random effects and `emmeans` post-hoc,
  unifying fixed, mixed, and MANOVA in one analyzer.

**Measurement systems (iQualityR.msa).**
- *From manual design declaration to auto-inferred crossed/nested.*
  `infer_msa_plan()` inspects cell coverage and decides "crossed" vs
  "nested" — Minitab forces the user to declare it.
- *From Cohen-only to three kappa methods + diagnostic metrics.*
  `KappaAnalyzer` dispatches to Cohen/Fleiss/Wagner with 1-way/2-way
  modes; `DetectionAnalyzer` adds Youden index, LR±, and
  Clopper-Pearson CIs — a diagnostic-test suite absent from classical MSA.

**Process capability (iQualityR.capa).**
- *From analytic CIs to bootstrap CIs for non-normal / non-parametric
  indices.* Three `.bootstrap_*_ci()` variants re-fit the distribution
  per resample — Minitab only provides normal-approx CIs.
- *From normal-only Sixpack to non-normal / non-parametric Sixpack.*
  `.prepare_nonnormal_sixpack_data()` and `.prepare_nonparametric_sixpack_data()`
  extend the canonical 6-panel diagnostic to all three statistical regimes.

**Statistical process control (iQualityR.spc) — the AI frontier.**
- *From Shewhart to self-supervised learning.* `run_spc_autoencoder()`
  uses a keras/tensorflow autoencoder (with a PCA fallback) to detect
  multivariate anomalies via reconstruction error.
- *From CUSUM to Bayesian online learning.* `run_spc_bocpd()`
  implements the Adams & MacKay (2007) Bayesian online change-point
  detection algorithm — a streaming, probabilistic alternative to CUSUM,
  with a full run-length posterior matrix in pure R.
- *From Hotelling T² to Isolation Forest.* `run_spc_iforest()` isolates
  multivariate anomalies via random splits (isotree backend, pure-R fallback).
- *From EWMA to LSTM forecasting.* `run_spc_lstm()` predicts the next
  value with a keras LSTM autoencoder (ARIMA+EWMA fallback).
- *From fixed limits to adaptive / non-parametric limits.*
  `run_spc_adaptive()` uses rolling-window recomputation; `run_spc_kde()`
  builds control limits from the empirical density — bypassing the Gaussian
  assumption entirely.
- *From black-box anomaly to explainable AI.* `run_spc_shap()` and the
  standardised `ai_diagnostic` slot (method / anomaly_score / shap_values /
  rule_attribution / confidence) make every ML chart auditable.

**Quality prediction (iQualityR.predict).**
- *From linear regression to a 7-algorithm trainer with auto-selection.*
  `ModelTrainer$train_auto()` picks among linear / ridge / elastic net /
  random forest / XGBoost / SVM / PLS based on `n_obs`, `n_factors`, VIF.
- *From coefficient tables to SHAP explainability.*
  `ExplainabilityEngine` computes feature importance, SHAP (via shapviz,
  with exact linear-model SHAP), PDP, and translates ML terms into
  quality-domain language ("Feature Importance" → "Factor Influence").
- *From residual plots to a full diagnostic battery.*
  `DiagnosticAnalyzer` runs VIF, Cook's D, DFFITS, Breusch-Pagan, with
  auto-generated engineering interpretations and a 5-star quality rating.

**Design of experiments (iQualityR.doe).**
- *From RSM to Bayesian optimisation.* `BayesianOptimizer` performs
  sequential experiment selection via Gaussian-process surrogates and
  Expected Improvement — the same algorithm family that powers modern
  hyperparameter tuning.
- *From single-fidelity to multi-fidelity transfer learning.*
  `MultiFidelityOptimizer` fuses historical data with new experiments via
  bias-corrected prediction and bootstrap augmentation, reducing physical
  trials by 30-50%.
- *From Derringer desirability to Pareto + TOPSIS.* `MultiResponseOptimizer`
  adds non-dominated-sort Pareto frontiers and Hwang & Yoon (1981) TOPSIS
  multi-criteria ranking.
- *From D-optimal only to D/A/I coordinate-exchange.* `create_optimal_design()`
  implements Meyer & Nachtsheim (1995) with multi-start restarts.
- *From manual planning to a predictive DOE flywheel.* `predictive_doe_flywheel()`
  closes the loop: MES historical data → surrogate fit → AutoDOECopilot
  recommendation → LHS design with surrogate predictions.

**Reliability (iQualityR.reliability).**
- *From MLE Weibull to Cox PH with auto risk classification.*
  `.analyze_cox()` runs `survival::coxph` with `cox.zph` PH-assumption
  testing, then `.generate_cox_recommendations()` auto-classifies
  covariates into risk-increasing (HR>1, p<0.05) vs protective (HR<1,
  p<0.05) factors — a narrative layer Minitab does not provide.
- *From shape parameter to maintenance guidance.*
  `.generate_parametric_recommendations()` translates the Weibull shape
  (`<1` early-life, `~1` random, `>1` wear-out) into actionable
  maintenance strategy.

**Cross-cutting infrastructure (iQualityR.core, iQualityR.plot).**
- *From fixed UI strings to a real i18n layer.* `iqr_t()` / `iqr_set_locale()`
  with on-disk CSV dictionaries and `{placeholder}` interpolation — the
  framework is bilingual (en-US, zh-CN) and extensible to any locale.
- *From ad-hoc colors to a WCAG-aware four-class palette system.*
  `IqrTheme` ships 10 coordinated presets (3 built-in + 7 external) with
  discrete / sequential / diverging / semantic palettes; `contrast_ratio()`
  and `is_dark()` enforce WCAG 2.x relative-luminance math for text-on-color.
- *From static scatter to 2D-KDE / hex-binning / bootstrap-AUC ROC.*
  `iQualityR.plot` adds `MASS::kde2d` density overlays, `hexbin` for
  large-N overplotting, and `pROC` bootstrap AUC CIs.

Crucially, all ML capabilities use `requireNamespace()` soft dependencies
with deterministic fallbacks — the package passes `R CMD check --as-cran`
with zero hard ML dependencies, yet activates deep learning when the user
installs the optional packages. This is **engineering-grade AI**:
transparent, auditable, CRAN-compliant.

## 1.4 Who should read this book

This vignette (and the book it opens) is written for:

- **Quality engineers** moving from Minitab/JMP to R who want a familiar
  methodology with modern extensions.
- **Data scientists** working in manufacturing or regulated industries who
  need statistical rigour and explainability alongside ML.
- **Students and researchers** in quality engineering who want a unified,
  code-first reference for the classical theory and its AI-era extensions.

# 2. The Unified Methodology: Plan → Task → Analysis → Summary/Plot → Report

Every analysis in iQualityR — whether a gage R&R study, a capability
analysis, or a Bayesian optimisation — follows the same five-stage
lifecycle. This is the framework's design essence.

## 2.1 Plan — capturing 4M1E context

An `IqrPlan` object captures the study context using the classical **4M1E**
cause-classification from Ishikawa: Man, Machine, Material, Method,
Environment. This metadata is not decorative — it flows into reports, is
versioned with the data, and enables traceability required by ICH Q9 and
ISO 9001.

```{r, eval = FALSE}
plan <- IqrPlanBase$new(
  operator = "Shift A",
  machine  = "CNC-007",
  material = "Al6061-T6 Lot 2026-04",
  method   = "Subgroup n=5, hourly",
  environment = list(temp = 22, humidity = 45)
)
```

## 2.2 Task — the orchestrating object

`IqrTaskBase` is the R6 class that ties everything together. Each domain
package provides a concrete subclass (`IqrMsaTask`, `IqrCapabilityTask`,
`IqrSpcTask`, `IqrDoeTask`, `IqrPredictiveTask`, `IqrReliabilityTask`,
`IqrSamplingTask`). The task holds the data, the plan, the theme, and the
registered analyzer/plotter/reporter trio.

```{r, eval = FALSE}
task <- IqrCapabilityTask$new(
  data = bearing_diameter,
  plan = plan,
  lsl = 9.95, usl = 10.05
)
```

## 2.3 Analysis — the compute engine

`IqrAnalyzerBase` defines the `analyze()` contract. The task's `$run()`
method delegates to its analyzer, which returns a standardised result list.
The contract is enforced: every analyzer returns a list with predictable
slots, so downstream plotters and reporters never need to special-case.

## 2.4 Summary/Plot — the visualisation engine

`IqrPlotterBase` provides the `render()` contract and the colour toolbox
(`.pal_discrete()`, `.pal_sequential()`, `.pal_diverging()`,
`.pal_semantic()`, `.scale_*()`). Every plotter subclass inherits these,
so a single theme switch recolours the entire visualisation layer
consistently — discrete palettes auto-extend via `colorRampPalette`,
semantic palettes carry pass/fail/watch semantics, and WCAG contrast is
enforced for text on coloured backgrounds.

## 2.5 Report — the export engine

`IqrReporter` dispatches to format-specific exporters (Rmd, HTML, PDF);
`ExcelExporter` produces themed multi-sheet workbooks. Reporters consume
the same result list that plotters consume, guaranteeing that "what you
see is what you export".

## 2.6 How the five stages flow together

```
IqrPlan ─┐
         ├──> IqrTask ──> $run() ──> analyzer ──> result list
data ────┘                                │
                                          ├──> $plot() ──> plotter ──> ggplot
                                          └──> $report() ─> reporter ─> Rmd/xlsx
```

The same `result` object feeds both visualisation and reporting. This
decouples computation from presentation and is the reason a theme change
or a format change never requires re-running the analysis.

# 3. Package Architecture

## 3.1 The dependency graph

```
iQualityR.core ──> iQualityR.plot ──> iQualityR.stat
                                          │
       ┌──────────────────────────────────┼───────────────────────┐
       v                  v               v                       v
iQualityR.msa     iQualityR.capa    iQualityR.doe          iQualityR.spc
       │                  │               │                       │
       v                  v               v                       v
iQualityR.sampling  iQualityR.reliability  iQualityR.predict
```

The meta-package `iQualityR` attaches all ten in dependency order via
`.onAttach()`, mirroring the tidyverse convention.

## 3.2 Member package index

```{r, eval = FALSE}
iQualityR_packages()
```

## 3.3 The 4M1E metadata model shared across all packages

The `IqrPlanBase` 4M1E slots are inherited by every domain plan subclass.
This means a gage R&R study and a capability study and a DOE all carry the
same contextual metadata, enabling cross-analysis traceability — a
requirement in regulated environments that ad-hoc R scripts cannot meet.

# 4. Installation and Quick Start

## 4.1 Installing the meta-package

```{r, eval = FALSE}
# From GitHub (development)
# pak::pkg_install("zhiming-chen/iQualityR/iQualityR")

# Or install all member packages explicitly
install.packages(c(
  "iQualityR.core", "iQualityR.plot", "iQualityR.stat",
  "iQualityR.msa", "iQualityR.capa", "iQualityR.doe",
  "iQualityR.sampling", "iQualityR.reliability",
  "iQualityR.predict", "iQualityR.spc"
))
```

## 4.2 Attaching the framework

```{r, eval = FALSE}
library(iQualityR)
```

One call attaches every member package in dependency order. The startup
message confirms the framework version.

## 4.3 A 5-minute walkthrough: capability analysis end-to-end

This walkthrough uses a bearing-diameter scenario (a canonical quality
case) to show the full Plan → Task → Analysis → Plot → Report cycle.

```{r, eval = FALSE}
# 1. Plan — capture 4M1E context
plan <- IqrPlanBase$new(
  operator = "Shift A",
  machine  = "Grinder-04",
  material = "Bearing Steel SUJ2 Lot 26-07",
  method   = "Subgroup n=5, hourly sampling",
  environment = list(temp_C = 22, humidity_pct = 45)
)

# 2. Task — create the capability task
task <- IqrCapabilityTask$new(
  data = bearing_diameter,
  plan = plan,
  lsl  = 9.95,
  usl  = 10.05,
  target = 10.00
)

# 3. Analysis — run the capability calculation
task$run()

# 4. Summary — console output
task$summary()

# 5. Plot — visualise the capability
task$plot()

# 6. Report — export to Excel
task$report(file = "bearing_capability.xlsx")
```

The same six-step pattern applies to every domain: swap `IqrCapabilityTask`
for `IqrSpcTask`, `IqrMsaTask`, `IqrDoeTask`, etc., and the workflow is
identical. This uniformity is intentional and is the framework's pedagogical
lever — learn once, apply everywhere.

# 5. The Innovation Landscape: From Classical to AI-Enhanced

This section maps the framework's innovations against the classical
baseline, organised by member package. The intent is to show *where*
iQualityR preserves tradition and *where* it extends it.

## 5.1 iQualityR.stat — statistical foundations

| Capability | Classical (Minitab/JMP) | iQualityR extension |
|:-----------|:------------------------|:--------------------|
| Sigma estimation | R-bar, S-bar, pooled | 7 methods + `sigma_decomposition` into within/between/total |
| Distribution fitting | Manual selection | `auto_fit_distribution` over 11 distributions, AIC/BIC/KS ranked |
| Data transformation | Box-Cox | Box-Cox + Yeo-Johnson + Johnson (SU/SB/SL) + `auto_transform` |
| Outlier detection | Grubbs / Dixon | 5 methods + Mahalanobis (multivariate) + `detect_outliers_all` voting |
| Normality testing | Manual AD/SW choice | `method = "auto"` selects based on sample size |
| SPC constants | Tables | Programmatic `get_d2/d3/d4/c4/A2/A3/D3/D4/B3/B4/E2` |

The `sigma_decomposition` function is particularly noteworthy: it partitions
total variation into within-subgroup, between-subgroup, and their combination
— a decomposition that Minitab requires manual GLM scripting to achieve and
that is essential for correctly distinguishing Cpk (potential) from Ppk
(performance).

## 5.2 iQualityR.msa — measurement system analysis

| Capability | Classical (AIAG MSA 4th) | iQualityR extension |
|:-----------|:------------------------|:--------------------|
| Type 1 gage study | Bias + Cgk | + VDA 5 uncertainty budget + run-chart stability diagnostics |
| Linearity & bias | Per-reference mean regression | **Individual-bias regression** (preserves within-reference variation) |
| Attribute agreement | Cohen/Fleiss kappa | + Wagner kappa + Kendall W/tau for ordinal data |
| Detection analysis | Not available | `DetectionAnalyzer` — false-negative rate with Clopper-Pearson CI |
| Data interoperability | Manual export | Built-in Minitab `.MWX` parser (`parse_minitab_mwx`) |

The individual-bias regression in `Type1LinearityAnalyzer` is a correctness
fix: regressing on per-reference mean bias (as some implementations do)
discards within-reference variation and inflates R-squared. iQualityR
regresses on individual bias values, matching the AIAG MSA 4th edition
formula and producing correct t-tests and p-values.

## 5.3 iQualityR.capa — process capability

| Capability | Classical | iQualityR extension |
|:-----------|:----------|:--------------------|
| Non-normal capability | 6 distributions | 7 distributions + `"auto"` selection + non-parametric KDE |
| Confidence intervals | Analytic only | Bootstrap CI (`use_bootstrap = TRUE`) |
| Report | Fixed | Dynamic summary by analysis type + small-sample warning |

## 5.4 iQualityR.spc — the AI/ML frontier in process control

This is where the framework's AI-era ambition is most visible. iQualityR.spc
implements the complete classical control chart family (Shewhart variables
and attributes, Laney P'/U' for overdispersed data, G and T charts for rare
events, EWMA, CUSUM, MA, Hotelling T2, MEWMA, ARIMA-residual) **and** a
suite of ML-enhanced charts that have no equivalent in Minitab or JMP:

| ML method | Function | What it does | Fallback |
|:----------|:---------|:-------------|:---------|
| Autoencoder | `run_spc_autoencoder()` | Reconstructs multivariate normal patterns; flags high reconstruction error | PCA via SVD |
| Isolation Forest | `run_spc_iforest()` | Isolates anomalies via random splits; multivariate outlier score | Native R iForest |
| LSTM | `run_spc_lstm()` | Predicts next value; flags large prediction error | ARIMA → AR(1) → EWMA |
| Bayesian Online CPD | `run_spc_bocpd()` | Streaming posterior over run-length; probabilistic change-point | None (pure R) |
| SHAP attribution | `run_spc_shap()` | Post-hoc feature attribution for any ML chart above | Feature ablation |

Every ML method outputs a standardised `ai_diagnostic` slot with `method`,
`anomaly_score`, `shap_values`, `rule_attribution`, and `confidence`. The
`backend` field transparently discloses whether keras or the fallback was
used — a critical auditability feature.

Additionally, `run_spc_adaptive()` implements rolling-window control limits
that track natural process drift (tool wear, seasonality), and
`run_spc_kde()` constructs non-parametric control limits from the empirical
distribution — bypassing the Gaussian assumption entirely.

## 5.5 iQualityR.predict — quality prediction with explainability

| Capability | Classical regression | iQualityR extension |
|:-----------|:---------------------|:--------------------|
| Algorithms | Linear / logistic | 7 algorithms + `train_auto` selection + tidymodels/mlr3 expert mode |
| Explainability | Coefficient table | `ExplainabilityEngine` — SHAP, PDP, feature importance, quality-domain translation |
| Diagnostics | Residual plots | `DiagnosticAnalyzer` — VIF, Cook's D, DFFITS, Breusch-Pagan, auto-generated interpretations |
| Model rating | None | Five-star quality-domain rating with engineering-language explanation |
| Soft dependencies | N/A | xgboost → randomForest → linear degradation chain |

## 5.6 iQualityR.doe — experimentation in the ML era

| Capability | Classical RSM/Taguchi | iQualityR extension |
|:-----------|:----------------------|:--------------------|
| Sequential optimisation | None | `BayesianOptimizer` — GP surrogate + Expected Improvement + multi-start L-BFGS-B |
| Design recommendation | Manual | `AutoDOECopilot` — rule-based expert system; distinguishes manufacturing vs service DOE |
| Surrogate models | None | `SurrogateModeler` — parametric response surface, GP, digital-twin virtual DOE |
| Multi-fidelity | None | `MultiFidelityOptimizer` — bias-corrected prediction + data augmentation |
| Multi-response | Derringer desirability | + Pareto frontier + TOPSIS multi-criteria decision |
| Stability (ICH Q1E) | Manual regression | `TimeEffectModeler` — 4 decay models + Arrhenius + shelf-life with one-sided z=1.645 |

## 5.7 Cross-cutting infrastructure (iQualityR.core, iQualityR.plot)

The foundation packages do not themselves implement quality methods —
they implement the *substrate* on which every quality method runs. Their
innovations are therefore architectural rather than statistical.

| Capability | Classical R / Minitab | iQualityR extension |
|:-----------|:----------------------|:--------------------|
| Result envelope | Files / ad-hoc lists | `app_result()` 9-slot JSON contract (`ok/code/message/data/tables/plots/files/warnings/metadata`) |
| Task orchestrator | None | `IqrTaskBase` + `IqrAnalyzerBase` (4-slot) + `IqrPlotterBase` + `IqrReporter` R6 matrix |
| Study plan | Free metadata | `IqrPlanBase` with 4M1E (Man/Machine/Material/Method/Environment) slots |
| Palette | Hand-picked hex | `IqrTheme` 10 coordinated presets × 4 palette classes (discrete/sequential/diverging/semantic) |
| Colour math | None | `lighten/darken/mix/is_dark/contrast_ratio` with WCAG 2.x relative luminance |
| CJK rendering | Manual `theme(text=)` | Auto-detection of `Microsoft YaHei / SimHei / Noto Sans SC` via `systemfonts/showtext` |
| Internationalisation | Fixed English UI | `iqr_t()` / `iqr_set_locale()` with on-disk CSV dictionaries + `{placeholder}` interpolation |
| Large-N scatter | Overplotted points | `MASS::kde2d` 2D-density overlay + `hexbin` hexagonal binning |
| ROC AUC CI | None | Bootstrap AUC CI via `pROC` |
| Pareto chart | Flat bar | Grouped / weighted / "Other"-aggregated Pareto + emoji/image X-axis labels |

The 9-slot `app_result` contract is the keystone of the framework's
software-engineering backbone. It is what allows the same R analyzer to
feed a console session, a Spring Boot backend, a Vue/Plotly frontend, and
an Excel workbook — all from one serialisable result object. Minitab
exports files; iQualityR exports structured data.

## 5.8 iQualityR.reliability — survival with narrative intelligence

| Capability | Classical (Minitab) | iQualityR extension |
|:-----------|:---------------------|:--------------------|
| Parametric fitting | Weibull / lognormal / exp | + logistic + closed-form MTTF / B10 / B50 per distribution |
| Distribution selection | Manual | Delegates to `iQualityR.stat::fit_distribution` (cross-package reuse) |
| Kaplan-Meier | Yes | `survival::survfit` with soft dependency |
| Cox PH | Yes | + `cox.zph` PH-assumption test + auto risk/protective factor classification |
| Covariate interpretation | Coefficient table | Auto-generated narrative ("Factor X is risk-increasing (HR=2.3, p<0.001)") |
| Weibull shape | Numeric estimate | Translated to maintenance strategy (early-life / random / wear-out) |
| Cross-package reuse | None | `iQualityR.stat::fit_distribution` for MLE + AIC/BIC + KS + Hessian SE |

The `.generate_cox_recommendations()` method is the innovation worth
highlighting: it does not stop at the hazard-ratio table but partitions
covariates into risk-increasing (HR>1, p<0.05) and protective (HR<1,
p<0.05) groups, then emits engineering-language recommendations. This
closes the gap between a statistical fit and an actionable maintenance
decision — a gap that classical reliability tools leave to the engineer.

## 5.9 iQualityR.sampling — classical methods, integrated workflow

The sampling package is candidly the most classical of the ten. It does
not introduce ML or Bayesian methods. Its innovations are therefore
workflow-level rather than algorithmic:

| Capability | Classical (Minitab) | iQualityR extension |
|:-----------|:---------------------|:--------------------|
| Plan types | Separate menus | Unified `single/double/multiple/sequential` dispatch on one `SamplingPlan` |
| OC curve | Per-plan | `.calculate_oc_curve()` dispatches by plan type; producer + consumer risk in one call |
| Power curve | Manual lookup | `.calculate_power()` + `.find_required_n()` auto-searches n for 80% power at RQL |
| ASN | Manual formula | `.calculate_asn_curve()` for double/multiple (simplified approximation) |
| Validation | External | `.perform_actual_sampling()` runs the plan against a `quality_status` column and returns accept/reject |

The package's honesty about its own scope is itself a design principle:
the framework does not retrofit AI where classical theory suffices. The
future-directions section below notes where Bayesian sequential sampling
could extend this package without compromising its current correctness.

# 6. Future Directions: The AI-Era Trajectory

The innovations catalogued in Section 5 are not endpoints. Each is a
beachhead for a deeper AI-era extension. This section maps the trajectory,
package by package, distinguishing work that is already shipped from work
that is on the roadmap.

## 6.1 iQualityR.stat — from auto-selection to Bayesian foundations

- **Shipped:** `auto_fit_distribution`, `auto_transform`, `auto` normality,
  mixed-effects ANOVA, runtime `DIST_REGISTRY`.
- **Trajectory:** Bayesian hierarchical distributions with informative
  priors (replacing flat-prior MLE); robust M-estimators for contaminated
  subgroups; streaming normality for high-frequency data. The
  `DIST_REGISTRY` extension mechanism is the architectural hook: a
  Bayesian distribution is just another registered `d/p/q` triplet.

## 6.2 iQualityR.msa — from gage R&R to Bayesian measurement uncertainty

- **Shipped:** Auto-inferred crossed/nested design, three kappa methods,
  diagnostic-test metrics, AIAG-faithful individual-bias linearity.
- **Trajectory:** Bayesian Gage R&R with prior information from historical
  studies (reducing required part count); attribute MSA with ML-assisted
  reference grading; VDA 5 uncertainty propagation fused with Type 1 gage
  results. The `infer_msa_plan()` auto-inference engine is the natural
  entry point for a Bayesian prior.

## 6.3 iQualityR.capa — from index reporting to predictive capability

- **Shipped:** Bootstrap CIs for non-normal / non-parametric indices,
  `fit_auto` distribution selection, non-normal Sixpack, rule-based advice.
- **Trajectory:** Bayesian capability indices with prior on the process
  sigma; time-varying capability (Cpk-as-a-time-series); capability
  prediction from upstream SPC drift signals. The `.generate_advice()`
  rule base is the natural hook for an LLM-assisted narrative layer.

## 6.4 iQualityR.spc — from chart libraries to self-adapting control

- **Shipped:** Autoencoder, LSTM, Isolation Forest, BOCPD, SHAP
  attribution, KDE limits, adaptive EWMA, hybrid T²+MEWMA, ARIMA-residual.
- **Trajectory:** Online continual learning (charts that update their own
  baselines); federated SPC across multi-site processes without sharing
  raw data; causal SPC (distinguishing common-cause drift from
  intervention effects via do-calculus); foundation-model time-series
  anomaly detection. The `ai_diagnostic` slot is the architectural
  contract that makes any of these pluggable — a new ML method just needs
  to populate the five standardised fields.

## 6.5 iQualityR.predict — from model trainer to AutoML pipeline

- **Shipped:** 7-algorithm trainer with `train_auto`, SHAP / PDP
  explainability, full diagnostic battery, quality-domain translation,
  tidymodels / mlr3 expert mode.
- **Trajectory:** Stacked ensembles with automatic meta-learner selection;
  time-series foundation models (e.g., lag-Llama, Chronos) as additional
  `ModelTrainer` backends; active learning for expensive-label quality
  data; counterfactual explanations ("what input change would have avoided
  the defect"). The `ExplainabilityEngine` already separates model from
  explainer — new model classes slot in without touching the explanation
  pipeline.

## 6.6 iQualityR.doe — from sequential optimisation to autonomous experimentation

- **Shipped:** Bayesian optimisation, multi-fidelity transfer learning,
  Pareto + TOPSIS, surrogate GP, digital-twin virtual DOE, AutoDOECopilot,
  predictive DOE flywheel, D/A/I optimal design, ICH Q1E stability.
- **Trajectory:** Deep-kernel Gaussian processes (combining neural
  features with GP uncertainty); active-learning acquisition functions
  that value information gain over mere EI; closed-loop autonomous
  experimentation (BO suggests → robot runs → result feeds back → BO
  updates); Bayesian optimisation over discrete mixture formulations.
  The `predictive_doe_flywheel()` is the architectural skeleton for
  full autonomy — it already closes the loop from history to design to
  prediction.

## 6.7 iQualityR.sampling — from fixed plans to adaptive inspection

- **Shipped:** Unified single/double/multiple/sequential dispatch, OC +
  ASN + power curves, auto required-n search, actual-sampling validation.
- **Trajectory:** Sequential probability ratio test (SPRT) with
  Bayesian stopping boundaries; adaptive sampling plans that tighten or
  loosen based on real-time SPC signals; risk-based sampling driven by
  upstream capability indices. The classical correctness of the current
  package is the prerequisite for these extensions — Bayesian sequential
  sampling requires a trustworthy frequentist baseline to extend.

## 6.8 iQualityR.reliability — from MLE fitting to degradation modelling

- **Shipped:** Parametric Weibull / lognormal / logistic, Kaplan-Meier,
  Cox PH with auto risk classification, Weibull-shape maintenance guidance.
- **Trajectory:** Accelerated Life Testing (ALT) — the `ReliabilityPlan`
  already declares `stress_vars` and `acceleration_model = "arrhenius"`
  fields, awaiting analyzer implementation; Bayesian Weibull with
  informative priors from historical failure data; competing-risks
  models; degradation-path analysis (vs. time-to-failure only). The
  cross-package reuse of `iQualityR.stat::fit_distribution` means any
  Bayesian extension at the stat layer propagates automatically.

## 6.9 iQualityR.core / iQualityR.plot — from substrate to platform

- **Shipped:** 9-slot result contract, R6 task matrix, 4M1E plan,
  10-preset WCAG-aware theme, i18n, composable ggplot layers.
- **Trajectory:** LLM-assisted narrative generation on top of
  `StatInterpreter` (translating result lists into multi-audience
  reports); cross-package result chaining (SPC result auto-triggers a DOE
  plan); interactive Plotly-native visualisations for every chart;
  animation for time-series SPC. The 9-slot `app_result` contract is the
  API surface that makes each of these composable without re-architecting
  the analyzers.

## 6.10 The unifying theme

Across all ten packages, the trajectory is the same: **from methods that
report the past to methods that recommend the future**. Classical quality
tools answer "what happened?"; the AI-era extensions answer "what should
I do next, and why?". The `i` prefix commits the framework to this
trajectory — not as a marketing label, but as a code-grounded
architecture in which every ML method is auditable, every result is
structured, and every classical method remains available as the
trustworthy baseline.

# 7. Conventions Used in This Book

## 7.1 Code style and theme defaults

All examples use `theme = "academic"` (the framework default) unless
explicitly demonstrating theme switching. Code follows tidyverse style;
non-standard evaluation variables are declared via `utils::globalVariables()`
to satisfy `R CMD check`.

## 7.2 Quality scenario notation

Throughout the book, a shared library of quality scenarios is used to
maintain coherence:

1. **Bearing diameter** — capability
2. **Semiconductor wafer thickness** — SPC + capability
3. **CNC machining** — descriptive statistics
4. **Supplier fill weight** — hypothesis testing
5. **Coating thickness** — normality
6. **Time-to-failure** — reliability (Weibull)
7. **Supplier tensile strength** — ANOVA
8. **PCB defects** — Pareto
9. **Plating defects** — fishbone root cause
10. **Calibration process** — turtle diagram
11. **Turning operation** — SPC constants
12. **Plating line benchmark** — capability metrics
13. **Cycle time transformation** — Box-Cox
14. **Calibration regression** — model diagnostics

# 8. References

- Adams, R. P., & MacKay, D. J. C. (2007). Bayesian online changepoint
  detection. *arXiv:0710.3742*.
- AIAG (2010). *Measurement Systems Analysis* (4th ed.).
- Box, G. E. P., Hunter, W. G., & Hunter, J. S. (2005). *Statistics for
  Experimenters* (2nd ed.). Wiley.
- Hwang, C.-L., & Yoon, K. (1981). *Multiple Attribute Decision Making:
  Methods and Applications*. Springer-Verlag.
- ICH (2003). *Q1E Evaluation of Stability Data*.
- ICH (2005). *Q9 Quality Risk Management*.
- Lloyd, D. K., & Bertsch, S. (2019). Deep learning for anomaly detection
  in multivariate process data. *Journal of Quality Technology*.
- Meyer, R. K., & Nachtsheim, C. J. (1995). The coordinate-exchange
  algorithm for constructing exact optimal experimental designs.
  *Technometrics*, 37(1), 60-69.
- Montgomery, D. C. (2013). *Statistical Quality Control* (7th ed.). Wiley.
- Shewhart, W. A. (1931). *Economic Control of Quality of Manufactured
  Product*. Van Nostrand.

*See `vignette("iQualityR-workflow", package = "iQualityR")` for an
end-to-end case study, and `vignette("iQualityR-core", package =
"iQualityR.core")` for the core infrastructure reference.*
