---
title: "Process Capability Analysis Report"
subtitle: "`r params$analysis_label %||% 'Capability Study'`"
date: "`r params$timestamp %||% Sys.time()`"
output:
  html_document:
    theme: readable
    highlight: tango
    toc: true
    toc_depth: 3
    number_sections: true
params:
  results: NULL
  plan: NULL
  theme_obj: NULL
  timestamp: NULL
  statistics: NULL
  diagnostics: NULL
  data_tables: NULL
  analysis_label: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 6, fig.align = "center"
)

# Local helpers (kept self-contained so the template does not depend on rlang
# being attached at render time).
`%||%` <- function(x, y) if (is.null(x)) y else x
fmt_num <- function(x, digits = 4) {
  if (is.null(x) || length(x) == 0 || all(is.na(x))) return("NA")
  sprintf(paste0("%.", digits, "f"), x)
}
fmt_sci <- function(x) {
  if (is.null(x) || length(x) == 0 || is.na(x)) return("NA")
  if (x < 0.001 && x > 0) formatC(x, format = "e", digits = 2) else sprintf("%.4f", x)
}
round_df <- function(x, digits = 4) {
  if (is.null(x) || !is.data.frame(x) || nrow(x) == 0) return(x)
  for (nm in names(x)) if (is.numeric(x[[nm]])) x[[nm]] <- round(x[[nm]], digits)
  x
}

results    <- params$results
plan       <- params$plan
theme_obj  <- params$theme_obj
stats      <- params$statistics  %||% results$statistics
diag       <- params$diagnostics %||% results$diagnostics
data_tbls  <- params$data_tables %||% results$data_tables

# Detect analysis type from plan or diagnostics.
analysis_type <- if (!is.null(plan) && !is.null(plan$analysis_type)) {
  plan$analysis_type
} else if (!is.null(diag$distribution)) {
  "nonnormal"
} else if (!is.null(diag$method) && diag$method %in% c("kernel", "empirical")) {
  "nonparametric"
} else {
  "normal"
}
analysis_label <- switch(analysis_type,
  normal       = "Normal Capability Analysis",
  nonnormal    = "Non-Normal Capability Analysis",
  nonparametric= "Non-Parametric Capability Analysis",
  "Capability Analysis"
)
```

# Study Overview

| Item | Value |
|---|---|
| Analysis type | `r analysis_label` |
| Specification LSL | `r fmt_num(plan$lsl)` |
| Specification USL | `r fmt_num(plan$usl)` |
| Target | `r fmt_num(plan$target %||% NA)` |
| Sample size (n) | `r stats$n` |
| Process mean | `r fmt_num(stats$mean)` |
| StdDev (Within)  | `r if (is.null(stats$sd_within)) "N/A" else fmt_num(stats$sd_within)` |
| StdDev (Overall) | `r fmt_num(stats$sd_overall)` |
| Confidence level | `r fmt_num(plan$conf_level)` |

```{r overview-note, results='asis'}
if (analysis_type == "nonnormal") {
  cat(sprintf(
    "Non-normal analysis fits a parametric distribution (**%s**) to the data and computes capability from the 0.135%% / 99.865%% quantiles (equivalent to $\\pm 3\\sigma$ for a normal distribution).\n",
    diag$distribution %||% "auto-selected"
  ))
} else if (analysis_type == "nonparametric") {
  cat(sprintf(
    "Non-parametric analysis uses the empirical distribution (method: **%s\"). Capability indices are based on empirical quantiles; no distributional assumption is made.\n",
    diag$method %||% "empirical"
  ))
} else {
  cat("Normal analysis assumes the data follow a normal distribution. Within-subgroup sigma is used for Cp/Cpk; overall sigma is used for Pp/Ppk.\n")
}
```

# Capability Indices

```{r indices-table}
idx_rows <- list(
  Index = c("Cp", "Cpk", "Cpl", "Cpu"),
  Value = c(stats$cp, stats$cpk, stats$cpl, stats$cpu)
)
if (analysis_type == "normal") {
  idx_rows$Index <- c(idx_rows$Index, "Pp", "Ppk", "Ppl", "Ppu")
  idx_rows$Value <- c(idx_rows$Value, stats$pp, stats$ppk, stats$ppl, stats$ppu)
} else {
  idx_rows$Index <- c(idx_rows$Index, "Pp", "Ppk", "Ppl", "Ppu")
  idx_rows$Value <- c(idx_rows$Value, stats$pp, stats$ppk, stats$ppl, stats$ppu)
}
if (!is.null(stats$cpm)) {
  idx_rows$Index <- c(idx_rows$Index, "Cpm (Taguchi)")
  idx_rows$Value <- c(idx_rows$Value, stats$cpm)
}
idx_df <- data.frame(Index = idx_rows$Index, Value = round(idx_rows$Value, 4),
                     stringsAsFactors = FALSE)
knitr::kable(idx_df, caption = "Capability indices summary")
```

## Capability Judgment

```{r judgment}
judgment <- diag$capability_judgment
if (!is.null(judgment)) {
  j_df <- data.frame(
    Metric = c("Cpk status", "Ppk status", "Cp status", "Pp status", "Overall verdict"),
    Status = c(judgment$cpk_status, judgment$ppk_status,
               judgment$cp_status,  judgment$pp_status,
               judgment$overall_verdict)
  )
  knitr::kable(j_df, caption = "Capability judgment")
}
```

# Expected Non-Conforming Rate (PPM)

```{r ppm-table}
if (!is.null(data_tbls$ppm_within) && nrow(data_tbls$ppm_within) > 0) {
  knitr::kable(round_df(data_tbls$ppm_within, 2),
               caption = "Expected PPM (within-subgroup, normal only)")
}
if (!is.null(data_tbls$ppm_overall) && nrow(data_tbls$ppm_overall) > 0) {
  knitr::kable(round_df(data_tbls$ppm_overall, 2),
               caption = "Expected PPM (overall)")
}
```

# Diagnostics

```{r diagnostics, results='asis'}
if (analysis_type == "normal") {
  method <- diag$normality_method %||% "Normality test"
  p_val  <- diag$normality_p_value
  if (is.null(p_val) || is.na(p_val)) {
    cat(sprintf("- Normality test: **%s** (p-value not computed, possibly large n)\n", method))
  } else {
    verdict <- if (p_val < 0.05) "rejected (data non-normal)" else "not rejected (data consistent with normality)"
    cat(sprintf("- Normality test: **%s** (p = %s) -> %s\n", method, fmt_sci(p_val), verdict))
  }
} else if (analysis_type == "nonnormal") {
  cat(sprintf("- Fitted distribution: **%s**\n", diag$distribution))
  if (!is.null(diag$aic)) cat(sprintf("- AIC: %.2f\n", diag$aic))
  if (!is.null(diag$bic)) cat(sprintf("- BIC: %.2f\n", diag$bic))
  if (!is.null(diag$ks_p_value)) {
    ks_verdict <- if (diag$ks_p_value < 0.05) "poor fit" else "acceptable fit"
    cat(sprintf("- Kolmogorov-Smirnov test: D = %.4f, p = %s -> %s\n",
                diag$ks_statistic, fmt_sci(diag$ks_p_value), ks_verdict))
  }
} else {
  cat(sprintf("- Method: **%s** (no distributional assumption)\n", diag$method))
  cat(sprintf("- Sample size: %d\n", diag$sample_size %||% stats$n))
}
```

```{r warnings, results='asis'}
if (length(diag$warnings) > 0) {
  cat("## Warnings\n\n")
  for (w in diag$warnings) cat(sprintf("- %s\n", w))
}
```

# Process Advice

```{r advice, results='asis'}
advice <- diag$process_advice
if (!is.null(advice) && length(advice) > 0) {
  for (a in advice) cat(sprintf("- %s\n", a))
} else {
  cat("- No specific advice; process is performing within expectations.\n")
}
```

# Visual Diagnostics

## Process Capability Sixpack

```{r sixpack, fig.width=12, fig.height=10}
if (!is.null(results) && !is.null(theme_obj)) {
  plotter <- iQualityR.capa::CapabilityPlotter$new()
  p <- plotter$render(results, theme_obj, type = "full", plan = plan)
  print(p)
} else {
  cat("Sixpack plot not available (results or theme missing).\n")
}
```

## Histogram

```{r histogram, fig.width=8, fig.height=5}
if (!is.null(results) && !is.null(theme_obj)) {
  plotter <- iQualityR.capa::CapabilityPlotter$new()
  print(plotter$render(results, theme_obj, type = "basic", plan = plan))
}
```

## Q-Q Plot

```{r qq, fig.width=8, fig.height=5}
if (!is.null(results) && !is.null(theme_obj)) {
  plotter <- iQualityR.capa::CapabilityPlotter$new()
  print(plotter$render(results, theme_obj, type = "qq", plan = plan))
}
```

## Capability Indices Bar Chart

```{r capbar, fig.width=8, fig.height=5}
if (!is.null(results) && !is.null(theme_obj)) {
  plotter <- iQualityR.capa::CapabilityPlotter$new()
  print(plotter$render(results, theme_obj, type = "capbar", plan = plan))
}
```

# Bootstrap Confidence Intervals

```{r bootstrap-ci}
boot <- data_tbls$bootstrap_ci
if (!is.null(boot) && nrow(boot) > 0) {
  knitr::kable(round_df(boot, 4), caption = "Bootstrap confidence intervals")
} else {
  cat("Bootstrap CI not computed (disabled or insufficient samples).\n")
}
```

---

Report generated by `iQualityR.capa` at `r format(params$timestamp %||% Sys.time(), "%Y-%m-%d %H:%M:%S")`.
