---
title: "Core Infrastructure: The R6 Base Classes and Workflow Engine"
author: "iQualityR.core Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Core Infrastructure: The R6 Base Classes and Workflow Engine}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

library(iQualityR.core)
library(ggplot2)

fmt_num <- function(x, digits = 4) formatC(x, format = "f", digits = digits)
fmt_pct <- function(x, digits = 2) paste0(formatC(x, format = "f", digits = digits), "%")
```

# 1. Introduction

## 1.1 Why a layered R6 architecture?

Quality engineering covers a broad surface — descriptive statistics,
hypothesis testing, measurement system analysis, process capability,
design of experiments, statistical process control, reliability, and
predictive modelling. Each domain has its own canonical methods, but
they all share the same *workflow*: capture context, execute a
computation, visualise the result, and export a report. Classical R
packages treat each analysis as a one-off function call; the engineer
stitches the pieces together with ad-hoc scripts, copy-pastes data
between tools, and rebuilds the same report template in every project.

`iQualityR.core` breaks this pattern by providing a small set of R6
abstract base classes that codify the shared workflow. Every domain
package (`iQualityR.msa`, `iQualityR.spc`, `iQualityR.capa`, ...)
subclasses these bases, so a practitioner who has learned the workflow
for gage R&R already knows the workflow for SPC, capability, and DOE.

The architectural trade-off is deliberate. R6 reference semantics are
not a universal fit for R, but for quality engineering they are the
right choice because:

- **Stateful lifecycle.** An analysis has a natural state — plan set,
  data loaded, computation run, results cached, plot rendered, report
  exported. R6 encapsulates this state without polluting the global
  environment.
- **Inheritance.** A concrete `IqrCapabilityTask` inherits the plan
  storage, theme plumbing, and reporter wiring from `IqrTaskBase`,
  overriding only the domain-specific compute logic. This eliminates
  boilerplate that would otherwise be duplicated across ten packages.
- **Encapsulation.** Private fields and methods hide implementation
  details; the public API (`$new()`, `$run()`, `$plot()`, `$report()`,
  `$summary()`) is stable and documented.
- **Reference identity.** A task object can be passed to a plotter and
  a reporter and they both see the same results — no copy-on-modify
  surprises.

## 1.2 The class hierarchy at a glance

| Layer | Base class | Responsibility | Key subclass examples |
|:------|:-----------|:---------------|:----------------------|
| Plan | `IqrPlanBase` | 4M1E metadata, criteria, params | `MSAPlan`, `CapabilityPlan`, `SpcPlan`, `DoePlan`, `ReliabilityPlan`, `SamplingPlan` |
| Task | `IqrTaskBase` | Orchestrates plan + data + executor trio | `IqrCapabilityTask`, `IqrSpcTask`, `IqrDoeTask`, `IqrReliabilityTask`, `IqrSamplingTask` |
| Analyzer | `IqrAnalyzerBase` | Computes statistics, 4-slot result container | `CapabilityAnalyzer`, `SpcAnalyzer`, `DoeAnalyzer`, `ReliabilityAnalyzer`, `MsaAnalyzer` |
| Plotter | `IqrPlotterBase` | Renders ggplot2, palette toolbox | `CapabilityPlotter`, `SpcPlotter`, `DoePlotter`, `ReliabilityPlotter` |
| Reporter | `IqrReporter` | Multi-format export dispatcher | (used directly; subclasses register templates) |
| Exporter | `ExcelExporter` | Themed multi-sheet xlsx | (used directly) |

Every domain package composes these bases into a concrete
`Plan → Task → Analyzer → Plotter → Reporter` quintet. The remainder
of this vignette walks through each layer in detail.

# 2. The Plan Layer — IqrPlanBase

## 2.1 The 4M1E metadata model

The `IqrPlanBase` class captures study context using the classical
**4M1E** cause-classification from Ishikawa: Man, Machine, Material,
Method, Environment. A sixth slot, `project`, carries project-level
identifiers. This metadata is not decorative — it flows into reports,
is versioned with the data, and enables the traceability required by
ICH Q9 and ISO 9001.

The 4M1E slots are initialised as empty lists and populated via
`$set_meta()`:

```{r plan-4m1e}
plan <- IqrPlanBase$new(
  task_tag  = "capability",
  conf_level = 0.95
)

plan$set_meta("man",         operator = "Shift A", experience_yrs = 5)
plan$set_meta("machine",     model = "CNC-007",  spindle_rpm = 12000)
plan$set_meta("material",    batch = "SUJ2-26-07", hardness_hrc = 62)
plan$set_meta("method",      procedure = "SOP-001", subgroup_n = 5)
plan$set_meta("environment", temp_C = 22, humidity_pct = 45)
plan$set_meta("project",     id = "Q-2026-001", customer = "ACME")

str(plan$meta_data, max.level = 1)
```

## 2.2 Creating and validating a plan

A plan carries three additional slots beyond 4M1E:

- `task_tag` — a short identifier (`"capability"`, `"msa"`, `"spc"`)
  used by the reporter to locate templates.
- `conf_level` — the confidence level (default 0.95), validated to be
  in (0, 1).
- `criteria` — named evaluation criteria, e.g. specification limits or
  acceptance thresholds, set via `$set_criteria()`.
- `stats_params` — domain-specific statistical parameters, populated
  via `...` at construction or merged later.

```{r plan-criteria}
plan$set_criteria(lsl = 9.95, usl = 10.05, target = 10.00, cpk_min = 1.33)
plan$criteria
plan$validate()  # returns invisibly; raises on invalid conf_level
```

## 2.3 Handing off to the analyzer

The `$to_list()` method serialises the plan into a plain list that the
analyzer consumes via `$setup()`. This decoupling lets the analyzer
accept either a full `IqrPlanBase` object or a bare list — useful for
testing and for programmatic pipelines.

```{r plan-to-list}
str(plan$to_list(), max.level = 1)
```

# 3. The Task Layer — IqrTaskBase

## 3.1 Task as the orchestrator

`IqrTaskBase` is the high-level coordinator. It holds three things:
the data, an `IqrTheme` object, and an `executor` list. A concrete
subclass (e.g. `IqrCapabilityTask`) populates the executor list with
the domain-specific analyzer, plotter, and reporter, then exposes the
five-stage lifecycle through virtual methods.

```{r task-lifecycle, eval = FALSE}
# Pseudo-code showing the lifecycle; real construction uses subclasses
task <- IqrTaskBase$new(
  data  = bearing_diameter,
  theme = "academic"
)
task$compute()   # virtual — subclass dispatches to analyzer$run()
task$summary()   # virtual — subclass prints domain-specific summary
task$plot()      # virtual — subclass calls plotter$render()
task$report()    # virtual — subclass calls reporter$export()
```

The base class itself raises informative errors if a virtual method is
called without being overridden — this catches missing implementations
early rather than producing silent no-ops.

## 3.2 Theme resolution at construction

The `IqrTaskBase$initialize()` method accepts three theme input types,
so downstream code can pass whichever is convenient:

```{r task-theme-types, eval = FALSE}
IqrTaskBase$new(data, theme = "academic")       # character: IqrTheme$new("academic")
IqrTaskBase$new(data, theme = my_theme_obj)     # IqrTheme: used directly
IqrTaskBase$new(data, theme = my_config_obj)    # ThemeConfig: wrapped in IqrTheme
```

This is the single point at which the active theme is bound; every
downstream plot inherits it without re-resolution.

## 3.3 The task registry helper

For packages that dispatch tasks by name (e.g. a backend HTTP handler),
`create_task_registry()` builds a mapping from task names to the
analyzer / plotter / reporter class names:

```{r task-registry}
registry <- create_task_registry(c("capability", "msa", "spc"))
str(registry, max.level = 2)
```

# 4. The Analysis Layer — IqrAnalyzerBase

## 4.1 The 4-slot result container

`IqrAnalyzerBase` defines a standardised result container with four
slots, enforced at construction:

| Slot | Type | Purpose |
|:-----|:-----|:--------|
| `statistics` | named list | Scalar summary metrics (mean, sd, Cpk, p-value, ...) |
| `diagnostics` | named list | Diagnostic objects (normality test, VIF, Cook's D, ...) |
| `data_tables` | named list of data.frames | Tabular outputs (subgroup summaries, violation tables, ...) |
| `raw_output` | arbitrary | The original model object (lm, aov, coxph, ...) for advanced users |

The container is reset at the start of every `$run()` call, so an
analyzer instance is reusable across datasets.

## 4.2 The setup / run / get_results protocol

```{r analyzer-protocol}
analyzer <- IqrAnalyzerBase$new()

# 1. Setup: accept a plan or a list
analyzer$setup(plan)

# 2. (In a real subclass) override .run_logic(dt) to compute results.
#    Here we set values directly to demonstrate the container.
analyzer$set_statistic("mean", 10.021)
analyzer$set_statistic("sd",   0.034)
analyzer$set_diagnostic("normality_p", 0.42)
analyzer$set_datatable("subgroups", data.frame(
  subgroup = 1:3,
  mean     = c(10.01, 10.03, 10.02),
  n        = c(5, 5, 5)
))
analyzer$set_raw_output("lm_object_placeholder")

# 3. Retrieve results
str(analyzer$get_results(), max.level = 2)
```

## 4.3 The app_result 9-slot contract

For results that cross the R-to-frontend boundary (e.g. a Spring Boot
backend serving a Vue/Plotly frontend), `app_result()` produces a
JSON-serialisable 9-slot envelope that subsumes the 4-slot analyzer
container:

| Slot | Type | Purpose |
|:-----|:-----|:--------|
| `ok` | logical | Success flag |
| `code` | character | Stable machine-readable code (`"success"`, `"analysis_failed"`, ...) |
| `message` | character | User-facing message |
| `data` | named list | Summary metrics (mirrors `statistics`) |
| `tables` | named list of data.frames | Tabular outputs (mirrors `data_tables`) |
| `plots` | named list | Plot file descriptors or Plotly JSON specs |
| `files` | named list | Output file paths (Excel, PDF, ...) |
| `warnings` | character vector | Non-fatal warnings |
| `metadata` | named list | Diagnostic metadata (mirrors `diagnostics`) |

Two factory functions cover the common cases:

```{r app-result-factories}
ok <- app_success(
  code = "success",
  data = list(cpk = 1.42, ppk = 1.38),
  tables = list(summary = data.frame(metric = "Cpk", value = 1.42))
)
str(ok, max.level = 1)

err <- app_error(
  code = "normality_failed",
  message = "Data failed Anderson-Darling test; non-normal capability recommended."
)
str(err, max.level = 1)
```

The 9-slot contract is the keystone of the framework's
software-engineering backbone: the same R analyzer feeds a console
session, a backend API, and an Excel workbook, all from one
serialisable object.

# 5. The Visualization Layer — IqrPlotterBase

## 5.1 The render() contract

`IqrPlotterBase` declares a single virtual method, `render(results, theme_obj, type = "full", ...)`.
Subclasses override it to produce a ggplot2 object (or a patchwork
composition). The contract is deliberately minimal — the heavy lifting
lives in the palette toolbox.

## 5.2 The color toolbox

The toolbox is the architectural piece that makes the theme system
work. It exposes palette accessors (`.pal_*`) and scale factories
(`.scale_*`) with a leading-dot naming convention (private by R6
convention, inheritable). Subclasses call these helpers from inside
`render()` and never touch `IqrTheme` internals directly.

| Toolbox method | Returns | Use when |
|:----------------|:--------|:---------|
| `.pal_discrete(theme, n = NULL)` | character vector | Categorical grouping; auto-extends to `n` |
| `.pal_sequential(theme)` | 3-stop gradient | Non-negative continuous magnitudes |
| `.pal_diverging(theme)` | 3-stop (low/mid/high) | Signed values around a midpoint |
| `.pal_semantic(theme, name = NULL)` | named vector or scalar | Verdict colors (pass/fail/watch) |
| `.pal_ui(theme, name, default)` | scalar hex | UI slots (primary, danger, ...) |
| `.contrast_text(bg)` | "#000000" or "#FFFFFF" | Readable text on a colored background |
| `.scale_fill_discrete(theme, ...)` | ggplot2 ScaleDiscrete | Bar/box fills by category |
| `.scale_color_discrete(theme, ...)` | ggplot2 ScaleDiscrete | Lines/points by category |
| `.scale_fill_color_paired(theme, ...)` | list(fill, color) | Coordinated fill + darkened outline |
| `.scale_fill_sequential(theme, ...)` | ggplot2 ScaleContinuous | Density / count gradients |
| `.scale_fill_diverging(theme, ...)` | ggplot2 ScaleContinuous | Residual / effect gradients |
| `.scale_fill_semantic(theme, labels)` | ggplot2 ScaleDiscrete | Pass/fail/watch verdicts |

## 5.3 Subclassing IqrPlotterBase

The pattern below is what a downstream plotter looks like. It inherits
the toolbox, calls the helpers, and stays agnostic of the active theme:

```{r minimal-plotter}
MyPlotter <- R6::R6Class("MyPlotter",
  inherit = IqrPlotterBase,
  public = list(
    render = function(results, theme_obj, type = "full", ...) {
      ggplot(results$data, aes(x = group, y = value, fill = group)) +
        geom_col() +
        theme_obj$theme_iqr() +
        self$.scale_fill_discrete(theme_obj) +
        labs(title = results$title)
    }
  )
)

sim_results <- list(
  title = "Operator means (Type 1 gage study)",
  data  = data.frame(
    group = factor(c("Op A", "Op B", "Op C")),
    value = c(10.02, 10.08, 9.98)
  )
)

MyPlotter$new()$render(sim_results, IqrTheme$new("workbench"))
```

Switching the theme argument recolours the plot without touching the
plotter code — this is the design goal of the toolbox.

# 6. The Report Layer — IqrReporter and ExcelExporter

## 6.1 IqrReporter: multi-format export dispatcher

`IqrReporter` is constructed from an `IqrTheme` and dispatches to
format-specific exporters. It supports five formats: `excel`, `html`,
`pdf`, `docx`, `pptx`. The reporter consumes the same result list that
plotters consume, guaranteeing that "what you see is what you export".

```{r reporter-construct, eval = FALSE}
theme   <- IqrTheme$new("academic")
reporter <- IqrReporter$new(theme)

# Register a custom Excel generator for the "capability" task
reporter$register(
  task_tag        = "capability",
  excel_generator = function(results, plan) {
    list(
      Summary = data.frame(metric = "Cpk", value = results$statistics$cpk),
      Data    = results$data_tables$subgroups
    )
  }
)

# Export
reporter$export(results = analyzer$get_results(),
                plan    = plan,
                task_tag = "capability",
                format  = "excel",
                path    = "capability_report.xlsx")
```

## 6.2 Template discovery

For R Markdown formats (html/pdf/docx/pptx), the reporter searches for
a template in three tiers, then fails with an actionable error:

1. A template registered via `$register(task_tag, rmd_template = ...)`.
2. `inst/templates/<task_tag>_template.Rmd` in `iQualityR.core` or any
   of the eight domain packages.
3. A glob match under `iQualityR.*/inst/templates/` in the working
   directory or its parent.

This convention lets each domain package ship its own templates
without the core package needing to know about them.

## 6.3 ExcelExporter: themed multi-sheet workbooks

`ExcelExporter` is constructed from a `ThemeConfig` (not the full
`IqrTheme`) and produces themed xlsx files with title rows, header
styles, zebra striping, and column-width heuristics. The styles are
derived from the theme's UI palette, so a workbook matches the plots
that accompany it.

```{r excel-export, eval = FALSE}
theme    <- IqrTheme$new("academic")
exporter <- ExcelExporter$new(theme$config)

sheets <- list(
  Summary = data.frame(
    metric = c("mean", "sd", "Cpk"),
    value  = c(10.021, 0.034, 1.42)
  ),
  Subgroups = data.frame(
    subgroup = 1:3,
    mean     = c(10.01, 10.03, 10.02)
  )
)

exporter$export_excel(
  data  = sheets,
  path  = tempfile(fileext = ".xlsx"),
  title = "Bearing Diameter Capability Study"
)
```

The exporter supports per-call style overrides via
`$set_excel_theme(...)` and a `$reset_excel_theme()` restore.

# 7. Utility Functions

The package exports a small set of dependency-free utilities that
downstream packages build on.

## 7.1 Validation

- `validate_metadata(metadata)` — ensures all six 4M1E slots exist.
- `validate_inputs(data, required_cols)` — checks data.frame-ness and
  required columns.
- `validate_output_dir(path, create = TRUE)` — ensures a writable
  directory, creating it if needed.

## 7.2 Formatting

- `format_p_value(p, context = "table"|"plot")` — significance stars
  for tables, `<0.001` notation for plots.
- `format_scientific(x, digits = 3)` — scientific notation below
  0.001, fixed notation otherwise.

```{r format-utils}
format_p_value(c(0.0001, 0.003, 0.04, 0.08, 0.5))
format_p_value(c(0.0001, 0.003, 0.04, 0.08, 0.5), context = "plot")
format_scientific(c(0.00005, 0.042, 1234.5))
```

## 7.3 Sigma and tolerance helpers

- `moving_range_stats(data, m_span)` — moving ranges, their mean and
  median, padded with `m_span - 1` NAs. Used by I-MR charts.
- `safe_tolerance(usl, lsl)` — returns `usl - lsl` or `NULL` when
  either limit is missing.

```{r sigma-utils}
moving_range_stats(c(10.2, 10.3, 10.1, 10.5, 10.2), m_span = 2)
safe_tolerance(10.05, 9.95)
safe_tolerance(10.05, NULL)
```

## 7.4 IDs and config

- `generate_anon_id(n, k = 8, prefix = "ID")` — unique randomised
  identifiers for anonymising part/subject IDs.
- `get_config(key, default)` — retrieves a value from the
  `iQualityR.config` option list.
- `create_task_registry(tasks, ...)` — builds the name-to-class map
  described in Section 3.3.
- `null_to_na(x)` — `NULL` to `NA` coercion for data-frame binding.

```{r id-utils}
head(generate_anon_id(6, k = 6))
```

# 8. Internationalization — iqr_t, iqr_locale, iqr_set_locale

The framework ships a real i18n layer with on-disk CSV dictionaries and
`{placeholder}` interpolation. Two locales are supported out of the
box: `en-US` and `zh-CN`.

```{r i18n}
iqr_t("common.missing_required_column", locale = "en-US", column = "X")
iqr_t("common.missing_required_column", locale = "zh-CN", column = "X")
```

Set the global locale once and every downstream `iqr_t()` call picks
it up:

```{r i18n-set}
iqr_set_locale("zh-CN")
iqr_t("common.missing_required_column", column = "MeasurementValue")
iqr_set_locale("en-US")
```

Dictionaries live in `inst/i18n/<locale>.csv` and are cached in an
environment after first load, so repeated lookups are fast. Extending
to a new locale is a matter of adding a CSV file — no code changes.

# 9. Best Practices for Extending .core

## 9.1 Subclassing checklist

When building a new domain task (e.g. `IqrFooTask`):

1. **Plan.** Subclass `IqrPlanBase`. Add domain-specific fields (e.g.
   `lsl`, `usl`, `target` for capability). Override `validate()` to
   check domain constraints; always call `super$validate()` first.
2. **Analyzer.** Subclass `IqrAnalyzerBase`. Implement
   `private$.run_logic(dt)`. Use `self$set_statistic()`,
   `self$set_diagnostic()`, `self$set_datatable()`,
   `self$set_raw_output()` to populate the 4-slot container.
3. **Plotter.** Subclass `IqrPlotterBase`. Implement
   `render(results, theme_obj, type, ...)`. Use the `.pal_*` and
   `.scale_*` toolbox — never call `IqrTheme` internals directly.
4. **Task.** Subclass `IqrTaskBase`. In `initialize()`, construct the
   plan, analyzer, plotter, and reporter. Wire `$compute()`,
   `$summary()`, `$plot()`, `$report()` to the executor trio.
5. **Reporter.** Register a task template via
   `reporter$register(task_tag, rmd_template, excel_generator)`. Ship
   the Rmd template at `inst/templates/<task_tag>_template.Rmd`.

## 9.2 Result-protocol discipline

- Always populate the 4-slot container — downstream plotters and
  reporters rely on predictable names.
- Use `app_result()` (not a bare list) when crossing the R-to-frontend
  boundary. The 9-slot envelope is the API contract.
- Never put non-serialisable objects (e.g. R6 instances, connections)
  in `data` / `tables` / `metadata` — they break JSON serialisation.
  Use `raw_output` for model objects.

## 9.3 Theme discipline

- Resolve the theme once, at task construction. Pass `theme_obj` down
  to plotters; do not re-resolve inside `render()`.
- Use the toolbox, not `IqrTheme` internals. If the toolbox is missing
  a helper you need, add it to `IqrPlotterBase` rather than bypassing
  it in a subclass.

# 10. References

- Chambers, W. (2019). *Object-Oriented Programming, Functional
  Programming, and R*. Statistical Science.
- ICH (2005). *Q9 Quality Risk Management*.
- ISO 9001 (2015). *Quality Management Systems — Requirements*.
- Ishikawa, K. (1985). *What is Total Quality Control? The Japanese
  Way*. Prentice-Hall.
- Montgomery, D. C. (2013). *Statistical Quality Control* (7th ed.).
  Wiley.
- WCAG 2.x (2018). *Web Content Accessibility Guidelines*.
  https://www.w3.org/TR/WCAG21/

*See `vignette("theme-system", package = "iQualityR.core")` for the
complete theme and color pipeline reference, and
`vignette("iQualityR", package = "iQualityR")` for the framework-level
methodology overview.*
