---
title: "Introduction to iQualityR.core"
subtitle: "Foundational infrastructure for the iQualityR framework"
author: "Zhiming Chen"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Introduction to iQualityR.core}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(iQualityR.core)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.path = "fig-iqr-core-"
)
```

# Overview

The `iQualityR.core` package provides the foundational infrastructure
for the iQualityR quality engineering framework. It defines R6 abstract
base classes for plans, tasks, analyzers, plotters, and reporters; a
unified ggplot2 theme system with four coordinated palette classes;
Excel export utilities; metadata management following a 4M1E convention
(Man, Machine, Material, Method, Environment); simple
internationalization (en-US, zh-CN); and utility functions for
validation, formatting, and configuration.

Downstream packages such as `iQualityR.plot`, `iQualityR.stat`,
`iQualityR.msa`, `iQualityR.spc`, and `iQualityR.doe` build on these
components. Users of those packages do not need to interact with
`iQualityR.core` directly, but understanding the theme system helps
when customizing analyses or building new plotter subclasses.

# Base classes

## IqrPlanBase

An `IqrPlanBase` object stores the configuration of an analysis: the
task tag, the confidence level, named criteria (e.g. specification
limits), and 4M1E metadata.

```{r plan}
plan <- IqrPlanBase$new(
  task_tag = "capability",
  conf_level = 0.95
)

plan$set_criteria(lsl = 0, usl = 10)
plan$set_meta("man",       operator = "John", experience = 5)
plan$set_meta("machine",   model = "Machine A")
plan$set_meta("material",  batch = "B-2025")
plan$set_meta("method",    procedure = "SOP-001")
plan$set_meta("environment", temperature = 23)
plan$set_meta("project",   id = "Q-001")

plan$conf_level
plan$criteria$usl
```

Convert the plan to a plain list:

```{r plan-list}
str(plan$to_list(), max.level = 2)
```

## IqrAnalyzerBase

`IqrAnalyzerBase` is the abstract executor that runs an analysis and
stores statistics, diagnostics, data tables, and raw output. Subclasses
override the private `.run_logic` method.

```{r analyzer}
analyzer <- IqrAnalyzerBase$new()

analyzer$setup(list(alpha = 0.05, method = "t-test"))

analyzer$set_statistic("mean", 10.1)
analyzer$set_diagnostic("normality", "OK")
analyzer$set_datatable("runs", data.frame(
  run = 1:3,
  value = c(9.8, 10.2, 10.3)
))

str(analyzer$get_results(), max.level = 2)
```

## IqrTaskBase, IqrPlotterBase, IqrReporter

- `IqrTaskBase` is the high-level coordinator. A subclass wires
  together a plan, data, an analyzer, a plotter, and a reporter.
- `IqrPlotterBase` is the abstract interface for plot rendering. It
  also exposes a toolbox of palette / scale accessors that subclasses
  inherit, so downstream plotters do not have to know the internals of
  `IqrTheme` (see the *Plotter toolbox* section).
- `IqrReporter` handles Excel and R Markdown export.

# Theme system

The theme system is the single point of control for visual consistency
across the entire framework. The design goal is that **users only need
to pick a theme preset**; every downstream package, every plot type,
and every analysis module then obtains a coordinated color scheme
automatically.

## Four palette classes

Every theme preset bundles four palette classes, each covering a
distinct visual scenario:

| Class | Type | Typical use | Example |
|-------|------|-------------|---------|
| `discrete` | Qualitative | Categorical grouping variables | SPC chart with multiple operators, DOE factor levels |
| `sequential` | Ordered gradient | Non-negative continuous magnitudes | Capability density, heatmaps, contour surfaces |
| `diverging` | Signed gradient | Values centered on a neutral midpoint | Residuals, standardized effects, correlations |
| `semantic` | Verdict-bound | Pass / fail / watch / good / bad / neutral | MSA verdict charts, capability traffic lights |

Using the right palette class for the right scenario is what produces a
coherent look across heterogeneous plots. A control chart (discrete),
a residual histogram (diverging), and a verdict bar (semantic) all
share the same theme even though they draw from different palette
classes.

## Three built-in presets

Three built-in presets are shipped with the package:

```{r presets-list}
names(IqrTheme$new("academic")$config$style_presets)
```

The first three are pure-built-in (no external dependencies):

- `"academic"` -- publication-ready, colorblind-safe qualitative palette
  (Paul Tol), ColorBrewer Blues / RdBu gradients, black axes. Default
  for reports and papers.
- `"workbench"` -- higher-saturation UI palette for interactive apps
  and dashboards; blue / teal / amber / red accents.
- `"tech"` -- GitHub-inspired cool palette with strong contrast;
  suitable for technical documentation and demos.

The remaining seven (`economist` / `wsj` / `gdocs` / `tufte` / `few` /
`solarized` / `prism`) are external theme presets -- see the next
section.

Switch presets at any time with `$set_theme()`:

```{r theme-switch, message = FALSE}
library(ggplot2)

theme <- IqrTheme$new("academic")

ggplot(mtcars, aes(x = factor(cyl), fill = factor(cyl))) +
  geom_bar(show.legend = FALSE) +
  theme$theme_iqr() +
  theme$scale_fill_iqr() +
  labs(title = "Academic preset")

theme$set_theme("workbench")

ggplot(mtcars, aes(x = factor(cyl), fill = factor(cyl))) +
  geom_bar(show.legend = FALSE) +
  theme$theme_iqr() +
  theme$scale_fill_iqr() +
  labs(title = "Workbench preset")
```

## External theme presets

In addition to the three built-in presets, `iQualityR.core` ships seven
external theme presets that mirror the look-and-feel of well-known
publications and tools. Each external preset is a **full preset** --
it carries its own UI colors and all four data palette classes
(discrete / sequential / diverging / semantic), so picking
`"economist"` recolors every plot to match *The Economist* style,
not just the axis frame.

The external ggplot2 theme function (e.g. `ggthemes::theme_economist`)
is layered on top of the preset's color configuration inside
`theme_iqr()`, so the final plot inherits both the publication's
typography / gridlines and its color palette.

| Preset | Source package | Look-and-feel | Discrete first 3 colors |
|--------|----------------|---------------|--------------------------|
| `economist` | ggthemes | *The Economist*: deep teal + slate blue + cyan | `#6794A7 #014D64 #7AD2F6` |
| `wsj` | ggthemes | *Wall Street Journal*: red + blue + gold on white | `#C72E29 #016392 #BE9C2E` |
| `gdocs` | ggthemes | Google Docs: Material Design blue / red / yellow / green | `#4285F4 #EA4335 #FBBC04` |
| `tufte` | ggthemes | Tufte minimal ink: grayscale only | `#000000 #595959 #8C8C8C` |
| `few` | ggthemes | Stephen Few's qualitative palette | `#5DA5DA #FAA43A #60BD68` |
| `solarized` | ggthemes | Solarized (Ethan Schoonover): 16-color perceptual scheme | `#268BD2 #B58900 #CB4B16` |
| `prism` | ggprism | GraphPad Prism "winter_bright" | `#077E97 #800080 #000080` |

External presets are available when the corresponding package is
installed (`ggthemes` and/or `ggprism`, declared in `Suggests`). When
the package is absent, the preset's color configuration is still
applied -- only the ggplot2 theme function is skipped, so the plots
still get the right colors with a fallback `academic` frame.

### Switching to an external preset

```{r external-switch, message = FALSE}
theme <- IqrTheme$new("economist")

ggplot(mtcars, aes(x = factor(cyl), fill = factor(cyl))) +
  geom_bar(show.legend = FALSE) +
  theme$theme_iqr() +
  theme$scale_fill_iqr() +
  labs(title = "Economist preset (full palette + theme)")
```

Note how both the bar colors and the axis / gridline style change
together -- this is the difference between the new full-preset
behavior and the old "external theme only" behavior, which left the
data colors on the academic palette.

### Comparing external palettes side by side

The four palette classes are coordinated within each preset, so the
same scenario plot looks coherent regardless of which external
preset is active:

```{r external-compare, fig.show = "hold", out.width = "33%"}
data(mtcars)
mtcars$cyl <- factor(mtcars$cyl)

for (style in c("wsj", "few", "solarized")) {
  theme <- IqrTheme$new(style)
  print(
    ggplot(mtcars, aes(x = wt, y = mpg, color = cyl)) +
      geom_point(size = 3) +
      theme$theme_iqr() +
      theme$scale_color_iqr() +
      labs(title = paste0(style, " preset"))
  )
}
```

### External preset palettes under the hood

You can inspect any preset's full color configuration directly:

```{r external-inspect}
theme <- IqrTheme$new("prism")

# Discrete palette (GraphPad Prism winter_bright)
theme$get_pal("discrete")

# Sequential gradient (single-hue cyan)
theme$get_pal("sequential")

# Diverging gradient (magenta -> white -> cyan)
theme$get_pal("diverging")

# Semantic verdict colors (pass / fail / watch)
theme$get_pal("semantic")
```

### Soft dependency behavior

`ggthemes` and `ggprism` are declared in `Suggests`, not `Imports`, so
the package installs cleanly even when they are absent. The runtime
contract has three tiers:

- **Package installed** (the common case): full preset + external
  ggplot2 theme function. The plot inherits both the publication's
  typography / gridlines and its color palette. This is the intended
  publication-quality look.
- **Package absent but preset name requested**: the preset's color
  configuration is still applied (the preset lives in `style_presets`,
  not in the external package); only the external ggplot2 theme
  function is skipped. The plot gets the right colors with a fallback
  `academic` frame.
- **Preset name not recognized at all**: `set_theme()` raises an
  error listing all valid names.

```{r external-soft-dep}
# Preset entry is always present in style_presets, regardless of
# whether ggthemes is installed:
"economist" %in% names(IqrTheme$new("academic")$config$style_presets)
```


## Accessing palettes directly

`get_pal(type, n, name, custom)` returns the raw color vector. This is
the lowest-level entry point, useful when a downstream plotter needs to
pass colors to a non-ggplot2 function (e.g. `plotly`, `openxlsx`).

```{r get-pal}
theme <- IqrTheme$new("academic")

# Discrete palette (base length = 10)
theme$get_pal("discrete")

# Sequential and diverging palettes (3-stop)
theme$get_pal("sequential")
theme$get_pal("diverging")

# Semantic palette (named vector)
theme$get_pal("semantic")

# Single semantic color by name
theme$get_pal("semantic", name = "pass")
```

### Discrete palette auto-extension

When the number of categories exceeds the base palette length, the
discrete palette is automatically interpolated via
`grDevices::colorRampPalette`:

```{r pal-extend}
theme <- IqrTheme$new("academic")
base <- theme$get_pal("discrete")
extended <- theme$get_pal("discrete", n = 15L)

length(base)
length(extended)
extended[1]   # first color is preserved
extended[15]  # last color matches the last base color
```

This is what every `scale_*_iqr()` call uses internally, so plots with
many categories stay readable without manual color picking.

### Resolution priority: custom > option > preset

`get_pal()` resolves colors in three layers, which lets users override
the preset globally (via `options()`) or per-call (via `custom`):

```{r pal-priority}
theme <- IqrTheme$new("academic")

# 1. Default: preset
head(theme$get_pal("discrete"), 3)

# 2. option() overrides preset globally
op <- options(iqr.custom_discrete = c("#111111", "#222222", "#333333"))
head(theme$get_pal("discrete"), 3)

# 3. custom argument overrides everything per-call
theme$get_pal("discrete", custom = c("#AAAAAA", "#BBBBBB"))

options(op)  # restore
```

## ggplot2 scales

`IqrTheme` exposes one scale family per palette class. All scales
return ready-to-add ggplot2 scale objects.

| Method | Returns | Use when |
|--------|---------|----------|
| `scale_fill_iqr()` / `scale_color_iqr()` | `ScaleDiscrete` or `ScaleContinuous` | General discrete or sequential, controlled by `discrete` flag |
| `scale_fill_sequential()` / `scale_color_sequential()` | `ScaleContinuous` | Continuous non-negative magnitudes |
| `scale_fill_diverging()` / `scale_color_diverging()` | `ScaleContinuous` | Signed continuous values around a midpoint |
| `scale_fill_semantic()` / `scale_color_semantic()` | `ScaleDiscrete` | Verdict / status factors (pass / fail / watch) |

### Discrete fill and color

```{r scale-discrete}
theme <- IqrTheme$new("academic")

ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(size = 3) +
  theme$theme_iqr() +
  theme$scale_color_iqr() +
  labs(title = "Discrete color scale (academic preset)")
```

### Paired fill and color

By default, `scale_fill_iqr()` and `scale_color_iqr()` use the same
colors, so a bar's fill and its legend key match exactly. For plots
where the fill has an outline (bars with `color = ...` mapped to the
same variable), use `style = "paired"` on **both** scales -- the color
scale becomes a darkened variant of the fill scale so edges stay
visible against the fill:

```{r scale-paired}
theme <- IqrTheme$new("workbench")

ggplot(iris, aes(x = Species, y = Sepal.Length,
                 fill = Species, color = Species)) +
  geom_boxplot(linewidth = 0.6, outlier.size = 1.5) +
  theme$theme_iqr() +
  theme$scale_fill_iqr(style = "paired") +
  theme$scale_color_iqr(style = "paired") +
  labs(title = "Paired fill and outline (workbench preset)")
```

Using `style = "paired"` on only one of the two scales has no visible
effect -- the pairing is a coordination contract between the fill and
color scales.

### Sequential gradient

For continuous non-negative magnitudes such as density, frequency, or
process capability:

```{r scale-sequential}
theme <- IqrTheme$new("academic")

ggplot(faithfuld, aes(x = waiting, y = eruptions)) +
  geom_raster(aes(fill = density)) +
  theme$theme_iqr() +
  theme$scale_fill_sequential(name = "Density") +
  labs(title = "Sequential gradient (density)")
```

### Diverging gradient

For signed values centered on a neutral midpoint, e.g. residuals from
a fitted model or standardized effects in a DOE:

```{r scale-diverging}
theme <- IqrTheme$new("academic")

set.seed(1)
df <- data.frame(
  x = seq(-3, 3, length.out = 60),
  y = seq(-3, 3, length.out = 60)
)
df$z <- with(df, sin(x) * cos(y))

ggplot(df, aes(x = x, y = y, fill = z)) +
  geom_raster() +
  theme$theme_iqr() +
  theme$scale_fill_diverging(midpoint = 0, name = "Effect") +
  labs(title = "Diverging gradient (signed effect)")
```

### Semantic verdict scale

For categorical outcomes that carry a verdict -- typical in MSA,
capability, and SPC acceptance charts:

```{r scale-semantic}
theme <- IqrTheme$new("academic")

verdicts <- data.frame(
  part = factor(1:6),
  result = factor(c("pass", "pass", "watch", "fail", "pass", "watch"),
                  levels = c("pass", "watch", "fail"))
)

ggplot(verdicts, aes(x = part, y = 1, fill = result)) +
  geom_col(width = 0.7) +
  theme$theme_iqr() +
  theme$scale_fill_semantic(labels = c("pass", "watch", "fail")) +
  labs(title = "Semantic verdict scale (MSA-style)") +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

The labels passed to `scale_fill_semantic()` must be a subset of the
semantic palette names (`pass`, `fail`, `watch`, `good`, `bad`,
`neutral`, `green`, `yellow`, `red`).

## Color utility functions

The package exports a small set of dependency-free color utilities so
downstream packages can derive coordinated shades from the same
primitives:

```{r color-utils}
# Lighten / darken a base color
lighten("#2563EB", 0.2)
darken("#2563EB", 0.3)

# Linear interpolation between two colors
mix("#FF0000", "#0000FF", 0.5)

# Luminance checks (WCAG-based)
is_dark("#000000")
is_dark("#FFFFFF")
contrast_ratio("#FFFFFF", "#000000")
```

These are useful when a plotter needs to derive, for example, a
lighter ribbon color from the discrete palette without reaching for
another package.

# Plotter toolbox for downstream packages

The most important architectural piece for downstream package authors
is `IqrPlotterBase`. It exposes a toolbox of palette accessors and
scale factories with a leading-dot naming convention (private by
convention, inheritable via R6). Subclasses call these helpers from
inside `render()` and never touch `IqrTheme` internals directly.

```{r plotter-toolbox}
plotter <- IqrPlotterBase$new()
theme   <- IqrTheme$new("academic")

# Palette accessors
plotter$.pal_discrete(theme)              # 10 colors
plotter$.pal_discrete(theme, n = 12L)     # auto-extended to 12
plotter$.pal_sequential(theme)            # 3-stop gradient
plotter$.pal_diverging(theme)             # 3 stops
plotter$.pal_semantic(theme)              # full named vector
plotter$.pal_semantic(theme, name = "fail")

# Scale factories
plotter$.scale_fill_discrete(theme)
plotter$.scale_color_discrete(theme)
plotter$.scale_fill_sequential(theme)
plotter$.scale_color_sequential(theme)
plotter$.scale_fill_diverging(theme)
plotter$.scale_color_diverging(theme)
plotter$.scale_fill_semantic(theme, labels = c("pass", "fail"))
plotter$.scale_color_semantic(theme, labels = c("pass", "fail"))

# Paired fill + color (returns a list of two scales)
paired <- plotter$.scale_fill_color_paired(theme)
names(paired)
```

The benefit is that switching the active theme at the task level is
enough to recolor every plot consistently -- the plotter code does
not change.

## Example: a minimal subpackage plotter

The pattern below is what a downstream plotter (e.g. in
`iQualityR.msa` or `iQualityR.spc`) looks like. It inherits
`IqrPlotterBase`, calls the toolbox helpers, and stays agnostic of the
active theme:

```{r minimal-plotter}
MyPlotter <- R6::R6Class("MyPlotter",
  inherit = IqrPlotterBase,
  public = list(
    render = function(results, theme_obj, type = "full", ...) {
      ggplot(results$data, aes(x = group, y = value, fill = group)) +
        geom_col() +
        theme_obj$theme_iqr() +
        self$.scale_fill_discrete(theme_obj) +
        labs(title = results$title)
    }
  )
)

# Simulate an SPC-style result
spc_results <- list(
  title = "Operator means (Type 1 study)",
  data  = data.frame(
    group = factor(c("Op A", "Op B", "Op C")),
    value = c(10.02, 10.08, 9.98)
  )
)

MyPlotter$new()$render(spc_results, IqrTheme$new("workbench"))
```

# Typical subpackage application scenarios

This section shows how the four palette classes map onto the standard
plot families used across the iQualityR subpackages. Every example
below uses only `iQualityR.core` plus `ggplot2`, so the same patterns
apply verbatim in `iQualityR.msa`, `iQualityR.spc`, `iQualityR.doe`,
and `iQualityR.stat`.

## Scenario 1 -- MSA verdict chart (semantic)

Measurement System Analysis produces explicit verdicts (pass / watch /
fail). Bind them to the semantic palette so a verdict is always the
same color, regardless of plot type:

```{r scenario-msa}
theme <- IqrTheme$new("academic")

msa_summary <- data.frame(
  characteristic = factor(c("Bias", "Linearity", "GRR", "NDC"),
                          levels = c("Bias", "Linearity", "GRR", "NDC")),
  verdict = factor(c("pass", "watch", "fail", "fail"),
                   levels = c("pass", "watch", "fail"))
)

ggplot(msa_summary, aes(x = characteristic, y = 1, fill = verdict)) +
  geom_col(width = 0.6) +
  theme$theme_iqr() +
  theme$scale_fill_semantic(labels = c("pass", "watch", "fail")) +
  labs(title = "MSA study verdicts (semantic palette)") +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        axis.title.y = element_blank())
```

The same `scale_fill_semantic()` call is reused by capability verdicts
and SPC stability flags, so a green bar always means the same thing.

## Scenario 2 -- SPC control chart with operators (discrete, paired)

When a control chart is faceted or colored by operator / machine, use
the discrete palette with `style = "paired"` so the point fill and
its connecting line stay legible:

```{r scenario-spc}
theme <- IqrTheme$new("workbench")

set.seed(7)
spc <- data.frame(
  run     = rep(1:20, times = 3),
  value   = c(rnorm(20, 10, 0.3),
              rnorm(20, 10.1, 0.3),
              rnorm(20, 9.95, 0.3)),
  operator = rep(c("A", "B", "C"), each = 20)
)

ggplot(spc, aes(x = run, y = value,
                color = operator, fill = operator)) +
  geom_hline(yintercept = 10, linetype = 2, color = "grey50") +
  geom_line(linewidth = 0.6) +
  geom_point(shape = 21, color = "black", size = 2) +
  theme$theme_iqr() +
  theme$scale_color_iqr(discrete = TRUE, style = "paired") +
  theme$scale_fill_iqr(discrete = TRUE, style = "paired") +
  labs(title = "SPC run chart by operator (discrete, paired)")
```

## Scenario 3 -- Capability density (sequential)

Process capability histograms overlay a density curve; the density
fill uses the sequential gradient so the bulk of the distribution
reads as the deepest color:

```{r scenario-cap}
theme <- IqrTheme$new("academic")

set.seed(3)
meas <- rnorm(500, mean = 10, sd = 0.5)
cap_df <- data.frame(value = meas)

ggplot(cap_df, aes(x = value)) +
  geom_histogram(aes(y = after_stat(density),
                     fill = after_stat(count)),
                 bins = 30, color = "white", linewidth = 0.2) +
  geom_density(aes(y = after_stat(density)),
               color = theme$get_pal("discrete")[1],
               linewidth = 0.8, fill = NA) +
  theme$theme_iqr() +
  theme$scale_fill_sequential(name = "Count") +
  labs(title = "Capability histogram with density (sequential)")
```

## Scenario 4 -- DOE residuals (diverging)

Residuals from a fitted DOE model are signed; the diverging gradient
makes the sign and magnitude immediately readable:

```{r scenario-doe}
theme <- IqrTheme$new("academic")

set.seed(11)
doe_res <- data.frame(
  run = 1:24,
  residual = rnorm(24, 0, 1.2)
)

ggplot(doe_res, aes(x = run, y = residual, fill = residual)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.4) +
  theme$theme_iqr() +
  theme$scale_fill_diverging(midpoint = 0, name = "Residual") +
  labs(title = "DOE residuals (diverging palette)")
```

## Scenario 5 -- DOE main effects with paired outlines (discrete, paired)

Main-effects plots group runs by factor level. The paired style keeps
the level color consistent across the fill (bar) and the outline /
point:

```{r scenario-doe-main}
theme <- IqrTheme$new("tech")

main_eff <- data.frame(
  level = factor(c("Low", "Center", "High"),
                 levels = c("Low", "Center", "High")),
  mean  = c(8.2, 10.1, 11.8)
)

ggplot(main_eff, aes(x = level, y = mean, fill = level)) +
  geom_col(color = "black", linewidth = 0.4, width = 0.6) +
  theme$theme_iqr() +
  theme$scale_fill_iqr(style = "paired") +
  labs(title = "DOE main effects (discrete, paired)",
       x = "Factor level", y = "Mean response")
```

## Scenario 6 -- Multi-response heatmap (sequential)

When a DOE or capability study reports a continuous response across a
two-dimensional grid, the sequential gradient gives a single
perceptually uniform axis:

```{r scenario-heatmap}
theme <- IqrTheme$new("workbench")

grid <- expand.grid(x = seq(0, 1, 0.05), y = seq(0, 1, 0.05))
grid$response <- with(grid, 30 + 10 * x - 8 * y + 4 * x * y)

ggplot(grid, aes(x = x, y = y, fill = response)) +
  geom_raster() +
  theme$theme_iqr() +
  theme$scale_fill_sequential(name = "Response") +
  labs(title = "Response surface heatmap (sequential)")
```

# Internationalization

Two locales are supported. `iqr_t` looks up a message key and
interpolates named placeholders.

```{r i18n}
iqr_t("common.missing_required_column", locale = "en-US", column = "X")
iqr_t("common.missing_required_column", locale = "zh-CN", column = "X")
```

The active locale can be set globally with `iqr_set_locale`:

```{r set-locale}
iqr_set_locale("zh-CN")
iqr_t("common.missing_required_column", column = "MeasurementValue")
iqr_set_locale("en-US")
```

# Utility functions

Common formatting and validation helpers:

- `format_p_value` summarizes statistical significance
- `moving_range_stats` computes moving ranges and their mean/median
- `safe_tolerance` returns the span `usl - lsl`, or `NULL` if missing
- `generate_anon_id` produces unique randomized identifiers
- `validate_metadata`, `validate_inputs`, `validate_output_dir`
  ensure data structure

```{r utils}
format_p_value(c(0.0001, 0.005, 0.05, 0.1, 0.5))
moving_range_stats(c(10.2, 10.3, 10.1, 10.5, 10.2), m_span = 2)
safe_tolerance(10, 5)
safe_tolerance(10, NULL)
head(generate_anon_id(6, k = 6))
```

# Excel export

`ExcelExporter` writes themed multi-sheet workbooks:

```{r excel-export, eval = TRUE}
theme <- IqrTheme$new("academic")
exporter <- ExcelExporter$new(theme$config)

sheets <- list(
  Summary = data.frame(
    statistic = c("mean", "sd", "n"),
    value = c(10.1, 0.4, 50)
  ),
  Raw = data.frame(run = 1:5, value = rnorm(5, 10, 0.5))
)

out <- tempfile(fileext = ".xlsx")
exporter$export_excel(sheets, path = out, title = "Analysis Report")
file.exists(out)
```

# Architecture notes

- All base classes are R6 reference classes. Subclasses typically
  override a small set of private methods (e.g. `.run_logic`) and
  inherit the rest.
- The theme system follows a three-layer architecture:
  `ThemeConfig` (palette storage and resolution) -> `PlotTheme`
  (ggplot2 scale and theme factories) -> `IqrTheme` (unified facade).
  Downstream code only touches `IqrTheme`.
- The package ships **ten theme presets**: three built-in
  (`academic` / `workbench` / `tech`) and seven external
  (`economist` / `wsj` / `gdocs` / `tufte` / `few` / `solarized` /
  `prism`). Each preset is a full configuration covering UI colors
  and all four data palette classes; external presets additionally
  layer the source publication's ggplot2 theme function on top.
- `IqrPlotterBase` exposes a private-by-convention toolbox
  (`.pal_*`, `.scale_*`) so downstream plotters stay decoupled from
  the theme implementation. Switching the active theme at the task
  level recolors every plot consistently.
- Dependencies are kept minimal: R6, ggplot2, grDevices, openxlsx, and
  rmarkdown are the only heavy imports. Optional packages
  (`systemfonts`, `sysfonts`, `showtext`, `ggthemes`, `ggprism`)
  provide on-demand CJK font handling and external theme presets.
- Result objects are plain lists so they serialize to JSON, Excel,
  or R Markdown without friction.
