---
title: "Introduction to iQualityR.core"
subtitle: "Foundational infrastructure for the iQualityR framework"
author: "Zhiming Chen"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Introduction to iQualityR.core}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(iQualityR.core)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.path = "fig-iqr-core-"
)
```

# Overview

The `iQualityR.core` package provides the foundational infrastructure
for the iQualityR quality engineering framework. It defines R6 abstract
base classes for plans, tasks, analyzers, plotters, and reporters; a
unified ggplot2 theme system; Excel export utilities; metadata
management following a 4M1E convention (Man, Machine, Material, Method,
Environment); simple internationalization (en-US, zh-CN); and utility
functions for validation, formatting, and configuration.

Downstream packages such as `iQualityR.plot`, `iQualityR.stat`, and
`iQualityR.msa` build on these components. Users of those packages do
not need to interact with `iQualityR.core` directly, but understanding
the architecture helps when extending or customizing analyses.

# Base classes

## IqrPlanBase

An `IqrPlanBase` object stores the configuration of an analysis: the
task tag, the confidence level, named criteria (e.g. specification
limits), and 4M1E metadata.

```{r plan}
plan <- IqrPlanBase$new(
  task_tag = "capability",
  conf_level = 0.95
)

plan$set_criteria(lsl = 0, usl = 10)
plan$set_meta("man",       operator = "John", experience = 5)
plan$set_meta("machine",   model = "Machine A")
plan$set_meta("material",  batch = "B-2025")
plan$set_meta("method",    procedure = "SOP-001")
plan$set_meta("environment", temperature = 23)
plan$set_meta("project",   id = "Q-001")

plan$conf_level
plan$criteria$usl
```

Convert the plan to a plain list:

```{r plan-list}
str(plan$to_list(), max.level = 2)
```

## IqrAnalyzerBase

`IqrAnalyzerBase` is the abstract executor that runs an analysis and
stores statistics, diagnostics, data tables, and raw output. Subclasses
override the private `.run_logic` method.

```{r analyzer}
analyzer <- IqrAnalyzerBase$new()

analyzer$setup(list(alpha = 0.05, method = "t-test"))

analyzer$set_statistic("mean", 10.1)
analyzer$set_diagnostic("normality", "OK")
analyzer$set_datatable("runs", data.frame(
  run = 1:3,
  value = c(9.8, 10.2, 10.3)
))

str(analyzer$get_results(), max.level = 2)
```

## IqrTaskBase, IqrPlotterBase, IqrReporter

- `IqrTaskBase` is the high-level coordinator. A subclass wires
  together a plan, data, an analyzer, a plotter, and a reporter.
- `IqrPlotterBase` is the abstract interface for plot rendering.
- `IqrReporter` handles Excel and R Markdown export.

# Theme and plots

`IqrTheme` exposes three built-in visual styles (`academic`,
`workbench`, `tech`) and ggplot2 scales.

```{r theme-ggplot, message = FALSE}
library(ggplot2)

theme <- IqrTheme$new(theme_style = "tech")

ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width,
                 color = Species)) +
  geom_point() +
  theme$theme_iqr() +
  theme$scale_color_iqr() +
  labs(title = "Sepal dimensions by species")
```

Switch styles at any time:

```{r theme-switch, message = FALSE}
theme$set_theme("workbench")

ggplot(mtcars, aes(x = factor(cyl), fill = factor(cyl))) +
  geom_bar(show.legend = FALSE) +
  theme$theme_iqr() +
  theme$scale_fill_iqr() +
  labs(title = "Engine cylinder distribution")
```

# Internationalization

Two locales are supported. `iqr_t` looks up a message key and
interpolates named placeholders.

```{r i18n}
iqr_t("common.missing_required_column", locale = "en-US", column = "X")
iqr_t("common.missing_required_column", locale = "zh-CN", column = "X")
```

The active locale can be set globally with `iqr_set_locale`:

```{r set-locale}
iqr_set_locale("zh-CN")
iqr_t("common.missing_required_column", column = "MeasurementValue")
iqr_set_locale("en-US")
```

# Utility functions

Common formatting and validation helpers:

- `format_p_value` summarizes statistical significance
- `moving_range_stats` computes moving ranges and their mean/median
- `safe_tolerance` returns the span `usl - lsl`, or `NULL` if missing
- `generate_anon_id` produces unique randomized identifiers
- `validate_metadata`, `validate_inputs`, `validate_output_dir`
  ensure data structure

```{r utils}
format_p_value(c(0.0001, 0.005, 0.05, 0.1, 0.5))
moving_range_stats(c(10.2, 10.3, 10.1, 10.5, 10.2), m_span = 2)
safe_tolerance(10, 5)
safe_tolerance(10, NULL)
head(generate_anon_id(6, k = 6))
```

# Excel export

`ExcelExporter` writes themed multi-sheet workbooks:

```{r excel-export, eval = TRUE}
theme <- IqrTheme$new("academic")
exporter <- ExcelExporter$new(theme$config)

sheets <- list(
  Summary = data.frame(
    statistic = c("mean", "sd", "n"),
    value = c(10.1, 0.4, 50)
  ),
  Raw = data.frame(run = 1:5, value = rnorm(5, 10, 0.5))
)

out <- tempfile(fileext = ".xlsx")
exporter$export_excel(sheets, path = out, title = "Analysis Report")
file.exists(out)
```

# Architecture notes

- All base classes are R6 reference classes. Subclasses typically
  override a small set of private methods (e.g. `.run_logic`) and
  inherit the rest.
- Dependencies are kept minimal: R6, ggplot2, openxlsx, and rmarkdown
  are the only heavy imports. Optional packages (`systemfonts`,
  `sysfonts`, `showtext`) provide on-demand CJK font handling.
- Result objects are plain lists so they serialize to JSON, Excel,
  or R Markdown without friction.
