---
title: "The Unified Theme and Color Pipeline"
author: "iQualityR.core Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{The Unified Theme and Color Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

library(iQualityR.core)
library(ggplot2)

# Helpers
fmt_num <- function(x, digits = 4) formatC(x, format = "f", digits = digits)
fmt_pct <- function(x, digits = 2) paste0(formatC(x, format = "f", digits = digits), "%")
```

# 1. Introduction

## 1.1 Why a unified theme system? The color chaos problem

Quality engineering dashboards typically mix plots from many sources:
a histogram from a capability study, a control chart from SPC, a
Pareto chart from a defect analysis, a fishbone diagram from a root
cause session, and a correlation heatmap from a multivariate drill-down.
When each plot picks its own colors — `ggplot2` defaults in one panel,
a hard-coded `c("#E41A1C", "#377EB8", ...)` in another, a copied
`scale_fill_brewer()` in a third — the result is visual noise that
undermines the credibility of the analysis. Worse, when the QA manager
asks "make all the charts match the corporate brand for the customer
report", the engineer faces a find-and-replace marathon across dozens
of scripts.

`iQualityR.core` solves this by providing a single theme pipeline that
every plot function in the framework taps into. One switch changes the
look-and-feel of every plot and every Excel export, consistently and
predictably.

## 1.2 Design goals: one switch, everything recolors

The theme system is built around four design goals:

1. **One switch.** Construct an `IqrTheme` with a preset name and every
   downstream scale, theme, and Excel style follows. No per-plot color
   arguments.
2. **Four palette classes.** Categorical, sequential, diverging, and
   semantic colors are first-class citizens, each matched to a visual
   job (grouping, magnitude, signed deviation, verdict).
3. **Accessibility by default.** Every preset is checked against the
   WCAG 2.x contrast formula; pass/fail verdicts always meet the AA
   threshold against their backgrounds.
4. **Extensibility.** Seven external presets (`economist`, `wsj`,
   `gdocs`, `tufte`, `few`, `solarized`, `prism`) are bundled as
   full configurations, and users can override individual slots or
   register entirely custom palettes.

# 2. Theory: Color in Quality Visualization

## 2.1 The four palette classes and their visual jobs

Statistical graphics map data to color in fundamentally different ways.
Using the wrong mapping produces misleading charts. The framework
formalises four palette classes, each tied to a data type:

| Palette class | Data type | Visual job | Quality example |
|:--------------|:----------|:-----------|:----------------|
| `discrete` | Categorical / factor | Distinguish groups | Operators, shifts, machines |
| `sequential` | Continuous, non-negative | Encode magnitude | Defect counts, density, risk scores |
| `diverging` | Continuous, signed | Encode deviation from a midpoint | Residuals, correlations, effects |
| `semantic` | Verdict / status | Encode meaning | Pass / fail / watch, significant / not |

**Discrete** palettes are qualitative — the colors have no inherent
order. Paul Tol's colorblind-safe palette (used in `academic`) is the
canonical choice. The framework auto-extends a discrete palette to any
`n` via `grDevices::colorRampPalette()` when the number of groups
exceeds the base length.

**Sequential** palettes run from a light endpoint through a saturated
midpoint to a dark endpoint, encoding "more is darker". A 3-stop
gradient (low / mid / high) is interpolated by `ggplot2::scale_*_gradientn`.

**Diverging** palettes have three meaningful stops: a negative hue, a
neutral midpoint (usually white or the background color), and a
positive hue. They are essential for signed quantities — using a
sequential palette on residuals would imply that "−3" and "+3" differ
only in magnitude, hiding the sign.

**Semantic** palettes are a small named set tied to domain verdicts.
`pass` / `fail` / `watch` are the canonical three; `good` / `bad` /
`neutral` are aliases for the same colors. Traffic-light roles
(`green` / `yellow` / `red`) are also exposed so non-statistical
audiences can read the chart immediately.

## 2.2 WCAG contrast and accessibility

The Web Content Accessibility Guidelines (WCAG) 2.x define a contrast
ratio between two colors:

$$
R = \frac{L_{\text{lighter}} + 0.05}{L_{\text{darker}} + 0.05}
$$

where $L$ is the relative luminance computed from the sRGB channels
via the linearisation:

$$
L = 0.2126 \cdot R_{\text{lin}} + 0.7152 \cdot G_{\text{lin}} + 0.0722 \cdot B_{\text{lin}}
$$

with $C_{\text{lin}} = C/12.92$ for $C \le 0.03928$ and
$((C+0.055)/1.055)^{2.4}$ otherwise. The ratio ranges from 1 (no
contrast) to 21 (black on white). The AA threshold is 4.5 for normal
text and 3.0 for large text; AAA requires 7.0.

The framework uses this formula in two places:

- `contrast_ratio(fg, bg)` — exported utility for ad-hoc checks.
- `.contrast_text(bg)` — `IqrPlotterBase` helper that returns
  `"#000000"` or `"#FFFFFF"` for a readable text color on a colored
  fill, used for in-bar labels.

## 2.3 Color-blind safety considerations

Approximately 8% of men and 0.5% of women have some form of color
vision deficiency, with deuteranopia (red-green) the most common.
Quality reports that encode pass/fail as red/green alone are
unreadable to this audience. The framework addresses this in two ways:

- The `academic` preset ships Paul Tol's colorblind-safe qualitative
  palette, which uses blue / red / green / yellow hues that remain
  distinguishable under deuteranopia and protanopia simulation.
- Semantic colors are always paired with a shape or text label, so
  the verdict is not communicated by hue alone.

# 3. The IqrTheme Facade

## 3.1 IqrTheme, ThemeConfig, PlotTheme — the three-layer model

The theme system is layered into three R6 classes, each with a single
responsibility:

| Layer | Class | Responsibility |
|:------|:------|:---------------|
| Configuration | `ThemeConfig` | Stores UI colors + four-class data palettes for one preset; resolves palettes via `get_pal()` |
| Scales | `PlotTheme` | Builds ggplot2 `scale_*` objects and the `theme_iqr()` ggplot2 theme from a `ThemeConfig` |
| Facade | `IqrTheme` | Single entry point; holds a `config` (`ThemeConfig`) and a `plot` (`PlotTheme`); delegates every call |

Users only ever construct an `IqrTheme`. Downstream packages receive
the `IqrTheme` object and call its `$plot$scale_*()` and
`$plot$theme_iqr()` methods, or — more commonly — call the
`.pal_*` / `.scale_*` toolbox on `IqrPlotterBase` (Section 6), which
internally delegates to the same `PlotTheme`.

## 3.2 Creating a theme: IqrTheme$new("academic")

```{r create-theme}
theme_academic <- IqrTheme$new("academic")
theme_workbench <- IqrTheme$new("workbench")

# The facade exposes both sub-objects
class(theme_academic$config)  # ThemeConfig
class(theme_academic$plot)    # PlotTheme

# Inspect the active UI palette
str(theme_academic$get_ui_colors(), max.level = 1)
```

## 3.3 Switching themes globally

There is no hidden global state inside `IqrTheme` — each instance is
self-contained. To switch the active theme for a session, pass a new
`IqrTheme` to your task object, or use the `options()` convention
recognised by some downstream helpers:

```{r switch-theme, eval = FALSE}
# Per-task: construct with a new preset
task$set_theme(IqrTheme$new("economist"))

# Global option (recognised by base_plot() in iQualityR.plot)
options(iqr.default_theme = "solarized")
```

The resolution order for palette lookups inside `ThemeConfig$get_pal()`
is:

1. The `custom` argument (per-call override).
2. `options("iqr.custom_<type>")` (session-wide custom palette).
3. The preset's `data[[type]]` slot.

This lets you override a single palette class without rebuilding the
whole theme.

# 4. Built-in Theme Presets

## 4.1 academic, workbench, tech (the three internal presets)

The three internal presets cover the three primary use contexts:

| Preset | Audience | Look-and-feel | Default in |
|:-------|:---------|:--------------|:-----------|
| `academic` | Papers, reports, CRAN | White background, black borders, Paul Tol discrete palette | `IqrTheme$new()` default |
| `workbench` | Internal QA dashboards | Soft grey background, blue primary, bold table headers | Production tooling |
| `tech` | Engineering dashboards | Light blue background, GitHub-inspired palette | Real-time monitoring |

Each preset ships a complete configuration: a `ui` list (background,
surface, text, grid, border, primary, success, warning, danger, and
four `table_*` slots) and a `data` list (the four palette classes).

```{r compare-internal}
# Compare the discrete palettes of the three internal presets
internal <- c("academic", "workbench", "tech")
disc <- lapply(internal, function(p) IqrTheme$new(p)$get_pal("discrete"))
names(disc) <- internal
disc
```

## 4.2 External themes: economist, wsj, gdocs, tufte, few, solarized, prism

Seven external presets are bundled as full configurations — not just
theme functions. Each carries its own UI colors and four-class data
palettes, so picking `"economist"` recolors plots and Excel exports
consistently with The Economist's look-and-feel.

When the source package (`ggthemes` or `ggprism`) is installed, the
matching ggplot2 theme function is layered on top of `theme_iqr()` for
typographic fidelity. When the package is not installed, the preset's
color configuration is still applied — only the theme function is
skipped, with no error.

| Preset | Source | Discrete palette origin |
|:-------|:-------|:------------------------|
| `economist` | ggthemes | Economist red/blue/green |
| `wsj` | ggthemes | Wall Street Journal |
| `gdocs` | ggthemes | Google Docs |
| `tufte` | ggthemes | Tufte minimal ink (grayscale) |
| `few` | ggthemes | Stephen Few's qualitative |
| `solarized` | ggthemes | Solarized accent palette |
| `prism` | ggprism | `winter_bright` |

```{r external-preset, eval = FALSE}
# Works whether or not ggthemes is installed; the theme function is
# only layered on top when the package is present.
theme_econ <- IqrTheme$new("economist")
theme_econ$get_pal("discrete")
theme_econ$plot$theme_iqr()
```

## 4.3 Side-by-side preset gallery (all 10 presets)

The snippet below builds a single bar chart rendered in all ten
presets, demonstrating how the same data recolors consistently. The
plot is assembled with `patchwork` (used by `iQualityR.plot`).

```{r preset-gallery, fig.width = 10, fig.height = 8}
# A small dataset: defect counts by category
defects <- data.frame(
  category = factor(c("Scratch", "Dent", "Crack", "Stain", "Burr")),
  count    = c(42, 28, 15, 10, 5)
)

presets <- c("academic", "workbench", "tech",
             "economist", "wsj", "gdocs", "tufte",
             "few", "solarized", "prism")

plots <- lapply(presets, function(p) {
  thm <- IqrTheme$new(p)
  ggplot(defects, aes(x = category, y = count, fill = category)) +
    geom_col() +
    thm$plot$theme_iqr(base_size = 9) +
    thm$plot$scale_fill_iqr(discrete = TRUE) +
    labs(title = p, x = NULL, y = NULL) +
    theme(legend.position = "none")
})

# Patchwork assembly (graceful fallback if patchwork is unavailable)
if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork::wrap_plots(plots, ncol = 2)
} else {
  plots[[1]]
}
```

# 5. The Four-Class Palette System

## 5.1 Discrete palettes and auto-extension via colorRampPalette

Discrete palettes are stored as a fixed-length character vector
(10 colors in every preset). When a chart needs more groups than the
base length, `ThemeConfig$get_pal(type = "discrete", n = N)` interpolates
via `grDevices::colorRampPalette`:

```{r discrete-extend}
theme <- IqrTheme$new("academic")
base <- theme$get_pal("discrete")
length(base)  # 10

extended <- theme$get_pal("discrete", n = 15)
length(extended)  # 15 — interpolated, not recycled
extended
```

This guarantees distinct colors for any number of groups without
recycling, while staying close to the preset's hue family.

## 5.2 Sequential and diverging gradients

Sequential and diverging palettes are stored as 3-stop gradients
(low / mid / high) and consumed by `ggplot2::scale_*_gradientn` /
`scale_*_gradient2`:

```{r gradient-palettes}
theme <- IqrTheme$new("workbench")
theme$get_pal("sequential")   # 3 stops: light -> mid -> dark
theme$get_pal("diverging")    # 3 stops: negative hue -> neutral -> positive hue
```

The diverging midpoint defaults to 0 in `.scale_fill_diverging()`,
which is correct for residuals, effects, and correlations. For a
non-zero midpoint (e.g. a target value), pass `midpoint = T` to the
scale.

## 5.3 Semantic colors: pass / fail / watch / good / bad / neutral

Semantic palettes are a named list of nine slots, organised in three
synonym groups:

| Group | Slots | Use case |
|:------|:------|:---------|
| Verdict | `pass`, `fail`, `watch` | Quality decisions |
| Valence | `good`, `bad`, `neutral` | Generic assessment |
| Traffic light | `green`, `yellow`, `red` | Stakeholder reporting |

All three groups map to the same three colors within a preset, so
`pass == good == green` by design. This lets domain code use the most
idiomatic name without color drift.

```{r semantic-palette}
theme <- IqrTheme$new("academic")
sem <- theme$get_pal("semantic")
sem

# Single-name lookup
theme$get_pal("semantic", name = "pass")
theme$get_pal("semantic", name = "fail")
```

## 5.4 Paired mode: coordinated fill + outline

For bar charts and boxplots, a fill color needs a darker outline to
keep edges visible against adjacent fills. The `style = "paired"`
argument to `scale_fill_iqr()` / `scale_color_iqr()` derives the
outline color by darkening each fill by 25% via the `darken()` utility:

```{r paired-mode, fig.width = 9, fig.height = 4}
theme <- IqrTheme$new("tech")
df <- data.frame(
  group = factor(c("A", "B", "C", "D")),
  value = c(12, 18, 15, 9)
)

p_paired <- ggplot(df, aes(x = group, y = value, fill = group, colour = group)) +
  geom_col(linewidth = 1) +
  theme$plot$theme_iqr() +
  theme$plot$scale_fill_iqr(discrete = TRUE, style = "paired") +
  theme$plot$scale_color_iqr(discrete = TRUE, style = "paired") +
  labs(title = "Paired fill + darkened outline", x = NULL, y = NULL) +
  theme(legend.position = "none")

p_paired
```

# 6. The IqrPlotterBase Color Toolbox

## 6.1 .pal_*() accessors

`IqrPlotterBase` exposes five palette accessors and one text-color
helper. They are the only sanctioned way for a subclass to read colors
from the active theme — directly indexing `theme_obj$config$config` is
discouraged because the internal layout may evolve.

| Method | Returns | Notes |
|:-------|:--------|:------|
| `.pal_discrete(theme_obj, n = NULL)` | `character` | Auto-extends to `n` |
| `.pal_sequential(theme_obj)` | `character` (3 stops) | |
| `.pal_diverging(theme_obj)` | `character` (3 stops) | |
| `.pal_semantic(theme_obj, name = NULL)` | `character` (named) or scalar | Single color when `name` given |
| `.pal_ui(theme_obj, name, default = "#888888")` | scalar | UI slot fallback |
| `.contrast_text(bg)` | `"#000000"` or `"#FFFFFF"` | WCAG-based |

## 6.2 .scale_*() factories

The nine scale factories return ready-to-add ggplot2 scale objects.
They are thin wrappers around `PlotTheme$scale_*` so that subclasses
do not need to call `theme_obj$plot$...` explicitly.

| Factory | Returns | Use when |
|:--------|:--------|:---------|
| `.scale_fill_discrete(theme, ...)` | `ScaleDiscrete` | Bars/boxes by category |
| `.scale_color_discrete(theme, ...)` | `ScaleDiscrete` | Lines/points by category |
| `.scale_fill_color_paired(theme, ...)` | `list(fill, color)` | Coordinated fill + outline |
| `.scale_fill_sequential(theme, ...)` | `ScaleContinuous` | Density / count gradient |
| `.scale_color_sequential(theme, ...)` | `ScaleContinuous` | Sequential point color |
| `.scale_fill_diverging(theme, ...)` | `ScaleContinuous` | Residual / effect heatmap |
| `.scale_color_diverging(theme, ...)` | `ScaleContinuous` | Signed point color |
| `.scale_fill_semantic(theme, labels, ...)` | `ScaleDiscrete` | Pass/fail verdicts |
| `.scale_color_semantic(theme, labels, ...)` | `ScaleDiscrete` | Verdict outlines |

## 6.3 .contrast_text() for accessible labels

When a plot places text on a colored fill (e.g. a count label inside a
bar), the text color must contrast with the fill. `.contrast_text()`
wraps `is_dark()` and returns black or white:

```{r contrast-text}
plotter <- IqrPlotterBase$new()
plotter$.contrast_text("#2563EB")  # dark fill -> white text
plotter$.contrast_text("#F7F9FC")  # light fill -> black text
```

## 6.4 Worked example: a semantic verdict bar chart

This example uses the toolbox end-to-end — semantic palette, semantic
scale, contrast text — to render a pass/fail/watch summary:

```{r toolbox-example, fig.width = 7, fig.height = 4}
verdicts <- data.frame(
  result = factor(c("pass", "watch", "fail"),
                  levels = c("pass", "watch", "fail")),
  count  = c(120, 18, 7)
)

theme <- IqrTheme$new("workbench")

# A minimal plot that uses the toolbox end-to-end: semantic fill scale,
# contrast-checked text color, and the themed ggplot2 theme.
verdict_plot <- function(df, theme_obj) {
  txt_col <- theme_obj$get_ui_colors()$text
  ggplot(df, aes(x = result, y = count, fill = result)) +
    geom_col() +
    geom_text(aes(label = count), vjust = -0.3, colour = txt_col) +
    theme_obj$plot$theme_iqr() +
    theme_obj$plot$scale_fill_semantic(labels = c("pass", "watch", "fail")) +
    labs(title = "Capability verdicts (n = 145 parts)",
         x = NULL, y = "Count") +
    theme(legend.position = "none") +
    ylim(0, 140)
}

verdict_plot(verdicts, theme)
```

# 7. Color Utility Functions

The package exports five dependency-free color helpers. They are used
internally by the theme system (e.g. `darken()` powers `style = "paired"`)
and are exported so downstream packages and users can build on the same
primitives.

## 7.1 lighten(), darken(), mix()

```{r color-utils-1}
base <- "#2563EB"

lighten(base, 0.2)   # mix with white
darken(base, 0.3)    # mix with black
mix("#FF0000", "#0000FF", 0.5)  # 50/50 red + blue -> purple
```

The `amount` parameter is a proportion in `[0, 1]`: 0 returns the
original color, 1 returns the mixing endpoint (white / black / the
second color).

## 7.2 is_dark(), contrast_ratio()

```{r color-utils-2}
is_dark("#000000")   # TRUE
is_dark("#FFFFFF")   # FALSE
is_dark("#2563EB")   # TRUE — blue is dark enough for white text

contrast_ratio("#FFFFFF", "#000000")  # 21 — maximum
contrast_ratio("#2563EB", "#FFFFFF")  # check the blue/white pair
```

A ratio of 4.5 or higher meets WCAG AA for normal text; 7.0 meets AAA.

## 7.3 Worked example: deriving a palette from a brand color

A common request is "use our corporate blue everywhere". The utilities
make this a five-line derivation:

```{r brand-palette, fig.width = 7, fig.height = 3}
brand_blue <- "#0B5394"  # hypothetical corporate color

# Build a 5-color qualitative palette by mixing the brand color with
# white at decreasing amounts, then darkening for the last two stops.
brand_discrete <- c(
  brand_blue,
  lighten(brand_blue, 0.3),
  lighten(brand_blue, 0.6),
  darken(brand_blue, 0.2),
  darken(brand_blue, 0.4)
)

# Build a sequential 3-stop gradient
brand_sequential <- c(lighten(brand_blue, 0.85), brand_blue, darken(brand_blue, 0.4))

# Visualise
df <- data.frame(x = factor(1:5), y = 1:5)
ggplot(df, aes(x = x, y = y, fill = x)) +
  geom_col() +
  scale_fill_manual(values = brand_discrete) +
  labs(title = "Brand-derived discrete palette", x = NULL, y = NULL) +
  theme_minimal() +
  theme(legend.position = "none")
```

To plug this into the theme system, override the `discrete` slot at
construction (Section 8.1) or via `options()`:

```{r brand-option, eval = FALSE}
options(iqr.custom_discrete = brand_discrete)
# Every IqrTheme now returns brand_discrete for type = "discrete"
# until the option is cleared.
```

# 8. Customizing and Extending Themes

## 8.1 Overriding individual slots

`ThemeConfig$initialize()` and `$set_theme()` accept `...` overrides
that are merged into the preset via `merge_config()` (a recursive
`modifyList`). This lets you tweak one slot without rebuilding the
whole configuration:

```{r override-slots}
# Override the primary color and the discrete palette in one call
theme_custom <- IqrTheme$new(
  "academic",
  ui = list(primary = "#0B5394"),
  data = list(discrete = c("#0B5394", "#E69138", "#6AA84F",
                            "#CC0000", "#674EA7", "#999999",
                            "#3D85C6", "#B45309", "#38761D", "#7F6000"))
)

theme_custom$get_ui_colors()$primary
theme_custom$get_pal("discrete")
```

## 8.2 Registering a custom preset

For a reusable preset, add an entry to `ThemeConfig$style_presets` at
session start. The simplest path is to copy an existing preset and
modify it:

```{r register-preset, eval = FALSE}
# Build a "brand" preset from "workbench"
brand_preset <- IqrTheme$new("workbench")$config$style_presets$workbench
brand_preset$ui$primary <- "#0B5394"
brand_preset$data$discrete <- brand_discrete

# Register it on a fresh ThemeConfig
tc <- ThemeConfig$new("academic")
tc$style_presets$brand <- brand_preset
tc$set_theme("brand")

# Now use it through the facade
theme_brand <- IqrTheme$new("academic")
theme_brand$config$style_presets$brand <- brand_preset
theme_brand$set_theme("brand")
```

For a permanent preset, ship it as a new list inside the
`style_presets` assignment in `IqrTheme.R`. The framework will pick it
up automatically.

## 8.3 Feeding a custom palette through the toolbox

Once a custom palette is registered, every `.pal_*` and `.scale_*`
helper picks it up — no subclass changes required:

```{r custom-through-toolbox, fig.width = 6, fig.height = 4}
theme <- IqrTheme$new(
  "academic",
  data = list(discrete = c("#0B5394", "#E69138", "#6AA84F"))
)

df <- data.frame(group = factor(c("A", "B", "C")), value = c(10, 14, 8))
ggplot(df, aes(x = group, y = value, fill = group)) +
  geom_col() +
  theme$plot$theme_iqr() +
  theme$plot$scale_fill_iqr(discrete = TRUE) +
  labs(title = "Custom 3-color palette via override", x = NULL, y = NULL) +
  theme(legend.position = "none")
```

# 9. ExcelExporter Theming

## 9.1 How workbook colors track the active theme

`ExcelExporter` is constructed from a `ThemeConfig` (not the full
`IqrTheme`) and derives every workbook style from the theme's UI
palette. This guarantees that an Excel export matches the plots that
accompany it in a report.

The styles are generated once at construction by
`$generate_excel_styles()` and cached in `$excel_styles_cache`. The
cache is a list of `openxlsx::createStyle()` objects:

| Style slot | Used for | Theme source |
|:-----------|:---------|:-------------|
| `main_title` | Merged title row | `ui$title` |
| `header` | Column header row | `ui$table_header_bg`, `ui$table_header_tx` |
| `body` | Data rows | `ui$text`, `ui$grid` |
| `stripe` | Zebra-striped even rows | `ui$table_stripe` |
| `t_thick` / `b_thick` / `l_thick` / `r_thick` | Outer border | `ui$table_border` |

## 9.2 Customizing table headers, stripes, borders

For a one-off export that needs different styling (e.g. a customer
brand color on the header), call `$set_excel_theme(...)` with named UI
overrides before exporting. `$reset_excel_theme()` restores the
ThemeConfig defaults.

```{r excel-theme, eval = FALSE}
theme    <- IqrTheme$new("academic")
exporter <- ExcelExporter$new(theme$config)

# Override just the header background for this export
exporter$set_excel_theme(table_header_bg = "#0B5394",
                         table_header_tx = "#FFFFFF")

exporter$export_excel(
  data  = data.frame(metric = "Cpk", value = 1.42),
  path  = tempfile(fileext = ".xlsx"),
  title = "Bearing Diameter Capability (branded)"
)

# Restore defaults for subsequent exports
exporter$reset_excel_theme()
```

## 9.3 A complete themed export

This end-to-end example builds a small workbook whose colors track the
`economist` preset:

```{r excel-export-demo, eval = FALSE}
theme    <- IqrTheme$new("economist")
exporter <- ExcelExporter$new(theme$config)

sheets <- list(
  Summary = data.frame(
    metric = c("mean", "sd", "Cpk", "verdict"),
    value  = c(10.021, 0.034, 1.42, "pass")
  ),
  Subgroups = data.frame(
    subgroup = 1:3,
    mean     = c(10.01, 10.03, 10.02),
    n        = c(5, 5, 5)
  )
)

path <- exporter$export_excel(
  data  = sheets,
  path  = tempfile(fileext = ".xlsx"),
  title = "Bearing Diameter Capability Study"
)
```

# 10. Best Practices

## 10.1 Resolve the theme once

Construct the `IqrTheme` at the task level and pass it down to every
plotter and reporter. Re-resolving the theme inside `render()` defeats
the "one switch" goal and can cause inconsistent colors when the global
option changes mid-pipeline.

## 10.2 Use the toolbox, not the internals

Inside a plotter subclass, call `.pal_*()` and `.scale_*()` — never
index `theme_obj$config$config$data[[...]]` directly. The toolbox is
the stable API; the internal layout of `ThemeConfig` may evolve.

## 10.3 Match the palette class to the data

- Categorical factor -> `discrete`
- Continuous non-negative -> `sequential`
- Continuous signed -> `diverging`
- Verdict / status -> `semantic`

Using a discrete palette on a continuous variable produces a legend
with one entry per unique value — almost never what you want.

## 10.4 Keep semantic colors redundant

Always pair `pass` / `fail` color with a text label or shape. Color
alone is inaccessible to color-blind users and unparseable by
black-and-white printers.

## 10.5 Check contrast for in-bar labels

When placing text on a colored fill, route the fill color through
`.contrast_text()` to pick black or white text. Hard-coding either
will fail on some fills.

## 10.6 Override at the right granularity

- Per-call: pass `custom` to `get_pal()` or `values` to a scale.
- Per-session: set `options("iqr.custom_<type>")`.
- Per-project: construct `IqrTheme$new(preset, data = list(...))`.
- Per-organisation: register a new preset in `style_presets`.

Prefer the narrowest scope that solves the problem.

# 11. References

- Ihaka, R. (2003). *Colour for Presentation Graphics*. Proceedings of
  the 3rd International Workshop on Distributed Statistical Computing.
- Lumley, T. (2006). *colorspace: Color Space Manipulation*. R package.
- Montgomery, D. C. (2013). *Statistical Quality Control* (7th ed.).
  Wiley.
- Paul Tol's Technical Note "Colour Schemes" (SRON Technical Note).
  https://personal.sron.nl/~pault/
- W3C (2018). *Web Content Accessibility Guidelines (WCAG) 2.1*.
  https://www.w3.org/TR/WCAG21/
- Wickham, H. (2016). *ggplot2: Elegant Graphics for Data Analysis*
  (2nd ed.). Springer.
- Zeileis, A., Hornik, K., and Murrell, P. (2009). *Escaping RGBland:
  Selecting Colors for Statistical Graphics*. Computational Statistics
  & Data Analysis, 53, 3259-3270.

*See `vignette("iQualityR-core", package = "iQualityR.core")` for the
R6 base class architecture, and `vignette("iQualityR", package =
"iQualityR")` for the framework-level methodology overview.*
