---
title: "Attribute Agreement Analysis: Kappa, Wagner and Kendall Statistics"
author: "iQualityR MSA Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Attribute Agreement Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

library(iQualityR.msa)

fmt_num <- function(x, digits = 4) {
  x <- suppressWarnings(as.numeric(x))
  if (length(x) == 0 || all(is.na(x))) return("NA")
  formatC(x, format = "f", digits = digits)
}
fmt_pct <- function(x, digits = 2) {
  x <- suppressWarnings(as.numeric(x))
  if (length(x) == 0 || all(is.na(x))) return("NA")
  paste0(formatC(100 * x, format = "f", digits = digits), "%")
}
fmt_p <- function(x, digits = 4) {
  x <- suppressWarnings(as.numeric(x))
  if (length(x) == 0 || all(is.na(x))) return("NA")
  if (x < 10^-digits) return(paste0("<", formatC(10^-digits, format = "f", digits = digits)))
  formatC(x, format = "f", digits = digits)
}
round_df <- function(x, digits = 4) {
  x <- as.data.frame(x)
  num_cols <- vapply(x, is.numeric, logical(1))
  x[num_cols] <- lapply(x[num_cols], round, digits = digits)
  x
}
```

## 1. Business Context

A textile dyeing company produces printed fabric rolls. Before each roll is released to the customer, a team of visual inspectors rates the print quality on an **ordinal 1–5 scale**, where `1` represents the worst print quality and `5` represents the best. The quality engineering team needs to validate that the inspection system — the four inspectors combined — produces consistent and accurate ratings.

The study follows the classical Minitab *Attribute Agreement Analysis* example design:

| Parameter | Value |
|---|---|
| Inspectors (appraisers) | Amanda, Britt, Eric, Mike |
| Fabric samples | 50 |
| Trials per inspector | 2 |
| Total rating events | 4 × 50 × 2 = 400 |
| Rating scale | Ordinal, 1 (worst) — 5 (best) |
| Known standard | Yes, one engineering master rating per sample |
| Kappa method | Fleiss (multiple raters, long format) |
| Scale type | Ordinal (enables Kendall statistics) |

Each inspector rates every sample twice, so the study supports three complementary questions:

1. **Within-appraiser repeatability** — does the same inspector give the same rating on trial 1 and trial 2?
2. **Between-appraiser agreement** — do different inspectors agree with each other on the same sample?
3. **Appraiser-vs-standard accuracy** — do the inspectors' ratings agree with the known engineering standard?

Because the scale is ordinal, the analysis reports both Kappa statistics (exact agreement) and Kendall statistics (ordered association), giving a complete picture of inspection system capability.

## 2. Method and Formulas

### Observed and expected agreement

Let $n$ be the number of comparison units and let $r_i$ and $c_j$ be the row and column marginals of the $n \times n$ classification table.

$$
P_o = \frac{\#\text{ matches}}{\#\text{ comparisons}}, \qquad
P_e = \frac{\sum_i r_i \, c_j}{n^2}
$$

### Cohen's kappa (two raters)

$$
\kappa = \frac{P_o - P_e}{1 - P_e}
$$

Cohen's kappa applies to exactly two raters (or two rating streams) classifying the same set of subjects. The standard error is

$$
\mathrm{SE}(\kappa) = \sqrt{\frac{P_o(1 - P_o)}{n(1 - P_e)^2}},
$$

and the test statistic $Z = \kappa / \mathrm{SE}(\kappa)$ is evaluated against the standard normal distribution.

### Fleiss' kappa (multiple raters)

For $k$ raters each classifying $n$ subjects into $c$ categories, let $n_{ij}$ be the count of raters assigning subject $i$ to category $j$, and $p_j = \sum_i n_{ij} / (nk)$ the overall proportion of category $j$.

$$
P_o = \frac{1}{n}\sum_{i=1}^{n}\frac{\sum_j n_{ij}^2 - n_{i\cdot}}{n_{i\cdot}(n_{i\cdot}-1)},
\qquad
P_e = \sum_{j=1}^{c} p_j^2,
\qquad
\kappa = \frac{P_o - P_e}{1 - P_e}
$$

### Kendall's W (concordance coefficient)

For ordinal data, Kendall's $W$ measures how consistently $k$ raters rank $n$ subjects:

$$
W = \frac{12 \sum_{i=1}^{n}(R_i - \bar R)^2}{k^2(n^3 - n) - k \sum T}
$$

where $R_i$ is the sum of ranks for subject $i$ and $T$ is a tie correction. $W$ ranges from 0 (no concordance) to 1 (perfect concordance).

### Kendall's tau-b (appraiser vs standard)

Kendall's tau-b measures the rank correlation between each appraiser's ordered ratings and the known standard:

$$
\tau_b = \frac{n_c - n_d}{\sqrt{(n_0 - n_1)(n_0 - n_2)}}
$$

where $n_c$ and $n_d$ are concordant and discordant pairs, and $n_1, n_2$ are tie corrections.

### Wagner two-way method

The Wagner method compares two raters directly. $V$ is the concordance proportion (exact match rate) and $BP$ is the mean absolute distance between ratings:

$$
V = \frac{\#\text{ concordant}}{n}, \qquad
BP = \frac{1}{n}\sum_{i=1}^{n}|x_i - y_i|
$$

### Kappa interpretation bands

| Kappa range | Interpretation |
|---|---|
| $< 0$ | Poor |
| $0.00$ — $0.19$ | Slight |
| $0.20$ — $0.39$ | Fair |
| $0.40$ — $0.59$ | Moderate |
| $0.60$ — $0.74$ | Good |
| $0.75$ — $0.89$ | Strong |
| $\geq 0.90$ | Excellent |

**Acceptance criteria:** $\kappa \geq 0.90$ is strong enough for system release; $\kappa \geq 0.75$ is good and may be conditionally accepted depending on risk. For the Wagner method, $V \geq 0.75$ is Acceptable, $0.50 \leq V < 0.75$ is Conditionally Acceptable, and $V < 0.50$ is Unacceptable.

## 3. Data Preparation

The fabric print quality data ships with the package as a UTF-8 CSV file. It contains 400 rows and 5 columns with Chinese headers that we rename to English for programmatic access.

```{r load-data}
fabric_path <- system.file("extdata/minitab_msa/织物印刷质量.csv", package = "iQualityR.msa")
fabric_data <- read.csv(fabric_path, fileEncoding = "UTF-8")
names(fabric_data) <- c("Appraiser", "Rating", "Sample", "Trial", "Standard")

dim(fabric_data)
head(fabric_data, 12)
```

```{r data-summary}
data.frame(
  Appraisers = paste(sort(unique(fabric_data$Appraiser)), collapse = ", "),
  Samples = length(unique(fabric_data$Sample)),
  Trials = length(unique(fabric_data$Trial)),
  Rating_Levels = paste(sort(unique(fabric_data$Rating)), collapse = ", "),
  Total_Rows = nrow(fabric_data),
  Has_Standard = all(!is.na(fabric_data$Standard))
)
```

## 4. Quick Start (One-Liner)

For a multi-rater ordinal study in long format, `iqr_fleiss_kappa()` runs the full analysis in a single call — no explicit plan object required.

```{r quick-start}
task <- iqr_fleiss_kappa(
  fabric_data,
  sample_col = "Sample",
  rater_col = "Appraiser",
  rating_col = "Rating",
  standard_col = "Standard",
  trial_col = "Trial",
  scale_type = "ordinal",
  category_order = as.character(1:5)
)

task$summary()
```

The returned `task` object holds the complete kappa and Kendall results. The rest of this vignette walks through the same analysis using the explicit Plan/Task workflow and then drills into each output table and plot.

## 5. Step-by-Step Workflow

The explicit workflow gives full control over study parameters and produces a reproducible protocol. The flow is: **Plan → validate → protocol → Task → compute → summary → plot → report**.

```{r step-plan}
plan <- AttrGagePlan$new(
  plan_name = "Fabric Print Quality Inspection",
  objectives = "Validate visual inspection agreement on ordinal 1-5 print quality scale",
  appraisers = c("Amanda", "Britt", "Eric", "Mike"),
  samples = 1:50,
  standards = NULL,
  categories = as.character(1:5),
  category_order = as.character(1:5),
  scale_type = "ordinal",
  trials = 2,
  comparison_mode = "one_way",
  kappa_method = "fleiss",
  conf_level = 0.95
)

plan$set_meta(
  "data",
  sample_col = "Sample",
  rater_col = "Appraiser",
  rating_col = "Rating",
  standard_col = "Standard",
  trial_col = "Trial"
)

plan$validate()
plan$generate_protocol()
```

```{r step-task}
task <- AttrGageTask$new(
  data = fabric_data,
  plan = plan,
  theme = "academic",
  mode = "kappa",
  kappa_method = "fleiss"
)

task$compute()
task$summary()
```

## 6. Cohen Kappa (Two Raters)

Cohen's kappa is the right tool when exactly two raters classify the same subjects. To demonstrate it, we subset the fabric data to **Amanda vs Britt** on trial 1 and reshape to wide format.

```{r cohen-data}
ab_trial1 <- fabric_data[
  fabric_data$Appraiser %in% c("Amanda", "Britt") & fabric_data$Trial == 1,
  c("Sample", "Appraiser", "Rating")
]
cohen_wide <- reshape(ab_trial1, idvar = "Sample", timevar = "Appraiser", direction = "wide")
names(cohen_wide) <- sub("^Rating\\.", "", names(cohen_wide))
head(cohen_wide, 10)
```

```{r cohen-run}
cohen_task <- iqr_cohen_kappa(
  cohen_wide,
  eval1_col = "Amanda",
  eval2_col = "Britt",
  theme = "academic"
)

cohen_task$summary()

ck <- cohen_task$kappa_results$raw_output
data.frame(
  Statistic = c("Kappa", "SE", "Z", "P(>0)", "Po", "Pe", "Interpretation"),
  Value = c(
    fmt_num(ck$kappa, 4),
    fmt_num(ck$se, 4),
    fmt_num(ck$z, 4),
    fmt_p(ck$p_value, 4),
    fmt_pct(ck$Po, 2),
    fmt_pct(ck$Pe, 2),
    ck$interpretation
  )
)

knitr::kable(
  round_df(ck$response_table, 4),
  caption = "Response-level agreement: Amanda vs Britt (Cohen)"
)
```

```{r cohen-plot}
print(cohen_task$plot(type = "kappa_benchmark"))
```

## 7. Wagner Two-Way Method

The Wagner method evaluates two raters through the concordance proportion $V$ and the bias proportion $BP$. It is especially useful for ordinal scales because $BP$ captures the *magnitude* of disagreement, not just whether ratings match exactly.

```{r wagner-run}
wagner_task <- iqr_wagner(
  cohen_wide,
  eval1_col = "Amanda",
  eval2_col = "Britt",
  theme = "academic"
)

wagner_task$summary()

wk <- wagner_task$kappa_results$raw_output
data.frame(
  Metric = c("V (concordance)", "BP (bias)", "N", "Concordant", "SE_V", "Z", "P(>0)", "Grade"),
  Value = c(
    fmt_pct(wk$V, 4),
    fmt_num(wk$BP, 4),
    wk$n,
    wk$concordant,
    fmt_num(wk$se_V, 4),
    fmt_num(wk$z, 4),
    fmt_p(wk$p_value, 4),
    wk$grade
  )
)
```

The Wagner grade thresholds are:

- $V \geq 0.75$ — **Acceptable**
- $0.50 \leq V < 0.75$ — **Conditionally Acceptable**
- $V < 0.50$ — **Unacceptable**

```{r wagner-plot}
print(wagner_task$plot(type = "summary"))
```

## 8. Fleiss Kappa Deep-Dive

The Fleiss kappa analysis produces a rich set of tables that answer every agreement question in the study. All tables below are extracted from `task$kappa_results$raw_output`.

### Overall kappa

```{r fleiss-overall}
kr <- task$kappa_results$raw_output
ci <- kr$ci

data.frame(
  Statistic = c(
    "Method", "Scale type", "Kappa", "SE", "Z", "P(>0)",
    "95% CI lower", "95% CI upper",
    "Observed agreement (Po)", "Expected agreement (Pe)",
    "N subjects", "N raters per subject", "N appraisers", "N categories",
    "Interpretation"
  ),
  Value = c(
    kr$method,
    kr$scale_type,
    fmt_num(kr$kappa, 4),
    fmt_num(kr$se, 4),
    fmt_num(kr$z, 4),
    fmt_p(kr$p_value, 4),
    fmt_num(ci[1], 4),
    fmt_num(ci[2], 4),
    fmt_pct(kr$Po, 2),
    fmt_pct(kr$Pe, 2),
    kr$n_subjects,
    kr$n_raters,
    kr$n_appraisers,
    kr$n_categories,
    kr$interpretation
  )
)
```

### Each appraiser vs standard

This table compares every inspector against the known engineering master. A low kappa here flags an inspector who needs retraining.

```{r fleiss-vs-standard}
knitr::kable(
  round_df(kr$appraiser_vs_standard, 4),
  caption = "Each appraiser vs known standard: matches, percent agreement, kappa"
)
```

### Between-appraiser agreement

Pairwise agreement across all appraiser combinations, averaged over trials.

```{r fleiss-pairwise}
knitr::kable(
  round_df(kr$pairwise_appraisers, 4),
  caption = "Between-appraiser pairwise agreement"
)
```

### Within-appraiser repeatability

Each appraiser's trial-1 vs trial-2 self-consistency.

```{r fleiss-within}
knitr::kable(
  round_df(kr$within_appraiser, 4),
  caption = "Within-appraiser repeatability across two trials"
)
```

### Response-level kappa

Each rating level (1–5) is collapsed into a binary "this level vs all others" question and kappa is recomputed. This reveals whether a particular rating level is unstable even when the overall kappa looks acceptable.

```{r fleiss-response-kappa}
knitr::kable(
  round_df(kr$response_kappa, 4),
  caption = "Response-level Fleiss kappa (each level vs all others)"
)

knitr::kable(
  round_df(kr$response_kappa_vs_standard, 4),
  caption = "Response-level kappa vs standard"
)
```

### Sample disagreement table (top disputed samples)

Samples are ranked by the number of discordant ratings, making it easy to identify borderline fabric rolls that drive inspector disagreement.

```{r fleiss-sample-disagreement}
knitr::kable(
  round_df(head(kr$sample_disagreement, 15), 4),
  caption = "Top 15 samples by disagreement count"
)
```

## 9. Kendall Ordinal Statistics

Kappa measures *exact* agreement: a rating of 4 vs 5 counts as a complete mismatch, even though the two levels are adjacent on an ordinal scale. For ordinal data this is overly strict. Kendall's $W$ (concordance coefficient) and Kendall's tau-b (rank correlation) use the *order* of the categories, so a 4-vs-5 disagreement is penalised far less than a 1-vs-5 disagreement.

These outputs appear only when `scale_type = "ordinal"` is set on the plan.

### Kendall's W — between appraisers

```{r kendall-between}
knitr::kable(
  round_df(kr$ordinal$between_appraisers, 4),
  caption = "Kendall's coefficient of concordance (between appraisers)"
)
```

### Kendall's W — within appraiser

```{r kendall-within}
knitr::kable(
  round_df(kr$ordinal$within_appraiser, 4),
  caption = "Kendall's W within each appraiser (trial-to-trial concordance)"
)
```

### Kendall's tau-b — each appraiser vs standard

```{r kendall-vs-standard}
knitr::kable(
  round_df(kr$ordinal$appraiser_vs_standard, 4),
  caption = "Kendall's tau-b: each appraiser vs known standard"
)
```

### Kendall's tau-b — all appraisers vs standard

```{r kendall-all-vs-standard}
knitr::kable(
  round_df(kr$ordinal$all_appraisers_vs_standard, 4),
  caption = "Kendall's tau-b: all appraisers pooled vs known standard"
)
```

## 10. Visualization Gallery

The `AttrGagePlotter` supports the following plot types. Each is rendered via `print(task$plot(type = "..."))`.

```{r plot-catalogue}
plot_catalogue <- data.frame(
  Plot_Type = c(
    "summary", "kappa_benchmark", "pairwise_agreement",
    "appraiser_standard", "kendall_ordinal", "response_agreement",
    "sample_disagreement", "confusion_matrix"
  ),
  Purpose = c(
    "Combined overview panel of all available kappa graphics",
    "Overall kappa value against interpretation bands",
    "Between-appraiser pairwise kappa and percent agreement",
    "Each appraiser vs known standard (kappa and match rate)",
    "Kendall W and tau-b coefficients for ordinal studies",
    "Response-level match-to-standard rate by rating level",
    "Top disputed samples ranked by discordant rating count",
    "Rating count matrix or rater-vs-rater confusion matrix"
  )
)
knitr::kable(plot_catalogue, caption = "Available attribute agreement plot types")
```

```{r plot-summary, fig.height=6}
print(task$plot(type = "summary"))
```

```{r plot-kappa-benchmark}
print(task$plot(type = "kappa_benchmark"))
```

```{r plot-pairwise}
print(task$plot(type = "pairwise_agreement"))
```

```{r plot-appraiser-standard}
print(task$plot(type = "appraiser_standard"))
```

```{r plot-kendall}
print(task$plot(type = "kendall_ordinal"))
```

```{r plot-response}
print(task$plot(type = "response_agreement"))
```

```{r plot-sample-disagreement}
print(task$plot(type = "sample_disagreement"))
```

```{r plot-confusion}
print(task$plot(type = "confusion_matrix"))
```

## 11. Report Export

The task exports a self-contained report in Excel, HTML, or PDF format. The Excel workbook contains one sheet per result table (overall kappa, appraiser-vs-standard, pairwise, within-appraiser, response-level, sample disagreement, Kendall tables, and raw data).

```{r report-export}
xlsx_path <- tempfile("fabric_attribute_agreement_", fileext = ".xlsx")
html_path <- tempfile("fabric_attribute_agreement_", fileext = ".html")

task$report(format = "excel", path = xlsx_path)
task$report(format = "html", path = html_path)

data.frame(
  Format = c("Excel", "HTML"),
  Path = c(xlsx_path, html_path),
  Exists = c(file.exists(xlsx_path), file.exists(html_path)),
  Size_KB = round(c(file.info(xlsx_path)$size, file.info(html_path)$size) / 1024, 1)
)
```

## 12. Interpretation and Decision

Synthesising the tables above, the engineering conclusion for the fabric print quality inspection system is:

**Within-appraiser repeatability.** All four inspectors show high trial-to-trial consistency, with kappa values well above the 0.75 acceptance threshold. This means each inspector is internally stable when re-rating the same fabric sample.

**Between-appraiser agreement.** Pairwise kappa values between inspector pairs are in the Strong range (0.75–0.89) or above, confirming that different inspectors apply the 1–5 scale consistently.

**Appraiser vs standard.** This is the most demanding test. **Amanda** shows the strongest agreement with the known standard ($\kappa \approx 1.0$, near-perfect), while **Eric** is the weakest ($\kappa \approx 0.82$, still in the Strong band). All four inspectors exceed the 0.75 acceptance threshold, and the pooled appraiser-vs-standard kappa confirms system-level accuracy.

**Kendall ordinal statistics.** Kendall's W between appraisers and Kendall's tau-b vs standard both fall in the 0.95–0.98 range, indicating that even when inspectors disagree on the exact integer rating, their *ordering* of fabric quality is almost perfectly aligned with the engineering standard. This is critical for a 1–5 ordinal scale where a one-level difference (e.g., 4 vs 5) is operationally minor.

**Overall decision.** The inspection system is **acceptable** for release. All inspectors demonstrate good-to-strong agreement with the standard, Kendall coefficients confirm strong ordinal consistency, and no individual rating level shows instability. Eric's slightly lower kappa suggests a minor calibration opportunity, but it does not block system acceptance.

## 13. Reference

- Minitab Help, *Attribute Agreement Analysis* — kappa statistics, appraiser agreement tables, and Kendall coefficients.
- AIAG, *Measurement Systems Analysis (MSA)*, 4th edition, Chapter III — Attribute Measurement Systems Study.
- Cohen, J. (1960). "A coefficient of agreement for nominal scales." *Educational and Psychological Measurement*, 20(1), 37–46.
- Fleiss, J. L. (1971). "Measuring nominal scale agreement among many raters." *Psychological Bulletin*, 76(5), 378–382.
- Wagner, C. H. (1982). "Simpson's paradox in real life." *The American Statistician* — two-way concordance method for paired raters.
- Kendall, M. G. (1948). *Rank Correlation Methods*. Charles Griffin, London.
