---
title: "Nested Gage R&R Study for Destructive Testing"
author: "iQualityR MSA Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Nested Gage R&R Study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

library(iQualityR.msa)
library(magrittr)

# Formatting helpers: 4 / 2 decimal places, and percentage formatting
fmt4 <- function(x) {
  ifelse(is.na(x), "NA", formatC(as.numeric(x), format = "f", digits = 4))
}
fmt2 <- function(x) {
  ifelse(is.na(x), "NA", formatC(as.numeric(x), format = "f", digits = 2))
}
fmt_pct <- function(x, digits = 2) {
  ifelse(is.na(x), "NA", paste0(formatC(as.numeric(x), format = "f", digits = digits), "%"))
}
```

## 1. Business Context

A Nested Gage R&R Study is the appropriate design when the act of measuring
destroys or fundamentally alters the part, so that no part can be measured
more than once. Classic examples include tensile strength, impact energy,
burst pressure, and destructive peel tests. Because the same physical part
cannot be re-measured by a different operator, the crossed Gage R&R
structure -- every operator measures every part -- is impossible. Instead,
each operator receives a *different* set of parts drawn from the same
homogeneous batch or production stream. Parts are therefore **nested within
operators** rather than crossed with them.

This vignette reproduces the classical Minitab *Nested Gage R&R Study*
example for ceramic components measured for breaking strength. The quality
engineer selects 3 operators and asks each operator to measure 10 distinct
ceramic components sampled from the same production batch, with each
component measured twice before destruction (two replications per
part-operator cell). Because each operator consumes their own 10 parts, the
study uses 30 distinct ceramic components in total, producing
`3 x 10 x 2 = 60` measurements.

| Study Parameter | Value |
|---|---:|
| Characteristic | Ceramic component breaking strength |
| Operators | 1, 2, 3 (3 operators) |
| Parts per operator | 10 (disjoint sets, 30 distinct parts overall) |
| Replications per part | 2 |
| Design | Nested (parts nested within operators) |
| Total measurements | 60 |
| Process tolerance | not specified (analysis uses %Study Var) |
| Study variation multiplier (sigma) | 6 |
| Interaction significance level (`alpha_lim`) | 0.25 (retained for API consistency only; no interaction term in a nested model) |

**Key contrast with the crossed design.** In a crossed Gage R&R study every
operator measures every part, so the Part factor is crossed with the
Operator factor and the Operator x Part interaction is estimable. In a
nested design each part is seen by exactly one operator, so the interaction
is confounded with the nested part effect -- it cannot be estimated, and
the model omits it by construction.

## 2. Method and Formulas

### 2.1 Nested two-way model

For a nested Gage R&R study with $o$ operators, $p$ parts *per operator*
and $r$ replications per part, the linear model is:

$$
y_{ijk} = \mu + O_j + P(O)_{j(i)} + \varepsilon_{ijk}
$$

where $O_j$ is the effect of operator $j$, $P(O)_{j(i)}$ is the effect of
part $i$ **nested within** operator $j$, and $\varepsilon_{ijk}$ is the
random error (repeatability). The nesting notation $P(O)_{j(i)}$
emphasises that part $i$ appears only under operator $j$ -- it is not a
separate crossed factor.

### 2.2 No interaction term

Because each part is measured by only one operator, the Operator x Part
interaction is completely confounded with the nested Part(Operator)
effect. The model therefore has **no interaction term**; reproducibility
reduces to the Operator component alone.

### 2.3 Variance components

The expected mean squares for the nested model give the following
unbiased variance component estimators (with $p$ = parts per operator and
$r$ = replications per part):

| Source | Variance component estimator |
|---|---|
| Repeatability | $\sigma^2_e = MS_{\text{Repeatability}}$ |
| Part(Operator) | $\sigma^2_{P(O)} = \max\!\left(\dfrac{MS_{P(O)} - MS_e}{r},\, 0\right)$ |
| Operator | $\sigma^2_O = \max\!\left(\dfrac{MS_O - MS_{P(O)}}{p \cdot r},\, 0\right)$ |

The combined quantities are:

$$
\sigma^2_{\text{Total Gage R\&R}} = \sigma^2_{\text{Repeatability}} + \sigma^2_{\text{Operator}},
\quad
\sigma^2_{\text{Reproducibility}} = \sigma^2_{\text{Operator}},
\quad
\sigma^2_{\text{Part-to-Part}} = \sigma^2_{P(O)}
$$

$$
\sigma^2_{\text{Total Variation}} = \sigma^2_{\text{Total Gage R\&R}} + \sigma^2_{\text{Part-to-Part}}
$$

### 2.4 Derived metrics

- **Study Variation** = $6 \times \text{SD}$ (where SD is the standard deviation of each component)
- **%Study Var (%SV)** = $\dfrac{\text{component SD}}{\text{total SD}} \times 100$
- **%Tolerance** = $\dfrac{\text{component Study Var}}{\text{tolerance}} \times 100$ (only when a tolerance is supplied)
- **%Contribution** = $\dfrac{\text{component Variance}}{\text{total Variance}} \times 100$
- **NDC (Number of Distinct Categories)** = $\left\lfloor \sqrt{2}\,\dfrac{\sigma_{\text{Part}}}{\sigma_{\text{Gage}}} \right\rfloor$

### 2.5 AIAG acceptance criteria

| Criterion | Accept | Marginal | Reject |
|---|---|---|---|
| %Study Var (or %Contribution) | < 10% | 10% -- 30% | > 30% |
| NDC | >= 5 | 4 | < 4 |

The overall verdict takes the worst of the two independent assessments.

## 3. Data Preparation

The example data ship with the package as a UTF-8 encoded CSV
(`陶瓷部件.csv`). It contains 60 rows and three columns in Chinese:
`部件` (part, 1--30), `操作员` (operator, 1/2/3) and `响应`
(measurement). The API expects the English column names `Part`,
`Operator` and `MeasurementValue`, so we rename them after loading.

```{r load-data}
ceramic_path <- system.file("extdata/minitab_msa/陶瓷部件.csv", package = "iQualityR.msa")
ceramic_data <- read.csv(ceramic_path, fileEncoding = "UTF-8")
names(ceramic_data) <- c("Part", "Operator", "MeasurementValue")

nrow(ceramic_data)
head(ceramic_data, 9)
```

Inspecting the structure first confirms the nested layout -- each part is
associated with exactly one operator, and each part-operator cell holds
exactly two replications.

```{r data-structure}
str(ceramic_data)
summary(ceramic_data)

# Each part appears with exactly one operator -> nested design
op_per_part <- table(ceramic_data$Part, ceramic_data$Operator)
n_ops_per_part <- rowSums(op_per_part > 0)
cat("Distinct operators per part (unique):",
    paste(unique(n_ops_per_part), collapse = ", "), "\n")
cat("Distinct parts overall:", length(unique(ceramic_data$Part)), "\n")
cat("Distinct operators:", paste(unique(ceramic_data$Operator), collapse = ", "), "\n")

cell_counts <- as.data.frame(table(ceramic_data$Operator, ceramic_data$Part))
names(cell_counts) <- c("Operator", "Part", "Count")
cat("Replications per part-operator cell (unique):",
    paste(unique(cell_counts$Count[cell_counts$Count > 0]), collapse = ", "), "\n")
cat("Overall mean:", fmt4(mean(ceramic_data$MeasurementValue)), "\n")
cat("Overall SD:", fmt4(sd(ceramic_data$MeasurementValue)), "\n")
cat("Measurement range:", fmt4(min(ceramic_data$MeasurementValue)), "--",
    fmt4(max(ceramic_data$MeasurementValue)), "\n")
```

The observed measurement range is about
`r fmt2(max(ceramic_data$MeasurementValue) - min(ceramic_data$MeasurementValue))`
units, distributed across 30 parts drawn from the same batch. Because the
source data set does not specify an engineering tolerance for the
breaking-strength characteristic, the analyses below use
`tolerance = NULL`, which omits the `%Tolerance` column gracefully. If a
tolerance is known (for example, `tolerance = 10`), it can be passed to
`gage_rr_study()` or registered on the plan via `plan$add_specification()`
to enable `%Tolerance` reporting.

## 4. Quick Start (One-Liner)

The fastest path to a full nested Gage R&R analysis is the
`gage_rr_study()` entry function with `method = "nested"`. The function
builds a nested `MSAPlan` internally, fits `MeasurementValue ~ Operator +
Part`, and decomposes the variance using the nested ANOVA formulas.

```{r quick-start}
task_anova <- gage_rr_study(
  ceramic_data,
  method = "nested",
  tolerance = NULL
)
task_anova$summary()
```

The summary prints the variance components table with `%Contribution` and
`%Study Var` (no `%Tolerance` column because no tolerance was supplied),
the key decision metrics (Total Gage R&R `%Contribution`, NDC) and the
AIAG-based interpretation verdict. Note that the `Operator:Part` row is
absent from the ANOVA decomposition -- that interaction is not estimable
in a nested design.

## 5. Step-by-Step Workflow

For traceable, auditable studies it is better to define an explicit
`MSAPlan` first and then bind it to an `MsaTask`. The plan records the
study objectives, operator list, number of parts per operator,
replications and design method, and is saved alongside the results in
the report.

```{r create-plan}
plan <- create_msa_plan(
  plan_name = "Ceramic Strength Nested Gage R&R",
  objectives = "Destructive strength measurement on ceramic parts",
  operators = c("1", "2", "3"),
  parts = 10,
  measurements = 2,
  method = "nested"
)
plan$print()
plan$validate()
```

We then build the task and run the ANOVA computation. `alpha_lim` is
retained in the API for consistency with the crossed workflow but has
no effect for nested designs, because no interaction term is ever
fitted.

```{r build-task}
task <- MsaTask$new(data = ceramic_data, plan = plan, analysis_method = "anova")
task$compute(alpha_lim = 0.25, sigma = 6)
task$summary()
```

### 5.1 Inferring the plan from data

When the data has already been collected, `infer_msa_plan()` can
auto-detect the design (crossed vs nested), the operator list, the part
list and the number of replications directly from the data frame.

```{r infer-plan}
inferred <- infer_msa_plan(ceramic_data)
cat("Inferred design method:", inferred$method, "\n")
cat("Number of operators:", inferred$num_operators, "\n")
cat("Operator names:", paste(inferred$operator_names, collapse = ", "), "\n")
cat("Number of parts (overall):", inferred$num_parts, "\n")
cat("Replications per cell:", inferred$num_measurements, "\n")
```

The inferred plan can be passed directly to `run_msa_analysis()` or to
`MsaTask$new()` exactly like a manually created plan. Note that the
inferred `num_parts` is the *total* number of distinct parts (30),
because each part appears under only one operator in a nested design.

## 6. Nested vs Crossed Comparison

The choice between crossed and nested designs is dictated by the
physics of the measurement, not by statistics.

| Aspect | Crossed Gage R&R | Nested Gage R&R |
|---|---|---|
| Each part measured by every operator | Yes | No |
| Each part is destroyed by measurement | No (allowed) | Yes |
| Part factor | Crossed with Operator | Nested within Operator |
| Operator x Part interaction | Estimable | Confounded with Part(Operator) |
| Reproducibility decomposition | Operator + Operator:Part | Operator only |
| Model formula | `y ~ Part * Operator` (or `Part + Operator`) | `y ~ Operator + Part` (Part nested) |
| Typical application | Non-destructive dimensional checks | Destructive strength, impact, burst tests |

When the parts within an operator's batch are homogeneous (sampled from
the same production stream), the nested `Part(Operator)` component is a
reasonable proxy for the part-to-part variation that *would* have been
measured in a crossed study. When the batches themselves differ, the
`Part(Operator)` component absorbs both part-to-part and batch-to-batch
variation, and the nested design cannot separate them.

A side-by-side comparison of the variance decomposition makes the
structural difference visible. To avoid forcing a crossed model onto
nested data, we compare against the package's bundled crossed example
(`量具数据.csv`, see `vignette("gage-rr-crossed", package = "iQualityR.msa")`):

```{r nested-vs-crossed}
# Nested variance components (ceramic data, correct analysis)
vc_nested <- task_anova$variance_components %>%
  dplyr::mutate(Source = trimws(.data$Source)) %>%
  dplyr::select(Source, VarComp_Nested = .data$Variance,
                PctContr_Nested = .data$PercentContribution)

# Crossed variance components (shaft-diameter data, correct crossed analysis)
crossed_path <- system.file("extdata/minitab_msa/量具数据.csv", package = "iQualityR.msa")
crossed_data <- read.csv(crossed_path, fileEncoding = "UTF-8")
names(crossed_data) <- c("Part", "Operator", "MeasurementValue")
task_crossed <- gage_rr_study(crossed_data, method = "crossed", tolerance = 8.0)

vc_crossed <- task_crossed$variance_components %>%
  dplyr::mutate(Source = trimws(.data$Source)) %>%
  dplyr::select(Source, VarComp_Crossed = .data$Variance,
                PctContr_Crossed = .data$PercentContribution)

comparison <- vc_nested %>%
  dplyr::full_join(vc_crossed, by = "Source") %>%
  dplyr::mutate(
    VarComp_Nested  = fmt4(.data$VarComp_Nested),
    VarComp_Crossed = fmt4(.data$VarComp_Crossed),
    PctContr_Nested  = fmt2(.data$PctContr_Nested),
    PctContr_Crossed = fmt2(.data$PctContr_Crossed)
  )
comparison
```

The two tables share the same `Source` rows, but the meaning of each row
differs. In the nested column, `Part-to-Part` is the variance of
`Part(Operator)` -- the spread of part means *within* each operator's
batch. In the crossed column, `Part-to-Part` is the variance of the
marginal `Part` factor estimated across all operators. The crossed
table may also report a non-zero `Operator:Part` sub-row under
Reproducibility (visible when the interaction term is retained at
`alpha_lim = 0.25`); the nested model can never produce such a row,
because the interaction is structurally confounded with `Part(Operator)`.

Applying a crossed model directly to nested data is statistically
incorrect and the package will refuse to estimate the interaction
term -- the `aov()` call drops the aliased `Part:Operator` columns.
Use `method = "nested"` whenever each part is measured by only one
operator.

## 7. ANOVA Deep-Dive

The ANOVA results live on three public fields of the task:

- `task$anova_results` -- the ANOVA table and metadata (`has_interaction`
  is always `FALSE` for nested designs, `with_interaction` is `NULL`)
- `task$variance_components` -- the variance components tibble
- `task$statistics` -- the summary statistics (NDC, %Contribution, %Tolerance)

### 7.1 ANOVA table

```{r anova-table}
anova_info <- task$anova_results
cat("Has interaction term:", anova_info$has_interaction, "\n")
cat("alpha_lim used (no effect for nested):", anova_info$alpha_lim, "\n")
cat("Design method:", anova_info$design_method, "\n")

anova_tbl <- task$results$data_tables$anova_table
anova_tbl
```

Each row corresponds to a source of variation. The `Operator` row tests
whether operator means differ (against `MS_Part`); the `Part` row tests
whether the nested part-within-operator effect is non-zero (against
`MS_Repeatability`); the `Repeatability` row is the residual (pure error
within part-operator cells); the `Total` row aggregates all degrees of
freedom and sums of squares. Because R fits `Part` as a single
crossed-style factor in the underlying `aov()` call, the printed `Part`
row corresponds to the *nested* `Part(Operator)` effect -- its degrees of
freedom equal $o \cdot (p - 1) = 3 \cdot 9 = 27$, the signature of a
nested design.

### 7.2 Variance components

```{r variance-components}
vc <- task$variance_components
vc_display <- data.frame(
  Source = trimws(vc$Source),
  VarComp = fmt4(vc$Variance),
  Pct_Contribution = fmt2(vc$PercentContribution),
  StdDev = fmt4(vc$StdDev),
  StudyVar = fmt4(vc$StudyVar)
)
if ("PercentStudyVar" %in% names(vc)) {
  vc_display$Pct_StudyVar <- fmt2(vc$PercentStudyVar)
}
if ("PercentTolerance" %in% names(vc)) {
  vc_display$Pct_Tolerance <- fmt2(vc$PercentTolerance)
}
vc_display
```

Reading the table:

- **Total Gage R&R** = Repeatability + Reproducibility. It is the
  total measurement-system variation.
- **Repeatability** (equipment variation, EV) is the within-cell
  variation -- the gage's own noise, including any microstructure
  differences between the two halves of a destructively split sample.
- **Reproducibility** (appraiser variation, AV) equals the **Operator**
  component only. There is no `Operator:Part` sub-row in a nested model.
- **Part-to-Part** here is the variance of `Part(Operator)` -- the
  spread of part means *within* each operator's batch, averaged over
  operators. It mixes genuine part-to-part variation with any
  batch-to-batch differences between operators' samples.
- **Total Variation** = Total Gage R&R + Part-to-Part.

### 7.3 Gage evaluation table

The gage evaluation table is the Minitab-style compact view of the same
components.

```{r gage-evaluation}
gage_eval <- task$results$data_tables$gage_evaluation
gage_eval
```

### 7.4 Key statistics

```{r key-stats}
stats_df <- data.frame(
  Metric = c(
    "Total Gage R&R %Contribution",
    "Total Gage R&R %Study Var",
    "Total Gage R&R %Tolerance",
    "Number of Distinct Categories (NDC)",
    "Verdict status",
    "Verdict message"
  ),
  Value = c(
    fmt_pct(task$statistics$total_gage_rr_percent, 2),
    fmt_pct(task$variance_components$PercentStudyVar[1], 2),
    ifelse(is.na(task$statistics$tolerance_percent),
           "NA (no tolerance supplied)",
           fmt_pct(task$statistics$tolerance_percent, 2)),
    as.character(task$statistics$ndc),
    task$interpretation$status,
    task$interpretation$message
  )
)
stats_df
```

## 8. Visualization Gallery

The Minitab-style summary plot for nested designs is a **five-panel**
layout: components of variation, Xbar chart by operator, response by
operator, R chart by operator, and response by part. The `interaction`
panel that appears in the crossed layout is intentionally omitted,
because the Operator x Part interaction is not estimable when parts are
nested within operators.

### 8.1 Five-panel summary

```{r plot-summary, fig.width=10, fig.height=8}
print(task$plot(type = "summary"))
```

### 8.2 Components of variation

```{r plot-variation}
print(task$plot(type = "variation"))
```

The dashed reference lines mark the 10% (accept) and 30% (marginal)
AIAG thresholds. The Part-to-Part bar should be much larger than the
Total Gage R&R bar when the gage is acceptable.

### 8.3 Response by part

```{r plot-by-part}
print(task$plot(type = "by_part"))
```

The diamond markers and connecting line show the mean per part. Because
each part is owned by a single operator, the points within a part are
coloured by that one operator -- the colour pattern thus visually
reinforces the nested structure (each cluster of same-colour points
belongs to one operator's batch).

### 8.4 Response by operator

```{r plot-by-operator}
print(task$plot(type = "by_operator"))
```

Boxplots are centred on each operator. Closely aligned medians and
means indicate small reproducibility bias; a systematically high or low
operator reveals a reproducibility problem.

### 8.5 Xbar chart by operator

```{r plot-xbar}
print(task$plot(type = "xbar_chart"))
```

The Xbar chart is faceted by operator; each point is the mean of a part
within that operator. Most points falling outside the control limits is
**desirable** here -- it means the parts within each operator's batch
are sufficiently different to be distinguished by the measurement system.

### 8.6 R chart by operator

```{r plot-r}
print(task$plot(type = "r_chart"))
```

The R chart monitors within-cell repeatability. Points inside the
control limits indicate that the operators are consistent in their
repeatability; an out-of-control point would flag an unstable
measurement process for that operator-part combination.

### 8.7 Interaction plot is not available

Requesting `type = "interaction"` on a nested task raises an
informative error rather than producing a chart, because the
Operator x Part interaction is structurally confounded with the nested
`Part(Operator)` effect:

```{r plot-interaction-na}
tryCatch(
  print(task$plot(type = "interaction")),
  error = function(e) cat("Expected error for nested design:",
                          conditionMessage(e), "\n")
)
```

## 9. Report Export

The task can export a self-contained report in Excel, HTML, PDF or Word
format. Excel is always available and writes each result table to its
own worksheet.

```{r report-excel}
xlsx_path <- tempfile("gage_rr_nested_", fileext = ".xlsx")
task$report(format = "excel", path = xlsx_path)

data.frame(
  Format = "Excel",
  Path = xlsx_path,
  Exists = file.exists(xlsx_path),
  Size_KB = fmt4(file.info(xlsx_path)$size / 1024)
)
```

The exported workbook contains the following sheets:

```{r report-sheets}
names(task$build_excel_sheets())
```

HTML, PDF and Word reports are produced by rendering an Rmd template and
should be generated in an interactive R session (they are not
evaluated during vignette build to avoid nesting rmarkdown calls):

```{r report-other, eval=FALSE}
task$report(format = "html", path = "gage_rr_nested_report.html")
task$report(format = "pdf",  path = "gage_rr_nested_report.pdf")
task$report(format = "docx", path = "gage_rr_nested_report.docx")
```

## 10. Interpretation and Decision

Pulling the verdict together for the ANOVA analysis:

```{r verdict}
verdict_df <- data.frame(
  Criterion = c(
    "Total Gage R&R %Study Var",
    "Total Gage R&R %Contribution",
    "Total Gage R&R %Tolerance",
    "NDC",
    "Interaction retained",
    "Overall status"
  ),
  Value = c(
    fmt_pct(task$variance_components$PercentStudyVar[1], 2),
    fmt_pct(task$statistics$total_gage_rr_percent, 2),
    ifelse(is.na(task$statistics$tolerance_percent),
           "NA (no tolerance supplied)",
           fmt_pct(task$statistics$tolerance_percent, 2)),
    as.character(task$statistics$ndc),
    as.character(task$anova_results$has_interaction),
    task$interpretation$status
  )
)
verdict_df
cat("Message:", task$interpretation$message, "\n")
```

Engineering conclusions for this study:

1. **Acceptance**. The Total Gage R&R `%Study Var` for the destructive
   ceramic-strength measurement is below the 10% acceptance threshold
   and the NDC is at least 5, so the measurement system is acceptable
   for monitoring ceramic component strength despite the nested
   (destructive) design.
2. **R chart**. All R-chart points fall within the control limits, so
   the three operators are consistent in their repeatability -- no
   special-cause instability is present in the destructive test
   procedure.
3. **Xbar chart**. Most part means fall outside the Xbar control limits
   within each operator's facet, which confirms that the 10 parts
   sampled per operator span a wide enough range and that the test
   apparatus can distinguish between them.
4. **By Operator**. Operator means are close to each other, indicating
   small reproducibility bias; no operator is systematically high or
   low across the destructive test.
5. **Interaction**. No interaction is estimable in a nested design; the
   `Operator:Part` row is structurally absent from the ANOVA
   decomposition, and the `interaction` plot type is not available.
6. **Components of Variation**. The Part-to-Part bar (which here is
   `Part(Operator)`) dominates the Total Gage R&R bar, which is the
   desired pattern: most of the observed variation comes from the
   parts, not from the destructive measurement system.

If the `%Study Var` had fallen in the 10%--30% marginal band, the
recommended actions would be to investigate the dominant component
(Repeatability points to the test fixture or the specimen grip;
Reproducibility points to operator training or sample-preparation
differences between operators' batches) and to re-run the study after
corrective action. If the nested `Part(Operator)` component looks
suspiciously large, that is often a sign that the operators' batches are
not truly homogeneous -- a batch-to-batch effect has been absorbed into
the part-to-part estimate, and a more carefully controlled sampling plan
is needed.

## 11. Reference

- Minitab Help: *Nested Gage R&R Study* -- example data and
  interpretation guidance for destructive-measurement studies.
- Minitab Help: *All statistics and graphs for Nested Gage R&R Study* --
  definitions of VarComp, %Contribution, %Study Var, %Tolerance and
  NDC for the nested model.
- AIAG Measurement Systems Analysis (MSA) Reference Manual, 4th
  edition -- nested ANOVA method, variance component estimators under
  nesting, and the acceptance criteria for %GRR and NDC.
- Montgomery, D. C. (2017). *Design and Analysis of Experiments*, 9th
  ed. -- source of the nested two-way model and the
  expected-mean-square derivations used by the variance component
  formulas.
