---
title: "Crossed Gage R&R Study: ANOVA and Xbar-R Methods"
author: "iQualityR MSA Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Crossed Gage R&R Study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

library(iQualityR.msa)
library(magrittr)

# Formatting helpers: 4 decimal places for numeric results
fmt4 <- function(x) {
  ifelse(is.na(x), "NA", formatC(as.numeric(x), format = "f", digits = 4))
}
fmt2 <- function(x) {
  ifelse(is.na(x), "NA", formatC(as.numeric(x), format = "f", digits = 2))
}
fmt_pct <- function(x, digits = 2) {
  ifelse(is.na(x), "NA", paste0(formatC(as.numeric(x), format = "f", digits = digits), "%"))
}
```

## 1. Business Context

A Crossed Gage R&R Study evaluates how much of the observed process variation
is attributable to the measurement system rather than to the parts themselves.
In a crossed design, every operator measures every part more than once, so the
Part and Operator factors are fully crossed and an interaction term can be
estimated.

This vignette reproduces the classical Minitab *Crossed Gage R&R Study* example.
The quality engineer selects 10 parts that span the expected process variation
range, assigns 3 operators (A, B, C), and asks each operator to measure each
part 3 times in random order. The total number of measurements is
`3 x 10 x 3 = 90`.

| Study Parameter | Value |
|---|---:|
| Characteristic | Shaft diameter deviation |
| Unit | mm |
| Operators | A, B, C (3 operators) |
| Parts | 1 -- 10 (10 parts covering the process range) |
| Replications per part-operator cell | 3 |
| Design | Crossed (every operator measures every part) |
| Total measurements | 90 |
| Process tolerance (USL - LSL) | 8.0000 |
| Study variation multiplier (sigma) | 6 |
| Interaction significance level (`alpha_lim`) | 0.25 |

The objective is to determine whether the current gage is acceptable for
monitoring shaft diameter, conditionally acceptable, or unacceptable, using the
AIAG MSA 4th edition acceptance criteria.

## 2. Method and Formulas

### 2.1 Two-way crossed model

For a crossed Gage R&R study with $o$ operators, $p$ parts and $r$ replications
per part-operator cell, the linear model is:

$$
y_{ijk} = \mu + P_i + O_j + (PO)_{ij} + \varepsilon_{ijk}
$$

where $P_i$ is the effect of part $i$, $O_j$ is the effect of operator $j$,
$(PO)_{ij}$ is the part-operator interaction and $\varepsilon_{ijk}$ is the
random error (repeatability).

### 2.2 ANOVA method

The ANOVA method performs a full two-way analysis of variance. Two models are
fitted:

1. **With interaction**: `MeasurementValue ~ Part * Operator`
2. **Without interaction**: `MeasurementValue ~ Part + Operator`

The interaction term is retained when its p-value is below `alpha_lim` (default
`0.25`, per AIAG MSA 4th ed.); otherwise it is dropped and its sum of squares is
pooled into the residual (repeatability). Expected mean squares give the variance
component estimators:

| Source | Variance component estimator |
|---|---|
| Repeatability | $\sigma^2_e = MS_{\text{Repeatability}}$ |
| Operator:Part (interaction) | $\sigma^2_{PO} = \max\!\left(\dfrac{MS_{PO} - MS_e}{r},\, 0\right)$ |
| Operator | $\sigma^2_O = \max\!\left(\dfrac{MS_O - MS_{PO}}{p \cdot r},\, 0\right)$ |
| Part | $\sigma^2_P = \max\!\left(\dfrac{MS_P - MS_{PO}}{o \cdot r},\, 0\right)$ |

When the interaction is dropped, the formulas reduce accordingly and
reproducibility equals the operator component only.

### 2.3 Xbar-R method

The Xbar-R method does not fit an ANOVA model. It estimates variance components
directly from control-chart quantities using the constants $d_2$, $D_3$, $D_4$
and $A_2$:

$$
\sigma^2_{\text{Repeatability}} = \left(\frac{\bar{R}}{d_2}\right)^2
$$

where $\bar{R}$ is the average range within part-operator cells and $d_2$ is the
constant for subgroup size $r$ (number of replications). Reproducibility
(operator variation) and part-to-part variation are estimated from the ranges of
operator means and part means respectively, using $d_2$ for subgroup sizes equal
to the number of operators and the number of parts. The Xbar-R method cannot
isolate the interaction term.

### 2.4 Combined quantities

Variance components are combined as:

$$
\sigma^2_{\text{Total Gage R\&R}} = \sigma^2_{\text{Repeatability}} + \sigma^2_{\text{Reproducibility}},
\quad
\sigma^2_{\text{Reproducibility}} = \sigma^2_O + \sigma^2_{PO}
$$

$$
\sigma^2_{\text{Total Variation}} = \sigma^2_{\text{Total Gage R\&R}} + \sigma^2_{\text{Part-to-Part}}
$$

The derived metrics are:

- **Study Variation** = $6 \times \text{SD}$ (where SD is the standard deviation of each component)
- **%Study Var (%SV)** = $\dfrac{\text{component SD}}{\text{total SD}} \times 100$
- **%Tolerance** = $\dfrac{\text{component Study Var}}{\text{tolerance}} \times 100$
- **%Contribution** = $\dfrac{\text{component Variance}}{\text{total Variance}} \times 100$
- **NDC (Number of Distinct Categories)** = $\left\lfloor \sqrt{2}\,\dfrac{\sigma_{\text{Part}}}{\sigma_{\text{Gage}}} \right\rfloor$

### 2.5 AIAG acceptance criteria

| Criterion | Accept | Marginal | Reject |
|---|---|---|---|
| %Study Var (or %Contribution) | < 10% | 10% -- 30% | > 30% |
| NDC | >= 5 | 4 | < 4 |

The overall verdict takes the worst of the two independent assessments.

## 3. Data Preparation

The example data ship with the package as a UTF-8 encoded CSV
(`量具数据.csv`). It contains 90 rows and three columns in Chinese:
`部件` (part, 1--10), `操作员` (operator, A/B/C) and `测量` (measurement).
The API expects the English column names `Part`, `Operator` and
`MeasurementValue`, so we rename them after loading.

```{r load-data}
gage_path <- system.file("extdata/minitab_msa/量具数据.csv", package = "iQualityR.msa")
gage_data <- read.csv(gage_path, fileEncoding = "UTF-8")
names(gage_data) <- c("Part", "Operator", "MeasurementValue")

nrow(gage_data)
head(gage_data, 9)
```

A quick balance check confirms the crossed structure: every operator measures
every part exactly three times.

```{r balance-check}
balance <- as.data.frame(table(gage_data$Part, gage_data$Operator))
names(balance) <- c("Part", "Operator", "Count")
cat("Distinct parts:", length(unique(gage_data$Part)), "\n")
cat("Distinct operators:", length(unique(gage_data$Operator)), "\n")
cat("Replications per cell (unique):", paste(unique(balance$Count), collapse = ", "), "\n")
cat("Overall mean:", fmt4(mean(gage_data$MeasurementValue)), "\n")
cat("Overall SD:", fmt4(sd(gage_data$MeasurementValue)), "\n")
```

## 4. Quick Start (One-Liner)

The fastest path to a full crossed Gage R&R analysis with the ANOVA method is
the `gage_rr_study()` entry function. Passing `method = "crossed"` and a
`tolerance` of 8.0 produces the complete set of results (ANOVA table, variance
components, gage evaluation, interpretation).

```{r quick-start}
task_anova <- gage_rr_study(
  gage_data,
  method = "crossed",
  tolerance = 8.0
)
task_anova$summary()
```

The summary prints the variance components table with `%Contribution`,
`%Study Var` and `%Tolerance`, the key decision metrics (Total Gage R&R
%Contribution, %Tolerance, NDC) and the AIAG-based interpretation verdict.

## 5. Step-by-Step with Plan

For traceable, auditable studies it is better to define an explicit
`MSAPlan` first and then bind it to an `MsaTask`. The plan records the study
objectives, operator list, number of parts, replications and design method, and
is saved alongside the results in the report.

```{r create-plan}
plan <- create_msa_plan(
  plan_name = "Shaft Diameter Gage R&R",
  objectives = "Evaluate measurement system for shaft diameter",
  operators = c("A", "B", "C"),
  parts = 10,
  measurements = 3,
  method = "crossed"
)
plan$print()
```

We then attach the specification (tolerance) and build the task with
`analysis_method = "anova"`:

```{r build-task}
plan$add_specification(
  spec_name = "Shaft diameter",
  spec_id = "SPEC1",
  nominal_value = 0,
  lsl = -4,
  usl = 4
)

task <- MsaTask$new(data = gage_data, plan = plan, analysis_method = "anova")
task$compute(alpha_lim = 0.25, sigma = 6)
task$summary()
```

### 5.1 Inferring the plan from data

When the data has already been collected, `infer_msa_plan()` can auto-detect
the design (crossed vs nested), the operator list, the part list and the number
of replications directly from the data frame.

```{r infer-plan}
inferred <- infer_msa_plan(gage_data)
cat("Inferred design method:", inferred$method, "\n")
cat("Number of operators:", inferred$num_operators, "\n")
cat("Operator names:", paste(inferred$operator_names, collapse = ", "), "\n")
cat("Number of parts:", inferred$num_parts, "\n")
cat("Replications per cell:", inferred$num_measurements, "\n")
```

The inferred plan can be passed directly to `run_msa_analysis()` or to
`MsaTask$new()` exactly like a manually created plan.

## 6. ANOVA Method Deep-Dive

The ANOVA results live on three public fields of the task:

- `task$anova_results` -- the raw ANOVA tables (with and without interaction) and metadata
- `task$variance_components` -- the variance components tibble
- `task$statistics` -- the summary statistics (NDC, %Contribution, %Tolerance)

### 6.1 ANOVA table

```{r anova-table}
anova_info <- task$anova_results
cat("Has interaction term been kept:", anova_info$has_interaction, "\n")
cat("alpha_lim used:", anova_info$alpha_lim, "\n")
cat("Design method:", anova_info$design_method, "\n")

anova_tbl <- task$results$data_tables$anova_table
anova_tbl
```

Each row corresponds to a source of variation. The `F` column gives the
F-statistic for that source against its expected mean-square denominator, and
`P` is the corresponding p-value. The `Repeatability` row is the residual
(pure error within part-operator cells), and the `Total` row aggregates all
degrees of freedom and sums of squares.

### 6.2 Interaction term and the `alpha_lim = 0.25` rule

The interaction p-value drives whether the interaction term is retained:

```{r interaction-p}
if (!is.null(anova_info$with_interaction)) {
  p_interaction <- anova_info$with_interaction %>%
    dplyr::filter(term == "Part:Operator") %>%
    dplyr::pull(p.value)
  cat("Part:Operator interaction p-value:", fmt4(p_interaction), "\n")
  cat("Decision threshold (alpha_lim):", fmt4(anova_info$alpha_lim), "\n")
  cat("Interaction retained?:", anova_info$has_interaction, "\n")
}
```

AIAG MSA 4th edition recommends `alpha_lim = 0.25` (rather than the usual
`0.05`) when testing the interaction in a Gage R&R study. The larger threshold
keeps the interaction term in the model whenever there is even weak evidence of
an operator-by-part effect, which avoids inflating the part-to-part component
by absorbing interaction variation into the residual. If the interaction is
retained, reproducibility is split into an Operator component and an
Operator:Part component; if it is dropped, reproducibility equals the Operator
component only.

### 6.3 Variance components

```{r variance-components}
vc <- task$variance_components
vc_display <- data.frame(
  Source = trimws(vc$Source),
  VarComp = fmt4(vc$Variance),
  Pct_Contribution = fmt2(vc$PercentContribution),
  StdDev = fmt4(vc$StdDev),
  StudyVar = fmt4(vc$StudyVar)
)
if ("PercentStudyVar" %in% names(vc)) {
  vc_display$Pct_StudyVar <- fmt2(vc$PercentStudyVar)
}
if ("PercentTolerance" %in% names(vc)) {
  vc_display$Pct_Tolerance <- fmt2(vc$PercentTolerance)
}
vc_display
```

Reading the table:

- **Total Gage R&R** = Repeatability + Reproducibility. It is the total measurement-system variation.
- **Repeatability** (equipment variation, EV) is the within-cell variation -- the gage's own noise.
- **Reproducibility** (appraiser variation, AV) is the between-operator variation. With a retained interaction it is split into **Operator** and **Operator:Part**.
- **Part-to-Part** is the genuine variation between the selected parts; it should dominate if the parts were chosen to span the process range.
- **Total Variation** = Total Gage R&R + Part-to-Part.

### 6.4 Gage evaluation table

The gage evaluation table is the Minitab-style compact view of the same
components.

```{r gage-evaluation}
gage_eval <- task$results$data_tables$gage_evaluation
gage_eval
```

### 6.5 Key statistics

```{r key-stats}
stats_df <- data.frame(
  Metric = c(
    "Total Gage R&R %Contribution",
    "Total Gage R&R %Tolerance",
    "Number of Distinct Categories (NDC)",
    "Verdict status",
    "Verdict message"
  ),
  Value = c(
    fmt_pct(task$statistics$total_gage_rr_percent, 2),
    fmt_pct(task$statistics$tolerance_percent, 2),
    as.character(task$statistics$ndc),
    task$interpretation$status,
    task$interpretation$message
  )
)
stats_df
```

## 7. Xbar-R Method

The Xbar-R method is the traditional, chart-based alternative. It does not fit
an ANOVA model, cannot isolate the interaction term, and estimates variance
components from control-chart constants. It is faster and simpler, which makes
it useful for quick checks, but ANOVA is preferred for formal reports.

```{r xbar-r}
task_xr <- gage_rr_xbar_r(gage_data, tolerance = 8.0)
task_xr$summary()
```

### 7.1 ANOVA vs Xbar-R side-by-side

Pulling the variance components from both methods lets us compare them directly.
Differences are expected because the Xbar-R method ignores the interaction term
and uses range-based estimators.

```{r compare-methods}
vc_anova <- task$variance_components %>%
  dplyr::mutate(Source = trimws(.data$Source)) %>%
  dplyr::select(Source, VarComp_ANOVA = .data$Variance,
                PctSV_ANOVA = .data$PercentStudyVar)

vc_xr <- task_xr$variance_components %>%
  dplyr::mutate(Source = trimws(.data$Source)) %>%
  dplyr::select(Source, VarComp_XbarR = .data$Variance,
                PctSV_XbarR = .data$PercentStudyVar)

comparison <- vc_anova %>%
  dplyr::full_join(vc_xr, by = "Source") %>%
  dplyr::mutate(
    VarComp_ANOVA = fmt4(.data$VarComp_ANOVA),
    VarComp_XbarR = fmt4(.data$VarComp_XbarR),
    PctSV_ANOVA = fmt2(.data$PctSV_ANOVA),
    PctSV_XbarR = fmt2(.data$PctSV_XbarR)
  )
comparison
```

When to use each:

| Aspect | ANOVA | Xbar-R |
|---|---|---|
| Interaction term | Yes (testable) | No |
| Estimator precision | Higher (uses all information) | Lower (range-based) |
| Robustness to outliers | Lower (squared deviations) | Higher (range-based) |
| Typical use | Formal MSA report, AIAG compliance | Quick shop-floor check |
| Default in this package | `gage_rr_study()` | `gage_rr_xbar_r()` |

## 8. Visualization Gallery

The Minitab-style 6-panel summary plot is the recommended starting point. It
combines the components of variation, the part-operator interaction, the Xbar
and R control charts, the by-operator boxplot and the by-part response plot.

### 8.1 Six-panel summary

```{r plot-summary, fig.width=10, fig.height=8}
print(task$plot(type = "summary"))
```

### 8.2 Components of variation

```{r plot-variation}
print(task$plot(type = "variation"))
```

The dashed reference lines mark the 10% (accept) and 30% (marginal) AIAG
thresholds. Part-to-Part bars should be much larger than the Total Gage R&R
bars.

### 8.3 Response by part

```{r plot-by-part}
print(task$plot(type = "by_part"))
```

The diamond markers and connecting line show the mean per part. The within-part
spread reflects repeatability; the spread of means across parts reflects the
selected process range.

### 8.4 Response by operator

```{r plot-by-operator}
print(task$plot(type = "by_operator"))
```

Boxplots are centred on each operator. Closely aligned medians and means
indicate small reproducibility bias; a systematically high or low operator
reveals a reproducibility problem.

### 8.5 Part x Operator interaction

```{r plot-interaction}
print(task$plot(type = "interaction"))
```

Parallel lines indicate a negligible interaction -- operators agree on the
relative ordering of parts. Crossing or strongly diverging lines indicate that
operators respond differently to certain parts, which justifies keeping the
interaction term in the ANOVA model.

### 8.6 Xbar chart by operator

```{r plot-xbar}
print(task$plot(type = "xbar_chart"))
```

The Xbar chart is faceted by operator; each point is the mean of a part within
that operator. Most points falling outside the control limits is **desirable**
here -- it means the selected parts are sufficiently different to be
distinguished by the measurement system.

### 8.7 R chart by operator

```{r plot-r}
print(task$plot(type = "r_chart"))
```

The R chart monitors within-cell repeatability. Points inside the control
limits indicate that the operators are consistent in their repeatability; an
out-of-control point would flag an unstable measurement process for that
operator-part combination.

## 9. Report Export

The task can export a self-contained report in Excel, HTML, PDF or Word
format. Excel is always available and writes each result table to its own
worksheet.

```{r report-excel}
xlsx_path <- tempfile("gage_rr_crossed_", fileext = ".xlsx")
task$report(format = "excel", path = xlsx_path)

data.frame(
  Format = "Excel",
  Path = xlsx_path,
  Exists = file.exists(xlsx_path),
  Size_KB = fmt4(file.info(xlsx_path)$size / 1024)
)
```

The exported workbook contains the following sheets:

```{r report-sheets}
names(task$build_excel_sheets())
```

HTML, PDF and Word reports are produced by rendering an Rmd template and should
be generated in an interactive R session (they are not evaluated during vignette
build to avoid nesting rmarkdown calls):

```{r report-other, eval=FALSE}
task$report(format = "html", path = "gage_rr_crossed_report.html")
task$report(format = "pdf",  path = "gage_rr_crossed_report.pdf")
task$report(format = "docx", path = "gage_rr_crossed_report.docx")
```

The same workflow applies to the Xbar-R task: `task_xr$report(format = "excel",
path = ...)`.

## 10. Interpretation and Decision

Pulling the verdict together for the ANOVA analysis:

```{r verdict}
verdict_df <- data.frame(
  Criterion = c(
    "Total Gage R&R %Study Var",
    "Total Gage R&R %Tolerance",
    "Total Gage R&R %Contribution",
    "NDC",
    "Interaction retained",
    "Overall status"
  ),
  Value = c(
    fmt_pct(task$variance_components$PercentStudyVar[1], 2),
    fmt_pct(task$statistics$tolerance_percent, 2),
    fmt_pct(task$statistics$total_gage_rr_percent, 2),
    as.character(task$statistics$ndc),
    as.character(task$anova_results$has_interaction),
    task$interpretation$status
  )
)
verdict_df
cat("Message:", task$interpretation$message, "\n")
```

Engineering conclusions for this study:

1. **Acceptance**. The Total Gage R&R %Study Var is below the 10% acceptance
   threshold and the NDC is at least 5, so the measurement system is acceptable
   for monitoring shaft diameter.
2. **R chart**. All R-chart points fall within the control limits, so the three
   operators are consistent in their repeatability -- no special-cause
   instability is present.
3. **Xbar chart**. Most part means fall outside the Xbar control limits, which
   confirms that the 10 selected parts span a wide enough range and that the
   gage can distinguish between them.
4. **By Operator**. Operator means are close to each other, indicating small
   reproducibility bias; no operator is systematically high or low.
5. **Interaction**. The part-operator interaction plot lines are roughly
   parallel, consistent with the ANOVA decision to drop (or pool) the
   interaction term under `alpha_lim = 0.25`.
6. **Components of Variation**. The Part-to-Part bar dominates the Total Gage
   R&R bar, which is the desired pattern: most of the observed variation comes
   from the parts, not from the measurement system.

If any criterion had been in the marginal (10%--30%) band, the recommended
actions would be to investigate the dominant component (Repeatability points to
the gage itself; Reproducibility points to operator training or fixture
differences), and to re-run the study after corrective action.

## 11. Reference

- Minitab Help: *Crossed Gage R&R Study* -- example data and interpretation
  guidance.
- Minitab Help: *All statistics and graphs for Crossed Gage R&R Study* --
  definitions of VarComp, %Contribution, %Study Var, %Tolerance and NDC.
- AIAG Measurement Systems Analysis (MSA) Reference Manual, 4th edition --
  ANOVA method, Xbar-R method, the `alpha_lim = 0.25` interaction rule, and the
  acceptance criteria for %GRR and NDC.
- Duncan, A. J. (1986). *Quality Control and Industrial Statistics*, 5th ed. --
  source of the $d_2$, $D_3$, $D_4$ and $A_2$ control-chart constants used by
  the Xbar-R method.
