---
title: "Gage Linearity and Bias Study"
author: "iQualityR MSA Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Gage Linearity and Bias Study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

library(iQualityR.msa)

fmt_num <- function(x, digits = 6) formatC(x, format = "f", digits = digits)
fmt_pct <- function(x, digits = 2) paste0(formatC(x, format = "f", digits = digits), "%")
```

## 1. Business Context

A Gage Linearity and Bias Study answers a question that a single-reference
Type 1 study cannot: **does the bias of the gage stay constant across the
operating range of the characteristic?** If the bias changes systematically
with the magnitude of the measurement, the gage has a *linearity* problem,
and a single bias correction will not make it acceptable.

This vignette reproduces the canonical Minitab **Gage Linearity and Bias
Study** example. An engineer measures the **diameter of bearings** using a
micrometer-style gage. Five reference parts (main levels) are selected to
span the operating range from 2.0 mm to 10.0 mm. Each reference part has a
known master value, traceable to a calibration standard, and is measured 12
times by a single operator under repeatability conditions. The 60
measurements are then regressed against their reference values to expose any
magnitude-dependent bias.

The study parameters are summarized below.

| Item                          | Value                          |
|-------------------------------|--------------------------------|
| Characteristic                | Bearing diameter               |
| Unit                          | mm                             |
| Reference values (main levels)| 2.0, 4.0, 6.0, 8.0, 10.0       |
| Lower specification limit     | 0.0                            |
| Upper specification limit     | 12.0                           |
| Tolerance ($T = USL - LSL$)   | 12.0                           |
| Process variation ($PV=6\sigma$) | 16.5368                       |
| Number of reference parts     | 5                              |
| Measurements per reference    | 12                             |
| Total measurements            | 60                             |
| Confidence level              | 0.95                           |

The tolerance ($T = USL - LSL = 12$ mm) and the **process variation**
($PV = 16.5368$ mm, taken from a historical capability study as $6\sigma$)
are both recorded as study inputs. Per Minitab's *Methods and formulas for
Gage Linearity and Bias Study*, the linearity and bias percentages are
expressed relative to the **process variation**, not the tolerance; the
tolerance is retained only for context and for the slope-magnitude
acceptance criterion.

## 2. Method and Formulas

The analysis has three stages: per-reference bias summarisation, a linear
regression of *all individual measurement biases* on the reference value,
and a pass/fail evaluation against acceptance thresholds. The formulas
below follow Minitab's *Gage Linearity and Bias Study* methods and the
AIAG MSA 4th edition convention.

**Per-reference bias.** For each distinct reference value $r$, the bias is
the difference between the mean of the $n_r$ repeated measurements and the
reference value:

$$
\text{bias}_r = \bar{x}_r - r, \qquad
\bar{x}_r = \frac{1}{n_r}\sum_{i=1}^{n_r} x_{ir}
$$

A two-sided $t$-test of $H_0: \text{bias}_r = 0$ is run per reference with
$t_r = \text{bias}_r / (\text{sd}_r / \sqrt{n_r})$ on $n_r - 1$ degrees of
freedom. The aggregate metrics are the average bias (across all
$n$ individual biases) and the maximum absolute per-reference bias, each
expressed as a percent of the process variation:

$$
\%\text{Bias}_{avg} = \frac{|\overline{\text{bias}}|}{PV}\cdot 100, \qquad
\%\text{Bias}_{max} = \frac{\max_r|\text{bias}_r|}{PV}\cdot 100
$$

where $PV$ is the process variation supplied by the user (typically
$6\sigma$ from a capability study). An overall average-bias $t$-test is
also computed across all $n$ individual biases.

**Linearity regression.** Following Minitab, the regression is performed on
**all $n$ individual measurement biases** (not on the per-reference means).
For each measurement $x_{ir}$ on reference $r$ the individual bias is
$b_{ir} = x_{ir} - r$, and the ordinary-least-squares fit is

$$
b_{ir} = \beta_0 + \beta_1 \cdot r + \varepsilon_{ir}
$$

With $n = 60$ individual biases for this example, the regression has
$n - 2 = 58$ degrees of freedom for the slope. The slope $\beta_1$ is the
**linearity coefficient**: the change in bias per unit increase in the
reference value. The intercept $\beta_0$ is the extrapolated bias at
reference = 0. The linearity metric is the absolute slope scaled by the
process variation:

$$
\text{Linearity} = |\hat{\beta}_1| \cdot PV
$$

and its percent of process variation is

$$
\%\text{Linearity} = \frac{\text{Linearity}}{PV}\cdot 100 = |\hat{\beta}_1| \cdot 100
$$

A $100(1-\alpha)\%$ confidence interval for the linearity is obtained by
propagating the slope CI through the same scaling,
$\text{CI}_{\text{Linearity}} = |\text{CI}_{\hat{\beta}_1}| \cdot PV$.

**Slope significance.** Under $H_0: \beta_1 = 0$ the test statistic is
$t = \hat{\beta}_1 / \mathrm{SE}(\hat{\beta}_1)$ on $\nu = n - 2$ degrees of
freedom (where $n$ is the number of individual bias points, here 60). The
$100(1-\alpha)\%$ confidence interval for the slope is

$$
\hat{\beta}_1 \pm t_{1-\alpha/2,\,\nu}\cdot \mathrm{SE}(\hat{\beta}_1)
$$

**Acceptance criteria** used by `iQualityR.msa` (consistent with AIAG MSA
4th ed. and the Minitab Gage Linearity and Bias Study guidance):

- $|\hat{\beta}_1| < 0.1$ (slope magnitude below the linearity tolerance)
- $\%\text{Bias}_{max} < 10\%$ of $PV$ (worst single-reference bias within
  process variation)
- Slope not statistically significant ($p \geq \alpha$): no detectable
  magnitude-dependent bias

$R^2$ is **not** an acceptance criterion: a high $R^2$ actually indicates a
*strong* linearity effect (the linear model explains the bias pattern well),
so using $R^2 \geq 0.95$ as a pass threshold would be backwards. $R^2$ is
reported as a diagnostic indicator only.

The $\%\text{Bias}$ and $\%\text{Linearity}$ metrics are reported relative
to the **process variation** $PV$, not the tolerance $T$. A gage is judged
to have a linearity problem when the slope is statistically significant
and/or $|\hat{\beta}_1|$ exceeds the 0.1 slope tolerance, even if the
percent-of-process-variation metrics are modest.

## 3. Data Preparation

The bearing-diameter dataset ships with the package at
`inst/extdata/minitab_msa/轴承直径.csv`. It contains 60 rows and 3 columns
(`部件` = part id, `主水平` = reference main level, `响应` = response
measurement).

```{r data-prep}
bearing_path <- system.file("extdata/minitab_msa/轴承直径.csv", package = "iQualityR.msa")
bearing_data <- read.csv(bearing_path, fileEncoding = "UTF-8")

cat("Rows:", nrow(bearing_data), " Cols:", ncol(bearing_data), "\n")
cat("Original column names:", names(bearing_data), "\n")

# Rename to English for readability
names(bearing_data) <- c("Part", "Reference", "Measurement")
head(bearing_data)
```

Before building the analysis, inspect the actual reference values present in
the data — do **not** assume them.

```{r ref-values}
(unique_refs <- unique(bearing_data$Reference))
```

The five reference values are `r paste(unique_refs, collapse = ", ")`, which
match the planned main levels (2, 4, 6, 8, 10 mm).

The analyzer API expects lower-case `reference` and `measurement` column
names (the matching is case-sensitive). We therefore build an
analysis-ready copy with those names:

```{r analysis-ready}
bearing_data <- data.frame(
  part        = bearing_data$Part,
  reference   = bearing_data$Reference,
  measurement = bearing_data$Measurement
)

str(bearing_data)
```

The data frame now has 60 rows, 3 columns, and the two columns the analyzer
requires.

## 4. Quick Start (One-Liner)

The fastest path is the `iqr_linearity_bias()` convenience wrapper. It
auto-detects the reference values if omitted, constructs the analyzer, runs
the full pipeline, and returns a populated `Type1Task`. Here we pass the
verified reference values and the **process variation** explicitly. The
process variation ($PV = 16.5368$ mm, taken as $6\sigma$ from a historical
capability study) is required by Minitab's linearity formula
$\text{Linearity} = |\hat{\beta}_1| \cdot PV$.

```{r quick-start}
ref_vals <- unique(bearing_data$reference)

task <- iqr_linearity_bias(
  bearing_data,
  reference_values = ref_vals,
  lsl = 0,
  usl = 12,
  process_variation = 16.5368
)

task$summary()
```

The summary mirrors the Minitab Gage Linearity and Bias Study layout: an
overall statistics block (with percent-of-process-variation column) and a
regression block carrying the slope, its 95% confidence interval, the
linearity and its CI, and the $R^2$ with PASS/FAIL decisions.

## 5. Step-by-Step Workflow

For auditability and reproducibility, the same analysis can be expressed
explicitly as Plan → Task → compute → summary → plot → report. This is the
recommended pattern for regulated environments where every step must be
recorded.

```{r step-by-step}
plan <- Type1Plan$new(
  study_type       = "linearity",
  reference_values = ref_vals,
  spec_limits      = list(lsl = 0, usl = 12),
  process_variation = 16.5368
)
plan$validate()
plan$generate_protocol()

task <- Type1Task$new(data = bearing_data, plan = plan, study_type = "linearity")
task$compute()
task$summary()
```

`Type1Plan$validate()` enforces that `reference_values` and both spec limits
are present and that `usl > lsl`. `process_variation` is a required input
for linearity studies (per Minitab's formula
$\text{Linearity} = |\hat{\beta}_1| \cdot PV$); it may be a numeric value
(typically $6\sigma$ from a capability study), the string `"from_study"`
to use $6 \cdot \text{sd}(\text{measurements})$, or `NULL` to fall back to
that same value with a warning. `generate_protocol()` prints a short,
audit-ready summary of the study configuration. The `task$results` structure
produced here is identical to the one produced by the convenience wrapper.

## 6. Public Method Deep-Dive (Refactored API)

The analysis pipeline is exposed as three discrete public methods on
`Type1LinearityAnalyzer`:

1. `compute_per_reference_summary(dt)` — per-reference counts, means, sds, biases and aggregate metrics
2. `fit_linearity_regression()` — `lm(bias ~ reference)`, linearity, %linearity, slope CI, $R^2$
3. `evaluate_linearity_criteria()` — pass/fail diagnostics against the plan criteria

Each step is independently callable, returns its results as a named list,
and stores the same values in `self$results$statistics` (or
`self$results$diagnostics` for step 3). This is useful when building custom
workflows, debugging an unexpected result, or unit-testing one step at a
time.

### Step 1 — `compute_per_reference_summary()`

```{r step-1}
analyzer <- Type1LinearityAnalyzer$new()
analyzer$setup(plan$to_list())

s1 <- analyzer$compute_per_reference_summary(data.table::as.data.table(bearing_data))

# Per-reference table (also stored in results$data_tables$ref_summary).
# Each row carries n, mean, sd, bias, se_bias, t_stat, df, p_value and the
# 95% CI for the bias at that reference (Minitab per-reference bias test).
s1$ref_summary

# Aggregate metrics returned by this step
data.frame(
  Metric = c("n_total", "n_ref_points", "tolerance", "process_variation",
             "avg_bias", "max_bias",
             "percent_avg_bias (% of PV)", "percent_max_bias (% of PV)",
             "avg_bias_t", "avg_bias_df", "avg_bias_p",
             "avg_sd", "avg_cv"),
  Value = c(s1$n_total, s1$n_ref_points, fmt_num(s1$tolerance),
            fmt_num(s1$process_variation),
            fmt_num(s1$avg_bias), fmt_num(s1$max_bias),
            fmt_pct(s1$percent_avg_bias), fmt_pct(s1$percent_max_bias),
            fmt_num(s1$avg_bias_t, 4), s1$avg_bias_df,
            formatC(s1$avg_bias_p, format = "f", digits = 6),
            fmt_num(s1$avg_sd), fmt_num(s1$avg_cv))
)
```

After this call, `analyzer$results$statistics` carries `n_total`,
`n_ref_points`, `tolerance`, `process_variation`, `avg_bias`, `max_bias`,
`percent_avg_bias`, `percent_max_bias` (all `%` metrics are relative to the
process variation $PV$), the overall average-bias $t$-test fields
(`avg_bias_se`, `avg_bias_t`, `avg_bias_df`, `avg_bias_p`, `avg_bias_ci`),
`avg_sd`, and `avg_cv`. `analyzer$results$data_tables$ref_summary` holds
the per-reference table (with per-reference `t_stat`, `p_value`, `df`,
`ci_lower`, `ci_upper`), and `analyzer$results$data_tables$bias_long`
holds the $n = 60$ individual-bias table that feeds the regression.

### Step 2 — `fit_linearity_regression()`

```{r step-2}
s2 <- analyzer$fit_linearity_regression()

data.frame(
  Term = c("Intercept", "Slope", "SE(slope)", "t(slope)", "p(slope)", "df(slope)",
           "Slope 95% CI low", "Slope 95% CI high",
           "R-squared", "Adj. R-squared", "S (regression)",
           "Linearity", "%Linearity",
           "Linearity 95% CI low", "Linearity 95% CI high"),
  Value = c(fmt_num(s2$intercept), fmt_num(s2$slope),
            fmt_num(s2$se_slope), fmt_num(s2$t_slope, 4), formatC(s2$p_slope, format = "f", digits = 6),
            s2$df_slope,
            fmt_num(s2$ci_slope[1]), fmt_num(s2$ci_slope[2]),
            fmt_num(s2$r_squared, 6), fmt_num(s2$adj_r_squared, 6),
            fmt_num(s2$s_regression, 6),
            fmt_num(s2$linearity), fmt_pct(s2$percent_linearity),
            fmt_num(s2$ci_linearity[1]), fmt_num(s2$ci_linearity[2]))
)
```

The regression is fit on the $n = 60$ individual measurement biases, so
the slope has `r s2$df_slope` degrees of freedom. The linearity CI is
derived from the slope CI via $\text{CI}_{\text{Linearity}} =
|\text{CI}_{\hat{\beta}_1}| \cdot PV$. The fitted model object `lm_model`
is also stored in `analyzer$results$raw_output$lm_model` (after a full
`run()`) and returned here as `s2$lm_model` for further diagnostics
(see Section 7).

### Step 3 — `evaluate_linearity_criteria()`

```{r step-3}
s3 <- analyzer$evaluate_linearity_criteria()

s3
```

The diagnostic verdict is summarised below.

| Criterion                       | Threshold     | Observed                      | Decision |
|---------------------------------|---------------|-------------------------------|:--------:|
| Slope magnitude                 | $< 0.1$       | `r fmt_num(s2$slope)`           | `r ifelse(s3$slope_ok, "PASS", "FAIL")` |
| $R^2$ (diagnostic only)         | —             | `r fmt_num(s2$r_squared, 6)`    | info |
| %Max bias of process variation  | $< 10\%$      | `r fmt_pct(s1$percent_max_bias)` | `r ifelse(s3$percent_max_bias_ok, "PASS", "FAIL")` |

### Equivalence: full `run()` vs step-by-step pipeline

The inherited `IqrAnalyzerBase$run()` method delegates to
`private$.run_logic()`, which simply calls the three public step methods in
order. The two paths must produce identical `statistics` and `diagnostics` —
verified below:

```{r equivalence}
dt <- data.table::as.data.table(bearing_data)

# Full pipeline
a_full <- Type1LinearityAnalyzer$new()
a_full$setup(plan$to_list())
a_full$run(dt)

# Step-by-step
a_step <- Type1LinearityAnalyzer$new()
a_step$setup(plan$to_list())
a_step$compute_per_reference_summary(dt)
a_step$fit_linearity_regression()
a_step$evaluate_linearity_criteria()

identical(a_full$results$statistics,  a_step$results$statistics)
identical(a_full$results$diagnostics, a_step$results$diagnostics)
```

Both checks return `TRUE`, confirming that the refactored public API and the
orchestrated `run()` produce byte-identical results. (The full `run()`
additionally assembles a `raw_output` bundle consumed by the plotter; the
step-by-step path skips that bookkeeping step.)

## 7. Interpreting the Regression

The regression of the $n = 60$ individual measurement biases on the
reference value is the heart of the study. Four quantities carry the
engineering meaning:

- **Slope ($\hat{\beta}_1$ = `r fmt_num(s2$slope)`)** — the constant change
  in bias per unit increase in the reference value. A negative slope means
  the gage increasingly *under-reads* as the part size grows. The slope
  magnitude (`r fmt_num(abs(s2$slope))`) exceeds the 0.1 tolerance, so the
  linearity effect is non-negligible.
- **Intercept ($\hat{\beta}_0$ = `r fmt_num(s2$intercept)`)** — the
  extrapolated bias at reference = 0. It is large and positive here because
  the regression line is fit over 2–10 mm; the intercept should not be
  interpreted as a physical bias at zero, only as the anchor of the line.
- **$R^2$ (`r fmt_num(s2$r_squared, 6)`)** — the fraction of the
  variation in individual bias explained by the linear trend on reference
  value. At `r fmt_num(s2$r_squared, 6)` the linear model captures most but
  not all of the bias structure; it falls well short of the 0.95 threshold.
- **Slope 95% CI (`r paste0("[", fmt_num(s2$ci_slope[1]), ", ", fmt_num(s2$ci_slope[2]), "]")`)** —
  the interval excludes 0, so the slope is statistically significant at the
  5% level ($t$ = `r fmt_num(s2$t_slope, 4)`, $p$ = `r formatC(s2$p_slope, format = "f", digits = 6)`,
  on `r s2$df_slope` df). This is the formal evidence of a linearity
  effect.

Because $\text{Linearity} = |\hat{\beta}_1| \cdot PV$, the linearity and
its 95% CI inherit the slope's significance: linearity =
`r fmt_num(s2$linearity)`, 95% CI
`r paste0("[", fmt_num(s2$ci_linearity[1]), ", ", fmt_num(s2$ci_linearity[2]), "]")`,
which is `r fmt_pct(s2$percent_linearity)` of the process variation.

The underlying `lm` object is available at `task$results$raw_output$lm_model`
for any further diagnostics not wrapped by the analyzer. The chunk below
pulls it out and inspects the coefficient table, the residual vector, and a
normal QQ plot of the residuals.

```{r lm-diagnostics, fig.width=7, fig.height=5}
lm_model <- task$results$raw_output$lm_model

# Coefficient table with t- and p-values
summary(lm_model)$coefficients

# Residuals (one per individual measurement, n = 60)
residuals(lm_model)

# Normal QQ plot of the regression residuals
qqnorm(residuals(lm_model), main = "Normal Q-Q Plot of Linearity Residuals")
qqline(residuals(lm_model), col = "red", lwd = 2)
```

Because the regression uses all 60 individual biases, the QQ plot has
enough points to be a meaningful check on the normality assumption
underlying the slope $t$-test; a strong departure from normality would
undermine the slope significance test.

## 8. Visualization Gallery

The `Type1Plotter` exposes five plot types for a linearity study:

| `type`          | Description                                                        |
|-----------------|--------------------------------------------------------------------|
| `summary`       | Combined patchwork of linearity + ref_biases + repeatability       |
| `list`          | List of the three individual ggplot objects                        |
| `linearity`     | Bias vs. reference with fitted regression line and zero-bias line  |
| `ref_biases`    | Bar chart of bias at each reference value                          |
| `repeatability` | Bar chart of within-reference standard deviation                   |

Use `print(task$plot(type = "..."))` to guarantee the ggplot/patchwork
object is rendered inside an R Markdown document.

```{r plot-summary, fig.width=8, fig.height=9}
print(task$plot(type = "summary"))
```

The summary panel stacks the three individual charts. The linearity panel
(top) is the most diagnostic: the fitted line slopes clearly downward,
crossing zero near the middle of the reference range.

```{r plot-linearity}
print(task$plot(type = "linearity"))
```

The standalone linearity plot makes the magnitude dependence explicit:
bias is positive at the low end (2 mm) and negative at the high end
(10 mm), a classic gage-linearity signature.

```{r plot-ref-biases}
print(task$plot(type = "ref_biases"))
```

The per-reference bias bar chart quantifies what the regression averages
over. The bias swings from about `r fmt_num(max(s1$ref_summary$bias))` mm
down to about `r fmt_num(min(s1$ref_summary$bias))` mm across the range.

```{r plot-repeatability}
print(task$plot(type = "repeatability"))
```

The repeatability chart shows the within-reference standard deviation at
each main level. Repeatability is uneven — the reference at 10 mm is
markedly more scattered than the one at 8 mm — which is a separate concern
from linearity but worth flagging for the gage owner.

```{r plot-list}
# `list` returns the three ggplot objects without combining them
pl <- task$plot(type = "list")
names(pl)
```

## 9. Report Export

`Type1Task$report()` supports Excel, HTML, PDF, and Word output. The Excel
export writes a multi-sheet workbook (a Statistics sheet with all metrics
plus a Reference_Summary sheet with the per-reference table). The HTML
export embeds all plots and is suitable for sharing with reviewers who do
not have R installed.

```{r report-excel}
xlsx_path <- tempfile(fileext = ".xlsx")
task$report(format = "excel", path = xlsx_path)

data.frame(
  Format = "Excel",
  Path   = xlsx_path,
  Exists = file.exists(xlsx_path),
  Size_KB = round(file.info(xlsx_path)$size / 1024, 1)
)
```

```{r report-html}
html_path <- tempfile(fileext = ".html")
task$report(format = "html", path = html_path)

data.frame(
  Format = "HTML",
  Path   = html_path,
  Exists = file.exists(html_path),
  Size_KB = round(file.info(html_path)$size / 1024, 1)
)
```

```{r report-pdf, eval=FALSE}
# PDF requires a working LaTeX installation; uncomment when needed.
# pdf_path <- tempfile(fileext = ".pdf")
# task$report(format = "pdf", path = pdf_path)
```

## 10. Interpretation and Decision

For the bearing-diameter data, with `process_variation = 16.5368` and
`tolerance = 12`, the analyzer produces:

| Quantity                              | Value                          |
|---------------------------------------|--------------------------------|
| Total measurements                    | `r task$results$statistics$n_total`               |
| Reference points                      | `r task$results$statistics$n_ref_points`           |
| Tolerance $T$                         | `r fmt_num(task$results$statistics$tolerance)`     |
| Process variation $PV$                | `r fmt_num(task$results$statistics$process_variation)` |
| Average bias                          | `r fmt_num(task$results$statistics$avg_bias)`      |
| Max absolute bias                     | `r fmt_num(task$results$statistics$max_bias)`      |
| %Avg bias of process variation        | `r fmt_pct(task$results$statistics$percent_avg_bias)`  |
| %Max bias of process variation        | `r fmt_pct(task$results$statistics$percent_max_bias)`  |
| Linearity ($|\hat{\beta}_1| \cdot PV$)            | `r fmt_num(task$results$statistics$linearity)`     |
| Linearity 95% CI                      | `r paste0("[", fmt_num(task$results$statistics$ci_linearity[1]), ", ", fmt_num(task$results$statistics$ci_linearity[2]), "]")` |
| %Linearity                            | `r fmt_pct(task$results$statistics$percent_linearity)` |
| Intercept $\hat{\beta}_0$             | `r fmt_num(task$results$statistics$intercept)`     |
| Slope $\hat{\beta}_1$                 | `r fmt_num(task$results$statistics$slope)`         |
| Slope df                              | `r task$results$statistics$df_slope`               |
| Slope 95% CI                          | `r paste0("[", fmt_num(task$results$statistics$ci_slope[1]), ", ", fmt_num(task$results$statistics$ci_slope[2]), "]")` |
| $R^2$                                 | `r fmt_num(task$results$statistics$r_squared, 6)`  |
| S (regression residual std dev)       | `r fmt_num(task$results$statistics$s_regression, 6)` |

**Engineering conclusion.** The gage has a **linearity problem**. The slope
of bias on reference is `r fmt_num(task$results$statistics$slope)` mm/mm:
for every additional millimetre of part diameter the gage under-reads by
about `r fmt_num(abs(task$results$statistics$slope))` mm. The slope is
statistically significant ($p$ = `r formatC(s2$p_slope, format = "f", digits = 6)`,
95% CI `r paste0("[", fmt_num(task$results$statistics$ci_slope[1]), ", ", fmt_num(task$results$statistics$ci_slope[2]), "]")`,
which excludes 0), and its magnitude
(`r fmt_num(abs(task$results$statistics$slope))`) exceeds the 0.1 linearity
tolerance, so the `slope_ok` criterion fails. The linearity is
`r fmt_num(task$results$statistics$linearity)` mm, which is
`r fmt_pct(task$results$statistics$percent_linearity)` of the process
variation — well above the 10% level that would normally be considered
acceptable. The $R^2$ of `r fmt_num(task$results$statistics$r_squared, 6)`
indicates that the linear model explains about
`r fmt_pct(task$results$statistics$r_squared * 100, 1)` of the variation in
individual biases; this is a diagnostic indicator (a high $R^2$ would
indicate a *strong* linearity effect, not a pass), not an acceptance
criterion. The percent-max-bias
criterion (`r fmt_pct(task$results$statistics$percent_max_bias)` $< 10\%$
of $PV$) passes, but only because the process variation is large relative
to the per-reference biases; it does not exonerate the gage, since the
linearity slope itself is the decisive failure.

The practical implication is that a single bias correction cannot make this
gage acceptable across the full 2–10 mm range: a correction that removes
the positive bias at 2 mm would deepen the negative bias at 10 mm, and vice
versa.

**Recommended actions** (in order of preference):

1. **Recalibrate the gage** across the operating range. The systematic,
   magnitude-dependent bias suggests a calibration curve that has drifted
   or was never established correctly for this gage.
2. **Restrict the operating range.** If recalibration is not immediately
   feasible, limit the gage to the part of the range where the bias is
   acceptable (here, the middle references near 6 mm), and use a different
   gage for the extremes.
3. **Compensate with a calibration curve.** Apply the fitted regression
   ($\hat{\text{bias}} = `r fmt_num(task$results$statistics$intercept)` `r fmt_num(task$results$statistics$slope)` \cdot r$)
   as a software correction to each measured value. This is a valid
   stop-gap but shifts the burden to maintaining the correction table; it
   should be re-verified periodically.
4. **Investigate repeatability at the high end.** The 10 mm reference also
   shows the largest within-reference scatter; the linearity and
   repeatability issues may share a root cause (probe wear, fixture
   geometry, or contact-pressure variation with part size).

**Verification against Minitab.** The Minitab Gage Linearity and Bias
Study help uses this same bearing-diameter example with $PV = 16.5368$. The
table below compares the `iQualityR.msa` output against the values Minitab
documents for this dataset; all match to display precision.

| Quantity                         | iQualityR.msa                                          | Minitab   | Match |
|----------------------------------|--------------------------------------------------------|-----------|:-----:|
| Regression sample size $n$       | `r task$results$statistics$n_total` (individual biases) | 60        | ✓     |
| Intercept $\hat{\beta}_0$        | `r fmt_num(task$results$statistics$intercept)`         | 0.736667  | ✓     |
| Slope $\hat{\beta}_1$            | `r fmt_num(task$results$statistics$slope)`             | -0.131667 | ✓     |
| $R^2$                            | `r fmt_num(task$results$statistics$r_squared, 4)`      | 0.7143    | ✓     |
| Linearity $=|\hat{\beta}_1|\cdot PV$ | `r fmt_num(task$results$statistics$linearity)`      | 1.966     | ✓     |
| %Linearity $=|\hat{\beta}_1|\cdot 100$ | `r fmt_pct(task$results$statistics$percent_linearity)` | 13.17% | ✓     |
| Avg bias                         | `r fmt_num(task$results$statistics$avg_bias)`          | -0.053333 | ✓     |

Both tools agree that the regression slope is significant, the gage
exhibits a clear linearity effect, and the recommendation is to
investigate the gage calibration rather than to apply a single bias
offset. Because `iQualityR.msa` now uses the same Minitab formulas
(regression on individual biases, $\text{Linearity} = |\hat{\beta}_1|
\cdot PV$, %Linearity and %Bias relative to $PV$), the conclusion is
identical in both tools: the gage is not linear across the range and must
be recalibrated, range-restricted, or software-compensated before it can
be released for routine use.

## 11. Reference

- Minitab Support, *Gage Linearity and Bias Study overview* —
  https://support.minitab.com/en-us/minitab/help-and-how-how/quality-and-process-improvement/measurement-system-analysis/how-to/gage-linearity-and-bias-study/
- Minitab Support, *Methods and formulas for Gage Linearity and Bias Study* —
  https://support.minitab.com/en-us/minitab/help-and-how-how/quality-and-process-improvement/measurement-system-analysis/how-to/gage-linearity-and-bias-study/methods-and-formulas/
- AIAG **Measurement Systems Analysis** (MSA) Reference Manual, 4th edition,
  Chapter III, Section B: Gage Linearity and Bias Study.
- VDA **Volume 5**, *Capability of Measurement Processes*, 3rd edition —
  measurement system uncertainty and linearity considerations.
- ISO 22568-7 / ISO/TS 21748 — complementary guidance on measurement
  uncertainty and bias evaluation.
