---
title: "Batch Gage R&R Study for Multi-Specification Parts"
author: "iQualityR MSA Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Batch Gage R&R Study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

library(iQualityR.msa)

# Formatting helpers
fmt4 <- function(x) {
  ifelse(is.na(x), "NA", formatC(as.numeric(x), format = "f", digits = 4))
}
fmt2 <- function(x) {
  ifelse(is.na(x), "NA", formatC(as.numeric(x), format = "f", digits = 2))
}
fmt_pct <- function(x, digits = 2) {
  ifelse(is.na(x), "NA", paste0(formatC(as.numeric(x), format = "f", digits = digits), "%"))
}
```

## 1. Business Context

A CNC machining shop produces a complex mechanical assembly with multiple
critical dimensions on the same part. The customer requires the measurement
system for **each** critical dimension to be independently qualified before
serial production can be released. The dimensions span very different scales:

- **Small precision features** (1--10 mm range): a 5.0 mm bore diameter and a
  3.0 mm pin diameter, both held to tight unilateral tolerances.
- **Medium features** (20--50 mm range): a 32.0 mm hole-center distance and a
  40.0 mm width.
- **Large features** (100--200 mm range): a 150.0 mm overall length and a
  180.0 mm height.

Each dimension is measured on the same coordinate measuring machine (CMM) by
three operators (Op1, Op2, Op3). Ten parts are sampled from the production run,
and each operator measures each part three times for **each** dimension. The
customer requires a single Gage R&R report covering all six dimensions in one
batch.

The six critical dimensions and their specifications are summarised below.

| SpecID | Name | Nominal (mm) | LSL | USL | Tolerance | Range Category |
|:---:|:---|---:|---:|---:|---:|:---|
| D1 | Bore diameter | 5.0 | 4.95 | 5.05 | 0.10 | Small (1-10) |
| D2 | Pin diameter | 3.0 | 2.98 | 3.02 | 0.04 | Small (1-10) |
| D3 | Hole center distance | 32.0 | 31.90 | 32.10 | 0.20 | Medium (20-50) |
| D4 | Width | 40.0 | 39.85 | 40.15 | 0.30 | Medium (20-50) |
| D5 | Overall length | 150.0 | 149.50 | 150.50 | 1.00 | Large (100-200) |
| D6 | Height | 180.0 | 179.60 | 180.40 | 0.80 | Large (100-200) |

The tolerances range from 0.04 mm (pin diameter) to 1.00 mm (overall length) --
a 25x spread. Because the same CMM is used for every dimension, the central
engineering question is whether a single instrument can qualify both the
tightest and the loosest tolerance in one study.

## 2. Method and Formulas

### 2.1 Batch analysis strategy

The batch analysis is performed by `multi_spec_msa()`. It splits the input data
frame by the `SpecID` column and runs a separate crossed ANOVA Gage R&R study
for each specification. Each sub-study returns a full `MsaTask` object -- with
its own ANOVA table, variance components, gage evaluation, plots and report --
so the per-dimension results are statistically independent and can be compared
on equal footing.

A key implementation detail: `multi_spec_msa()` creates a study plan without
specification limits for each spec, so the per-spec tolerance (needed for the
**%Tolerance** metric) must be attached to each task's plan after the initial
call, and `compute()` re-run. This is shown in section 4.

### 2.2 Crossed Gage R&R variance decomposition

Each per-spec study uses the two-way crossed ANOVA model described in detail in
the `vignette("gage-rr-crossed", package = "iQualityR.msa")` vignette. In
brief, for $o$ operators, $p$ parts and $r$ replications per cell:

$$
y_{ijk} = \mu + P_i + O_j + (PO)_{ij} + \varepsilon_{ijk}
$$

The ANOVA method estimates variance components for Repeatability
($\sigma^2_e$), the Operator:Part interaction ($\sigma^2_{PO}$), the Operator
($\sigma^2_O$) and the Part ($\sigma^2_P$). These combine into:

$$
\sigma^2_{\text{Total Gage R\&R}} = \sigma^2_{\text{Repeatability}} + \sigma^2_{\text{Reproducibility}}, \quad
\sigma^2_{\text{Total Variation}} = \sigma^2_{\text{Total Gage R\&R}} + \sigma^2_{\text{Part-to-Part}}
$$

The derived comparison metrics are:

- **%Study Var (%SV)** = $\dfrac{6 \cdot \sigma_{\text{component}}}{6 \cdot \sigma_{\text{total}}} \times 100$
- **%Tolerance** = $\dfrac{6 \cdot \sigma_{\text{component}}}{\text{USL} - \text{LSL}} \times 100$
- **NDC** = $\left\lfloor \sqrt{2}\,\dfrac{\sigma_{\text{Part}}}{\sigma_{\text{Gage}}} \right\rfloor$

### 2.3 Why %Tolerance is the right metric for multi-spec parts

When several dimensions are measured with the same instrument, **%SV** tends to
be similar across dimensions if the part-to-part variation scales with the
feature size (which it usually does, because larger features have larger
manufacturing variation). This makes %SV a poor discriminator for deciding
whether the gage is fit for each individual tolerance.

**%Tolerance**, by contrast, normalises the gage study variation against the
*engineering* tolerance of each dimension. Because each dimension has its own
LSL/USL, %Tolerance is the meaningful metric for cross-dimension comparison:
a 6-sigma gage spread of 0.024 mm is 24% of a 0.10 mm tolerance (marginal) but
only 2.4% of a 1.00 mm tolerance (acceptable). The batch study therefore
reports both metrics, and the engineering verdict for each dimension is driven
primarily by %Tolerance and %SV together with the NDC.

### 2.4 AIAG acceptance criteria

| Criterion | Accept | Marginal | Reject |
|---|---|---|---|
| %Study Var (%SV) | < 10% | 10% -- 30% | > 30% |
| NDC | >= 5 | 4 | < 4 |

The per-spec `Status` column used in this vignette classifies each dimension by
the %SV band above, so that the reader can see at a glance which dimensions
pass, are marginal, or fail the gage study.

## 3. Data Simulation

Because no public Minitab dataset covers a multi-specification batch Gage R&R
scenario, we simulate a realistic CNC machining scenario. **This is the only
vignette in the package that uses simulated data** -- the other vignettes use
the public Minitab MSA datasets shipped with the package.

The simulation models a CMM with a **fixed absolute repeatability** of
approximately 4 micrometres (`r 0.004` mm) and a small fixed operator bias of
approximately 1 micrometre. This is the physically correct model for a CMM: its
probe resolution and probing repeatability do not scale with the feature size.
Part-to-part variation, by contrast, is proportional to each dimension's
tolerance, because the manufacturing process is controlled relative to the
specification.

```{r simulate-data}
set.seed(20260725)

# Define 6 critical dimensions
dims <- data.frame(
  SpecID = c("D1", "D2", "D3", "D4", "D5", "D6"),
  Name = c("Bore diameter", "Pin diameter", "Hole center distance",
           "Width", "Overall length", "Height"),
  Nominal = c(5.0, 3.0, 32.0, 40.0, 150.0, 180.0),
  LSL = c(4.95, 2.98, 31.90, 39.85, 149.50, 179.60),
  USL = c(5.05, 3.02, 32.10, 40.15, 150.50, 180.40),
  RangeCategory = c("Small (1-10)", "Small (1-10)", "Medium (20-50)",
                    "Medium (20-50)", "Large (100-200)", "Large (100-200)"),
  stringsAsFactors = FALSE
)
dims$Tolerance <- dims$USL - dims$LSL

# Simulate measurements
operators <- c("Op1", "Op2", "Op3")
parts <- paste0("P", 1:10)
n_rep <- 3

sim_data <- do.call(rbind, lapply(1:nrow(dims), function(i) {
  d <- dims[i, ]
  # Part-to-part variation scales with tolerance (~25% of tol, manufacturing spread)
  part_bias <- rnorm(length(parts), 0, 0.25 * d$Tolerance)
  # Operator bias: fixed ~1 um (CMM positioning/probe approach, NOT feature-scaled)
  op_bias <- rnorm(length(operators), 0, 0.001)
  # CMM repeatability: fixed ~4 um absolute precision (probe resolution)
  rep_sd <- 0.004

  rows <- expand.grid(Part = parts, Operator = operators, Trial = 1:n_rep)
  rows$MeasurementValue <- d$Nominal +
    part_bias[match(rows$Part, parts)] +
    op_bias[match(rows$Operator, operators)] +
    rnorm(nrow(rows), 0, rep_sd)
  rows$SpecID <- d$SpecID
  rows[, c("SpecID", "Part", "Operator", "MeasurementValue")]
}))

head(sim_data)
```

A quick balance check confirms the crossed structure within each spec: 3
operators x 10 parts x 3 trials = 90 rows per dimension, 540 rows in total.

```{r balance-check}
cat("Total rows:", nrow(sim_data), "\n")
cat("SpecIDs:", paste(unique(sim_data$SpecID), collapse = ", "), "\n")
balance <- as.data.frame(table(sim_data$SpecID))
names(balance) <- c("SpecID", "Rows")
balance
cat("\nRows per SpecID x Operator x Part cell (unique):",
    paste(unique(as.data.frame(table(sim_data$SpecID,
                                     sim_data$Operator,
                                     sim_data$Part))$Freq), collapse = ", "),
    "\n")
```

## 4. Quick Start (One-Liner)

The batch analysis is a single call to `multi_spec_msa()`. The function returns
a named list of `MsaTask` objects -- one per `SpecID`.

```{r quick-start}
# Step 1: run the crossed ANOVA Gage R&R for each spec in one batch
batch_results <- multi_spec_msa(sim_data, spec_id_col = "SpecID")
names(batch_results)

# Step 2: attach per-spec tolerance (LSL/USL) and re-compute so that
# %Tolerance is populated for every dimension. multi_spec_msa() does not
# carry specification limits through the spec_id split, so we bind them
# onto each task's plan and re-run compute().
for (spec in names(batch_results)) {
  d <- dims[dims$SpecID == spec, ]
  batch_results[[spec]]$plan$add_specification(
    spec_name      = d$Name,
    spec_id        = spec,
    nominal_value  = d$Nominal,
    lsl            = d$LSL,
    usl            = d$USL
  )
  batch_results[[spec]]$compute(alpha_lim = 0.25, sigma = 6)
}

# Step 3: per-spec summary
for (spec in names(batch_results)) {
  cat("\n=== ", spec, " (", dims$Name[dims$SpecID == spec], ") ===\n", sep = "")
  batch_results[[spec]]$summary()
}
```

Each summary prints the variance-components table with `%Contribution`,
`%Study Var` and `%Tolerance`, the key decision metrics (Total Gage R&R
%Contribution, %Tolerance, NDC) and the AIAG-based interpretation verdict for
that single dimension.

The public fields available on every `batch_results[[spec]]` task are:

```{r task-fields}
# The statistics list carries the headline numbers
names(batch_results[[1]]$statistics)

# The variance_components tibble carries the full breakdown
names(batch_results[[1]]$variance_components)

# The interpretation list carries the AIAG verdict
names(batch_results[[1]]$interpretation)
```

## 5. Consolidated Results Table

To compare all six dimensions side by side, we extract the headline metrics
from each task's `variance_components` and `statistics` fields into a single
data frame. The `Status` column is derived from the %Study Var band (< 10%
Accept, 10--30% Marginal, > 30% Reject), per AIAG MSA 4th edition.

```{r consolidated-table}
# Helper: pull %Study Var for a given source from a task's variance components
get_pctSV <- function(task, source) {
  vc <- task$variance_components
  idx <- which(trimws(vc$Source) == source)
  if (length(idx) == 0) NA_real_ else vc$PercentStudyVar[idx[1]]
}

summary_df <- do.call(rbind, lapply(names(batch_results), function(spec) {
  t <- batch_results[[spec]]
  s <- t$statistics
  pSV <- get_pctSV(t, "Total Gage R&R")
  data.frame(
    SpecID              = spec,
    Dimension           = dims$Name[dims$SpecID == spec],
    Range               = dims$RangeCategory[dims$SpecID == spec],
    Tolerance           = dims$Tolerance[dims$SpecID == spec],
    TotalGageRR_pctSV   = round(pSV, 2),
    Repeatability_pctSV = round(get_pctSV(t, "Repeatability"), 2),
    Reproduc_pctSV      = round(get_pctSV(t, "Reproducibility"), 2),
    PartToPart_pctSV    = round(get_pctSV(t, "Part-to-Part"), 2),
    NDC                 = s$ndc,
    TotalGageRR_pctTol  = round(s$tolerance_percent, 2),
    Status              = dplyr::case_when(
      pSV < 10  ~ "Accept",
      pSV <= 30 ~ "Marginal",
      TRUE      ~ "Reject"
    ),
    stringsAsFactors = FALSE
  )
}))
rownames(summary_df) <- NULL
summary_df
```

Rendered as a kable for the report:

```{r consolidated-kable}
knitr::kable(summary_df,
             col.names = c("SpecID", "Dimension", "Range", "Tol (mm)",
                           "%SV\nGage R&R", "%SV\nRepeat", "%SV\nReprod",
                           "%SV\nPart", "NDC", "%Tol\nGage R&R", "Status"),
             align = c("c", "l", "l", "r", "r", "r", "r", "r", "c", "r", "c"),
             caption = "Consolidated batch Gage R&R results across 6 critical dimensions (simulated CNC CMM data).")
```

The table tells the story at a glance: the two small-feature dimensions (D1,
D2) have the highest %SV and %Tolerance, while the large features (D5, D6) are
well under the 10% acceptance threshold. D2 -- the pin diameter with the
tightest 0.04 mm tolerance -- is the only dimension that fails outright.

## 6. Comparative Analysis

### 6.1 Why the same CMM gives different verdicts per dimension

The six dimensions are measured with the same CMM, by the same three
operators, on the same ten parts. Yet the Gage R&R verdicts differ
dramatically. The reason lies in the interaction between the CMM's **fixed
absolute precision** and the **different tolerance widths**:

1. **%SV is broadly similar across the small and medium dimensions** (D1--D4
   range from ~4% to ~36%) but collapses to ~1--2% for the large dimensions.
   This is because %SV normalises against total variation, and the large
   features have very large part-to-part variation relative to the fixed gage
   noise. The CMM's 4 um repeatability is invisible next to 250 um of
   part-to-part spread on the 150 mm overall length.

2. **%Tolerance varies dramatically** because each dimension has its own
   tolerance. The same 6-sigma gage spread of roughly 0.024 mm is:
   - 67.9% of D2's 0.04 mm tolerance (reject),
   - 29.6% of D1's 0.10 mm tolerance (marginal),
   - 6.9% of D4's 0.30 mm tolerance (accept),
   - only 2.6% of D5's 1.00 mm tolerance (accept).

3. **NDC follows the same pattern**: D2 distinguishes only ~3 distinct
   categories (below the AIAG threshold of 5), while D5 distinguishes over 100.
   The CMM can clearly separate parts on the large features but cannot
   reliably separate them on the pin diameter.

4. **Small features may FAIL while large features PASS**. This is the central
   finding of the batch study: a single instrument is not automatically
   qualified for every dimension just because it passes on one. The
   tightest-tolerance dimension (D2) drives the qualification decision.

### 6.2 Visual comparison

The grouped bar chart below plots %SV and %Tolerance for Total Gage R&R side by
side across all six dimensions, with the 10% (accept) and 30% (marginal)
reference lines overlaid.

```{r comparative-plot, fig.width=9, fig.height=5}
plot_df <- summary_df |>
  dplyr::select(Dimension, SpecID, Range, TotalGageRR_pctSV, TotalGageRR_pctTol) |>
  tidyr::pivot_longer(
    cols      = c(TotalGageRR_pctSV, TotalGageRR_pctTol),
    names_to  = "Metric",
    values_to = "Percent"
  ) |>
  dplyr::mutate(
    Metric = factor(dplyr::recode(Metric,
      TotalGageRR_pctSV  = "% Study Var",
      TotalGageRR_pctTol = "% Tolerance"
    ), levels = c("% Study Var", "% Tolerance")),
    Dimension = factor(Dimension, levels = dims$Name)
  )

ggplot2::ggplot(plot_df, ggplot2::aes(x = Dimension, y = .data$Percent, fill = .data$Metric)) +
  ggplot2::geom_col(position = ggplot2::position_dodge(width = 0.8), alpha = 0.9, width = 0.7) +
  ggplot2::geom_hline(yintercept = 10, color = "#228833", linetype = "dashed", linewidth = 0.5) +
  ggplot2::geom_hline(yintercept = 30, color = "#EE6677", linetype = "dashed", linewidth = 0.5) +
  ggplot2::annotate("text", x = 0.6, y = 11, label = "10% accept", color = "#228833", size = 3, hjust = 0) +
  ggplot2::annotate("text", x = 0.6, y = 31, label = "30% reject", color = "#EE6677", size = 3, hjust = 0) +
  ggplot2::scale_fill_manual(values = c("#4477AA", "#EE6677")) +
  ggplot2::labs(
    title = "Total Gage R&R: %Study Var vs %Tolerance by dimension",
    subtitle = "Same CMM, 6 critical dimensions -- %Tolerance is the discriminating metric",
    x = NULL, y = "Percent"
  ) +
  ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = c(0, 0.05))) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    legend.position = "top",
    legend.title    = ggplot2::element_blank(),
    axis.text.x     = ggplot2::element_text(angle = 25, hjust = 1),
    plot.subtitle   = ggplot2::element_text(size = 9, color = "grey30"),
    panel.grid.major.x = ggplot2::element_blank()
  )
```

Two patterns are visible:

- The **%Study Var** bars (blue) decrease monotonically from D1 to D5 as the
  feature size grows, because the fixed gage noise becomes a smaller fraction
  of the increasing part-to-part variation.
- The **%Tolerance** bars (red) decrease even more steeply, because both the
  numerator (gage study variation, roughly constant) and the denominator
  (tolerance, growing) push the ratio down. D2 is the only bar clearing the
  30% reject line on both metrics.

## 7. Decision-Making

Based on the consolidated table, the engineering recommendations are:

```{r decision-table}
decision_df <- summary_df |>
  dplyr::mutate(
    Action = dplyr::case_when(
      Status == "Reject"  ~ "Improve gage / switch instrument",
      Status == "Marginal" ~ "Conditional -- monitor, tighten gage if possible",
      Status == "Accept"   ~ "No action required"
    )
  ) |>
  dplyr::select(SpecID, Dimension, Range, Tolerance,
                TotalGageRR_pctSV, TotalGageRR_pctTol, NDC, Status, Action)
knitr::kable(decision_df,
             col.names = c("SpecID", "Dimension", "Range", "Tol (mm)",
                           "%SV", "%Tol", "NDC", "Status", "Recommended action"),
             align = c("c", "l", "l", "r", "r", "r", "c", "c", "l"),
             caption = "Per-dimension engineering recommendations from the batch Gage R&R study.")
```

**Summary of the verdict:**

- **PASS (Accept, %SV < 10%)**: D4 (Width), D5 (Overall length), D6 (Height).
  The CMM is fully qualified for these medium and large features. The gage
  noise is a small fraction of both the observed variation and the engineering
  tolerance.

- **MARGINAL (10% <= %SV <= 30%)**: D1 (Bore diameter, %SV `r fmt2(summary_df$TotalGageRR_pctSV[1])`%)
  and D3 (Hole center distance, %SV `r fmt2(summary_df$TotalGageRR_pctSV[3])`%).
  The CMM is conditionally acceptable for these dimensions. Production can
  proceed, but the gage should be monitored and the study re-run if the process
  tightens. For D1 the %Tolerance is close to the 30% line, which is the
  warning signal.

- **FAIL (Reject, %SV > 30%)**: D2 (Pin diameter, %SV `r fmt2(summary_df$TotalGageRR_pctSV[2])`%,
  %Tol `r fmt2(summary_df$TotalGageRR_pctTol[2])`%, NDC `r summary_df$NDC[2]`).
  The CMM cannot adequately distinguish parts on the 3.0 mm pin diameter with
  its 0.04 mm tolerance. The 4 um repeatability produces a 6-sigma gage spread
  that is roughly two-thirds of the tolerance, and the NDC of
  `r summary_df$NDC[2]` is below the AIAG minimum of 5.

**Engineering recommendation**: use a higher-precision instrument for the small
features -- for example a dedicated air gauge, a laser micrometre, or a CMM
with a higher-resolution probe and tighter temperature control. The current CMM
is acceptable for the medium features (D3, D4) and excellent for the large
features (D5, D6). The pin diameter (D2) should not be released to production
on the current CMM without gage improvement.

## 8. Visualization per Spec

The six-panel Minitab-style summary plot is available for every dimension via
`batch_results[[spec]]$plot(type = "summary")`. Below we show two
representative dimensions: one that passes (D6, Height -- large feature) and
one that fails (D1, Bore diameter -- small feature, marginal). Replacing the
SpecID with `"D2"` shows the failing pin diameter.

### 8.1 Passing dimension -- D6 (Height, large feature)

```{r plot-d6, fig.width=10, fig.height=8}
# Large feature: CMM is well within acceptance -- PASS
print(batch_results[["D6"]]$plot(type = "summary"))
```

For D6 the components-of-variation panel shows a tiny Total Gage R&R bar
against a dominant Part-to-Part bar. The R chart is in control, the Xbar chart
shows most points outside the limits (the gage can clearly distinguish the
parts), and the operator boxplots are closely aligned (small reproducibility
bias).

### 8.2 Marginal dimension -- D1 (Bore diameter, small feature)

```{r plot-d1, fig.width=10, fig.height=8}
# Small feature with tight tolerance -- MARGINAL (close to the 30% line on %Tol)
print(batch_results[["D1"]]$plot(type = "summary"))
```

For D1 the Total Gage R&R bar is visibly larger relative to Part-to-Part, and
the %Tolerance column approaches the 30% reference line. The Xbar chart still
shows most points outside the limits (the gage can still distinguish parts),
but the margin is much thinner than for D6. This is the visual signature of a
marginal measurement system.

### 8.3 Individual plot types

Any single panel can also be rendered on its own. The available plot types are
`"variation"`, `"by_part"`, `"by_operator"`, `"interaction"`, `"xbar_chart"`
and `"r_chart"`. For example, the components-of-variation panel for the failing
dimension D2:

```{r plot-d2-variation}
print(batch_results[["D2"]]$plot(type = "variation"))
```

For D2 the Total Gage R&R bar is comparable to the Part-to-Part bar, and the
%Tolerance column clears the 30% reject line -- the visual confirmation that
the CMM is not adequate for the pin diameter tolerance.

## 9. Report Export

Every per-spec task can export a self-contained Excel report. The loop below
writes all six reports to temporary files and prints their paths. Each workbook
contains a `Variance_Components` sheet, a `Gage_Evaluation` sheet, a `Data`
sheet and the ANOVA tables.

```{r report-export}
exported_paths <- character(length(names(batch_results)))
names(exported_paths) <- names(batch_results)

for (spec in names(batch_results)) {
  out_path <- tempfile(pattern = paste0("gage_rr_", spec, "_"), fileext = ".xlsx")
  batch_results[[spec]]$report(format = "excel", path = out_path)
  exported_paths[spec] <- out_path
}

export_df <- data.frame(
  SpecID    = names(exported_paths),
  Dimension = dims$Name[match(names(exported_paths), dims$SpecID)],
  Path      = exported_paths,
  Exists    = file.exists(exported_paths),
  Size_KB   = round(file.info(exported_paths)$size / 1024, 2),
  stringsAsFactors = FALSE
)
export_df
```

The sheets inside each workbook are:

```{r report-sheets}
names(batch_results[["D2"]]$build_excel_sheets())
```

HTML, PDF and Word reports can be produced by changing the `format` argument.
Because those formats render an Rmd template, they are best generated in an
interactive R session and are not evaluated during vignette build:

```{r report-other, eval=FALSE}
for (spec in names(batch_results)) {
  batch_results[[spec]]$report(format = "html",
    path = tempfile(pattern = paste0("gage_rr_", spec, "_"), fileext = ".html"))
}
```

## 10. Reference

- AIAG Measurement Systems Analysis (MSA) Reference Manual, 4th edition --
  ANOVA method, the `alpha_lim = 0.25` interaction rule, the %GRR / NDC
  acceptance criteria, and the multi-characteristic study guidance.
- `vignette("gage-rr-crossed", package = "iQualityR.msa")` -- the companion
  vignette covering the crossed ANOVA and Xbar-R methods in detail, including
  the full variance-component formulas, the interaction-term decision rule and
  the six-panel summary plot interpretation.
- `vignette("msa-overview", package = "iQualityR.msa")` -- package-level
  overview of all MSA study types (Type 1, Gage R&R, attribute agreement).
- Minitab Help: *Crossed Gage R&R Study* -- definitions of VarComp,
  %Contribution, %Study Var, %Tolerance and NDC, which the per-spec task
  objects follow.
