---
title: "Type 1 Gage Study: Bias, Repeatability, Cg/Cgk and VDA5"
author: "iQualityR MSA Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Type 1 Gage Study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

library(iQualityR.msa)

fmt_num <- function(x, digits = 6) formatC(x, format = "f", digits = digits)
fmt_pct <- function(x, digits = 2) paste0(formatC(x, format = "f", digits = digits), "%")
```

## 1. Business Context

This vignette reproduces the canonical Minitab **Type 1 Gage Study** example:
ultrasonic thickness measurement of a protective hardcoat on a painted door
panel. An engineer has a reference sample with a known thickness of 0.025 inch,
and a single operator measures that reference 50 times in a row. The goal is to
decide whether the gage itself — independent of operator or part variation — is
capable of supporting the tolerance before any further MSA work is done.

A Type 1 study is the simplest MSA: it isolates **bias** and **repeatability**
of one gage against one reference. There are no operators, no parts, no
interactions. It is the gate that must be passed before a Gage R&R study makes
sense.

The study parameters are summarized below.

| Item                       | Value        |
|----------------------------|-------------:|
| Characteristic             | Hardcoat thickness |
| Unit                       | inch         |
| Reference value ($x_{ref}$)| 0.025000     |
| Lower specification limit  | 0.024650     |
| Upper specification limit  | 0.025350     |
| Tolerance ($T = USL - LSL$) | 0.000700     |
| Number of measurements ($n$) | 50         |
| Gage resolution            | 0.000001     |
| Default $k$ factor         | 0.2          |

At this scale (tens of micro-inches) the number of decimal places is
engineering-meaningful. Throughout this vignette, thickness, tolerance, and
standard deviation are printed with 6 decimal places, while summary statistics
(Cg, Cgk, t, p) are printed with 4.

## 2. Method and Formulas

A Type 1 study examines **only the gage itself**. The analysis has three
pillars: bias hypothesis testing, capability indices (Cg/Cgk), and VDA 5
measurement-system uncertainty.

**Bias and its hypothesis test.**

$$
\text{Bias} = \bar{x} - x_{ref}
$$

Under $H_0: \text{Bias} = 0$ the test statistic is
$t = \text{Bias} / (s/\sqrt{n})$ on $\nu = n - 1$ degrees of freedom, with the
usual two-sided $p$-value and $100(1-\alpha)\%$ confidence interval
$\text{Bias} \pm t_{1-\alpha/2,\,\nu}\cdot s/\sqrt{n}$.

**Capability indices (AIAG MSA 4th ed., Minitab Type 1).**

$$
C_g = \frac{k \cdot T}{6s}, \qquad
C_{gk} = \frac{kT/2 - |\text{Bias}|}{3s}
$$

with default $k = 0.2$. $C_g$ evaluates repeatability alone; $C_{gk}$ penalises
for bias simultaneously. The corresponding variation ratios are:

$$
\%\text{Var}(\text{Repeatability}) = \frac{6s}{T} \cdot 100 = \frac{k \cdot 100}{C_g}, \qquad
\%\text{Var}(\text{Repeatability and Bias}) = \frac{k \cdot 100}{C_{gk}}
$$

The two `%Var` metrics are the direct reciprocals of $C_g$ and $C_{gk}$ (scaled
by $k \times 100$). Note that $\%\text{Var}(\text{Repeatability and Bias})$ is
**not** $(6s + |\text{Bias}|)/T \times 100$; the correct relationship follows
the symmetric form $k \cdot 100 / C_{gk}$, which is verified against the
Minitab Type 1 example (hardcoat data: $k=0.2$, $C_{gk}=0.421$ gives
$0.2 \times 100 / 0.421 = 47.52\%$, matching Minitab's output).

**VDA 5 uncertainty (VDA volume 5, 3rd edition, §5.3).** A Type 1 Gage
Study characterises the **measurement system** (a single reading in
production), so the standard uncertainties are defined at the
single-measurement level:

$$
u_{EVR} = s \quad (\text{repeatability, single measurement; VDA 5 §5.3.1})
$$

$$
u_{RE} = \frac{\text{resolution}}{2\sqrt{3}} \quad (\text{resolution, rectangular; VDA 5 §5.3.3})
$$

$$
u_{BI} = |\text{Bias}| \quad (\text{uncorrected systematic bias; VDA 5 §5.3.2})
$$

Two points are deliberate and worth flagging. First, $u_{EVR}$ uses $s$
(not $s/\sqrt{n}$): the gage makes one reading in production, so its
repeatability contribution is the standard deviation of a single
measurement. Using $s/\sqrt{n}$ would understate $u_{MS}$ by a factor of
$\sqrt{n}$ and make $\%Q_{MS}$ inconsistent with $C_g$ (which uses
$6s$). Second, $u_{BI}$ uses $|\text{Bias}|$ (not $s/\sqrt{n}$): in a
Type 1 study the bias is being *evaluated*, not corrected, so its full
magnitude contributes to uncertainty. The statistical uncertainty of the
bias *estimate* ($s/\sqrt{n}$) is already captured in $u_{EVR}$ and must
not be double-counted.

The combined standard uncertainty of the measurement system is the root
sum of squares of its components:

$$
u_{MS} = \sqrt{u_{EVR}^2 + u_{RE}^2 + u_{BI}^2 + u_{LIN}^2 + u_{CAL}^2 + \sum u_{REST}^2}
$$

where $u_{LIN} = 0$ for a Type 1 study (linearity is assessed separately
by a Gage Linearity and Bias Study), $u_{CAL}$ is the calibration
uncertainty, and $u_{REST}$ is any remaining contributor. The ratio
against tolerance is

$$
\%Q_{MS} = \frac{2 u_{MS}}{T} \cdot 100
$$

(the factor 2 expands the standard uncertainty to a $2\sigma$ "expanded"
uncertainty consistent with VDA 5 reporting).

**Acceptance criteria** used by `iQualityR.msa` (consistent with AIAG MSA 4th
ed. and VDA 5):

- $C_g \geq 1.33$
- $C_{gk} \geq 1.33$
- $\%$Bias of tolerance $< 10\%$
- $\%Q_{MS} < 15\%$
- Bias 95% CI must contain 0

## 3. Data Preparation

The hardcoat dataset ships with the package at
`inst/extdata/minitab_msa/硬盖厚度.csv`. It is a single column of 50
thickness readings (column name `厚度`, "thickness" in Chinese).

```{r data-prep}
hardcoat_path <- system.file("extdata/minitab_msa/硬盖厚度.csv", package = "iQualityR.msa")
hardcoat <- read.csv(hardcoat_path, fileEncoding = "UTF-8")

cat("Rows:", nrow(hardcoat), " Cols:", ncol(hardcoat), "\n")
cat("Column names:", names(hardcoat), "\n")

head(hardcoat)

summary(hardcoat$厚度)
```

The 50 measurements span roughly 0.02489 to 0.02506 inch. The sample mean is
slightly below the 0.025 reference, which is the first hint of a negative
bias — formalised in the next section.

## 4. Quick Start (One-Liner)

The fastest path is the `iqr_type1_bias()` convenience wrapper. It constructs
the plan, runs the analyzer, and returns a fully populated `Type1Task`:

```{r quick-start}
task <- iqr_type1_bias(
  hardcoat$厚度,
  reference_value = 0.025,
  lsl = 0.024650,
  usl = 0.025350,
  resolution = 0.000001
)

task$summary()
```

The summary mirrors the Minitab Type 1 layout: basic statistics, gage
capability table with PASS/FAIL decisions, and VDA 5 uncertainty components.
Two capability indices (`Cg`, `Cgk`) and the VDA 5 `%QMS` carry the formal
accept/reject verdicts; everything else is diagnostic context.

## 5. Step-by-Step Workflow

For auditability and reproducibility, the same analysis can be expressed
explicitly as Plan → Task → compute → summary → plot → report. This is the
recommended pattern for regulated environments where every step must be
recorded.

```{r step-by-step}
plan <- Type1Plan$new(
  study_type     = "bias",
  reference_value = 0.025,
  spec_limits    = list(lsl = 0.024650, usl = 0.025350),
  k_factor       = 0.2,
  resolution     = 0.000001
)
plan$validate()
plan$generate_protocol()

task <- Type1Task$new(data = hardcoat$厚度, plan = plan, study_type = "bias")
task$compute()
task$summary()
```

The `Type1Plan$validate()` call enforces that a reference value and both spec
limits are present; `generate_protocol()` prints a short, audit-ready summary
of the study configuration. The same `task$results` structure is produced
whether the task is built from a plan or from the convenience wrapper.

## 6. Public Method Deep-Dive (Refactored API)

After the P1-4 refactor the analysis pipeline is exposed as four discrete
public methods on `Type1Analyzer`:

1. `compute_bias(dt)` — sample statistics and bias t-test
2. `compute_capability()` — Cg, Cgk, %Var repeatability
3. `compute_vda5()` — VDA 5 uncertainty components and %QMS
4. `evaluate_criteria()` — pass/fail diagnostics against the plan criteria

Each step is independently callable, returns its results as a named list,
and stores the same values in `self$results$statistics` (or
`self$results$diagnostics` for step 4). This is useful when building custom
workflows, debugging an unexpected result, or unit-testing one step at a
time.

### Step 1 — `compute_bias()`

```{r step-bias}
analyzer <- Type1Analyzer$new()
analyzer$setup(plan$to_list())

bias_out <- analyzer$compute_bias(data.table::as.data.table(hardcoat$厚度))

bias_out
```

After this call, `analyzer$results$statistics` carries `n`, `reference_value`,
`mean_meas`, `sd_meas`, `bias`, `percent_bias`, `t_stat`, `p_value`,
`ci_bias`, `tolerance`, and `spec_limits`. The 95% bias confidence interval is
`r sprintf("[%.6f, %.6f]", task$results$statistics$ci_bias[1], task$results$statistics$ci_bias[2])`,
which does **not** contain 0 — already foreshadowing the bias verdict.

### Step 2 — `compute_capability()`

```{r step-capability}
cap_out <- analyzer$compute_capability()

cap_out
```

`Cg = r round(task$results$statistics$Cg, 3)` and
`Cgk = r round(task$results$statistics$Cgk, 3)` — both well below the 1.33
threshold, confirming that the gage consumes too much of the tolerance even
before bias is accounted for.

### Step 3 — `compute_vda5()`

```{r step-vda5}
vda5_out <- analyzer$compute_vda5()

vda5_out

analyzer$results$data_tables$vda5_uncertainty
```

The VDA 5 table breaks down the combined uncertainty into its source
components. Per VDA 5 §5.3, $u_{EVR} = s$ (single-measurement
repeatability) and $u_{BI} = |\text{Bias}|$ (uncorrected systematic
bias), so for the hardcoat data repeatability dominates the uncertainty
budget (the gage's scatter is much larger than its bias), while
resolution is negligible (the gage resolution is 1 µin, far below the
observed scatter). `%QMS = r round(task$results$statistics$vda5_qms_percent, 2)%`
is now consistent with the failed $C_g = r round(task$results$statistics$Cg, 2)`:
both say the gage consumes a large fraction of the tolerance. `%QMS`
still passes the 15% VDA 5 threshold, but only narrowly — unlike the
$C_g$ verdict, which fails decisively. This is the expected
relationship between the two standards: VDA 5's $\%Q_{MS}$ (based on
$2u_{MS}/T$) is structurally more lenient than AIAG's $C_g \geq 1.33$
(based on $kT/(6s)$), so a gage can pass VDA 5 while failing AIAG.

### Step 4 — `evaluate_criteria()`

```{r step-criteria}
diag_out <- analyzer$evaluate_criteria()

diag_out
```

The diagnostic verdict is summarised below.

| Criterion                | Threshold | Observed           | Decision |
|--------------------------|-----------|--------------------|:--------:|
| $C_g$                    | $\geq 1.33$ | `r round(task$results$statistics$Cg, 3)` | **FAIL** |
| $C_{gk}$                 | $\geq 1.33$ | `r round(task$results$statistics$Cgk, 3)` | **FAIL** |
| Bias 95% CI contains 0   | yes       | excludes 0         | **FAIL** |
| %Bias of tolerance       | $< 10\%$    | `r paste0(round(task$results$statistics$percent_bias, 2), "%")` | PASS |
| %QMS                     | $< 15\%$    | `r paste0(round(task$results$statistics$vda5_qms_percent, 2), "%")` | PASS |

### Equivalence: full `run()` vs step-by-step pipeline

The internal `IqrAnalyzerBase$run()` method delegates to `private$.run_logic()`,
which simply calls the four public step methods in order. The two paths must
produce identical results — verified below on a synthetic dataset:

```{r equivalence}
set.seed(42)
dt <- data.table::data.table(measurement = rnorm(25, mean = 10.05, sd = 0.25))
params <- list(
  reference_value = 10, spec_limits = list(lsl = 7, usl = 13),
  k_factor = 0.2, conf_level = 0.95, resolution = 0.005,
  u_cal = 0.02, u_rest = list(),
  criteria = list(Cg_min = 1.33, Cgk_min = 1.33,
                  percent_bias_max = 10, vda5_qms_max = 15)
)

# Full pipeline
a_full <- Type1Analyzer$new(); a_full$setup(params); a_full$run(dt)

# Step-by-step
a_step <- Type1Analyzer$new(); a_step$setup(params)
a_step$compute_bias(dt)
a_step$compute_capability()
a_step$compute_vda5()
a_step$evaluate_criteria()

identical(a_full$results$statistics,  a_step$results$statistics)
identical(a_full$results$diagnostics, a_step$results$diagnostics)
```

Both checks return `TRUE`, confirming that the refactored public API and the
orchestrated `run()` produce byte-identical results.

## 7. Visualization Gallery

The `Type1Plotter` exposes five plot types for a Type 1 bias study:

| `type`       | Description                                                  |
|--------------|--------------------------------------------------------------|
| `summary`    | Combined patchwork of run chart + Cg/Cgk + VDA 5 components  |
| `list`       | List of the three individual ggplot objects                 |
| `run_chart`  | Run chart of measurements with reference and mean lines     |
| `capability` | Bar chart of Cg and Cgk against the 1.33 threshold           |
| `vda5`       | Stacked/bar chart of VDA 5 uncertainty contributions         |

Use `print(task$plot(type = "..."))` to guarantee the ggplot/patchwork
object is rendered inside an R Markdown document.

```{r plot-summary, fig.width=8, fig.height=7}
print(task$plot(type = "summary"))
```

```{r plot-runchart}
print(task$plot(type = "run_chart"))
```

The run chart shows the 50 measurements oscillating tightly around the mean
(solid green) and slightly below the dashed-red reference line. The bias CI
in the subtitle does not straddle zero — visually consistent with the
`p = 0.021` verdict.

```{r plot-capability}
print(task$plot(type = "capability"))
```

The capability plot is the single most actionable chart: both bars sit well
below the 1.33 dashed threshold, by a factor of roughly 2.5×. No further MSA
work should be done until this is fixed.

```{r plot-vda5}
print(task$plot(type = "vda5"))
```

The VDA 5 chart shows that repeatability ($u_{EVR} = s$) dominates the
combined uncertainty for this gage, with the uncorrected bias
($u_{BI} = |\text{Bias}|$) a smaller but non-negligible contributor;
resolution is essentially invisible at this gage setting. This is the
expected profile for a gage that fails $C_g$ but has only a modest bias.

## 8. Report Export

`Type1Task$report()` supports Excel, HTML, PDF, and Word output. The Excel
export writes a multi-sheet workbook (statistics + VDA 5 breakdown). The
HTML export embeds all plots and is suitable for sharing with reviewers who
do not have R installed.

```{r report-excel}
xlsx_path <- tempfile(fileext = ".xlsx")
task$report(format = "excel", path = xlsx_path)

data.frame(
  Format = "Excel",
  Path   = xlsx_path,
  Exists = file.exists(xlsx_path),
  Size_KB = round(file.info(xlsx_path)$size / 1024, 1)
)
```

```{r report-html}
html_path <- "./.html"
task$report(format = "html", path = html_path)

data.frame(
  Format = "HTML",
  Path   = html_path,
  Exists = file.exists(html_path),
  Size_KB = round(file.info(html_path)$size / 1024, 1)
)
```

```{r report-pdf, eval=FALSE}
# PDF requires a working LaTeX installation; uncomment when needed.
# pdf_path <- tempfile(fileext = ".pdf")
# task$report(format = "pdf", path = pdf_path)
```

## 9. Interpretation and Decision

For the hardcoat data, the analyzer produces:

| Quantity                     | Value            |
|------------------------------|-----------------:|
| Mean $\bar{x}$               | 0.024985         |
| StDev $s$                    | 0.000044         |
| Bias                         | -0.000015        |
| t (df = 49)                  | -2.3847          |
| p-value                      | 0.0210           |
| 95% CI for Bias              | [-0.000027, -0.000002] |
| $C_g$                        | 0.533            |
| $C_{gk}$                     | 0.421            |
| %Var (Repeatability)         | 37.50%           |
| %Var (Repeatability + Bias)  | 47.52%           |
| %Bias of tolerance           | 2.11%            |
| %QMS                         | 2.50%            |

**Engineering conclusion.** The gage is **NOT acceptable** for this
characteristic. The dominant failure mode is repeatability: $C_g = 0.533$ is
roughly 40% of the 1.33 acceptance threshold, and the %Var (Repeatability)
consumes 37.5% of the 0.0007-inch tolerance — well above the 10% rule of
thumb. Bias is statistically significant ($p = 0.021$, CI excludes 0) but
small in absolute terms (2.11% of tolerance), so eliminating it alone would
not rescue the study; $C_{gk} = 0.421$ is only marginally lower than $C_g =
0.533$.

The VDA 5 uncertainty view is more forgiving — `%QMS = 2.50%` is well under
15% — because VDA 5's $u_{EVR} = s/\sqrt{n}$ shrinks with sample size, while
$C_g$ uses the full $6s$ of an individual measurement. This divergence is
expected and is exactly why the AIAG and VDA 5 frameworks are reported side
by side: VDA 5 answers "is the measurement uncertainty acceptable for
tolerance conformance?", while $C_g$/$C_{gk}$ answer "is the gage capable
of single-measurement discrimination?".

**Verification against Minitab output.** The Minitab Type 1 Gage Study
help uses this same hardcoat dataset. The following table compares the
`iQualityR.msa` results with the Minitab session window output, confirming
that all four capability formulas match:

| Metric        | iQualityR.msa | Minitab | Match |
|---------------|---------------|---------|-------|
| $C_g$         | 0.5333        | 0.53    | ✓     |
| $C_{gk}$      | 0.4208        | 0.42    | ✓     |
| %Var (Rep)    | 37.50%        | 37.50%  | ✓     |
| %Var (Rep&Bias) | 47.52%      | 47.52%  | ✓     |

The formulas used are:

- $C_g = \frac{K \cdot T}{6s}$,  $C_{gk} = \frac{K \cdot T / 2 - |\text{Bias}|}{3s}$
- %Var (Rep) $= \frac{6s}{T} \times 100 = \frac{K \cdot 100}{C_g}$
- %Var (Rep&Bias) $= \frac{K \cdot 100}{C_{gk}}$

where $K = 0.2$ (default), $T$ = tolerance, $s$ = sample standard deviation.
The %Var (Repeatability and Bias) formula is symmetric to %Var (Repeatability)
through $C_{gk}$, exactly as specified in the Minitab *Methods and Formulas*
documentation.

**Recommended actions** (consistent with Minitab's guidance):

1. Do **not** release the gage for routine inspection of this characteristic.
2. Do **not** proceed to a Gage R&R study — it will inherit the same
   repeatability problem and produce misleading reproducibility numbers.
3. Identify and remove the dominant repeatability cause (probe, fixture,
   vibration, operator technique), then re-run the Type 1 study until
   $C_g \geq 1.33$ and $C_{gk} \geq 1.33$.
4. Once Type 1 passes, perform the Gage Linearity & Bias Study (multiple
   references) and finally the Gage R&R study.

## 10. Reference

- Minitab Support, *Type 1 Gage Study overview* —
  https://support.minitab.com/en-us/minitab/help-and-how-how/quality-and-process-improvement/measurement-system-analysis/how-to/type-1-gage-study/
- Minitab Support, *Methods and formulas for Type 1 Gage Study* —
  https://support.minitab.com/en-us/minitab/help-and-how-how/quality-and-process-improvement/measurement-system-analysis/how-to/type-1-gage-study/methods-and-formulas/
- AIAG **Measurement Systems Analysis** (MSA) Reference Manual, 4th edition,
  Chapter III-C: Type 1 Gage Study.
- VDA **Volume 5**, *Capability of Measurement Processes*, 3rd edition,
  Section 4: Measurement system uncertainty.
- ISO 22568-7 / ISO/TS 21748 — complementary guidance on measurement
  uncertainty in capability analysis.
