---
title: "ML Enhancement Charts: LSTM, Autoencoder, Isolation Forest, BOCPD, SHAP"
author: "iQualityR.spc"
date: "2026-07-06"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ML Enhancement Charts: LSTM, Autoencoder, Isolation Forest, BOCPD, SHAP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE,
  error = TRUE
)
```

```{r setup, echo = FALSE}
library(iQualityR.spc)
library(ggplot2)
set.seed(2026)
```

## Overview

This vignette covers the **AI/ML enhancement charts** introduced in
`iQualityR.spc` v0.6. These charts extend the SPC toolkit with deep
learning and Bayesian methods for anomaly detection and structural
change detection:

- **LSTM anomaly detection** -- recurrent neural network for univariate
  time series with non-linear dynamics
- **Autoencoder (AE)** -- deep neural network for multivariate
  reconstruction-error anomaly detection
- **Isolation Forest (IForest)** -- tree-based ensemble for multivariate
  anomaly detection with no distributional assumptions
- **BOCPD** -- Bayesian Online Change Point Detection for structural
  breaks in a univariate stream
- **SHAP attribution** -- post-hoc Shapley value explanation of the
  anomaly score from any of the above charts

All functions return an `IqrSpcTask` object with the standard
`summary()`, `plot()`, and `results` accessors, so they integrate
seamlessly with the rest of the iQualityR ecosystem.

### Soft dependencies

The ML charts are designed to work **with or without** their heavy ML
backends. Each chart auto-detects the active backend and reports it in
`task$results$diagnostics$backend`:

| Chart | Primary backend | Fallback backend |
|-------|-----------------|------------------|
| LSTM | `keras` (TensorFlow) | `arima_ess` (ARIMA + EWMA) |
| Autoencoder | `keras` (TensorFlow) | `pca` (SVD reconstruction) |
| Isolation Forest | `isotree` (C++) | `native` (in-house R trees) |
| BOCPD | (none required) | pure R |
| SHAP | `fastshap` (KernelSHAP) | `ablation` (drop-column delta) |

To enable the deep-learning backends, install the optional packages:

```r
# Not required; fallbacks work without them.
install.packages(c("isotree", "fastshap"))
# For keras/tensorflow, see https://keras.posit.co/ for setup.
```

CRAN submission does not require any of these packages; the Suggests
field declares them as soft dependencies, and every call site uses
`requireNamespace()` for graceful degradation.

## 1. LSTM Anomaly Detection Chart

The LSTM chart trains a recurrent neural network to predict the next
observation from a sliding window of past values. After training, the
prediction error (residual) is converted to an anomaly score on a
z-scale; values exceeding `lstm_threshold` are flagged as anomalies.

LSTM is well suited to **non-linear, autocorrelated** time series where
classical ARIMA residuals may miss subtle regime changes.

```{r}
df <- data.frame(measurement = c(rnorm(60, 100, 1),
                                   rnorm(15, 103, 1)))
task <- run_spc_lstm(df, "measurement",
                     lstm_window = 10, lstm_epochs = 5,
                     lstm_threshold = 2.5)
task$summary()
```

The points data table augments the standard SPC columns with model
outputs:

```{r}
head(task$results$data_tables$points[, c("index", "value", "fit",
                                          "reconstruction_error",
                                          "anomaly_score", "is_anomaly")])
```

**Tuning guidance**:

- `lstm_window`: lookback length. Shorter windows react faster but
  lose long-range context. Typical: 5--20.
- `lstm_units`: capacity of the recurrent layer. Larger units fit
  more complex dynamics but require more data. Typical: 16--64.
- `lstm_epochs`: more epochs reduce training loss but risk overfitting
  on small datasets. Typical: 10--50.
- `lstm_threshold`: z-scale cutoff. Lower for higher sensitivity.

## 2. Autoencoder Anomaly Detection Chart

The Autoencoder (AE) chart compresses multivariate input to a
low-dimensional latent representation and reconstructs it. Observations
with large reconstruction error are flagged as anomalies. AE is
particularly effective at capturing **correlation structure** between
variables -- a single variable can be in-control marginally yet
anomalous jointly.

```{r}
df <- data.frame(
  x1 = c(rnorm(40, 50, 1), rnorm(10, 55, 1)),
  x2 = c(rnorm(40, 30, 0.8), rnorm(10, 33, 0.8)),
  x3 = c(rnorm(40, 12, 0.5), rnorm(10, 14, 0.5))
)
task <- run_spc_autoencoder(df, ae_encoding_dim = 1, ae_epochs = 10)
task$summary()
```

Per-feature reconstruction errors are stored in the
`feature_contributions` data table and provide a first-look
explanation of *which variable drove the anomaly*:

```{r}
head(task$results$data_tables$feature_contributions)
```

**Tuning guidance**:

- `ae_encoding_dim`: bottleneck size. Smaller than the number of
  variables, typically 1--3. A smaller bottleneck is more sensitive
  but less able to capture normal correlation structure.
- `ae_threshold_quantile`: quantile of the reconstruction error used as
  the control limit. Default `0.99` flags the top 1% of training data
  as anomalous.

## 3. Isolation Forest Anomaly Detection Chart

The Isolation Forest (IForest) chart isolates observations by random
recursive partitioning. Anomalous points require fewer splits to
isolate, so they have a shorter average path length. The anomaly score
`s(x) = 2^(-E[h(x)]/c(n))` is in (0, 1), with higher values indicating
more anomalous observations.

IForest is **non-parametric** and works well on high-dimensional data
with no distributional assumptions, making it a robust baseline for
multivariate anomaly detection.

```{r}
df <- data.frame(
  x1 = c(rnorm(50, 50, 1), rnorm(10, 58, 1)),
  x2 = c(rnorm(50, 30, 0.8), rnorm(10, 35, 0.8))
)
task <- run_spc_iforest(df, iforest_ntree = 100,
                         iforest_threshold = 0.6)
task$summary()
```

**Tuning guidance**:

- `iforest_ntree`: more trees reduce variance of the anomaly score.
  Typical: 100--500.
- `iforest_sample_size`: subsample size per tree. Smaller values
  increase sensitivity to local anomalies. Default 256.
- `iforest_threshold`: anomaly score cutoff in (0, 1). Higher for
  fewer false positives.

## 4. Bayesian Online Change Point Detection (BOCPD)

BOCPD (Adams & MacKay, 2007) maintains a posterior distribution over
the **run length** -- the number of observations since the last
changepoint. At each step, it reports the probability that the current
observation is a changepoint.

The model assumes Gaussian observations with a Normal-Inverse-Gamma
conjugate prior. The **hazard rate** controls the prior expected run
length: a hazard of 1/250 means a changepoint is expected roughly
every 250 observations.

```{r}
df <- data.frame(measurement = c(rnorm(40, 100, 1),
                                   rnorm(40, 105, 1)))
task <- run_spc_bocpd(df, "measurement", bocpd_hazard = 1 / 50)
task$summary()
```

The points data table contains the run-length posterior summary:

```{r}
head(task$results$data_tables$points[, c("index", "value",
                                           "changepoint_prob",
                                           "run_length")])
```

Detected changepoints (where the posterior peaks) are also exposed as
a dedicated data table:

```{r}
print(task$results$data_tables$changepoints)
```

**Tuning guidance**:

- `bocpd_hazard`: in (0, 1). Lower values detect longer regimes;
  higher values detect more frequent changepoints. Start with
  `1 / expected_run_length`.
- `bocpd_prior_mu`, `bocpd_prior_kappa`, `bocpd_prior_alpha`,
  `bocpd_prior_beta`: conjugate prior parameters. For a process
  centered at `mu0` with prior sample weight `kappa0`, set
  `bocpd_prior_mu = mu0`, `bocpd_prior_kappa = kappa0`. Weakly
  informative defaults (`kappa = alpha = beta = 1`) work for most
  standardized data.

## 5. SHAP Attribution

`run_spc_shap()` provides post-hoc **Shapley value** attribution of
the anomaly score for any multivariate ML chart. It explains *which
variables contributed most* to each point's anomaly score, enabling
root-cause analysis.

```{r}
set.seed(123)
df <- data.frame(
  x1 = c(rnorm(40, 50, 1), rnorm(10, 55, 1)),
  x2 = c(rnorm(40, 30, 0.8), rnorm(10, 33, 0.8))
)
task <- run_spc_iforest(df, iforest_ntree = 100)
shap_result <- run_spc_shap(task, nsample = 30)
head(shap_result$shap_values)
shap_result$backend
```

The returned list contains:

- `shap_values`: a data frame with one column per feature and one row
  per observation, giving the Shapley value contribution.
- `backend`: `"fastshap"` (KernelSHAP) or `"ablation"` (drop-column
  fallback).
- `feature_names`: character vector of feature names.

**Tuning guidance**:

- `nsample`: number of Monte Carlo samples for KernelSHAP. Larger
  values give more stable estimates at the cost of runtime. Typical:
  30--100.

## 6. AI Diagnostic Structure

Every ML chart populates `task$results$diagnostics$ai_diagnostic`,
a unified list for downstream tooling:

| Field | Type | Description |
|-------|------|-------------|
| `method` | character | `"lstm"`, `"autoencoder"`, `"iforest"`, or `"bocpd"` |
| `anomaly_score` | numeric vector | Per-observation anomaly score |
| `shap_values` | data.frame | SHAP-like feature contributions |
| `rule_attribution` | character | Human-readable rule summary |
| `confidence` | numeric in [0, 1] | Model confidence, scaled by data size |
| `extras` | list | Backend-specific metadata (backend, threshold, etc.) |

```{r}
task <- run_spc_iforest(df, iforest_ntree = 100)
ai <- task$results$diagnostics$ai_diagnostic
names(ai)
ai$method
ai$confidence
```

## 7. Choosing an ML Chart

| Data characteristic | Recommended chart |
|---------------------|-------------------|
| Univariate, non-linear, autocorrelated | LSTM |
| Multivariate, correlated, joint anomalies | Autoencoder |
| Multivariate, no distributional assumptions, high-dim | Isolation Forest |
| Univariate, structural breaks, online | BOCPD |
| Any chart, root-cause attribution needed | SHAP (post-hoc) |

## 8. Combining ML Charts with Classical SPC

A typical end-to-end workflow:

1. Run `run_spc_changepoint` (v0.2) to identify historical regimes.
2. Within each regime, run `run_spc_iforest` on multivariate data to
   detect joint anomalies.
3. For each detected anomaly, run `run_spc_shap` to attribute the
   anomaly to specific variables.
4. For univariate critical-to-quality (CTQ) variables, run
   `run_spc_lstm` for high-sensitivity detection.
5. For online monitoring of slow drifts, run `run_spc_bocpd` and
   alert when the changepoint probability exceeds 0.5.

## References

- Adams, R. P., & MacKay, D. J. C. (2007). Bayesian online
  changepoint detection. *arXiv preprint arXiv:0710.3742*.
- Hochreiter, S., & Schmidhuber, J. (1997). Long short-term memory.
  *Neural Computation*, 9(8), 1735-1780.
- Hinton, G. E., & Salakhutdinov, R. R. (2006). Reducing the
  dimensionality of data with neural networks. *Science*,
  313(5786), 504-507.
- Liu, F. T., Ting, K. M., & Zhou, Z. H. (2008). Isolation forest.
  In *2008 Eighth IEEE International Conference on Data Mining*,
  pp. 413-422.
- Lundberg, S. M., & Lee, S. I. (2017). A unified approach to
  interpreting model predictions. In *Advances in Neural Information
  Processing Systems*, pp. 4765-4774.
