---
title: "Shewhart Control Charts: Variables and Attributes"
author: "iQualityR.spc"
date: "2026-07-05"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Shewhart Control Charts: Variables and Attributes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE,
  error = TRUE
)
```

```{r setup, echo = FALSE}
library(iQualityR.spc)
library(ggplot2)
set.seed(2026)

# Helper: save plot via ggsave() and display via include_graphics().
# This avoids knitr's inline graphics device which can crash in R CMD check.
iqr_plot <- function(task, type = "full", name) {
  dir.create("figures", showWarnings = FALSE)
  path <- file.path("figures", paste0(name, ".png"))
  p <- task$plot(type = type)
  ggplot2::ggsave(path, p, width = 7, height = 4.5, dpi = 100, bg = "white")
  knitr::include_graphics(path)
}
```

## Overview

This vignette demonstrates the **Shewhart control charts** implemented in
`iQualityR.spc` for variables data (Xbar-R, Xbar-S, I-MR) and attributes
data (P, NP, U, C, Laney P', Laney U'). These charts are the foundation of
Statistical Process Control (SPC) and are aligned with the Minitab
`Stat > Control Charts` menu and ISO 7870-2:2023.

The package reuses the sigma estimation, Nelson rules detection, and control
limit constants from `iQualityR.stat`, so results are consistent with the
rest of the iQualityR ecosystem.

## 1. Variables Charts

### 1.1 Xbar-R Chart

Use Xbar-R when data can be grouped into rational subgroups of size 2-10.

```{r}
df <- data.frame(
  measurement = rnorm(50, mean = 100, sd = 2),
  subgroup = rep(1:10, each = 5)
)
task <- run_spc_xbar_r(df, "measurement",
                       subgroup = "subgroup", subgroup_size = 5)
task$summary()
iqr_plot(task, "full", "spc_shewhart_xbar_r")
```

### 1.2 Xbar-S Chart

Xbar-S is preferred when subgroup sizes are larger than 10, because the
sample standard deviation is a more efficient estimator than the range.

```{r}
df <- data.frame(
  measurement = rnorm(60, mean = 50, sd = 1.5),
  subgroup = rep(1:10, each = 6)
)
task <- run_spc_xbar_s(df, "measurement",
                       subgroup = "subgroup", subgroup_size = 6)
task$summary()
```

### 1.3 I-MR Chart

I-MR (Individuals - Moving Range) is used when each subgroup has size 1.

```{r}
df <- data.frame(measurement = rnorm(30, mean = 100, sd = 1))
task <- run_spc_imr(df, "measurement")
task$summary()
iqr_plot(task, "full", "spc_shewhart_imr")
```

### 1.4 Out-of-Control Detection

When a point violates a Nelson rule, it is highlighted in red on the chart:

```{r}
df <- data.frame(measurement = c(rnorm(29, 100, 1), 105))
task <- run_spc_imr(df, "measurement")
iqr_plot(task, "summary", "spc_shewhart_ooc")
```

## 2. Attributes Charts

### 2.1 P Chart

For monitoring the proportion of defective items when sample sizes may vary:

```{r}
df <- data.frame(
  defectives = rbinom(20, 100, 0.05),
  sample_size = rep(100, 20)
)
task <- run_spc_p(df, "defectives", "sample_size")
task$summary()
```

### 2.2 NP Chart

For counts of defectives with equal sample sizes:

```{r}
df <- data.frame(
  defectives = rbinom(15, 50, 0.04),
  sample_size = rep(50, 15)
)
task <- run_spc_np(df, "defectives", "sample_size")
```

### 2.3 U Chart

For defects per unit:

```{r}
df <- data.frame(
  defects = rpois(20, lambda = 4),
  sample_size = rep(1, 20)
)
task <- run_spc_u(df, "defects", "sample_size")
```

### 2.4 C Chart

For total counts of defects:

```{r}
df <- data.frame(defects = rpois(20, lambda = 5))
task <- run_spc_c(df, "defects")
```

### 2.5 Laney P' and U' Charts

When sample sizes are very large, traditional P and U charts produce too
many false alarms because they ignore between-subgroup variation. The
Laney correction (Laney 2002) adjusts for this over-dispersion and is
recommended by AIAG-VDA SPC (2026).

```{r}
df <- data.frame(
  defectives = rbinom(25, 500, 0.05),
  sample_size = rep(500, 25)
)
task <- run_spc_p_laney(df, "defectives", "sample_size")
task$summary()
```

## 3. Nelson Rules

All charts apply the 8 Nelson rules by default. Use the `nelson_rules`
argument to enable specific rules:

```{r}
df <- data.frame(measurement = rnorm(30, 100, 1))
task <- run_spc_imr(df, "measurement", nelson_rules = c(1, 2, 3))
```

The rules are:

| Rule | Description |
|------|-------------|
| 1 | 1 point beyond 3 sigma |
| 2 | 9 consecutive points on same side of center |
| 3 | 6 consecutive points increasing or decreasing |
| 4 | 14 consecutive points alternating up and down |
| 5 | 2 of 3 points beyond 2 sigma (same side) |
| 6 | 4 of 5 points beyond 1 sigma (same side) |
| 7 | 15 consecutive points within 1 sigma (stratification) |
| 8 | 8 consecutive points outside 1 sigma (mixture) |

## 4. Sigma Estimation Methods

The package supports 7 sigma estimation methods via `iQualityR.stat`:

| Method | Code | Default for |
|--------|------|-------------|
| Range | `r_bar` | Xbar-R |
| Standard deviation | `s_bar` | Xbar-S |
| Pooled SD | `pooled_s` | I-MR-R/S |
| Moving range (mean) | `mr_bar` | I-MR, EWMA, CUSUM, MA |
| Moving range (median) | `mr_median` | Robust alternative |
| Mean successive squared diff | `mssd` | Short-run SPC |
| Total SD | `total` | Attributes, rare events |

```{r}
df <- data.frame(measurement = rnorm(30, 100, 1))
task_mrb <- run_spc_imr(df, "measurement", sigma_method = "mr_bar")
task_mssd <- run_spc_imr(df, "measurement", sigma_method = "mssd")
cat("MR-bar sigma:", round(task_mrb$results$statistics$sigma, 4), "\n")
cat("MSSD sigma: ", round(task_mssd$results$statistics$sigma, 4), "\n")
```

## References

- Montgomery, D. C. (2013). *Introduction to Statistical Quality Control* (7th ed.).
- ISO 7870-2:2023. *Shewhart control charts*.
- AIAG-VDA SPC (2026). *Statistical Process Control*.
- Laney, D. B. (2002). Improved control charts for attributes. *Quality Engineering*.
