---
title: "Advanced SPC: Time-Weighted, Multivariate, and Rare-Event Charts"
author: "iQualityR.spc"
date: "2026-07-05"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced SPC: Time-Weighted, Multivariate, and Rare-Event Charts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE,
  error = TRUE
)
```

```{r setup, echo = FALSE}
library(iQualityR.spc)
library(ggplot2)
set.seed(2026)

# Helper: save plot via ggsave() and display via include_graphics().
# This avoids knitr's inline graphics device which can crash in R CMD check.
iqr_plot <- function(task, type = "full", name) {
  dir.create("figures", showWarnings = FALSE)
  path <- file.path("figures", paste0(name, ".png"))
  p <- task$plot(type = type)
  ggplot2::ggsave(path, p, width = 7, height = 4.5, dpi = 100, bg = "white")
  knitr::include_graphics(path)
}
```

## Overview

This vignette covers the advanced control charts in `iQualityR.spc`:

- **Time-weighted charts**: EWMA, CUSUM, Moving Average
- **Multivariate charts**: Hotelling T2, MEWMA
- **Rare-event charts**: G chart, T chart
- **Phase / stage analysis**

These charts extend the Shewhart family to scenarios where traditional
variables / attributes charts are insufficient: small sustained shifts,
correlated measurements, or rare defects.

## 1. Time-Weighted Charts

### 1.1 EWMA

The Exponentially Weighted Moving Average (EWMA) chart is sensitive to
small sustained shifts (1 to 1.5 sigma). The smoothing parameter `lambda`
controls the trade-off between sensitivity and noise:

```{r}
df <- data.frame(
  measurement = c(rnorm(20, 100, 1), rnorm(10, 102, 1))
)
task <- run_spc_ewma(df, "measurement", lambda = 0.2)
task$summary()
iqr_plot(task, "summary", "spc_advanced_ewma")
```

Smaller `lambda` (e.g. 0.1) detects smaller shifts but reacts slower.
Larger `lambda` (e.g. 0.4) is more responsive but less sensitive.

### 1.2 CUSUM

The Cumulative Sum chart accumulates deviations from target. Parameters:

- `k`: reference value (sigma units). Default 0.5 detects 0.5 sigma shifts.
- `h`: decision interval (sigma units). Default 4.77 gives ARL0 ~ 370.

```{r}
df <- data.frame(
  measurement = c(rnorm(20, 50, 1), rnorm(10, 51.5, 1))
)
task <- run_spc_cusum(df, "measurement", k = 0.5, h = 4.77)
task$summary()
```

### 1.3 Moving Average

The MA chart averages the most recent `ma_window` observations:

```{r}
df <- data.frame(measurement = rnorm(30, 100, 1))
task <- run_spc_ma(df, "measurement", ma_window = 3)
```

## 2. Multivariate Charts

### 2.1 Hotelling T2

Use T2 when monitoring multiple correlated variables simultaneously:

```{r}
df <- data.frame(
  x1 = c(rnorm(25, 50, 1), rnorm(5, 53, 1)),
  x2 = c(rnorm(25, 30, 0.8), rnorm(5, 32.5, 0.8))
)
task <- run_spc_t2(df)
task$summary()
iqr_plot(task, "summary", "spc_advanced_t2")
```

### 2.2 MEWMA

The Multivariate EWMA (Lowry et al. 1992) is the multivariate extension
of EWMA and is sensitive to small sustained shifts in the mean vector:

```{r}
df <- data.frame(
  x1 = rnorm(30, 50, 1),
  x2 = rnorm(30, 30, 0.8)
)
task <- run_spc_mewma(df, lambda = 0.2)
```

## 3. Rare-Event Charts

### 3.1 G Chart

For monitoring the number of opportunities between rare defects
(geometric distribution):

```{r}
df <- data.frame(opportunities = rgeom(30, prob = 0.05) + 1)
task <- run_spc_g(df, "opportunities")
task$summary()
```

### 3.2 T Chart

For monitoring time between rare events (exponential distribution):

```{r}
df <- data.frame(time_between = rexp(30, rate = 0.1))
task <- run_spc_t(df, "time_between")
task$summary()
```

## 4. Phase / Stage Analysis

Phase analysis splits the data at specified boundaries and computes
control limits separately for each phase. This is useful for assessing
whether a process improvement has shifted the mean or reduced variation:

```{r}
df <- data.frame(
  measurement = c(rnorm(20, 100, 1), rnorm(15, 102, 1))
)
phases <- run_spc_phase(df, "measurement", "imr",
                        phase_boundaries = c(21))
print(phases)
```

The output shows center, sigma, UCL/LCL, and violations for each phase.
Compare phases to assess whether the improvement was effective.

## 5. Choosing a Chart

| Scenario | Recommended Chart |
|----------|--------------------|
| Continuous data, subgroups of 2-10 | Xbar-R |
| Continuous data, subgroups > 10 | Xbar-S |
| Continuous data, individual values | I-MR |
| Proportion defective, equal n | NP |
| Proportion defective, unequal n | P |
| Defects per unit, equal n | C |
| Defects per unit, unequal n | U |
| Large sample P chart over-dispersion | Laney P' |
| Large sample U chart over-dispersion | Laney U' |
| Small sustained mean shifts | EWMA or CUSUM |
| Correlated multivariate data | T2 or MEWMA |
| Rare defects (low rate) | G chart |
| Rare events (time-based) | T chart |

## References

- Lowry, C. A., Woodall, W. H., Champ, C. W., & Rigdon, S. E. (1992). A
  multivariate exponentially weighted moving average control chart.
  *Technometrics*, 34(1), 46-53.
- Lucas, J. M. (1982). Combined Shewhart-CUSUM quality control schemes.
  *Journal of Quality Technology*, 14(2), 51-61.
- Montgomery, D. C. (2013). *Introduction to Statistical Quality Control*
  (7th ed.).
- NIST/SEMATECH e-Handbook of Statistical Methods, Section 6.3.
