---
title: "Statistical Enhancement Charts: Adaptive, ARIMA, AEWMA, Change-Point, KDE, T2+MEWMA"
author: "iQualityR.spc"
date: "2026-07-05"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Statistical Enhancement Charts: Adaptive, ARIMA, AEWMA, Change-Point, KDE, T2+MEWMA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE,
  error = TRUE
)
```

```{r setup, echo = FALSE}
library(iQualityR.spc)
library(ggplot2)
set.seed(2026)

# Helper: save plot via ggsave() and display via include_graphics().
# This avoids knitr's inline graphics device which can crash in R CMD check.
iqr_plot <- function(task, type = "full", name) {
  dir.create("figures", showWarnings = FALSE)
  path <- file.path("figures", paste0(name, ".png"))
  p <- task$plot(type = type)
  ggplot2::ggsave(path, p, width = 7, height = 4.5, dpi = 100, bg = "white")
  knitr::include_graphics(path)
}
```

## Overview

This vignette covers the **statistical enhancement charts** introduced in
`iQualityR.spc` v0.2. These charts extend the Shewhart, time-weighted, and
multivariate families to handle real-world complications that classical
charts struggle with:

- **Adaptive rolling-window chart** -- processes with naturally drifting mean
- **ARIMA residual chart** -- autocorrelated processes
- **Adaptive EWMA (AEWMA)** -- small sustained shifts with adaptive smoothing
- **Change-point detection** -- structural breaks in mean or variance
- **KDE nonparametric chart** -- non-normal underlying distributions
- **T2 + MEWMA hybrid** -- multivariate processes needing both shift types

All functions return an `IqrSpcTask` object with the standard `summary()`,
`plot()`, and `results` accessors, so they integrate seamlessly with the
rest of the iQualityR ecosystem.

## 1. Adaptive Rolling-Window Chart

The adaptive chart recomputes the center line and sigma over a rolling
window of the most recent `window_size` observations. It is useful for
processes with naturally drifting mean where fixed limits would produce
excessive false alarms.

For each point `i`, the center is the mean of the previous `window_size`
observations (including point `i`), and sigma is estimated via
`iQualityR.stat::sigma_estimate` on the same window.

```{r}
df <- data.frame(
  measurement = c(rnorm(40, 100, 1), rnorm(20, 102, 1))
)
task <- run_spc_adaptive(df, "measurement", window_size = 15)
task$summary()
iqr_plot(task, "summary", "spc_enhancement_adaptive")
```

**Tuning guidance**:

- `window_size` >= 5. Smaller windows react faster but are noisier.
- Larger windows approximate a standard I-MR chart.
- For drift-plus-shift processes, pair with `run_spc_changepoint` (Section 4)
  to identify regime boundaries.

## 2. ARIMA Residual Chart

When measurements are autocorrelated (e.g., chemical processes, sensor
readings), the independence assumption of standard Shewhart charts is
violated, leading to misleading control limits. The ARIMA residual chart
fits an ARIMA(p,d,q) model and applies an I-MR chart to the residuals.

```{r}
# Simulate an AR(1) process with mean 100
ar1 <- stats::arima.sim(list(ar = 0.6), n = 60, sd = 1) + 100
df <- data.frame(measurement = as.numeric(ar1))
task <- run_spc_arima_resid(df, "measurement", arima_order = c(1, 0, 0))
task$summary()
```

The fitted ARIMA model coefficients, sigma2, and AIC are stored in
`task$results$diagnostics` for traceability:

```{r}
task$results$diagnostics$arima_coef
task$results$diagnostics$arima_sigma2
task$results$diagnostics$arima_aic
```

**Choosing `arima_order`**:

- `c(1, 0, 0)` -- AR(1), suitable for most slowly-drifting sensor data.
- `c(0, 1, 1)` -- IMA(1,1), equivalent to exponentially smoothed residuals.
- `c(1, 1, 1)` -- ARIMA(1,1,1), for trending plus autocorrelated data.

Always inspect the residual ACF/PACF before reporting. If residuals still
show significant autocorrelation, the order is too low.

## 3. Adaptive EWMA (AEWMA)

Standard EWMA uses a fixed smoothing parameter `lambda`, which forces a
trade-off between sensitivity (small lambda) and responsiveness (large
lambda). The AEWMA chart varies `lambda` over time based on the magnitude
of the recent forecast error:

\deqn{\lambda_i = \lambda_0 + (1 - \lambda_0) \cdot \min(1, k \cdot |e_i| / \sigma)}{lambda_i = lambda_0 + (1 - lambda_0) * min(1, k * |e_i| / sigma)}

where `e_i = x_i - z_{i-1}` is the one-step forecast error. Large deviations
increase `lambda`, making the chart responsive to sudden shifts while
retaining the smoothing benefit of standard EWMA in steady state.

```{r}
df <- data.frame(
  measurement = c(rnorm(30, 100, 1), rnorm(15, 103, 1))
)
task <- run_spc_aewma(df, "measurement", aewma_lambda = 0.2, aewma_k = 1.5)
task$summary()
iqr_plot(task, "summary", "spc_enhancement_aewma")
```

**Parameter guidance**:

- `aewma_lambda` in (0, 1]. Base smoothing; 0.2 is a common starting point.
- `aewma_k` > 0. Larger values trigger faster adaptation. Typical range 1-2.
- Use AEWMA instead of standard EWMA when the process is subject to
  occasional large shocks that should be reflected quickly.

## 4. Change-Point Detection

Change-point analysis identifies structural breaks in the process mean or
variance. Unlike control charts (which monitor for ongoing stability),
change-point detection is retrospective: it segments the historical data
into regimes with different parameters.

```{r}
df <- data.frame(
  measurement = c(rnorm(30, 100, 1), rnorm(30, 103, 1))
)
task <- run_spc_changepoint(df, "measurement")
task$summary()
```

Detected change-points are reported with the segment means before and after:

```{r}
task$results$data_tables$change_points
```

**Backend selection**:

- If the `changepoint` package is installed, the PELT algorithm is used via
  `requireNamespace("changepoint", quietly = TRUE)`.
- Otherwise, an in-house binary-segmentation routine based on cumulative
  sums is used as a fallback.
- `cp_method` accepts `"PELT"`, `"AMOC"`, `"SegNeigh"`, or `"BinSeg"`
  (case-insensitive). `cp_penalty` controls sensitivity; `"MBIC"` is the
  default and is robust for most applications.

## 5. KDE Nonparametric Chart

When the underlying process distribution is non-normal (e.g., skewed,
bounded, or multimodal), the standard 3-sigma limits of Shewhart charts are
inappropriate. The KDE chart derives control limits from the empirical
distribution via Kernel Density Estimation:

```{r}
# Non-normal data (exponential)
df <- data.frame(measurement = rexp(50, rate = 0.1))
task <- run_spc_kde(df, "measurement")
task$summary()
iqr_plot(task, "summary", "spc_enhancement_kde")
```

The control limits are the empirical 0.135th and 99.865th percentiles
(equivalent to +-3 sigma under normality), computed from the KDE via
quantile integration.

**Bandwidth selection**:

- `kde_bandwidth = NULL` (default) uses the Sheather-Jones selector
  (`bw.SJ`), which is robust for most unimodal distributions.
- For multimodal data, a smaller bandwidth may be needed to resolve modes.
- The bandwidth actually used is stored in `task$results$diagnostics`.

```{r}
task$results$diagnostics$bandwidth
```

## 6. T2 + MEWMA Hybrid Multivariate Chart

When monitoring multiple correlated variables, the Hotelling T2 chart
detects large mean shifts while MEWMA detects small sustained shifts. The
hybrid chart runs both statistics on the same data and flags a violation
when either exceeds its upper control limit:

```{r}
df <- data.frame(
  x1 = c(rnorm(25, 50, 1), rnorm(10, 52, 1)),
  x2 = c(rnorm(25, 30, 0.8), rnorm(10, 31, 0.8))
)
task <- run_spc_t2_mewma(df, lambda = 0.2)
task$summary()
```

The combined statistic displayed on the chart is the maximum of
`T2 / UCL_T2` and `MEWMA / UCL_MEWMA`; a value > 1 indicates a violation.
Both raw statistics are stored in the `points` data table:

```{r}
head(task$results$data_tables$points)
```

**When to use which multivariate chart**:

| Scenario | Recommended Chart |
|----------|--------------------|
| Large mean shift detection | T2 alone |
| Small sustained mean shift detection | MEWMA alone |
| Both shift types, single dashboard | T2 + MEWMA hybrid |
| Diagnosing which variable shifted | Pair with `run_spc_t2` per-variable |

## 7. Choosing an Enhancement Chart

| Data characteristic | Recommended chart |
|---------------------|-------------------|
| Drifting mean | Adaptive |
| Autocorrelated residuals | ARIMA residual |
| Occasional large shocks | AEWMA |
| Historical regime changes | Change-point |
| Non-normal distribution | KDE |
| Multiple correlated vars, mixed shifts | T2 + MEWMA |

## 8. Combining Enhancement Charts

In practice, multiple enhancement charts are often used together. A
typical workflow for an unfamiliar process:

1. Run `run_spc_changepoint` to identify historical regime boundaries.
2. Within each regime, run `run_spc_kde` to assess distributional shape.
3. If non-normal, use KDE chart; if normal but autocorrelated, use ARIMA
   residual; otherwise standard Shewhart or EWMA.
4. For real-time monitoring of slow drifts with occasional jumps, prefer
   AEWMA over standard EWMA.
5. For multivariate processes, always pair T2 with MEWMA (use the hybrid)
   unless one shift type is known to be irrelevant.

## References

- Capizzi, G., & Masarotto, G. (2003). An adaptive exponentially weighted
  moving average control chart. *Technometrics*, 45(3), 199-207.
- Hotelling, H. (1947). Multivariate quality control. *Techniques of
  Statistical Analysis*, 111-184.
- Killick, R., & Eckley, I. A. (2014). changepoint: An R package for
  changepoint analysis. *Journal of Statistical Software*, 58(3), 1-19.
- Lowry, C. A., Woodall, W. H., Champ, C. W., & Rigdon, S. E. (1992). A
  multivariate exponentially weighted moving average control chart.
  *Technometrics*, 34(1), 46-53.
- Montgomery, D. C. (2013). *Introduction to Statistical Quality Control*
  (7th ed.).
- Sheather, S. J., & Jones, M. C. (1991). A reliable data-based bandwidth
  selection method for kernel density estimation. *Journal of the Royal
  Statistical Society, Series B*, 53(3), 683-690.
