---
title: "描述性统计分析报告"
author: "iQualityR"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

# 分析概览

```{r overview}
cat(sprintf("置信水平: %.0f%%\n", params$conf_level * 100))
cat(sprintf("分析变量数: %d\n", length(params$target_vars)))
if (!is.null(params$group_var)) {
  cat(sprintf("分组变量: %s\n", params$group_var))
}
```

# 统计摘要

## 描述统计表

| 变量 | N | 均值 | 标准差 | 中位数 | 变异系数(%) | 偏度 | 峰度 | P值 |
|------|---|------|--------|--------|-------------|------|------|-----|
`r for (var in names(results)) { s <- results[[var]]; cat(sprintf("| %s | %d | %.4f | %.4f | %.4f | %.2f | %.3f | %.3f | %s |", var, s$n, s$mean, s$stdev, s$median, s$cv, s$skew, s$kurt, if (s$p_value < 0.001) "< 0.001" else sprintf("%.3f", s$p_value))) }`

## 分位数详情

| 变量 | 最小值 | Q1 | 均值 | Q3 | 最大值 | 范围 | IQR |
|------|--------|-----|------|-----|--------|------|-----|
`r for (var in names(results)) { s <- results[[var]]; cat(sprintf("| %s | %.4f | %.4f | %.4f | %.4f | %.4f | %.4f | %.4f |", var, s$min, s$q1, s$mean, s$q3, s$max, s$range, s$iqr)) }`

# 正态性检验

Anderson-Darling正态性检验结果：

| 变量 | AD统计量 | P值 | 结论 |
|------|----------|-----|------|
`r for (var in names(results)) { s <- results[[var]]; conclusion <- if (s$p_value < 0.05) "拒绝正态性" else "不能拒绝正态性"; cat(sprintf("| %s | %.3f | %s | %s |", var, s$ad_stat, if (s$p_value < 0.001) "< 0.001" else sprintf("%.3f", s$p_value), conclusion)) }`

# 图表

## 直方图

```{r hist, fig.width=10, fig.height=6}
for (var in names(results)) {
  s <- results[[var]]
  df <- data.frame(v = s$raw)
  lims <- c(min(df$v) - diff(range(df$v)) * 0.15, max(df$v) + diff(range(df$v)) * 0.15)

  p <- ggplot(df, aes(x = v)) +
    geom_histogram(aes(y = after_stat(density)), fill = "#A9C4E3", color = "white", bins = 20) +
    stat_function(fun = dnorm, args = list(mean = s$mean, sd = s$stdev), color = "red", linewidth = 1) +
    scale_x_continuous(limits = lims) +
    labs(title = sprintf("直方图: %s", var), x = NULL, y = "密度") +
    theme_minimal()

  print(p)
}
```

## 箱线图

```{r boxplot, fig.width=10, fig.height=6}
for (var in names(results)) {
  s <- results[[var]]
  df <- data.frame(v = s$raw)
  lims <- c(min(df$v) - diff(range(df$v)) * 0.15, max(df$v) + diff(range(df$v)) * 0.15)

  p <- ggplot(df, aes(x = v, y = "")) +
    geom_boxplot(fill = "#F2F2F2", outlier.color = "red", width = 0.6) +
    scale_x_continuous(limits = lims) +
    labs(title = sprintf("箱线图: %s", var), x = "值", y = NULL) +
    theme_minimal()

  print(p)
}
```

---

*报告由 iQualityR.stat 生成*
