---
title: "Non-Parametric Capability & Multi-Variable Considerations"
subtitle: "Distribution-free capability and handling multiple CTQs"
author: "iQualityR capa Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Non-Parametric Capability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7, fig.height = 4.5, fig.align = "center",
  warning = FALSE, message = FALSE
)

options(iqr.default_theme = "academic")
library(iQualityR.capa)
library(ggplot2)

fmt4 <- function(x) ifelse(is.na(x), "NA", formatC(as.numeric(x), format = "f", digits = 4))
fmt2 <- function(x) ifelse(is.na(x), "NA", formatC(as.numeric(x), format = "f", digits = 2))
```

## 1. When to Use This Path

Use **non-parametric capability** when:

1. No parametric distribution fits the data (KS p < 0.05 for every candidate
   in `capability_nonnormal()`).
2. The sample is small and fitting a parametric family would be unstable.
3. The data is multi-modal, mixture-distributed, or has a shape no standard
   distribution captures.
4. You need the *observed* defect rate rather than a model-based prediction.

The non-parametric path makes **no distributional assumption**: the
empirical (or kernel-smoothed) CDF is the model. The trade-off is that you
lose the within/overall σ distinction and the Cpm/Cpmk indices (which require
a parametric mean / target relationship), and the PPM is the **observed**
exceedance rate, which is only meaningful when the sample is large enough to
contain out-of-spec observations.

## 2. Method and Formulas

### 2.1 Capability indices (empirical quantile method)

The non-parametric path estimates the distribution quantiles directly from
the data via `stats::quantile()` (type 7, the default) or via a
kernel-smoothed CDF, then applies the same quantile-method formulas as the
non-normal path:

$$
\sigma_{eq} = \frac{Q_{0.99865} - Q_{0.00135}}{6}
$$

$$
P_p = \frac{USL - LSL}{6\sigma_{eq}}, \quad
P_{pk} = \min\left(\frac{USL - Q_{0.5}}{3\sigma_{eq}}, \frac{Q_{0.5} - LSL}{3\sigma_{eq}}\right)
$$

where $Q_{0.00135}$, $Q_{0.5}$, $Q_{0.99865}$ are the empirical quantiles.

### 2.2 PPM (observed)

$$
PPM_{observed} = \frac{\#\{x < LSL\} + \#\{x > USL\}}{n} \times 10^6
$$

This is the **actual observed** defect rate, not a model-based prediction.
It is therefore a *lower bound* on the true PPM (you only see what happened
in the sample). For small samples with no out-of-spec observations, the
observed PPM will be 0 — which means "no observed defects", not "process is
defect-free".

### 2.3 Two methods

| Method | How Q_p is estimated | Best for |
|---|---|---|
| `"empirical"` | `stats::quantile(x, p)` (plain order statistics) | Discrete data, small samples, or when you want no smoothing. |
| `"kernel"` | Kernel-smoothed CDF (`stats::density` + integrate) | Continuous data; smoother quantiles at extreme tail probabilities. |

The default is `"kernel"` because the 0.135% / 99.865% quantiles are extreme
tail probabilities where plain order statistics are unstable; kernel smoothing
borrows strength from neighbouring observations.

## 3. Data Preparation

We reuse the bundled right-skewed dataset. The same data is used in
`vignette("capa-nonnormal")` where a Weibull fits it well — here we run the
non-parametric path to show how the two paths compare on the same data.

```{r load-data}
df <- read.csv(
  system.file("extdata", "capability_nonnormal_weibull.csv",
              package = "iQualityR.capa"),
  fileEncoding = "UTF-8"
)
str(df)
```

## 4. Non-Parametric Capability — Quick Start

```{r np-quick}
task <- capability_nonparametric(
  data = df, measurement = "measurement",
  lsl = 5, usl = 80,
  nonparametric_method = "kernel"
)
task$summary()
```

## 5. Inspecting the Results

### 5.1 Capability indices

```{r np-indices}
stats <- task$results$statistics
data.frame(
  Index = c("Cp (= Pp)", "Cpk (= Ppk)", "Cpl", "Cpu", "Mean", "n"),
  Value = c(stats$cp, stats$cpk, stats$cpl, stats$cpu, stats$mean, stats$n)
)
```

### 5.2 PPM (observed)

```{r np-ppm}
task$results$data_tables$ppm_overall
```

The PPM here is the **observed** exceedance rate. Compare with the model-based
PPM from the non-normal path on the same data (see
`vignette("capa-nonnormal")`). If the two agree, the fitted distribution is a
good model; if they differ widely, the parametric model is mis-specified.

### 5.3 Diagnostics

```{r np-diag}
diag <- task$results$diagnostics
data.frame(
  Metric = c("Method", "Sample size", "Verdict"),
  Value = c(diag$method, diag$sample_size,
            diag$capability_judgment$overall_verdict)
)
```

## 6. Visualization

### 6.1 The non-parametric Sixpack

```{r np-plot-full, fig.width=10, fig.height=8}
print(task$plot(type = "full"))
```

The non-parametric Sixpack differs from the normal Sixpack in the Q-Q panel:
instead of a normal Q-Q or a fitted-distribution Q-Q, it shows an **empirical
Q-Q (rankit plot)** — sample order statistics against expected normal order
statistics. This verifies whether the empirical shape deviates from normality
without committing to any parametric family. A straight line means the data
is already normal (and you should have used the normal path); curvature
confirms non-normality, and the non-parametric indices are the honest report.

### 6.2 Empirical Q-Q plot

```{r np-plot-qq}
print(task$plot(type = "qq"))
```

### 6.3 Histogram with density

```{r np-plot-basic}
print(task$plot(type = "basic"))
```

## 7. Empirical vs Kernel Method

```{r np-method-compare}
task_emp <- capability_nonparametric(
  data = df, measurement = "measurement",
  lsl = 5, usl = 80,
  nonparametric_method = "empirical"
)

data.frame(
  Method = c("kernel", "empirical"),
  Cp  = c(task$results$statistics$cp,  task_emp$results$statistics$cp),
  Cpk = c(task$results$statistics$cpk, task_emp$results$statistics$cpk),
  PPM_total = c(task$results$data_tables$ppm_overall$PPM[3],
                task_emp$results$data_tables$ppm_overall$PPM[3])
)
```

For continuous data the two methods usually agree closely. Use `"empirical"`
when you want the pure order-statistic answer (no smoothing artefacts at
all); use `"kernel"` when the data is continuous and you want smoother
extreme-tail quantiles.

## 8. Bootstrap Confidence Intervals

Non-parametric capability benefits particularly from the bootstrap, because
the empirical quantiles at 0.135% / 99.865% are noisy in finite samples.

```{r np-bootstrap}
task_boot <- capability_nonparametric(
  data = df, measurement = "measurement",
  lsl = 5, usl = 80,
  nonparametric_method = "kernel",
  use_bootstrap = TRUE, bootstrap_samples = 200
)
task_boot$results$data_tables$bootstrap_ci
```

## 9. Multi-Variable Capability Considerations

Real products have **multiple CTQs** (critical-to-quality characteristics):
a shaft has diameter, surface roughness, and length; a circuit board has
thickness, impedance, and solder volume. The question "is this product
capable?" requires assessing all CTQs together, not just one at a time.

### 9.1 The practical approach: per-CTQ capability + a summary view

The standard industrial workflow (and the JMP Process Capability platform's
default) is:

1. Run a univariate capability analysis on each CTQ.
2. Collect each CTQ's `Ppk` (the long-term, centering-aware index).
3. Visualise all CTQs on a single **Ppk bar chart** or **Goal Plot** to see
   which CTQs are capable and which are not.

This is a *multi-variable* (multiple univariate) analysis, not a
*multivariate* (joint distribution) analysis — it treats CTQs as independent.
When CTQs are correlated (they usually are), this is conservative but
defensible for routine reporting.

#### Worked example: two CTQs

We analyse two CTQs from the bundled data: the normal-shaft diameter and the
non-normal surface roughness. Each gets its own capability path.

```{r multi-var-setup}
df_n <- read.csv(
  system.file("extdata", "capability_normal_subgroup.csv", package = "iQualityR.capa"),
  fileEncoding = "UTF-8"
)
df_nn <- read.csv(
  system.file("extdata", "capability_nonnormal_weibull.csv", package = "iQualityR.capa"),
  fileEncoding = "UTF-8"
)

# CTQ 1: shaft diameter (normal, subgrouped)
task_diameter <- capability_normal(
  data = df_n, measurement = "measurement",
  lsl = 44, usl = 56, target = 50, subgroup = "subgroup"
)

# CTQ 2: surface roughness (non-normal)
task_roughness <- capability_nonnormal(
  data = df_nn, measurement = "measurement",
  lsl = 5, usl = 80, distribution = "auto"
)
```

#### The multi-CTQ summary table

```{r multi-var-summary}
multi_summary <- data.frame(
  CTQ = c("Diameter (mm)", "Surface roughness (µm)"),
  Path = c("Normal", paste0("Non-normal (",
           task_roughness$results$diagnostics$distribution, ")")),
  Cp  = c(task_diameter$results$statistics$cp,
          task_roughness$results$statistics$cp),
  Cpk = c(task_diameter$results$statistics$cpk,
          task_roughness$results$statistics$cpk),
  Ppk = c(task_diameter$results$statistics$ppk,
          task_roughness$results$statistics$ppk),
  PPM_overall = c(task_diameter$results$data_tables$ppm_overall$PPM[3],
                  task_roughness$results$data_tables$ppm_overall$PPM[3]),
  Verdict = c(task_diameter$results$diagnostics$capability_judgment$overall_verdict,
              task_roughness$results$diagnostics$capability_judgment$overall_verdict)
)
multi_summary
```

#### The multi-CTQ Ppk bar chart (poor-man's Goal Plot)

```{r multi-var-goal-plot, fig.width=7, fig.height=4}
goal_df <- data.frame(
  CTQ = c("Diameter", "Surface roughness"),
  Ppk = c(task_diameter$results$statistics$ppk,
          task_roughness$results$statistics$ppk),
  Cp  = c(task_diameter$results$statistics$cp,
          task_roughness$results$statistics$cp)
)
goal_df$Status <- ifelse(goal_df$Ppk >= 1.33, "pass",
                         ifelse(goal_df$Ppk >= 1.00, "watch", "fail"))

ggplot(goal_df, aes(x = reorder(CTQ, Ppk), y = Ppk, fill = Status)) +
  geom_col(show.legend = FALSE) +
  geom_hline(yintercept = c(1.00, 1.33), linetype = "dashed", color = "grey50") +
  geom_text(aes(label = sprintf("%.2f", Ppk)), vjust = -0.5) +
  scale_fill_manual(values = c(pass = "#2E7D32", watch = "#F9A825", fail = "#C62828")) +
  labs(title = "Multi-CTQ Capability Summary",
       subtitle = "Dashed lines: 1.00 (marginal) and 1.33 (acceptable)",
       x = NULL, y = "Ppk (long-term)") +
  coord_flip() +
  theme_minimal()
```

This bar chart is the simplest multi-variable summary: the product is only as
capable as its weakest CTQ. If any CTQ is red, the product is not capable —
even if the others are green.

### 9.2 The JMP Goal Plot

JMP's signature multi-variable view is the **Goal Plot**: a scatter of
process spread (vertical, 3σ as a fraction of half-tolerance) vs. mean
shift (horizontal, |μ − T| as a fraction of half-tolerance). CTQs in the
lower-left "goal box" are well-centred and capable; CTQs in the top-right
are off-target and/or too variable. This separates "off-target" problems
from "too much variation" problems at a glance — information the per-CTQ
Ppk bar above cannot convey.

`iQualityR.plot` now ships `plot_goal_plot()` (plus the faceted
`plot_capability_snapshot()` for per-CTQ mini-histograms). Both are
demonstrated in `vignette("capa-advanced")` §4.10.

### 9.3 When CTQs are correlated

The per-CTQ approach treats CTQs as independent. When CTQs are correlated
(e.g. diameter and length on the same shaft), the joint defect probability is
*not* the product of the marginal probabilities. Two ways to handle this:

1. **Conservative (default)** — report the worst marginal Ppk. This
   over-estimates the joint defect rate but is safe.
2. **Joint modelling (roadmap)** — fit a multivariate normal (or copula) to
   the CTQ vector and compute the joint probability of falling outside the
   spec hyper-rectangle. This is the multivariate capability problem.

## 10. Multivariate Capability Indices (see `vignette("capa-advanced")`)

True multivariate capability indices quantify how well a *vector* of CTQs
fits inside a *region* of acceptance. The three families in the literature
are:

| Index | Author | Idea |
|---|---|---|
| **MCPV** (Multivariate Capability Process Vector) | Taam et al. (1993) | Ratio of the volume of the modified spec region (shrunk for centering) to the volume of the 99.73% probability ellipsoid of the process. |
| **HPCI** (Hotelling $T^2$ Process Capability Index) | Shahriari et al. (1995) | A three-component vector: (1) ratio of spec volume to process volume, (2) Hotelling $T^2$ test for centering, (3) tail-area test for shape. |
| **MCpm** | Taam / Pearn-Kotz | Multivariate extension of Cpm using a target vector. |

### Status in `iQualityR.capa`

MCPV (Taam) and HPCI (Shahriari) **are now implemented** — see
`vignette("capa-advanced")` for the full worked example with engine
cylinder bore & stroke data. The implementation uses `mvtnorm::pmvnorm`
for the joint probability content of the spec hyper-rectangle and a
regularized covariance inverse (SVD trick from `iQualityR.spc`) for robust
Mahalanobis distances.

This fills a gap that **even Minitab v21 and JMP v19 do not fully cover** —
neither ships dedicated MCPV / HPCI indices as capability outputs.

For the simpler multi-variable (multiple univariate) view used by JMP's Goal
Plot, see §9 above — it treats CTQs as independent and is the routine
fallback when you only need a multi-CTQ summary, not a joint verdict.

## 11. Common Pitfalls

1. **Reporting observed PPM = 0 as "zero defects"** — it means "no defects
   *observed in this sample*", not "process is defect-free". Use the
   bootstrap CI to bound the true PPM.
2. **Small samples** — non-parametric quantiles at 0.135% / 99.865% need
   n ≥ 100 to be stable; below that, prefer the bootstrap or report only
   the observed PPM.
3. **Using non-parametric when a parametric fit is available** — the
   parametric path is more efficient. Reserve non-parametric for when no
   distribution fits.
4. **Treating CTQs as independent when they are not** — see §9.3. The
   conservative worst-CTQ rule is safe but may over-state defect rates.
5. **Confusing multi-variable with multivariate** — multiple univariate
   analyses (§9) is *multi-variable*; a joint distribution analysis (§10) is
   *multivariate*. The former is routine; the latter is research-grade.

## 12. References

1. Conover, W. J. (1999). *Practical Nonparametric Statistics* (3rd ed.).
   Wiley. — Empirical quantiles, kernel CDF.
2. Silverman, B. W. (1986). *Density Estimation for Statistics and Data
   Analysis*. Chapman & Hall. — Kernel smoothing.
3. Taam, W., Subbaiah, P., & Liddy, J. W. (1993). *A note on multivariate
   capability indices*. Journal of Applied Statistics, 20(3), 339–351. — MCPV.
4. Shahriari, H., Hubele, N. F., & Lawrence, F. P. (1995). *A multivariate
   process capability vector*. Proc. IIE Annual Conference, 304–309. — HPCI.
5. Pearn, W. L., & Kotz, S. (2006). *Encyclopedia and Handbook of Process
   Capability Indices*. World Scientific. — Multivariate indices survey.
6. JMP. *Process Capability platform: Goal Plot*, v19.
7. `iQualityR.stat::capability_to_ppm` — PPM computation backend.
