---
title: "Non-Normal Capability"
subtitle: "Parametric capability analysis for skewed and bounded data"
author: "iQualityR capa Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Non-Normal Capability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7, fig.height = 4.5, fig.align = "center",
  warning = FALSE, message = FALSE
)

options(iqr.default_theme = "academic")
library(iQualityR.capa)

fmt4 <- function(x) ifelse(is.na(x), "NA", formatC(as.numeric(x), format = "f", digits = 4))
fmt2 <- function(x) ifelse(is.na(x), "NA", formatC(as.numeric(x), format = "f", digits = 2))
```

## 1. When to Use This Path

Use **non-normal capability** when the data is clearly non-normal *and* a
parametric non-normal distribution fits it well. Typical cases:

- **Cycle time, service time, repair time** — right-skewed; Weibull, Gamma,
  or Lognormal.
- **Surface roughness, particle counts** — right-skewed with a natural zero
  bound; Weibull or Exponential.
- **Tensile strength, breaking load** — left-skewed; smallest extreme value.
- **Bounded proportions** — data in (0, 1); Beta.

The non-normal path is *not* appropriate when:

- The data *is* normal — use `capability_normal()` (see
  `vignette("capa-normal")`). The normal path gives tighter, more efficient
  estimates when its assumption holds.
- No parametric distribution fits (KS p < 0.05 for all candidates) — use
  `capability_nonparametric()` (see `vignette("capa-nonparametric")`).
- You can normalise the data with Box-Cox/Johnson and report on the
  transformed scale — see §6 below for the transformation path, which is a
  legitimate alternative to the quantile method.

This vignette covers both the **quantile method** (the Minitab non-normal
path) and the **transformation path** (Box-Cox / Johnson), because the two
compete for the same use case and the analyst must choose between them.

## 2. Method and Formulas

### 2.1 The quantile method (Clements / Minitab convention)

When the data follows a non-normal distribution with CDF $F$ and quantile
function $Q = F^{-1}$, capability is computed from the distribution
quantiles rather than the mean/SD:

$$
\sigma_{eq} = \frac{Q_{0.99865} - Q_{0.00135}}{6}
$$

where $Q_{0.00135}$ and $Q_{0.99865}$ are the 0.135% and 99.865% quantiles of
the fitted distribution (the same tail probabilities a ±3σ normal would
cover). The indices are then:

$$
P_p = \frac{USL - LSL}{6\sigma_{eq}}, \quad
P_{pk} = \min\left(\frac{USL - Q_{0.5}}{3\sigma_{eq}}, \frac{Q_{0.5} - LSL}{3\sigma_{eq}}\right)
$$

For non-normal data there is **no within/overall distinction** (there are no
subgroups in this path), so `Cp = Pp` and `Cpk = Ppk`.

### 2.2 PPM (non-normal)

PPM is computed from the fitted distribution's tail areas:

$$
PPM_{below} = F(LSL) \cdot 10^6, \quad
PPM_{above} = (1 - F(USL)) \cdot 10^6
$$

This is delegated to `iQualityR.stat::capability_to_ppm` with the fitted
distribution function, so the arithmetic is identical across the ecosystem.

### 2.3 Distribution selection

The auto-fitter (`distribution = "auto"`) compares candidate distributions by
**AIC** and reports the winner. The candidate set is: **Weibull, Lognormal,
Gamma, Exponential, Logistic**. Goodness-of-fit is verified with the
**Kolmogorov-Smirnov** test (delegated to `iQualityR.stat::fit_distribution`).

If you know the distribution from domain knowledge (e.g. Weibull for
strength), pass it explicitly with `distribution = "weibull"`.

### 2.4 The transformation path (alternative)

Instead of the quantile method, you can transform the data to normality with
Box-Cox or Johnson, run the normal analysis on the transformed scale, and
report the indices on that scale. The spec limits are transformed with the
**same** fitted parameters so the indices remain comparable.

| Aspect | Quantile method | Transformation path |
|---|---|---|
| Output scale | Original (mm, s, MPa) | Transformed (dimensionless) |
| Within/overall split | No | Yes (if subgrouped) |
| Best when | A specific non-normal family fits well | Data is positive and monotone-normalisable |
| Downside | No within σ; no Cpm/Cpmk | Indices on unfamiliar scale; harder to communicate |

`iQualityR.capa` supports both. Use the quantile method when you need to
report on the original scale; use the transformation path when you need the
within/overall distinction or Cpm.

## 3. Data Preparation

The bundled `capability_nonnormal_weibull.csv` is a right-skewed
surface-roughness sample (Ra, µm) with a natural lower bound at 0. The
engineering specification is **LSL = 5 µm, USL = 80 µm** (no target).

```{r load-data}
df <- read.csv(
  system.file("extdata", "capability_nonnormal_weibull.csv",
              package = "iQualityR.capa"),
  fileEncoding = "UTF-8"
)
str(df)
summary(df$measurement)
```

The summary already hints at skewness: the mean exceeds the median, and the
range is asymmetric.

## 4. Non-Normal Capability — Quick Start

```{r nonnormal-quick}
task <- capability_nonnormal(
  data = df, measurement = "measurement",
  lsl = 5, usl = 80,
  distribution = "auto"
)
task$summary()
```

## 5. Distribution Diagnostics

### 5.1 Selected distribution and goodness-of-fit

```{r nonnormal-diag}
diag <- task$results$diagnostics
data.frame(
  Metric = c("Selected distribution", "AIC", "BIC",
             "KS statistic", "KS p-value", "Verdict"),
  Value = c(diag$distribution,
            round(diag$aic, 2),
            round(diag$bic, 2),
            fmt4(diag$ks_statistic),
            fmt4(diag$ks_p_value),
            diag$capability_judgment$overall_verdict)
)
```

**Decision rule:** If `KS p-value ≥ 0.05`, the selected distribution is an
acceptable fit and the quantile-method indices are defensible. If
`KS p-value < 0.05`, no candidate fits — fall back to
`capability_nonparametric()`.

### 5.2 Capability indices

```{r nonnormal-indices}
stats <- task$results$statistics
data.frame(
  Index = c("Cp (= Pp)", "Cpk (= Ppk)", "Cpl", "Cpu", "Mean", "n"),
  Value = c(stats$cp, stats$cpk, stats$cpl, stats$cpu, stats$mean, stats$n)
)
```

Note: for non-normal capability there is no within/overall distinction, so
`Cp = Pp` and `Cpk = Ppk`. The single-sided `Cpl`/`Cpu` use the 0.5 and
tail quantiles of the fitted distribution.

### 5.3 PPM (observed vs expected)

```{r nonnormal-ppm}
task$results$data_tables$ppm_overall
```

The PPM here is **model-based** (computed from the fitted distribution's tail
areas), not the raw observed exceedance rate. Compare with the
non-parametric path's observed PPM to sanity-check the fit.

## 6. Visualization

### 6.1 The non-normal Sixpack

```{r nonnormal-plot-full, fig.width=10, fig.height=8}
print(task$plot(type = "full"))
```

The non-normal Sixpack differs from the normal Sixpack in two panels:

- **Histogram (middle-left)** overlays the fitted distribution density in red,
  so you can visually verify the fit.
- **Q-Q plot (middle-right)** is a *fitted-distribution* Q-Q: theoretical
  quantiles come from the fitted distribution (not standard normal). Points
  hugging the y = x line confirm the fit.

### 6.2 The fitted-distribution Q-Q plot

```{r nonnormal-plot-qq}
print(task$plot(type = "qq"))
```

The subtitle reports the selected distribution and KS p-value. If you see
systematic curvature (e.g. S-shape, bowing), the distribution is wrong even
if KS passes — try another candidate or move to the non-parametric path.

### 6.3 Histogram with fitted density overlay

```{r nonnormal-plot-basic}
print(task$plot(type = "basic"))
```

## 7. Specifying a Distribution Explicitly

If domain knowledge dictates a particular distribution, pass it explicitly.
This bypasses AIC selection and reports the fit quality for that single
distribution.

```{r nonnormal-explicit}
task_lognormal <- capability_nonnormal(
  data = df, measurement = "measurement",
  lsl = 5, usl = 80,
  distribution = "lognormal"
)
diag_ln <- task_lognormal$results$diagnostics
data.frame(
  Metric = c("Distribution", "AIC", "KS p-value"),
  Value = c(diag_ln$distribution,
            round(diag_ln$aic, 2),
            fmt4(diag_ln$ks_p_value))
)
```

Comparing the explicit-Weibull, explicit-Lognormal, and auto-selected fits
side-by-side is the recommended way to choose a defensible model — do not
blindly trust AIC alone, always inspect the Q-Q plot.

## 8. The Box-Cox / Johnson Transformation Path

When the data is positive and monotone-normalisable, the transformation path
is a legitimate alternative to the quantile method. `capability_normal()`
with a `transform` argument fits the transform, transforms data **and** spec
limits with the same parameters, runs the normal analysis on the transformed
scale, and annotates every panel.

### 8.1 Box-Cox

```{r transform-boxcox}
task_bc <- capability_normal(
  data = df, measurement = "measurement",
  lsl = 5, usl = 80,
  transform = "box_cox"
)
task_bc$summary()
```

### 8.2 Transform diagnostics

```{r transform-diag}
diag_bc <- task_bc$results$diagnostics
data.frame(
  Metric = c("Transform applied", "Lambda", "Normality before (p)",
             "Normality after (p)", "Transformed LSL", "Transformed USL"),
  Value = c(diag_bc$transform_applied,
            fmt4(diag_bc$transform_lambda),
            fmt4(diag_bc$transform_normality_before),
            fmt4(diag_bc$transform_normality_after),
            fmt4(diag_bc$transformed_lsl),
            fmt4(diag_bc$transformed_usl))
)
```

A successful transform shows `normality_after` ≫ `normality_before` (both are
p-values). If `normality_after < 0.05`, the transform did not achieve
normality — abandon this path and use the quantile method or the
non-parametric path.

### 8.3 Transformed-scale Sixpack

```{r transform-plot-full, fig.width=10, fig.height=8}
print(task_bc$plot(type = "full"))
```

Note that every panel's subtitle annotates the transform (`Box-Cox
(lambda = …)`) and the spec labels show the **transformed** spec values
(formatted to 4 significant digits). This makes it unambiguous that the
reported indices are on the transformed scale.

### 8.4 Johnson transform

The Johnson family (SU / SB / SL) handles a wider range of shapes than
Box-Cox, including bounded data. Use `transform = "johnson"`:

```{r transform-johnson, eval = FALSE}
task_jh <- capability_normal(
  data = df, measurement = "measurement",
  lsl = 5, usl = 80,
  transform = "johnson"
)
```

Use `transform = "auto"` to let the analyzer pick the best of Box-Cox /
Johnson / Yeo-Johnson / log / sqrt / reciprocal (delegated to
`iQualityR.stat::auto_transform`).

## 9. Bootstrap Confidence Intervals

For defensible uncertainty quantification on the non-normal indices, enable
the bootstrap:

```{r nonnormal-bootstrap}
task_boot <- capability_nonnormal(
  data = df, measurement = "measurement",
  lsl = 5, usl = 80,
  distribution = "auto",
  use_bootstrap = TRUE, bootstrap_samples = 200
)
task_boot$results$data_tables$bootstrap_ci
```

The bootstrap re-fits the distribution on each resample, so it captures both
the parameter-estimation uncertainty and the index-estimation uncertainty.

## 10. Quantile Method vs Transformation Path — How to Choose

| Criterion | Choose quantile method | Choose transformation path |
|---|---|---|
| Need to report on original scale | ✓ | |
| Need within/overall split | | ✓ (if subgrouped) |
| Need Cpm / Cpmk | | ✓ (requires target) |
| Data is bounded (0, 1) | Beta | Johnson SB |
| Data is right-skewed, positive | Weibull/Gamma/Lognormal | Box-Cox |
| Data is heavily skewed | | Johnson SU |
| Distribution must be defensible to auditor | KS test on original scale | Normality test on transformed scale |

A common workflow is to run both and compare: if the two give materially
different Cpk, investigate which assumption is weaker (the distribution fit
or the transform's normality).

## 11. Report Export

```{r nonnormal-report}
xlsx_path <- tempfile(fileext = ".xlsx")
task$report(format = "excel", path = xlsx_path)

data.frame(
  Format = "excel",
  Path = basename(xlsx_path),
  Exists = file.exists(xlsx_path),
  Size_KB = round(file.size(xlsx_path) / 1024, 1)
)
```

The Excel report adds a Diagnostics sheet with the selected distribution,
AIC/BIC, and KS results, so the choice of distribution is auditable.

## 12. Common Pitfalls

1. **Trusting AIC alone** — a distribution can win on AIC but still fail KS.
   Always check the KS p-value and the Q-Q plot.
2. **Forgetting the natural bound** — surface roughness, time, and counts have
   a natural zero; fitting a normal (or a Lognormal without shift) is wrong.
3. **Reporting transformed-scale indices without annotation** — always keep
   the transform note in the subtitle; auditors must know the scale.
4. **Using non-normal when normal is fine** — the normal path is more
   efficient. Only switch when normality genuinely fails.
5. **Small samples** — non-normal fitting needs n ≥ 50 to be stable; below
   that, prefer the non-parametric path.

## 13. References

1. Clements, J. A. (1989). *Process capability calculations for non-normal
   distributions*. Quality Progress, 22(9), 95–100. — Quantile method.
2. Kotz, S., & Lovelace, N. L. (1998). *Process Capability Indices in Theory
   and Practice*. Arnold. — Non-normal capability.
3. Box, G. E. P., & Cox, D. R. (1964). *An analysis of transformations*.
   JRSS-B, 26(2), 211–252. — Box-Cox.
4. Johnson, N. L. (1949). *Systems of frequency curves generated by methods
   of translation*. Biometrika, 36(1/2), 149–176. — Johnson family.
5. Minitab LLC. *Nonnormal Capability documentation*, v21.
6. `iQualityR.stat::fit_distribution` — distribution fitting backend.
7. `iQualityR.stat::box_cox_transform` / `johnson_transform` /
   `auto_transform` — transformation backends.
