---
title: "Normal & Between/Within Capability"
subtitle: "Subgroup-based process capability under the normal assumption"
author: "iQualityR capa Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Normal Capability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7, fig.height = 4.5, fig.align = "center",
  warning = FALSE, message = FALSE
)

options(iqr.default_theme = "academic")
library(iQualityR.capa)

fmt4 <- function(x) ifelse(is.na(x), "NA", formatC(as.numeric(x), format = "f", digits = 4))
fmt2 <- function(x) ifelse(is.na(x), "NA", formatC(as.numeric(x), format = "f", digits = 2))
```

## 1. When to Use This Path

Use **normal capability** when the measured characteristic is well-modelled by
a normal distribution and the process is in statistical control. This is the
default path for machined dimensions, fill weights, electrical parameters,
and most continuous CTQs (critical-to-quality characteristics).

Use **between/within capability** — the Minitab B/W path — when you have
**rational subgroups** and you suspect **between-subgroup sources of variation**
(setup changes, operator-to-operator differences, lot-to-lot drift). The B/W
path decomposes total σ into four components and uses the between-within σ for
Cp/Cpk, giving a more honest short-term capability estimate than the plain
within σ when subgroups drift.

This vignette covers both, because they share the same normal-distribution
assumption and differ only in the σ decomposition.

## 2. Method and Formulas

### 2.1 σ decomposition (the heart of the normal path)

`iQualityR.capa` delegates all σ estimation to
`iQualityR.stat::sigma_decomposition`, which produces:

| Component | Symbol | Estimator |
|---|---|---|
| Within-subgroup σ | σ_within | R̄/d₂ (subgroups of size ≥ 2) or MSSD (individuals). |
| Between-subgroup σ | σ_between | √(max(0, MSSD_between − within_var / m)) where m = subgroup size. |
| Between-within σ | σ_BW | √(σ²_within + σ²_between) |
| Overall σ | σ_overall | Sample SD of all observations (long-term). |

For the plain **normal** path, Cp/Cpk use σ_within and Pp/Ppk use σ_overall.

For the **between/within** path, Cp/Cpk use σ_BW (which captures both
within-subgroup and between-subgroup variation) and Pp/Ppk still use σ_overall.
This is exactly the Minitab B/W convention.

### 2.2 Capability indices

$$
C_p = \frac{USL - LSL}{6\sigma_{within}}, \quad
C_{pk} = \min\left(\frac{USL - \mu}{3\sigma_{within}}, \frac{\mu - LSL}{3\sigma_{within}}\right)
$$

$$
P_p = \frac{USL - LSL}{6\sigma_{overall}}, \quad
P_{pk} = \min\left(\frac{USL - \mu}{3\sigma_{overall}}, \frac{\mu - LSL}{3\sigma_{overall}}\right)
$$

One-sided indices `Cpl`, `Cpu`, `Ppl`, `Ppu` are the two terms inside the
`min()`.

### 2.3 Taguchi indices (require a target T)

$$
C_{pm} = \frac{USL - LSL}{6\tau}, \quad \tau = \sqrt{\frac{1}{n}\sum (x_i - T)^2}
$$

$$
C_{pmk} = \min\left(\frac{USL - \mu}{3\tau}, \frac{\mu - LSL}{3\tau}\right)
$$

`Cpm` penalises off-target processes even when Cpk looks fine. Provide a
`target` to `capability_normal()` to enable Cpm/Cpmk.

### 2.4 Z-based indices

$$
Z_{LSL} = \frac{\mu - LSL}{\sigma}, \quad
Z_{USL} = \frac{USL - \mu}{\sigma}, \quad
Z_{Bench} = \Phi^{-1}(1 - P_{total})
$$

where $P_{total} = \Phi(-Z_{LSL}) + (1 - \Phi(Z_{USL}))$ is the total
non-conforming probability. Z.Bench is the "sigma quality level" used in
Six Sigma (delegated to `iQualityR.stat::z_bench`).

### 2.5 PPM (parts-per-million non-conforming)

PPM is computed from the fitted normal tail areas via
`iQualityR.stat::capability_to_ppm`. Two flavours are reported:

- **PPM (within)** — uses σ_within (short-term prediction).
- **PPM (overall)** — uses σ_overall (long-term prediction).

Each is split into below-LSL, above-USL, and total.

### 2.6 Acceptance thresholds (automotive)

| Index | Acceptable | Marginal | Unacceptable |
|---|---|---|---|
| Cpk / Ppk | ≥ 1.33 | 1.00 – 1.33 | < 1.00 |
| PPM (overall) | < 63 | 63 – 6210 | > 6210 |

These thresholds are applied automatically via
`iQualityR.stat::capability_interpret` and surfaced in the
`capability_judgment` diagnostic.

## 3. Data Preparation

The bundled `capability_normal_subgroup.csv` contains 125 diameter
measurements from a precision-machining line, organised as 25 subgroups of
size 5. The engineering specification is **LSL = 44 mm, USL = 56 mm**, target
**T = 50 mm**.

```{r load-data}
df <- read.csv(
  system.file("extdata", "capability_normal_subgroup.csv",
              package = "iQualityR.capa"),
  fileEncoding = "UTF-8"
)
str(df)
head(df)

# Subgroup structure
table(df$subgroup)
```

## 4. Normal Capability — Quick Start

```{r normal-quick}
task <- capability_normal(
  data = df, measurement = "measurement",
  lsl = 44, usl = 56, target = 50,
  subgroup = "subgroup"
)
task$summary()
```

## 5. Step-by-Step Workflow

For full control over the plan, build the `CapabilityPlan` explicitly:

```{r normal-step}
plan <- CapabilityPlan$new(
  lsl = 44, usl = 56, target = 50,
  subgroup = "subgroup",
  conf_level = 0.95,
  analysis_type = "normal"
)
plan$validate()

task2 <- IqrCapabilityTask$new(df, "measurement", plan, theme = "academic")
task2$compute()
task2$summary()
```

## 6. Inspecting the Results

### 6.1 Capability indices table

```{r normal-indices}
stats <- task$results$statistics
data.frame(
  Index = c("Cp", "Cpk", "Cpl", "Cpu",
            "Pp", "Ppk", "Ppl", "Ppu",
            "Cpm", "Cpmk",
            "Z.LSL", "Z.USL", "Z.Bench"),
  Value = c(stats$cp, stats$cpk, stats$cpl, stats$cpu,
            stats$pp, stats$ppk, stats$ppl, stats$ppu,
            stats$cpm, stats$cpmk,
            stats$z_lsl, stats$z_usl, stats$z_bench_within)
)
```

### 6.2 σ components

```{r normal-sigma}
data.frame(
  Component = c("Within", "Overall", "Mean", "n", "n_subgroups"),
  Value = c(stats$sd_within, stats$sd_overall,
            stats$mean, stats$n, stats$n_subgroups)
)
```

### 6.3 PPM (within vs overall)

```{r normal-ppm}
task$results$data_tables$ppm_within
task$results$data_tables$ppm_overall
```

The *within* PPM is the short-term prediction (what the process *could* do if
it stayed centred); the *overall* PPM is the long-term prediction (what it
*actually* does, including drift). A large gap between the two signals
instability — investigate the I-chart and MR-chart panels.

### 6.4 Normality diagnostics

```{r normal-diag}
diag <- task$results$diagnostics
data.frame(
  Check = c("Normality method", "p-value", "Verdict"),
  Value = c(diag$normality_method,
            fmt4(diag$normality_p_value),
            diag$capability_judgment$overall_verdict)
)
```

## 7. Visualization Gallery

| `type` | Description |
|---|---|
| `"basic"` | Histogram with density and spec limits. |
| `"qq"` | Normal Q-Q plot. |
| `"capbar"` | Capability indices bar chart (Cp/Cpk/Pp/Ppk). |
| `"individual"` | Individual values chart (I chart). |
| `"mr"` | Moving range chart. |
| `"trend"` | Cpk trend / last subgroups. |
| `"full"` | **Sixpack** — all six panels in a 3×2 grid (default). |

### 7.1 The Sixpack

```{r normal-plot-full, fig.width=12, fig.height=16}
print(task$plot(type = "full"))
```

**How to read the Sixpack:**

- **I chart (top-left)** — points should stay within the spec lines (red
  dashed). Out-of-spec points are immediate red flags.
- **MR chart (top-right)** — points should stay below the UCL (red dashed).
  This verifies short-term stability.
- **Histogram (middle-left)** — the density curve should be symmetric and
  bell-shaped. Spec lines mark where the tolerance band sits relative to the
  data.
- **Q-Q plot (middle-right)** — points should hug the reference line. Curvature
  signals non-normality; the subtitle reports the Shapiro-Wilk p-value.
- **Capability bar chart (bottom-left)** — green (≥1.33), amber (1.00–1.33),
  red (<1.00). The dashed line is the 1.33 reference.
- **Cpk trend (bottom-right)** — if subgrouped, shows subgroup means; a flat
  line indicates stability.

### 7.2 Individual panels

```{r normal-plot-panels, fig.width=8, fig.height=4}
print(task$plot(type = "basic"))
```

```{r normal-plot-qq}
print(task$plot(type = "qq"))
```

```{r normal-plot-capbar}
print(task$plot(type = "capbar"))
```

## 8. Between/Within Capability

The same dataset has 25 subgroups. If we suspect between-subgroup drift
(e.g. tool wear, setup changes between shifts), the plain `capability_normal`
under-states the short-term σ because it only sees within-subgroup
variation. The **between/within** path adds σ_between and uses
σ_between_within for Cp/Cpk.

```{r bw-quick}
task_bw <- capability_between_within(
  data = df, measurement = "measurement",
  lsl = 44, usl = 56, target = 50,
  subgroup = "subgroup"
)
task_bw$summary()
```

### 8.1 The four σ components

```{r bw-sigma}
stats_bw <- task_bw$results$statistics
data.frame(
  Component = c("Within", "Between", "Between-Within", "Overall"),
  Sigma = c(stats_bw$sd_within, stats_bw$sd_between,
            stats_bw$sd_between_within, stats_bw$sd_overall)
)
```

- **σ_within** — variation *inside* subgroups (the "noise floor").
- **σ_between** — variation *between* subgroup means (drift / setup effects).
- **σ_between_within** — the B/W combination used for Cp/Cpk in this path.
- **σ_overall** — the total sample SD, used for Pp/Ppk.

When σ_between is close to zero, the B/W path collapses to the plain normal
path. When σ_between is large relative to σ_within, the B/W Cp/Cpk will be
lower than the plain Cp/Cpk — a honest correction.

### 8.2 B/W Sixpack

```{r bw-plot-full, fig.width=12, fig.height=16}
print(task_bw$plot(type = "full"))
```

### 8.3 Normal vs Between/Within comparison

```{r compare-normal-bw}
comparison <- data.frame(
  Path = c("Normal (within σ)", "Between/Within (B/W σ)"),
  Cp  = c(stats$cp,  stats_bw$cp),
  Cpk = c(stats$cpk, stats_bw$cpk),
  Pp  = c(stats$pp,  stats_bw$pp),
  Ppk = c(stats$ppk, stats_bw$ppk)
)
comparison
```

**Interpretation:** If the B/W Cpk is materially lower than the normal Cpk,
the process has significant between-subgroup variation that the plain within σ
hides. Investigate the subgroup means on the trend panel and address the
upstream source of drift (tooling, setup SOP, operator training) rather than
reporting the inflated within-only Cpk.

## 9. Confidence Intervals via Bootstrap

For defensible uncertainty quantification on the indices, enable the
bootstrap. This is the JMP-style approach (analytical CIs à la Bissell are on
the roadmap).

```{r normal-bootstrap}
task_boot <- capability_normal(
  data = df, measurement = "measurement",
  lsl = 44, usl = 56, target = 50, subgroup = "subgroup",
  use_bootstrap = TRUE, bootstrap_samples = 500
)
task_boot$results$data_tables$bootstrap_ci
```

The bootstrap CI table reports the lower and upper bounds for each index at
the configured `conf_level` (default 0.95). If the lower bound of Cpk is
below 1.33, the process is *not* statistically proven capable even if the
point estimate looks fine.

## 10. Box-Cox / Johnson Transformation Path

When the normality diagnostic fails (Shapiro-Wilk p < 0.05) but the data is
the kind that *can* be normalised (positive, monotone), you can request a
transform. The analyzer fits the transform, transforms both the data **and**
the spec limits with the same parameters, runs the normal analysis on the
transformed scale, and annotates every panel so you know the indices are
reported on the transformed scale.

```{r normal-transform, eval = FALSE}
# Example: a right-skewed dimension that needs Box-Cox
task_bc <- capability_normal(
  data = df_skewed, measurement = "measurement",
  lsl = 5, usl = 80,
  transform = "box_cox"   # or "johnson", or "auto"
)
```

See `vignette("capa-nonnormal")` for a full worked example — the
transformation path is documented there because it competes with the
non-normal parametric path.

## 11. Report Export

```{r normal-report}
xlsx_path <- tempfile(fileext = ".xlsx")
task$report(format = "excel", path = xlsx_path)

data.frame(
  Format = "excel",
  Path = basename(xlsx_path),
  Exists = file.exists(xlsx_path),
  Size_KB = round(file.size(xlsx_path) / 1024, 1)
)
```

The Excel report mirrors the Sixpack layout: Overview, Indices, Judgment,
PPM (within & overall), Diagnostics, Warnings, Advice, Bootstrap CI, and
Raw Data sheets.

## 12. Common Pitfalls

1. **No subgroup column** — `capability_normal()` without `subgroup` uses MSSD
   for σ_within (individuals logic). This is correct for I-MR data but you
   lose the within/overall distinction if the data was actually subgrouped.
2. **Ignoring normality** — always check the Q-Q panel and the Shapiro-Wilk
   p-value. If p < 0.05, switch to the non-normal or non-parametric path.
3. **Reporting Cpk without Cpm** — a process can have Cpk = 1.5 but be
   off-target, hurting the customer. Always report Cpm when a target exists.
4. **Confusing within and overall PPM** — within PPM is the *potential*;
   overall PPM is the *reality*. A 10× gap means instability.
5. **B/W when subgroups are not rational** — σ_between only makes sense if
   subgroups represent logical production units. Random subgroups inflate
   σ_between artificially.

## 13. References

1. Montgomery, D. C. (2013). *Introduction to Statistical Quality Control*
   (7th ed.). Wiley. — Cp/Cpk, Pp/Ppk, σ decomposition.
2. AIAG (2005). *Statistical Process Control (SPC) Reference Manual* (2nd
   ed.). — Within vs. overall σ, rational subgroups.
3. Minitab LLC. *Between/Within Capability documentation*, v21.
4. `iQualityR.stat::sigma_decomposition` — σ estimator backend.
5. `iQualityR.stat::capability_interpret` — verdict thresholds.
