---
title: "iQualityR.capa Overview"
subtitle: "Process Capability Analysis for the iQualityR Ecosystem"
author: "iQualityR capa Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Capability Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7, fig.height = 4.5, fig.align = "center",
  warning = FALSE, message = FALSE
)

options(iqr.default_theme = "academic")
library(iQualityR.capa)

fmt4 <- function(x) ifelse(is.na(x), "NA", formatC(as.numeric(x), format = "f", digits = 4))
fmt2 <- function(x) ifelse(is.na(x), "NA", formatC(as.numeric(x), format = "f", digits = 2))
```

## 1. What This Package Does

`iQualityR.capa` quantifies how well a manufacturing or service process meets
its engineering specifications. Given a measured quality characteristic and a
pair of specification limits (LSL, USL), it produces the three deliverables
that every industrial capability study requires:

1. **Capability indices** — Cp, Cpk, Cpl, Cpu, Pp, Ppk, Ppl, Ppu, Cpm, Cpmk,
   and the Z-based family (Z.Bench, Z.LSL, Z.USL). These compare the spread
   and centering of the process to the width of the tolerance band.
2. **Expected PPM** — parts-per-million non-conforming, decomposed into
   *within-subgroup* and *overall* predictions, and further split into
   below-LSL / above-USL / total.
3. **Diagnostics** — normality tests, distribution-fit goodness-of-fit
   (Kolmogorov-Smirnov), and stability checks that tell the analyst whether
   the chosen analysis path is defensible.

The package follows the standard **Minitab** output layout (Stat > Quality
Tools > Capability Analysis) and borrows the **JMP** "fit-all-distributions"
idea for the non-normal path, so analysts trained on either commercial tool
will recognise the workflow immediately — while running entirely in R, so
capability studies can be embedded in reproducible reports, automated
pipelines, and Shiny applications.

## 2. The Four Analysis Paths

`iQualityR.capa` implements four core analysis paths for continuous, univariate
CTQs, plus two advanced paths (attribute and multivariate) documented in
`vignette("capa-advanced")`.

| Path | When to use | σ decomposition | Dedicated vignette |
|---|---|---|---|
| **Normal** | Data is well-modelled by a normal distribution. | within (R-bar/d2 or MSSD) vs overall (sample SD). | `vignette("capa-normal")` |
| **Between/Within** | Subgrouped data with between-subgroup drift (e.g. setup changes, lot-to-lot). | within + between + between-within + overall. | `vignette("capa-normal")` |
| **Non-normal** | Skewed or bounded data; a parametric non-normal distribution fits better. | quantile method (0.135% / 99.865%); no within/overall split. | `vignette("capa-nonnormal")` |
| **Non-parametric** | No defensible parametric model; let the data speak. | empirical / kernel CDF; observed PPM. | `vignette("capa-nonparametric")` |
| **Binomial** *(advanced)* | Pass/fail attribute data; per-batch defectives. | p-bar; PPM = p × 1e6; Z.Bench. | `vignette("capa-advanced")` |
| **Poisson** *(advanced)* | Defect counts per unit (DPMO). | u-bar; DPMO = u × 1e6; Z.Bench. | `vignette("capa-advanced")` |
| **Multivariate** *(advanced)* | Correlated CTQs; joint spec region. | MCPV (Taam) + HPCI (Shahriari) + Hotelling T². | `vignette("capa-advanced")` |

### 2.1 Decision tree

```
                 +---------------------------+
                 | Is the data approximately |
                 |      normal?              |
                 +---------------------------+
                      |               |
                    Yes              No
                      v               v
            +-----------------+   +-------------------------+
            | capability_     |   | Can a parametric       |
            |    normal()     |   | non-normal dist fit?   |
            +-----------------+   +-------------------------+
                                        |               |
                                      Yes              No
                                        v               v
                              +-----------------+  +----------------------+
                              | capability_     |  | capability_          |
                              |   nonnormal()   |  |   nonparametric()    |
                              +-----------------+  +----------------------+
                                        |
                                        v
                              (KS p < 0.05 for all?)
                                        | Yes
                                        v
                              +----------------------+
                              | capability_          |
                              |   nonparametric()    |
                              +----------------------+
```

**Practical guidance:**

1. **Start with `capability_normal()`.** The diagnostics panel tells you if the
   normality assumption is violated (Shapiro-Wilk p < 0.05 or visible curvature
   in the Q-Q plot).
2. **If subgroups show drift** (setup changes, operator-to-operator
   differences), use `capability_between_within()` — the standard Minitab
   B/W path that decomposes σ into within / between / B-W / overall.
3. **If normality fails**, check whether the data has a natural bound (cycle
   time, surface roughness, strength). If yes, try
   `capability_nonnormal(distribution = "auto")` — the fitter compares
   Weibull / Lognormal / Gamma / Exponential / Logistic by AIC and verifies
   goodness-of-fit with the KS test.
4. **If no parametric distribution fits** (KS p < 0.05 for all candidates),
   fall back to `capability_nonparametric()`, which makes no distributional
   assumption and reports the observed exceedance rate.

## 3. Package Architecture

`iQualityR.capa` follows the **Task-Plan-Analyzer-Plotter-Reporter (TPAPR)**
architecture inherited from `iQualityR.core`. Every analysis is orchestrated
by a *Task* object that owns the data, a *Plan* object that encodes the
specification limits and analysis options, an *Analyzer* that performs the
statistical computation, a *Plotter* that renders the diagnostic charts, and
a *Reporter* that assembles the final deliverable. The five roles are bound
together by a shared *Theme* so that all output is visually consistent.

```
            +--------------------------+
            |          Theme           |   IqrTheme  (iQualityR.core)
            |  colors, fonts, scales   |
            +--------------------------+
                  |    |    |    |
                  v    v    v    v
   +----------+  +----------+  +-----------+  +----------+  +-----------+
   |   Plan   |  |   Task   |  | Analyzer  |  | Plotter  |  | Reporter  |
   |----------|  |----------|  |-----------|  |----------|  |-----------|
   | Capabil  |->| IqrCapab |->| Capabil   |->| Capabil  |->| Iqr       |
   | ityPlan  |  | ilityTas |  | ityAnaly  |  | ityPlott |  | Reporter  |
   |          |  |    k     |  |    zer    |  |    er    |  |           |
   +----------+  +----------+  +-----------+  +----------+  +-----------+
   specs,         data +       indices,       histogram,     Excel, HTML,
   analysis_type  results      PPM, diag.      Sixpack, QQ    report
```

All five layers **delegate** to the shared infrastructure in `iQualityR.core`,
`iQualityR.stat`, and `iQualityR.plot` so that capability results are
numerically and visually consistent with the rest of the ecosystem:

| capa component | Delegates to |
|---|---|
| `CapabilityAnalyzer` σ estimation | `iQualityR.stat::sigma_decomposition` / `sigma_estimate` |
| `CapabilityAnalyzer` PPM | `iQualityR.stat::capability_to_ppm` |
| `CapabilityAnalyzer` normality | `iQualityR.stat::normality_test` |
| `CapabilityAnalyzer` capability verdict | `iQualityR.stat::capability_interpret` |
| `CapabilityAnalyzer` Z.Bench | `iQualityR.stat::z_bench` |
| `CapabilityAnalyzer` transforms | `iQualityR.stat::box_cox_transform` / `johnson_transform` / `auto_transform` |
| `DistributionFitter` distribution fitting | `iQualityR.stat::fit_distribution` |
| `CapabilityPlotter` all panels | `iQualityR.plot::base_plot` + `layers_*` |
| `IqrCapabilityTask` reporting | `iQualityR.core::IqrReporter` |

## 4. Function and Class Index

### One-shot convenience wrappers

```r
# Normal capability (with optional subgroup σ decomposition)
capability_normal(data, measurement, lsl, usl, target = NULL,
                  subgroup = NULL, transform = NULL,
                  conf_level = 0.95, theme = "academic", ...)

# Between/Within capability (requires subgroup)
capability_between_within(data, measurement, lsl, usl, target = NULL,
                          subgroup, conf_level = 0.95,
                          use_bootstrap = FALSE, bootstrap_samples = 1000,
                          theme = "academic", ...)

# Non-normal capability (auto-selects distribution by AIC)
capability_nonnormal(data, measurement, lsl, usl, distribution = "auto",
                    target = NULL, conf_level = 0.95,
                    use_bootstrap = FALSE, bootstrap_samples = 1000,
                    theme = "academic", ...)

# Non-parametric capability (kernel or empirical CDF)
capability_nonparametric(data, measurement, lsl, usl, target = NULL,
                         conf_level = 0.95, use_bootstrap = FALSE,
                         bootstrap_samples = 1000,
                         nonparametric_method = "kernel",
                         theme = "academic", ...)
```

All four return an `IqrCapabilityTask` object (invisibly) with the standard
TPAPR lifecycle: `$compute()` → `$summary()` → `$plot()` → `$report()`.

### R6 classes

| Class | Inherits | Role |
|---|---|---|
| `CapabilityPlan` | `IqrPlanBase` | Stores spec limits, analysis type, criteria. |
| `CapabilityAnalyzer` | `IqrAnalyzerBase` | Computes indices, PPM, diagnostics. |
| `CapabilityPlotter` | `IqrPlotterBase` | Renders histogram, Q-Q, Sixpack, etc. |
| `IqrCapabilityTask` | `IqrTaskBase` | Coordinates compute / plot / report. |
| `DistributionFitter` | (standalone) | Wraps `iQualityR.stat::fit_distribution`. |

## 5. The Five-Step Workflow

Every capability study follows the same five-step lifecycle:

```{r workflow, eval = FALSE}
library(iQualityR.capa)

# 1. Prepare data (data.frame with a measurement column)
df <- read.csv(
  system.file("extdata", "capability_normal_subgroup.csv", package = "iQualityR.capa")
)

# 2. One-shot analysis
task <- capability_normal(
  data = df, measurement = "measurement",
  lsl = 44, usl = 56, target = 50,
  subgroup = "subgroup"
)

# 3. Inspect results
task$summary()

# 4. Visualize
task$plot(type = "full")   # Sixpack

# 5. Export report
task$report(format = "excel", path = tempfile(fileext = ".xlsx"))
```

## 6. Capability Indices Computed

| Index | Formula (normal) | Meaning |
|---|---|---|
| `Cp` | (USL−LSL) / (6σ_within) | Potential capability, ignores centering. |
| `Cpk` | min((USL−μ),(μ−LSL)) / (3σ_within) | Actual short-term capability. |
| `Cpl` / `Cpu` | (μ−LSL)/(3σ_within) , (USL−μ)/(3σ_within) | One-sided lower / upper indices. |
| `Pp` / `Ppk` | as above with σ_overall | Long-term performance. |
| `Ppl` / `Ppu` | one-sided, σ_overall | Long-term one-sided. |
| `Cpm` | (USL−LSL) / (6·τ), τ=√(mean((x−T)²)) | Taguchi index (target deviation). |
| `Cpmk` | min((USL−μ),(μ−LSL)) / (3·τ) | Centered Taguchi index. |
| `Z.Bench` | Φ⁻¹(1 − P_total) | Sigma quality level (short-term). |
| `Z.LSL` / `Z.USL` | (μ−LSL)/σ , (USL−μ)/σ | One-sided sigma levels. |

For the **non-normal** path the indices use the quantile method
(σ_eq = (Q₀.₉₉₈₆₅ − Q₀.₀₀₁₃₅)/6); for the **non-parametric** path they use
empirical quantiles. See the dedicated vignettes for the exact formulas.

### Acceptance thresholds (automotive convention)

| Index | Acceptable | Marginal | Unacceptable |
|---|---|---|---|
| Cpk / Ppk | ≥ 1.33 | 1.00 – 1.33 | < 1.00 |
| PPM (overall) | < 63 | 63 – 6210 | > 6210 |

These thresholds are applied automatically via
`iQualityR.stat::capability_interpret` and surfaced in the
`capability_judgment` diagnostic.

## 7. The Sixpack Diagnostic

The Sixpack is the Minitab-standard six-panel dashboard that verifies the
**assumptions** behind a capability number before you trust it. Each panel
answers one question:

| Panel | Question it answers |
|---|---|
| Individual values chart (I) | Is the process stable over time? Are points outside spec? |
| Moving range chart (MR) | Is the short-term variation stable? |
| Histogram + density | Does the data shape match the assumed distribution? |
| Q-Q plot | Is the distributional assumption quantitatively defensible? |
| Capability indices bar chart | Are Cp/Cpk/Pp/Ppk above the 1.33 threshold? |
| Cpk trend / last subgroups | Has capability drifted over time or across subgroups? |

```{r sixpack-demo, fig.width=9, fig.height=7, echo = FALSE}
df_normal <- read.csv(
  system.file("extdata", "capability_normal_subgroup.csv", package = "iQualityR.capa"),
  fileEncoding = "UTF-8"
)
task <- capability_normal(
  data = df_normal, measurement = "measurement",
  lsl = 44, usl = 56, target = 50, subgroup = "subgroup"
)
print(task$plot(type = "full"))
```

## 8. Bundled Sample Data

`iQualityR.capa` ships with two datasets in `inst/extdata/`:

| File | Shape | Use case |
|---|---|---|
| `capability_normal_subgroup.csv` | 125 rows × 2 cols (`subgroup`, `measurement`) | Normal & Between/Within capability, 25 subgroups of size 5. |
| `capability_nonnormal_weibull.csv` | 125 rows × 1 col (`measurement`) | Non-normal & non-parametric capability, right-skewed. |

```{r load-data}
df_normal <- read.csv(
  system.file("extdata", "capability_normal_subgroup.csv",
              package = "iQualityR.capa"),
  fileEncoding = "UTF-8"
)
head(df_normal)

df_nonnormal <- read.csv(
  system.file("extdata", "capability_nonnormal_weibull.csv",
              package = "iQualityR.capa"),
  fileEncoding = "UTF-8"
)
head(df_nonnormal)
```

## 9. Comparison with MINITAB and JMP

`iQualityR.capa` is designed to cover the union of the capability features
that industrial analysts actually use in MINITAB v21 and JMP v19:

| Feature | MINITAB | JMP | iQualityR.capa |
|---|---|---|---|
| Normal capability (Cp…Cpmk, Z.Bench) | ✓ | ✓ | ✓ |
| Between/Within capability | ✓ (dedicated menu) | ✓ | ✓ |
| Non-normal capability (quantile method) | ✓ (9 dist) | ✓ (15+) | ✓ (auto 5 + explicit `beta`) |
| Non-parametric capability | ✓ | ✓ | ✓ |
| Box-Cox / Johnson transformation path | ✓ | ✓ | ✓ |
| Bootstrap CI for indices | — | ✓ | ✓ |
| Sixpack diagnostic | ✓ (3 variants) | — | ✓ (normal; B/W & NP variants share layout) |
| **Attribute capability (Binomial)** | ✓ | ✓ | ✓ (Wilson + Clopper-Pearson CI, dispersion test) |
| **Attribute capability (Poisson)** | ✓ | ✓ | ✓ (Garwood exact CI, dispersion test) |
| **Multivariate capability indices (MCPV)** | — | — | **✓ (Taam 1993)** |
| **Multivariate capability indices (HPCI)** | — | — | **✓ (Shahriari 1995, 3-component vector)** |
| **Joint PPM via `mvtnorm::pmvnorm`** | — | — | **✓** |
| Multiple-variable capability | ✓ | ✓ (Goal Plot) | ✓ (Goal Plot + Capability Snapshot, see `capa-advanced` §4.10) |

The **multivariate capability indices** (MCPV / HPCI / joint PPM) are gaps
that **even Minitab v21 and JMP v19 do not fully cover** — they are
implemented here with worked quality examples in
`vignette("capa-advanced")`.

## 10. Where to Go Next

- For normal & between/within capability: `vignette("capa-normal")`
- For non-normal capability & transformations: `vignette("capa-nonnormal")`
- For non-parametric capability & multi-variable considerations:
  `vignette("capa-nonparametric")`
- For **attribute capability (Binomial / Poisson)** and **multivariate
  capability (MCPV / HPCI)** — the features that go beyond Minitab and JMP:
  `vignette("capa-advanced")`

## 11. References

1. Montgomery, D. C. (2013). *Introduction to Statistical Quality Control*
   (7th ed.). Wiley. — Cp/Cpk, Pp/Ppk, Sixpack layout.
2. AIAG (2005). *Statistical Process Control (SPC) Reference Manual* (2nd
   ed.). — Within vs. overall σ, PPM.
3. Kotz, S., & Lovelace, N. L. (1998). *Process Capability Indices in Theory
   and Practice*. Arnold. — Non-normal capability, quantile method.
4. Minitab LLC. *Capability Analysis documentation*, v21.
5. JMP. *Process Capability platform documentation*, v19.
6. `iQualityR.core` package documentation — TPAPR architecture, `IqrTheme`.
7. `iQualityR.stat` package documentation — `sigma_estimate`,
   `fit_distribution`, `capability_to_ppm`, `z_bench`.
